### 2023-11-06 tex/state-learner-pub-version.tex
new repo

\documentclass[a4,danish]{article}
\usepackage[a4paper, margin=3.5cm]{geometry}
\input{./latex-settings/standard-settings.tex}
\input{./latex-settings/standard-commands.tex}
\input{./latex-settings/theorem-env.tex}
\input{./latex-settings/title-compact.tex}

\fxsetup{status=draft}
\usepackage[colorlinks=true,citecolor=blue]{hyperref} % Doesn't work for org, so included here

%% Specific libraries
\usepackage{booktabs}

%% Specific commands
\newcommand{\data}{\ensuremath{\mathcal{D}}}

\begin{document}

% \articletype{ARTICLE TEMPLATE}% Specify the article type or omit as appropriate

\title{The state learner \\ {-- a super learner for right-censored data}}

\author{Anders Munch and Thomas A. Gerds
  % \name{Anders Munch\textsuperscript{a}\thanks{CONTACT: Anders Munch. Email: a.munch@sund.ku.dk}
  %   and Thomas A. Gerds\textsuperscript{a}}
  % \affil{\textsuperscript{a} Section of Biostatistics, Department of Public Health, University of Copenhagen}
}

\maketitle

\begin{abstract}
  In survival analysis, prediction models are needed as stand-alone tools and in
  applications of causal inference to estimate nuisance parameters. The super
  learner is a machine learning algorithm which combines a library of prediction
  models into a meta learner based on cross-validated loss. Unfortunately, the
  commonly used partial likelihood loss is not suited for super learning, and
  inverse probability of censoring weighted loss functions require a
  pre-specified estimator of the censoring distribution. To relax this, we
  introduce the state learner, a new super learner for survival analysis, which
  evaluates the loss based on the observed data simultaneously using libraries
  of predictions models for the event(s) of interest and the censoring
  distribution. We establish an oracle inequality for the state learner and
  investigate its performance through numerical experiments. We illustrate how
  the state learner allows us to estimate causal effects in a competing risks
  setting without having to pre-specify models for neither the cause-specific
  hazards of interest nor the censoring distribution.
  % In survival analysis, prediction models are needed as stand-alone tools and in
  % applications of causal inference to estimate nuisance parameters. The super
  % learner is a machine learning algorithm which combines a library of prediction
  % models into a meta learner based on cross-validated loss. Unfortunately, the
  % commonly used partial likelihood loss is not suited for super learning, and
  % inverse probability of censoring weighted loss functions require a
  % pre-specified estimator of the censoring distribution. To relax this, we
  % introduce the state learner, a new super learner for survival analysis, which
  % evaluates the loss based on the observed data simultaneously for two libraries
  % of prediction models, one for the event time distribution and one for the
  % censoring distribution. We establish an oracle inequality for the state
  % learner and investigate its performance through numerical experiments.
\end{abstract}

% \begin{keywords}
%   Loss based estimation; right-censored data; competing risks; super learner;
%   cross-validation
% \end{keywords}

% \begin{amscode}
%   % Use 2010 Mathematics Subject Classification keywords
%   62N02
% \end{amscode}

\section{Introduction}
\label{sec:introduction}

A super learner is a machine learning algorithm that combines a finite
set of learners into a meta learner by estimating prediction
performance in hold-out samples using a pre-specified loss function
\citep{van2007super}. When the aim is to make a prediction model,
super learners typically combine strong learners, such as Cox
regression models and random survival forests
\citep{gerds2021medical}.
% For targeted learning of a low dimensional
% target parameter in presence of a high-dimensional nuisance parameter,
% a super learner will usually also include the highly adaptive lasso
% \citep{van2011targeted, benkeser2016highly,van2017generally}.
While the general idea of combining strong learners based on cross-validation
data stems from earlier work \citep{wolpert1992stacked,breiman1996stacked}, the
name super learner is justified by an oracle inequality
\citep{van2003unicv,vaart2006oracle}: The super learner is guaranteed to perform
almost as well as the model which minimizes the expected performance, i.e., the
model we would select if we could evaluate the prediction performance in an
infinite hold-out sample.

We are concerned with the choice of the loss function for super learning in
survival analysis. Existing super learner algorithms for right-censored data use
partial log-likelihood loss or inverse probability of censoring weighted loss
\citep{polley2011-sl-cens,keles2004asymptotically,golmakani2020super,westling2021inference}.
The use of the partial log-likelihood loss restricts the class of learners and
excludes for example simple Kaplan-Meier based learners and also more complex
random survival forest algorithms. For this reason \cite{golmakani2020super}
restrict their learners to Cox proportional hazard models. A lesser known fact
is that a super learner constructed with the negative partial log-likelihood
loss implicitly depends on the censoring distribution
\citep{hjort1992inference,whitney2019comment}. A disadvantage of inverse
probability of censoring weighted loss functions is that they requires a
pre-specified model for the censoring distribution. \cite{westling2021inference}
tackle this challenge by iterating between super learning of the censoring
distribution and the event time distribution.

In this article we define the state learner, a new super learner for
right-censored data, which simultaneously evaluates the loss for learners of the
event time distribution and the censoring distribution. The loss function which
is used to define the state learner is only based on observable
quantities. The state learner can be applied to all types of survival
estimators, works in the presence of competing risks, and does not require a
single pre-specified estimator of the conditional censoring distribution. To
analyze the theoretical properties of the state learner we focus on the
so-called discrete super learner which `combines' the library of
learners by picking the one that minimizes the cross-validated loss. The state
learner uses separate libraries to model each competing event and the censoring
distribution. We show that the oracle selector of the state learner is
consistent if all libraries contain a consistent learner and prove a finite
sample oracle inequality.

The state learner can be used to select a model which predicts the probability
of an event based on covariates in the presence of competing risks. Another
application is in targeted learning where conditional event probabilities occur
as high-dimensional nuisance parameters which need to be estimated at a certain
rate \citep{van2011targeted, rytgaard2021estimation, rytgaard2022targeted}. We
show how a targeted estimator can be obtained from the state learner, and that a
second order product structure for the asymptotic bias term of the targeted
estimator is retained when the state learner is used to estimate nuisance
parameters.

% For machine learning with right-censored data, the negative partial
% log-likelihood is a popular loss function
% \citep[e.g.,][]{li2016regularized,
%   yao2017deep,lee2018deephit,katzman2018deepsurv,
%   gensheimer2019scalable,lee2021boosted, golmakani2020super}.
% However, there are two major problems when attempting to use the partial log-likelihood for super
% learning. The first is that the partial
% log-likelihood cannot be used to evaluate cross-validated loss for
% survival learners which are not absolutely continuous in the time
% argument \citep{golmakani2020super}. The second problem is that the
% expected value of the partial log-likelihood depends on the censoring
% distribution. Both problems are further discussed in Section
% \ref{sec:relat-liter-exist}.

% Inverse probability of censoring weighted (IPCW) estimators are often
% used to resolve the dependence of the expected loss on the censoring
% distribution \citep{graf1999assessment,
%   molinaro2004tree,keles2004asymptotically,hothorn2006survival,gerds2006consistent,gonzalez2021stacked}.
% However, commonly these methods require a consistent estimator of the
% conditional censoring distribution given covariates. An obvious idea
% would be to estimate the conditional censoring distribution by a super
% learner. However, that would require a loss function to evaluate the
% performance of candidate learners of the censoring distribution and
% the expected value of that loss function would in turn depend on the
% event time distribution. To resolve this dilemma
% \cite{westling2021inference} and \cite{han2021inverse} suggested to
% iterate between super learning of the censoring distribution and the
% event time distribution, where both super learners use the respective
% cross-validated IPCW loss. Our approach is different. The state
% learner uses all combinations of learners for the event time
% distribution with a learner for the censoring distribution to
% parameterize a set of `observable' cumulative distribution
% functions. Specifically, the cross-validated integrated Brier score
% for each of the cumulative distribution functions then only depends on
% the observed data distribution and hence can be estimated empirically
% without further modeling.

The article is organized as follows. We introduce our notation and framework in
Section~\ref{sec:framework}. In Section~\ref{sec:super-learning} we define super
learning in general with right-censored data, and in
Section~\ref{sec:super-learner-simple} we introduce the state learner.
Section~\ref{sec:theor-results-prop} provides theoretical guarantees for the
state learner. In Section~\ref{sec:targeted-learning} we discuss the use of the
state learner in the context of targeted learning. We report a numerical study
in Section~\ref{sec:numer-exper}, and analyze a prostate cancer data set in
Section~\ref{sec:real-data-appl}. Finally, we relate the state learner to
existing approaches in Section~\ref{sec:discussion} and discuss some limitations
of our proposal. Appendices~\ref{sec:proof-proposition}
and~\ref{sec:state-learner-with} contain proofs.

\section{Notation and framework}
\label{sec:framework}

In a competing risk framework \citep{andersen2012statistical}, let \( T\) be a
time to event variable, \(D\in\{1,2\}\) the cause of the event, and
$X \in \mathcal{X}$ a vector of baseline covariates taking values in a bounded
subset \( \mathcal{X} \subset \R^p \), \( p\in\N \). Let $\tau< \infty$ be the
maximal length of follow-up. We use \( \mathcal{Q} \) to denote the collection
of all probability measures on \( [0,\tau] \times \{1,2\}\times \mathcal{X} \)
such that \( (T, D, X) \sim Q \) for some unknown \( Q \in \mathcal{Q} \). For
\(j\in\{1,2\}\), the cause-specific conditional cumulative hazard functions are
defined by
\( \Lambda_{j} \colon [0, \tau] \times \mathcal{X} \rightarrow \R_+ \) such that
\begin{equation*}
  % \label{eq:cum-haz}
  \Lambda_{j}(t \mid x) = \int_0^t\frac{  Q(T \in \diff s, D=j \mid X=x )}{Q(T \geq s \mid X=x )}.
\end{equation*}
For ease of notation we assume throughout that \( \Lambda_j(\blank \mid x) \) is
continuous for all \( x \) and \( j \). We denote by \(S\) the conditional
event-free survival function,
\begin{equation}
  \label{eq:surv-def}
  S(t \mid x)=\exp\left\{-\Lambda_{1}(t \mid x)-\Lambda_{2}(t \mid x)\right\}.
\end{equation}
Let \( \mathcal{M} \) denote the space of all conditional cumulative hazard
functions on \( [0,\tau] \times\mathcal{X}\). Any distribution
\( Q \in \mathcal{Q} \) can be characterized by
\begin{equation*}
  \label{eq:parametrizeQ}
  \begin{split}
    Q(\diff t,j,\diff x)&= \left\{S(t- \mid x)\Lambda_1(\diff t \mid x)H(\diff x)\right\}^{1_{\{j=1\}}}\\
                        &\quad\left\{S(t- \mid x)\Lambda_2(\diff t \mid x)H(\diff x)\right\}^{1_{\{j=2\}}},
  \end{split}
\end{equation*}
where \(\Lambda_{j} \in \mathcal{M}\) for \(j=1,2\) and \(H\) is the marginal
distribution of the covariates.

We consider the usual right-censored setting in which we observe data
\(O = (\tilde{T},\tilde D, X)\), where $\tilde T = \min(T,C)$ for a
right-censoring time \(C\), $\Delta = \1{\{T \leq C\}}$, and
\(\tilde D=\Delta D\). Let \(\mathcal{P}\) denote a set of probability measures
on the sample space
\(\mathcal{O} = [0, \tau] \times \{0, 1, 2\} \times \mathcal{X}\) such that
\(O \sim P \) for some unknown \(P\in \mathcal{P}\). We assume that the event
time and the censoring time are conditionally independent given covariates,
\( T \independent C \mid X \). This implies that any distribution
\( P \in \mathcal{P} \) is characterized by a distribution
\( Q \in \mathcal{Q} \) and a conditional cumulative hazard function for \( C \)
given \( X \) \citep[c.f.,][]{begun1983information,gill1997coarsening}. We use
\(\Gamma\in\mathcal M\) to denote the conditional cumulative hazard function for
censoring. We assume that \( \Gamma(\blank \mid x) \) is continuous for all
\( x \), and let \(G(t \mid x)=\exp\left\{-\Gamma(t \mid x)\right\}\) denote the
survival function of the conditional censoring distribution. In our setting with
competing risks, this yields
\begin{equation}\label{eq:parametrizeP}
  \begin{split}
    P(\diff t, j, \diff x) &= \left\{G(t- \mid x)S(t- \mid x)\Lambda_1(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=1\}}}}\\
                           &\quad\left\{G(t- \mid x)S(t- \mid x)\Lambda_2(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=2\}}}}\\
                           &\quad\left\{G(t- \mid x)S(t- \mid x)\Gamma(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=0\}}}}\\
                           &=\left\{G(t- \mid x)Q(\diff t,j,\diff x)\right\}^{\1{{\{j\ne 0\}}}}\\    
                           &\quad\left\{G(t- \mid x)S(t- \mid x)\Gamma(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=0\}}}}.
  \end{split}
\end{equation}
Hence, we may write
\( \mathcal{P} = \{ P_{Q, \Gamma} : Q \in \mathcal{Q}, \Gamma \in
\mathcal{G} \} \) for some \( \mathcal{G} \subset \mathcal{M} \). We
also have
\begin{equation*}
P(\tilde T>t \mid X=x) = S(t \mid x)G(t \mid x) = \exp\left\{-\Lambda_{1}(t \mid x)-\Lambda_{2}(t \mid x)-\Gamma(t \mid x) \right\}.
\end{equation*}
We further assume that there exists \(\kappa<\infty\) such that
\(\Lambda_{j}(\tau- \mid x)<\kappa \), for \(j\in\{1,2\}\), and
\(\Gamma(\tau- \mid x)<\kappa\) for almost all \(x\in\mathcal X\). Note that this
implies that \(G(\tau- \mid x)\) is bounded away from zero for almost all \(x\in\mathcal X\).
Under these assumptions, the conditional cumulative hazard functions
\(\Lambda_{j}\) and \(\Gamma\) can be identified from \(P\) by
\begin{align}
  \Lambda_{j}(t \mid x) &= \int_0^t\frac{  P(\tilde T \in \diff s, \tilde D=j \mid X=x )}{P(\tilde T \geq s \mid X=x )}, \label{eq:lambdaj}\\
  \Gamma(t \mid x) &= \int_0^t\frac{  P(\tilde T \in \diff s, \tilde D=0 \mid X=x )}{P(\tilde T \geq s \mid X=x )}\label{eq:gamma}.
\end{align}
Thus, we can consider $\Lambda_j$ and \(\Gamma\) as operators which map from
\( \mathcal{P} \) to \(\mathcal M\).
% and we shall sometimes write
% \( \Lambda^{P} \) to emphasize that \( \Gamma^{P} \) is associated with a
% particular data-generating distribution \( P \).

\section{Super learning with right-censored survival data}
\label{sec:super-learning}

A super learner estimates a parameter $\Psi$ which can be identified from the
observed data distribution \(P\in\mathcal P\). In this section, to introduce the
concept of super learning, we simply consider estimation of the function
\( \Lambda_{j}\). The parameter \(\Psi:\mathcal P\to\mathcal M\) is then
identified via equation \eqref{eq:lambdaj} by \(\Psi(P)=\Lambda_j\).

As input to the super learner we need a sample \( \data_n=\{O_i\}_{i=1}^n \) of
i.i.d.\ observations from some unknown \( P \in \mathcal{P} \) and a finite
collection of candidate learners $\mathcal{A}$. Each learner
\(a \in \mathcal{A}\) is a map
\( a \colon \mathcal{O}^n \rightarrow \mathcal{M}\) which takes a data set as
input and returns an estimate $a(\data_n) \in \mathcal{M}$ of $\Lambda_{j}$.
% Formally, the domain of $a$ depends on \(n\) but we suppress this in the notation.
In what follows, we use the short-hand notation
\(P[f] = \int f(o) P(\diff o) \). A super learner evaluates the
performance of \(a \in \mathcal{A}\) using a loss function
\(L\colon \mathcal{M} \times \mathcal{O} \rightarrow \R_+\) by
estimating the expected loss \(P[L(a(\data_n), \blank)]\) using
cross-validation. Specifically, the expected loss of $a\in\mathcal A$
is estimated by splitting the data set $\data_n$ into $K$ disjoint
approximately equally sized subsets
\(\data_n^1, \data_n^2, \dots, \data_n^K \) and then calculating the
cross-validated loss
\begin{equation*}
  % \label{eq:cv-risk-est}
  \hat{R}_n(a; L) =
  \frac{1}{K}\sum_{k=1}^{K}
  % \empmeas^k{[L {(a{ (\data_n^{-k})} , \blank) }]},
  \frac{1}{| \data_n^{k} |}\sum_{O_i \in \data_n^{k}}
  L
  {
    \left(
      a{ (\data_n^{-k})}
      , O_i
    \right)
  },
  \quad \text{with} \quad
  \data_n^{-k} = \data_n \setminus \data_n^{k}.
\end{equation*}
% where \( \empmeas^{k} \) is the empirical measure of \( \data_n^{k} \). 
The subset \(\data_n^{-k}\) is referred to as the \(k\)'th training
sample, while \(\data_n^{k}\) is referred to as the \(k\)'th test or
hold-out sample.
The discrete super learner is defined as
\begin{equation*}
\hat{a}_n = \argmin_{a\in\mathcal A}\hat{R}_n(a; L).
\end{equation*}
The final estimator of \(\Psi(P)=\Lambda_j\) is then the selected
learner applied to the full data set, i.e., \(\hat{a}_n(\data_n)\).
The oracle learner is defined as the learner that minimizes the
average loss according to the data-generating distribution \( P \),
i.e.,
\begin{equation*}
  \tilde{a}_n =
  \argmin_{a \in \mathcal{A}}
  \tilde{R}_n(a; L),
  \quad \text{with} \quad 
  \tilde{R}_n(a; L)=
  \frac{1}{K}\sum_{k=1}^{K} 
  P{
    \left[
      L
      {
        \left(
          a{ (\data_n^{-k})}
          , \blank
        \right)
      }
    \right]}
  .
\end{equation*}
Note that both the discrete super learner and the oracle learner depend on the
library of learners and on the number of folds \(K\), and that the oracle
learner is a function of the data and the unknown data-generating distribution.
These dependencies are suppressed in the notation.

\section{The state learner}
\label{sec:super-learner-simple}

The problem with most existing super learners for right-censored data is that
they depend on a pre-specified estimator of the censoring distribution. The main
idea of the state learner is to jointly use learners of \( \Lambda_1 \),
\( \Lambda_2 \), and \( \Gamma \), and the relations in
equation~(\ref{eq:parametrizeP}), to learn a feature of the observed data
distribution \( P \). The discrete state learner ranks a tuple of learners of
\( (\Lambda_1, \Lambda_2, \Gamma) \) based on how well they jointly model the
observed data. To formally introduce the state learner, we define the
multi-state process
\begin{equation*}
  \eta(t) = \1\{\tilde{T} \leq t, \tilde D=1\} + 2\,\1\{\tilde{T} \leq t, \tilde D=2\}+ 3\,\1\{\tilde{T} \leq t, \tilde D=0\},
  \quad \text{for} \quad t \in [0, \tau].
\end{equation*}
Note that at time \(t\), we observe that each observation is in one of four
mutually exclusive states (Figure \ref{fig:multi-state-process}).
\begin{figure}[h]
  \centering
  \includegraphics[width=.5\textwidth]{./figures/figure-multi-state-process.pdf}
  \caption{Illustration of the multi-state process \(\eta\) used by
    the state learner. Note that `censored' is a state, hence the
    process is always observed at any time.}
  \label{fig:multi-state-process}
\end{figure}
The conditional distribution of \( \eta(t) \) given \( X=x \) is
determined by the function
\begin{equation}
  \label{eq:F-def}
  F(t, k, x) = P(\eta(t) = k \mid X=x),
  \quad \text{for all} \quad
  t \in [0,\tau],
  k \in \{0,1,2,3\},
  x \in \mathcal{X}.
\end{equation}
The function \( F \) describes the conditional state occupation
probabilities corresponding to the observed multi-state process
\(\eta\).

We propose to construct a super learner for \( F \), i.e., the target of this
super learner is $\Psi(P) = F$ where the parameter is identified through
equation~(\ref{eq:F-def}). Because each quadruple
$(\Lambda_{1}, \Lambda_{2}, \Gamma, H)$ characterizes a \(P\in\mathcal P\) which
in turn determines \( (F, H) \), a learner for \( F \) can be constructed from
learners of \( \Lambda_1 \), \( \Lambda_2 \), and $\Gamma$ as follows:
\begin{equation}\label{eq:transition}
  \begin{split}
  F(t, 1, x)
  & = P(\tilde{T} \leq t, \Delta=1 \mid X=x)
    = \int_0^t e^{\{-\Lambda_{1}(s \mid x)-\Lambda_{2}(s \mid x) - \Gamma(s \mid x)\} }  \Lambda_{1}(\diff s \mid x),
  \\
  F(t, 2, x)
  & = P(\tilde{T} \leq t, \Delta=2 \mid X=x)
    = \int_0^t e^{\{-\Lambda_{1}(s \mid x)-\Lambda_{2}(s \mid x) - \Gamma(s \mid x)\} }  \Lambda_{2}(\diff s \mid x),
  \\
  F(t, 3, x)
  & =
    P(\tilde{T} \leq t, \Delta=0 \mid X=x)
    = \int_0^t e^{\{-\Lambda_{1}(s \mid x)-\Lambda_{2}(s \mid x) - \Gamma(s \mid x)\} }  \Gamma(\diff s \mid x),
  \\
  F(t, 0, x)
  &
    = P(\tilde{T} > t \mid X= x)
    = 1- F(t, 1, x) - F(t, 2, x)- F(t, 3, x).
  \end{split}
\end{equation}
The state learner requires three libraries of learners, \(\mathcal{A}_1\),
\( \mathcal{A}_2 \), and \( \mathcal{B} \), where \(\mathcal{A}_1\) and
\( \mathcal{A}_2\) contain learners of the conditional cause-specific cumulative
hazard functions of the event time distribution \(\Lambda_1\) and
\( \Lambda_2\), respectively, and \(\mathcal{B}\) contains learners of the
conditional cumulative hazard function of the censoring
distribution. % We further
% define \(\mathcal{H}\) as the set of all probability distributions on
% \( \mathcal{X} \) and
% \begin{equation*}
%   \mathcal{H}_n=\left\{h\colon\mathcal{O}^n\longrightarrow\mathcal{H} \midd
%     h(\data_n)=H_n = \frac 1 n \sum_{i=1}^n \delta_{X_i} \right\}
% \end{equation*}
% as the library of learners which consists of a single learner, the empirical
% distribution function.
Based on the Cartesian product of
libraries of learners for \((\Lambda_1,\Lambda_2,\Gamma)\) we construct a library
$\mathcal{F}(\mathcal{A}_1, \mathcal{A}_2, \mathcal{B})$ of learners
for \( F \):
\begin{align*}
  \mathcal{F}(\mathcal{A}_1, \mathcal{A}_2, \mathcal{B})
  &= \{ \phi_{a_1,a_2, b} : a_1 \in \mathcal{A}_1, a_2 \in \mathcal{A}_2, b \in \mathcal{B}\},
    \intertext{where in correspondance with  the relations in equation \eqref{eq:transition},} 
    \phi_{a_1,a_2, b}(\data_n)(t,1,x) &= \int_0^t e^{\{-a_1(\data_n)(s \mid x)-a_2(\data_n)(s \mid x) - b(\data_n)(s \mid x)\} }  a_1(\data_n)(\diff s \mid x),\\
  \phi_{a_1,a_2, b}(\data_n)(t,2,x) &= \int_0^t e^{\{-a_1(\data_n)(s \mid x)-a_2(\data_n)(s \mid x) - b(\data_n)(s \mid x)\} }  a_2(\data_n)(\diff s \mid x),\\
  \phi_{a_1,a_2, b}(\data_n)(t,3,x) &= \int_0^t e^{\{-a_1(\data_n)(s \mid x)-a_2(\data_n)(s \mid x) - b(\data_n)(s \mid x)\} }  b(\data_n)(\diff s \mid x),\\
  \phi_{a_1,a_2, b}(\data_n)(t,0,x) &= 1-  \sum_{j=1}^3\phi_{a_1,a_2, b}(\data_n)(t,j,x).
\end{align*}
To evaluate how well a function \( F \) predicts the observed
multi-state process we use the integrated Brier score
\( \bar B_\tau( F,O) = \int_0^{\tau} B_t(F,O) \diff t \), where \( B_t \) is the
Brier score \citep{brier1950verification} at time \( t \in [0, \tau] \),
\begin{equation*}
  B_t(F,O) = \sum_{j=0}^{3}
  \left(
      F(t,j,X) - \1{\{\eta(t)=j\}}
  \right)^2.
\end{equation*}
As described in Section~\ref{sec:super-learning}, each learner
\( \phi_{a_1, a_2, b} \) in the library
\( \mathcal{F}(\mathcal{A}_1, \mathcal{A}_2, \mathcal{B}) \) is evaluated using
the cross-validated loss,
\begin{equation*}
  \hat{R}_{n}(\phi_{a_1,a_2,b} ; \bar{B}_{\tau}) =
  \frac{1}{K}\sum_{k=1}^{K}
  \frac{1}{| \data_n^{k} |}\sum_{O_i \in \data_n^{k}}
  \bar B_\tau
  {
    \left(
      \phi_{a_1,a_2,b}{ (\data_n^{-k})}
      , O_i
    \right)
  },
\end{equation*}
and the discrete state learner is
\begin{align*}\label{eq:discrete-state-learner}
  \hat{\phi}_n
  &=  \argmin_{(a_1,a_2,b)\in \mathcal{A}_1\times\mathcal{A}_2\times\mathcal{B}}
    \hat{R}_{n}(\phi_{a_1,a_2,b} ; \bar{B}_{\tau}).
\end{align*}
% Note that depending on the target parameter of the learning task, the
% state learner may or may not depend on the distribution of the
% covariates \(H_n\). For example, for the parameter
% \(\tilde\Psi(\Lambda_{1}, \Lambda_{2}, \Gamma, H)=\Lambda_{1}\) the
% state learner would not depend on \(H_n\) but for the parameter
% \(\tilde\Psi(\Lambda_{1}, \Lambda_{2}, \Gamma, H)=\int S(t \mid x)H(\diff
% x)\) it would.

\section{Theoretical results for the state learner}
\label{sec:theor-results-prop}

In this section we establish theoretical guarantees for the state learner.
% We let \( \E_P \) denote
% expectation under \( P \) and use \( \Theta \) to denote the
% collection of all conditional state-occupation probability functions.
Proposition~\ref{prop:stric-prop} can be derived from the fact that the
integrated Brier score (also called the continuous ranked probability score) is
a strictly proper scoring rule \citep{gneiting2007strictly}. This implies that
if we minimize the average loss of the integrated Brier score, we recover the
parameters of the data-generating distribution. In particular, the oracle of a
state learner will be consistent if the library of learners contains at least
one learner that is consistent for estimation of \( F \). Recall that the
function \(F\) implicitly depends on the data generating probability measure
\(P\in\mathcal P\) but that this was suppressed in the notation. We now make
this dependence explicit by writing \(F_0\) for the function which is obtained
by substituting a specific \(P_0\in\mathcal{P}\) for \(P\) in equation
\eqref{eq:transition}.

\begin{proposition}
  \label{prop:stric-prop}
  For \(P_0\in\mathcal{P}\) define
  \begin{equation*}
    F^* = \argmin_{F} P_0{[\bar{B}_\tau(F, \blank)]},
  % F^* = \argmin_{(\Lambda_1,\Lambda_2,\Gamma) \in \mathcal{M}\otimes\mathcal{M}\otimes\mathcal{M}} P_0{[\bar B_\tau(\phi_{(\Lambda_1,\Lambda_2,\Gamma)}, \blank)]},
\end{equation*}
where the minimum is taken over all \( F \), such that \( F \) is a conditional
state occupation probability function for some measure \( P \) as defined in equation~(\ref{eq:F-def}).
Then \( F^*(t,j,\blank) = F_0(t,j,\blank) \) \( H \)-almost surely for
any \( j\in \{0,1,2,3\} \) and almost any \( t \in [0, \tau]\).
\end{proposition}
\begin{proof}
  See Appendix~\ref{sec:proof-proposition}.
\end{proof}

We establish a finite sample oracle result for the state learner.
Our Corollary~\ref{cor:oracle-prop} is in essence a special case of a
general cross-validation result by \cite{vaart2006oracle}. We assume
that we split the data into equally sized folds, and for simplicity of
presentation we take \( n \) to be such that
\( |\data_n^{-k}| = n/K \) with \( K \) fixed. We will allow the
number of learners to grow with \( n \) and write
\( \mathcal{F}_n=\mathcal{F}(\mathcal{A}_{1,n}, \mathcal{A}_{2,n},
\mathcal{B}_n)\) as short-hand notation and to emphasize the
dependence on \( n \).
% The discrete super $\hat{\phi}_n$ and the
% oracle $\tilde{\phi}_n$ are defined as in Section~\ref{sec:framework}
% but now with respect to the library \( \mathcal{F}_n \) and the loss
% \( B \).
In the following we let \( \| \blank \|_{P_0} \) denote the norm
\begin{equation}
  \label{eq:norm}
  \| F \|_{P_0} = 
  \left\{
    \sum_{j=0}^{3}\int_{\mathcal{X}} \int_0^{\tau} F(t, j, x)^2 \diff t H_0( \diff x)
  \right\}^{1/2}.
\end{equation}
\begin{corollary}
  \label{cor:oracle-prop}
  For all \(P_0\in\mathcal{P}\), \( n \in \N \), \( k \in \{1, \dots, K\} \),
  and $\delta>0$,
  \begin{align*}
    \E_{P_0}{\left[ \Vert \hat{\phi}_n(\data_n^{-k}) - F_0 \Vert_{P_0}^2 \right]}
    & \leq (1 + 2\delta)
      \E_{P_0}{\left[ \Vert \tilde{\phi}_n(\data_n^{-k}) - F_0 \Vert_{P_0}^2 \right]}
    \\
    & \quad
      + (1+ \delta) 16   K \tau
      \left(
      13 + \frac{12}{\delta}
      \right)
      \frac{\log(1 + |\mathcal{F}_n|)}{n}.
  \end{align*}
\end{corollary}
\begin{proof}
  See Appendix~\ref{sec:proof-proposition}.
\end{proof}

Corollary~\ref{cor:oracle-prop} has the following asymptotic consequences.

\begin{corollary}
  \label{cor:asymp-cons}
  Assume that \( |\mathcal{F}_n| = \bigO(n^q)\), for some \( q \in \N \)
  and that there exists a sequence \( \phi_n \in \mathcal{F}_n \),
  \( n \in \N \), such that
  \( \E_{P_0}{\left[ \Vert \phi_n(\data_n^{-k}) - F_{0} \Vert_{P_0}^2 \right]} =
  \bigO(n^{-\alpha}) \), for some \( \alpha\leq 1 \).
  \begin{enumerate}[label=(\roman*), topsep=0pt]
  \item If $\alpha=1$ then
    \( \E_{P_0}{\left[ \Vert \hat{\phi}_n(\data_n^{-k}) - F_0 \Vert_{P_0}^2
      \right]} = \bigO(\log(n)n^{-1}) \).
  \item If $\alpha<1$ then
    \( \E_{P_0}{\left[ \Vert \hat{\phi}_n(\data_n^{-k}) - F_0 \Vert_{P_0}^2 \right]} =
    \bigO(n^{-\alpha}) \).
  \end{enumerate}
\end{corollary}
\begin{proof}
  See Appendix~\ref{sec:proof-proposition}.
\end{proof}

In Section~\ref{sec:targeted-learning} we demonstrate how the state learner can
be used for targeted learning. A targeted estimator is typically obtained from
estimators of the nuisance parameters $\Lambda_1$, $\Lambda_2$, and $\Gamma$. By
equations~(\ref{eq:lambdaj}) and~(\ref{eq:gamma}) and the definition of \( F \),
we have
\begin{equation}
  \label{eq:7}
  \Gamma(t , x) 
  = \int_0^t  \frac{F(\diff s, 3, x )}{F(s-, 0, x )},
  \quad \text{and} \quad
  \Lambda_j(t , x) 
  = \int_0^t  \frac{F(\diff s, j, x )}{F(s-, 0, x )},
  \quad j \in \{1,2\},
\end{equation}
and thus a targeted estimator can also be obtained from an estimator of $F$
using equation~(\ref{eq:7}). The key feature of a targeted estimator is that it
is asymptotically equivalent to a sum of i.i.d.\ random variables plus a second
order remainder term \citep{bickel1993efficient,fisher2021visually}. For our
setting of competing risks, the remainder term is typically dominated by terms
of the form
\begin{equation}
  \label{eq:dr-term}
  P{\left[
      \int_0^{\tau} w_n(s, \blank)
      \hat{M}_{1,n}(s \mid  \blank)
      \hat{M}_{2,n}(\diff s \mid  \blank)
    \right]},
\end{equation}
where \( (\hat{M}_{1,n}, \hat{M}_{2,n}) \) is any of the nine combinations of
\( \hat{M}_{1,n} \in \{[\Gamma -\hat{\Gamma}_n], [\Lambda_1
-\hat{\Lambda}_{1,n}], [\Lambda_2 -\hat{\Lambda}_{2,n}]\} \) and
\( \hat{M}_{2,n} \in \{[\Gamma -\hat{\Gamma}_n], [\Lambda_1
-\hat{\Lambda}_{1,n}], [\Lambda_2 -\hat{\Lambda}_{2,n}]\} \), and \( w_n \) is
some data-dependent function with domain \([0,\tau]\times\mathcal X \)
\citep{van2003unified}. In particular, a targeted estimator will be
asymptotically linear if the `products' of the estimation errors
\( \hat{M}_{1,n} \) and \( \hat{M}_{2,n} \) in equation~(\ref{eq:dr-term}) are
\( \smallO_P{(n^{-1/2})}\). Proposition~\ref{prop:dr-structure} states that if
equation~(\ref{eq:dr-term}) holds for the a targeted estimator based on
estimator $\hat{\Lambda}_{1,n}$, $\hat{\Lambda}_{2,n}$, and $\hat{\Gamma}_{n}$,
then a similar product structure holds for a targeted estimator based on
\( \hat{F}_n \). We state the result for the special case that
\(\hat{M}_{1,n}= \Gamma-\hat{\Gamma}_n \) and
\(\hat{M}_{2,n} =\Lambda_1-\hat{\Lambda}_{1,n} \), but similar results holds for
any combinations of \( \Gamma-\hat{\Gamma}_n\),
\( \Lambda_1-\hat{\Lambda}_{1,n} \), and \( \Lambda_2-\hat{\Lambda}_{2,n} \).
\begin{proposition}
  \label{prop:dr-structure}
  Assume that \( w(s,x)\leq c \), \( F(s, 0, x) \geq 1/c \) and
  \( \hat{F}_n(s, 0, x) \geq 1/c \) for some \( c>0 \) for all
  \( s \in [0, \tau] \) and \( x \in \mathcal{X} \). Then there are real-valued
  uniformly bounded functions \( w^a_n \), \( w^b_n \), \( w^c_n \), and
  \( w^d_n \) with domain \( [0,\tau]^2 \times \mathcal{X} \) such that
  \begin{align*}
    & P_0{\left[
      \int_0^{\tau} w(s, \blank)
      \left\{
      \Gamma_0(s,\blank) -\hat{\Gamma}_n(s,\blank)
      \right\}
      [\Lambda_0-\hat{\Lambda}_n]
      (\diff s, \blank)
      \right]}
    \\
    & =
      P_0{\left[
      \int_0^{\tau} \int_0^{s} w^a_n(s,u,\blank) [F_0 - \hat{F}_n](u-, 0, \blank)[F_0 - \hat{F}_n](s-, 0, \blank) F_0(\diff u, 2, \blank ) F_0 ( \diff s, 1, \blank)
      \right]}
    \\
    & \quad +
      P_0{\left[
      \int_0^{\tau} \int_0^{s} w^b_n(s,u,\blank) [F_0 - \hat{F}_n](u-, 0, \blank)
      F_0(\diff u, 2, \blank ) [F_0 - \hat{F}_n](\diff s, 1, \blank)
      \right]}
    \\
    & \quad +
      P_0{\left[
      \int_0^{\tau} \int_0^{s} w^c_n(s,u,\blank) [F_0 - \hat{F}_n](\diff u, 2, \blank)
      [F_0 - \hat{F}_n](s-, 0, \blank)
      F_0(\diff s, 1, \blank ) 
      \right]}
    \\
    & \quad +
      P_0{\left[
      \int_0^{\tau} \int_0^{s} w^d_n(s,u,\blank) [F_0 - \hat{F}_n](\diff u, 2, \blank)
      [F_0 - \hat{F}_n](\diff s, 1, \blank)
      \right]}.
  \end{align*}
\end{proposition}
\begin{proof}
  See Appendix~\ref{sec:state-learner-with}.
\end{proof}

\section{Targeted learning}
\label{sec:targeted-learning}

Features of the observed data distribution \( P \in \mathcal{P} \) are rarely of
interest. We are instead interested in a parameter
\( \theta \colon \mathcal{Q} \rightarrow \Theta \) that expresses a property of
the uncensored population governed by the measure \( Q \in \mathcal{Q} \). The
parameter space $\Theta$ can be a subset of \(\R^d\) or a subset of a function
space, such as \(\mathcal{M}\). In subsection~\ref{sec:cause-spec-aver} we give
a concrete example from causal inference where $\theta$ is the average treatment
effect and \( \Theta = [-1,1] \). Under the assumption of conditional
independent censoring and positivity, $\theta$ is identifiable from
\( \mathcal{P} \) which means that there exists an operator
\( \Psi \colon \mathcal{P} \rightarrow \Theta \) such that
\( \theta(Q) = \Psi(P_{Q, \Gamma}) \) for all $\Gamma \in \mathcal{M}$. By
equation~(\ref{eq:parametrizeP}) this imply that we may write
\begin{equation*}
  \theta(Q) = \Psi(P) = \tilde{\Psi}^0(\Lambda_1, \Lambda_2, H),
\end{equation*}
for some operator \( \tilde{\Psi}^0 \). The state learner provides a ranking of
tuples
\( (a_1, a_2, b) \in \mathcal{A}_1 \times \mathcal{A}_2 \times \mathcal{B} \),
and we use \( \hat{a}_{1,n} \), \( \hat{a}_{2,n} \), and \( \hat{b}_n \) to
denote the learners corresponding to the discrete state learner
\( \hat{\phi}_n \). Letting \( H(\data_n) \) denote the empirical measure of
\( \{X_1, \dots, X_n\} \), we can obtain a simple plug-in an estimator of
$\theta$ as
\begin{equation}
  \label{eq:2}
  \hat{\Psi}^0(\data_n) =
  \tilde{\Psi}^0(\hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n), H(\data_n)). 
\end{equation}

The asymptotic distribution of \( \hat{\Psi}_{\tau, j}^0 \) is difficult to
analyze due to the model selection step involved when estimating the nuisance
parameters $\Lambda_1$ and $\Lambda_2$. In addition, the estimator will
typically have an asymptotic bias that vanishes at a sub-optimal rate. Using
tools from semi-parametric efficiency theory, it is possible to construct a
so-called targeted or debiased estimator with smaller asymptotic bias and an
asymptotic distribution which we know how to estimate
\citep{bickel1993efficient,van2011targeted,chernozhukov2018double}. A targeted
estimator is based on the efficient influence function for the parameter
$\tilde{\Psi}^0$ and relies on an estimator of $\Gamma$ in addition to
estimators of \( \Lambda_1 \) and $\Lambda_2$. The efficient influence function
is a \( P \)-zero mean and square integrable function indexed by the nuisance
parameters \( (\Lambda_1, \Lambda_2, \Gamma) \), which we denote by
\( \psi(\blank ; \Lambda_1, \Lambda_2, \Gamma) \). The name is justified because
any regular asymptotically linear estimator that has \( \psi \) as its influence
function is efficient, meaning that it has smallest asymptotic variance among
all regular asymptotically linear estimators.

An example of a targeted estimator is the \fxnote*{also TMLE, depending on
  implementation}{one-step estimator}, defined as
\begin{equation}
    \label{eq:one-step-def}
    \hat{\Psi}_{\text{OS}}(\data_n)
    =
    \tilde{\Psi}^0(\hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
    H(\data_n))
    + \empmeas{[\psi(\blank; \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
      \hat{b}_n(\data_n) )]},
\end{equation}
where \( \empmeas \) is the empirical measure of a sample \(\{O_i\}_{i=1}^n\).
We can make the following asymptotic expansion of the one-step estimator
\citep{pfanzagl1985contributions,fisher2021visually,kennedy2022semiparametric},
\begin{equation*}
  \hat{\Psi}_{\text{OS}}(\data_n)- \Psi(P)
  =  \empmeas{[\psi(\blank ; \Lambda_1, \Lambda_2, \Gamma)]}
  +\mathrm{Rem}{(\hat{\Lambda}_{1,n},\hat{\Lambda}_{2,n},  \hat{\Gamma}_n, P)} + \smallO_{P}(n^{-1/2}),
\end{equation*}
where the remainder term is typically such that
\citep{van2003unified,rytgaard2022continuous}
\begin{equation}
  \label{eq:4}
  \mathrm{Rem}{(\hat{\Lambda}_{1,n},\hat{\Lambda}_{2,n},  \hat{\Gamma}_n, P)}
  = \mathcal{O}_P{
    \left\{
      (\Lambda_1-\hat{\Lambda}_{1,n})^2
      +
      (\Lambda_2-\hat{\Lambda}_{2,n})^2
      +
      (\Gamma-\hat{\Gamma}_{n})^2
    \right\}
  }.
\end{equation}
Hence, under a suitable Donsker class regularity condition
\citep{bickel1993efficient,kennedy2016semiparametric}, when
equation~(\ref{eq:4}) holds and the nuisance parameters $\Lambda_1$,
$\Lambda_2$, and $\Gamma$ are consistently estimated at rate
\( \smallO_P{(n^{-1/4})} \) then
\begin{equation}
  \label{eq:3}
  \sqrt{n}(\hat{\Psi}_{\text{OS}}(\data_n)- \Psi(P)) \weakly \mathcal{N}(0,
  P{[\psi(\blank; \Lambda_1, \Lambda_2, \Gamma)^2]}).
\end{equation}
In particular, equation~(\ref{eq:3}) and Slutsky's lemma imply that we can
obtain asymptotically valid \((1-\alpha)\cdot100\%\) confidence intervals by
calculating
\begin{equation*}
  \left[
    \hat{\Psi}_{\text{OS}}(\data_n) - q_{\alpha/2} \hat{\sigma}(\data_n) ,
    \;
    \hat{\Psi}_{\text{OS}}(\data_n) + q_{\alpha/2} \hat{\sigma}(\data_n)
  \right],
\end{equation*}
where \( q_{\alpha} \) is the \( (1-\alpha) \)-quantile or the standard normal
distribution, and
\begin{equation*}
  \hat{\sigma}(\data_n) = 
  \left(
    \empmeas{
      \left[
        \psi(\blank; \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
        \hat{b}_n(\data_n))^2
      \right]}
  \right)^{1/2}.
\end{equation*}

% When the nuisance parameters are consistently estimated at rate
% \( \smallO_P{(n^{-1/4})} \) and
% \begin{equation*}
%   P{\left(
%       \psi(\blank; \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
%       \hat{b}_n(\data_n))
%       \in \mathcal{R}
%     \right)} \longrightarrow 1,
%   \quad \text{as} \quad   n \longrightarrow \infty , 
% \end{equation*}
% for some Donsker class of functions \( \mathcal{R} \), it holds that
% \citep{bickel1993efficient,kennedy2022semiparametric}

% According to equation~(\ref{eq:parametrizeP}) and equation~(\ref{eq:7}), any
% \( P \) is characterized by both \((\Lambda_1, \Lambda_2, \Gamma, H)\) and
% \( (F, H) \), which means that there exist operators $\tilde\Psi$ and $\bar\Psi$
% such that
% \begin{equation*}
%   % \label{eq:state-learner-identity}
%   \tilde\Psi(\Lambda_{1,0},\Lambda_{2,0}, \Gamma_0, H_0)=\Psi(P_0)=\bar\Psi(F_0,H_0).
% \end{equation*}
% Hence, any parameter $\Psi$ defined on \( \mathcal{P} \) can be estimated using
% either estimators \( (\hat{\Lambda}_n, \hat{\Gamma}_n, \hat{H}_n) \) or
% \( (\hat{F}_n, \hat{H}_n) \). When we use the state learner the final estimator
% of the target parameter is $\bar\Psi(\hat\phi_n(\data_n), \hat{H}_n)$, where
% \( \hat{H}_n \) can often be taken to be the marginal empirical measure.

% For the rest of this section we consider the case where $\Theta = \R$, and
% $\Psi$ is estimable by a regular asymptotically linear estimator. Specifically,
% we assume that it is possible to construct an estimator
% \( \tilde\Psi(\hat\Lambda_n, \hat\Gamma_n, H_n) \) such that the the following
% expansion holds:
% \begin{equation}
%   \label{eq:ral}
%   \tilde\Psi(\hat\Lambda_n, \hat\Gamma_n, H_n) - \Psi(P) = \empmeas{[\psi_P]} +
%   \mathrm{Rem}{(\hat{\Lambda}_{n},  \hat{\Gamma}_n, P)} + \smallO_{P}(n^{-1/2}),
% \end{equation}
% where \( \empmeas \) is the empirical measure of a sample \(\{O_i\}_{i=1}^n\),
% $\psi_P$ a zero-mean function with \( P[\psi_P^2]< \infty \), and
% \( \mathrm{Rem}{(\hat{\Lambda}_{n}, \hat{\Gamma}_n, P)} \) a second order
% remainder term \citep{bickel1993efficient,fisher2021visually}.

% ...

% The \( \smallO_P{(n^{-1/2})} \) bound on the products of estimation errors is
% typically a weaker condition than requiring \( \Gamma \), $\Lambda_1$, and
% $\Lambda_2$ to be estimated independently at rate \( n^{-1/2} \). This is
% particularly important when data-adaptive estimation methods are used as these
% methods rarely provide \( n^{-1/2} \)-rate convergence for function-valued
% nuisance parameters.

\subsection{Cause-specific average treatment effect}
\label{sec:cause-spec-aver}

We now demonstrate how the state learner and the general estimation strategy
outlined above can be used to estimate a concrete parameter of interest in the
competing risk setting. We assume that the covariate vector \( X \in \R^p \) can
be separated into a binary treatment indicator \( A \) and a vector of potential
confounders, \( W \in \mathcal{W}\subset \R^{p-1} \). We abuse notation slightly
by writing \( \Lambda_j(t \mid x) = \Lambda_j(t \mid a, w) \) and
\( S(t \mid x) = S(t \mid a, w) \) when \( x=(a, w) \). We use $\mu$ to denote
the marginal distribution of \( W \) and $\pi$ to denote the conditional
probability of treatment,
\begin{equation*}
  \pi(w) = P(A=1 \mid W=w).
\end{equation*}
We assume throughout that $\pi$ is uniformly bounded away from \( 0 \) and
\( 1 \) on \( \mathcal{W} \). As both \( A \) and \( W \) are fully observed for
all individuals we can use a super learner to estimate $\pi$
\fxnote*{add}{[ref]}, and we denote this estimator by $\hat{\pi}_n$. We use the
empirical measure of \( \{W_1, \dots, W_n\} \) to estimate $\mu$, and denote
this estimator by the $\hat{\mu}_n$. As parameter of interest we consider the
\fxnote*{what is this called?}{standardized difference in absolute risk of cause
  \( 1 \) before time $\tau$},
\begin{equation*}
  \theta_{{\tau}}(Q) = \int_{\mathcal{W}} 
  \left\{
    \int_0^{\tau}
    S(s- \mid w, 1)  \Lambda_1(\diff s \mid w, 1)
    -
    \int_0^{\tau}
    S(s- \mid w, 0)  \Lambda_1(\diff s \mid w, 0)
  \right\}
  \mu(\diff w).
\end{equation*}
Under a set of
additional assumptions, \( \theta_{{\tau}} \) can be given the causal
interpretation
\begin{equation*}
  \theta_{{\tau}}(Q) =
  \E{[T^{A=1} \leq \tau, D^{A=1}=1]}-
  \E{[T^{A=0} \leq \tau, D^{A=0}=1]},
\end{equation*}
where \( (T^A, D^A) \) denotes potential outcomes \citep{hernanRobinsWhatIf}. In
this case, the interpretation of $\theta_{\tau}$ is the difference in the
average risk of cause \( 1 \) occurring before time \( \tau \) in the population
if everyone had been given treatment (\( A=1 \)) compared to if no one had been
given treatment.

Using equation~(\ref{eq:surv-def}), we may write
\( \theta_{{\tau}}(Q) = \tilde{\Psi}_{t}^0(\Lambda_1, \Lambda_2, \mu) \),
where
\begin{equation}
  \label{eq:1}    
  \begin{split}
  \tilde{\Psi}_{t}^0(\Lambda_1, \Lambda_2, \mu) & =
  \int_{\mathcal{W}} 
  \int_0^{\tau}
  e^{-\Lambda_1(s- \mid w, 1)-\Lambda_2(s- \mid w, 1)}  \Lambda_1(\diff s \mid
  w, 1)
  \mu(\diff w)
  \\
  &  \quad
  -\int_{\mathcal{W}} 
  \int_0^{\tau}
  e^{-\Lambda_1(s- \mid w, 0)-\Lambda_2(s- \mid w, 0)}  \Lambda_1(\diff s \mid w, 0)
  \mu(\diff w).
  \end{split}
\end{equation}
The efficient influence function for the parameter $\tilde{\Psi}_{\tau}$
depends on the set \( (\Lambda_1, \Lambda_2, \Gamma, \pi) \) of nuisance
parameters, and is given by \fxnote*{other refs?}{\citep{rytgaard2022targeted}}
\begin{equation}
  \label{eq:5}
  \begin{split}
    & \psi_{\tau}(O; \Lambda_1, \Lambda_2, \Gamma, \pi)
    \\
    & =
      \sum_{l=1}^{2} \int_0^{\tau}
      \left[
      h_{\tau,l,1}(t, O  ; \Lambda_1, \Lambda_2, \Gamma, \pi)
      - h_{\tau,l,0}(t, O  ; \Lambda_1, \Lambda_2, \Gamma, \pi)
      \right]
    \\
    & \quad\qquad  \qquad
      \times
      \left(
      N_l(\diff t) - \1{\{\tilde{T} \geq t\}}\Lambda_l(\diff t \mid A, W)
      \right)
    \\
    & \quad
      -\sum_{a=0}^{1} 
      \left\{
      (-1)^a
      \int_0^{\tau}
      e^{-\Lambda_1(s- \mid W, a)-\Lambda_2(s- \mid W, a)}  \Lambda_1(\diff s \mid
      W, a)
      \right\}
      - \tilde{\Psi}_{t}^0(\Lambda_1, \Lambda_2, \mu),
    % & \quad+ \int_0^{\tau} e^{-\Lambda_1(s- \mid W, 1)-\Lambda_2(s- \mid W,
    % 1)} \Lambda_1(\diff s \mid W, 1)
    % \\
    % & \quad - \int_0^{\tau} e^{-\Lambda_1(s- \mid W, 0)-\Lambda_2(s- \mid W,
    % 0)} \Lambda_1(\diff s \mid W, 0)
  \end{split}
\end{equation}
where
\begin{align*}
  h_{\tau,l,a}(t, O  ; \Lambda_1, \Lambda_2, \Gamma, \pi)
  =&  \frac{\1{\{A=a\}} e^{\Gamma(t-\mid A, W)}}{\pi(W)^{a}(1-\pi(W))^{1-a}}
  \\
   & \quad \times
     \left(
     1- \frac{\int_t^{\tau} e^{-\Lambda_1(s- \mid A, W)-\Lambda_2(s- \mid A, W)}
     \Lambda_1(\diff s \mid A, W)}
     {e^{-\Lambda_1(t \mid A, W)-\Lambda_2(t \mid A, W)}}
     \right)^{\1{\{l=1\}}}
  \\
   & \quad\times
     \left(
     - \frac{\int_t^{\tau} e^{-\Lambda_1(s- \mid A, W)-\Lambda_2(s- \mid A, W)}
     \Lambda_1(\diff s \mid A, W)}
     {e^{-\Lambda_1(t \mid A, W)-\Lambda_2(t \mid A, W)}}
     \right)^{\1{\{l=2\}}}
\end{align*}
Equation~(\ref{eq:1}) and~(\ref{eq:5}) then allow us to construct a one-step
estimator by using the definition given in equation~(\ref{eq:one-step-def}),
which gives the estimator
\begin{equation}
  \label{eq:one-step-comp-ate}
  \tilde{\Psi}_{t, j}^0(\hat{a}_1(\data_n), \hat{a}_2(\data_n), \hat{\mu}_n(\data_n))
\end{equation}

\section{Numerical experiments}
\label{sec:numer-exper}

We compare the state learner with two other discrete super learners and an
oracle selector in a simulation study without a competing event. The two other
super learners are based on inverse probability of censoring weighted (IPCW)
Brier scores \citep{graf1999assessment,gerds2006consistent}, and we refer to
these as IPCW super learners. These super learners depend on an estimator of the
censoring distribution, and we consider IPCW super learners that use either the
Kaplan-Meier estimator (IPCW(KM)) or a correctly specified Cox model (IPCW(Cox))
to estimate the censoring distribution. Both IPCW super learners are fitted
using the \texttt{R}-package \texttt{riskRegression}
\citep{Gerds_Ohlendorff_Ozenne_2023}. Each discrete super learner provides a
learner for the cumulative hazard function for the outcome of interest, and from
this a risk prediction model can be obtained. We measure performance of each
super learner in terms of the Brier score of the provided risk prediction model
at a specific time horizon. The Brier score is approximated using a large
(\( n = 20,000 \)) independent data set of uncensored data. The oracle selector
uses the large data set of uncensored event times to select the learner with the
lowest expected Brier score. The expected Brier score of the oracle selector
serves as a lower benchmark value. For all super learners we split the data into
five folds for training and testing.

Note that given a learner for the cumulative hazard function of the
outcome event, we can typically use the same method to construct a
learner of the cumulative hazard function of the censoring
distribution. This would typically work by training the learner on the
data set \( \data_n^c \), where \( \data_n^c = \{O_i^c\}_{i=1}^n \)
with \( O_i^c = (\tilde{T}_i, 1-\Delta, X_i) \). When we say that we
use a learner for the cumulative hazard function of the outcome to
learn the cumulative hazard function of the censoring time, we mean
that the learner is trained on \( \data_n^c \).

\paragraph{Scenario 1}

We first generate a simple dataset to demonstrate that an IPCW super learner can
perform poorly when the censoring model is misspecified. We start by generating a
binomial baseline covariate \( A \) with success probability \( 30\% \). We then
generate outcome and censoring variables according to a Cox-Weibull distribution
with hazard rates of approximately \( 0.5 \) and \( 2.5 \), respectively. We use
a library \( \mathcal{A} \) consisting of two learners, the (marginal)
Kaplan-Meier estimator and the Kaplan-Meier estimator stratified on the binary
covariate. For the state learner we use the same learners to construct the
library \( \mathcal{B} \).

The results are shown in Figure~\ref{fig:ipcw-fail}. We see that the IPCW super
learner based on a misspecified censoring model (IPCW(KM)) has a larger Brier
score than the other super learners and that its performance does not improve
much with sample size. % When the censoring model is misspecified, the
% IPCW(KM) super learner fails to identify the learner corresponding to the
% data-generating mechanism, even though this learner is included in the
% library.
The performance of the state learner is comparable to or slightly better than
the IPCW super learner based on a correctly pre-specified censoring model. The
performance of both the IPCW(Cox) super learner and the state learner are close
to that of the oracle for most sample sizes.

\begin{figure}
  \centerline{\includegraphics[width=1\linewidth]{./figures/ipcw-fail.pdf}}
  \caption[]{Results for scenario~1 of the simulation study. For the learner
    selected by each of the four discrete super learners, the Brier score
    calculated in a large independent data set without censoring is plotted
    against sample size. The results are based on 200 repetitions.}
  \label{fig:ipcw-fail}
\end{figure}

\paragraph{Scenario 2}

We next generate data according to a more complex distribution motivated from a
real dataset in which censoring depends on the baseline covariates. We simulate
data based on a prostate cancer study described in
\citep{kattan2000pretreatment}. The outcome of interest was the time to tumor
recurrence, and five baseline covariates were used to predict outcome:
prostate-specific antigen (PSA, ng/mL), Gleason score sum (GSS, values between 6
and 10), radiation dose (RD), hormone therapy (HT, yes/no) and clinical stage
(CS, six values). The study was designed such that a patient's radiation dose
depended on when the patient entered the study \citep{gerds2013estimating}. This
in turn implied that the time of censoring depended on the radiation dose.
% It was found that almost all covariates were predictive of censoring.
The data were re-analyzed in \citep{gerds2013estimating} where a sensitivity
analysis was conducted based on simulated data. We use the same simulation
setup, where event and censoring times are generated according to parametric
Cox-Weibull models estimated from the original data, and the covariates are
generated according to either marginal Gaussian normal or binomial distributions
estimated from the original data
\citep[c.f.,][Section~4.6]{gerds2013estimating}. We use the library consisting
of the nine learners described in Table~\ref{tab:zel-library}. For the state
learner we use the same library to learn the censoring distribution.

\begin{table}[h]
  \centering
  \begin{tabular}{ lll }
    \toprule
    Family & Model & Description \\
    \midrule
    Marginal & \texttt{KM} & The Kaplan-Meier estimator \\ 
    Cox & \texttt{Cox} & All five covariates included with additive effects \\
           & \( \texttt{Cox strata CS} \)  & Cox model stratified on CS  \\
           & \( \texttt{Cox strata HT} \)  & Cox model stratified on HT \\
           & \( \texttt{Cox spline} \)  & PSA and RD modeled with splines \\ 
    Penalized Cox & \texttt{Lasso} & Cox model with \( L_1 \)-norm penalty   \\
           & \texttt{Ridge} & Cox model with \( L_2 \)-norm penalty \\
           & \texttt{Elastic} & Cox model with \( L_1 \)- and \( L_2 \)-norm penalty \\
    Random forest & \texttt{RF} & Random forest with 50 trees and default settings \\
    \bottomrule 
  \end{tabular}
  \caption[]{Overview of the nine learners used in scenario~2 of the simulation
    study. The Kaplan-Meier estimator was fitted using the package
    \texttt{prodlim} \citep{Gerds_2019prodlim}. All Cox models included all five
    covariates in the model and were fitted using the package \texttt{survival}
    \citep{survival-package}. All penalized Cox models included all five
    covariates as linear predictors and were fitted using the package
    \texttt{glmnet} \citep{glmnet-cox,glmnet-glm}. The random forest was fitted
    with the package \texttt{randomForestSRC} \citep{rfsrc-paclage}.}
  \label{tab:zel-library}
\end{table}

The results are shown in Figure~\ref{fig:zelefski}. We see that the Brier score
of the IPCW super learner based on a misspecified censoring model (IPCW(KM))
decreased with sample size, but is larger than that of the other super learners
for all sample sizes. The state learner and the IPCW super learner based on a
correctly pre-specified censoring model demonstrate similar performance for all
sample sizes. The performance of both the IPCW(Cox) super learner and the state
learner approaches the benchmark provided by the oracle selector for large sample
sizes.
\begin{figure}
  \centerline{\includegraphics[width=1\linewidth]{./figures/zelefski-sim.pdf}}
  \caption[]{Results for scenario~2 of the simulation study. For the learner
    selected by each of the four discrete super learners, the Brier score
    calculated in a large independent data set without censoring is plotted
    against sample size. The results are based on 200 repetitions.}
  \label{fig:zelefski}
\end{figure}

\section{Real data application}
\label{sec:real-data-appl}

The original prostate cancer data analyzed by \cite{kattan2000pretreatment},
which we introduced in Section~\ref{sec:numer-exper}, include a competing event
in the form of death without tumor recurrence. To illustrate our method we fit
the state learner to the original data set consisting of 1,042 patients. We
consider death without tumor recurrence and recurrence of tumor as two competing
events of interest. We include the fives learner \texttt{KM}, \texttt{Cox strata
  CS}, \texttt{Lasso}, \texttt{Elastic}, and \texttt{RF} which are described in
Table~\ref{tab:zel-library}. We use the same library of learners to learn
\( \Lambda_1 \), \( \Lambda_2 \), and $\Gamma$. In this case, \( \Lambda_1 \)
denotes the cause-specific cumulative hazard function of tumor recurrence, and
\( \Lambda_2 \) denotes the cause-specific cumulative hazard function of death
without tumor recurrence.

This gives a library consisting of \( 5^3 = 125 \) learners for the conditional
state occupation probability function \( F \) defined in
equation~(\ref{eq:F-def}). We use five folds for training and testing the
models, and we repeat training and evaluation five times with different splits.
The integrated Brier score for all learners are shown in
Figure~\ref{fig:zelefski-real}, and the top 10 combinations of learners are
displayed in Table~\ref{tab:zelefski-real}. We see that the prediction
performance is mostly affected by the choice of learner for the censoring
distribution. Several combinations of learners give similar performance as
measured by the integrated Brier score, as long as a random forest is used to
model the censoring distribution.

\begin{figure}
  \centerline{\includegraphics[width=1\linewidth]{./figures/zelefski-real-data.pdf}}
  \caption[]{The results of applying the 125 combinations of learners to the
    prostate cancer data set. The learners are \texttt{KM} (KM), \texttt{Cox
      strata CS} (strata), \texttt{Lasso} (lasso), \texttt{Elastic} (elastic),
    and \texttt{RF} (RF) as described Table~\ref{tab:zel-library}. The error
    bars are based on five repetitions using different splits. We refer to
    learners of \( \Lambda_1 \), \( \Lambda_2 \), and $\Gamma$ as `Tumor
    learner', `Death learner', and `Censoring learner', respectively.}
  \label{fig:zelefski-real}
\end{figure}

\begin{table}[ht]
  \centering \input{./figures/zel-tab.tex}
  \caption{The 10 best performing models in terms of integrated Brier score. The
    reported standard errors are based on five repetitions using different
    splits. The models are described in Table~\ref{tab:zel-library}. We refer to
    learners of \( \Lambda_1 \), \( \Lambda_2 \), and $\Gamma$ as `Tumor
    learner', `Death leaner', and `Censoring learner', respectively.}
  \label{tab:zelefski-real}
\end{table}
\section{Discussion}
\label{sec:discussion}

We have proposed a new super learner that can be used with right-censored data
and competing events. In this section, we compare our proposal to existing super
learners and discuss avenues for further research.

\subsection{Existing super learners for right-censored data}
\label{sec:relat-liter-exist}

Machine learning based on right-censored data commonly use the negative partial
log-likelihood as loss function
\citep[e.g.,][]{li2016regularized,yao2017deep,lee2018deephit,katzman2018deepsurv,gensheimer2019scalable,lee2021boosted,kvamme2021continuous}.
However, this loss function is unsuited for super learning, because many
canonical survival learners (e.g., the Kaplan-Meier estimator, random survival
forest, and semi-parametric Cox models) provide cumulative hazard functions that
are piece-wise constant in the time argument, and hence assign zero probability
to event times not observed in the training data. This implies that when data
are observed in continuous time, any of these learners will almost surely have
infinite loss in any independent hold-out sample according to the negative
partial log-likelihood loss. When a proportional hazards model is assumed, the
baseline hazard function can be profiled out of the likelihood to give a new
partial log-likelihood loss \citep{cox1972regression}, which has been suggested
as a loss function for super learning
\citep{golmakani2020super,verweij1993cross}. While this allows the library of
learners to include Cox' proportional hazard models, the drawback is that the
library is in fact \textit{only} allowed to include these models. The advantage
of the state learner is that it does not require evaluation of the density of
\( F(\blank, j, x) \) and does not assume a particular semi-parametric structure
for $\Lambda_j$ but can be used with any library of learners.

Another approach for super learning with right-censored data is to use an
inverse probability of censoring weighted (IPCW) loss function
\citep{graf1999assessment,van2003unicv,molinaro2004tree,keles2004asymptotically,hothorn2006survival,gerds2006consistent,gonzalez2021stacked}.
An IPCW loss function is attractive because the associated risk does not depend
on the censoring distribution but describes a feature of the population of
interested governed by the measure \( Q \in \mathcal{Q} \). Similar results can
be obtained using censoring unbiased transformations
\citep{fan1996local,steingrimsson2019censoring} or pseudo-values
\citep{andersen2003generalised,mogensen2013random,sachs2019ensemble}. All these
methods rely on an estimator of the censoring distribution, and their drawback
is that this estimator has to be pre-specified. When the data-generating
mechanism is complex and not well-understood, pre-specification of the censoring
distribution is a challenge. The advantage of using the state learner is that a
censoring distribution need not be pre-specified but is selected automatically
based on the provided library \( \mathcal{B} \).

To the best of our knowledge, the only existing attempt at avoiding the need to
pre-specify a censoring model is a recent proposal suggested independently by
\cite{han2021inverse} and \cite{westling2021inference}. The authors do not
consider competing risks but suggest to iterate between learning \( \Lambda \)
and $\Gamma$ using IPCW loss functions and select the final learner when the
iterative procedure has converged.
% Using the estimator \( \hat{b}^{(0)}(\data_n^c) \) of $\Gamma$ based on some
% initial learner \( \hat{b}^{(0)} \in \mathcal{B} \), we calculate the
% estimated IPCW risk for each leaner in \( \mathcal{A} \). Picking
% \( a^{(1)} \) as the discrete super learner according to this IPCW loss
% function, we then estimate weight
No general theoretical guarantees exist for this procedure, but it would be
interesting to compare its performance to that of the state learner in a
simulation study.

% In a non-survival targeted learning setting \cite{sun2022selective}, based on
% ideas presented by \cite{robins2007comment}, proposed to use as a model
% selection criteria the so-called `doubly robustness' property that some
% targeted estimators enjoy. In many special cases the remainder term
% \( \mathrm{Rem}{(\hat{\Lambda}_{n}, \hat{\Gamma}_n, P)} \) defined in
% equation~(\ref{eq:ral}) has the property that
% \begin{equation}
%   \label{eq:dr-property}
%   \mathrm{Rem}{(\Lambda^*, \Gamma_0, P_0)}
%   = \mathrm{Rem}{(\Lambda_0, \Gamma^*, P_0)} = 0,
%   \quad \text{for any }
%   \Lambda^*, \Gamma^* \in \mathcal{M}.
% \end{equation}
% The property in equation~(\ref{eq:dr-property}) is known as `doubly
% robustness' because it ensures that
% $\tilde{\Psi}(\hat{\Lambda}_{n}, \hat{\Gamma}_n, H_n)$ defined in
% equation~(\ref{eq:ral}) is consistent if just one of $\Lambda$ or $\Gamma$ is
% estimated consistently
% \citep{van2003unified,bang2005doubly,kang2007demystifying}. For
% $a \in \mathcal{A}$ and \( b \in \mathcal{B} \), define
% \( \hat{\Psi}^k_n(a, b) = \tilde{\Psi}(a(\data_n^k), b(\data_n^k), \empmeas^k) \)
% where \( \empmeas^{k} \) is the empirical measure of \( \data_n^{k} \).
% \cite{sun2022selective} suggest to select $a \in \mathcal{A}$ as the minimizer
% of the estimated `fluctuation pseudo-risk',
% \begin{equation*}
%   \hat R_{\mathrm{f}}(a) =
%   \max_{b_1, b_2 \in \mathcal{B}}
%   \frac{1}{K}\sum_{k=1}^{K}
%   \left\{
%     \hat{\Psi}^k_n(a, b_1)
%     - \hat{\Psi}^k_n(a, b_2)
%   \right\}^2.
% \end{equation*}
% The idea is that the doubly robustness property
% (equation~(\ref{eq:dr-property})) implies that \(\hat{R}_{\mathrm{f}}(a)\) will
% be zero (asymptotically) if the learner \(a\) correctly estimates $\Lambda$. The
% authors establish finite sample results guarantying that the selected learner
% \(a\) will be robust against changing the learner of the other nuisance
% parameter across $\mathcal{B}$. It seems unclear to what extend this robustness
% property also guarantees consistency of the estimator of the target parameters
% when we use the fluctuation risk \(\hat{R}_{\mathrm{f}}\) to select the learners
% of the nuisance parameters. It would be interesting to explore the performance
% of this procedure in a survival setting and compare it to targeted estimators
% obtained using the state learner as outlined in
% Section~\ref{sec:targeted-learning}.

\subsection{A performance measure of interest}
\label{sec:perf-meas-inter}

A major advantage of the state learner is that performance of each combination
of learners is measured in terms of observable quantities. This means that no
additional nuisance parameters need to be estimated to evaluate the loss. The
drawback with this approach is that we are rarely interested in features of the
observed data distribution when the data are right-censored. The finite sample
oracle inequality in Corollary~\ref{cor:oracle-prop} concerns the function
\( F \), which is a feature of \( P \in \mathcal{P} \), while what we are
typically interested in is \( \Lambda_j \) or \( S \), which are features of
\( Q \in \mathcal{Q} \). We emphasize that while the state learner provides us
with estimates of \( \Lambda_j \) and $\Gamma$ based on libraries
\( \mathcal{A}_j \) and \( \mathcal{B} \), performance are not assessed directly
for these parameters, but only jointly for estimation of the parameter \( F \).
For settings without a competing risk, our numerical studies suggest that
measuring performance with respect to estimation of \( F \) also leads to good
performance for estimation of \( S \). Further research on this topic, both
numerical and theoretical, is warranted.

\fxnote*{mention this?}{[Double robustness maybe lacking]}

% In the context of targeted learning, a drawback of the state learner is that the
% doubly robustness property defined in equation~(\ref{eq:dr-property}) seem to be
% lost because we only use a single nuisance parameter estimator. As defined here,
% however, the state learner is build using libraries for the conditional
% cause-specific cumulative hazard functions, so some doubly robustness might be
% preserved.

\subsection{Implementation}
\label{sec:software}

Our proposed super learner can be implemented with a broad library of learners
using existing software, for instance the \texttt{R}-package
\texttt{riskRegression} \citep{Gerds_Ohlendorff_Ozenne_2023}. Furthermore, while
the library \( \mathcal{F}(\mathcal{A}_1,\mathcal{A}_2,\mathcal{B}) \) consists
of \( |\mathcal{A}_1||\mathcal{A}_2||\mathcal{B}| \) learners, as long as we
have sufficient memory we need only fit
\( |\mathcal{A}_1| +|\mathcal{A}_2| + |\mathcal{B}| \) learners in each fold. To
evaluate the performance of each learner we need to perform
\( |\mathcal{A}_1||\mathcal{A}_2||\mathcal{B}| \) operations to calculate the
integrated Brier score in each hold-out sample, one for each combination of the
fitted models, but these operations are often negligible compared to fitting the
models. Hence the state learner is essentially no more computationally demanding
than any procedure that uses super learning to learn $\Lambda_1$, $\Lambda_2$,
and $\Gamma$ separately. While our proposal is based on constructing the library
\( \mathcal{F} \) from libraries for learning \( \Lambda_1 \), $\Lambda_2$, and
$\Gamma$, it would also be of interest to consider learners that estimate
\( F \) directly.

In our numerical studies, we only considered learners of $\Lambda_j$ and
$\Gamma$ that provide cumulative hazard functions which are piece-wise constant
in the time argument. This simplifies the calculation of \( F \) as the
integrals in equation~(\ref{eq:transition}) reduce to sums. When $\Lambda_j$ or
\( \Gamma \) are absolutely continuous in the time argument, calculating \( F \)
is more involved, but we expect that a good approximation can be achieved by
discretization. In the future, we intend to investigate the performance of the
state learner when using a broader library of learners in more comprehensive
simulation studies.

% \bibliography{bib.bib}

\appendix

\section{Theoretical guarantees for the state learner}
\label{sec:proof-proposition}

In this section we provide proofs of the results stated in
Section~\ref{sec:theor-results-prop}.

Define
\( \bar{B}_{\tau,0}(F, o) = \bar{B}_{\tau}(F, o) - \bar{B}_{\tau}(F_0, o) \) and
\( R_{0}(F) = P_0{[\bar{B}_{\tau,0}(F, \blank)]} \).
\begin{lemma}
  \label{lemma:norm}
  \( R_{0}(F) = \Vert F - F_0 \Vert_{P_0}^2 \), where \( \Vert \blank \Vert_{P_0}\) is defined
  in equation~(\ref{eq:norm}).
\end{lemma}
\begin{proof}
  For any \( t \in [0, \tau] \) and \( k\in \{0,1,2\} \) we have
  \begin{align*}
    & \E_{P_0}{\left[ (F(t, k, X) - \1{\{\eta(t) = k \}})^2 \right]}
    \\
    & =    \E_{P_0}{\left[ (F(t, k, X) - F_0(t, k, X) + F_0(t, k, X) - \1{\{\eta(t) = k
      \}})^2 \right]}
    \\
    & =    \E_{P_0}{\left[ (F(t, k, X) - F_0(t, k, X))^2\right]}
      + \E_{P_0}{\left[ (F_0(t, k, X) - \1{\{\eta(t) = k \}})^2\right]}
    \\
    & \quad
      + 2\E_{P_0}{\left[ (F(t, k, X) - F_0(t, k, X))(F_0(t, k, X) - \1{\{\eta(t) = k
      \}})\right]}
    \\
    & =    \E_{P_0}{\left[ (F(t, k, X) - F_0(t, k, X))^2\right]}
      + \E_{P_0}{\left[ (F_0(t, k, X) - \1{\{\eta(t) = k \}})^2\right]},
  \end{align*}
  where the last equality follows from the tower property. Hence, using Fubini,
  we have
  \begin{equation*}
    P{[\bar{B}_{\tau}(F, \blank)]}
    = \Vert F - F_0 \Vert_{P_0}^2 + P_0{[\bar{B}_{\tau}(F_0, \blank)]}. \qedhere
  \end{equation*}
\end{proof}

\begin{proof}[Proof of Proposition~\ref{prop:stric-prop}]
  The result follows from Lemma~\ref{lemma:norm}.
\end{proof}

In the following, let $\Theta$ denote the function space consisting of all
conditional state occupation probability functions for some measure \( P \) as
defined in equation~(\ref{eq:F-def}).

\begin{proof}[Proof of Corollary~\ref{cor:oracle-prop}]
  First note that minimising the loss \( \bar{B}_{\tau} \) is equivalent to
  minimising the loss \( \bar{B}_{\tau,0} \), so the discrete super learner and
  oracle according to \( \bar{B}_{\tau} \) and \( \bar{B}_{\tau,0} \) are
  identical. By Lemma~\ref{lemma:norm}, \( R_0(F) \geq 0 \) for any
  \( F \in \Theta \), and so using Theorem 2.3 from \citep{vaart2006oracle} with
  \( p=1 \), we have that for all \( \delta >0 \),
\begin{align*}
  \E_{P_0}{\left[ R_0(\hat{\phi}_n(\data_n^{-k})) \right]}
  \leq
  &(1+2\delta)\E_{P_0}{\left[ R_0(\tilde{\phi}_n(\data_n^{-k})) \right]}
  \\
  & \quad + (1+\delta) \frac{16 K}{n}
    \log(1 + |\mathcal{F}_n|)\sup_{F \in \Theta}
    \left\{
    M(F) + \frac{v(F)}{R_0(F)}
    \left(
    \frac{1}{\delta} + 1
    \right)
    \right\}
\end{align*}
where for each \( F \in \Theta \), \( (M(F), v(F)) \) is some Bernstein pair for
the function \(o \mapsto \bar{B}_{\tau,0}(F, o) \). As
\( \bar{B}_{\tau,0}(F, \blank) \) is uniformly bounded by \( \tau \) for any
\( F \in \Theta \), it follows from section 8.1 in \citep{vaart2006oracle} that
\( (\tau, 1.5 P_0{[\bar{B}_{\tau,0}(F, \blank)^2]}) \) is a Bernstein pair for
\( \bar{B}_{\tau,0}(F, \blank) \). Now, for any \( a,b,c \in \R \) we have
\begin{align*}
  (a-c)^2 - (b-c)^2
  & = (a-b+b-c)^2 - (b-c)^2
  \\
  & = (a-b)^2 + (b-c)^2 +2(b-c)(a-b) - (b-c)^2
  \\
  & = (a-b)
    \left\{
    (a-b) +  2(b-c)
    \right\}
  \\
  & = (a-b)
    \left\{
     a + b -2c
    \right\},
\end{align*}
so using this with \( a=F(t, k, x) \), \( b=F_0(t, k, x) \), and
\( c = \1{\{\eta(t) = k\}} \), we have by Jensen's inequality
\begin{align*}
  & P_0{[\bar{B}_{\tau,0}(F, \blank)^2]}
  \\
  & \leq
    2\tau\E_{P_0}{\left[
    \sum_{k=0}^{3} \int_0^{\tau}
    \left\{
    \left(
    F(t, k, X) - \1{\{\eta(t) = k\}}
    \right)^2
    -
    \left(
    F_0(t, k, X) - \1{\{\eta(t) = k\}}
    \right)^2
    \right\}^2
    \diff t 
    \right]}
  \\
  & =2\tau
    \E_{P_0}\Bigg[
    \sum_{k=0}^{3} \int_0^{\tau}
    \left(
    F(t, k, X) - F_0(t, k, X)
    \right)^2
  \\
  & \quad \quad \quad\quad \quad \quad \times
    \left\{
    F(t, k, X) +  F_0(t, k, X)-2 \1{\{\eta(t) = k\}}
    \right\}^2
    \diff t 
    \Bigg]
  \\
  & \leq
    8\tau \E_{P_0}{\left[
    \sum_{k=0}^{3} \int_0^{\tau}
    \left(
    F(t, k, X) - F_0(t, k, X)
    \right)^2
    \diff t 
    \right]}.
  \\
  & =
    8\tau \Vert F - F_0 \Vert_{P_0}^2.
\end{align*}
Thus when \( v(F) = 1.5 P_0{[\bar{B}_{\tau,0}(F, \blank)^2]} \) we have by
Lemma~\ref{lemma:norm}
\begin{equation*}
  \frac{v(F)}{R_0(F)}
  = 1.5 \frac{P_0{[\bar{B}_{\tau,0}(F, \blank)^2]}}{P_0{[\bar{B}_{\tau,0}(F, \blank)]}}
  \leq 12 \tau,
\end{equation*}
and so using the Bernstein pairs \( (\tau, 1.5 P_0{[\bar{B}_{\tau,0}(F, \blank)^2]}) \) we have
\begin{equation*}
  \sup_{F \in \Theta}
  \left\{
    M(F) + \frac{v(F)}{R_0(F)}
    \left(
      \frac{1}{\delta} + 1
    \right)
  \right\}
  \leq \tau
  \left(
    13 + \frac{12}{\delta}
  \right),
\end{equation*}
For all $\delta>0$ we thus have
\begin{align*}
  \E_{P_0}{\left[ R_0(\hat{\phi}_n(\data_n^{-k})) \right]}
  \leq
  &(1+2\delta)\E_{P_0}{\left[ R_0(\tilde{\phi}_n(\data_n^{-k})) \right]}
  \\
  & \quad
    + (1+\delta)\log(1 + |\mathcal{F}_n|) \tau \frac{16 K}{n}
    \left(
    13 + \frac{12}{\delta}
    \right),
\end{align*}
and then the final result follows from Lemma~\ref{lemma:norm}.
\end{proof}

\begin{proof}[Proof of Corollary~\ref{cor:asymp-cons}]
  By definition of the oracle and Lemma~\ref{lemma:norm},
  \( \E_{P_0}{\left[ \Vert \tilde{\phi}_n(\data_n^{-k}) - F_0 \Vert_{P_0}^2 \right]}
  \leq \E_{P_0}{\left[ \Vert \phi_n(\data_n^{-k}) - F_0 \Vert_{P_0}^2 \right]} \) for
  all \( n \in \N \). The results then follows from
  Corollary~\ref{cor:oracle-prop}.
\end{proof}

\section{The state learner with targeted learning}
\label{sec:state-learner-with}

In this section show that a product structure is preserved when the estimator
$\bar\Psi(\hat{F}_n, \hat{H}_n)$ is used instead of
$\tilde\Psi(\hat{\Lambda}_n, \hat{\Gamma}_n, \hat{H}_n)$.

\begin{proof}[Proof of Proposition~\ref{prop:dr-structure}]
  For notational convenience we suppress \( X \) in the following. The final
  result can be obtained by adding the argument \( X \) to all functions and
  averaging. We use the relations from equation~(\ref{eq:7}) to write
  \begin{align*}
    & \int_0^{\tau} w(s) 
      \left\{
      \Gamma(s) - \hat{\Gamma}_n(s)
      \right\}
      [\Lambda - \hat{\Lambda}_n](\diff s)
    \\
    & =
      \int_0^{\tau} w(s) 
      \left\{
      \int_0^s \frac{F(\diff u, 2)}{F(u-, 0)} -
      \int_0^s \frac{\hat{F}_n(\diff u, 2)}{\hat{F}_n(u-, 0)}  -
      \right\}
      \left[
      \frac{F(\diff s, 1)}{F(s-, 0)}
      - \frac{\hat{F}_n(\diff s, 1)}{\hat{F}_n(s-, 0)}
      \right]
    \\
    & =
      \int_0^{\tau} w(s) 
      \Bigg\{
      \int_0^s 
      \left(
      \frac{1}{F(u-, 0)} -  \frac{1}{\hat{F}_n(u-, 0)}
      \right) F(\diff u, 2)
    \\
    & \qquad\qquad \qquad
      +
      \int_0^s \frac{1}{\hat{F}_n(u-, 0)} 
      \left[
      F(\diff u, 2) - \hat{F}_n(\diff u, 2)
      \right]
      \Bigg\}
    \\
    & \qquad\qquad \times
      \left[
      \left(
      \frac{1}{F(s-, 0)} -
      \frac{1}{\hat{F}_n(s-, 0)}
      \right)F(\diff s, 1)
       + \frac{1}{\hat{F}_n(s-, 0)}
      \left(
      F(\diff s, 1) -
      \hat{F}_n(\diff s, 1)
      \right)
      \right]
    \\
    &
      = \int_0^{\tau} 
      \int_0^s
      w(s) 
      \left(
      \frac{1}{F(u-, 0)} -  \frac{1}{\hat{F}_n(u-, 0)}
      \right) 
      \left(
      \frac{1}{F(s-, 0)} -
      \frac{1}{\hat{F}_n(s-, 0)}
      \right)F(\diff u, 2)F(\diff s, 1)
    \\
    & \quad +
      \int_0^{\tau}
      \int_0^s
      w(s) 
      \left(
      \frac{1}{F(u-, 0)} -  \frac{1}{\hat{F}_n(u-, 0)}
      \right) \frac{F(\diff u, 2) }{\hat{F}_n(u-,0)}
      \left(
      F(\diff s, 1) -
      \hat{F}_n(\diff s, 1)
      \right)
    \\
    & \quad +
      \int_0^{\tau} 
      \int_0^s      
      \frac{w(s) }{\hat{F}_n(u-, 0)} 
      \left[
      F(\diff u, 2) - \hat{F}_n(\diff u, 2)
      \right]
      \left(
      \frac{1}{F(s-, 0)} -
      \frac{1}{\hat{F}_n(s-, 0)}
      \right)F(\diff s, 1)
    \\
    & \quad +
      \int_0^{\tau} 
      \int_0^s      
      \frac{w(s) }{\hat{F}_n(u-, 0)} 
      \left[
      F(\diff u, 2) - \hat{F}_n(\diff u, 2)
      \right]
      \frac{1}{\hat{F}_n(s-, 0)}
      \left(
      F(\diff s, 1) -
      \hat{F}_n(\diff s, 1)
      \right).
  \end{align*}
  Consider the first term on the right hand side. Defining
  \begin{equation*}
    w_n^*(t)  = 
    \left(
      F(t-, 0)
      - \hat{F}_n(t-, 0)
    \right)
    \left(
      \frac{1}{F(t-, 0)}
      - \frac{1}{\hat{F}_n(t-, 0)}
    \right),
  \end{equation*}
  we can write
  \begin{align*}
    & \int_0^{\tau} 
      \int_0^s
      w(s) 
      \left(
      \frac{1}{F(u-, 0)} -  \frac{1}{\hat{F}_n(u-, 0)}
      \right)      
      \left(
      \frac{1}{F(s-, 0)} -
      \frac{1}{\hat{F}_n(s-, 0)}
      \right)F(\diff u, 2)F(\diff s, 1)
    \\
    & =
      \int_0^{\tau} 
      \int_0^s
      w(s)
      w_n^*(u) 
      \left(
      F(u-, 0) - \hat{F}_n(u-, 0)
      \right)
    \\
    & \qquad \qquad \quad
      \times
      w_n^*(s) 
      \left(
      F(s-, 0) - \hat{F}_n(s-, 0)
      \right)       
      F(\diff u, 2)F(\diff s, 1)
    \\
    & =
      \int_0^{\tau} 
      \int_0^s
      w_n^a(s,u)
      \left(
      F(u-, 0) - \hat{F}_n(u-, 0)
      \right)
      \left(
      F(s-, 0) - \hat{F}_n(s-, 0)
      \right)       
      F(\diff u, 2)F(\diff s, 1),
  \end{align*}
  where we have defined \( w_n^a(s,u) = w(s)w^*_n(s)w^*_n(u) \). By assumption,
  \( w_n^a(s,u) \) is uniformly bounded. The same approach can be applied to the
  three remaining terms which gives the result.
\end{proof}

\bibliography{bib.bib}

\end{document}

### 2023-11-06 R-code/functions/abs-risk-ate.R

@@ -1,0 +1,474 @@
+### martingale-int.R --- 
+#----------------------------------------------------------------------
+## Author: Anders Munch
+## Created: Oct 27 2023 (16:08) 
+## Version: 
+## Last-Updated: Nov  3 2023 (11:43) 
+##           By: Anders Munch
+##     Update #: 376
+#----------------------------------------------------------------------
+## 
+### Commentary: 
+## 
+### Change Log:
+#----------------------------------------------------------------------
+## 
+### Code:
+library(here)
+library(data.table)
+source(here("R-code/functions", "integral-calc.R"))
+
+## naiv (data, t, Lambda1, Lambda2, jump_points)
+naiv <- function(data, t, Lambda1, Lambda2, jump_points, chunks = 1){
+    Lambda1_ = minus_t_fun(Lambda1,jump_points)
+    Lambda2_ = minus_t_fun(Lambda2,jump_points)
+    A_fun = function(newdata,times) exp(-(Lambda1_(newdata,times) + Lambda2_(newdata,times)))
+    data_treat0 = copy(data)[, A := 0]
+    data_treat1 = copy(data)[, A := 1]
+    preds0 = integral1(A = A_fun,B = Lambda1,data = data_treat0,end = t,jump_points = jump_points, chunks = chunks)
+    preds1 = integral1(A = A_fun,B = Lambda1,data = data_treat1,end = t,jump_points = jump_points, chunks = chunks)
+    out = list("A=1" = preds1, "A=0" = preds0)
+    return(out)
+}
+
+
+## Checking:
+termW <- function(data, pi){
+    pi_fit = as.numeric(pi(data))
+    out1 = numeric(nrow(data))
+    A1_ind = data[, A == 1]
+    out1[A1_ind] = 1/pi_fit[A1_ind]    
+    out0 = numeric(nrow(data))
+    out0[!A1_ind] = 1/(1-pi_fit[!A1_ind])
+    out = list("A=1" = out1, "A=0" = out0)
+    return(out)
+}
+termA <- function(data, t, Lambda1, Gamma, jump_points,chunks = 1){
+    Gamma_ = minus_t_fun(Gamma,jump_points)
+    ## Counting term
+    out_count = count_integral(A = function(newdata,times) exp(Gamma_(newdata,times)),t = t,times = data[, time],events = data[, (status == 1)],data = data,chunks = chunks)
+    ## Compensator
+    A_fun = function(newdata,times) {exp(Gamma_(newdata,times))*at_risk_fun(newdata, times)}
[…]

### 2023-11-06 R-code/functions/colon-data-sim.R

@@ -1,0 +1,108 @@
+### validate-estimator.R --- 
+#----------------------------------------------------------------------
+## Author: Anders Munch
+## Created: Nov  2 2023 (12:07) 
+## Version: 
+## Last-Updated: Nov  3 2023 (15:17) 
+##           By: Anders Munch
+##     Update #: 196
+#----------------------------------------------------------------------
+## 
+### Commentary: 
+## 
+### Change Log:
+#----------------------------------------------------------------------
+## 
+### Code:
+
+library(survival)
+library(data.table)
+library(here)
+library(parallel)
+source(here("R-code/functions", "hely-colon-sim.R"))
+
+## ## form1=Surv(time, event==1)~sex+differ+age+adhere
+## form1=Surv(time, event==1)~rx+sex+differ+age+adhere
+## ## form2=Surv(time, event==2)~sex+nodes+differ+age
+## form2=Surv(time, event==2)~rx+sex+nodes+differ+age
+## form0=Surv(time, event==0)~rx+sex
+## formA=rx~sex+age+nodes+differ
+
+fit_estimator <- function(data, nuisance_form, est_times){
+    ## Fit nuisance models
+    cause1_fit <- coxph(nuisance_form[["formula_1"]],data = data,x = TRUE, y = TRUE)
+    cause2_fit <- coxph(nuisance_form[["formula_2"]],data = data,x = TRUE, y = TRUE)
+    cens_fit <- coxph(nuisance_form[["formula_cens"]],data = data,x = TRUE, y = TRUE)
+    treat_fit <- glm(nuisance_form[["formula_treat"]], family = binomial(), data = data)
+    ## Setup on correct format
+    L1 <- construct_pred_fun(cause1_fit)
+    L2 <- construct_pred_fun(cause2_fit)
+    L0 <- construct_pred_fun(cens_fit)
+    pi <- construct_pred_fun(treat_fit)
+    ## Fit target parameters
+    out = do.call(rbind, lapply(est_times, function(est_tt){
+        tar_est = abs_risk_ate(data = data,
+                               t = est_tt,
+                               Lambda1 = L1,
+                               Lambda2 = L2,
+                               Gamma = L0,
+                               pi = pi,
+                               jump_points = sort(unique(data[, time])), collapse = 0)
+
+        effect_names = c("A=1","A=0","ATE")
+        naiv1 = mean(tar_est$naiv1_i)
+        naiv0 = mean(tar_est$naiv0_i)
+        naiv_dt = data.table(type = "naiv",
[…]

### 2023-11-06 R-code/functions/hely-colon-sim.R

@@ -1,0 +1,283 @@
+### hely-colon-sim.R --- 
+#----------------------------------------------------------------------
+## Author: Anders Munch
+## Created: Oct 30 2023 (09:07) 
+## Version: 
+## Last-Updated: Nov  3 2023 (11:25) 
+##           By: Anders Munch
+##     Update #: 31
+#----------------------------------------------------------------------
+## 
+### Commentary: 
+## 
+### Change Log:
+#----------------------------------------------------------------------
+## 
+### Code:
+library(data.table)
+library(survival)
+
+
+fit.categorical <- function(formula,data,pseudo.out=FALSE,...){
+    pseudo.data <- get.pseudodata(formula,data)
+    # introduce all interactions
+    pseudo.formula <- update(formula,"pseudo.outcome~(.)*pseudo.factor")
+    # fit pseudo model
+    fit <- glm(pseudo.formula, data=pseudo.data[pseudo.data$at.risk==1,], family=binomial())
+    if (pseudo.out) {
+        pseudo.data$pred <- predict(fit, type="response", newdata=pseudo.data)
+        predicted <- cbind(0, do.call("cbind", lapply(unique(pseudo.data$pseudo.aa), function(aa) {
+            pseudo.data[pseudo.data$pseudo.aa==aa, "pred"]
+        })))
+        predicted2 <- do.call("cbind", lapply(2:ncol(predicted), function(jj) {
+            apply(1-predicted[, 1:(jj-1), drop=FALSE], 1, prod)*predicted[, jj]
+        }))
+        colnames(predicted2) <- unique(pseudo.data$pseudo.aa)
+    }
+    # add the call
+    fit$catfit.call <- match.call()
+    # add the outcome values
+    fit$y <- attr(pseudo.data,"outcome.values")
+    # add a new S3 class
+    class(fit) <- c("catfit",class(fit))
+    # return the fit
+    if (pseudo.out) {
+        return(list(fit, predicted2))
+    } else {
+        return(list(fit))
+    }
+}
+
+get.pseudodata <- function(formula,data,atrisk=TRUE){
+    d <- model.frame(formula,data)
+    # should check that outcome is categorical 
+    N <- NROW(d)
+    d$id <- 1:N
+    Y <- d[[1]]
+    outcome <- names(d)[[1]]
+    # HOV: need to deal with factors and factor levels
+    # that cannot be coerced to numeric
+    if (is.factor(Y))
+        outcome.values <- levels(Y)
+    else
[…]

### 2023-11-06 R-code/functions/integral-calc.R

@@ -1,0 +1,393 @@
+### integral-calc.R --- 
+#----------------------------------------------------------------------
+## Author: Anders Munch
+## Created: Oct 18 2023 (15:10) 
+## Version: 
+## Last-Updated: Nov  3 2023 (10:43) 
+##           By: Anders Munch
+##     Update #: 525
+#----------------------------------------------------------------------
+## 
+### Commentary:
+##
+## These functions calculate varioues integrals
+## based on piece-wise constant functions.
+## 
+### Change Log:
+#----------------------------------------------------------------------
+##
+
+## Notes:
+## Later on, consider implementing the integral functions where we supply matrices.
+## Then they can be reused later on.
+## Should minimize the number of times we need to call predict function.
+## But then we need to export things in the end.
+## Could we somehow write a completely functional expression, so that R figures out
+## when to evaluate the supplied functions. 
+## Thus only provide functions once, and number of chuncks once!
+## All other functions should the be symboblic combinations of the supplied functions.
+## Would also make stuff easier to read.
+## And then everything the combiner (chunk) function could be exported to C++.
+
+### Code:
+library(data.table)
+library(prodlim)
+library(riskRegression)
+
+## Meta functions:
+minus_t_fun <- function(Lambda, jump_points){
+    t_minus = min(diff(sort(unique(jump_points))))/2
+    fun0 = function(newdata, times){
+        times0 = pmax(0, times-t_minus)
+        Lambda(newdata, times0)
+    }
+    return(fun0)
+}
+construct_pred_fun <- function(object, ...){
+    UseMethod("construct_pred_fun",object)
+}
+construct_pred_fun.coxph <- function(model, ...){
+    out = function(newdata,times)
+        matrix(predictCox(model, newdata = newdata, times = times, type = "cumhazard")$cumhazard, ncol = length(times))
+    return(out)
+}
+construct_pred_fun.glm <- function(model, ...){
+    out = function(newdata)
+        matrix(predict(model, newdata = newdata, type = "response"), ncol = 1)
+    return(out)
+    
+}
+#### at risk funcion
+at_risk_fun <- function(newdata,times,time_name = "time"){
+    spring_times = newdata[[time_name]]
+    grid_mat = matrix(times,ncol = length(times),nrow = length(spring_times),byrow = TRUE)
[…]

### 2023-11-06 R-code/pilot-sim-study/_targets.R

@@ -1,0 +1,74 @@
+### _targets.R --- 
+#----------------------------------------------------------------------
+## Author: Anders Munch
+## Created: Nov  3 2023 (09:35) 
+## Version: 
+## Last-Updated: Nov  6 2023 (09:06) 
+##           By: Anders Munch
+##     Update #: 103
+#----------------------------------------------------------------------
+## 
+### Commentary: 
+## 
+### Change Log:
+#----------------------------------------------------------------------
+## 
+### Code:
+try(setwd("~/Documents/phd/statelearner/R-code/pilot-sim-study/"))
+library(here)
+library(targets)
+library(tarchetypes)
+library(parallel)
+tar_source(here("R-code/functions"))
+
+## DGM
+dgm_settings <- list(
+    formula_1 = Surv(time, event==1)~treat+sex+differ+age+adhere+nodes+nodes.squared+treat*sex+treat*perfor,
+    formula_2 = Surv(time, event==2)~treat+sex+nodes+differ+age,
+    formula_cens = Surv(time, event==0)~treat+sex+nodes+treat*sex,
+    formula_treat = treat~sex+age+nodes+differ+nodes.squared)
+
+## Sim settings
+true_settings <- list(
+    formula_1 = Surv(time, status==1)~factor(A)+sex+differ+age+adhere+nodes+nodes.squared+factor(A)*sex+factor(A)*perfor,
+    formula_2 = Surv(time, status==2)~factor(A)+sex+nodes+differ+age,
+    formula_cens = Surv(time, status==0)~factor(A)+sex+nodes+factor(A)*sex,
+    formula_treat = A~sex+age+nodes+differ+nodes.squared)
+nuisance_settings <- rep(list(true_settings), 5)
+## names(nuisance_settings) <- c("all_correct", "mis_treat", "mis_out")
+## names(nuisance_settings) <- c("all_correct", "mis_treat", "mis_cens", "mis_out")
+names(nuisance_settings) <- c("all_correct", "mis_treat", "mis_cens", "mis_out", "mis_all")
+nuisance_settings[["mis_treat"]][["formula_treat"]] <- A~1
+nuisance_settings[["mis_cens"]][["formula_cens"]] <- Surv(time, status==0)~1
+nuisance_settings[["mis_out"]][["formula_1"]] <- Surv(time, status==1)~factor(A)
+nuisance_settings[["mis_out"]][["formula_2"]] <- Surv(time, status==1)~factor(A) + perfor
[…]

### 2023-11-06 R-code/pilot-sim-study/pilot-study-sandbox.R

@@ -1,0 +1,170 @@
+### pilot-study-sandbox.R --- 
+#----------------------------------------------------------------------
+## Author: Anders Munch
+## Created: Nov  3 2023 (11:50) 
+## Version: 
+## Last-Updated: Nov  3 2023 (19:07) 
+##           By: Anders Munch
+##     Update #: 28
+#----------------------------------------------------------------------
+## 
+### Commentary: 
+## 
+### Change Log:
+#----------------------------------------------------------------------
+## 
+### Code:
+
+
+
+fit.colon.cr <- fit.colon.fun(      
+    ## formula.1=Surv(time, event==1)~rx+sex+differ+age+nodes.squared+obstruct+perfor+adhere+extent+surg+rx*sex+rx*perfor+rx*age,
+    ## formula.2=Surv(time, event==2)~rx+sex+nodes+differ+age+obstruct+adhere+extent+surg,
+    ## formula.0=Surv(time, event==0)~rx+sex+nodes+differ+age+obstruct+perfor+adhere+extent+surg,
+    ## formula.treat=rx~sex+age+nodes+differ+obstruct+perfor+adhere+extent+surg,
+    formula.1=form1,
+    formula.2=form2,
+    formula.0=form0,
+    formula.treat=formA,
+    d=colon_binary)
+#-- here we simulate data (fixed seed); 
+## set.seed(31)      
+sim.colon.cr <- synthesize.colon.fun(fit.colon=fit.colon.cr,
+                                     d=colon_binary,
+                                     name.treat="treat",
+                                     event.name="status",
+                                     n = 1000)
+## Setup data on right format
+sim.colon.cr[, A := as.numeric(treat)-1][, ":="(treat.num = NULL, treat = NULL)]
+
+## Get true stuff:
+if(FALSE){
+    (true_ate <- synthesize.colon.fun(fit.colon=fit.colon.cr,
+                                      get.true.value = 1,
+                                      d=colon_binary,
+                                      name.treat="treat",
+                                      event.name="status",
+                                      n = 1000000,
+                                      tau = 1000)["F1"]-
+         synthesize.colon.fun(fit.colon=fit.colon.cr,
[…]

### 2023-11-06 R-code/sandbox.R

@@ -1,0 +1,525 @@
+### sandbox.R --- 
+#----------------------------------------------------------------------
+## Author: Anders Munch
+## Created: Oct 27 2023 (15:40) 
+## Version: 
+## Last-Updated: Nov  2 2023 (12:06) 
+##           By: Anders Munch
+##     Update #: 216
+#----------------------------------------------------------------------
+## 
+### Commentary: 
+## 
+### Change Log:
+#----------------------------------------------------------------------
+## 
+### Code:
+
+library(targets)
+library(riskRegression)
+library(survival)
+library(prodlim)
+tar_source("./functions/")
+
+## Sandbox
+## Get something to play with
+
+
+## Reproduce CSC functionality first:
+set.seed(44)
+d <- sampleData(300,outcome="competing.risks")
+aa <- CSC(list(Hist(time,event)~X8, Hist(time,event)~X1+X2+X8), data=d)
+d[, state1 := 1*(event == 1)]
+d[, state2 := 1*(event == 2)]
+cox1 <- coxph(Surv(time,event = state1)~X8, data=d, x = TRUE, y = TRUE)
+cox2 <- coxph(Surv(time,state2)~X8, data=d, x = TRUE, y = TRUE)
+time_jumps <- d[time <= .5, sort(time)]
+## time_jumps <- c(0,d[time <= .5, sort(time)])
+cumHazard1 <- predictCox(cox1, newdata = d[1:5], times = time_jumps, type = "cumhazard")$cumhazard
+cumHazard2 <- predictCox(cox2, newdata = d[1:5], times = time_jumps, type = "cumhazard")$cumhazard
+Hazard1 <- predictCox(cox1, newdata = d[1:5], times = time_jumps, type = "hazard")$hazard
+Hazard2 <- predictCox(cox2, newdata = d[1:5], times = time_jumps, type = "hazard")$hazard
+## Slight differences are explained by difference between S(t) and S(t-)
+as.numeric(predictRisk(aa, newdata = d[1:5], times = .5, cause = 1, product.limit = 0))
+## rowSums(exp(- cumHazard1-cumHazard2)*Hazard1)
+rowSums(cbind(1,exp(- cumHazard1[,-ncol(cumHazard1), drop = FALSE]-cumHazard2[,-ncol(cumHazard2), drop = FALSE]))*Hazard1)
+
+## Testing:
+## Example functions:
+fun1 <- function(newdata, times){
[…]

### 2023-11-06 empirical-study/_targets.R

@@ -1,0 +1,174 @@
+### _targets.R --- 
+#----------------------------------------------------------------------
+## Author: Thomas Alexander Gerds
+## Created: Apr 27 2023 (11:06) 
+## Version: 
+## Last-Updated: Aug 21 2023 (14:08) 
+##           By: Anders Munch
+##     Update #: 171
+#----------------------------------------------------------------------
+## 
+### Commentary: 
+## 
+### Change Log:
+#----------------------------------------------------------------------
+## 
+### Code:
+try(setwd("~/research/SuperVision/Anders/survival-loss/statelearner/empirical-study/"))
+library(targets)
+library(tarchetypes)
+library(parallel)
+tar_option_set(packages = c("glmnet","lava","foreach","data.table","prodlim","survival","riskRegression","ranger"))
+tar_source("functions")
+list(
+    tar_target(test_data,draw_data(484)),
+    tar_target(test_fit,{
+        outcome_models <- list(cox0 = coxph(Surv(time,event = event)~1, data=test_data, x = TRUE, y = TRUE),
+                               cox1 = coxph(Surv(time,event = event)~X1+X2, data=test_data, x = TRUE, y = TRUE),
+                               cox2 = coxph(Surv(time,event = event)~X6+X7, data=test_data, x = TRUE, y = TRUE),
+                               cox_dgm = coxph(Surv(time,event = event)~ X1 + X2 + X5 + X6 + X7 + X8 + X11 + X12 + quad.X3 + quad.X4 + quad.X5 + quad.X6,data = test_data,x = TRUE,y = TRUE),
+                               rf = ranger(Surv(time, event = event) ~ X1+X2+X3+X4+X5+X6+X7+X8+X9+X10+X11+X12, data = test_data,num.tree = 50)
+                               ## rf_try = ranger(Surv(time, event = event) ~ X1+X2+X6+X7, mtry = 4, data = test_data),
+                               ## rf_deep = ranger(Surv(time, event = event) ~ X1+X2+X6+X7, mtry = 4, min.node.size = 1, data = test_data)
+                               )
+        cens_models <- list(cox1 = coxph(Surv(time,event = cens_status)~X1+X2, data=test_data, x = TRUE, y = TRUE),
+                            cox2 = coxph(Surv(time,event = cens_status)~X6+X7, data=test_data, x = TRUE, y = TRUE),
[…]

### 2023-11-06 empirical-study/functions/draw_data.R

@@ -1,0 +1,34 @@
+### draw_data.R --- 
+#----------------------------------------------------------------------
+## Author: Thomas Alexander Gerds
+## Created: Jul 11 2023 (13:07) 
+## Version: 
+## Last-Updated: Jul 11 2023 (14:15) 
+##           By: Thomas Alexander Gerds
+##     Update #: 5
+#----------------------------------------------------------------------
+## 
+### Commentary: 
+## 
+### Change Log:
+#----------------------------------------------------------------------
+## 
+### Code:
+draw_data <- function(n){
+    ## set effects
+    beta1 <- rep(c(1,-1,0,0,1,-1),2) * log(2)
+    q.beta1 <- c(1,-1,0,0,1,-1) * log(2)
+    q.gamma <- c(0,0,1,-1,1,-1) * log(2)
+    gamma <- rep(c(0,0,1,-1,1,-1),2) * log(2)
+    ## create data generating model 
+    qmodel <- quadratic_model(beta1=beta1,q.vars=paste("X",1:6,sep=""),q.beta1=q.beta1,q.gamma=q.gamma,gamma=gamma,competing.risks = FALSE)
+    d = sim(qmodel,n)
+    # add censoring status
+    setDT(d)
+    d[,cens_status := 1*(event != 0)]
+    d[]
+}
+
+
+######################################################################
+### draw_data.R ends here

### 2023-11-06 empirical-study/functions/quadratic_model.R

@@ -1,0 +1,63 @@
+### quadratic_model.R --- 
+#----------------------------------------------------------------------
+## Author: Thomas Alexander Gerds
+## Created: Jul 11 2023 (13:08) 
+## Version: 
+## Last-Updated: Jul 11 2023 (13:18) 
+##           By: Thomas Alexander Gerds
+##     Update #: 4
+#----------------------------------------------------------------------
+## 
+### Commentary: 
+## 
+### Change Log:
+#----------------------------------------------------------------------
+## 
+### Code:
+quadratic_model <- function(beta1,
+                            beta2,
+                            gamma,
+                            q.vars,
+                            q.beta1,
+                            q.beta2,
+                            q.gamma,
+                            competing.risks = TRUE){
+    require(lava)
+    m <- lvm()
+    X <- paste("X",seq(12),sep="")
+    timecause1 <- "timecause1"
+    if (competing.risks[[1]]) timecause2 <- "timecause2"
+    censtime <- "censtime"
+    if (competing.risks[[1]])
+        latent(m) <- c(timecause1,timecause2,censtime)
+    else
+        latent(m) <- c(timecause1,censtime)
+    if (competing.risks[[1]])
+        addvar(m) <- c(timecause1,timecause2,censtime,X)
+    else
+        addvar(m) <- c(timecause1,censtime,X)
+    exogenous(m) <- c(X,q.vars)
+    distribution(m,X[1:6]) <- normal.lvm()
+    distribution(m,X[7:12]) <- binomial.lvm()
+    distribution(m,timecause1) <- coxWeibull.lvm(scale=1/100)
+    if (competing.risks[[1]]) distribution(m,timecause2) <- coxWeibull.lvm(scale=1/100)
+    distribution(m,censtime) <- coxWeibull.lvm(scale=1/100)
+    regression(m,to=timecause1,from=X) <- beta1
+    if (competing.risks[[1]]) regression(m,to=timecause2,from=X) <- beta2
+    regression(m,to=censtime,from=X) <- gamma
+    ## quadratic terms
+    for (v in q.vars){
+        form.v <- formula(paste("quad.",v,"~",v,sep=""))
+        transform(m,form.v) <- function(x)x^2
+    }
+    regression(m,to=timecause1,from=paste("quad.",q.vars,sep="")) <- q.beta1
+    if (competing.risks[[1]]) regression(m,to=timecause2,from=paste("quad.",q.vars,sep="")) <- q.beta2
[…]

### 2023-11-06 empirical-study/functions/run_simulation.R

@@ -1,0 +1,51 @@
+### run_simulation.R --- 
+#----------------------------------------------------------------------
+## Author: Thomas Alexander Gerds
+## Created: Jul 18 2023 (10:49) 
+## Version: 
+## Last-Updated: Jul 18 2023 (15:22) 
+##           By: Thomas Alexander Gerds
+##     Update #: 10
+#----------------------------------------------------------------------
+## 
+### Commentary: 
+## 
+### Change Log:
+#----------------------------------------------------------------------
+## 
+### Code:
+run_simulation <- function(n,nlarge = 100000,time_horizon){
+    learn_data = draw_data(n = n)
+    outcome_models <- list(cox0 = coxph(Surv(time,event = event)~1, data=learn_data, x = TRUE, y = TRUE),
+                           cox1 = coxph(Surv(time,event = event)~X1+X2, data=learn_data, x = TRUE, y = TRUE),
+                           cox2 = coxph(Surv(time,event = event)~X6+X7, data=learn_data, x = TRUE, y = TRUE),
+                           ## cox_dgm = coxph(Surv(time,event = event)~ X1 + X2 + X5 + X6 + X7 + X8 + X11 + X12 + quad.X3 + quad.X4 + quad.X5 + quad.X6,data = learn_data,x = TRUE,y = TRUE),
+                           rf = ranger(Surv(time, event = event) ~ X1+X2+X3+X4+X5+X6+X7+X8+X9+X10+X11+X12, data = learn_data,num.tree = 50))
+    cens_models <- list(cox0 = coxph(Surv(time,event = event)~1, data=learn_data, x = TRUE, y = TRUE),
+                        cox1 = coxph(Surv(time,event = cens_status)~X1+X2, data=learn_data, x = TRUE, y = TRUE),
+                        cox2 = coxph(Surv(time,event = cens_status)~X6+X7, data=learn_data, x = TRUE, y = TRUE),
+                        ## cox_dgm = coxph(Surv(time,event = cens_status)~ X1 + X2 + X5 + X6 + X7 + X8 + X11 + X12 + quad.X3 + quad.X4 + quad.X5 + quad.X6,data = learn_data,x = TRUE,y = TRUE),
+                        rf = ranger(Surv(time, event = cens_status) ~ X1+X2+X6+X7, data = learn_data,num.tree = 50))
+    sl = statelearner(learners = list(state = outcome_models,censoring = cens_models),data = learn_data,B = 4,times = time_horizon)
[…]

### 2023-11-06 empirical-study/functions/simZelefsky.R

@@ -1,0 +1,69 @@
+### zelefsky-empirical-simulate-data.R ---
+#----------------------------------------------------------------------
+## Author: Thomas Alexander Gerds
+## Created: Nov 9 2018 (9:49)
+## Version:
+#----------------------------------------------------------------------
+##
+### Commentary: Function simulation data similar to the zelefsky data
+##
+### Change Log:
+#----------------------------------------------------------------------
+##
+### Code:
+simZelefsky <- function(n,censoring = TRUE,simulation_input, ...){
+    m <- lvm()
+    input.surv = simulation_input$surv
+    input.cens = simulation_input$cens
+    zelefsky = simulation_input$zelefsky
+    distribution(m,~time.recur) <- coxWeibull.lvm(scale=exp(-input.surv$coef["(Intercept)"]/input.surv$scale),shape=1/input.surv$scale)
+    distribution(m,~time.cens) <- coxWeibull.lvm(scale=exp(-input.cens$coef["(Intercept)"]/input.cens$scale),shape=(1/input.cens$scale))
+    ## put code to generate covariates: psa, stage, ggtot, dose100, hormones
+    ## alike the real data distributions, maybe even with
+    ## regression statements between them to make them correlated
+
+    distribution(m,~logPSA) <- normal.lvm(mean=mean(log(zelefsky$psa)),sd=sd(log(zelefsky$psa)))
+    m <- categorical(m,~stage, K=6, p=c(table(zelefsky$stage)[-6]/nrow(zelefsky)), labels=c("T1c","T2a","T2b","T2c","T3ab","T3c"))
+    transform(m,stageT2a~stage) <- function(x){1*(x=="T2a")}
+    transform(m,stageT2b~stage) <- function(x){1*(x=="T2b")}
+    transform(m,stageT2c~stage) <- function(x){1*(x=="T2c")}
+    transform(m,stageT3ab~stage) <- function(x){1*(x=="T3ab")}
+    transform(m,stageT3c~stage) <- function(x){1*(x=="T3c")}
+    distribution(m,~ggtot) <- normal.lvm(mean=mean(zelefsky$ggtot),sd=sd(zelefsky$ggtot))
+    distribution(m,~sDose) <- normal.lvm(mean=mean(zelefsky$sDose),sd=sd(zelefsky$sDose))    
[…]

### 2023-11-06 empirical-study/functions/statelearner-v2.R

@@ -1,0 +1,263 @@
+### statelearner-v2.R --- 
+#----------------------------------------------------------------------
+## Author: Anders Munch
+## Created: Aug 18 2023 (10:26) 
+## Version: 
+## Last-Updated: Oct 13 2023 (16:45) 
+##           By: Anders Munch
+##     Update #: 363
+#----------------------------------------------------------------------
+## 
+### Commentary: 
+## 
+### Change Log:
+#----------------------------------------------------------------------
+## 
+### Code:
+
+
+## Set data on canonical form
+form_data <- function(data, max_time = Inf, time = "time", status = "status", cause_codes = c("1" = 1, "2" = 2, "c" = 0), vars = NULL){
+    if(!is.null(vars))
+        wd = copy(data)[, c(time, status, vars), with = FALSE]
+    else
+        wd = copy(data)
+    setnames(wd, old = c(time, status), new = c("time", "status"))
+    ## Make cause of interest = 1, everything else = 2
+    wd[, cause1 := 1+1*(status != cause_codes[["1"]] & time <= max_time)]
+    wd[, censor := 1+1*(status != cause_codes[["c"]] & time <= max_time)]
+    if(wd[, length(unique(status))>2])
+        wd[, cause2 := 1+1*(status != cause_codes[["2"]] & time <= max_time)]
+    wd[, status := NULL]
+    return(wd[])
+}
+## Fit model on canonical data form
+fit_cause_model <- function(model, data, cause, x_form = NULL, ...){
+    cause_names = c("cause1", "cause2", "censor")
+    cause_names = cause_names[cause_names %in% names(data)]
+    stopifnot(cause %in% cause_names)
+    wd = copy(data)[, -cause_names[cause_names != cause], with = FALSE]
+    setnames(wd, old = cause, new = "event")
+    ## not so elegant
+    if(is.null(x_form))
+        form = Surv(time, event) ~ .
+    else
+        form = update(x_form, Surv(time, event) ~ .)
+    if(model == "cox"){
+        wd[, event := 1*(event == 1)] ## To make this cause of interest
+        out = coxph(form, data = wd,x = TRUE,y = TRUE, ...)
+    }
+    if(model == "GLMnet"){
+        wd[, event := 1*(event == 1)] ## To make this cause of interest
+        out = GLMnet(form, data = wd, ...)
+    }
+    if(model == "rfsrc"){
+        out = rfsrc(form, data = wd, ...)
+    }
[…]

### 2023-11-06 empirical-study/functions/zelefsky-data-learners.R

@@ -1,0 +1,275 @@
+### zelefsky-data-learners.R --- 
+#----------------------------------------------------------------------
+## Author: Anders Munch
+## Created: Jul 23 2023 (14:00) 
+## Version: 
+## Last-Updated: Aug  2 2023 (15:52) 
+##           By: Anders Munch
+##     Update #: 135
+#----------------------------------------------------------------------
+## 
+### Commentary: 
+## 
+### Change Log:
+#----------------------------------------------------------------------
+## 
+### Code:
+library(data.table)
+library(prodlim)
+library(riskRegression)
+library(lava)
+library(survival)
+library(randomForestSRC)
+source("./functions/simZelefsky.R")
+sim_learners_zelefsky <- function(ns,simulation_input, eval_time = 36, B = 1, split.method = "cv5"){
+    do.call(rbind, lapply(ns, function(nn){
+        sim_dt = simZelefsky(n = nn,censoring = TRUE,simulation_input = simulation_input)
+        sim_dt[, ":="(time = dmos, event = status)]
+        out_mms = list(km = coxph(Surv(time,event)~1, data=sim_dt, x = TRUE, y = TRUE),
+                       cox_full = coxph(Surv(time,event)~logPSA+stage+ggtot+sDose+hormones, data=sim_dt, x = TRUE, y = TRUE),
+                       cox_int = coxph(Surv(time,event)~pspline(logPSA) + sDose + stage + ggtot+hormones, data=sim_dt, x = TRUE, y = TRUE),
+                       cox_spline = coxph(Surv(time,event)~pspline(logPSA)+stage+ggtot+pspline(sDose)+hormones, data=sim_dt, x = TRUE, y = TRUE),
+                       cox_small = coxph(Surv(time,event)~logPSA+sDose+stage, data=sim_dt, x = TRUE, y = TRUE)
+                       )
+        cens_mms = list(km = coxph(Surv(time,event==0)~1, data=sim_dt, x = TRUE, y = TRUE),
+                        cox_full = coxph(Surv(time,event==0)~logPSA+stage+ggtot+sDose+hormones, data=sim_dt, x = TRUE, y = TRUE),
+                        cox_int = coxph(Surv(time,event==0)~pspline(logPSA) + sDose + stage + ggtot+hormones, data=sim_dt, x = TRUE, y = TRUE),
[…]

### 2023-11-06 empirical-study/ipcw-fail-example.R

@@ -1,0 +1,72 @@
+### ipcw-fail-example.R --- 
+#----------------------------------------------------------------------
+## Author: Anders Munch
+## Created: Jul 21 2023 (16:27) 
+## Version: 
+## Last-Updated: Jul 23 2023 (11:46) 
+##           By: Anders Munch
+##     Update #: 34
+#----------------------------------------------------------------------
+## 
+### Commentary: 
+## 
+### Change Log:
+#----------------------------------------------------------------------
+## 
+### Code:
+library(data.table)
+library(prodlim)
+library(riskRegression)
+library(targets)
+library(lava)
+library(data.table)
+library(riskRegression)
+library(survival)
+library(ranger)
+
+try(setwd("~/research/SuperVision/Anders/survival-loss/statelearner/empirical-study/"))
+try(setwd("~/Documents/phd/survival-loss-function/statelearner/empirical-study/"))
+tar_source("functions")
+
+## Trying with binary
+sim2 <- function(n){
+    m <- lvm()
+    formula_event <- ~f(A,-.8)
+    formula_cens <- ~f(A,.9)
+    event_scale <- 1/500
+    cens_scale <- 1/500
+    lava::distribution(m,~A) <- lava::binomial.lvm(p = .3)
+    lava::distribution(m,~censtime) <- lava::coxWeibull.lvm(scale=event_scale)
+    lava::distribution(m,~eventtime) <- lava::coxWeibull.lvm(scale=cens_scale)
+    m <- lava::eventTime(m, time ~ min(eventtime = 1, censtime = 0), "event")
+    lava::regression(m) <- stats::update(formula_event, "eventtime~.")
+    lava::regression(m) <- stats::update(formula_cens, "censtime~.")
+    out <- setDT(sim(m,n))
+    out[,dummy := 1]
+    return(out[])
+}
+
+eval_time <- 20
+dlarge <- sim2(1000)
+par(mfrow = c(1,2))
+plot(prodlim(Hist(time,event)~A, data = dlarge), xlim = c(0,eval_time), type = "risk")
+plot(prodlim(Hist(time,event)~A, data = dlarge, reverse = TRUE), xlim = c(0,eval_time))
+plot(prodlim(Hist(time,event)~1, data = dlarge, reverse = TRUE), xlim = c(0,eval_time), add = TRUE, col = "red")
+par(mfrow = c(1,1))
+mms <- list(dgm = coxph(Surv(time, event)~strata(A), data = dlarge, x = TRUE, y = TRUE),
[…]

### 2023-11-06 empirical-study/run.R

@@ -1,0 +1,200 @@
+### run.R --- 
+#----------------------------------------------------------------------
+## Author: Thomas Alexander Gerds
+## Created: Jul 11 2023 (14:07) 
+## Version: 
+## Last-Updated: Jul 23 2023 (19:48) 
+##           By: Anders Munch
+##     Update #: 173
+#----------------------------------------------------------------------
+## 
+### Commentary: 
+## 
+### Change Log:
+#----------------------------------------------------------------------
+## 
+### Code:
+## library(here)
+try(setwd("~/research/SuperVision/Anders/survival-loss/statelearner/empirical-study/"))
+try(setwd("/home/amnudn/Documents/phd/survival-loss-function/statelearner/empirical-study"))
+library(targets)
+library(lava)
+library(data.table)
+library(riskRegression)
+library(survival)
+library(ranger)
+library(ggplot2)
+tar_make()
+
+## IPCW fail example:
+tar_load(ipcw_fail)
+ipcw_fail[loss_type == "sl-brier", learner := "statelearner"]
+ipcw_fail[loss_type == "brier", learner := paste0("IPCW(", cens_model , ")")]
+winners <- rbind(ipcw_fail[!is.na(loss), .SD[min(loss) == loss, .(winner = out_model[1], oracle_loss = oracle_loss[1])], .(learner, n, n_sim)],
+                 ipcw_fail[learner == "IPCW(dgm)", .SD[min(oracle_loss) == oracle_loss, .(winner = out_model[1], oracle_loss = oracle_loss[1])], .(learner, n, n_sim)][, learner := "oracle"])
+winners_summ <- winners[, .(correct = mean(winner == "dgm")), .(learner,n)]
+
+## Compare oracle risk
+ggplot(winners[,.(oracle_risk = 100*mean(oracle_loss)), .(learner, n)], aes(x = n, y = oracle_risk, col = learner)) +
+    geom_line() + geom_point() + 
+    theme_bw()
+
+## See how many times we get things right
+ggplot(winners_summ, aes(x = n, y = correct, col = learner)) +
+    geom_line() + geom_point() + 
+    theme_bw() + 
+    ylim(c(0,1))
+
+
+## Simulate data based on zelefsky data
+tar_load(sim_zel_learners)
[…]

### 2023-11-06 empirical-study/sandbox/another.R

@@ -1,0 +1,51 @@
+library(riskRegression)
+library(survival)
+library(lava)
+library(data.table)
+library(prodlim)
+try(setwd("~/research/SuperVision/Anders/survival-loss/statelearner/empirical-study/"))
+tar_source("functions")
+tar_source("~/Documents/phd/survival-loss-function/statelearner/empirical-study/functions")
+
+m <- lvm()
+# B discriminates only when A == 1
+formula_event <- ~f(A,0)+f(B,0)+f(C,0)+f(AB,-1.5)+f(AC,-1.3)
+formula_cens <- ~f(A,0)+f(B,2)+f(C,0)
+event_scale <- 1/500
+cens_scale <- 1/500
+lava::distribution(m,~A) <- lava::binomial.lvm(p = .4)
+lava::distribution(m,~B) <- lava::binomial.lvm(p = .3)
+lava::distribution(m,~C) <- lava::binomial.lvm(p = .35)
+transform(m,AB~A+B) <- prod
+transform(m,AC~A+C) <- prod
+lava::distribution(m,~censtime) <- lava::coxWeibull.lvm(scale=event_scale)
+lava::distribution(m,~eventtime) <- lava::coxWeibull.lvm(scale=cens_scale)
+m <- lava::eventTime(m, time ~ min(eventtime = 1, censtime = 0), "event")
+lava::regression(m) <- stats::update(formula_event, "eventtime~.")
+lava::regression(m) <- stats::update(formula_cens, "censtime~.")
+d <- setDT(sim(m,10000))
+d[,table(event)]
+dlarge <- setDT(sim(m,10000))
+dlarge[,dummy := 1]
+## pdf("~/tmp/test.pdf",width = 12,height = 6)
+pdf("~/tmp/test.pdf",width = 10,height = 10)
+par(mfrow = c(2,2))
+plot(prodlim(Hist(time,event)~A+B,data = d),xlim = c(0,20),newdata = expand.grid(A =c(0,1),B =c(0,1)),type = "risk",plot.main = "Event probability",confint = FALSE,legend.cex = 0.5)
+plot(prodlim(Hist(time,event)~A+C,data = d),xlim = c(0,20),newdata = expand.grid(A = c(0,1),C = c(0,1)),type = "risk",plot.main = "Event probability",confint = FALSE,legend.cex = 0.5)
[…]

### 2023-11-06 figures/figure-multi-state-process.R

@@ -1,0 +1,26 @@
+### figure-multi-state-process.R --- 
+#----------------------------------------------------------------------
+## Author: Thomas Alexander Gerds
+## Created: Jul 27 2023 (10:04) 
+## Version: 
+## Last-Updated: Oct  9 2023 (18:22) 
+##           By: Anders Munch
+##     Update #: 2
+#----------------------------------------------------------------------
+## 
+### Commentary: 
+## 
+### Change Log:
+#----------------------------------------------------------------------
+## 
+### Code:
+library(prodlim)
+nTrans <- 3
+stateLabels = c("Initial","Cause 1", "Cause 2", "Censored")
+crHist <- Hist(time = 1:nTrans, event = list(from = rep("1", nTrans), to = stateLabels[-1]))
+pdf("~/research/SuperVision/Anders/survival-loss/statelearner/figure-multi-state-process.pdf")
+plot(crHist,stateLabels = c("Initial","Cause 1", "Cause 2", "Censored"),arrowLabels = FALSE)
+dev.off()
+
+######################################################################
+### figure-multi-state-process.R ends here

### 2023-11-21 experiments/tmp-summarize-results.tex
simulation studies

% Created 2023-11-21 Tue 08:42
% Intended LaTeX compiler: pdflatex
\documentclass[a4paper,danish]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{capt-of}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{color}
\usepackage{amsmath}
\usepackage{array}
\usepackage[T1]{fontenc}
\usepackage{natbib}
\usepackage[margin=4cm]{geometry}
\usepackage{dsfont, pgfpages, tikz,amssymb, amsmath,xcolor, caption, subcaption}
\input{./latex-settings/standard-settings.tex}
\input{./latex-settings/standard-commands.tex}
\input{./latex-settings/org-settings.tex}
\input{./latex-settings/title-compact.tex}
\author{Anders Munch}
\date{\today}
\title{Summary of some simulation results}
\begin{document}

\maketitle
\section{Super learners}
\label{sec:orgd3f81f9}
In all simulation studies, we compare five super learners, which are listed
below. To evaluate performance super learner, we use an independent data set of
10.000 uncensored samples and calculate the integrated Brier score in this data
set. All results are based on 500 simulated data sets.

\begin{enumerate}
\item The state learner (referred to as \texttt{statelearner}).
\item The super learner proposed in \citep{westling2021inference} (referred to as
\texttt{survSL}).
\item A super learner based on the estimated integrated Brier score, where the
censoring mechanism is estimated with the Kaplan-Meier estimator (referred to
as \texttt{ipcw\_km}).
\item A super learner based on the estimated integrated Brier score, where the
censoring mechanism is estimated with a Cox model that includes all available
covariates as main effects (referred to as \texttt{ipcw\_cox}).
\item The (discrete) oracle super learner, which picks the model that minimizes the
Brier score in the independent data of 10.000 uncensored samples (referred to
as \texttt{oracle}).
\end{enumerate}

Only the super learners 1., 2., and 3. provides estimates of the censoring distribution.

\section{Zelefsky based simulation}
\label{sec:orgf2b767a}
We generate data in four different ways:

\begin{enumerate}
\item Data as generated in \citep{gerds2013estimating} (referred to as \texttt{original}).
\item As in 1., but where censoring in completely independent of covariates
(referred to as \texttt{indep\_cens}).
\item Same censoring mechanism as in 1., but where the outcome depend only on one
of the covariates (referred to as \texttt{simple\_effect})
\item As in 1., but we add 5 independent standard Guassian covariates with no
effect on neither outcome nor censoring (referred to as \texttt{noise}).
\end{enumerate}

\clearpage

\subsection{Kaplan-Meier, Cox, and random forest}
\label{sec:orgdf8f8f8}
In this setting, we include the following learners in all libraries:

\begin{itemize}
\item The Kaplan-Meier estimator
\item A Cox model with main effects
\item A random forest based on 50 trees
\end{itemize}

\begin{figure}[htbp]
\centering
\includegraphics[width=1\linewidth]{/tmp/babel-IUB42s/figure-EtEIk3.pdf}
Zelefsky simulation setting using library consisting of Kaplan-Meier, Cox model, and random forests
\end{figure}

\clearpage

\subsection{Add LASSO}
\label{sec:org25907ea}
In this setting we add a learner to all libraries, so that all libraries include
the learners:

\begin{itemize}
\item The Kaplan-Meier estimator
\item A Cox model with main effects
\item A random forest based on 50 trees
\item A penalized Cox model with main effects, where the \(\|\blank\|_1\) penalty
(LASSO) is used and the penalty parameter is selected using cross-validation
based on Cox' partial likelihood
\end{itemize}

\begin{figure}[htbp]
\centering
\includegraphics[width=1\linewidth]{/tmp/babel-IUB42s/figure-Ylaz8E.pdf}
Zelefsky simulation setting including LASSO into the library
\end{figure}

\clearpage

\section{Effect of number of variables}
\label{sec:orged3af38}
We generate data in three different way:

\begin{enumerate}
\item Outcome and censoring depends on one binary covariate (\(X_1\)). Another
continuous covariates (\(X_2\)) that is correlated with \(X_1\) is generated.
\item Same as in 1., but we also add 4 independent Gaussian covariates (\(X_3,
   \dots X_6\)).
\item Same as in 1., but we also add 9 independent Gaussian covariates (\(X_3,
   \dots X_{11}\)).
\end{enumerate}

\subsection{Kaplan-Meier, Cox, and random forest}
\label{sec:org493c472}
In this setting, we include the following learners in all libraries:

\begin{itemize}
\item The Kaplan-Meier estimator
\item A Cox model with main effects
\item A random forest based on 50 trees
\end{itemize}

\begin{figure}[htbp]
\centering
\includegraphics[width=1\linewidth]{/tmp/babel-IUB42s/figure-bpNXlw.pdf}
Zelefsky simulation setting using library consisting of Kaplan-Meier, Cox model, and random forests
\end{figure}

\clearpage

\subsection{Add LASSO}
\label{sec:org8fb1981}
In this setting we add a learner to all libraries, so that all libraries include
the learners:

\begin{itemize}
\item The Kaplan-Meier estimator
\item A Cox model with main effects
\item A random forest based on 50 trees
\item A penalized Cox model with main effects, where the \(\|\blank\|_1\) penalty
(LASSO) is used and the penalty parameter is selected using cross-validation
based on Cox' partial likelihood
\end{itemize}

\begin{figure}[htbp]
\centering
\includegraphics[width=1\linewidth]{/tmp/babel-IUB42s/figure-p5pluf.pdf}
Zelefsky simulation setting including LASSO into the library
\end{figure}

\section{References}
\label{sec:org92fb677}
\renewcommand{\section}[2]{} 
\bibliography{./latex-settings/default-bib.bib}
\end{document}

### 2023-11-21 tex/state-learner-pub-version.tex
simulation studies

@@
   \centering
-  \includegraphics[width=.5\textwidth]{./figures/figure-multi-state-process.pdf}
+  \includegraphics[width=.5\textwidth]{../figures/figure-multi-state-process.pdf}
   \caption{Illustration of the multi-state process \(\eta\) used by
@@
 We can make the following asymptotic expansion of the one-step estimator
-\citep{pfanzagl1985contributions,fisher2021visually,kennedy2022semiparametric},
+\citep{pfanzagl1985contributions,van2003unified,fisher2021visually,kennedy2022semiparametric,rytgaard2022continuous},
 \begin{equation*}
@@
 where the remainder term is typically such that
-\citep{van2003unified,rytgaard2022continuous}
 \begin{equation}
@@
 $\Lambda_2$, and $\Gamma$ are consistently estimated at rate
-\( \smallO_P{(n^{-1/4})} \) then
+\( \smallO_P{(n^{-1/4})} \), then
 \begin{equation}
@@
 \end{equation}
-The efficient influence function for the parameter $\tilde{\Psi}_{\tau}$
-depends on the set \( (\Lambda_1, \Lambda_2, \Gamma, \pi) \) of nuisance
-parameters, and is given by \fxnote*{other refs?}{\citep{rytgaard2022targeted}}
+The efficient influence function for the parameter $\tilde{\Psi}_{\tau}$ depends
+on the set \( (\Lambda_1, \Lambda_2, \Gamma, \pi) \) of nuisance parameters.
+% \begin{equation}
+%   \label{eq:5}
+%   \begin{split}
+%     & \psi_{\tau}(O; \Lambda_1, \Lambda_2, \Gamma, \pi)
+%     \\
+%     & =
+%       \sum_{l=1}^{2} \int_0^{\tau}
+%       \left[
+%       h_{\tau,l,1}(t, O  ; \Lambda_1, \Lambda_2, \Gamma, \pi)
+%       - h_{\tau,l,0}(t, O  ; \Lambda_1, \Lambda_2, \Gamma, \pi)
+%       \right]
+%     \\
+%     & \quad\qquad  \qquad
+%       \times
+%       \left(
+%       N_l(\diff t) - \1{\{\tilde{T} \geq t\}}\Lambda_l(\diff t \mid A, W)
+%       \right)
+%     \\
+%     & \quad
+%       -\sum_{a=0}^{1} 
+%       \left\{
+%       (-1)^a
+%       \int_0^{\tau}
+%       e^{-\Lambda_1(s- \mid W, a)-\Lambda_2(s- \mid W, a)}  \Lambda_1(\diff s \mid
+%       W, a)
+%       \right\}
+%       - \tilde{\Psi}_{t}^0(\Lambda_1, \Lambda_2, \mu),
+%     % & \quad+ \int_0^{\tau} e^{-\Lambda_1(s- \mid W, 1)-\Lambda_2(s- \mid W,
+%     % 1)} \Lambda_1(\diff s \mid W, 1)
+%     % \\
+%     % & \quad - \int_0^{\tau} e^{-\Lambda_1(s- \mid W, 0)-\Lambda_2(s- \mid W,
+%     % 0)} \Lambda_1(\diff s \mid W, 0)
+%   \end{split}
+% \end{equation}
+% where
+% \begin{align*}
+%   h_{\tau,l,a}(t, O  ; \Lambda_1, \Lambda_2, \Gamma, \pi)
+%   =&  \frac{\1{\{A=a\}} e^{\Gamma(t-\mid A, W)}}{\pi(W)^{a}(1-\pi(W))^{1-a}}
+%   \\
+%    & \quad \times
+%      \left(
+%      1- \frac{\int_t^{\tau} e^{-\Lambda_1(s- \mid A, W)-\Lambda_2(s- \mid A, W)}
+%      \Lambda_1(\diff s \mid A, W)}
+%      {e^{-\Lambda_1(t \mid A, W)-\Lambda_2(t \mid A, W)}}
+%      \right)^{\1{\{l=1\}}}
+%   \\
+%    & \quad\times
+%      \left(
+%      - \frac{\int_t^{\tau} e^{-\Lambda_1(s- \mid A, W)-\Lambda_2(s- \mid A, W)}
+%      \Lambda_1(\diff s \mid A, W)}
+%      {e^{-\Lambda_1(t \mid A, W)-\Lambda_2(t \mid A, W)}}
+%      \right)^{\1{\{l=2\}}}
+% \end{align*}
+Define the quantities
+\begin{align*}
+  \omega_a(A,W; \pi)
+  &=  \frac{% (-1)^{a+1}
+    \1{\{A=a\}}}{\pi(W)^{a}(1-\pi(W))^{1-a}},
+  \\
+  g(t, A, W; \Lambda_1, \Lambda_2)
+  & = \int_0^{t}
+    e^{-\Lambda_1(s- \mid W, A)-\Lambda_2(s- \mid W, A)}  \Lambda_1(\diff s \mid
+    W, A),
+  \\  
+  M_j(\diff t \mid A, W;  \Lambda_j  )
+  & = N_j(\diff t) -
+    \1{\{\tilde{T} \geq t\}} \Lambda_j(\diff t \mid W, A),
+    \quad j \in \{1,2\},
+  \intertext{and}
+  M(\diff t \mid A, W;  \Lambda_1, \Lambda_2  )
+  & = M_1(\diff t \mid A, W;  \Lambda_1  ) +
+    M_2(\diff t \mid A, W;  \Lambda_2  ).
+\end{align*}
+The efficient influence function can then be written as
+\citep{van2003unified,jewell2007non,rytgaard2022targeted},
+\begin{equation*}
+  \psi_{\tau}(O; \Lambda_1, \Lambda_2, \Gamma, \pi)
+  = \psi_{\tau}^1(O; \Lambda_1, \Lambda_2, \Gamma, \pi)
+  - \psi_{\tau}^0(O; \Lambda_1, \Lambda_2, \Gamma, \pi)
+  -\tilde{\Psi}_{t}^0(\Lambda_1, \Lambda_2, \mu),
+\end{equation*}
+where
 \begin{equation}
@@
   \begin{split}
-    & \psi_{\tau}(O; \Lambda_1, \Lambda_2, \Gamma, \pi)
+    & \psi_{\tau}^a(O; \Lambda_1, \Lambda_2, \Gamma, \pi)
     \\
     & =
-      \sum_{l=1}^{2} \int_0^{\tau}
-      \left[
-      h_{\tau,l,1}(t, O  ; \Lambda_1, \Lambda_2, \Gamma, \pi)
-      - h_{\tau,l,0}(t, O  ; \Lambda_1, \Lambda_2, \Gamma, \pi)
-      \right]
-    \\
-    & \quad\qquad  \qquad
-      \times
-      \left(
-      N_l(\diff t) - \1{\{\tilde{T} \geq t\}}\Lambda_l(\diff t \mid A, W)
-      \right)
+      \omega_a(A,W; \pi)
+      \int_0^{\tau} e^{\Gamma(t- \mid A, W)}   
+      M_1(\diff t \mid A, W; \Lambda_1)
     \\
     & \quad
-      -\sum_{a=0}^{1} 
-      \left\{
-      (-1)^a
+      -
+      \omega_a(A,W; \pi)
+      g(\tau, A, W; \Lambda_1, \Lambda_2)
       \int_0^{\tau}
-      e^{-\Lambda_1(s- \mid W, a)-\Lambda_2(s- \mid W, a)}  \Lambda_1(\diff s \mid
-      W, a)
-      \right\}
-      - \tilde{\Psi}_{t}^0(\Lambda_1, \Lambda_2, \mu),
-    % & \quad+ \int_0^{\tau} e^{-\Lambda_1(s- \mid W, 1)-\Lambda_2(s- \mid W,
-    % 1)} \Lambda_1(\diff s \mid W, 1)
-    % \\
-    % & \quad - \int_0^{\tau} e^{-\Lambda_1(s- \mid W, 0)-\Lambda_2(s- \mid W,
-    % 0)} \Lambda_1(\diff s \mid W, 0)
+      e^{[\Gamma+\Lambda_1 + \Lambda_2](t- \mid A, W)}
+      M(\diff t \mid A, W; \Lambda_1, \Lambda_2)
+    \\
+    & \quad
+      +
+      \omega_a(A,W; \pi)      
+      \int_0^{\tau}
+      g(t, A, W; \Lambda_1, \Lambda_2)
+      e^{[\Gamma+\Lambda_1 + \Lambda_2](t- \mid A, W)}
+      M(\diff t \mid A, W; \Lambda_1, \Lambda_2)
+    \\
+    & \quad + g(\tau, a, W; \Lambda_1, \Lambda_2).
   \end{split}
 \end{equation}
-where
-\begin{align*}
-  h_{\tau,l,a}(t, O  ; \Lambda_1, \Lambda_2, \Gamma, \pi)
-  =&  \frac{\1{\{A=a\}} e^{\Gamma(t-\mid A, W)}}{\pi(W)^{a}(1-\pi(W))^{1-a}}
-  \\
-   & \quad \times
-     \left(
-     1- \frac{\int_t^{\tau} e^{-\Lambda_1(s- \mid A, W)-\Lambda_2(s- \mid A, W)}
-     \Lambda_1(\diff s \mid A, W)}
-     {e^{-\Lambda_1(t \mid A, W)-\Lambda_2(t \mid A, W)}}
-     \right)^{\1{\{l=1\}}}
-  \\
-   & \quad\times
-     \left(
-     - \frac{\int_t^{\tau} e^{-\Lambda_1(s- \mid A, W)-\Lambda_2(s- \mid A, W)}
-     \Lambda_1(\diff s \mid A, W)}
-     {e^{-\Lambda_1(t \mid A, W)-\Lambda_2(t \mid A, W)}}
-     \right)^{\1{\{l=2\}}}
-\end{align*}
-Equation~(\ref{eq:1}) and~(\ref{eq:5}) then allow us to construct a one-step
+
+Equations~(\ref{eq:1}) and~(\ref{eq:5}) allow us to construct a one-step
 estimator by using the definition given in equation~(\ref{eq:one-step-def}),
@@
   \label{eq:one-step-comp-ate}
-  \tilde{\Psi}_{t, j}^0(\hat{a}_1(\data_n), \hat{a}_2(\data_n), \hat{\mu}_n(\data_n))
+  \begin{split}
+    \hat{\Psi}_{t,\text{OS}}(\data_n)
+    = &
+        \tilde{\Psi}_{t}^0(\hat{a}_1(\data_n), \hat{a}_2(\data_n),
+        \hat{\mu}_n(\data_n))
+    \\
+      &
+        +
+        \empmeas{[\psi_{\tau}(\blank; \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
+        \hat{b}_n(\data_n), \hat{\pi}_n(\data_n))]}
+    \\
+    = &
+        \empmeas{[\psi_{\tau}^1(\blank; \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
+        \hat{b}_n(\data_n), \hat{\pi}_n(\data_n))]}
+    \\
+      &
+        - \empmeas{[\psi_{\tau}^0(\blank; \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
+        \hat{b}_n(\data_n), \hat{\pi}_n(\data_n))]}
+  \end{split}
 \end{equation}
@@
 \label{sec:numer-exper}
+
+%% IPCW with potentially misspecified censoring model.
+%%
+%% Applying SL, concrete, and Westling on classical data set. (Where to put
+%% this?)
 
@@
 \begin{figure}
-  \centerline{\includegraphics[width=1\linewidth]{./figures/ipcw-fail.pdf}}
+  \centerline{\includegraphics[width=1\linewidth]{../figures/ipcw-fail.pdf}}
   \caption[]{Results for scenario~1 of the simulation study. For the learner
@@
 \begin{figure}
-  \centerline{\includegraphics[width=1\linewidth]{./figures/zelefski-sim.pdf}}
+  \centerline{\includegraphics[width=1\linewidth]{../figures/zelefski-sim.pdf}}
   \caption[]{Results for scenario~2 of the simulation study. For the learner
@@
 \end{figure}
+
+\section{Case study: Bystander efforts in out-of-hospital cardiac arrest}
+\label{sec:case-study:-byst}
 
@@
 \begin{figure}
-  \centerline{\includegraphics[width=1\linewidth]{./figures/zelefski-real-data.pdf}}
+  \centerline{\includegraphics[width=1\linewidth]{../figures/zelefski-real-data.pdf}}
   \caption[]{The results of applying the 125 combinations of learners to the
@@
 \begin{table}[ht]
-  \centering \input{./figures/zel-tab.tex}
+  \centering \input{../figures/zel-tab.tex}
   \caption{The 10 best performing models in terms of integrated Brier score. The

### 2023-11-21 R-code/functions/abs-risk-ate.R

@@ -3,9 +3,9 @@
 ## Author: Anders Munch
 ## Created: Oct 27 2023 (16:08) 
 ## Version: 
-## Last-Updated: Nov  7 2023 (11:22) 
+## Last-Updated: Nov 17 2023 (14:01) 
 ##           By: Anders Munch
-##     Update #: 426
+##     Update #: 441
 #----------------------------------------------------------------------
 ## 
 ### Commentary:
@@ -154,18 +154,22 @@ raw_os_abs_risk_ate <- function(data, t, Lambda1, Lambda2, Gamma, pi, jump_point
     }
     return(out)
 }
-os_abs_risk_ate <- function(data, eval_times, fit_1, fit_2, fit_cens, fit_treat, jump_points = data[, time],chunks = 1){
+os_abs_risk_ate <- function(data, eval_times, fit_1, fit_2, fit_cens, fit_treat, jump_points = data[, sort(unique(time))],chunks = 1){
+    ## TREATMENT VARIABLE SHOULD BE NAMED A!!!
     L1 = function(newdata,times) predictCHF(fit_1, newdata, times)
     L2 = function(newdata,times) predictCHF(fit_2, newdata, times)
     G = function(newdata,times) predictCHF(fit_cens, newdata, times)
     pi = function(newdata) predictTreat(fit_treat, newdata)
     cause_est = list(cause1 = L1, cause2 = L2)
     out = do.call(rbind, lapply(eval_times, function(tt){
-        do.call(rbind, lapply(1:length(cause_est), function(ii){
+        do.call(rbind, lapply(1:2, function(ii){
             cause_interest = names(cause_est)[ii]
             L1_ii = cause_est[[ii]]
             L2_ii = cause_est[[1+(ii %% 2)]]
-            raw0 = raw_os_abs_risk_ate(data = data,
+            cause_data = copy(data)
+            if(ii == 2)
+                cause_data[status != 0, status := 1+status %% 2]
+            raw0 = raw_os_abs_risk_ate(data = cause_data,
                                        t = tt,
                                        Lambda1 = L1_ii,
                                        Lambda2 = L2_ii,
@@ -182,11 +186,19 @@ os_abs_risk_ate <- function(data, eval_times, fit_1, fit_2, fit_cens, fit_treat,
             see_0 = with(raw0, sd(naiv0_i + W0_i*(A_i - B_i + C_i))/sqrt(nrow(data)))
[…]

### 2023-11-21 R-code/functions/construct-pred-funs.R

@@ -3,9 +3,9 @@
 ## Author: Anders Munch
 ## Created: Nov  6 2023 (12:03) 
 ## Version: 
-## Last-Updated: Nov  7 2023 (11:40) 
+## Last-Updated: Nov 17 2023 (08:37) 
 ##           By: Anders Munch
-##     Update #: 23
+##     Update #: 32
 #----------------------------------------------------------------------
 ## 
 ### Commentary:
@@ -21,6 +21,9 @@
 library(prodlim)
 library(data.table)
 library(riskRegression)
+library(survival)
+library(randomForestSRC)
+
 
 ## Cumulative hazard functions
 predictCHF <- function(object, newdata, ...){
@@ -64,6 +67,11 @@ predictTreat <- function(object, newdata, ...){
     out = matrix(predictRisk(object,newdata), ncol = 1)
     return(out)
 }
+predictRisk.SuperLearner <- function(object, newdata){
+    rel_newdata = newdata[, object$varNames, with = FALSE]
+    predict(object = object, newdata = rel_newdata, onlySL = TRUE)$pred
+}
+
 ## predictTreat <- function(object, newdata, ...){
 ##     UseMethod("predictTreat",object)
 ## }

### 2023-11-21 R-code/functions/eval-super-learners.R

@@ -1,0 +1,246 @@
+### eval-super-learners.R --- 
+#----------------------------------------------------------------------
+## Author: Anders Munch
+## Created: Nov 10 2023 (14:28) 
+## Version: 
+## Last-Updated: Nov 20 2023 (13:54) 
+##           By: Anders Munch
+##     Update #: 283
+#----------------------------------------------------------------------
+## 
+### Commentary: 
+## 
+### Change Log:
+#----------------------------------------------------------------------
+## 
+### Code:
+library(riskRegression)
+library(survSuperLearner)
+
+## Make small rfsrc sl for SurvSL... (must be a better way...)
+## survSL.rfsrc_small <- function(time, event, X, newX, new.times, obsWeights, id, ...)
+##     survSL.rfsrc(time = time, event = event, X = X, newX = newX, new.times = new.times, obsWeights = obsWeights, id = id, ntree = 50, ...)
+survSL.rfsrc_small <- function(time, event, X, newX, new.times, obsWeights, id, ...) {
+    data <- data.frame(time, event)
+    data <- cbind(data, X)
+    fit.rfsrc <- randomForestSRC::rfsrc.fast(Surv(time, event) ~ ., data=data, importance = FALSE, case.wt = obsWeights, forest = TRUE, ntree = 50, ...)
+    survs <- predict(fit.rfsrc, newdata=newX, importance='none')$survival
+    pred <- t(sapply(1:nrow(survs), function(i) {
+        stats::approx(c(0,fit.rfsrc$time.interest), c(1,survs[i,]), method='constant', xout = new.times, rule = 2)$y
+    }))
+
+  fit <- list(object = fit.rfsrc)
+  class(fit) <- c("survSL.rfsrc")
+  out <- list(pred = pred, fit = fit)
+  return(out)
+}
+predict.survSL.rfsrc_small <- function(object, newX, new.times, ...) 
+    predict.survSL.rfsrc(object = object, newX = newX, new.times = new.times, ...)
+
+survSL.elastic <- function(time, event, X, newX, new.times, obsWeights, id, ...) {
+    data <- data.table(time = time, event = event)
+    data <- cbind(data, X)
+    form0 <- formula(paste0("Surv(time, event)~", paste(names(X), collapse = "+")))    
+    ## fit.GLMnet <- GLMnet(Surv(time, event) ~ ., data=data, alpha = .5, weights = obsWeights, ...)
[…]

### 2023-11-21 R-code/functions/gonzalez-data-gen.R

@@ -1,0 +1,310 @@
+#' @title Simulation settings from Gonzalez Ginestet et al. (2019+). "Stacked IPCW Bagging bagging: a case study in the HIV care registry"
+#' @description Function that generates data of the different simulation studies
+#' presented in the accompanying paper. This function requires the
+#'   \code{gam} package to be installed.
+#' @param J number of simulated data sets
+#' @param n number of sample size
+#' @param frac.train percentange train data set. A number between 0 and 1.
+#' @param tao time point of interest
+#' @param simulation study indicator. It takes on 1 and 2. 
+#' @param scenario scenario indicator. It takes on 1, 2, 3 and 4.
+#' @return A list with the following elements: \describe{ \item{train.data}{ simulated train data set}
+#' \item{test.data}{simulated  test data set} } and the true AUC
+#' @examples
+#' DT=datagenPaper(J = 1 , n = 1250, frac.train = 0.8, simulation=1, scenario=4)
+#' @export
+
+# scenario=1 : Independent censoring. Censoring time does not depend on any covariate.
+# scenario=2 : Independent censoring, non-informative. Censoring time depends on a
+# disjoint subset of the covariates from those associated with the failure outcome.
+# Scenario=3 : Informative, dependent censoring. Censoring time depends on the same 
+#subset of the covariates from those associated with the failure outcome
+# scenario=4 : Informative, dependent censoring. Censoring time and the competing
+#event time depend on the same subset of the covariates from those associated with the
+#failure outcome. 
+
+# g1 = scale parameter type event 1
+# g2 = shape parameter type event 1
+# k1 = scale parameter type event 2
+# k2 = shape parameter type event 2
+
+# cenper = percentage level of censoring
[…]

### 2023-11-21 R-code/functions/ipcw-fail-sim-data.R

@@ -1,0 +1,50 @@
+### ipcw-fail-sim-data.R --- 
+#----------------------------------------------------------------------
+## Author: Anders Munch
+## Created: Nov 15 2023 (16:30) 
+## Version: 
+## Last-Updated: Nov 15 2023 (16:45) 
+##           By: Anders Munch
+##     Update #: 22
+#----------------------------------------------------------------------
+## 
+### Commentary: 
+## 
+### Change Log:
+#----------------------------------------------------------------------
+## 
+### Code:
+library(data.table)
+library(lava)
+
+ipcw_fail_sim_data <- function(n, p = 0){
+    m <- lvm()
+    formula_event <- ~f(X1,-.8)
+    formula_cens <- ~f(X1,.9)
+    event_scale <- 1/500
+    cens_scale <- 1/500
+    lava::distribution(m,~X1) <- lava::binomial.lvm(p = .3)
+    lava::distribution(m,~censtime) <- lava::coxWeibull.lvm(scale=event_scale)
+    lava::distribution(m,~eventtime) <- lava::coxWeibull.lvm(scale=cens_scale)
+    m <- lava::eventTime(m, time ~ min(eventtime = 1, censtime = 0), "status")
+    lava::regression(m) <- stats::update(formula_event, "eventtime~.")
+    lava::regression(m) <- stats::update(formula_cens, "censtime~.")
+    out <- setDT(sim(m,n))
+    setnames(out, c("censtime", "eventtime"), c("cens_time", "true_time"))
+    ## add covariates
+    if(p>0){
+        X2 = data.table(X2 = rnorm(n, mean = 0.5*out[, X1]))
+        if(p>1){
+            Xp = as.data.table(matrix(rnorm(n*(p-1)), n, p-1))
+            names(Xp) = paste0("X", 1+(p:2))
+            X = cbind(Xp, X2)
+        }else{
+            X = X2
+        }
+        out = cbind(X, out)
+    }        
+    return(out[])
+}
+
+######################################################################
+### ipcw-fail-sim-data.R ends here

### 2023-11-21 R-code/functions/sim_zelefsky.R

@@ -1,0 +1,137 @@
+### sim_zelefsky.R --- 
+#----------------------------------------------------------------------
+## Author: Anders Munch
+## Created: Nov 12 2023 (15:55) 
+## Version: 
+## Last-Updated: Nov 20 2023 (09:38) 
+##           By: Anders Munch
+##     Update #: 51
+#----------------------------------------------------------------------
+## 
+### Commentary: 
+## 
+### Change Log:
+#----------------------------------------------------------------------
+## 
+### Code:
+library(survival)
+library(data.table)
+library(lava)
+
+zelefsky = setDT(get(load(here("data/zelefsky.rda"))))
+zelefsky$status=as.numeric(as.character(factor(zelefsky$recur,levels=c("No","Yes"),labels=c("0","1"))))
+zelefsky$logPSA <- log(zelefsky$psa)
+zelefsky$sDose <- as.vector(scale(zelefsky$dose))
+zelefsky$hormonesYes <- zelefsky$hormones=="Yes"
+for (v in levels(zelefsky$stage)){
+    zelefsky[,paste("stage",v,sep="")] <- as.numeric(zelefsky$stage==v)
+}
+zelefsky[]
+
+zWeib.surv <- survreg(formula = Surv(dmos, recur != "No") ~ log(psa) + stage + ggtot + sDose + hormones,data = zelefsky)
+# censoring hazards
+zWeib.cens <- survreg(formula = Surv(dmos, recur == "No") ~ log(psa) + stage + ggtot + sDose + hormones, data = zelefsky)
+zelefsky_sim_input <- list(surv = list(coef = coef(zWeib.surv),
+                                       scale = zWeib.surv$scale),
+                           cens = list(coef = coef(zWeib.cens),
+                                       scale = zWeib.cens$scale),
+                           zelefsky = zelefsky)
+
+
+simZelefsky <- function(n,censoring = TRUE,simulation_input = zelefsky_sim_input, ...){
+    m <- lvm()
+    input.surv = simulation_input$surv
+    input.cens = simulation_input$cens
+    zelefsky = simulation_input$zelefsky
[…]

### 2023-11-21 R-code/functions/statelearner.R

@@ -3,9 +3,9 @@
 ## Author: Anders Munch
 ## Created: Aug 18 2023 (10:26) 
 ## Version: 
-## Last-Updated: Nov  7 2023 (10:02) 
+## Last-Updated: Nov 20 2023 (11:06) 
 ##           By: Anders Munch
-##     Update #: 373
+##     Update #: 402
 #----------------------------------------------------------------------
 ## 
 ### Commentary: 
@@ -14,7 +14,6 @@
 #----------------------------------------------------------------------
 ## 
 ### Code:
-
 
 ## Set data on canonical form
 form_data <- function(data, max_time = Inf, time = "time", status = "status", cause_codes = c("1" = 1, "2" = 2, "c" = 0), vars = NULL){
@@ -53,6 +52,9 @@ fit_cause_model <- function(model, data, cause, x_form = NULL, ...){
     }
     if(model == "rfsrc"){
         out = rfsrc(form, data = wd, ...)
+    }
+    if(model == "rfsrc.fast"){
+        out = rfsrc.fast(form, data = wd, forest = TRUE, ...)
     }
     return(out)
 }
@@ -114,7 +116,6 @@ statelearner <- function(learners,
             super_list_ch = lapply(names(learners), function(l_name){
                 lapply(learners[[l_name]], function(mm){
                     fit = do.call(fit_cause_model,c(mm, list(data = train, cause = l_name)))
-                    ## browser()
                     ch = predictCHF(fit, newdata = test, times = eval_times)
                     return(ch)
                 })
@@ -206,7 +207,16 @@ statelearner <- function(learners,
     else
         cv_return <- cv_fits
     setkey(cv_return,loss)
-    return(cv_return)
+    ## Fit winner models
+    cause1_fit = do.call(fit_cause_model, c(learners$cause1[[ave_cv_fit[1, cause1]]], list(data = wd, cause = "cause1")))
+    censor_fit = do.call(fit_cause_model, c(learners$censor[[ave_cv_fit[1, censor]]], list(data = wd, cause = "censor")))
+    if(comp_event_present){
+        cause2_fit = do.call(fit_cause_model, c(learners$cause2[[ave_cv_fit[1, cause2]]], list(data = wd, cause = "cause2")))
+        winners = list(cause1 = cause1_fit,cause2 = cause2_fit,censor = censor_fit)
+    }else{
+        winners = list(cause1 = cause1_fit,censor = censor_fit)
+    }
+    return(list(cv_fit = cv_return, fitted_winners = winners))
 }
 
 

### 2023-11-21 R-code/functions/survival-simulator.R

@@ -1,0 +1,175 @@
+### survival-simulator.R --- 
+#----------------------------------------------------------------------
+## Author: Anders Munch
+## Created: Aug 19 2022 (13:51) 
+## Version: 
+## Last-Updated: Nov 12 2023 (14:59) 
+##           By: Anders Munch
+##     Update #: 92
+#----------------------------------------------------------------------
+## 
+### Commentary: Flexible setup for simulating from general survival models
+## 
+### Change Log:
+#----------------------------------------------------------------------
+## 
+### Code:
+library(here)
+library(data.table)
+library(prodlim)
+library(riskRegression)
+library(ggplot2)
+library(gridExtra)
+## ----------------- Main function -----------------
+## Main function to build survival simulator from smaller functions below
+build_sim <- function(outcome, cens, covars = sim_covars_truncated_norm()){
+    sim_fun = function(n){
+        V = covars(n)
+        event_time = outcome(covars = V)
+        cens_time = cens(covars = V)
+        dd = data.table(V)
+        names(dd) = paste0("X", 1:ncol(dd))
+        dd[, ":=" (true_time = event_time, cens_time = cens_time)]
+        dd[, time := pmin(true_time, cens_time)]
+        dd[, status := 1*(true_time <= time)]
+        return(dd[])
+    }
+    return(sim_fun)
+}
+## ----------------- Simulator functions -----------------
+## The actual simulators.
+## Simulate baseline covariates:
+sim_covars_truncated_norm <- function(p = 3, sd = .8, mean = 0, trunc = 1, shift = 0){
+    sim_fun = function(n){
+        V = matrix(rnorm(n*p, mean = mean, sd = sd), nrow = n, ncol = p)
+        V[abs(V)>trunc] = trunc
+        V = V+shift
+        return(V)
+    }
+    return(sim_fun)
+}
+## Functions for simulating conditional event times given covariates
+## Simulate data from a model with Weibul as baseline... Using this:
+## https://cran.r-project.org/web/packages/survMS/vignettes/how-to-simulate-survival-models.html
+sim_coxWeibull <- function(beta = 0, base_shape = 0.7, base_scale = 0.3){
+    sim_fun = function(covars){
+        nsim = nrow(covars)
+        p = ncol(covars)
+        if(p>length(beta)) ## Pad with trailing zeros
[…]

### 2023-11-21 R-code/sandbox-alt-approaches.R

@@ -1,0 +1,334 @@
+### sandbox-alt-approaches.R --- 
+#----------------------------------------------------------------------
+## Author: Anders Munch
+## Created: Nov  7 2023 (14:30) 
+## Version: 
+## Last-Updated: Nov 14 2023 (14:11) 
+##           By: Anders Munch
+##     Update #: 187
+#----------------------------------------------------------------------
+## 
+### Commentary: 
+## 
+### Change Log:
+#----------------------------------------------------------------------
+## 
+### Code:
+
+library(survSuperLearner)
+library(data.table)
+library(here)
+library(targets)
+tar_source(here("R-code/functions/"))
+
+sim_dat <- function(n){
+    X <- data.frame(X1 = rnorm(n), X2 = rbinom(n, size = 1, prob = 0.5))     
+    S0 <- function(t, x) pexp(t, rate = exp(-2 + x[,1] - x[,2] + .5 * x[,1] * x[,2]), lower.tail = FALSE)
+    T <- rexp(n, rate = exp(-2 + X[,1] - X[,2] + .5 *  X[,1] * X[,2]))    
+    G0 <- function(t, x) {
+        as.numeric(t < 15) * .9 * pexp(t, rate = exp(-2 -.5 * x[,1] - .25 * x[,2] + .5 * x[,1] * x[,2]), lower.tail=FALSE)
+    }
+    C0 <- rbinom(n, 1, .1)
+    C <- rexp(n, exp(-2 -.5 * X[,1] - .25 * X[,2] + .5 * X[,1] * X[,2]))
+    ## C[C0 == 1] <- 0 ## This makes no sense?!?
+    C[C > 15] <- 15     
+    time <- pmin(T, C)
+    event <- as.numeric(T <= C)
+    setDT(X)
+    X[, ":="(time = time, status = event, true_time = T, cens_time = C)]
+    return(X[])
+}
+
+## Westling approach
+dt0 <- sim_dat(200)
+event.SL.library <- cens.SL.library <- lapply(c("survSL.km", "survSL.coxph", "survSL.rfsrc"), function(alg) {
+    ## c(alg, "survscreen.glmnet", "survscreen.marg", "All")
+    c(alg, "All")
+})
+
+event.SL.library <- cens.SL.library <- list(
+                        km = c("survSL.km", "All"),
+                        cox = c("survSL.coxph", "All"),
+                        rf = c("survSL.rfsrc", "All"))
+
[…]

### 2023-11-21 R-code/sandbox-statelearner.R

@@ -1,11 +1,11 @@
-### sandbox-statelearner.R --- 
+o### sandbox-statelearner.R --- 
 #----------------------------------------------------------------------
 ## Author: Anders Munch
 ## Created: Nov  6 2023 (15:48) 
 ## Version: 
-## Last-Updated: Nov  7 2023 (11:45) 
+## Last-Updated: Nov 12 2023 (15:45) 
 ##           By: Anders Munch
-##     Update #: 4
+##     Update #: 7
 #----------------------------------------------------------------------
 ## 
 ### Commentary: 
@@ -93,6 +93,9 @@ treat_fit <- glm(A ~ sex + age, family = binomial, data = wd_dat)
 ## predictTreat(treat_fit, Melanoma)
 
 ## Fit with build in function:
+
+## Testing survival simulator
+
 
 
 

### 2023-11-21 data-applications/sandbox.R

@@ -1,0 +1,242 @@
+### sandbox.R --- 
+#----------------------------------------------------------------------
+## Author: Anders Munch
+## Created: Nov 16 2023 (13:40) 
+## Version: 
+## Last-Updated: Nov 18 2023 (12:43) 
+##           By: Anders Munch
+##     Update #: 230
+#----------------------------------------------------------------------
+## 
+### Commentary: 
+## 
+### Change Log:
+#----------------------------------------------------------------------
+## 
+### Code:
+library(here)
+library(targets)
+tar_source(here("R-code/functions"))
+
+library(SuperLearner)
+library(concrete)
+
+## Testing data
+library(riskRegression)
+data(Melanoma, package="riskRegression")
+setDT(Melanoma)
+
+## Make list of data sets:
+pbc <- as.data.table(survival::pbc)[!is.na(trt), ][, A := trt - 1][, c("time", "status", "A", "age", "sex", "albumin")]
+dat_list0 <- list(Melanoma=list(data = copy(Melanoma)[,!"event"][, A := as.numeric(ulcer)-1][, ulcer := NULL][],
+                                time = 5*365.25,
+                                CR=TRUE),
+                  pbc = list(data = copy(pbc),
+                             time = 1500,
+                             CR = TRUE))
+
+target_with_statelearner <- function(dl){
+    stopifnot(c("time", "status", "A") %in% names(dl$data))
+    dat_stateL = copy(dl$data)
+    covars = names(dat_stateL)[!(names(dat_stateL) %in% c("time", "status", "A"))]
+    x_form0 = formula(paste("~", paste(c(covars, "A"), collapse = "+")))
+    stateL_learners = list(
+        km = list(model = "cox", x_form = ~1),
+        km_strat = list(model = "cox", x_form = ~strata(A)),
+        ## Get stratified cox to work...
+        cox = list(model = "cox", x_form = x_form0),
+        rfsrc = list(model = "rfsrc", x_form = x_form0, ntree = 200),
+        lasso = list(model = "GLMnet", x_form = x_form0),
+        elastic = list(model = "GLMnet", x_form = x_form0, alpha = .5)
+    )
+    stateL_fit = statelearner(learners = list(cause1 = stateL_learners,
+                                              cause2 = stateL_learners,
+                                              censor = stateL_learners),
[…]

### 2023-11-21 empirical-study/functions/zelefsky-data-learners.R

@@ -3,9 +3,9 @@
 ## Author: Anders Munch
 ## Created: Jul 23 2023 (14:00) 
 ## Version: 
-## Last-Updated: Aug  2 2023 (15:52) 
+## Last-Updated: Nov 12 2023 (18:18) 
 ##           By: Anders Munch
-##     Update #: 135
+##     Update #: 136
 #----------------------------------------------------------------------
 ## 
 ### Commentary: 

### 2023-11-21 experiments/_targets.R

@@ -1,0 +1,49 @@
+### _targets.R --- 
+#----------------------------------------------------------------------
+## Author: Anders Munch
+## Created: Nov 14 2023 (09:52) 
+## Version: 
+## Last-Updated: Nov 21 2023 (08:33) 
+##           By: Anders Munch
+##     Update #: 38
+#----------------------------------------------------------------------
+## 
+### Commentary: 
+## 
+### Change Log:
+#----------------------------------------------------------------------
+## 
+### Code:
+library(targets)
+library(data.table)
+
+combine_results <- function(job_name, results_dir){
+    fls = list.files(results_dir)
+    do.call(rbind,
+            lapply(fls[grepl(paste0(job_name, "_taskID_"), fls)],
+                   function(fl) fread(paste0("./", results_dir, "/", fl))))
+}
+
+
+list(
+    tar_target(results_dir, "results", format = "file"), ## Make it look for changes to results dir
+    tar_target(zel_sim1_results, combine_results("zel_sim1", results_dir = results_dir)),
+    tar_target(zel_indep_cens_sim_results, combine_results("zel-indep-cens-sim", results_dir = results_dir)),
+    tar_target(zel_all_results, combine_results("zel-all-settings", results_dir = results_dir)),
+    tar_target(zel_all0_results, combine_results("zel-all-settings0", results_dir = results_dir)),
+    tar_target(zel_all2_results, combine_results("zel-all-settings2", results_dir = results_dir)),
+    tar_target(ipcw_fail_sim, {
+        run0 = combine_results("ipcw-fail-sim", results_dir = results_dir)
+        state_more = combine_results("ipcw-fail-sim-more-state-learners", results_dir = results_dir)
+        state_more[, SL := "statelearner_many"]
+        rbind(run0, state_more)
+    }),
+    tar_target(ipcw_fail_sim0_more, combine_results("ipcw-fail-sim-more-state-learners", results_dir = results_dir)),    
+    tar_target(ipcw_fail_sim0, combine_results("ipcw-fail-sim", results_dir = results_dir)),
+    tar_target(ipcw_fail_sim2, combine_results("ipcw-fail-sim2", results_dir = results_dir)),
+    tar_target(ipcw_fail_sim_without_rf, combine_results("ipcw-fail-sim-without-rf", results_dir = results_dir))
+)
+
+
+######################################################################
+### _targets.R ends here

### 2023-11-21 experiments/ipcw-fail-sim-more-state-learners.R

@@ -1,0 +1,76 @@
+### ipcw-fail-sim-more-state-learners.R --- 
+#----------------------------------------------------------------------
+## Author: Anders Munch
+## Created: Nov 16 2023 (12:47) 
+## Version: 
+## Last-Updated: Nov 16 2023 (12:48) 
+##           By: Anders Munch
+##     Update #: 3
+#----------------------------------------------------------------------
+## 
+### Commentary: 
+## 
+### Change Log:
+#----------------------------------------------------------------------
+## 
+### Code:
+library(data.table)
+options(rf.cores = 1) ## Keep RF on one core
+setDTthreads(1) ## Avoid DT spawning
+
+## Get input
+task_id <- as.numeric(Sys.getenv("SLURM_ARRAY_TASK_ID"))
+max_task_id <- as.numeric(Sys.getenv("SLURM_ARRAY_TASK_MAX"))
+job_name <- as.character(Sys.getenv("SLURM_JOB_NAME"))
+
+## Get seeds
+set.seed(12324)
+seeds <- sample(1:1e6, size = max_task_id, replace = FALSE)
+
+## Load libraries here:
+library(here)
+library(targets)
+tar_source(here("R-code/functions"))
+
+sim_fun <- function(seed){
+    ## Write sim setup here
+    start_t = 1
+    end_t = 20
+    time_inc = (end_t-start_t)/100
+    eval_times = seq(start_t, end_t, time_inc)
+    ns = c(300, 600, 1200, 2400)
+    ps = c(1, 5, 10)
+    set.seed(seed)
+    raw_calc = do.call(rbind, lapply(ps, function(pp){
+        dat_sim = function(n) ipcw_fail_sim_data(n = n, p = pp)
+        true_test = dat_sim(10000)[, !c("time", "status")]
+        do.call(rbind, lapply(ns, function(nn){
+            train0 <- dat_sim(nn)[, !c("true_time", "cens_time")]
+            x_form0 = formula(paste("~",paste(names(train0)[grepl("X", names(train0))], collapse = "+")))
+            out = eval_sl(train0, true_test, eval_time = eval_times,
+                          stateL_learners = list(
+                              km = list(model = "cox", x_form = ~1),
+                              cox = list(model = "cox", x_form = x_form0),
+                              lasso = list(model = "GLMnet", x_form = x_form0),
+                              elastic = list(model = "GLMnet", x_form = x_form0, alpha = .5),
[…]

### 2023-11-21 experiments/ipcw-fail-sim-without-rf.R

@@ -1,0 +1,83 @@
+### ipcw-fail-sim-without-rf.R --- 
+#----------------------------------------------------------------------
+## Author: Anders Munch
+## Created: Nov 15 2023 (20:45) 
+## Version: 
+## Last-Updated: Nov 15 2023 (20:58) 
+##           By: Anders Munch
+##     Update #: 17
+#----------------------------------------------------------------------
+## 
+### Commentary: 
+## 
+### Change Log:
+#----------------------------------------------------------------------
+## 
+### Code:
+library(data.table)
+options(rf.cores = 1) ## Keep RF on one core
+setDTthreads(1) ## Avoid DT spawning
+
+## Get input
+task_id <- as.numeric(Sys.getenv("SLURM_ARRAY_TASK_ID"))
+max_task_id <- as.numeric(Sys.getenv("SLURM_ARRAY_TASK_MAX"))
+job_name <- as.character(Sys.getenv("SLURM_JOB_NAME"))
+
+## Get seeds
+set.seed(12324)
+seeds <- sample(1:1e6, size = max_task_id, replace = FALSE)
+
+## Load libraries here:
+library(here)
+library(targets)
+tar_source(here("R-code/functions"))
+
+sim_fun <- function(seed){
+    ## Write sim setup here
+    start_t = 1
+    end_t = 20
+    time_inc = (end_t-start_t)/100
+    eval_times = seq(start_t, end_t, time_inc)
+    ns = c(300, 600, 1200, 2400)
+    ps = c(0, 1, 5, 10)
+    set.seed(seed)
+    raw_calc = do.call(rbind, lapply(ps, function(pp){
+        dat_sim = function(n) ipcw_fail_sim_data(n = n, p = pp)
+        true_test = dat_sim(10000)[, !c("time", "status")]
+        do.call(rbind, lapply(ns, function(nn){
+            train0 <- dat_sim(nn)[, !c("true_time", "cens_time")]
+            out = eval_sl(train0, true_test, eval_time = eval_times,
+                          stateL_learners = list(
+                              km = list(model = "cox", x_form = ~1),
+                              cox = list(model = "cox")),
+                          SurvSL_learners = list(
+                              c("survSL.km", "All"),
+                              c("survSL.coxph", "All")),
+                          ipcw_learners = list(
+                              km = coxph(Surv(time,status)~1, data=train0, x = TRUE, y = TRUE),
+                              cox = coxph(Surv(time,status)~., data=train0, x = TRUE, y = TRUE))
+                          )
[…]

### 2023-11-21 experiments/ipcw-fail-sim.R

@@ -1,0 +1,66 @@
+### ipcw-fail-sim.R --- 
+#----------------------------------------------------------------------
+## Author: Anders Munch
+## Created: Nov 15 2023 (16:46) 
+## Version: 
+## Last-Updated: Nov 15 2023 (17:10) 
+##           By: Anders Munch
+##     Update #: 18
+#----------------------------------------------------------------------
+## 
+### Commentary: 
+## 
+### Change Log:
+#----------------------------------------------------------------------
+## 
+### Code:
+library(data.table)
+options(rf.cores = 1) ## Keep RF on one core
+setDTthreads(1) ## Avoid DT spawning
+
+## Get input
+task_id <- as.numeric(Sys.getenv("SLURM_ARRAY_TASK_ID"))
+max_task_id <- as.numeric(Sys.getenv("SLURM_ARRAY_TASK_MAX"))
+job_name <- as.character(Sys.getenv("SLURM_JOB_NAME"))
+
+## Get seeds
+set.seed(12324)
+seeds <- sample(1:1e6, size = max_task_id, replace = FALSE)
+
+## Run sim function with current seed
+library(here)
+library(targets)
+tar_source(here("R-code/functions"))
+
+sim_fun <- function(seed){
+    start_t = 1
+    end_t = 20
+    time_inc = (end_t-start_t)/100
+    eval_times = seq(start_t, end_t, time_inc)
+    ns = c(300, 600, 1200, 2400)
+    ps = c(0, 1, 5, 10)
+    set.seed(seed)
+    raw_calc = do.call(rbind, lapply(ps, function(pp){
+        dat_sim = function(n) ipcw_fail_sim_data(n = n, p = pp)
+        true_test = dat_sim(10000)[, !c("time", "status")]
+        do.call(rbind, lapply(ns, function(nn){
+            train0 <- dat_sim(nn)[, !c("true_time", "cens_time")]
+            out = eval_sl(train0, true_test, eval_time = eval_times)
+            out[, ":="(n_obs = nn, n_covar = pp, seed = seed)]
+            return(out)
+        }))
+    }))
+    out = raw_calc[!(type == "cens" & grepl("ipcw", SL)),
+                   .(scaled_int_brier = sum(100*brier)*time_inc/end_t),
+                   .(n_obs, n_covar, type, SL, seed)]
+    return(out)
+}
+
+result <- sim_fun(seeds[task_id])
+
+# Save the results for this task as an individual file in the output folder
[…]

### 2023-11-21 experiments/ipcw-fail-sim2.R

@@ -1,0 +1,74 @@
+library(data.table)
+options(rf.cores = 1) ## Keep RF on one core
+setDTthreads(1) ## Avoid DT spawning
+
+## Get input
+task_id <- as.numeric(Sys.getenv("SLURM_ARRAY_TASK_ID"))
+max_task_id <- as.numeric(Sys.getenv("SLURM_ARRAY_TASK_MAX"))
+job_name <- as.character(Sys.getenv("SLURM_JOB_NAME"))
+
+## Get seeds
+set.seed(12324)
+seeds <- sample(1:1e6, size = max_task_id, replace = FALSE)
+
+## Run sim function with current seed
+library(here)
+library(targets)
+tar_source(here("R-code/functions"))
+
+sim_fun <- function(seed){
+    start_t = 1
+    end_t = 20
+    time_inc = (end_t-start_t)/100
+    eval_times = seq(start_t, end_t, time_inc)
+    ns = c(300, 600, 1200, 2400)
+    ps = c(1, 5, 10)
+    set.seed(seed)
+    raw_calc = do.call(rbind, lapply(ps, function(pp){
+        dat_sim = function(n) ipcw_fail_sim_data(n = n, p = pp)
+        true_test = dat_sim(10000)[, !c("time", "status")]
+        do.call(rbind, lapply(ns, function(nn){
+            train0 = dat_sim(nn)[, !c("true_time", "cens_time")]
+            x_form0 = formula(paste("~",paste(names(train0)[grepl("X", names(train0))], collapse = "+")))
+            ## Unnice hack...
+            if(pp == 0)
+                ipcw_glm0 = GLMnet(Surv(time, status)~X1, data = train0)
+            if(pp == 1)
+                ipcw_glm0 = GLMnet(Surv(time, status)~X1+X2, data = train0)
+            if(pp == 5)
+                ipcw_glm0 = GLMnet(Surv(time, status)~X1+X2+X3+X4+X5+X6, data = train0)
+            if(pp == 10)
+                ipcw_glm0 = GLMnet(Surv(time, status)~X1+X2+X3+X4+X5+X6+X7+X8+X9+X10+X11, data = train0)
+            out0 = eval_sl(train0, true_test, eval_time = eval_times,
+                           stateL_learners = list(
+                               km = list(model = "cox", x_form = ~1),
+                               cox = list(model = "cox", x_form = x_form0),
+                               lasso = list(model = "GLMnet", x_form = x_form0),
[…]

### 2023-11-21 experiments/ipcw-fail-sim3.R

@@ -1,0 +1,85 @@
+library(data.table)
+options(rf.cores = 1) ## Keep RF on one core
+setDTthreads(1) ## Avoid DT spawning
+
+## Get input
+task_id <- as.numeric(Sys.getenv("SLURM_ARRAY_TASK_ID"))
+max_task_id <- as.numeric(Sys.getenv("SLURM_ARRAY_TASK_MAX"))
+job_name <- as.character(Sys.getenv("SLURM_JOB_NAME"))
+
+## Get seeds
+set.seed(12324)
+seeds <- sample(1:1e6, size = max_task_id, replace = FALSE)
+
+## Run sim function with current seed
+library(here)
+library(targets)
+tar_source(here("R-code/functions"))
+
+sim_fun <- function(seed){
+    start_t = 1
+    end_t = 20
+    time_inc = (end_t-start_t)/100
+    eval_times = seq(start_t, end_t, time_inc)
+    ns = c(300, 600, 1200, 2400)
+    ps = c(1, 5, 10)
+    set.seed(seed)
+    raw_calc = do.call(rbind, lapply(ps, function(pp){
+        dat_sim = function(n) ipcw_fail_sim_data(n = n, p = pp)
+        true_test = dat_sim(10000)[, !c("time", "status")]
+        do.call(rbind, lapply(ns, function(nn){
+            train0 = dat_sim(nn)[, !c("true_time", "cens_time")]
+            x_form0 = formula(paste("~",paste(names(train0)[grepl("X", names(train0))], collapse = "+")))
+            ## Unnice hack...
+            if(pp == 0){
+                ipcw_lasso = GLMnet(Surv(time, status)~X1, data = train0)
+                ipcw_elastic = GLMnet(Surv(time, status)~X1, data = train0, alpha = .5)
+            }
+            if(pp == 1){
+                ipcw_lasso = GLMnet(Surv(time, status)~X1+X2, data = train0)
+                ipcw_elastic = GLMnet(Surv(time, status)~X1+X2, data = train0, alpha = .5)
+            }
+            if(pp == 5){
+                ipcw_lasso = GLMnet(Surv(time, status)~X1+X2+X3+X4+X5+X6, data = train0)
+                ipcw_elastic = GLMnet(Surv(time, status)~X1+X2+X3+X4+X5+X6, data = train0, alpha = .5)
+            }
+            if(pp == 10){
[…]

### 2023-11-21 experiments/slurm-zel_sim1.R

@@ -1,0 +1,21 @@
+library(data.table)
+options(rf.cores = 1) ## Keep RF on one core
+setDTthreads(1) ## Avoid DT spawning
+
+## Get input
+number_of_tasks <- as.numeric(Sys.getenv("SLURM_ARRAY_TASK_COUNT"))
+task_id <- as.numeric(Sys.getenv("SLURM_ARRAY_TASK_ID"))
+
+## Get seeds
+set.seed(12324)
+seeds <- sample(1:1e6, size = 1e4, replace = FALSE)
+
+## Run sim function with current seed
+source("zel-based-sim.R")
+result <- zel_sim1(seeds[task_id])
+
+# Save the results for this task as an individual file in the output folder
+## if(!dir.exists("./results")) dir.create("./results")
+fwrite(result, file = paste0('./results/zel_sim1-', sprintf("%05d", task_id), '.txt'))
+
+## TODO: make it end with -task-id

### 2023-11-21 experiments/submit_ipcw-fail-sim-more-state-learners.sh

@@ -1,0 +1,12 @@
+#!/bin/bash
+
+#SBATCH --job-name=ipcw-fail-sim-more-state-learners
+#SBATCH --array=1-500%40
+#SBATCH --output=log/stdout-%x-%a.out
+
+# Enable Additional Software
+module load gcc/11.2.0
+module load R/4.1.2
+
+# Run the job commands
+Rscript --vanilla $SLURM_JOB_NAME.R

### 2023-11-21 experiments/submit_ipcw-fail-sim-without-rf.sh

@@ -1,0 +1,12 @@
+#!/bin/bash
+
+#SBATCH --job-name=ipcw-fail-sim-without-rf
+#SBATCH --array=1-500%40
+#SBATCH --output=log/stdout-%x-%a.out
+
+# Enable Additional Software
+module load gcc/11.2.0
+module load R/4.1.2
+
+# Run the job commands
+Rscript --vanilla $SLURM_JOB_NAME.R

### 2023-11-21 experiments/submit_ipcw-fail-sim.sh

@@ -1,0 +1,12 @@
+#!/bin/bash
+
+#SBATCH --job-name=ipcw-fail-sim
+#SBATCH --array=1-500%40
+#SBATCH --output=log/stdout-%x-%a.out
+
+# Enable Additional Software
+module load gcc/11.2.0
+module load R/4.1.2
+
+# Run the job commands
+Rscript --vanilla $SLURM_JOB_NAME.R

### 2023-11-21 experiments/submit_ipcw-fail-sim2.sh

@@ -1,0 +1,12 @@
+#!/bin/bash
+
+#SBATCH --job-name=ipcw-fail-sim2
+#SBATCH --array=1-500%60
+#SBATCH --output=log/stdout-%x-%a.out
+
+# Enable Additional Software
+module load gcc/11.2.0
+module load R/4.1.2
+
+# Run the job commands
+Rscript --vanilla $SLURM_JOB_NAME.R

### 2023-11-21 experiments/submit_ipcw-fail-sim3.sh

@@ -1,0 +1,12 @@
+#!/bin/bash
+
+#SBATCH --job-name=ipcw-fail-sim3
+#SBATCH --array=1-500%80
+#SBATCH --output=log/stdout-%x-%a.out
+
+# Enable Additional Software
+module load gcc/11.2.0
+module load R/4.1.2
+
+# Run the job commands
+Rscript --vanilla $SLURM_JOB_NAME.R

### 2023-11-21 experiments/submit_zel-all-settings.sh

@@ -1,0 +1,12 @@
+#!/bin/bash
+
+#SBATCH --job-name=zel-all-settings
+#SBATCH --array=1-300%60
+#SBATCH --output=log/stdout-%x-%a.out
+
+# Enable Additional Software
+module load gcc/11.2.0
+module load R/4.1.2
+
+# Run the job commands
+Rscript --vanilla $SLURM_JOB_NAME.R

### 2023-11-21 experiments/submit_zel-all-settings0.sh

@@ -1,0 +1,12 @@
+#!/bin/bash
+
+#SBATCH --job-name=zel-all-settings0
+#SBATCH --array=1-500%50
+#SBATCH --output=log/stdout-%x-%a.out
+
+# Enable Additional Software
+module load gcc/11.2.0
+module load R/4.1.2
+
+# Run the job commands
+Rscript --vanilla $SLURM_JOB_NAME.R

### 2023-11-21 experiments/submit_zel-all-settings2.sh

@@ -1,0 +1,12 @@
+#!/bin/bash
+
+#SBATCH --job-name=zel-all-settings2
+#SBATCH --array=1-500%50
+#SBATCH --output=log/stdout-%x-%a.out
+
+# Enable Additional Software
+module load gcc/11.2.0
+module load R/4.1.2
+
+# Run the job commands
+Rscript --vanilla $SLURM_JOB_NAME.R

### 2023-11-21 experiments/submit_zel-indep-cens-sim.sh

@@ -1,0 +1,12 @@
+#!/bin/bash
+
+#SBATCH --job-name=zel-indep-cens-sim
+#SBATCH --array=1-500%40
+#SBATCH --output=log/stdout-%x-%a.out
+
+# Enable Additional Software
+module load gcc/11.2.0
+module load R/4.1.2
+
+# Run the job commands
+Rscript --vanilla zel-indep-cens-sim.R

### 2023-11-21 experiments/submit_zel_sim1.sh

@@ -1,0 +1,12 @@
+#!/bin/bash
+
+#SBATCH --job-name=zel_sim1
+#SBATCH --array=1-500%40
+#SBATCH --output=log/stdout-%x-%a.out
+
+# Enable Additional Software
+module load gcc/11.2.0
+module load R/4.1.2
+
+# Run the job commands
+Rscript --vanilla slurm-zel_sim1.R

### 2023-11-21 experiments/zel-all-settings.R

@@ -1,0 +1,104 @@
+### zel-all-settings.R --- 
+#----------------------------------------------------------------------
+## Author: Anders Munch
+## Created: Nov 20 2023 (09:39) 
+## Version: 
+## Last-Updated: Nov 20 2023 (21:16) 
+##           By: Anders Munch
+##     Update #: 75
+#----------------------------------------------------------------------
+## 
+### Commentary: 
+## 
+### Change Log:
+#----------------------------------------------------------------------
+## 
+### Code:
+library(data.table)
+options(rf.cores = 1) ## Keep RF on one core
+setDTthreads(1) ## Avoid DT spawning
+
+## Get input
+task_id <- as.numeric(Sys.getenv("SLURM_ARRAY_TASK_ID"))
+max_task_id <- as.numeric(Sys.getenv("SLURM_ARRAY_TASK_MAX"))
+job_name <- as.character(Sys.getenv("SLURM_JOB_NAME"))
+
+## Get seeds
+set.seed(284219)
+seeds <- sample(1:1e6, size = max_task_id, replace = FALSE)
+
+## Load libraries here:
+library(here)
+library(targets)
+tar_source(here("R-code/functions"))
+
+sim_fun <- function(seed){
+    ## Write sim setup here
+    start_t = 1
+    end_t = 36
+    time_inc = (end_t-start_t)/100
+    eval_times = seq(start_t, end_t, time_inc)
+    ns = c(300, 600, 1200, 2400)
+    sim_sets = list(original = simZelefsky_wrapper,
+                    indep_cens = simZelefsky_indep_cens_wrapper,
+                    simple_effect = simZelefsky_simple_effect_wrapper,
+                    noise = simZelefsky_noise_wrapper)
+    set.seed(seed)
+    raw_calc = do.call(rbind, lapply(seq_along(sim_sets), function(pp){
+        sim_set_name = names(sim_sets)[pp]
+        dat_sim = function(n) sim_sets[[pp]](n = n)
+        true_test = dat_sim(10000)[, !c("time", "status")]
+        ## full_form0 = formula(paste("Surv(time,status)~", paste(paste0("X", 1:10), collapse = "+")))
+        do.call(rbind, lapply(ns, function(nn){
+            train0 <- dat_sim(nn)[, !c("true_time", "cens_time")]
+            x_form0 = formula(paste("~",paste(names(train0)[grepl("X", names(train0))], collapse = "+")))
[…]

### 2023-11-21 experiments/zel-all-settings0.R

@@ -1,0 +1,52 @@
+library(data.table)
+options(rf.cores = 1) ## Keep RF on one core
+setDTthreads(1) ## Avoid DT spawning
+
+## Get input
+task_id <- as.numeric(Sys.getenv("SLURM_ARRAY_TASK_ID"))
+max_task_id <- as.numeric(Sys.getenv("SLURM_ARRAY_TASK_MAX"))
+job_name <- as.character(Sys.getenv("SLURM_JOB_NAME"))
+
+## Get seeds
+set.seed(284219)
+seeds <- sample(1:1e6, size = max_task_id, replace = FALSE)
+
+## Load libraries here:
+library(here)
+library(targets)
+tar_source(here("R-code/functions"))
+
+sim_fun <- function(seed){
+    ## Write sim setup here
+    start_t = 1
+    end_t = 36
+    time_inc = (end_t-start_t)/100
+    eval_times = seq(start_t, end_t, time_inc)
+    ns = c(300, 600, 1200, 2400)
+    sim_sets = list(original = simZelefsky_wrapper,
+                    indep_cens = simZelefsky_indep_cens_wrapper,
+                    simple_effect = simZelefsky_simple_effect_wrapper,
+                    noise = simZelefsky_noise_wrapper)
+    set.seed(seed)
+    raw_calc = do.call(rbind, lapply(seq_along(sim_sets), function(pp){
+        sim_set_name = names(sim_sets)[pp]
+        dat_sim = function(n) sim_sets[[pp]](n = n)
+        true_test = dat_sim(10000)[, !c("time", "status")]
+        ## full_form0 = formula(paste("Surv(time,status)~", paste(paste0("X", 1:10), collapse = "+")))
+        do.call(rbind, lapply(ns, function(nn){
+            train0 <- dat_sim(nn)[, !c("true_time", "cens_time")]
+            x_form0 = formula(paste("~",paste(names(train0)[grepl("X", names(train0))], collapse = "+")))
+            out0 = eval_sl(train0, true_test, eval_time = eval_times)
+            out0[, ":="(n_obs = nn, sim_set = sim_set_name, seed = seed)]
+            return(out0)
+        }))
+    }))
+    out = raw_calc[!(type == "cens" & grepl("ipcw", SL)),
+                   .(scaled_int_brier = sum(100*brier)*time_inc/end_t),
+                   .(n_obs, sim_set, type, SL, seed)]
+    return(out[])
+}
+
[…]

### 2023-11-21 experiments/zel-all-settings2.R

@@ -1,0 +1,75 @@
+library(data.table)
+options(rf.cores = 1) ## Keep RF on one core
+setDTthreads(1) ## Avoid DT spawning
+
+## Get input
+task_id <- as.numeric(Sys.getenv("SLURM_ARRAY_TASK_ID"))
+max_task_id <- as.numeric(Sys.getenv("SLURM_ARRAY_TASK_MAX"))
+job_name <- as.character(Sys.getenv("SLURM_JOB_NAME"))
+
+## Get seeds
+set.seed(284219)
+seeds <- sample(1:1e6, size = max_task_id, replace = FALSE)
+
+## Load libraries here:
+library(here)
+library(targets)
+tar_source(here("R-code/functions"))
+
+sim_fun <- function(seed){
+    ## Write sim setup here
+    start_t = 1
+    end_t = 36
+    time_inc = (end_t-start_t)/100
+    eval_times = seq(start_t, end_t, time_inc)
+    ns = c(300, 600, 1200, 2400)
+    sim_sets = list(original = simZelefsky_wrapper,
+                    indep_cens = simZelefsky_indep_cens_wrapper,
+                    simple_effect = simZelefsky_simple_effect_wrapper,
+                    noise = simZelefsky_noise_wrapper)
+    set.seed(seed)
+    raw_calc = do.call(rbind, lapply(seq_along(sim_sets), function(pp){
+        sim_set_name = names(sim_sets)[pp]
+        dat_sim = function(n) sim_sets[[pp]](n = n)
+        true_test = dat_sim(10000)[, !c("time", "status")]
+        ## full_form0 = formula(paste("Surv(time,status)~", paste(paste0("X", 1:10), collapse = "+")))
+        do.call(rbind, lapply(ns, function(nn){
+            train0 <- dat_sim(nn)[, !c("true_time", "cens_time")]
+            x_form0 = formula(paste("~",paste(names(train0)[grepl("X", names(train0))], collapse = "+")))
+            ## Unnice hack...
+            if(sim_set_name == "noise")
+                ipcw_glm0 = GLMnet(Surv(time, status)~X6+X7+X8+X9+X10+X1+X2+X3+X4+X5, data = train0)
+            else
+                ipcw_glm0 = GLMnet(Surv(time, status)~X1+X2+X3+X4+X5, data = train0)            
+            out0 = eval_sl(train0, true_test, eval_time = eval_times,
+                           stateL_learners = list(
+                               km = list(model = "cox", x_form = ~1),
[…]

### 2023-11-21 experiments/zel-based-sim.R

@@ -1,0 +1,40 @@
+### zel-based-sim.R --- 
+#----------------------------------------------------------------------
+## Author: Anders Munch
+## Created: Nov 14 2023 (09:06) 
+## Version: 
+## Last-Updated: Nov 15 2023 (13:04) 
+##           By: Anders Munch
+##     Update #: 36
+#----------------------------------------------------------------------
+## 
+### Commentary: 
+## 
+### Change Log:
+#----------------------------------------------------------------------
+## 
+### Code:
+library(here)
+library(data.table)
+library(targets)
+tar_source(here("R-code/functions"))
+
+sl_sim_est <- function(seed, dat_sim, eval_times, ns = c(300, 600, 1200, 2400)){
+    set.seed(seed)
+    true_test = dat_sim(10000)[, !c("time", "status")]
+    do.call(rbind, lapply(ns, function(nn){
+        train0 <- dat_sim(nn)[, !c("true_time", "cens_time")]
+        out = eval_sl(train0, true_test, eval_time = eval_times)
+        out[, ":="(n_obs = nn, seed = seed)]
+        return(out)
+    }))
+}
+
+zel_sim1 <- function(seed){
+    sl_sim_est(seed,
+               simZelefsky_wrapper,
+               eval_times = seq(1, 36, length.out = 100))
+}
+
+######################################################################
+### zel-based-sim.R ends here

### 2023-11-21 experiments/zel-indep-cens-sim.R

@@ -1,0 +1,56 @@
+### zel-indep-cens-sim.R --- 
+#----------------------------------------------------------------------
+## Author: Anders Munch
+## Created: Nov 15 2023 (13:05) 
+## Version: 
+## Last-Updated: Nov 15 2023 (13:48) 
+##           By: Anders Munch
+##     Update #: 15
+#----------------------------------------------------------------------
+## 
+### Commentary: 
+## 
+### Change Log:
+#----------------------------------------------------------------------
+## 
+### Code:
+library(data.table)
+options(rf.cores = 1) ## Keep RF on one core
+setDTthreads(1) ## Avoid DT spawning
+
+## Get input
+task_id <- as.numeric(Sys.getenv("SLURM_ARRAY_TASK_ID"))
+max_task_id <- as.numeric(Sys.getenv("SLURM_ARRAY_TASK_MAX"))
+job_name <- as.character(Sys.getenv("SLURM_JOB_NAME"))
+
+## Get seeds
+set.seed(12324)
+seeds <- sample(1:1e6, size = max_task_id, replace = FALSE)
+
+## Run sim function with current seed
+library(here)
+library(targets)
+tar_source(here("R-code/functions"))
+
+sim_fun <- function(seed){
+    dat_sim = simZelefsky_indep_cens_wrapper
+    eval_times = seq(1, 36, length.out = 100)
+    ns = c(300, 600, 1200, 2400)
+    set.seed(seed)
+    true_test = dat_sim(10000)[, !c("time", "status")]
+    do.call(rbind, lapply(ns, function(nn){
+        train0 <- dat_sim(nn)[, !c("true_time", "cens_time")]
+        out = eval_sl(train0, true_test, eval_time = eval_times)
+        out[, ":="(n_obs = nn, seed = seed)]
+        return(out)
+    }))
+}
+
+result <- sim_fun(seeds[task_id])
+
+# Save the results for this task as an individual file in the output folder
+fwrite(result, file = paste0('./results/', job_name, '_taskID_', sprintf("%05d", task_id), '.txt'))
+
+
+######################################################################
+### zel-indep-cens-sim.R ends here

### 2023-11-21 experiments/tmp-summarize-results.tex
typo

@@
-% Created 2023-11-21 Tue 08:42
+% Created 2023-11-21 Tue 15:20
 % Intended LaTeX compiler: pdflatex
@@
 \section{Super learners}
-\label{sec:orgd3f81f9}
+\label{sec:orgfb13ca2}
 In all simulation studies, we compare five super learners, which are listed
@@
 \section{Zelefsky based simulation}
-\label{sec:orgf2b767a}
+\label{sec:org1123149}
 We generate data in four different ways:
@@
 \subsection{Kaplan-Meier, Cox, and random forest}
-\label{sec:orgdf8f8f8}
+\label{sec:org0516d66}
 In this setting, we include the following learners in all libraries:
@@
 \subsection{Add LASSO}
-\label{sec:org25907ea}
+\label{sec:orgd0d0620}
 In this setting we add a learner to all libraries, so that all libraries include
@@
 \centering
-\includegraphics[width=1\linewidth]{/tmp/babel-IUB42s/figure-Ylaz8E.pdf}
+\includegraphics[width=1\linewidth]{/tmp/babel-IUB42s/figure-KMJCXV.pdf}
 Zelefsky simulation setting including LASSO into the library
@@
 \section{Effect of number of variables}
-\label{sec:orged3af38}
+\label{sec:orgf46a6e0}
 We generate data in three different way:
@@
 \item Outcome and censoring depends on one binary covariate (\(X_1\)). Another
-continuous covariates (\(X_2\)) that is correlated with \(X_1\) is generated.
+continuous covariate (\(X_2\)) that is correlated with \(X_1\) is generated.
 \item Same as in 1., but we also add 4 independent Gaussian covariates (\(X_3,
@@
 \subsection{Kaplan-Meier, Cox, and random forest}
-\label{sec:org493c472}
+\label{sec:org4b465da}
 In this setting, we include the following learners in all libraries:
@@
 \centering
-\includegraphics[width=1\linewidth]{/tmp/babel-IUB42s/figure-bpNXlw.pdf}
+\includegraphics[width=1\linewidth]{/tmp/babel-IUB42s/figure-TImPsh.pdf}
 Zelefsky simulation setting using library consisting of Kaplan-Meier, Cox model, and random forests
@@
 \subsection{Add LASSO}
-\label{sec:org8fb1981}
+\label{sec:org79abb0b}
 In this setting we add a learner to all libraries, so that all libraries include
@@
 \centering
-\includegraphics[width=1\linewidth]{/tmp/babel-IUB42s/figure-p5pluf.pdf}
+\includegraphics[width=1\linewidth]{/tmp/babel-IUB42s/figure-ozs9vG.pdf}
 Zelefsky simulation setting including LASSO into the library
@@
 \section{References}
-\label{sec:org92fb677}
+\label{sec:org53b63a4}
 \renewcommand{\section}[2]{} 

### 2023-11-21 experiments/_targets.R

@@ -3,9 +3,9 @@
 ## Author: Anders Munch
 ## Created: Nov 14 2023 (09:52) 
 ## Version: 
-## Last-Updated: Nov 21 2023 (08:33) 
+## Last-Updated: Nov 21 2023 (10:43) 
 ##           By: Anders Munch
-##     Update #: 38
+##     Update #: 39
 #----------------------------------------------------------------------
 ## 
 ### Commentary: 
@@ -41,6 +41,7 @@ list(
     tar_target(ipcw_fail_sim0_more, combine_results("ipcw-fail-sim-more-state-learners", results_dir = results_dir)),    
     tar_target(ipcw_fail_sim0, combine_results("ipcw-fail-sim", results_dir = results_dir)),
     tar_target(ipcw_fail_sim2, combine_results("ipcw-fail-sim2", results_dir = results_dir)),
+    tar_target(ipcw_fail_sim3, combine_results("ipcw-fail-sim3", results_dir = results_dir)),
     tar_target(ipcw_fail_sim_without_rf, combine_results("ipcw-fail-sim-without-rf", results_dir = results_dir))
 )
 

### 2023-11-21 experiments/ipcw-fail-sim3.R

@@ -45,7 +45,7 @@ sim_fun <- function(seed){
             }
             if(pp == 10){
                 ipcw_lasso = GLMnet(Surv(time, status)~X1+X2+X3+X4+X5+X6+X7+X8+X9+X10+X11, data = train0)
-                ipcw_elastic = GLMnet(Surv(time, status)~X1+X2+X3+X4+X5+X6+X7+X8+X9+X10+X11, data = train0, .5)
+                ipcw_elastic = GLMnet(Surv(time, status)~X1+X2+X3+X4+X5+X6+X7+X8+X9+X10+X11, data = train0, alpha = .5)
             }
             out0 = eval_sl(train0, true_test, eval_time = eval_times,
                            stateL_learners = list(

### 2023-11-28 experiments/tmp-summarize-results.tex
updated readme with simulation setup

@@
-% Created 2023-11-21 Tue 15:20
+% Created 2023-11-22 Wed 08:53
 % Intended LaTeX compiler: pdflatex
@@
 \section{Super learners}
-\label{sec:orgfb13ca2}
+\label{sec:org4d4629b}
 In all simulation studies, we compare five super learners, which are listed
@@
 \section{Zelefsky based simulation}
-\label{sec:org1123149}
+\label{sec:orgdc8a899}
 We generate data in four different ways:
@@
 \subsection{Kaplan-Meier, Cox, and random forest}
-\label{sec:org0516d66}
+\label{sec:org3ebe586}
 In this setting, we include the following learners in all libraries:
@@
 \subsection{Add LASSO}
-\label{sec:orgd0d0620}
+\label{sec:org98a9507}
 In this setting we add a learner to all libraries, so that all libraries include
@@
 \section{Effect of number of variables}
-\label{sec:orgf46a6e0}
+\label{sec:org2e42afe}
 We generate data in three different way:
@@
 \subsection{Kaplan-Meier, Cox, and random forest}
-\label{sec:org4b465da}
+\label{sec:org7f2bb4f}
 In this setting, we include the following learners in all libraries:
@@
 \subsection{Add LASSO}
-\label{sec:org79abb0b}
+\label{sec:orgf3d57c9}
 In this setting we add a learner to all libraries, so that all libraries include
@@
 
+\subsection{Add LASSO and elastic net}
+\label{sec:org062c3d9}
+In this setting we add a learner to all libraries, so that all libraries include
+the learners:
+
+\begin{itemize}
+\item The Kaplan-Meier estimator
+\item A Cox model with main effects
+\item A random forest based on 50 trees
+\item A penalized Cox model with main effects, where the \(\|\blank\|_1\) penalty
+(LASSO) is used and the penalty parameter is selected using cross-validation
+based on Cox' partial likelihood
+\item A penalized Cox model with main effects, where the \(\|\blank\|_1 +
+  \|\blank\|_2\) penalty (elastic net) is used and the penalty parameter is
+selected using cross-validation based on Cox' partial likelihood
+\end{itemize}
+
+\begin{figure}[htbp]
+\centering
+\includegraphics[width=1\linewidth]{/tmp/babel-IUB42s/figure-xTZcFV.pdf}
+Zelefsky simulation setting including LASSO into the library
+\end{figure}
+
 \section{References}
-\label{sec:org53b63a4}
+\label{sec:org3ba31f0}
 \renewcommand{\section}[2]{} 

### 2023-11-28 R-code/functions/sim_zelefsky.R

@@ -3,9 +3,9 @@
 ## Author: Anders Munch
 ## Created: Nov 12 2023 (15:55) 
 ## Version: 
-## Last-Updated: Nov 20 2023 (09:38) 
+## Last-Updated: Nov 28 2023 (09:57) 
 ##           By: Anders Munch
-##     Update #: 51
+##     Update #: 74
 #----------------------------------------------------------------------
 ## 
 ### Commentary: 
@@ -17,49 +17,61 @@
 library(survival)
 library(data.table)
 library(lava)
+library(here)
 
-zelefsky = setDT(get(load(here("data/zelefsky.rda"))))
-zelefsky$status=as.numeric(as.character(factor(zelefsky$recur,levels=c("No","Yes"),labels=c("0","1"))))
-zelefsky$logPSA <- log(zelefsky$psa)
-zelefsky$sDose <- as.vector(scale(zelefsky$dose))
-zelefsky$hormonesYes <- zelefsky$hormones=="Yes"
-for (v in levels(zelefsky$stage)){
-    zelefsky[,paste("stage",v,sep="")] <- as.numeric(zelefsky$stage==v)
+if(FALSE){
+    zelefsky = setDT(get(load(here("data/zelefsky.rda"))))
+    zelefsky$status=as.numeric(as.character(factor(zelefsky$recur,levels=c("No","Yes"),labels=c("0","1"))))
+    zelefsky$logPSA <- log(zelefsky$psa)
+    zelefsky$sDose <- as.vector(scale(zelefsky$dose))
+    zelefsky$hormonesYes <- zelefsky$hormones=="Yes"
+    for (v in levels(zelefsky$stage)){    
+        zelefsky[,paste("stage",v,sep="")] <- as.numeric(zelefsky$stage==v)
+    }
+    zelefsky[]
+
+    zWeib.surv <- survreg(formula = Surv(dmos, recur != "No") ~ log(psa) + stage + ggtot + sDose + hormones,data = zelefsky)
+    # censoring hazards
+    zWeib.cens <- survreg(formula = Surv(dmos, recur == "No") ~ log(psa) + stage + ggtot + sDose + hormones, data = zelefsky)
+    zelefsky_sim_input <- list(surv = list(coef = coef(zWeib.surv),
+                                           scale = zWeib.surv$scale),
+                               cens = list(coef = coef(zWeib.cens),
+                                           scale = zWeib.cens$scale),
[…]

### 2023-11-29 experiments/tmp-summarize-results.tex
simulation results extracted and simplified

--- 2023-11-21 experiments/tmp-summarize-results.tex
@@
-% Created 2023-11-21 Tue 15:20
+% Created 2023-11-29 Wed 11:31
 % Intended LaTeX compiler: pdflatex
@@
 \maketitle
-\section{Super learners}
-\label{sec:orgfb13ca2}
-In all simulation studies, we compare five super learners, which are listed
-below. To evaluate performance super learner, we use an independent data set of
-10.000 uncensored samples and calculate the integrated Brier score in this data
-set. All results are based on 500 simulated data sets.
+\section{The state learner vs IPCW super learners}
+\label{sec:org4c83351}
+
+We compare the state learner to two IPCW-based super learners using different
+estimators of the censoring distribution. As benchmark we use the (discrete)
+oracle. This gives the following 4 super learners.
 
@@
 \item The state learner (referred to as \texttt{statelearner}).
-\item The super learner proposed in \citep{westling2021inference} (referred to as
-\texttt{survSL}).
 \item A super learner based on the estimated integrated Brier score, where the
@@
 \item The (discrete) oracle super learner, which picks the model that minimizes the
-Brier score in the independent data of 10.000 uncensored samples (referred to
+Brier score in an independent data set of 10.000 uncensored samples (referred to
 as \texttt{oracle}).
@@
 
-Only the super learners 1., 2., and 3. provides estimates of the censoring distribution.
-
-\section{Zelefsky based simulation}
-\label{sec:org1123149}
-We generate data in four different ways:
-
+We used data generated in two different ways:
 \begin{enumerate}
@@
 (referred to as \texttt{indep\_cens}).
-\item Same censoring mechanism as in 1., but where the outcome depend only on one
-of the covariates (referred to as \texttt{simple\_effect})
-\item As in 1., but we add 5 independent standard Guassian covariates with no
-effect on neither outcome nor censoring (referred to as \texttt{noise}).
 \end{enumerate}
+
+To evaluate performance super learner, we use an independent data set of 10.000
+uncensored samples and calculate the integrated Brier score in this data set.
+All results are based on 500 simulated data sets.
+
+\begin{center}
+\includegraphics[width=.9\linewidth]{/tmp/tmp-fig-sl-plot1.pdf}
+\end{center}
 
@@
 
-\subsection{Kaplan-Meier, Cox, and random forest}
-\label{sec:org0516d66}
-In this setting, we include the following learners in all libraries:
+\section{The state learner vs SurvSL}
+\label{sec:org09707c3}
 
-\begin{itemize}
-\item The Kaplan-Meier estimator
-\item A Cox model with main effects
-\item A random forest based on 50 trees
-\end{itemize}
-
-\begin{figure}[htbp]
-\centering
-\includegraphics[width=1\linewidth]{/tmp/babel-IUB42s/figure-EtEIk3.pdf}
-Zelefsky simulation setting using library consisting of Kaplan-Meier, Cox model, and random forests
-\end{figure}
-
-\clearpage
-
-\subsection{Add LASSO}
-\label{sec:orgd0d0620}
-In this setting we add a learner to all libraries, so that all libraries include
-the learners:
-
-\begin{itemize}
-\item The Kaplan-Meier estimator
-\item A Cox model with main effects
-\item A random forest based on 50 trees
-\item A penalized Cox model with main effects, where the \(\|\blank\|_1\) penalty
-(LASSO) is used and the penalty parameter is selected using cross-validation
-based on Cox' partial likelihood
-\end{itemize}
-
-\begin{figure}[htbp]
-\centering
-\includegraphics[width=1\linewidth]{/tmp/babel-IUB42s/figure-KMJCXV.pdf}
-Zelefsky simulation setting including LASSO into the library
-\end{figure}
-
-\clearpage
-
-\section{Effect of number of variables}
-\label{sec:orgf46a6e0}
-We generate data in three different way:
+We compare the state learner to the super learner proposed by
+\cite{westling2021inference}. As benchmark we use again the (discrete) oracle.
+This gives the following 3 super learners.
 
 \begin{enumerate}
-\item Outcome and censoring depends on one binary covariate (\(X_1\)). Another
-continuous covariate (\(X_2\)) that is correlated with \(X_1\) is generated.
-\item Same as in 1., but we also add 4 independent Gaussian covariates (\(X_3,
-   \dots X_6\)).
-\item Same as in 1., but we also add 9 independent Gaussian covariates (\(X_3,
-   \dots X_{11}\)).
+\item The state learner (referred to as \texttt{statelearner}).
+\item The super learner proposed in \citep{westling2021inference} (referred to as
+\texttt{survSL}).
+\item The (discrete) oracle super learner, which picks the model that minimizes the
+Brier score in an independent data set of 10.000 uncensored samples (referred to
+as \texttt{oracle}).
 \end{enumerate}
 
-\subsection{Kaplan-Meier, Cox, and random forest}
-\label{sec:org4b465da}
-In this setting, we include the following learners in all libraries:
+We use the same data-generating mechanisms and method of evaluation as described
+in the section above.
 
-\begin{itemize}
-\item The Kaplan-Meier estimator
-\item A Cox model with main effects
-\item A random forest based on 50 trees
-\end{itemize}
-
-\begin{figure}[htbp]
-\centering
-\includegraphics[width=1\linewidth]{/tmp/babel-IUB42s/figure-TImPsh.pdf}
-Zelefsky simulation setting using library consisting of Kaplan-Meier, Cox model, and random forests
-\end{figure}
-
-\clearpage
-
-\subsection{Add LASSO}
-\label{sec:org79abb0b}
-In this setting we add a learner to all libraries, so that all libraries include
-the learners:
-
-\begin{itemize}
-\item The Kaplan-Meier estimator
-\item A Cox model with main effects
-\item A random forest based on 50 trees
-\item A penalized Cox model with main effects, where the \(\|\blank\|_1\) penalty
-(LASSO) is used and the penalty parameter is selected using cross-validation
-based on Cox' partial likelihood
-\end{itemize}
-
-\begin{figure}[htbp]
-\centering
-\includegraphics[width=1\linewidth]{/tmp/babel-IUB42s/figure-ozs9vG.pdf}
-Zelefsky simulation setting including LASSO into the library
-\end{figure}
+\begin{center}
+\includegraphics[width=.9\linewidth]{/tmp/tmp-fig-sl-plot2.pdf}
+\end{center}
 
 \section{References}
-\label{sec:org53b63a4}
+\label{sec:orgdc8dbd8}
 \renewcommand{\section}[2]{} 

### 2024-03-20 R-code/functions/abs-risk-ate.R
prostate cancer ate example and target setup

@@ -3,9 +3,9 @@
 ## Author: Anders Munch
 ## Created: Oct 27 2023 (16:08) 
 ## Version: 
-## Last-Updated: Nov 17 2023 (14:01) 
+## Last-Updated: Jan 23 2024 (15:28) 
 ##           By: Anders Munch
-##     Update #: 441
+##     Update #: 443
 #----------------------------------------------------------------------
 ## 
 ### Commentary:
@@ -204,322 +204,6 @@ os_abs_risk_ate <- function(data, eval_times, fit_1, fit_2, fit_cens, fit_treat,
     out[, ":="(lower = est-1.96*see, upper = est+1.96*see)]
     return(out[])
 }
-
-
-#### Old versions
-## naiv_est <- function(data, t, Lambda1, Lambda2, jump_points, collapse = TRUE){
-##     Lambda1_ = minus_t_fun(Lambda1,jump_points)
-##     Lambda2_ = minus_t_fun(Lambda2,jump_points)
-##     A_fun = function(newdata,times) {-(Lambda1_(newdata,times) + Lambda2_(newdata,times))}
-##     data_treat0 = copy(data)[, A := 0]
-##     data_treat1 = copy(data)[, A := 1]
-##     preds0 = internal_intergral1(A = A_fun,C = Lambda1,newdata = data_treat0,end = t,jump_points = jump_points)
-##     preds1 = internal_intergral1(A = A_fun,C = Lambda1,newdata = data_treat1,end = t,jump_points = jump_points)
-##     if(collapse)
-##         return(mean(preds1 - preds0))
-##     else
-##         return(data.table(Y1_init_est = preds1, Y1_init_est = preds0))
-## }
-
-## Q1_fun <- function(data, t, Lambda1, Lambda2, Gamma, jump_points, collapse = TRUE, chunks = 1){
-##     Q_out = numeric(nrow(data))
-##     rel_ind = data[,(status>0) & (time <= t)]
-##     if(sum(rel_ind) == 0)
-##         return(Q_out)
-##     wd = copy(data[rel_ind])
-##     Lambda1_ = minus_t_fun(Lambda1,jump_points)
-##     Lambda2_ = minus_t_fun(Lambda2,jump_points)
-##     Gamma_ = minus_t_fun(Gamma,jump_points)
-##     A_fun = function(newdata,times) {-(Lambda1_(newdata,times) + Lambda2_(newdata,times))}
[…]

### 2024-03-20 R-code/functions/at_risk_fun.R

@@ -1,35 +1,0 @@
-### at_risk_fun.R --- 
-#----------------------------------------------------------------------
-## Author: Anders Munch
-## Created: Nov  6 2023 (12:07) 
-## Version: 
-## Last-Updated: Nov  6 2023 (12:10) 
-##           By: Anders Munch
-##     Update #: 4
-#----------------------------------------------------------------------
-## 
-### Commentary:
-##
-## In newdata, use time_name to calculate who are at risk at
-## each time point in time.
-##
-## Returns matrix of dimension nrow(newdata) * length(times)
-## 
-### Change Log:
-#----------------------------------------------------------------------
-## 
-### Code:
-library(data.table)
-
-at_risk_fun <- function(newdata,times,time_name = "time"){
-    spring_times = newdata[[time_name]]
-    grid_mat = matrix(times,ncol = length(times),nrow = length(spring_times),byrow = TRUE)
-    ## event_mat = apply(grid_mat, 2, function(x) 1*(spring_times >= x))
-    event_mat = do.call(cbind, lapply(1:ncol(grid_mat), function(xx) 1*(spring_times >= grid_mat[, xx])))
-    return(event_mat)
-}
-
-
-
-######################################################################
-### at_risk_fun.R ends here

### 2024-03-20 R-code/functions/construct-pred-funs.R

@@ -3,9 +3,9 @@
 ## Author: Anders Munch
 ## Created: Nov  6 2023 (12:03) 
 ## Version: 
-## Last-Updated: Nov 17 2023 (08:37) 
+## Last-Updated: Jan 23 2024 (14:52) 
 ##           By: Anders Munch
-##     Update #: 32
+##     Update #: 33
 #----------------------------------------------------------------------
 ## 
 ### Commentary:
@@ -71,27 +71,6 @@ predictRisk.SuperLearner <- function(object, newdata){
     rel_newdata = newdata[, object$varNames, with = FALSE]
     predict(object = object, newdata = rel_newdata, onlySL = TRUE)$pred
 }
-
-## predictTreat <- function(object, newdata, ...){
-##     UseMethod("predictTreat",object)
-## }
-## predictTreat.glm <- function(object, newdata, ...){
-##     out = matrix(predict(object, newdata = newdata, type = "response"), ncol = 1)
-##     return(out)    
-## }
-
-
-## ## alternative stuff
-## construct_pred_fun.coxph <- function(model, ...){
-##     out = function(newdata,times)
-##         matrix(predictCox(model, newdata = newdata, times = times, type = "cumhazard")$cumhazard, ncol = length(times))
-##     return(out)
-## }
-## construct_pred_fun.glm <- function(model, ...){
-##     out = function(newdata)
-##         matrix(predict(model, newdata = newdata, type = "response"), ncol = 1)
-##     return(out)    
-## }
 
 
 ######################################################################

### 2024-03-20 R-code/functions/eval-super-learners.R

@@ -3,9 +3,9 @@
 ## Author: Anders Munch
 ## Created: Nov 10 2023 (14:28) 
 ## Version: 
-## Last-Updated: Nov 20 2023 (13:54) 
+## Last-Updated: Jan  5 2024 (15:33) 
 ##           By: Anders Munch
-##     Update #: 283
+##     Update #: 296
 #----------------------------------------------------------------------
 ## 
 ### Commentary: 
@@ -95,11 +95,10 @@ eval_sl <- function(train_data,
     X_test = uncens_test_data[, !c("true_time", "cens_time")]
     event_ind = at_risk_fun(uncens_test_data, eval_times, "true_time")
     cens_ind = at_risk_fun(uncens_test_data, eval_times, "cens_time")
-    ## Fit SurvSL from Westling et als and eval performance
-    inner_eval_fun = function(surv_pred, cens_pred){
+    inner_eval_fun = function(surv_pred, cens_pred, eval_measure = "brier"){
         out = data.table(time = rep(eval_times,2),
-                         type = rep(c("event", "cens"), each = length(eval_times)),
-                         brier = as.numeric(NA))
+                         type = rep(c("event", "cens"), each = length(eval_times)))
+        out[, brier := as.numeric(NA)]
         if(!is.null(surv_pred)){
             eval_event = colSums((event_ind-surv_pred)^2)/nrow(uncens_test_data)
             out[type == "event", brier := eval_event]
@@ -110,6 +109,7 @@ eval_sl <- function(train_data,
         }
         return(out[])
     }
+    ## Fit SurvSL from Westling et als and eval performance
     if(is.null(SurvSL_learners)){
         eval_survSL = NULL
     }else{

### 2024-03-20 R-code/functions/integral-calc.R

@@ -3,9 +3,9 @@
 ## Author: Anders Munch
 ## Created: Oct 18 2023 (15:10) 
 ## Version: 
-## Last-Updated: Nov  6 2023 (12:11) 
+## Last-Updated: Jan 24 2024 (12:32) 
 ##           By: Anders Munch
-##     Update #: 527
+##     Update #: 530
 #----------------------------------------------------------------------
 ## 
 ### Commentary:
@@ -214,147 +214,23 @@ count_integral <- function(A,
     return(out)
 }
 
+## Support/meta funcions:
+at_risk_fun <- function(newdata,times,time_name = "time"){
+    spring_times = newdata[[time_name]]
+    grid_mat = matrix(times,ncol = length(times),nrow = length(spring_times),byrow = TRUE)
+    ## event_mat = apply(grid_mat, 2, function(x) 1*(spring_times >= x))
+    event_mat = do.call(cbind, lapply(1:ncol(grid_mat), function(xx) 1*(spring_times >= grid_mat[, xx])))
+    return(event_mat)
+}
 
-#### Old versions
-## internal_integral1 <- function(A,
-##                                 B,
-##                                 end,
-##                                 start = 0,
-##                                 data,
-##                                 jump_points,
-##                                 integrate = TRUE){
-##     ## Calculate the integral/sum:
-##     ## \int_start^end A(s) B(ds)
-##     ##
-##     jump_points = sort(unique(c(start, jump_points, end)))
-##     stopifnot(length(jump_points)>1)
-##     wtimes = jump_points[start <= jump_points & jump_points <= end]
-##     n_t = length(wtimes)
-##     A0 = A(newdata = data,times = wtimes)
-##     B0 = B(newdata = data,times = wtimes)
-##     ## B0_diff = t(apply(B0, 1, diff))
-##     B0_diff = do.call(rbind, lapply(1:nrow(B0), function(xx) diff(B0[xx,])))    
-##     ## Why not:
-##     integrands = A0[, -1] * B0_diff
-##     if(integrate)
-##         out = rowSums(integrands)
-##     else
-##         out = integrands
-##     return(out)
-## }
-
-## internal_integral2 <- function(A, B, C, D, end, start = 0, data, jump_points, integrate = TRUE){
-##     ## Calculate the double integral/sum:
-##     ## \int_start^end C(t) [\int_0^t A(s) B(ds)] D(dt)
-##     ##
[…]

### 2024-03-20 R-code/functions/minus_t_fun.R

@@ -1,38 +1,0 @@
-### minus_t_fun.R --- 
-#----------------------------------------------------------------------
-## Author: Anders Munch
-## Created: Nov  6 2023 (12:11) 
-## Version: 
-## Last-Updated: Nov  6 2023 (12:13) 
-##           By: Anders Munch
-##     Update #: 4
-#----------------------------------------------------------------------
-## 
-### Commentary:
-##
-## From
-##    t \mapsto Lambda(newdata, t),
-## construct the function
-##    t \mapsto Lambda(newdata, t-).
-##
-## It is assumed that Lambda only jumps at jump_points.
-## 
-## Returns a function: out=function(newdata,time)
-## 
-### Change Log:
-#----------------------------------------------------------------------
-## 
-### Code:
-## Meta functions:
-minus_t_fun <- function(Lambda, jump_points){
-    t_minus = min(diff(sort(unique(jump_points))))/2
-    fun0 = function(newdata, times){
-        times0 = pmax(0, times-t_minus)
-        Lambda(newdata, times0)
-    }
-    return(fun0)
-}
-
-
-######################################################################
-### minus_t_fun.R ends here

### 2024-03-20 empirical-study/_targets.R

@@ -1,174 +1,0 @@
-### _targets.R --- 
-#----------------------------------------------------------------------
-## Author: Thomas Alexander Gerds
-## Created: Apr 27 2023 (11:06) 
-## Version: 
-## Last-Updated: Nov  6 2023 (14:56) 
-##           By: Anders Munch
-##     Update #: 172
-#----------------------------------------------------------------------
-## 
-### Commentary: 
-## 
-### Change Log:
-#----------------------------------------------------------------------
-## 
-### Code:
-try(setwd("~/research/SuperVision/Anders/survival-loss/statelearner/empirical-study/"))
-library(targets)
-library(tarchetypes)
-library(parallel)
-tar_option_set(packages = c("glmnet","lava","foreach","data.table","prodlim","survival","riskRegression","ranger"))
-tar_source("functions")
-list(
-    tar_target(test_data,draw_data(484)),
-    tar_target(test_fit,{
-        outcome_models <- list(cox0 = coxph(Surv(time,event = event)~1, data=test_data, x = TRUE, y = TRUE),
-                               cox1 = coxph(Surv(time,event = event)~X1+X2, data=test_data, x = TRUE, y = TRUE),
-                               cox2 = coxph(Surv(time,event = event)~X6+X7, data=test_data, x = TRUE, y = TRUE),
-                               cox_dgm = coxph(Surv(time,event = event)~ X1 + X2 + X5 + X6 + X7 + X8 + X11 + X12 + quad.X3 + quad.X4 + quad.X5 + quad.X6,data = test_data,x = TRUE,y = TRUE),
-                               rf = ranger(Surv(time, event = event) ~ X1+X2+X3+X4+X5+X6+X7+X8+X9+X10+X11+X12, data = test_data,num.tree = 50)
-                               ## rf_try = ranger(Surv(time, event = event) ~ X1+X2+X6+X7, mtry = 4, data = test_data),
-                               ## rf_deep = ranger(Surv(time, event = event) ~ X1+X2+X6+X7, mtry = 4, min.node.size = 1, data = test_data)
-                               )
-        cens_models <- list(cox1 = coxph(Surv(time,event = cens_status)~X1+X2, data=test_data, x = TRUE, y = TRUE),
-                            cox2 = coxph(Surv(time,event = cens_status)~X6+X7, data=test_data, x = TRUE, y = TRUE),
[…]

### 2024-03-20 empirical-study/functions/draw_data.R

@@ -1,34 +1,0 @@
-### draw_data.R --- 
-#----------------------------------------------------------------------
-## Author: Thomas Alexander Gerds
-## Created: Jul 11 2023 (13:07) 
-## Version: 
-## Last-Updated: Jul 11 2023 (14:15) 
-##           By: Thomas Alexander Gerds
-##     Update #: 5
-#----------------------------------------------------------------------
-## 
-### Commentary: 
-## 
-### Change Log:
-#----------------------------------------------------------------------
-## 
-### Code:
-draw_data <- function(n){
-    ## set effects
-    beta1 <- rep(c(1,-1,0,0,1,-1),2) * log(2)
-    q.beta1 <- c(1,-1,0,0,1,-1) * log(2)
-    q.gamma <- c(0,0,1,-1,1,-1) * log(2)
-    gamma <- rep(c(0,0,1,-1,1,-1),2) * log(2)
-    ## create data generating model 
-    qmodel <- quadratic_model(beta1=beta1,q.vars=paste("X",1:6,sep=""),q.beta1=q.beta1,q.gamma=q.gamma,gamma=gamma,competing.risks = FALSE)
-    d = sim(qmodel,n)
-    # add censoring status
-    setDT(d)
-    d[,cens_status := 1*(event != 0)]
-    d[]
-}
-
-
-######################################################################
-### draw_data.R ends here

### 2024-03-20 empirical-study/functions/quadratic_model.R

@@ -1,63 +1,0 @@
-### quadratic_model.R --- 
-#----------------------------------------------------------------------
-## Author: Thomas Alexander Gerds
-## Created: Jul 11 2023 (13:08) 
-## Version: 
-## Last-Updated: Jul 11 2023 (13:18) 
-##           By: Thomas Alexander Gerds
-##     Update #: 4
-#----------------------------------------------------------------------
-## 
-### Commentary: 
-## 
-### Change Log:
-#----------------------------------------------------------------------
-## 
-### Code:
-quadratic_model <- function(beta1,
-                            beta2,
-                            gamma,
-                            q.vars,
-                            q.beta1,
-                            q.beta2,
-                            q.gamma,
-                            competing.risks = TRUE){
-    require(lava)
-    m <- lvm()
-    X <- paste("X",seq(12),sep="")
-    timecause1 <- "timecause1"
-    if (competing.risks[[1]]) timecause2 <- "timecause2"
-    censtime <- "censtime"
-    if (competing.risks[[1]])
-        latent(m) <- c(timecause1,timecause2,censtime)
-    else
-        latent(m) <- c(timecause1,censtime)
-    if (competing.risks[[1]])
-        addvar(m) <- c(timecause1,timecause2,censtime,X)
-    else
-        addvar(m) <- c(timecause1,censtime,X)
-    exogenous(m) <- c(X,q.vars)
-    distribution(m,X[1:6]) <- normal.lvm()
-    distribution(m,X[7:12]) <- binomial.lvm()
-    distribution(m,timecause1) <- coxWeibull.lvm(scale=1/100)
-    if (competing.risks[[1]]) distribution(m,timecause2) <- coxWeibull.lvm(scale=1/100)
-    distribution(m,censtime) <- coxWeibull.lvm(scale=1/100)
-    regression(m,to=timecause1,from=X) <- beta1
-    if (competing.risks[[1]]) regression(m,to=timecause2,from=X) <- beta2
-    regression(m,to=censtime,from=X) <- gamma
-    ## quadratic terms
-    for (v in q.vars){
-        form.v <- formula(paste("quad.",v,"~",v,sep=""))
-        transform(m,form.v) <- function(x)x^2
-    }
-    regression(m,to=timecause1,from=paste("quad.",q.vars,sep="")) <- q.beta1
-    if (competing.risks[[1]]) regression(m,to=timecause2,from=paste("quad.",q.vars,sep="")) <- q.beta2
[…]

### 2024-03-20 empirical-study/functions/statelearner-v2.R

@@ -1,263 +1,0 @@
-### statelearner-v2.R --- 
-#----------------------------------------------------------------------
-## Author: Anders Munch
-## Created: Aug 18 2023 (10:26) 
-## Version: 
-## Last-Updated: Oct 13 2023 (16:45) 
-##           By: Anders Munch
-##     Update #: 363
-#----------------------------------------------------------------------
-## 
-### Commentary: 
-## 
-### Change Log:
-#----------------------------------------------------------------------
-## 
-### Code:
-
-
-## Set data on canonical form
-form_data <- function(data, max_time = Inf, time = "time", status = "status", cause_codes = c("1" = 1, "2" = 2, "c" = 0), vars = NULL){
-    if(!is.null(vars))
-        wd = copy(data)[, c(time, status, vars), with = FALSE]
-    else
-        wd = copy(data)
-    setnames(wd, old = c(time, status), new = c("time", "status"))
-    ## Make cause of interest = 1, everything else = 2
-    wd[, cause1 := 1+1*(status != cause_codes[["1"]] & time <= max_time)]
-    wd[, censor := 1+1*(status != cause_codes[["c"]] & time <= max_time)]
-    if(wd[, length(unique(status))>2])
-        wd[, cause2 := 1+1*(status != cause_codes[["2"]] & time <= max_time)]
-    wd[, status := NULL]
-    return(wd[])
-}
-## Fit model on canonical data form
-fit_cause_model <- function(model, data, cause, x_form = NULL, ...){
-    cause_names = c("cause1", "cause2", "censor")
-    cause_names = cause_names[cause_names %in% names(data)]
-    stopifnot(cause %in% cause_names)
-    wd = copy(data)[, -cause_names[cause_names != cause], with = FALSE]
-    setnames(wd, old = cause, new = "event")
-    ## not so elegant
-    if(is.null(x_form))
-        form = Surv(time, event) ~ .
-    else
-        form = update(x_form, Surv(time, event) ~ .)
-    if(model == "cox"){
-        wd[, event := 1*(event == 1)] ## To make this cause of interest
-        out = coxph(form, data = wd,x = TRUE,y = TRUE, ...)
-    }
-    if(model == "GLMnet"){
-        wd[, event := 1*(event == 1)] ## To make this cause of interest
-        out = GLMnet(form, data = wd, ...)
-    }
-    if(model == "rfsrc"){
-        out = rfsrc(form, data = wd, ...)
-    }
[…]

### 2024-03-20 empirical-study/functions/zelefsky-data-learners.R

@@ -1,275 +1,0 @@
-### zelefsky-data-learners.R --- 
-#----------------------------------------------------------------------
-## Author: Anders Munch
-## Created: Jul 23 2023 (14:00) 
-## Version: 
-## Last-Updated: Nov 12 2023 (18:18) 
-##           By: Anders Munch
-##     Update #: 136
-#----------------------------------------------------------------------
-## 
-### Commentary: 
-## 
-### Change Log:
-#----------------------------------------------------------------------
-## 
-### Code:
-library(data.table)
-library(prodlim)
-library(riskRegression)
-library(lava)
-library(survival)
-library(randomForestSRC)
-source("./functions/simZelefsky.R")
-sim_learners_zelefsky <- function(ns,simulation_input, eval_time = 36, B = 1, split.method = "cv5"){
-    do.call(rbind, lapply(ns, function(nn){
-        sim_dt = simZelefsky(n = nn,censoring = TRUE,simulation_input = simulation_input)
-        sim_dt[, ":="(time = dmos, event = status)]
-        out_mms = list(km = coxph(Surv(time,event)~1, data=sim_dt, x = TRUE, y = TRUE),
-                       cox_full = coxph(Surv(time,event)~logPSA+stage+ggtot+sDose+hormones, data=sim_dt, x = TRUE, y = TRUE),
-                       cox_int = coxph(Surv(time,event)~pspline(logPSA) + sDose + stage + ggtot+hormones, data=sim_dt, x = TRUE, y = TRUE),
-                       cox_spline = coxph(Surv(time,event)~pspline(logPSA)+stage+ggtot+pspline(sDose)+hormones, data=sim_dt, x = TRUE, y = TRUE),
-                       cox_small = coxph(Surv(time,event)~logPSA+sDose+stage, data=sim_dt, x = TRUE, y = TRUE)
-                       )
-        cens_mms = list(km = coxph(Surv(time,event==0)~1, data=sim_dt, x = TRUE, y = TRUE),
-                        cox_full = coxph(Surv(time,event==0)~logPSA+stage+ggtot+sDose+hormones, data=sim_dt, x = TRUE, y = TRUE),
-                        cox_int = coxph(Surv(time,event==0)~pspline(logPSA) + sDose + stage + ggtot+hormones, data=sim_dt, x = TRUE, y = TRUE),
[…]

### 2024-03-20 empirical-study/sandbox/another.R

@@ -1,51 +1,0 @@
-library(riskRegression)
-library(survival)
-library(lava)
-library(data.table)
-library(prodlim)
-try(setwd("~/research/SuperVision/Anders/survival-loss/statelearner/empirical-study/"))
-tar_source("functions")
-tar_source("~/Documents/phd/survival-loss-function/statelearner/empirical-study/functions")
-
-m <- lvm()
-# B discriminates only when A == 1
-formula_event <- ~f(A,0)+f(B,0)+f(C,0)+f(AB,-1.5)+f(AC,-1.3)
-formula_cens <- ~f(A,0)+f(B,2)+f(C,0)
-event_scale <- 1/500
-cens_scale <- 1/500
-lava::distribution(m,~A) <- lava::binomial.lvm(p = .4)
-lava::distribution(m,~B) <- lava::binomial.lvm(p = .3)
-lava::distribution(m,~C) <- lava::binomial.lvm(p = .35)
-transform(m,AB~A+B) <- prod
-transform(m,AC~A+C) <- prod
-lava::distribution(m,~censtime) <- lava::coxWeibull.lvm(scale=event_scale)
-lava::distribution(m,~eventtime) <- lava::coxWeibull.lvm(scale=cens_scale)
-m <- lava::eventTime(m, time ~ min(eventtime = 1, censtime = 0), "event")
-lava::regression(m) <- stats::update(formula_event, "eventtime~.")
-lava::regression(m) <- stats::update(formula_cens, "censtime~.")
-d <- setDT(sim(m,10000))
-d[,table(event)]
-dlarge <- setDT(sim(m,10000))
-dlarge[,dummy := 1]
-## pdf("~/tmp/test.pdf",width = 12,height = 6)
-pdf("~/tmp/test.pdf",width = 10,height = 10)
-par(mfrow = c(2,2))
-plot(prodlim(Hist(time,event)~A+B,data = d),xlim = c(0,20),newdata = expand.grid(A =c(0,1),B =c(0,1)),type = "risk",plot.main = "Event probability",confint = FALSE,legend.cex = 0.5)
-plot(prodlim(Hist(time,event)~A+C,data = d),xlim = c(0,20),newdata = expand.grid(A = c(0,1),C = c(0,1)),type = "risk",plot.main = "Event probability",confint = FALSE,legend.cex = 0.5)
[…]

### 2024-03-20 empirical-study/sandbox/easy.R

@@ -1,50 +1,0 @@
-library(riskRegression)
-library(survival)
-library(lava)
-library(data.table)
-library(prodlim)
-try(setwd("~/research/SuperVision/Anders/survival-loss/statelearner/empirical-study/"))
-tar_source("functions")
-m <- lvm()
-formula_event <- ~f(age,-1)+f(sex,1)+f(agesex,0.5)
-formula_cens <- ~f(age,-0.8)+f(sex,1.3)
-event_scale <- 1/1000
-cens_scale <- 1/1000
-lava::distribution(m, ~age) <- lava::binomial.lvm(p = .4)
-lava::distribution(m,~sex) <- lava::binomial.lvm(p = .3)
-transform(m,agesex~age+sex) <- prod
-lava::distribution(m,~censtime) <- lava::coxWeibull.lvm(scale=event_scale)
-lava::distribution(m,~eventtime) <- lava::coxWeibull.lvm(scale=cens_scale)
-m <- lava::eventTime(m, time ~ min(eventtime = 1, censtime = 0), "event")
-lava::regression(m) <- stats::update(formula_event, "eventtime~.")
-lava::regression(m) <- stats::update(formula_cens, "censtime~.")
-d <- setDT(sim(m,20000))
-d[,table(event)]
-dlarge <- setDT(sim(m,5000))
-dlarge[,dummy := 1]
-## pdf("~/tmp/test.pdf",width = 12,height = 6)
-pdf("~/tmp/test.pdf",width = 10,height = 10)
-par(mfrow = c(2,2))
-plot(prodlim(Hist(time,event)~sex,data = d),xlim = c(0,50),newdata = expand.grid(sex =c(0,1)),type = "risk",plot.main = "Event probability",confint = FALSE,legend.cex = 0.5)
-plot(prodlim(Hist(time,event)~age,data = d),xlim = c(0,50),newdata = expand.grid(age = c(0,1)),type = "risk",plot.main = "Event probability",confint = FALSE,legend.cex = 0.5)
-plot(prodlim(Hist(censtime,dummy)~age,data = dlarge),xlim = c(0,50),newdata = expand.grid(age =c(0,1)),plot.main = "Censoring probability",confint = FALSE,legend.cex = 0.5,type = "risk")
[…]

### 2024-03-20 empirical-study/sandbox/play.R

@@ -1,46 +1,0 @@
-### play.R --- 
-#----------------------------------------------------------------------
-## Author: Thomas Alexander Gerds
-## Created: Jul 18 2023 (15:39) 
-## Version: 
-## Last-Updated: Jul 18 2023 (16:08) 
-##           By: Thomas Alexander Gerds
-##     Update #: 1
-#----------------------------------------------------------------------
-## 
-### Commentary: 
-## 
-### Change Log:
-#----------------------------------------------------------------------
-## 
-### Code:
-library(riskRegression)
-library(lava)
-library(data.table)
-library(prodlim)
-m <- lvm()
-formula_event <- ~f(age,0)
-formula_cens <- ~f(age,0)
-event_scale <- 1/1000
-cens_scale <- 1/1000
-lava::distribution(m, ~age) <- lava::normal.lvm(mean = 60,sd = 11)
-lava::distribution(m, ~biomarker) <- lava::normal.lvm(mean = 0,sd = 1)
-lava::regression(m) <- eventtime ~ 0.01*age+0.4*sex
-lava::regression(m) <- censtime ~ 0.03*age-0.4*sex
-lava::regression(m) <- S_event ~ exp(0.4*como-0.5*sex)
-lava::regression(m) <- S_cens ~ exp(-.4*como+0.8*sex)
-lava::distribution(m,~sex+como) <- lava::binomial.lvm()
-lava::distribution(m,~censtime) <- lava::coxWeibull.lvm(scale=event_scale,shape = ~S_cens)
-lava::distribution(m,~eventtime) <- lava::coxWeibull.lvm(scale=cens_scale,shape=~S_event)
-m <- lava::eventTime(m, time ~ min(eventtime = 1, censtime = 0), "event")
-## lava::regression(m) <- stats::update(formula_event, "eventtime~.")
-## lava::regression(m) <- stats::update(formula_cens, "censtime~.")
-d <- setDT(sim(m,1000))
-d[,table(event)]
-pdf("~/tmp/test.pdf",width = 12,height = 6)
-par(mfrow = c(1,2))
-plot(prodlim(Hist(time,event)~sex+age,data = d),xlim = c(0,50),newdata = expand.grid(age = c(20,60,90),sex =c(0,1)),type = "risk",plot.main = "Event probability",confint = FALSE,legend.cex = 0.5)
[…]

### 2024-03-20 empirical-study/sandbox/sandbox-sl2-comp.R

@@ -1,183 +1,0 @@
-### sandbox-v2.R --- 
-#----------------------------------------------------------------------
-## Author: Anders Munch
-## Created: Aug 18 2023 (16:18) 
-## Version: 
-## Last-Updated: Aug 21 2023 (14:06) 
-##           By: Anders Munch
-##     Update #: 81
-#----------------------------------------------------------------------
-## 
-### Commentary: 
-## 
-### Change Log:
-#----------------------------------------------------------------------
-## 
-### Code:
-## Get CHF for rf
-
-library(randomForestSRC)
-library(data.table)
-data(wihs, package = "randomForestSRC")
-setDT(wihs)
-wihs_fix <- copy(wihs)
-wihs_fix[, time := time + abs(rnorm(.N, sd = 0.01))]
-## dd <- form_data(wihs_fix)
-
-learners1 <- list(cox =  list("cox"),
-                  cox_small =  list("cox", x_form = ~ageatfda + cd4nadir),
-                  rf = list("rfsrc", ntree = 50)
-                  )
-learner_all <- list(cause1 = learners1, cause2 = learners1, censor = learners1)
-
-sl0 <- statelearner2(learner_all, data = wihs_fix, time = 5, integrate = FALSE, verbose = TRUE)
-sl0
-
-zelefsky
-zelefsky[, time := dmos]
-zelefsky[status == 0 & vital == "Dead", status := 2]
-use_dat <- copy(zelefsky)[, .(time,status,logPSA,stage,ggtot,sDose,hormones)]
-zel_learner <- list(
-    ## cox =  list("cox"),
-    cox_lasso =  list("GLMnet", x_form = ~logPSA+stage+ggtot+sDose+hormones), ## Does not work without explicit formula?
-    ## cox_ridge =  list("GLMnet", x_form = ~logPSA+stage+ggtot+sDose+hormones, alpha = 0),
-    cox_elastic =  list("GLMnet", x_form = ~logPSA+stage+ggtot+sDose+hormones, alpha = 0.5),
-    cox_strata_stage = list("cox", x_form = ~logPSA+strata(stage)+ggtot+sDose+hormones),
-    km =  list("cox", x_form = ~ 1),                    
[…]

### 2024-03-20 empirical-study/sandbox/simple.R

@@ -1,47 +1,0 @@
-library(riskRegression)
-library(survival)
-library(lava)
-library(data.table)
-library(prodlim)
-m <- lvm()
-formula_event <- ~f(age,-0.01)+f(sex,0.3)+f(quad.age,0.01)
-formula_cens <- ~f(age,0)+f(sex,1.8)
-event_scale <- 1/1000
-cens_scale <- 1/1000
-lava::distribution(m, ~age) <- lava::normal.lvm(mean = 60,sd = 11)
-transform(m,quad.age~age) <- function(x)(x$age-60)^2
-lava::distribution(m,~sex) <- lava::binomial.lvm(p = .3)
-lava::distribution(m,~censtime) <- lava::coxWeibull.lvm(scale=event_scale)
-lava::distribution(m,~eventtime) <- lava::coxWeibull.lvm(scale=cens_scale)
-m <- lava::eventTime(m, time ~ min(eventtime = 1, censtime = 0), "event")
-lava::regression(m) <- stats::update(formula_event, "eventtime~.")
-lava::regression(m) <- stats::update(formula_cens, "censtime~.")
-d <- setDT(sim(m,1000))
-d[,table(event)]
-dlarge <- setDT(sim(m,10000))
-dlarge[,dummy := 1]
-pdf("~/tmp/test.pdf",width = 12,height = 6)
-par(mfrow = c(1,2))
-plot(prodlim(Hist(time,event)~sex+age,data = d),xlim = c(0,50),newdata = expand.grid(sex =c(0,1),age = c(20,50,90)),type = "risk",plot.main = "Event probability",confint = FALSE,legend.cex = 0.5)
-plot(prodlim(Hist(censtime,dummy)~sex,data = dlarge),xlim = c(0,50),newdata = expand.grid(sex =c(0,1)),plot.main = "Censoring probability",confint = FALSE,legend.cex = 0.5,type = "risk")
-plot(prodlim(Hist(time,event)~1,data = d,rev = TRUE),xlim = c(0,50),plot.main = "Censoring probability",confint = FALSE,legend.cex = 0.5,add = TRUE,type = "risk",lty = 3)
-dev.off()
-d[,sex_age := 1]
-d[sex == 1,sex_age := age]
-d[,age_sex := 1]
-d[sex == 0,age_sex := age]
-dlarge[,sex_age := 1]
[…]

### 2024-03-20 experiments/_targets.R

@@ -3,9 +3,9 @@
 ## Author: Anders Munch
 ## Created: Nov 14 2023 (09:52) 
 ## Version: 
-## Last-Updated: Nov 21 2023 (10:43) 
+## Last-Updated: Jan  5 2024 (14:13) 
 ##           By: Anders Munch
-##     Update #: 39
+##     Update #: 40
 #----------------------------------------------------------------------
 ## 
 ### Commentary: 
@@ -32,6 +32,7 @@ list(
     tar_target(zel_all_results, combine_results("zel-all-settings", results_dir = results_dir)),
     tar_target(zel_all0_results, combine_results("zel-all-settings0", results_dir = results_dir)),
     tar_target(zel_all2_results, combine_results("zel-all-settings2", results_dir = results_dir)),
+    tar_target(zel_sim2_1, combine_results("zel-sim2-1", results_dir = results_dir)),
     tar_target(ipcw_fail_sim, {
         run0 = combine_results("ipcw-fail-sim", results_dir = results_dir)
         state_more = combine_results("ipcw-fail-sim-more-state-learners", results_dir = results_dir)

### 2024-03-20 experiments/sandbox.R

@@ -1,120 +1,0 @@
-### sandbox.R --- 
-#----------------------------------------------------------------------
-## Author: Anders Munch
-## Created: Nov 14 2023 (11:11) 
-## Version: 
-## Last-Updated: Nov 20 2023 (17:10) 
-##           By: Anders Munch
-##     Update #: 86
-#----------------------------------------------------------------------
-## 
-### Commentary: 
-## 
-### Change Log:
-#----------------------------------------------------------------------
-## 
-### Code:
-
-library(ggplot2)
-
-tar_load(zel_sim1_results, store = here("experiments/_targets"))
-tar_load(zel_indep_cens_sim_results, store = here("experiments/_targets"))
-
-time_inc <- zel_sim1_results[, diff(unique(time))[1]]
-max_time <- zel_sim1_results[, max(time)[1]]
-int_brier_zel_sim1 <- zel_sim1_results[!(type == "cens" & grepl("ipcw", SL)), .(int_brier = sum(100*brier)*time_inc/max_time), .(n_obs, type, SL, seed)]
-summ_zel_sim1 <- int_brier_zel_sim1[, .(ave_int_brier = mean(int_brier), se = sd(int_brier)/sqrt(.N)), .(n_obs, type, SL)]
-dd_ww <- 0.1
-
-ggplot(summ_zel_sim1,
-       aes(x = n_obs, y = ave_int_brier, col = SL)) +
-    theme_bw() +
-    geom_errorbar(position=position_dodge(width = dd_ww), aes(ymin = ave_int_brier-1.96*se, ymax = ave_int_brier+1.96*se), width = .05, alpha = .5) + 
-    geom_line(position=position_dodge(width = dd_ww)) + geom_point(position=position_dodge(width = dd_ww)) +
-    scale_x_continuous(trans='log2') +
-    facet_wrap(~type, scales = "free_y")
-
-time_inc <- zel_indep_cens_sim_results[, diff(unique(time))[1]]
-max_time <- zel_indep_cens_sim_results[, max(time)[1]]
-int_brier_zel_indep_cens <- zel_indep_cens_sim_results[!(type == "cens" & grepl("ipcw", SL)), .(int_brier = sum(100*brier)*time_inc/max_time), .(n_obs, type, SL, seed)]
[…]

### 2024-03-20 tex/jrss-version/state-learner-jrss-style.tex
jrss setup

--- 2023-11-06 tex/state-learner-pub-version.tex
@@
-\documentclass[a4,danish]{article}
-\usepackage[a4paper, margin=3.5cm]{geometry}
-\input{./latex-settings/standard-settings.tex}
-\input{./latex-settings/standard-commands.tex}
-\input{./latex-settings/theorem-env.tex}
-\input{./latex-settings/title-compact.tex}
-
+\documentclass{statsoc}
+\usepackage[a4paper]{geometry}
+
+\usepackage{natbib, dsfont, enumitem,
+  amssymb,soul,xcolor,amsmath,graphicx,verbatim,pgfplots,tikz,prodint,booktabs}
+
+%% Theorem environment
+% \theoremstyle{plain} % plain italic style
+\newtheorem{theorem}{Theorem}
+\numberwithin{theorem}{section}
+\newtheorem{lemma}[theorem]{Lemma}
+\newtheorem{proposition}[theorem]{Proposition}
+\newtheorem{corollary}[theorem]{Corollary}
+
+%% Only for comments:
+\usepackage[author=]{fixme}
+\fxusetheme{color}
+\definecolor{fxtarget}{rgb}{.5,.5,.5}
+\definecolor{fxnote}{rgb}{.5,.5,.5}
 \fxsetup{status=draft}
-\usepackage[colorlinks=true,citecolor=blue]{hyperref} % Doesn't work for org, so included here
-
-%% Specific libraries
-\usepackage{booktabs}
-
-%% Specific commands
+
+\bibliographystyle{rss}
+
+% New operators and commands
+\DeclareMathOperator{\E}{\mathbb{E}} % expectation
+\newcommand{\Z}{\mathbb{Z}}
+\newcommand{\R}{\mathbb{R}}
+\newcommand{\N}{\mathbb{N}}
+\newcommand{\C}{\mathbb{C}}
+\renewcommand{\S}{\mathbb{S}}
+\newcommand{\blank}{\makebox[1ex]{\textbf{$\cdot$}}}
+\newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
+\def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
+\renewcommand{\phi}{\varphi}
+\renewcommand{\epsilon}{\varepsilon}
+\newcommand*\diff{\mathop{}\!\mathrm{d}}
+\newcommand{\weakly}{\rightsquigarrow}
+\newcommand\smallO{\textit{o}}
+\newcommand\bigO{\textit{O}}
+\newcommand{\midd}{\; \middle|\;}
+\newcommand{\1}{\mathds{1}}
+\usepackage{ifthen} %% Empirical process with default argument
+% \newcommand{\G}[1][]{%
+%    \ifthenelse{ \equal{#1}{} }
+%       {\ensuremath{\mathbb{G}_n}}
+%       {\ensuremath{\mathbb{G}_{#1}}}
+% }
+% New version:
+\newcommand{\G}[2][n]{
+{\ensuremath{\mathbb{G}_{#1}}{\left[#2\right]}}
+}
+\DeclareMathOperator*{\argmin}{\arg\!\min}
+\DeclareMathOperator*{\argmax}{\arg\!\max}
+\newcommand{\empmeas}{\ensuremath{\mathbb{P}_n}} % empirical measure
+
 \newcommand{\data}{\ensuremath{\mathcal{D}}}
 
+\title[The state learner]{The state learner \\ {-- a super learner for right-censored data}}
+\author[Anders Munch {\it et al.}]{Anders Munch}
+\address{Section of Biostatistics, University of Copenhagen, Copenhagen, Denmark}
+\email{a.munch@sund.ku.dk}
+\author{Thomas A.\ Gerds}
+\address{Section of Biostatistics, University of Copenhagen, Copenhagen, Denmark}
+
 \begin{document}
-
-% \articletype{ARTICLE TEMPLATE}% Specify the article type or omit as appropriate
-
-\title{The state learner \\ {-- a super learner for right-censored data}}
-
-\author{Anders Munch and Thomas A. Gerds
-  % \name{Anders Munch\textsuperscript{a}\thanks{CONTACT: Anders Munch. Email: a.munch@sund.ku.dk}
-  %   and Thomas A. Gerds\textsuperscript{a}}
-  % \affil{\textsuperscript{a} Section of Biostatistics, Department of Public Health, University of Copenhagen}
-}
-
-\maketitle
 
@@
   hazards of interest nor the censoring distribution.
-  % In survival analysis, prediction models are needed as stand-alone tools and in
-  % applications of causal inference to estimate nuisance parameters. The super
-  % learner is a machine learning algorithm which combines a library of prediction
-  % models into a meta learner based on cross-validated loss. Unfortunately, the
-  % commonly used partial likelihood loss is not suited for super learning, and
-  % inverse probability of censoring weighted loss functions require a
-  % pre-specified estimator of the censoring distribution. To relax this, we
-  % introduce the state learner, a new super learner for survival analysis, which
-  % evaluates the loss based on the observed data simultaneously for two libraries
-  % of prediction models, one for the event time distribution and one for the
-  % censoring distribution. We establish an oracle inequality for the state
-  % learner and investigate its performance through numerical experiments.
 \end{abstract}
 
-% \begin{keywords}
-%   Loss based estimation; right-censored data; competing risks; super learner;
-%   cross-validation
-% \end{keywords}
-
-% \begin{amscode}
-%   % Use 2010 Mathematics Subject Classification keywords
-%   62N02
-% \end{amscode}
+\keywords{Loss based estimation, right-censored data, competing risks, super
+  learner, cross-validation}
 
@@
 
-% For machine learning with right-censored data, the negative partial
-% log-likelihood is a popular loss function
-% \citep[e.g.,][]{li2016regularized,
-%   yao2017deep,lee2018deephit,katzman2018deepsurv,
-%   gensheimer2019scalable,lee2021boosted, golmakani2020super}.
-% However, there are two major problems when attempting to use the partial log-likelihood for super
-% learning. The first is that the partial
-% log-likelihood cannot be used to evaluate cross-validated loss for
-% survival learners which are not absolutely continuous in the time
-% argument \citep{golmakani2020super}. The second problem is that the
-% expected value of the partial log-likelihood depends on the censoring
-% distribution. Both problems are further discussed in Section
-% \ref{sec:relat-liter-exist}.
-
-% Inverse probability of censoring weighted (IPCW) estimators are often
-% used to resolve the dependence of the expected loss on the censoring
-% distribution \citep{graf1999assessment,
-%   molinaro2004tree,keles2004asymptotically,hothorn2006survival,gerds2006consistent,gonzalez2021stacked}.
-% However, commonly these methods require a consistent estimator of the
-% conditional censoring distribution given covariates. An obvious idea
-% would be to estimate the conditional censoring distribution by a super
-% learner. However, that would require a loss function to evaluate the
-% performance of candidate learners of the censoring distribution and
-% the expected value of that loss function would in turn depend on the
-% event time distribution. To resolve this dilemma
-% \cite{westling2021inference} and \cite{han2021inverse} suggested to
-% iterate between super learning of the censoring distribution and the
-% event time distribution, where both super learners use the respective
-% cross-validated IPCW loss. Our approach is different. The state
-% learner uses all combinations of learners for the event time
-% distribution with a learner for the censoring distribution to
-% parameterize a set of `observable' cumulative distribution
-% functions. Specifically, the cross-validated integrated Brier score
-% for each of the cumulative distribution functions then only depends on
-% the observed data distribution and hence can be estimated empirically
-% without further modeling.
-
 The article is organized as follows. We introduce our notation and framework in
@@
 learning in general with right-censored data, and in
-Section~\ref{sec:super-learner-simple} we introduce the state learner.
-Section~\ref{sec:theor-results-prop} provides theoretical guarantees for the
-state learner. In Section~\ref{sec:targeted-learning} we discuss the use of the
-state learner in the context of targeted learning. We report a numerical study
-in Section~\ref{sec:numer-exper}, and analyze a prostate cancer data set in
-Section~\ref{sec:real-data-appl}. Finally, we relate the state learner to
-existing approaches in Section~\ref{sec:discussion} and discuss some limitations
+Section~\ref{sec:relat-liter-exist} we review existing methods.
+Section~\ref{sec:super-learner-simple} introduces our novel proposal, the state
+learner, and Section~\ref{sec:theor-results-prop} provides theoretical
+guarantees for the state learner. In Section~\ref{sec:targeted-learning} we
+discuss the use of the state learner in the context of targeted learning. We
+report some numerical experiments in Section~\ref{sec:numer-exper} and analyze a
+prostate cancer data set in Section~\ref{sec:real-data-appl}.
+Section~\ref{sec:discussion} contains a discussion of the merits and limitations
 of our proposal. Appendices~\ref{sec:proof-proposition}
@@
 \( \mathcal{P} \) to \(\mathcal M\).
-% and we shall sometimes write
-% \( \Lambda^{P} \) to emphasize that \( \Gamma^{P} \) is associated with a
-% particular data-generating distribution \( P \).
 
@@
 
+\section{Existing methods}
+\label{sec:relat-liter-exist}
+
+Machine learning based on right-censored data commonly use the negative partial
+log-likelihood as loss function
+\citep[e.g.,][]{li2016regularized,yao2017deep,lee2018deephit,katzman2018deepsurv,gensheimer2019scalable,lee2021boosted,kvamme2021continuous}.
+However, this loss function is unsuited for super learning, because many
+canonical survival learners (e.g., the Kaplan-Meier estimator, random survival
+forest, and semi-parametric Cox models) provide cumulative hazard functions that
+are piece-wise constant in the time argument, and hence assign zero probability
+to event times not observed in the training data. This implies that when data
+are observed in continuous time, any of these learners will almost surely have
+infinite loss in any independent hold-out sample according to the negative
+partial log-likelihood loss.
+
+When a proportional hazards model is assumed, the baseline hazard function can
+be profiled out of the likelihood to give a new partial log-likelihood loss
+\citep{cox1972regression}, which has been suggested as a loss function for super
+learning \citep{golmakani2020super,verweij1993cross}. While this allows the
+library of learners to include Cox' proportional hazard models, the drawback is
+that the library is now restricted to include only these models.
+
+Another approach for super learning with right-censored data is to use an
+inverse probability of censoring weighted (IPCW) loss function
+\citep{graf1999assessment,van2003unicv,molinaro2004tree,keles2004asymptotically,hothorn2006survival,gerds2006consistent,gonzalez2021stacked}.
+An IPCW loss function is attractive because the associated risk does not depend
+on the censoring distribution but describes a feature of the population of
+interested governed by the measure \( Q \in \mathcal{Q} \). Similar results can
+be obtained using censoring unbiased transformations
+\citep{fan1996local,steingrimsson2019censoring} or pseudo-values
+\citep{andersen2003generalised,mogensen2013random,sachs2019ensemble}. All these
+methods rely on an estimator of the censoring distribution, and their drawback
+is that this estimator has to be pre-specified. When the data-generating
+mechanism is complex and not well-understood, pre-specification of the censoring
+distribution is a challenge.
+
+As far as we are aware, the only existing attempt at avoiding the need to
+pre-specify a censoring model is a recent proposal suggested independently by
+\cite{han2021inverse} and \cite{westling2021inference}. The authors do not
+consider competing risks but suggest to iterate between learning \( \Lambda \)
+and $\Gamma$ using IPCW loss functions and select the final learner when the
+iterative procedure has converged. To the best of our knowledege, no general
+theoretical guarantees exist for this procedure. It seems unclear whether this
+iterative procedure will always be able to find a correctly specified model in a
+given library.
+
 \section{The state learner}
@@
   \centering
-  \includegraphics[width=.5\textwidth]{./figures/figure-multi-state-process.pdf}
+  \includegraphics[width=.5\textwidth]{./figure-multi-state-process.pdf}
   \caption{Illustration of the multi-state process \(\eta\) used by
@@
 \end{figure}
-The conditional distribution of \( \eta(t) \) given \( X=x \) is
-determined by the function
+The conditional distribution of \( \eta(t) \) given \( X=x \) is determined by
+the \fxnote*{reconsider letting censoring state be \( -1 \)?}{function}
 \begin{equation}
@@
 \end{align*}
-% Note that depending on the target parameter of the learning task, the
-% state learner may or may not depend on the distribution of the
-% covariates \(H_n\). For example, for the parameter
-% \(\tilde\Psi(\Lambda_{1}, \Lambda_{2}, \Gamma, H)=\Lambda_{1}\) the
-% state learner would not depend on \(H_n\) but for the parameter
-% \(\tilde\Psi(\Lambda_{1}, \Lambda_{2}, \Gamma, H)=\int S(t \mid x)H(\diff
-% x)\) it would.
 
@@
 equation~(\ref{eq:dr-term}) holds for the a targeted estimator based on
-estimator $\hat{\Lambda}_{1,n}$, $\hat{\Lambda}_{2,n}$, and $\hat{\Gamma}_{n}$,
+estimators $\hat{\Lambda}_{1,n}$, $\hat{\Lambda}_{2,n}$, and $\hat{\Gamma}_{n}$,
 then a similar product structure holds for a targeted estimator based on
@@
 
-An example of a targeted estimator is the \fxnote*{also TMLE, depending on
-  implementation}{one-step estimator}, defined as
+An example of a targeted estimator is the one-step estimator, defined as
 \begin{equation}
@@
     \left\{
-      (\Lambda_1-\hat{\Lambda}_{1,n})^2
+      \|\Lambda_1-\hat{\Lambda}_{1,n}\|^2
       +
-      (\Lambda_2-\hat{\Lambda}_{2,n})^2
+      \|\Lambda_2-\hat{\Lambda}_{2,n}\|^2
       +
-      (\Gamma-\hat{\Gamma}_{n})^2
+      \|\Gamma-\hat{\Gamma}_{n}\|^2
     \right\}
-  }.
+  },
 \end{equation}
-Hence, under a suitable Donsker class regularity condition
-\citep{bickel1993efficient,kennedy2016semiparametric}, when
+for some suitable norm \( \|\blank \| \), for instance the
+\( \mathcal{L}_{P}^2 \)-norm. Hence, under a suitable Donsker class regularity
+condition \citep{bickel1993efficient,kennedy2016semiparametric}, when
 equation~(\ref{eq:4}) holds and the nuisance parameters $\Lambda_1$,
 $\Lambda_2$, and $\Gamma$ are consistently estimated at rate
-\( \smallO_P{(n^{-1/4})} \) then
+\( \smallO_P{(n^{-1/4})} \), then
 \begin{equation}
@@
   \sqrt{n}(\hat{\Psi}_{\text{OS}}(\data_n)- \Psi(P)) \weakly \mathcal{N}(0,
-  P{[\psi(\blank; \Lambda_1, \Lambda_2, \Gamma)^2]}).
+  P{[\psi(\blank; \Lambda_1, \Lambda_2, \Gamma)^2]}),
 \end{equation}
+where we use \( \weakly \) to denote weak convergence \citep{van2000asymptotic}.
 In particular, equation~(\ref{eq:3}) and Slutsky's lemma imply that we can
@@
 
-% When the nuisance parameters are consistently estimated at rate
-% \( \smallO_P{(n^{-1/4})} \) and
-% \begin{equation*}
-%   P{\left(
-%       \psi(\blank; \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
-%       \hat{b}_n(\data_n))
-%       \in \mathcal{R}
-%     \right)} \longrightarrow 1,
-%   \quad \text{as} \quad   n \longrightarrow \infty , 
-% \end{equation*}
-% for some Donsker class of functions \( \mathcal{R} \), it holds that
-% \citep{bickel1993efficient,kennedy2022semiparametric}
-
-% According to equation~(\ref{eq:parametrizeP}) and equation~(\ref{eq:7}), any
-% \( P \) is characterized by both \((\Lambda_1, \Lambda_2, \Gamma, H)\) and
-% \( (F, H) \), which means that there exist operators $\tilde\Psi$ and $\bar\Psi$
-% such that
-% \begin{equation*}
-%   % \label{eq:state-learner-identity}
-%   \tilde\Psi(\Lambda_{1,0},\Lambda_{2,0}, \Gamma_0, H_0)=\Psi(P_0)=\bar\Psi(F_0,H_0).
-% \end{equation*}
-% Hence, any parameter $\Psi$ defined on \( \mathcal{P} \) can be estimated using
-% either estimators \( (\hat{\Lambda}_n, \hat{\Gamma}_n, \hat{H}_n) \) or
-% \( (\hat{F}_n, \hat{H}_n) \). When we use the state learner the final estimator
-% of the target parameter is $\bar\Psi(\hat\phi_n(\data_n), \hat{H}_n)$, where
-% \( \hat{H}_n \) can often be taken to be the marginal empirical measure.
-
-% For the rest of this section we consider the case where $\Theta = \R$, and
-% $\Psi$ is estimable by a regular asymptotically linear estimator. Specifically,
-% we assume that it is possible to construct an estimator
-% \( \tilde\Psi(\hat\Lambda_n, \hat\Gamma_n, H_n) \) such that the the following
-% expansion holds:
-% \begin{equation}
-%   \label{eq:ral}
-%   \tilde\Psi(\hat\Lambda_n, \hat\Gamma_n, H_n) - \Psi(P) = \empmeas{[\psi_P]} +
-%   \mathrm{Rem}{(\hat{\Lambda}_{n},  \hat{\Gamma}_n, P)} + \smallO_{P}(n^{-1/2}),
-% \end{equation}
-% where \( \empmeas \) is the empirical measure of a sample \(\{O_i\}_{i=1}^n\),
-% $\psi_P$ a zero-mean function with \( P[\psi_P^2]< \infty \), and
-% \( \mathrm{Rem}{(\hat{\Lambda}_{n}, \hat{\Gamma}_n, P)} \) a second order
-% remainder term \citep{bickel1993efficient,fisher2021visually}.
-
-% ...
-
-% The \( \smallO_P{(n^{-1/2})} \) bound on the products of estimation errors is
-% typically a weaker condition than requiring \( \Gamma \), $\Lambda_1$, and
-% $\Lambda_2$ to be estimated independently at rate \( n^{-1/2} \). This is
-% particularly important when data-adaptive estimation methods are used as these
-% methods rarely provide \( n^{-1/2} \)-rate convergence for function-valued
-% nuisance parameters.
-
-\subsection{Cause-specific average treatment effect}
+\subsection{Average treatment effect on the cumulative incidence}
 \label{sec:cause-spec-aver}
 
-We now demonstrate how the state learner and the general estimation strategy
-outlined above can be used to estimate a concrete parameter of interest in the
-competing risk setting. We assume that the covariate vector \( X \in \R^p \) can
-be separated into a binary treatment indicator \( A \) and a vector of potential
-confounders, \( W \in \mathcal{W}\subset \R^{p-1} \). We abuse notation slightly
-by writing \( \Lambda_j(t \mid x) = \Lambda_j(t \mid a, w) \) and
+In the following we demonstrate how the state learner and the general estimation
+strategy outlined above can be used to construct an estimator of the
+cause-specific average treatment effect. We assume that the covariate vector
+\( X \in \R^p \) can be separated into a binary treatment indicator \( A \) and
+a vector of potential confounders, \( W \in \mathcal{W}\subset \R^{p-1} \). We
+abuse notation slightly by writing
+\( \Lambda_j(t \mid x) = \Lambda_j(t \mid a, w) \) and
 \( S(t \mid x) = S(t \mid a, w) \) when \( x=(a, w) \). We use $\mu$ to denote
@@
 \( 1 \) on \( \mathcal{W} \). As both \( A \) and \( W \) are fully observed for
-all individuals we can use a super learner to estimate $\pi$
-\fxnote*{add}{[ref]}, and we denote this estimator by $\hat{\pi}_n$. We use the
-empirical measure of \( \{W_1, \dots, W_n\} \) to estimate $\mu$, and denote
-this estimator by the $\hat{\mu}_n$. As parameter of interest we consider the
-\fxnote*{what is this called?}{standardized difference in absolute risk of cause
-  \( 1 \) before time $\tau$},
+all individuals we can use an existing super learner to estimate $\pi$
+\citep{Polley_Ledell_Kennedy_Laan_2023_Superlearn}, and we denote this estimator
+by $\hat{\pi}_n$. We use the empirical measure of \( \{W_1, \dots, W_n\} \) to
+estimate $\mu$, and denote this estimator by the $\hat{\mu}_n$. As parameter of
+interest we consider the standardized difference in the cumulative incidence due
+to cause 1 at time $\tau$, which can be expressed as
 \begin{equation*}
@@
   \theta_{{\tau}}(Q) =
-  \E{[T^{A=1} \leq \tau, D^{A=1}=1]}-
-  \E{[T^{A=0} \leq \tau, D^{A=0}=1]},
+  P{(T^{A=1} \leq \tau, D^{A=1}=1)}-
+  P{(T^{A=0} \leq \tau, D^{A=0}=1)},
 \end{equation*}
@@
 \end{equation}
-The efficient influence function for the parameter $\tilde{\Psi}_{\tau}$
-depends on the set \( (\Lambda_1, \Lambda_2, \Gamma, \pi) \) of nuisance
-parameters, and is given by \fxnote*{other refs?}{\citep{rytgaard2022targeted}}
+The efficient influence function for the parameter $\tilde{\Psi}_{\tau}$ depends
+on the set \( (\Lambda_1, \Lambda_2, \Gamma, \pi) \) of nuisance parameters.
+Define the quantities
+\begin{align*}
+  \omega_a(A,W; \pi)
+  &=  \frac{% (-1)^{a+1}
+    \1{\{A=a\}}}{\pi(W)^{a}(1-\pi(W))^{1-a}},
+  \\
+  g(t, A, W; \Lambda_1, \Lambda_2)
+  & = \int_0^{t}
+    e^{-\Lambda_1(s- \mid W, A)-\Lambda_2(s- \mid W, A)}  \Lambda_1(\diff s \mid
+    W, A),
+  \\  
+  M_j(\diff t \mid A, W;  \Lambda_j  )
+  & = N_j(\diff t) -
+    \1{\{\tilde{T} \geq t\}} \Lambda_j(\diff t \mid W, A),
+    \quad j \in \{1,2\},
+  \intertext{and}
+  M(\diff t \mid A, W;  \Lambda_1, \Lambda_2  )
+  & = M_1(\diff t \mid A, W;  \Lambda_1  ) +
+    M_2(\diff t \mid A, W;  \Lambda_2  ).
+\end{align*}
+The efficient influence function can then be written as
+\citep{van2003unified,jewell2007non,rytgaard2022targeted},
+\begin{equation*}
+  \psi_{\tau}(O; \Lambda_1, \Lambda_2, \Gamma, \pi)
+  = \psi_{\tau}^1(O; \Lambda_1, \Lambda_2, \Gamma, \pi)
+  - \psi_{\tau}^0(O; \Lambda_1, \Lambda_2, \Gamma, \pi)
+  -\tilde{\Psi}_{t}^0(\Lambda_1, \Lambda_2, \mu),
+\end{equation*}
+where
 \begin{equation}
@@
   \label{eq:one-step-comp-ate}
-  \tilde{\Psi}_{t, j}^0(\hat{a}_1(\data_n), \hat{a}_2(\data_n), \hat{\mu}_n(\data_n))
+  \begin{split}
+    \hat{\Psi}_{t,\text{OS}}(\data_n)
+    = &
+        \tilde{\Psi}_{t}^0(\hat{a}_1(\data_n), \hat{a}_2(\data_n),
+        \hat{\mu}_n(\data_n))
+    \\
+      &
+        +
+        \empmeas{[\psi_{\tau}(\blank; \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
+        \hat{b}_n(\data_n), \hat{\pi}_n(\data_n))]}
+    \\
+    = &
+        \empmeas{[\psi_{\tau}^1(\blank; \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
+        \hat{b}_n(\data_n), \hat{\pi}_n(\data_n))]}
+    \\
+      &
+        - \empmeas{[\psi_{\tau}^0(\blank; \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
+        \hat{b}_n(\data_n), \hat{\pi}_n(\data_n))]}.
+  \end{split}
 \end{equation}
@@
 
-We compare the state learner with two other discrete super learners and an
-oracle selector in a simulation study without a competing event. The two other
-super learners are based on inverse probability of censoring weighted (IPCW)
-Brier scores \citep{graf1999assessment,gerds2006consistent}, and we refer to
-these as IPCW super learners. These super learners depend on an estimator of the
-censoring distribution, and we consider IPCW super learners that use either the
-Kaplan-Meier estimator (IPCW(KM)) or a correctly specified Cox model (IPCW(Cox))
-to estimate the censoring distribution. Both IPCW super learners are fitted
-using the \texttt{R}-package \texttt{riskRegression}
-\citep{Gerds_Ohlendorff_Ozenne_2023}. Each discrete super learner provides a
-learner for the cumulative hazard function for the outcome of interest, and from
-this a risk prediction model can be obtained. We measure performance of each
-super learner in terms of the Brier score of the provided risk prediction model
-at a specific time horizon. The Brier score is approximated using a large
-(\( n = 20,000 \)) independent data set of uncensored data. The oracle selector
-uses the large data set of uncensored event times to select the learner with the
-lowest expected Brier score. The expected Brier score of the oracle selector
-serves as a lower benchmark value. For all super learners we split the data into
-five folds for training and testing.
-
-Note that given a learner for the cumulative hazard function of the
-outcome event, we can typically use the same method to construct a
-learner of the cumulative hazard function of the censoring
-distribution. This would typically work by training the learner on the
-data set \( \data_n^c \), where \( \data_n^c = \{O_i^c\}_{i=1}^n \)
-with \( O_i^c = (\tilde{T}_i, 1-\Delta, X_i) \). When we say that we
-use a learner for the cumulative hazard function of the outcome to
-learn the cumulative hazard function of the censoring time, we mean
-that the learner is trained on \( \data_n^c \).
-
-\paragraph{Scenario 1}
-
-We first generate a simple dataset to demonstrate that an IPCW super learner can
-perform poorly when the censoring model is misspecified. We start by generating a
-binomial baseline covariate \( A \) with success probability \( 30\% \). We then
-generate outcome and censoring variables according to a Cox-Weibull distribution
-with hazard rates of approximately \( 0.5 \) and \( 2.5 \), respectively. We use
-a library \( \mathcal{A} \) consisting of two learners, the (marginal)
-Kaplan-Meier estimator and the Kaplan-Meier estimator stratified on the binary
-covariate. For the state learner we use the same learners to construct the
-library \( \mathcal{B} \).
-
-The results are shown in Figure~\ref{fig:ipcw-fail}. We see that the IPCW super
-learner based on a misspecified censoring model (IPCW(KM)) has a larger Brier
-score than the other super learners and that its performance does not improve
-much with sample size. % When the censoring model is misspecified, the
-% IPCW(KM) super learner fails to identify the learner corresponding to the
-% data-generating mechanism, even though this learner is included in the
-% library.
-The performance of the state learner is comparable to or slightly better than
-the IPCW super learner based on a correctly pre-specified censoring model. The
-performance of both the IPCW(Cox) super learner and the state learner are close
-to that of the oracle for most sample sizes.
-
-\begin{figure}
-  \centerline{\includegraphics[width=1\linewidth]{./figures/ipcw-fail.pdf}}
-  \caption[]{Results for scenario~1 of the simulation study. For the learner
-    selected by each of the four discrete super learners, the Brier score
-    calculated in a large independent data set without censoring is plotted
-    against sample size. The results are based on 200 repetitions.}
-  \label{fig:ipcw-fail}
-\end{figure}
-
-\paragraph{Scenario 2}
-
-We next generate data according to a more complex distribution motivated from a
-real dataset in which censoring depends on the baseline covariates. We simulate
-data based on a prostate cancer study described in
+As a proof of concept we conduct two small simulation studies in a setting
+without competing risks. We generate data according to a distribution motivated
+from a real dataset in which censoring depends on the baseline covariates. We
+simulate data based on a prostate cancer study described in
 \citep{kattan2000pretreatment}. The outcome of interest was the time to tumor
@@
 depended on when the patient entered the study \citep{gerds2013estimating}. This
-in turn implied that the time of censoring depended on the radiation dose.
-% It was found that almost all covariates were predictive of censoring.
-The data were re-analyzed in \citep{gerds2013estimating} where a sensitivity
+in turn implied that the time of censoring depended on the radiation dose. The
+data were re-analyzed in \citep{gerds2013estimating} where a sensitivity
 analysis was conducted based on simulated data. We use the same simulation
@@
 estimated from the original data
-\citep[c.f.,][Section~4.6]{gerds2013estimating}. We use the library consisting
-of the nine learners described in Table~\ref{tab:zel-library}. For the state
-learner we use the same library to learn the censoring distribution.
-
-\begin{table}[h]
+\citep[c.f.,][Section~4.6]{gerds2013estimating}. We refer to this simulation
+setting as ``dependent censoring''. We also considered a simulation setting
+where data were generated in the same way, except that censoring was generated
+completely independently. We refer to this simulation setting as ``independent
+censoring''.
+
+For all super learners we use a library consisting of three learners: The
+Kaplan-Meier estimator \citep{kaplan1958nonparametric,Gerds_2019prodlim}, a Cox
+model with main effects \citep{cox1972regression, survival-package}, and a
+random survival forest \citep{ishwaran2008random,rfsrc-paclage}. We use the same
+library to learn the outcome and the censoring mechanism. To be precise, note
+that given a learner for the cumulative hazard function of the outcome event, we
+can typically use the same method to construct a learner of the cumulative
+hazard function of the censoring distribution. This would typically work by
+training the learner on the data set \( \data_n^c \), where
+\( \data_n^c = \{O_i^c\}_{i=1}^n \) with
+\( O_i^c = (\tilde{T}_i, 1-\Delta, X_i) \). When we say that we use a learner
+for the cumulative hazard function of the outcome to learn the cumulative hazard
+function of the censoring time, we mean that the learner is trained on
+\( \data_n^c \).
+
+We first compare the state learner to two IPCW based super learners: The first
+uses a Cox model with main effects to estimate the censoring probabilities
+(IPCW(Cox)), while the second uses the Kaplan-Meier estimator to estimate the
+censoring probabilities (IPCW(KM)). The Cox model for the censoring distribution
+is correctly specified in both simulation settings while the Kaplan Meier
+estimator only estimates the censoring model correctly in the simulation setting
+where censoring is independent. Both IPCW super learners are fitted using the
+\texttt{R}-package \texttt{riskRegression} \citep{Gerds_Ohlendorff_Ozenne_2023}.
+Each super learner provides a learner for the cumulative hazard function for the
+outcome of interest, and from this a risk prediction model can be obtained. We
+measure performance of each super learner by calculating the index of prediction
+accuracy (IPA) \citep{kattan2018index} at a fixed time horizon (36 months) for
+the risk prediction model provided by the super learner. The IPA is 1 minus the
+ratio between the model's Brier score and the null model's Brier score, where
+the null model is the model that does not use any covariate information. The IPA
+is approximated using a large (\( n = 20,000 \)) independent data set of
+uncensored data. As a benchmark we calculate the performance of the risk
+prediction model chosen by the oracle selector. The oracle selector uses the
+large data set of uncensored event times to select the learner with the highest
+IPA.
+
+The results are shown in Figure~\ref{fig:ipcw-fail}. We see that when censoring
+is not independent, using the Kaplan-Meier estimator to estimate the censoring
+probabilities provides a risk prediction model with an IPA that is lower than
+the risk prediction model provided by the state learner. The performance of the
+risk prediction model selected by the state learner is similar to the risk
+prediction model selected by the IPCW(Cox) super learner which a priori uses a
+correctly specified model for the censoring distribution. Both these risk
+prediction models are close to the performance of the oracle, except for small
+sample sizes.
+
+\begin{figure}
+  \centerline{\includegraphics[width=1\linewidth]{./experiment-fig-sl-ipcw.pdf}}
+  \caption[]{For the risk prediction models provided by each of the super
+    learners, the IPA is plotted against sample size. The results are based on
+    1000 repetitions and the error bars are used to quantify the Monte Carlo
+    uncertainty.}
+  \label{fig:ipcw-fail}
+\end{figure}
+
+We next compare the state learner to the super learner \texttt{survSL}
+\citep{westling2021inference}, as this is the only existing super learner we
+know which works without a pre-specified censoring model. As both the state
+learner and \texttt{survSL} also provides a prediction model for the probability
+of being censored, we compare the performance of both the outcome and the
+censoring modeling using the IPA.
+
+The results are shown in Figure~\ref{fig:zelefski}. We see that the prediction
+models selected for both censoring and outcome have similar or higher API
+compared to the prediction models selected by \texttt{survSL} for most sample
+sizes.
+
+\begin{figure}
+  \centerline{\includegraphics[width=1\linewidth]{./experiment-fig-sl-survSL.pdf}}
+  \caption[]{For the risk prediction models of both the outcome and the
+    censoring probability provided by each of the super learners, the IPA is
+    plotted against sample size. The results are based on 1000 repetitions and
+    the error bars are used to quantify the Monte Carlo uncertainty.}
+  \label{fig:zelefski}
+\end{figure}
+
+\section{Example with prostate cancer study}
+\label{sec:real-data-appl}
+
+The prostate cancer data analyzed by \cite{kattan2000pretreatment}, which we
+introduced in Section~\ref{sec:numer-exper}, include a competing event in the
+form of death without tumor recurrence. For illustration, we fit the state
+learner to the original data set consisting of 1,042 patients. We consider death
+without tumor recurrence and recurrence of tumor as two competing events of
+interest. We include the fives learner \texttt{KM}, \texttt{Cox strata CS},
+\texttt{Lasso}, \texttt{Elastic}, and \texttt{RF} which are described in
+Table~\ref{tab:zel-library}. We use the same library of learners to learn
+\( \Lambda_1 \), \( \Lambda_2 \), and $\Gamma$. In this case, \( \Lambda_1 \)
+denotes the cause-specific cumulative hazard function of tumor recurrence, and
+\( \Lambda_2 \) denotes the cause-specific cumulative hazard function of death
+without tumor recurrence.
+
+\begin{table}
+  \caption{\label{tab:zel-library}Overview of the five learners used by the
+  state learner when applied to the prostate cancer data set. The Kaplan-Meier
+  estimator was fitted using the package \texttt{prodlim}
+  \citep{Gerds_2019prodlim}. All Cox models included all five covariates in
+  the model and were fitted using the package \texttt{survival}
+  \citep{survival-package}. All penalized Cox models included all five
+  covariates as linear predictors and were fitted using the package
+  \texttt{glmnet} \citep{glmnet-cox,glmnet-glm}. The random forest was fitted
+  with the package \texttt{randomForestSRC} \citep{rfsrc-paclage}.}
 \centering
@@
            & \( \texttt{Cox strata CS} \)  & Cox model stratified on CS  \\
-           & \( \texttt{Cox strata HT} \)  & Cox model stratified on HT \\
-           & \( \texttt{Cox spline} \)  & PSA and RD modeled with splines \\ 
+           % & \( \texttt{Cox strata HT} \)  & Cox model stratified on HT \\
+           % & \( \texttt{Cox spline} \)  & PSA and RD modeled with splines \\ 
     Penalized Cox & \texttt{Lasso} & Cox model with \( L_1 \)-norm penalty   \\
-           & \texttt{Ridge} & Cox model with \( L_2 \)-norm penalty \\
+           % & \texttt{Ridge} & Cox model with \( L_2 \)-norm penalty \\
            & \texttt{Elastic} & Cox model with \( L_1 \)- and \( L_2 \)-norm penalty \\
-    Random forest & \texttt{RF} & Random forest with 50 trees and default settings \\
+    Random forest & \texttt{RF} & Random forest with default settings \\
     \bottomrule 
   \end{tabular}
-  \caption[]{Overview of the nine learners used in scenario~2 of the simulation
-    study. The Kaplan-Meier estimator was fitted using the package
-    \texttt{prodlim} \citep{Gerds_2019prodlim}. All Cox models included all five
-    covariates in the model and were fitted using the package \texttt{survival}
-    \citep{survival-package}. All penalized Cox models included all five
-    covariates as linear predictors and were fitted using the package
-    \texttt{glmnet} \citep{glmnet-cox,glmnet-glm}. The random forest was fitted
-    with the package \texttt{randomForestSRC} \citep{rfsrc-paclage}.}
-  \label{tab:zel-library}
 \end{table}
-
-The results are shown in Figure~\ref{fig:zelefski}. We see that the Brier score
-of the IPCW super learner based on a misspecified censoring model (IPCW(KM))
-decreased with sample size, but is larger than that of the other super learners
-for all sample sizes. The state learner and the IPCW super learner based on a
-correctly pre-specified censoring model demonstrate similar performance for all
-sample sizes. The performance of both the IPCW(Cox) super learner and the state
-learner approaches the benchmark provided by the oracle selector for large sample
-sizes.
-\begin{figure}
-  \centerline{\includegraphics[width=1\linewidth]{./figures/zelefski-sim.pdf}}
-  \caption[]{Results for scenario~2 of the simulation study. For the learner
-    selected by each of the four discrete super learners, the Brier score
-    calculated in a large independent data set without censoring is plotted
-    against sample size. The results are based on 200 repetitions.}
-  \label{fig:zelefski}
-\end{figure}
-
-\section{Real data application}
-\label{sec:real-data-appl}
-
-The original prostate cancer data analyzed by \cite{kattan2000pretreatment},
-which we introduced in Section~\ref{sec:numer-exper}, include a competing event
-in the form of death without tumor recurrence. To illustrate our method we fit
-the state learner to the original data set consisting of 1,042 patients. We
-consider death without tumor recurrence and recurrence of tumor as two competing
-events of interest. We include the fives learner \texttt{KM}, \texttt{Cox strata
-  CS}, \texttt{Lasso}, \texttt{Elastic}, and \texttt{RF} which are described in
-Table~\ref{tab:zel-library}. We use the same library of learners to learn
-\( \Lambda_1 \), \( \Lambda_2 \), and $\Gamma$. In this case, \( \Lambda_1 \)
-denotes the cause-specific cumulative hazard function of tumor recurrence, and
-\( \Lambda_2 \) denotes the cause-specific cumulative hazard function of death
-without tumor recurrence.
 
@@
 models, and we repeat training and evaluation five times with different splits.
-The integrated Brier score for all learners are shown in
-Figure~\ref{fig:zelefski-real}, and the top 10 combinations of learners are
-displayed in Table~\ref{tab:zelefski-real}. We see that the prediction
-performance is mostly affected by the choice of learner for the censoring
-distribution. Several combinations of learners give similar performance as
-measured by the integrated Brier score, as long as a random forest is used to
-model the censoring distribution.
+The integrated Brier score (defined in Section~\ref{sec:super-learner-simple})
+for all learners are shown in Figure~\ref{fig:zelefski-real}, and the top 10
+combinations of learners are displayed in Table~\ref{tab:zelefski-real}. We see
+that the prediction performance is mostly affected by the choice of learner for
+the censoring distribution. Several combinations of learners give similar
+performance as measured by the integrated Brier score, as long as a random
+forest is used to model the censoring distribution.
 
 \begin{figure}
-  \centerline{\includegraphics[width=1\linewidth]{./figures/zelefski-real-data.pdf}}
+  \centerline{\includegraphics[width=1\linewidth]{./zelefski-real-data.pdf}}
   \caption[]{The results of applying the 125 combinations of learners to the
@@
 
-\begin{table}[ht]
-  \centering \input{./figures/zel-tab.tex}
-  \caption{The 10 best performing models in terms of integrated Brier score. The
+\begin{table}
+  \caption{\label{tab:zelefski-real}The 10 best performing models in terms of integrated Brier score. The
     reported standard errors are based on five repetitions using different
@@
     learners of \( \Lambda_1 \), \( \Lambda_2 \), and $\Gamma$ as `Tumor
-    learner', `Death leaner', and `Censoring learner', respectively.}
-  \label{tab:zelefski-real}
+    learner', `Death learner', and `Censoring learner', respectively.}
+  \centering
+  \input{./zel-tab.tex}
 \end{table}
+
+As an illustration, we use the learners of the three cumulative hazard functions
+selected by the state learner and the estimator defined in
+Section~\ref{sec:cause-spec-aver}, equation~(\ref{eq:one-step-comp-ate}), to
+estimate the average treatment effect of hormone therapy on risk of tumor
+recurrence and death. The results are shown in
+Figure~\ref{fig:zelefski-real-target} for 6 month intervals after baseline with
+point wise 95\% confidence intervals. We see that hormone therapy appears to
+decrease the risk of tumor recurrence and increase the risk of death without
+tumor recurrence, but that not of the estimated effects are significant.
+
+\begin{figure}
+  \centerline{\includegraphics[width=1\linewidth]{./zelefsky-data-target-par.pdf}}
+  \caption[]{Estimates of the average treatment effect of hormone therapy on the
+    risk of tumor recurrence and death obtained from the prostate cancer study
+    analyzed by \cite{kattan2000pretreatment}. The estimates are based on the
+    estimator defined in equation~(\ref{eq:one-step-comp-ate}) with point wise
+    95\% confidence intervals calculated as described in
+    Section~\ref{sec:targeted-learning}. The estimates of the nuisance
+    parameters are provided by the state learner.}
+  \label{fig:zelefski-real-target}
+\end{figure}
+
 \section{Discussion}
@@
 We have proposed a new super learner that can be used with right-censored data
-and competing events. In this section, we compare our proposal to existing super
-learners and discuss avenues for further research.
-
-\subsection{Existing super learners for right-censored data}
-\label{sec:relat-liter-exist}
-
-Machine learning based on right-censored data commonly use the negative partial
-log-likelihood as loss function
-\citep[e.g.,][]{li2016regularized,yao2017deep,lee2018deephit,katzman2018deepsurv,gensheimer2019scalable,lee2021boosted,kvamme2021continuous}.
-However, this loss function is unsuited for super learning, because many
-canonical survival learners (e.g., the Kaplan-Meier estimator, random survival
-forest, and semi-parametric Cox models) provide cumulative hazard functions that
-are piece-wise constant in the time argument, and hence assign zero probability
-to event times not observed in the training data. This implies that when data
-are observed in continuous time, any of these learners will almost surely have
-infinite loss in any independent hold-out sample according to the negative
-partial log-likelihood loss. When a proportional hazards model is assumed, the
-baseline hazard function can be profiled out of the likelihood to give a new
-partial log-likelihood loss \citep{cox1972regression}, which has been suggested
-as a loss function for super learning
-\citep{golmakani2020super,verweij1993cross}. While this allows the library of
-learners to include Cox' proportional hazard models, the drawback is that the
-library is in fact \textit{only} allowed to include these models. The advantage
-of the state learner is that it does not require evaluation of the density of
-\( F(\blank, j, x) \) and does not assume a particular semi-parametric structure
-for $\Lambda_j$ but can be used with any library of learners.
-
-Another approach for super learning with right-censored data is to use an
-inverse probability of censoring weighted (IPCW) loss function
-\citep{graf1999assessment,van2003unicv,molinaro2004tree,keles2004asymptotically,hothorn2006survival,gerds2006consistent,gonzalez2021stacked}.
-An IPCW loss function is attractive because the associated risk does not depend
-on the censoring distribution but describes a feature of the population of
-interested governed by the measure \( Q \in \mathcal{Q} \). Similar results can
-be obtained using censoring unbiased transformations
-\citep{fan1996local,steingrimsson2019censoring} or pseudo-values
-\citep{andersen2003generalised,mogensen2013random,sachs2019ensemble}. All these
-methods rely on an estimator of the censoring distribution, and their drawback
-is that this estimator has to be pre-specified. When the data-generating
-mechanism is complex and not well-understood, pre-specification of the censoring
-distribution is a challenge. The advantage of using the state learner is that a
-censoring distribution need not be pre-specified but is selected automatically
-based on the provided library \( \mathcal{B} \).
-
-To the best of our knowledge, the only existing attempt at avoiding the need to
-pre-specify a censoring model is a recent proposal suggested independently by
-\cite{han2021inverse} and \cite{westling2021inference}. The authors do not
-consider competing risks but suggest to iterate between learning \( \Lambda \)
-and $\Gamma$ using IPCW loss functions and select the final learner when the
-iterative procedure has converged.
-% Using the estimator \( \hat{b}^{(0)}(\data_n^c) \) of $\Gamma$ based on some
-% initial learner \( \hat{b}^{(0)} \in \mathcal{B} \), we calculate the
-% estimated IPCW risk for each leaner in \( \mathcal{A} \). Picking
-% \( a^{(1)} \) as the discrete super learner according to this IPCW loss
-% function, we then estimate weight
-No general theoretical guarantees exist for this procedure, but it would be
-interesting to compare its performance to that of the state learner in a
-simulation study.
-
-% In a non-survival targeted learning setting \cite{sun2022selective}, based on
-% ideas presented by \cite{robins2007comment}, proposed to use as a model
-% selection criteria the so-called `doubly robustness' property that some
-% targeted estimators enjoy. In many special cases the remainder term
-% \( \mathrm{Rem}{(\hat{\Lambda}_{n}, \hat{\Gamma}_n, P)} \) defined in
-% equation~(\ref{eq:ral}) has the property that
-% \begin{equation}
-%   \label{eq:dr-property}
-%   \mathrm{Rem}{(\Lambda^*, \Gamma_0, P_0)}
-%   = \mathrm{Rem}{(\Lambda_0, \Gamma^*, P_0)} = 0,
-%   \quad \text{for any }
-%   \Lambda^*, \Gamma^* \in \mathcal{M}.
-% \end{equation}
-% The property in equation~(\ref{eq:dr-property}) is known as `doubly
-% robustness' because it ensures that
-% $\tilde{\Psi}(\hat{\Lambda}_{n}, \hat{\Gamma}_n, H_n)$ defined in
-% equation~(\ref{eq:ral}) is consistent if just one of $\Lambda$ or $\Gamma$ is
-% estimated consistently
-% \citep{van2003unified,bang2005doubly,kang2007demystifying}. For
-% $a \in \mathcal{A}$ and \( b \in \mathcal{B} \), define
-% \( \hat{\Psi}^k_n(a, b) = \tilde{\Psi}(a(\data_n^k), b(\data_n^k), \empmeas^k) \)
-% where \( \empmeas^{k} \) is the empirical measure of \( \data_n^{k} \).
-% \cite{sun2022selective} suggest to select $a \in \mathcal{A}$ as the minimizer
-% of the estimated `fluctuation pseudo-risk',
-% \begin{equation*}
-%   \hat R_{\mathrm{f}}(a) =
-%   \max_{b_1, b_2 \in \mathcal{B}}
-%   \frac{1}{K}\sum_{k=1}^{K}
-%   \left\{
-%     \hat{\Psi}^k_n(a, b_1)
-%     - \hat{\Psi}^k_n(a, b_2)
-%   \right\}^2.
-% \end{equation*}
-% The idea is that the doubly robustness property
-% (equation~(\ref{eq:dr-property})) implies that \(\hat{R}_{\mathrm{f}}(a)\) will
-% be zero (asymptotically) if the learner \(a\) correctly estimates $\Lambda$. The
-% authors establish finite sample results guarantying that the selected learner
-% \(a\) will be robust against changing the learner of the other nuisance
-% parameter across $\mathcal{B}$. It seems unclear to what extend this robustness
-% property also guarantees consistency of the estimator of the target parameters
-% when we use the fluctuation risk \(\hat{R}_{\mathrm{f}}\) to select the learners
-% of the nuisance parameters. It would be interesting to explore the performance
-% of this procedure in a survival setting and compare it to targeted estimators
-% obtained using the state learner as outlined in
-% Section~\ref{sec:targeted-learning}.
-
-\subsection{A performance measure of interest}
-\label{sec:perf-meas-inter}
+and competing events. Compared to existing methods, the advantage of the state
+learner is that it does not depend on a pre-specified estimator of the censoring
+distribution, but selects one automatically based on the provided library
+\( \mathcal{B} \). Furthermore, the state learner neither requires evaluation of
+the density of \( \Lambda_j \) nor assumes a particular semi-parametric
+structure for $\Lambda_j$. In the remainder of this section we point to some
+limitations of our proposal and discuss avenues for further research.
 
@@
 
-\fxnote*{mention this?}{[Double robustness maybe lacking]}
-
 % In the context of targeted learning, a drawback of the state learner is that the
@@
 % preserved.
-
-\subsection{Implementation}
-\label{sec:software}
 
@@
 
-% \bibliography{bib.bib}
-
 \appendix
@@
     P{[\bar{B}_{\tau}(F, \blank)]}
-    = \Vert F - F_0 \Vert_{P_0}^2 + P_0{[\bar{B}_{\tau}(F_0, \blank)]}. \qedhere
+    = \Vert F - F_0 \Vert_{P_0}^2 + P_0{[\bar{B}_{\tau}(F_0, \blank)]}.
   \end{equation*}

### 2024-03-20 R-code/functions/sandbox.R

@@ -1,0 +1,314 @@
+
+#### Old versions
+## naiv_est <- function(data, t, Lambda1, Lambda2, jump_points, collapse = TRUE){
+##     Lambda1_ = minus_t_fun(Lambda1,jump_points)
+##     Lambda2_ = minus_t_fun(Lambda2,jump_points)
+##     A_fun = function(newdata,times) {-(Lambda1_(newdata,times) + Lambda2_(newdata,times))}
+##     data_treat0 = copy(data)[, A := 0]
+##     data_treat1 = copy(data)[, A := 1]
+##     preds0 = internal_intergral1(A = A_fun,C = Lambda1,newdata = data_treat0,end = t,jump_points = jump_points)
+##     preds1 = internal_intergral1(A = A_fun,C = Lambda1,newdata = data_treat1,end = t,jump_points = jump_points)
+##     if(collapse)
+##         return(mean(preds1 - preds0))
+##     else
+##         return(data.table(Y1_init_est = preds1, Y1_init_est = preds0))
+## }
+
+## Q1_fun <- function(data, t, Lambda1, Lambda2, Gamma, jump_points, collapse = TRUE, chunks = 1){
+##     Q_out = numeric(nrow(data))
+##     rel_ind = data[,(status>0) & (time <= t)]
+##     if(sum(rel_ind) == 0)
+##         return(Q_out)
+##     wd = copy(data[rel_ind])
+##     Lambda1_ = minus_t_fun(Lambda1,jump_points)
+##     Lambda2_ = minus_t_fun(Lambda2,jump_points)
+##     Gamma_ = minus_t_fun(Gamma,jump_points)
+##     A_fun = function(newdata,times) {-(Lambda1_(newdata,times) + Lambda2_(newdata,times))}
+##     not_at_risk0 = function(newdata, times) {1-at_risk_fun(newdata, times)}
+##     term_w = exp(diag(Lambda1_(wd,wd[,time]) + Lambda2_(wd,wd[,time]) + Gamma_(wd,wd[,time])))
+##     term_int = internal_intergral1(A = A_fun, B = not_at_risk0, C = Lambda1, end = t, newdata = wd, jump_points = jump_points)
+##     Q_out[rel_ind] =  term_w*term_int
+##     if(collapse)
+##         return(mean(Q_out))
+##     else
+##         return(Q_out)
+## }
+
+##
[…]

### 2024-03-20 experiments/sandbox/sandbox.R

@@ -1,0 +1,142 @@
+### sandbox.R --- 
+#----------------------------------------------------------------------
+## Author: Anders Munch
+## Created: Nov 14 2023 (11:11) 
+## Version: 
+## Last-Updated: Jan  6 2024 (11:05) 
+##           By: Anders Munch
+##     Update #: 92
+#----------------------------------------------------------------------
+## 
+### Commentary: 
+## 
+### Change Log:
+#----------------------------------------------------------------------
+## 
+### Code:
+
+library(ggplot2)
+
+tar_load(zel_sim1_results, store = here("experiments/_targets"))
+tar_load(zel_indep_cens_sim_results, store = here("experiments/_targets"))
+
+time_inc <- zel_sim1_results[, diff(unique(time))[1]]
+max_time <- zel_sim1_results[, max(time)[1]]
+int_brier_zel_sim1 <- zel_sim1_results[!(type == "cens" & grepl("ipcw", SL)), .(int_brier = sum(100*brier)*time_inc/max_time), .(n_obs, type, SL, seed)]
+summ_zel_sim1 <- int_brier_zel_sim1[, .(ave_int_brier = mean(int_brier), se = sd(int_brier)/sqrt(.N)), .(n_obs, type, SL)]
+dd_ww <- 0.1
+
+ggplot(summ_zel_sim1,
+       aes(x = n_obs, y = ave_int_brier, col = SL)) +
+    theme_bw() +
+    geom_errorbar(position=position_dodge(width = dd_ww), aes(ymin = ave_int_brier-1.96*se, ymax = ave_int_brier+1.96*se), width = .05, alpha = .5) + 
+    geom_line(position=position_dodge(width = dd_ww)) + geom_point(position=position_dodge(width = dd_ww)) +
+    scale_x_continuous(trans='log2') +
+    facet_wrap(~type, scales = "free_y")
+
+time_inc <- zel_indep_cens_sim_results[, diff(unique(time))[1]]
+max_time <- zel_indep_cens_sim_results[, max(time)[1]]
+int_brier_zel_indep_cens <- zel_indep_cens_sim_results[!(type == "cens" & grepl("ipcw", SL)), .(int_brier = sum(100*brier)*time_inc/max_time), .(n_obs, type, SL, seed)]
[…]

### 2024-03-20 experiments/submit_zel-sim2-1.sh

@@ -1,0 +1,12 @@
+#!/bin/bash
+
+#SBATCH --job-name=zel-sim2-1
+#SBATCH --array=1-1000
+#SBATCH --output=log/stdout-%x-%a.out
+
+# Enable Additional Software
+module load gcc/11.2.0
+module load R/4.1.2
+
+# Run the job commands
+Rscript --vanilla $SLURM_JOB_NAME.R

### 2024-03-20 experiments/zel-sim2-1.R

@@ -1,0 +1,54 @@
+library(data.table)
+options(rf.cores = 1) ## Keep RF on one core
+setDTthreads(1) ## Avoid DT spawning
+
+## Get input
+task_id <- as.numeric(Sys.getenv("SLURM_ARRAY_TASK_ID"))
+max_task_id <- as.numeric(Sys.getenv("SLURM_ARRAY_TASK_MAX"))
+job_name <- as.character(Sys.getenv("SLURM_JOB_NAME"))
+
+## Get seeds
+set.seed(284219)
+seeds <- sample(1:1e6, size = max_task_id, replace = FALSE)
+
+## Load libraries here:
+library(here)
+library(targets)
+tar_source(here("R-code/functions"))
+
+sim_fun <- function(seed){
+    ## Write sim setup here
+    eval_times = seq(6, 36, 6)
+    ns = c(300, 600, 1200, 2400)
+    sim_sets = list(original = simZelefsky_wrapper,
+                    indep_cens = simZelefsky_indep_cens_wrapper)
+    set.seed(seed)
+    raw_calc = do.call(rbind, lapply(seq_along(sim_sets), function(pp){
+        sim_set_name = names(sim_sets)[pp]
+        dat_sim = function(n) sim_sets[[pp]](n = n)
+        true_test = dat_sim(10000)[, !c("time", "status")]
+        dummy_true_test = copy(true_test)[, dummy := 1]
+        do.call(rbind, lapply(ns, function(nn){
+            train0 <- dat_sim(nn)[, !c("true_time", "cens_time")]
+            eval0 = eval_sl(train0, true_test, eval_time = eval_times)
+            eval0[, ":="(n_obs = nn, sim_set = sim_set_name, seed = seed)]
+            ## Get Brier from null model
+            out_null = prodlim(Hist(time, status)~1, data = train0)
+            cens_null = prodlim(Hist(time, !status)~1, data = train0)
+            out_sc = Score(list(out_null), data = dummy_true_test, formula = Hist(true_time, dummy)~1, times = eval_times,null.model = FALSE)$Brier$score
+            cens_sc = Score(list(cens_null), data = dummy_true_test, formula = Hist(cens_time, dummy)~1, times = eval_times,null.model = FALSE)$Brier$score
+            sc0 = rbind(out_sc[, .(time = times, null_brier = Brier, type = "event")],
[…]

### 2024-03-20 sandbox/zelefsky-study/_targets.R

@@ -1,0 +1,176 @@
+### _targets.R --- 
+#----------------------------------------------------------------------
+## Author: Thomas Alexander Gerds
+## Created: Apr 27 2023 (11:06) 
+## Version: 
+## Last-Updated: Mar 13 2024 (09:48) 
+##           By: Anders Munch
+##     Update #: 177
+#----------------------------------------------------------------------
+## 
+### Commentary: 
+## 
+### Change Log:
+#----------------------------------------------------------------------
+## 
+### Code:
+library(here)
+try(setwd(here("zelefsky-study")))
+try(setwd("~/research/SuperVision/Anders/survival-loss/statelearner/zelefsky-study/"))
+library(targets)
+library(tarchetypes)
+library(parallel)
+tar_option_set(packages = c("glmnet","lava","foreach","data.table","prodlim","survival","riskRegression","ranger"))
+tar_source("functions")
+list(
+    tar_target(test_data,draw_data(484)),
+    tar_target(test_fit,{
+        outcome_models <- list(cox0 = coxph(Surv(time,event = event)~1, data=test_data, x = TRUE, y = TRUE),
+                               cox1 = coxph(Surv(time,event = event)~X1+X2, data=test_data, x = TRUE, y = TRUE),
+                               cox2 = coxph(Surv(time,event = event)~X6+X7, data=test_data, x = TRUE, y = TRUE),
+                               cox_dgm = coxph(Surv(time,event = event)~ X1 + X2 + X5 + X6 + X7 + X8 + X11 + X12 + quad.X3 + quad.X4 + quad.X5 + quad.X6,data = test_data,x = TRUE,y = TRUE),
+                               rf = ranger(Surv(time, event = event) ~ X1+X2+X3+X4+X5+X6+X7+X8+X9+X10+X11+X12, data = test_data,num.tree = 50)
+                               ## rf_try = ranger(Surv(time, event = event) ~ X1+X2+X6+X7, mtry = 4, data = test_data),
+                               ## rf_deep = ranger(Surv(time, event = event) ~ X1+X2+X6+X7, mtry = 4, min.node.size = 1, data = test_data)
+                               )
+        cens_models <- list(cox1 = coxph(Surv(time,event = cens_status)~X1+X2, data=test_data, x = TRUE, y = TRUE),
+                            cox2 = coxph(Surv(time,event = cens_status)~X6+X7, data=test_data, x = TRUE, y = TRUE),
[…]

### 2024-03-20 sandbox/zelefsky-study/functions/ipcw-fail-sim.R

@@ -1,0 +1,94 @@
+### ipcw-fail-sim.R --- 
+#----------------------------------------------------------------------
+## Author: Anders Munch
+## Created: Jul 23 2023 (10:29) 
+## Version: 
+## Last-Updated: Aug  3 2023 (08:52) 
+##           By: Anders Munch
+##     Update #: 76
+#----------------------------------------------------------------------
+## 
+### Commentary: 
+## 
+### Change Log:
+#----------------------------------------------------------------------
+## 
+### Code:
+library(data.table)
+library(prodlim)
+library(riskRegression)
+library(lava)
+library(survival)
+
+sim_data <- function(n){
+    m <- lvm()
+    formula_event <- ~f(A,-.8)
+    formula_cens <- ~f(A,.9)
+    event_scale <- 1/500
+    cens_scale <- 1/500
+    lava::distribution(m,~A) <- lava::binomial.lvm(p = .3)
+    lava::distribution(m,~censtime) <- lava::coxWeibull.lvm(scale=event_scale)
+    lava::distribution(m,~eventtime) <- lava::coxWeibull.lvm(scale=cens_scale)
+    m <- lava::eventTime(m, time ~ min(eventtime = 1, censtime = 0), "event")
+    lava::regression(m) <- stats::update(formula_event, "eventtime~.")
+    lava::regression(m) <- stats::update(formula_cens, "censtime~.")
+    out <- setDT(sim(m,n))
+    out[,dummy := 1]
+    return(out[])
+}
+
+sim_ipcw_fail <- function(ns, eval_time = 20, B = 1, split.method = "cv5"){
+    do.call(rbind, lapply(ns, function(nn){
+        sim_dt = sim_data(n = nn)
+        mms = list(dgm = coxph(Surv(time, event)~strata(A), data = sim_dt, x = TRUE, y = TRUE),
+                   km = coxph(Surv(time, event)~1, data = sim_dt, x = TRUE, y = TRUE))
+        forms = list(km = Hist(time,event)~1,
+                     dgm = Hist(time,event)~strat(A))
+        ipcw_briers = lapply(seq_along(forms), function(ii){
+            try_message <- try({
+                ## What we want to happen
+                xx = Score(mms,data = sim_dt,formula = forms[[ii]], metrics = "brier",se.fit = FALSE,times = eval_time,contrasts = FALSE,null.model = FALSE, B = B, split.method = split.method, progress.bar = NULL)$Brier$score
[…]

### 2024-03-20 sandbox/zelefsky-study/functions/statelearner-v2.R

@@ -1,0 +1,263 @@
+### statelearner-v2.R --- 
+#----------------------------------------------------------------------
+## Author: Anders Munch
+## Created: Aug 18 2023 (10:26) 
+## Version: 
+## Last-Updated: Mar 13 2024 (09:14) 
+##           By: Anders Munch
+##     Update #: 365
+#----------------------------------------------------------------------
+## 
+### Commentary: 
+## 
+### Change Log:
+#----------------------------------------------------------------------
+## 
+### Code:
+
+
+## Set data on canonical form
+form_data <- function(data, max_time = Inf, time = "time", status = "status", cause_codes = c("1" = 1, "2" = 2, "c" = 0), vars = NULL){
+    if(!is.null(vars))
+        wd = copy(data)[, c(time, status, vars), with = FALSE]
+    else
+        wd = copy(data)
+    setnames(wd, old = c(time, status), new = c("time", "status"))
+    ## Make cause of interest = 1, everything else = 2
+    wd[, cause1 := 1+1*(status != cause_codes[["1"]] & time <= max_time)]
+    wd[, censor := 1+1*(status != cause_codes[["c"]] & time <= max_time)]
+    if(wd[, length(unique(status))>2])
+        wd[, cause2 := 1+1*(status != cause_codes[["2"]] & time <= max_time)]
+    wd[, status := NULL]
+    return(wd[])
+}
+## Fit model on canonical data form
+fit_cause_model <- function(model, data, cause, x_form = NULL, ...){
+    cause_names = c("cause1", "cause2", "censor")
+    cause_names = cause_names[cause_names %in% names(data)]
+    stopifnot(cause %in% cause_names)
+    wd = copy(data)[, -cause_names[cause_names != cause], with = FALSE]
+    setnames(wd, old = cause, new = "event")
+    ## not so elegant
+    if(is.null(x_form))
+        form = Surv(time, event) ~ .
+    else
+        form = update(x_form, Surv(time, event) ~ .)
+    if(model == "cox"){
+        wd[, event := 1*(event == 1)] ## To make this cause of interest
+        out = coxph(form, data = wd,x = TRUE,y = TRUE, ...)
+    }
+    if(model == "GLMnet"){
+        wd[, event := 1*(event == 1)] ## To make this cause of interest
+        out = GLMnet(form, data = wd, ...)
+    }
+    if(model == "rfsrc"){
+        out = rfsrc(form, data = wd, ...)
+    }
[…]

### 2024-03-20 sandbox/zelefsky-study/functions/zelefsky-data-learners.R

@@ -1,0 +1,275 @@
+### zelefsky-data-learners.R --- 
+#----------------------------------------------------------------------
+## Author: Anders Munch
+## Created: Jul 23 2023 (14:00) 
+## Version: 
+## Last-Updated: Nov 12 2023 (18:18) 
+##           By: Anders Munch
+##     Update #: 136
+#----------------------------------------------------------------------
+## 
+### Commentary: 
+## 
+### Change Log:
+#----------------------------------------------------------------------
+## 
+### Code:
+library(data.table)
+library(prodlim)
+library(riskRegression)
+library(lava)
+library(survival)
+library(randomForestSRC)
+source("./functions/simZelefsky.R")
+sim_learners_zelefsky <- function(ns,simulation_input, eval_time = 36, B = 1, split.method = "cv5"){
+    do.call(rbind, lapply(ns, function(nn){
+        sim_dt = simZelefsky(n = nn,censoring = TRUE,simulation_input = simulation_input)
+        sim_dt[, ":="(time = dmos, event = status)]
+        out_mms = list(km = coxph(Surv(time,event)~1, data=sim_dt, x = TRUE, y = TRUE),
+                       cox_full = coxph(Surv(time,event)~logPSA+stage+ggtot+sDose+hormones, data=sim_dt, x = TRUE, y = TRUE),
+                       cox_int = coxph(Surv(time,event)~pspline(logPSA) + sDose + stage + ggtot+hormones, data=sim_dt, x = TRUE, y = TRUE),
+                       cox_spline = coxph(Surv(time,event)~pspline(logPSA)+stage+ggtot+pspline(sDose)+hormones, data=sim_dt, x = TRUE, y = TRUE),
+                       cox_small = coxph(Surv(time,event)~logPSA+sDose+stage, data=sim_dt, x = TRUE, y = TRUE)
+                       )
+        cens_mms = list(km = coxph(Surv(time,event==0)~1, data=sim_dt, x = TRUE, y = TRUE),
+                        cox_full = coxph(Surv(time,event==0)~logPSA+stage+ggtot+sDose+hormones, data=sim_dt, x = TRUE, y = TRUE),
+                        cox_int = coxph(Surv(time,event==0)~pspline(logPSA) + sDose + stage + ggtot+hormones, data=sim_dt, x = TRUE, y = TRUE),
[…]

### 2024-03-20 sandbox/zelefsky-study/sandbox/play.R

@@ -1,0 +1,46 @@
+### play.R --- 
+#----------------------------------------------------------------------
+## Author: Thomas Alexander Gerds
+## Created: Jul 18 2023 (15:39) 
+## Version: 
+## Last-Updated: Jul 18 2023 (16:08) 
+##           By: Thomas Alexander Gerds
+##     Update #: 1
+#----------------------------------------------------------------------
+## 
+### Commentary: 
+## 
+### Change Log:
+#----------------------------------------------------------------------
+## 
+### Code:
+library(riskRegression)
+library(lava)
+library(data.table)
+library(prodlim)
+m <- lvm()
+formula_event <- ~f(age,0)
+formula_cens <- ~f(age,0)
+event_scale <- 1/1000
+cens_scale <- 1/1000
+lava::distribution(m, ~age) <- lava::normal.lvm(mean = 60,sd = 11)
+lava::distribution(m, ~biomarker) <- lava::normal.lvm(mean = 0,sd = 1)
+lava::regression(m) <- eventtime ~ 0.01*age+0.4*sex
+lava::regression(m) <- censtime ~ 0.03*age-0.4*sex
+lava::regression(m) <- S_event ~ exp(0.4*como-0.5*sex)
+lava::regression(m) <- S_cens ~ exp(-.4*como+0.8*sex)
+lava::distribution(m,~sex+como) <- lava::binomial.lvm()
+lava::distribution(m,~censtime) <- lava::coxWeibull.lvm(scale=event_scale,shape = ~S_cens)
+lava::distribution(m,~eventtime) <- lava::coxWeibull.lvm(scale=cens_scale,shape=~S_event)
+m <- lava::eventTime(m, time ~ min(eventtime = 1, censtime = 0), "event")
+## lava::regression(m) <- stats::update(formula_event, "eventtime~.")
+## lava::regression(m) <- stats::update(formula_cens, "censtime~.")
+d <- setDT(sim(m,1000))
+d[,table(event)]
+pdf("~/tmp/test.pdf",width = 12,height = 6)
+par(mfrow = c(1,2))
+plot(prodlim(Hist(time,event)~sex+age,data = d),xlim = c(0,50),newdata = expand.grid(age = c(20,60,90),sex =c(0,1)),type = "risk",plot.main = "Event probability",confint = FALSE,legend.cex = 0.5)
[…]

### 2024-03-20 sandbox/zelefsky-study/sandbox/sandbox-sl2-comp.R

@@ -1,0 +1,183 @@
+### sandbox-v2.R --- 
+#----------------------------------------------------------------------
+## Author: Anders Munch
+## Created: Aug 18 2023 (16:18) 
+## Version: 
+## Last-Updated: Aug 21 2023 (14:06) 
+##           By: Anders Munch
+##     Update #: 81
+#----------------------------------------------------------------------
+## 
+### Commentary: 
+## 
+### Change Log:
+#----------------------------------------------------------------------
+## 
+### Code:
+## Get CHF for rf
+
+library(randomForestSRC)
+library(data.table)
+data(wihs, package = "randomForestSRC")
+setDT(wihs)
+wihs_fix <- copy(wihs)
+wihs_fix[, time := time + abs(rnorm(.N, sd = 0.01))]
+## dd <- form_data(wihs_fix)
+
+learners1 <- list(cox =  list("cox"),
+                  cox_small =  list("cox", x_form = ~ageatfda + cd4nadir),
+                  rf = list("rfsrc", ntree = 50)
+                  )
+learner_all <- list(cause1 = learners1, cause2 = learners1, censor = learners1)
+
+sl0 <- statelearner2(learner_all, data = wihs_fix, time = 5, integrate = FALSE, verbose = TRUE)
+sl0
+
+zelefsky
+zelefsky[, time := dmos]
+zelefsky[status == 0 & vital == "Dead", status := 2]
+use_dat <- copy(zelefsky)[, .(time,status,logPSA,stage,ggtot,sDose,hormones)]
+zel_learner <- list(
+    ## cox =  list("cox"),
+    cox_lasso =  list("GLMnet", x_form = ~logPSA+stage+ggtot+sDose+hormones), ## Does not work without explicit formula?
+    ## cox_ridge =  list("GLMnet", x_form = ~logPSA+stage+ggtot+sDose+hormones, alpha = 0),
+    cox_elastic =  list("GLMnet", x_form = ~logPSA+stage+ggtot+sDose+hormones, alpha = 0.5),
+    cox_strata_stage = list("cox", x_form = ~logPSA+strata(stage)+ggtot+sDose+hormones),
+    km =  list("cox", x_form = ~ 1),                    
[…]

### 2024-03-20 sandbox/zelefsky-study/sandbox/simple.R

@@ -1,0 +1,47 @@
+library(riskRegression)
+library(survival)
+library(lava)
+library(data.table)
+library(prodlim)
+m <- lvm()
+formula_event <- ~f(age,-0.01)+f(sex,0.3)+f(quad.age,0.01)
+formula_cens <- ~f(age,0)+f(sex,1.8)
+event_scale <- 1/1000
+cens_scale <- 1/1000
+lava::distribution(m, ~age) <- lava::normal.lvm(mean = 60,sd = 11)
+transform(m,quad.age~age) <- function(x)(x$age-60)^2
+lava::distribution(m,~sex) <- lava::binomial.lvm(p = .3)
+lava::distribution(m,~censtime) <- lava::coxWeibull.lvm(scale=event_scale)
+lava::distribution(m,~eventtime) <- lava::coxWeibull.lvm(scale=cens_scale)
+m <- lava::eventTime(m, time ~ min(eventtime = 1, censtime = 0), "event")
+lava::regression(m) <- stats::update(formula_event, "eventtime~.")
+lava::regression(m) <- stats::update(formula_cens, "censtime~.")
+d <- setDT(sim(m,1000))
+d[,table(event)]
+dlarge <- setDT(sim(m,10000))
+dlarge[,dummy := 1]
+pdf("~/tmp/test.pdf",width = 12,height = 6)
+par(mfrow = c(1,2))
+plot(prodlim(Hist(time,event)~sex+age,data = d),xlim = c(0,50),newdata = expand.grid(sex =c(0,1),age = c(20,50,90)),type = "risk",plot.main = "Event probability",confint = FALSE,legend.cex = 0.5)
+plot(prodlim(Hist(censtime,dummy)~sex,data = dlarge),xlim = c(0,50),newdata = expand.grid(sex =c(0,1)),plot.main = "Censoring probability",confint = FALSE,legend.cex = 0.5,type = "risk")
+plot(prodlim(Hist(time,event)~1,data = d,rev = TRUE),xlim = c(0,50),plot.main = "Censoring probability",confint = FALSE,legend.cex = 0.5,add = TRUE,type = "risk",lty = 3)
+dev.off()
+d[,sex_age := 1]
+d[sex == 1,sex_age := age]
+d[,age_sex := 1]
+d[sex == 0,age_sex := age]
+dlarge[,sex_age := 1]
[…]

### 2024-03-20 sandbox/zelefsky-study/sandbox/yet_another.R

@@ -1,0 +1,50 @@
+library(riskRegression)
+library(targets)
+library(survival)
+library(lava)
+library(data.table)
+library(prodlim)
+try(setwd("~/research/SuperVision/Anders/survival-loss/statelearner/empirical-study/"))
+tar_source("functions")
+m <- lvm()
+# B discriminates only when A == 1
+formula_event <- ~f(A,0)+f(B,0)+f(AB,-1.5)+f(AC,-1.3)
+formula_cens <- ~f(A,0)+f(B,2)+f(C,0)
+event_scale <- 1/1000
+cens_scale <- 1/1000
+lava::distribution(m,~A) <- lava::binomial.lvm(p = .4)
+lava::distribution(m,~B) <- lava::normal.lvm()
+lava::distribution(m,~C) <- lava::normal.lvm()
+transform(m,AB~A+B) <- prod
+transform(m,AC~A+C) <- prod
+lava::distribution(m,~censtime) <- lava::coxWeibull.lvm(scale=event_scale)
+lava::distribution(m,~eventtime) <- lava::coxWeibull.lvm(scale=cens_scale)
+m <- lava::eventTime(m, time ~ min(eventtime = 1, censtime = 0), "event")
+lava::regression(m) <- stats::update(formula_event, "eventtime~.")
+lava::regression(m) <- stats::update(formula_cens, "censtime~.")
+d <- setDT(sim(m,5000))
+d[,table(event)]
+dlarge <- setDT(sim(m,5000))
+dlarge[,dummy := 1]
+## pdf("~/tmp/test.pdf",width = 12,height = 6)
+## pdf("~/tmp/test.pdf",width = 10,height = 10)
+## par(mfrow = c(2,2))
+## plot(prodlim(Hist(time,event)~A+B,data = d),xlim = c(0,20),newdata = expand.grid(A =c(0,1),B =c(0,1)),type = "risk",plot.main = "Event probability",confint = FALSE,legend.cex = 0.5)
+## plot(prodlim(Hist(time,event)~A+C,data = d),xlim = c(0,20),newdata = expand.grid(A = c(0,1),C = c(0,1)),type = "risk",plot.main = "Event probability",confint = FALSE,legend.cex = 0.5)
[…]

### 2024-03-20 zelefsky-case-study/_targets.R

@@ -1,0 +1,88 @@
+### _targets.R --- 
+#----------------------------------------------------------------------
+## Author: Anders Munch
+## Created: Mar 13 2024 (09:20) 
+## Version: 
+## Last-Updated: Mar 20 2024 (08:58) 
+##           By: Anders Munch
+##     Update #: 33
+#----------------------------------------------------------------------
+## 
+### Commentary: 
+## 
+### Change Log:
+#----------------------------------------------------------------------
+## 
+### Code:
+library(here)
+try(setwd(here("zelefsky-case-study")))
+try(setwd("~/research/SuperVision/Anders/survival-loss/statelearner/zelefsky-case-study/"))
+library(targets)
+library(tarchetypes)
+library(parallel)
+tar_option_set(packages = c("glmnet","lava","foreach","data.table","prodlim","survival","riskRegression","ranger"))
+tar_source(here("R-code/functions"))
+
+list(
+    tar_target(zelefsky,{
+        zelefsky = setDT(get(load(here("data/zelefsky.rda"))))
+        zelefsky$status=as.numeric(as.character(factor(zelefsky$recur,levels=c("No","Yes"),labels=c("0","1"))))
+        zelefsky$logPSA <- log(zelefsky$psa)
+        zelefsky$sDose <- as.vector(scale(zelefsky$dose))
+        zelefsky$hormonesYes <- zelefsky$hormones=="Yes"
+        for (v in levels(zelefsky$stage)){
+            zelefsky[,paste("stage",v,sep="")] <- as.numeric(zelefsky$stage==v)
+        }
+        zelefsky[]
+    }),
+    tar_target(use_data,{
+        zelefsky[, time := dmos]
+        out = copy(zelefsky)[, .(time,status,logPSA,stage,ggtot,sDose,hormones,vital)]
+        out[status == 0 & vital == "Dead", status := 2][, vital := NULL]
+        out
+    }),
+    tar_target(zelefsky_statelearner,{       
+        zelefsky[, time := dmos]
+        use_dat = copy(zelefsky)[, .(time,status,logPSA,stage,ggtot,sDose,hormones,vital)]
+        use_dat[status == 0 & vital == "Dead", status := 2][, vital := NULL]
+        zel_learner <- list(
+            cox_lasso =  list("GLMnet", x_form = ~logPSA+stage+ggtot+sDose+hormones), ## Does not work without explicit formula?
[…]

### 2024-03-21 tex/jrss-version/state-learner-jrss-style.tex
setup with and without figures and table at end

@@
   amssymb,soul,xcolor,amsmath,graphicx,verbatim,pgfplots,tikz,prodint,booktabs}
+
+\usepackage[tablesonly,heads,nolists]{endfloat}
 
@@
 \newtheorem{theorem}{Theorem}
-\numberwithin{theorem}{section}
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{proposition}[theorem]{Proposition}
-\newtheorem{corollary}[theorem]{Corollary}
+% \numberwithin{theorem}{section}
+\newtheorem{lemma}{Lemma}
+\newtheorem{proposition}{Proposition}
+\newtheorem{corollary}{Corollary}
 
@@
 \bibliographystyle{rss}
+
+%% Toggle with or without figures
+\def\nofig{1}
+
+%% TODO: Update figures with name 'figure-'.
+%% Write macro for including figure or not.
+
+% \newcommand{\includeFigCond}[2]{
+%   \ifx\nofig\undefined %
+%     \includegraphics[#1]{#2} %
+%   \else %
+%     \texttt{#2} %
+%   \fi %
+% }
+\newcommand{\includeFigCond}[2][]{
+  \ifx\nofig\undefined %
+    \includegraphics[#1]{#2} %
+  \else %
+    \texttt{#2} %
+  \fi %
+}
 
@@
 \title[The state learner]{The state learner \\ {-- a super learner for right-censored data}}
-\author[Anders Munch {\it et al.}]{Anders Munch}
+\author{Anders Munch}
 \address{Section of Biostatistics, University of Copenhagen, Copenhagen, Denmark}
 \email{a.munch@sund.ku.dk}
-\author{Thomas A.\ Gerds}
+\author[Munch {\it et al.}]{Thomas A.\ Gerds}
 \address{Section of Biostatistics, University of Copenhagen, Copenhagen, Denmark}
@@
 
-\keywords{Loss based estimation, right-censored data, competing risks, super
-  learner, cross-validation}
+\keywords{Competing risks, cross-validation, loss based estimation,
+  right-censored data, super learner}
 
@@
 \citep{van2003unicv,vaart2006oracle}: The super learner is guaranteed to perform
-almost as well as the model which minimizes the expected performance, i.e., the
+almost as well as the model which minimises the expected performance, i.e., the
 model we would select if we could evaluate the prediction performance in an
@@
 \citep{hjort1992inference,whitney2019comment}. A disadvantage of inverse
-probability of censoring weighted loss functions is that they requires a
+probability of censoring weighted loss functions is that they require a
 pre-specified model for the censoring distribution. \cite{westling2021inference}
@@
 so-called discrete super learner which `combines' the library of
-learners by picking the one that minimizes the cross-validated loss. The state
+learners by picking the one that minimises the cross-validated loss. The state
 learner uses separate libraries to model each competing event and the censoring
@@
 
-The article is organized as follows. We introduce our notation and framework in
+The article is organised as follows. We introduce our notation and framework in
 Section~\ref{sec:framework}. In Section~\ref{sec:super-learning} we define super
@@
 functions on \( [0,\tau] \times\mathcal{X}\). Any distribution
-\( Q \in \mathcal{Q} \) can be characterized by
+\( Q \in \mathcal{Q} \) can be characterised by
 \begin{equation*}
@@
 \( T \independent C \mid X \). This implies that any distribution
-\( P \in \mathcal{P} \) is characterized by a distribution
+\( P \in \mathcal{P} \) is characterised by a distribution
 \( Q \in \mathcal{Q} \) and a conditional cumulative hazard function for \( C \)
@@
 concept of super learning, we simply consider estimation of the function
-\( \Lambda_{j}\). The parameter \(\Psi:\mathcal P\to\mathcal M\) is then
-identified via equation \eqref{eq:lambdaj} by \(\Psi(P)=\Lambda_j\).
+\( \Lambda_{j}\). In this case, the parameter \(\Psi:\mathcal P\to\mathcal M\)
+is identified via equation \eqref{eq:lambdaj} by \(\Psi(P)=\Lambda_j\).
 
@@
 learner applied to the full data set, i.e., \(\hat{a}_n(\data_n)\).
-The oracle learner is defined as the learner that minimizes the
+The oracle learner is defined as the learner that minimises the
 average loss according to the data-generating distribution \( P \),
@@
 \begin{figure}[h]
-  \centering
-  \includegraphics[width=.5\textwidth]{./figure-multi-state-process.pdf}
+  \centering %
+  \includeFigCond[width=.5\textwidth]{figure-multi-state-process.pdf}
   \caption{Illustration of the multi-state process \(\eta\) used by
@@
 equation~(\ref{eq:F-def}). Because each quadruple
-$(\Lambda_{1}, \Lambda_{2}, \Gamma, H)$ characterizes a \(P\in\mathcal P\) which
+$(\Lambda_{1}, \Lambda_{2}, \Gamma, H)$ characterises a \(P\in\mathcal P\) which
 in turn determines \( (F, H) \), a learner for \( F \) can be constructed from
@@
 a strictly proper scoring rule \citep{gneiting2007strictly}. This implies that
-if we minimize the average loss of the integrated Brier score, we recover the
+if we minimise the average loss of the integrated Brier score, we recover the
 parameters of the data-generating distribution. In particular, the oracle of a
@@
 \( \mathcal{F}_n=\mathcal{F}(\mathcal{A}_{1,n}, \mathcal{A}_{2,n},
-\mathcal{B}_n)\) as short-hand notation and to emphasize the
+\mathcal{B}_n)\) as short-hand notation and to emphasise the
 dependence on \( n \).
@@
 where \( \empmeas \) is the empirical measure of a sample \(\{O_i\}_{i=1}^n\).
-We can make the following asymptotic expansion of the one-step estimator
+Under a suitable Donsker class regularity condition we can make the following
+asymptotic expansion of the one-step estimator
 \citep{pfanzagl1985contributions,van2003unified,fisher2021visually,kennedy2022semiparametric,rytgaard2022continuous},
@@
 for some suitable norm \( \|\blank \| \), for instance the
-\( \mathcal{L}_{P}^2 \)-norm. Hence, under a suitable Donsker class regularity
-condition \citep{bickel1993efficient,kennedy2016semiparametric}, when
-equation~(\ref{eq:4}) holds and the nuisance parameters $\Lambda_1$,
-$\Lambda_2$, and $\Gamma$ are consistently estimated at rate
-\( \smallO_P{(n^{-1/4})} \), then
+\( \mathcal{L}_{P}^2 \)-norm. Hence when equation~(\ref{eq:4}) holds and the
+nuisance parameters $\Lambda_1$, $\Lambda_2$, and $\Gamma$ are consistently
+estimated at rate \( \smallO_P{(n^{-1/4})} \), then
 \begin{equation}
@@
 \end{equation*}
-where \( q_{\alpha} \) is the \( (1-\alpha) \)-quantile or the standard normal
-distribution, and
-\begin{equation*}
-  \hat{\sigma}(\data_n) = 
-  \left(
-    \empmeas{
-      \left[
-        \psi(\blank; \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
-        \hat{b}_n(\data_n))^2
-      \right]}
-  \right)^{1/2}.
-\end{equation*}
+where \( q_{\alpha} \) is the \( (1-\alpha) \)-quantile of the standard normal
+distribution and
+\( \hat{\sigma}(\data_n)^2 = \empmeas{ \left[ \psi(\blank;
+    \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n), \hat{b}_n(\data_n))^2
+  \right]} \).
 
@@
 by $\hat{\pi}_n$. We use the empirical measure of \( \{W_1, \dots, W_n\} \) to
-estimate $\mu$, and denote this estimator by the $\hat{\mu}_n$. As parameter of
-interest we consider the standardized difference in the cumulative incidence due
+estimate $\mu$, and denote this estimator by $\hat{\mu}_n$. As parameter of
+interest we consider the standardised difference in the cumulative incidence due
 to cause 1 at time $\tau$, which can be expressed as
@@
 if everyone had been given treatment (\( A=1 \)) compared to if no one had been
-given treatment.
-
-Using equation~(\ref{eq:surv-def}), we may write
-\( \theta_{{\tau}}(Q) = \tilde{\Psi}_{t}^0(\Lambda_1, \Lambda_2, \mu) \),
-where
+given treatment \( (A=0) \).
+
+Using equation~(\ref{eq:surv-def}) we may write
+\( \theta_{{\tau}}(Q) = \tilde{\Psi}_{t}^0(\Lambda_1, \Lambda_2, \mu) \), where
 \begin{equation}
@@
 can typically use the same method to construct a learner of the cumulative
-hazard function of the censoring distribution. This would typically work by
-training the learner on the data set \( \data_n^c \), where
+hazard function of the censoring distribution. This works by training the
+learner on the data set \( \data_n^c \), where
 \( \data_n^c = \{O_i^c\}_{i=1}^n \) with
@@
 Each super learner provides a learner for the cumulative hazard function for the
-outcome of interest, and from this a risk prediction model can be obtained. We
-measure performance of each super learner by calculating the index of prediction
+outcome of interest, and from this a risk prediction model can be obtained
+(c.f., equation~(\ref{eq:surv-def}) with $\Lambda_2 = 0$). We measure
+performance of each super learner by calculating the index of prediction
 accuracy (IPA) \citep{kattan2018index} at a fixed time horizon (36 months) for
@@
 \begin{figure}
-  \centerline{\includegraphics[width=1\linewidth]{./experiment-fig-sl-ipcw.pdf}}
+  \centering %
+  \includeFigCond[width=1\linewidth]{experiment-fig-sl-ipcw.pdf}
   \caption[]{For the risk prediction models provided by each of the super
@@
 \begin{figure}
-  \centerline{\includegraphics[width=1\linewidth]{./experiment-fig-sl-survSL.pdf}}
+  \centering %
+  \includeFigCond[width=1\linewidth]{experiment-fig-sl-survSL.pdf}
   \caption[]{For the risk prediction models of both the outcome and the
@@
     the model and were fitted using the package \texttt{survival}
-  \citep{survival-package}. All penalized Cox models included all five
+    \citep{survival-package}. All penalised Cox models included all five
     covariates as linear predictors and were fitted using the package
@@
   \centering
-  \begin{tabular}{ lll }
+  \begin{tabular}{lll}
     \toprule
@@
            % & \( \texttt{Cox spline} \)  & PSA and RD modeled with splines \\ 
-    Penalized Cox & \texttt{Lasso} & Cox model with \( L_1 \)-norm penalty   \\
+    Penalised Cox & \texttt{Lasso} & Cox model with \( L_1 \)-norm penalty   \\
            % & \texttt{Ridge} & Cox model with \( L_2 \)-norm penalty \\
@@
 \begin{figure}
-  \centerline{\includegraphics[width=1\linewidth]{./zelefski-real-data.pdf}}
+  \centering %
+  \includeFigCond[width=1\linewidth]{zelefski-real-data.pdf}
   \caption[]{The results of applying the 125 combinations of learners to the
@@
 estimate the average treatment effect of hormone therapy on risk of tumor
-recurrence and death. The results are shown in
+recurrence and death. The propensity score is estimated with a lasso model that
+includes all levels of interaction. The results are shown in
 Figure~\ref{fig:zelefski-real-target} for 6 month intervals after baseline with
@@
 decrease the risk of tumor recurrence and increase the risk of death without
-tumor recurrence, but that not of the estimated effects are significant.
+tumor recurrence, but that none of the estimated effects are statistically
+significant from zero.
 
 \begin{figure}
-  \centerline{\includegraphics[width=1\linewidth]{./zelefsky-data-target-par.pdf}}
+  \centering%
+  \includeFigCond[width=1\linewidth]{zelefsky-data-target-par.pdf}
   \caption[]{Estimates of the average treatment effect of hormone therapy on the
@@
 typically interested in is \( \Lambda_j \) or \( S \), which are features of
-\( Q \in \mathcal{Q} \). We emphasize that while the state learner provides us
+\( Q \in \mathcal{Q} \). We emphasise that while the state learner provides us
 with estimates of \( \Lambda_j \) and $\Gamma$ based on libraries
@@
 is more involved, but we expect that a good approximation can be achieved by
-discretization. In the future, we intend to investigate the performance of the
+discretisation. In the future, we intend to investigate the performance of the
 state learner when using a broader library of learners in more comprehensive
@@
   By definition of the oracle and Lemma~\ref{lemma:norm},
-  \( \E_{P_0}{\left[ \Vert \tilde{\phi}_n(\data_n^{-k}) - F_0 \Vert_{P_0}^2 \right]}
-  \leq \E_{P_0}{\left[ \Vert \phi_n(\data_n^{-k}) - F_0 \Vert_{P_0}^2 \right]} \) for
-  all \( n \in \N \). The results then follows from
+  \begin{equation*}
+    \E_{P_0}{\left[ \Vert \tilde{\phi}_n(\data_n^{-k}) - F_0 \Vert_{P_0}^2
+      \right]} \leq \E_{P_0}{\left[ \Vert \phi_n(\data_n^{-k}) - F_0 \Vert_{P_0}^2
+      \right]}  
+  \end{equation*}
+  for all \( n \in \N \). The results then follows from
   Corollary~\ref{cor:oracle-prop}.
@@
 
-In this section show that a product structure is preserved when the estimator
+In this section show that a product structure is preserved when an estimator
 $\bar\Psi(\hat{F}_n, \hat{H}_n)$ is used instead of

### 2024-03-21 tex/jrss-version/state-learner-jrss-style.tex
changed notation to -1 for censoring
change k to j in appendix

@@
 %% Toggle with or without figures
-\def\nofig{1}
+% \def\nofig{1}
 
@@
 \begin{equation*}
-  \eta(t) = \1\{\tilde{T} \leq t, \tilde D=1\} + 2\,\1\{\tilde{T} \leq t, \tilde D=2\}+ 3\,\1\{\tilde{T} \leq t, \tilde D=0\},
+  \eta(t) = \1\{\tilde{T} \leq t, \tilde D=1\} + 2\,\1\{\tilde{T} \leq t, \tilde
+  D=2\} - \1\{\tilde{T} \leq t, \tilde D=0\},
   \quad \text{for} \quad t \in [0, \tau].
@@
 The conditional distribution of \( \eta(t) \) given \( X=x \) is determined by
-the \fxnote*{reconsider letting censoring state be \( -1 \)?}{function}
+the function
 \begin{equation}
@@
   t \in [0,\tau],
-  k \in \{0,1,2,3\},
+  k \in \{-1,0,1,2\},
   x \in \mathcal{X}.
@@
   \\
-  F(t, 3, x)
+  F(t, -1, x)
   & =
@@
     = P(\tilde{T} > t \mid X= x)
-    = 1- F(t, 1, x) - F(t, 2, x)- F(t, 3, x).
+    = 1- F(t, 1, x) - F(t, 2, x)- F(t, -1, x).
   \end{split}
@@
   \phi_{a_1,a_2, b}(\data_n)(t,2,x) &= \int_0^t e^{\{-a_1(\data_n)(s \mid x)-a_2(\data_n)(s \mid x) - b(\data_n)(s \mid x)\} }  a_2(\data_n)(\diff s \mid x),\\
-  \phi_{a_1,a_2, b}(\data_n)(t,3,x) &= \int_0^t e^{\{-a_1(\data_n)(s \mid x)-a_2(\data_n)(s \mid x) - b(\data_n)(s \mid x)\} }  b(\data_n)(\diff s \mid x),\\
-  \phi_{a_1,a_2, b}(\data_n)(t,0,x) &= 1-  \sum_{j=1}^3\phi_{a_1,a_2, b}(\data_n)(t,j,x).
+  \phi_{a_1,a_2, b}(\data_n)(t,-1,x) &= \int_0^t e^{\{-a_1(\data_n)(s \mid x)-a_2(\data_n)(s \mid x) - b(\data_n)(s \mid x)\} }  b(\data_n)(\diff s \mid x),\\
+  \phi_{a_1,a_2, b}(\data_n)(t,0,x) &= 1
+                                      - \phi_{a_1,a_2, b}(\data_n)(t,1,x)
+                                      - \phi_{a_1,a_2, b}(\data_n)(t,2,x)
+  \\
+  & \qquad
+    -\phi_{a_1,a_2, b}(\data_n)(t,-1,x)
 \end{align*}
@@
 \begin{equation*}
-  B_t(F,O) = \sum_{j=0}^{3}
+  B_t(F,O) = \sum_{j=-1}^{2}
   \left(
@@
 Then \( F^*(t,j,\blank) = F_0(t,j,\blank) \) \( H \)-almost surely for
-any \( j\in \{0,1,2,3\} \) and almost any \( t \in [0, \tau]\).
+any \( j\in \{-1,0,1,2\} \) and almost any \( t \in [0, \tau]\).
 \end{proposition}
@@
   \left\{
-    \sum_{j=0}^{3}\int_{\mathcal{X}} \int_0^{\tau} F(t, j, x)^2 \diff t H_0( \diff x)
+    \sum_{j=-1}^{2}\int_{\mathcal{X}} \int_0^{\tau} F(t, j, x)^2 \diff t H_0( \diff x)
   \right\}^{1/2}.
@@
   \Gamma(t , x) 
-  = \int_0^t  \frac{F(\diff s, 3, x )}{F(s-, 0, x )},
+  = \int_0^t  \frac{F(\diff s, -1, x )}{F(s-, 0, x )},
   \quad \text{and} \quad
@@
 \begin{proof}
-  For any \( t \in [0, \tau] \) and \( k\in \{0,1,2\} \) we have
+  For any \( t \in [0, \tau] \) and \( j\in \{-1,0,1,2\} \) we have
   \begin{align*}
-    & \E_{P_0}{\left[ (F(t, k, X) - \1{\{\eta(t) = k \}})^2 \right]}
-    \\
-    & =    \E_{P_0}{\left[ (F(t, k, X) - F_0(t, k, X) + F_0(t, k, X) - \1{\{\eta(t) = k
+    & \E_{P_0}{\left[ (F(t, j, X) - \1{\{\eta(t) = j \}})^2 \right]}
+    \\
+    & =    \E_{P_0}{\left[ (F(t, j, X) - F_0(t, j, X) + F_0(t, j, X) - \1{\{\eta(t) = j
       \}})^2 \right]}
     \\
-    & =    \E_{P_0}{\left[ (F(t, k, X) - F_0(t, k, X))^2\right]}
-      + \E_{P_0}{\left[ (F_0(t, k, X) - \1{\{\eta(t) = k \}})^2\right]}
+    & =    \E_{P_0}{\left[ (F(t, j, X) - F_0(t, j, X))^2\right]}
+      + \E_{P_0}{\left[ (F_0(t, j, X) - \1{\{\eta(t) = j \}})^2\right]}
     \\
     & \quad
-      + 2\E_{P_0}{\left[ (F(t, k, X) - F_0(t, k, X))(F_0(t, k, X) - \1{\{\eta(t) = k
+      + 2\E_{P_0}{\left[ (F(t, j, X) - F_0(t, j, X))(F_0(t, j, X) - \1{\{\eta(t) = j
       \}})\right]}
     \\
-    & =    \E_{P_0}{\left[ (F(t, k, X) - F_0(t, k, X))^2\right]}
-      + \E_{P_0}{\left[ (F_0(t, k, X) - \1{\{\eta(t) = k \}})^2\right]},
+    & =    \E_{P_0}{\left[ (F(t, j, X) - F_0(t, j, X))^2\right]}
+      + \E_{P_0}{\left[ (F_0(t, j, X) - \1{\{\eta(t) = j \}})^2\right]},
   \end{align*}
@@
 \end{align*}
-so using this with \( a=F(t, k, x) \), \( b=F_0(t, k, x) \), and
-\( c = \1{\{\eta(t) = k\}} \), we have by Jensen's inequality
+so using this with \( a=F(t, j, x) \), \( b=F_0(t, j, x) \), and
+\( c = \1{\{\eta(t) = j\}} \), we have by Jensen's inequality
 \begin{align*}
@@
     2\tau\E_{P_0}{\left[
-    \sum_{k=0}^{3} \int_0^{\tau}
+    \sum_{j=-1}^{2} \int_0^{\tau}
     \left\{
     \left(
-    F(t, k, X) - \1{\{\eta(t) = k\}}
+    F(t, j, X) - \1{\{\eta(t) = j\}}
     \right)^2
@@
     \left(
-    F_0(t, k, X) - \1{\{\eta(t) = k\}}
+    F_0(t, j, X) - \1{\{\eta(t) = j\}}
     \right)^2
@@
     \E_{P_0}\Bigg[
-    \sum_{k=0}^{3} \int_0^{\tau}
-    \left(
-    F(t, k, X) - F_0(t, k, X)
+    \sum_{j=-1}^{2} \int_0^{\tau}
+    \left(
+    F(t, j, X) - F_0(t, j, X)
     \right)^2
@@
     \left\{
-    F(t, k, X) +  F_0(t, k, X)-2 \1{\{\eta(t) = k\}}
+    F(t, j, X) +  F_0(t, j, X)-2 \1{\{\eta(t) = j\}}
     \right\}^2
@@
     8\tau \E_{P_0}{\left[
-    \sum_{k=0}^{3} \int_0^{\tau}
-    \left(
-    F(t, k, X) - F_0(t, k, X)
+    \sum_{j=-1}^{2} \int_0^{\tau}
+    \left(
+    F(t, j, X) - F_0(t, j, X)
     \right)^2

### 2024-04-29 tex/jrss-version/state-learner-jrss-style.tex

@@
 \begin{abstract}
-  In survival analysis, prediction models are needed as stand-alone tools and in
-  applications of causal inference to estimate nuisance parameters. The super
-  learner is a machine learning algorithm which combines a library of prediction
-  models into a meta learner based on cross-validated loss. Unfortunately, the
-  commonly used partial likelihood loss is not suited for super learning, and
-  inverse probability of censoring weighted loss functions require a
-  pre-specified estimator of the censoring distribution. To relax this, we
-  introduce the state learner, a new super learner for survival analysis, which
-  evaluates the loss based on the observed data simultaneously using libraries
-  of predictions models for the event(s) of interest and the censoring
-  distribution. We establish an oracle inequality for the state learner and
+  In survival analysis, prediction models are needed as stand-alone
+  tools and in applications of causal inference to estimate nuisance
+  parameters. The super learner is a machine learning algorithm which
+  combines a library of prediction models into a meta learner based on
+  cross-validated loss. In right censored data, loss needs to be estimated somehow.
+  % based on a model for the censoring distribution.
+  % Unfortunately, the
+  % commonly used partial likelihood loss is not suited for super learning, and
+  % inverse probability of censoring weighted loss functions require a
+  % pre-specified estimator of the censoring distribution. To relax this,
+  We introduce the state learner, a new super learner for survival analysis, which
+  simultaneously evaluates libraries
+  of prediction models for the event of interest and the censoring
+  distribution. The state learner can
+be applied to all types of survival models, works in the presence
+of competing risks, and does not require a single pre-specified
+estimator of the conditional censoring distribution.
+  We establish an oracle inequality for the state learner and
   investigate its performance through numerical experiments. We illustrate how
@@
 
-A super learner is a machine learning algorithm that combines a finite
-set of learners into a meta learner by estimating prediction
-performance in hold-out samples using a pre-specified loss function
-\citep{van2007super}. When the aim is to make a prediction model,
-super learners typically combine strong learners, such as Cox
-regression models and random survival forests
+Medical decision making is traditionally based on the discrepancy
+between intuitive jugdment and the law of chance
+\citep{redelmeier1995probability}. We are concerned with the
+statistical models that provide probabilistic predictions about the
+medical future of a patient. We have two different applications of
+these models in mind. One is as a stand-alone tool where the model
+provides information for a freshly diagnosed patient. The other is as
+input to an algorithm which estimates a causal target parameter such
+as an average treatment effect. The models are learned on data from
+previous patients. A super learner is a machine learning algorithm
+that combines a finite set of learners into a meta learner by
+estimating prediction performance in hold-out samples using a
+pre-specified loss function \citep{van2007super}. When the aim is to
+make a prediction model, super learners combine strong learners, such
+as Cox regression models and random survival forests
 \citep{gerds2021medical}.
@@
 % \citep{van2011targeted, benkeser2016highly,van2017generally}.
-While the general idea of combining strong learners based on cross-validation
-data stems from earlier work \citep{wolpert1992stacked,breiman1996stacked}, the
-name super learner is justified by an oracle inequality
-\citep{van2003unicv,vaart2006oracle}: The super learner is guaranteed to perform
-almost as well as the model which minimises the expected performance, i.e., the
-model we would select if we could evaluate the prediction performance in an
-infinite hold-out sample.
-
-We are concerned with the choice of the loss function for super learning in
-survival analysis. Existing super learner algorithms for right-censored data use
-partial log-likelihood loss or inverse probability of censoring weighted loss
+
+While the general idea of combining strong learners based on
+cross-validation data stems from earlier work
+\citep{wolpert1992stacked,breiman1996stacked}, the name super learner
+is justified by an oracle inequality
+\citep{van2003unicv,vaart2006oracle}: The super learner is guaranteed
+to perform almost as well as the model which minimises the expected
+performance, i.e., the model we would select if we could evaluate the
+prediction performance in an infinite hold-out sample. The choice of
+the loss function for super learning in survival analysis deserves a
+closer look. Existing super learner algorithms for right-censored data
+use partial log-likelihood loss or inverse probability of censoring
+weighted loss
 \citep{polley2011-sl-cens,keles2004asymptotically,golmakani2020super,westling2021inference}.
-The use of the partial log-likelihood loss restricts the class of learners and
-excludes for example simple Kaplan-Meier based learners and also more complex
-random survival forest algorithms. For this reason \cite{golmakani2020super}
-restrict their learners to Cox proportional hazard models. A lesser known fact
-is that a super learner constructed with the negative partial log-likelihood
-loss implicitly depends on the censoring distribution
-\citep{hjort1992inference,whitney2019comment}. A disadvantage of inverse
-probability of censoring weighted loss functions is that they require a
-pre-specified model for the censoring distribution. \cite{westling2021inference}
-tackle this challenge by iterating between super learning of the censoring
-distribution and the event time distribution.
+The use of the partial log-likelihood loss, however, restricts the
+class of learners and excludes for example simple Kaplan-Meier based
+learners and also more complex random survival forest algorithms. For
+this reason \cite{golmakani2020super} restrict their learners to Cox
+proportional hazard models. A lesser known fact is that a super
+learner constructed with the negative partial log-likelihood loss
+implicitly depends on the censoring distribution
+\citep{hjort1992inference,whitney2019comment}. Inverse probability of
+censoring weighted loss has been studied for decades
+\citep{graf1999assessment, hothorn2006survival, gerds2006consistent}. A disadvantage of inverse probability of
+censoring weighted loss estimators is that they require a
+pre-specified model for the censoring
+distribution. In \cite{westling2021inference} this problem is tackled by
+iterating between super learning of the censoring distribution and the
+event time
+distribution. % ANDERS: mangler afrundning efter sidste saetning. eller maaske kan saetningen flyttes laenger ned?
 
 In this article we define the state learner, a new super learner for
-right-censored data, which simultaneously evaluates the loss for learners of the
-event time distribution and the censoring distribution. The loss function which
-is used to define the state learner is only based on observable
-quantities. The state learner can be applied to all types of survival
-estimators, works in the presence of competing risks, and does not require a
-single pre-specified estimator of the conditional censoring distribution. To
-analyze the theoretical properties of the state learner we focus on the
-so-called discrete super learner which `combines' the library of
-learners by picking the one that minimises the cross-validated loss. The state
-learner uses separate libraries to model each competing event and the censoring
-distribution. We show that the oracle selector of the state learner is
-consistent if all libraries contain a consistent learner and prove a finite
-sample oracle inequality.
-
-The state learner can be used to select a model which predicts the probability
-of an event based on covariates in the presence of competing risks. Another
-application is in targeted learning where conditional event probabilities occur
-as high-dimensional nuisance parameters which need to be estimated at a certain
-rate \citep{van2011targeted, rytgaard2021estimation, rytgaard2022targeted}. We
-show how a targeted estimator can be obtained from the state learner, and that a
-second order product structure for the asymptotic bias term of the targeted
-estimator is retained when the state learner is used to estimate nuisance
-parameters.
-
-The article is organised as follows. We introduce our notation and framework in
-Section~\ref{sec:framework}. In Section~\ref{sec:super-learning} we define super
-learning in general with right-censored data, and in
-Section~\ref{sec:relat-liter-exist} we review existing methods.
-Section~\ref{sec:super-learner-simple} introduces our novel proposal, the state
-learner, and Section~\ref{sec:theor-results-prop} provides theoretical
-guarantees for the state learner. In Section~\ref{sec:targeted-learning} we
-discuss the use of the state learner in the context of targeted learning. We
-report some numerical experiments in Section~\ref{sec:numer-exper} and analyze a
-prostate cancer data set in Section~\ref{sec:real-data-appl}.
-Section~\ref{sec:discussion} contains a discussion of the merits and limitations
-of our proposal. Appendices~\ref{sec:proof-proposition}
+right-censored data, which simultaneously estimates the expected loss
+of learners of the event time distribution and the censoring
+distribution. The loss function which is used to define the state
+learner is based on the observable
+quantities. % ANDERS: hvad mener vi med observable quantities? distribution of the observed random variables?
+The state learner can be applied to all types of survival models,
+works in the presence of competing risks, and does not require a
+single pre-specified estimator of the conditional censoring
+distribution.  To analyze the theoretical properties of the state
+learner we focus on the discrete super learner which combines the
+library of learners by picking the one that minimises the
+cross-validated loss
+\citep{van2011targeted}. % ANDERS: er der bedre referencer fra Mark paper/bog for discrete super learner?
+In the presence of competing risks, our algorithm uses separate
+libraries to model the hazard rates of the competing risks and the
+censoring distribution. We show that the oracle selector of the state
+learner is consistent if all libraries contain a consistent learner
+and prove a finite sample oracle inequality. The result of the state
+learner algorithm is a model which predicts the probability of an
+event based on covariates in the presence of competing risks
+\citep{gerds2021medical}. The other application is in targeted
+learning where conditional event probabilities occur as
+high-dimensional nuisance parameters which need to be estimated at a
+certain rate \citep{van2011targeted, rytgaard2021estimation,
+  rytgaard2022targeted}. We show for a targeted estimator, which uses
+the state learner to estimate nuisance parameters, that a second order
+product structure holds for the asymptotic bias term.
+
+The article is organised as follows. We introduce our notation and
+framework in Section~\ref{sec:framework}. In
+Section~\ref{sec:super-learning} we define general super learning
+for right-censored data, and in Section~\ref{sec:relat-liter-exist}
+we review existing methods.  Section~\ref{sec:super-learner-simple}
+introduces our novel proposal, the state learner, and
+Section~\ref{sec:theor-results-prop} provides theoretical guarantees
+for the state learner. In Section~\ref{sec:targeted-learning} we
+discuss the use of the state learner in the context of targeted
+learning. We report results of our numerical experiments in
+Section~\ref{sec:numer-exper} and analyze a prostate cancer data set
+in Section~\ref{sec:real-data-appl}.  Section~\ref{sec:discussion}
+contains a discussion of the merits and limitations of our
+proposal. Appendices~\ref{sec:proof-proposition}
 and~\ref{sec:state-learner-with} contain proofs.
@@
 For ease of notation we assume throughout that \( \Lambda_j(\blank \mid x) \) is
-continuous for all \( x \) and \( j \). We denote by \(S\) the conditional
-event-free survival function,
+continuous for all \( x \) and \( j \). %ANDERS: need a note here which says that this is not a limitation?
+We denote by \(S\) the conditional
+event-free survival function:
 \begin{equation}
@@
 \end{equation}
-Let \( \mathcal{M} \) denote the space of all conditional cumulative hazard
+% ANDERS: I prefer to have the \tau explicit
+Let \( \mathcal{M}_{\tau}\) denote the space of all conditional cumulative hazard
 functions on \( [0,\tau] \times\mathcal{X}\). Any distribution
@@
 \end{equation*}
-where \(\Lambda_{j} \in \mathcal{M}\) for \(j=1,2\) and \(H\) is the marginal
+where \(\Lambda_{j} \in \mathcal{M}_{\tau}\) for \(j=1,2\) and \(H\) is the marginal
 distribution of the covariates.
 
-We consider the usual right-censored setting in which we observe data
+We consider the right-censored setting in which we observe data
 \(O = (\tilde{T},\tilde D, X)\), where $\tilde T = \min(T,C)$ for a
 right-censoring time \(C\), $\Delta = \1{\{T \leq C\}}$, and
-\(\tilde D=\Delta D\). Let \(\mathcal{P}\) denote a set of probability measures
-on the sample space
-\(\mathcal{O} = [0, \tau] \times \{0, 1, 2\} \times \mathcal{X}\) such that
-\(O \sim P \) for some unknown \(P\in \mathcal{P}\). We assume that the event
-time and the censoring time are conditionally independent given covariates,
-\( T \independent C \mid X \). This implies that any distribution
-\( P \in \mathcal{P} \) is characterised by a distribution
-\( Q \in \mathcal{Q} \) and a conditional cumulative hazard function for \( C \)
-given \( X \) \citep[c.f.,][]{begun1983information,gill1997coarsening}. We use
-\(\Gamma\in\mathcal M\) to denote the conditional cumulative hazard function for
-censoring. We assume that \( \Gamma(\blank \mid x) \) is continuous for all
-\( x \), and let \(G(t \mid x)=\exp\left\{-\Gamma(t \mid x)\right\}\) denote the
-survival function of the conditional censoring distribution. In our setting with
-competing risks, this yields
+\(\tilde D=\Delta D\). Let \(\mathcal{P}\) denote a set of probability
+measures on the sample space
+\(\mathcal{O} = [0, \tau] \times \{0, 1, 2\} \times \mathcal{X}\) such
+that \(O \sim P \) for some unknown \(P\in \mathcal{P}\). We assume
+that the event times and the censoring times are conditionally
+independent given covariates, \( T \independent C \mid X \). This
+implies that any distribution \( P \in \mathcal{P} \) is characterised
+by a distribution \( Q \in \mathcal{Q} \) and a conditional cumulative
+hazard function for \( C \) given \( X \)
+\citep[c.f.,][]{begun1983information,gill1997coarsening}. We use
+\(\Gamma\in\mathcal M\) to denote the conditional cumulative hazard
+function for censoring. % ANDERS: can we not allow Gamma to be discrete?
+For notational convenience, we assume that \( \Gamma(\blank \mid x) \) is
+continuous for all \( x \), and let
+\(G(t \mid x)=\exp\left\{-\Gamma(t \mid x)\right\}\) denote the
+survival function of the conditional censoring distribution. The distribution 
+is characterized by
 \begin{equation}\label{eq:parametrizeP}
@@
 \end{equation}
+% ANDERS: H is missing in the model?
 Hence, we may write
 \( \mathcal{P} = \{ P_{Q, \Gamma} : Q \in \mathcal{Q}, \Gamma \in
-\mathcal{G} \} \) for some \( \mathcal{G} \subset \mathcal{M} \). We
+\mathcal{G} \} \) for some \( \mathcal{G} \subset \mathcal{M}_{\tau} \). We
 also have
@@
 
-\section{Super learning with right-censored survival data}
+\section{The concept of super learning}
 \label{sec:super-learning}
 
-A super learner estimates a parameter $\Psi$ which can be identified from the
-observed data distribution \(P\in\mathcal P\). In this section, to introduce the
-concept of super learning, we simply consider estimation of the function
-\( \Lambda_{j}\). In this case, the parameter \(\Psi:\mathcal P\to\mathcal M\)
-is identified via equation \eqref{eq:lambdaj} by \(\Psi(P)=\Lambda_j\).
-
-As input to the super learner we need a sample \( \data_n=\{O_i\}_{i=1}^n \) of
-i.i.d.\ observations from some unknown \( P \in \mathcal{P} \) and a finite
-collection of candidate learners $\mathcal{A}$. Each learner
-\(a \in \mathcal{A}\) is a map
-\( a \colon \mathcal{O}^n \rightarrow \mathcal{M}\) which takes a data set as
-input and returns an estimate $a(\data_n) \in \mathcal{M}$ of $\Lambda_{j}$.
+In survival analysis, a super learner can be used to estimate a
+parameter $\Psi$ which can be identified from the observed data
+distribution \(P\in\mathcal P\). In this section, to introduce the
+discrete super learner and the oracle learner, we consider estimation of the
+function-valued parameter \(\Psi:\mathcal P\to\mathcal M\), given by
+\(\Psi(P)=\Lambda_j\). This parameter is identified via equation
+\eqref{eq:lambdaj} on \([0,\tau]\).
+
+As input to the super learner we need a data set
+\( \data_n=\{O_i\}_{i=1}^n \) of i.i.d.\ observations from \( P \in \mathcal{P} \) and a finite collection of candidate
+learners $\mathcal{A}$. Each learner \(a \in \mathcal{A}\) is a map
+\( a \colon \mathcal{O}^n \rightarrow \mathcal{M}_{\tau}\) which takes a data
+set as input and returns an estimate $a(\data_n) \in \mathcal{M}_{\tau}$ of
+$\Lambda_{j}$.
 % Formally, the domain of $a$ depends on \(n\) but we suppress this in the notation.
@@
 \(P[f] = \int f(o) P(\diff o) \). A super learner evaluates the
-performance of \(a \in \mathcal{A}\) using a loss function
-\(L\colon \mathcal{M} \times \mathcal{O} \rightarrow \R_+\) by
-estimating the expected loss \(P[L(a(\data_n), \blank)]\) using
+performance of \(a \in \mathcal{A}\) with a loss function
+\(L\colon \mathcal{M}_{\tau} \times \mathcal{O} \rightarrow \R_+\) and
+estimates the expected loss \(P[L(a(\data_n), \blank)]\) using
 cross-validation. Specifically, the expected loss of $a\in\mathcal A$
@@
 \end{equation*}
-The final estimator of \(\Psi(P)=\Lambda_j\) is then the selected
-learner applied to the full data set, i.e., \(\hat{a}_n(\data_n)\).
+% The final estimator of \(\Psi(P)=\Lambda_j\) is then the selected
+% learner applied to the full data set, i.e., \(\hat{a}_n(\data_n)\).
 The oracle learner is defined as the learner that minimises the
@@
 \end{equation*}
-Note that both the discrete super learner and the oracle learner depend on the
-library of learners and on the number of folds \(K\), and that the oracle
-learner is a function of the data and the unknown data-generating distribution.
-These dependencies are suppressed in the notation.
+Note that both the discrete super learner and the oracle learner
+depend on the library of learners and on the number of folds \(K\),
+and that the oracle learner is a function of the data and the unknown
+data-generating distribution. However, these dependencies are
+suppressed in the notation.
 
@@
 
-Machine learning based on right-censored data commonly use the negative partial
-log-likelihood as loss function
+Machine learning based on right-censored data commonly use the
+negative partial log-likelihood as loss function
 \citep[e.g.,][]{li2016regularized,yao2017deep,lee2018deephit,katzman2018deepsurv,gensheimer2019scalable,lee2021boosted,kvamme2021continuous}.
-However, this loss function is unsuited for super learning, because many
-canonical survival learners (e.g., the Kaplan-Meier estimator, random survival
-forest, and semi-parametric Cox models) provide cumulative hazard functions that
-are piece-wise constant in the time argument, and hence assign zero probability
-to event times not observed in the training data. This implies that when data
-are observed in continuous time, any of these learners will almost surely have
-infinite loss in any independent hold-out sample according to the negative
-partial log-likelihood loss.
-
-When a proportional hazards model is assumed, the baseline hazard function can
-be profiled out of the likelihood to give a new partial log-likelihood loss
-\citep{cox1972regression}, which has been suggested as a loss function for super
-learning \citep{golmakani2020super,verweij1993cross}. While this allows the
-library of learners to include Cox' proportional hazard models, the drawback is
-that the library is now restricted to include only these models.
+It seems however that this loss function is not perfectly suited for
+super learning, because it cannot be applied to important survival
+learners: The Kaplan-Meier estimator, the random survival forest, and
+the semi-parametric Cox regression model provide cumulative hazard
+functions that are piece-wise constant in the time argument, and hence
+assign zero probability to event times not observed in the training
+data. This implies that when event times are observed in continuous
+time, any of these learners will almost surely have infinite loss in
+an independent test set according to the negative partial
+log-likelihood loss.
+% ANDERS: would be good to have a formula for a the negative partial log-likelihood loss showing that the result is infitity
+
+When a proportional hazards model is assumed, the baseline hazard
+function can be profiled out of the likelihood to give a new partial
+log-likelihood loss \citep{cox1972regression}, which has been
+suggested as a loss function for super learning
+\citep{golmakani2020super,verweij1993cross}. While this allows the
+library of learners to include Cox' proportional hazard models, the
+drawback is that the library is now restricted to include only these
+models. % ANDERS: can the library really include multiple proportional hazards models? only one of them can be correct ...
+% also, would be good to merge this paragraph with the previous somehow
 
@@
 
-As far as we are aware, the only existing attempt at avoiding the need to
-pre-specify a censoring model is a recent proposal suggested independently by
-\cite{han2021inverse} and \cite{westling2021inference}. The authors do not
-consider competing risks but suggest to iterate between learning \( \Lambda \)
-and $\Gamma$ using IPCW loss functions and select the final learner when the
-iterative procedure has converged. To the best of our knowledege, no general
-theoretical guarantees exist for this procedure. It seems unclear whether this
-iterative procedure will always be able to find a correctly specified model in a
-given library.
+An attempt to avoid a pre-specified censoring model was proposed
+independently by \cite{han2021inverse} and
+\cite{westling2021inference}. In both articles, the authors suggest to
+iterate between learning of \( \Lambda \) and learning of $\Gamma$
+using IPCW loss functions. However, no general theoretical guarantees
+seem to exist for this procedure, and it has not yet been extended to
+the situation with competing risks.
 
@@
 
-The problem with most existing super learners for right-censored data is that
-they depend on a pre-specified estimator of the censoring distribution. The main
-idea of the state learner is to jointly use learners of \( \Lambda_1 \),
-\( \Lambda_2 \), and \( \Gamma \), and the relations in
-equation~(\ref{eq:parametrizeP}), to learn a feature of the observed data
-distribution \( P \). The discrete state learner ranks a tuple of learners of
-\( (\Lambda_1, \Lambda_2, \Gamma) \) based on how well they jointly model the
-observed data. To formally introduce the state learner, we define the
-multi-state process
+% The problem with most existing super learners for right-censored data is that
+% they depend on a pre-specified estimator of the censoring distribution.
+The main idea of the state learner is to jointly use learners of
+\( \Lambda_1 \), \( \Lambda_2 \), and \( \Gamma \), and the relations
+in equation~(\ref{eq:parametrizeP}), to learn a feature of the
+observed data distribution \( P \). The discrete state learner ranks a
+tuple of learners of \( (\Lambda_1, \Lambda_2, \Gamma) \) based on how
+well they jointly model the observed data. %ANDERS: would be good to add one sentence about how to obtain a prediction model from such a ranking.
+To formally introduce the
+state learner, we define the multi-state process
 \begin{equation*}
@@
 \end{equation*}
-Note that at time \(t\), we observe that each observation is in one of four
+At time \(t\), we observe that each individual is in one of four
 mutually exclusive states (Figure \ref{fig:multi-state-process}).
@@
 \end{figure}
-The conditional distribution of \( \eta(t) \) given \( X=x \) is determined by
+% ANDERS: Figuren er lavet med prodlim, ik? du kan angive nummer for boksene (0 = initial, 1 = cause 1, 2= cause 2 og jeg ville foretraekke 3 i stedet for -1 for censored)
+The conditional distribution of \( \eta(t) \) given \( X \) is determined by
 the function
@@
 The function \( F \) describes the conditional state occupation
-probabilities corresponding to the observed multi-state process
-\(\eta\).
-
-We propose to construct a super learner for \( F \), i.e., the target of this
-super learner is $\Psi(P) = F$ where the parameter is identified through
-equation~(\ref{eq:F-def}). Because each quadruple
-$(\Lambda_{1}, \Lambda_{2}, \Gamma, H)$ characterises a \(P\in\mathcal P\) which
-in turn determines \( (F, H) \), a learner for \( F \) can be constructed from
-learners of \( \Lambda_1 \), \( \Lambda_2 \), and $\Gamma$ as follows:
+probabilities of the multi-state process \(\eta\). We construct a
+super learner for \( F \). The target of this super learner is the
+function-valued parameter $\Psi(P) = F$ which is identified through
+equation~(\ref{eq:F-def}).
+% ANDERS: the next sentence requires conditional independent censoring and restriction to [0,\tau], right?
+Because each quadruple
+$(\Lambda_{1}, \Lambda_{2}, \Gamma, H)$ characterizes a distribution
+\(P\in\mathcal P\) which in turn determines \( (F, H) \),
+% ANDERS: and we should refer to equations (2) and (6) 
+a learner for \( F \) can be constructed from learners of \( \Lambda_1 \),
+\( \Lambda_2 \), and $\Gamma$ as follows:
 \begin{equation}\label{eq:transition}
@@
 \end{equation}
-The state learner requires three libraries of learners, \(\mathcal{A}_1\),
-\( \mathcal{A}_2 \), and \( \mathcal{B} \), where \(\mathcal{A}_1\) and
-\( \mathcal{A}_2\) contain learners of the conditional cause-specific cumulative
-hazard functions of the event time distribution \(\Lambda_1\) and
-\( \Lambda_2\), respectively, and \(\mathcal{B}\) contains learners of the
-conditional cumulative hazard function of the censoring
-distribution. % We further
+The state learner requires three libraries of learners,
+\(\mathcal{A}_1\), \( \mathcal{A}_2 \), and \( \mathcal{B} \), where
+\(\mathcal{A}_1\) and \( \mathcal{A}_2\) contain learners of the
+conditional cause-specific cumulative hazard functions \(\Lambda_1\)
+and \( \Lambda_2\), respectively, and \(\mathcal{B}\) contains
+learners of the conditional cumulative hazard function of the
+censoring distribution. % We further
 % define \(\mathcal{H}\) as the set of all probability distributions on
@@
 libraries of learners for \((\Lambda_1,\Lambda_2,\Gamma)\) we construct a library
-$\mathcal{F}(\mathcal{A}_1, \mathcal{A}_2, \mathcal{B})$ of learners
+$\mathcal{F}$ of learners
 for \( F \):
@@
 \end{equation*}
-As described in Section~\ref{sec:super-learning}, each learner
+Based on a split of a data set \(\data_n\) into $K$ disjoint
+approximately equally sized subsets (see Section \ref{sec:super-learning}), each learner
 \( \phi_{a_1, a_2, b} \) in the library
-\( \mathcal{F}(\mathcal{A}_1, \mathcal{A}_2, \mathcal{B}) \) is evaluated using
-the cross-validated loss,
+\( \mathcal{F}(\mathcal{A}_1, \mathcal{A}_2, \mathcal{B}) \) is
+evaluated using the cross-validated loss,
 \begin{equation*}
@@
 \end{equation*}
-and the discrete state learner is
+and the discrete state learner is given by
 \begin{align*}\label{eq:discrete-state-learner}
@@
 % collection of all conditional state-occupation probability functions.
-Proposition~\ref{prop:stric-prop} can be derived from the fact that the
-integrated Brier score (also called the continuous ranked probability score) is
-a strictly proper scoring rule \citep{gneiting2007strictly}. This implies that
-if we minimise the average loss of the integrated Brier score, we recover the
-parameters of the data-generating distribution. In particular, the oracle of a
-state learner will be consistent if the library of learners contains at least
-one learner that is consistent for estimation of \( F \). Recall that the
-function \(F\) implicitly depends on the data generating probability measure
-\(P\in\mathcal P\) but that this was suppressed in the notation. We now make
-this dependence explicit by writing \(F_0\) for the function which is obtained
-by substituting a specific \(P_0\in\mathcal{P}\) for \(P\) in equation
+Proposition~\ref{prop:stric-prop} can be derived from the fact that
+the integrated Brier score (also called the continuous ranked
+probability score) is a strictly proper scoring rule
+\citep{gneiting2007strictly}. This implies that if we minimise the
+average loss of the integrated Brier score, we recover the parameters
+of the data-generating distribution. Specifically, this implies that
+the oracle of a state learner is consistent if the library of learners
+contains at least one learner that is consistent for estimation of
+\( F \). Recall that the function \(F\) implicitly depends on the data
+generating probability measure \(P\in\mathcal P\) but that this was
+suppressed in the notation. We now make this dependence explicit by
+writing \(F_0\) for the function which is obtained by substituting a
+specific \(P_0\in\mathcal{P}\) for \(P\) in equation
 \eqref{eq:transition}.
@@
     F^* = \argmin_{F} P_0{[\bar{B}_\tau(F, \blank)]},
-  % F^* = \argmin_{(\Lambda_1,\Lambda_2,\Gamma) \in \mathcal{M}\otimes\mathcal{M}\otimes\mathcal{M}} P_0{[\bar B_\tau(\phi_{(\Lambda_1,\Lambda_2,\Gamma)}, \blank)]},
-\end{equation*}
+  % F^* = \argmin_{(\Lambda_1,\Lambda_2,\Gamma) \in \mathcal{M}_{\tau}\otimes\mathcal{M}_{\tau}\otimes\mathcal{M}_{\tau}} P_0{[\bar B_\tau(\phi_{(\Lambda_1,\Lambda_2,\Gamma)}, \blank)]},
+  \end{equation*}
+  %ANDERS: would be good to define the set over which F is minimized with a symbol rather than only words
 where the minimum is taken over all \( F \), such that \( F \) is a conditional
@@
 
-We establish a finite sample oracle result for the state learner.
-Our Corollary~\ref{cor:oracle-prop} is in essence a special case of a
-general cross-validation result by \cite{vaart2006oracle}. We assume
+We now establish a finite sample oracle result for the state learner.
+Our Corollary~\ref{cor:oracle-prop} is in essence a special case of Theorem 2.3
+in \cite{vaart2006oracle}. % ANDERS: is Theo 2.3 correct?
+We assume
 that we split the data into equally sized folds, and for simplicity of
@@
 In the following we let \( \| \blank \|_{P_0} \) denote the norm
+% ANDERS: norm on which space? 
 \begin{equation}
@@
 
-In Section~\ref{sec:targeted-learning} we demonstrate how the state learner can
-be used for targeted learning. A targeted estimator is typically obtained from
+% ANDERS: would be good to separate 2-3 subsection in this section. here is a vague proposal. above could be under \subsection{Oracle property of the state learner}
+\subsection{Transience of double-robustness}
+% In Section~\ref{sec:targeted-learning} we demonstrate how the state learner can
+% be used for targeted learning.
+A targeted estimator is typically obtained from
 estimators of the nuisance parameters $\Lambda_1$, $\Lambda_2$, and $\Gamma$. By
@@
 and thus a targeted estimator can also be obtained from an estimator of $F$
-using equation~(\ref{eq:7}). The key feature of a targeted estimator is that it
+using equation~(\ref{eq:7}). The key feature of a targeted estimator
+% ANDERS: at this point we have not defined "targeted estimator" does fisher 2021 define this? Bickel does not ...
+is that it
 is asymptotically equivalent to a sum of i.i.d.\ random variables plus a second
-order remainder term \citep{bickel1993efficient,fisher2021visually}. For our
-setting of competing risks, the remainder term is typically dominated by terms
+order remainder term \citep{bickel1993efficient,fisher2021visually}. For the
+setting with competing risks, the remainder term is dominated by terms
 of the form
@@
 
-Features of the observed data distribution \( P \in \mathcal{P} \) are rarely of
-interest. We are instead interested in a parameter
-\( \theta \colon \mathcal{Q} \rightarrow \Theta \) that expresses a property of
-the uncensored population governed by the measure \( Q \in \mathcal{Q} \). The
-parameter space $\Theta$ can be a subset of \(\R^d\) or a subset of a function
-space, such as \(\mathcal{M}\). In subsection~\ref{sec:cause-spec-aver} we give
-a concrete example from causal inference where $\theta$ is the average treatment
-effect and \( \Theta = [-1,1] \). Under the assumption of conditional
-independent censoring and positivity, $\theta$ is identifiable from
-\( \mathcal{P} \) which means that there exists an operator
-\( \Psi \colon \mathcal{P} \rightarrow \Theta \) such that
-\( \theta(Q) = \Psi(P_{Q, \Gamma}) \) for all $\Gamma \in \mathcal{M}$. By
-equation~(\ref{eq:parametrizeP}) this imply that we may write
+We are interested in a parameter
+\( \theta \colon \mathcal{Q} \rightarrow \Theta \) that expresses a
+property of the uncensored population governed by the measure
+\( Q \in \mathcal{Q} \). The parameter space $\Theta$ can be a subset
+of \(\R^d\) or a subset of a function space, for example a subset of
+\(\mathcal{M}_{\tau}\). In subsection~\ref{sec:cause-spec-aver} we
+discuss an example from causal inference where $\theta$ is the average
+treatment effect and \( \Theta = [-1,1] \).
+% ANDERS: need a sentence a la: Before we discuss the role of the
+% state learner for targeted learning we review some results from
+% semiparametric efficiency theory. The reader may want to skip this ...
+Under the assumption of
+conditional independent censoring and positivity, $\theta$ is
+identifiable from \( \mathcal{P} \) which means that there exists an
+operator \( \Psi \colon \mathcal{P} \rightarrow \Theta \) such that
+\( \theta(Q) = \Psi(P_{Q, \Gamma}) \) for all
+$\Gamma \in \mathcal{M}_{\tau}$. By equation~(\ref{eq:parametrizeP})
+this implies that we may write
 \begin{equation*}
@@
 \end{equation*}
-for some operator \( \tilde{\Psi}^0 \). The state learner provides a ranking of
-tuples
-\( (a_1, a_2, b) \in \mathcal{A}_1 \times \mathcal{A}_2 \times \mathcal{B} \),
-and we use \( \hat{a}_{1,n} \), \( \hat{a}_{2,n} \), and \( \hat{b}_n \) to
-denote the learners corresponding to the discrete state learner
-\( \hat{\phi}_n \). Letting \( H(\data_n) \) denote the empirical measure of
-\( \{X_1, \dots, X_n\} \), we can obtain a simple plug-in an estimator of
-$\theta$ as
+for some operator \( \tilde{\Psi}^0 \). The state learner provides a
+ranking of tuples
+\( (a_1, a_2, b) \in \mathcal{A}_1 \times \mathcal{A}_2 \times
+\mathcal{B} \), and we use \( \hat{a}_{1,n} \), \( \hat{a}_{2,n} \),
+and \( \hat{b}_n \) to denote the learners corresponding to the
+discrete state learner \( \hat{\phi}_n \). Letting \( H(\data_n) \)
+denote the empirical measure of \( \{X_1, \dots, X_n\} \), we obtain a
+simple plug-in estimator of $\theta$:
 \begin{equation}
@@
 
-The asymptotic distribution of \( \hat{\Psi}_{\tau, j}^0 \) is difficult to
-analyze due to the model selection step involved when estimating the nuisance
-parameters $\Lambda_1$ and $\Lambda_2$. In addition, the estimator will
-typically have an asymptotic bias that vanishes at a sub-optimal rate. Using
-tools from semi-parametric efficiency theory, it is possible to construct a
-so-called targeted or debiased estimator with smaller asymptotic bias and an
-asymptotic distribution which we know how to estimate
-\citep{bickel1993efficient,van2011targeted,chernozhukov2018double}. A targeted
-estimator is based on the efficient influence function for the parameter
-$\tilde{\Psi}^0$ and relies on an estimator of $\Gamma$ in addition to
-estimators of \( \Lambda_1 \) and $\Lambda_2$. The efficient influence function
-is a \( P \)-zero mean and square integrable function indexed by the nuisance
-parameters \( (\Lambda_1, \Lambda_2, \Gamma) \), which we denote by
-\( \psi(\blank ; \Lambda_1, \Lambda_2, \Gamma) \). The name is justified because
-any regular asymptotically linear estimator that has \( \psi \) as its influence
-function is efficient, meaning that it has smallest asymptotic variance among
-all regular asymptotically linear estimators.
+The asymptotic distribution of \( \hat{\Psi}_{\tau, j}^0 \) is
+difficult to analyze due to the cross-validated model selection
+involved in the estimation of the nuisance parameters $\Lambda_1$ and
+$\Lambda_2$. In addition, the estimator will typically have an
+asymptotic bias that vanishes at a too slow 
+% ANDERS: is sub-optimal = too slow?
+rate. Using tools from semi-parametric efficiency theory, it is
+possible to construct a so-called targeted or debiased estimator with
+smaller asymptotic bias and a asymptotic distribution which we know
+how to estimate
+\citep{bickel1993efficient,van2011targeted,chernozhukov2018double}. A
+targeted estimator is based on the efficient influence function for
+the parameter $\tilde{\Psi}^0$ and relies on an estimator of $\Gamma$
+in addition to estimators of \( \Lambda_1 \) and $\Lambda_2$. The
+efficient influence function is a \( P \)-zero mean and square
+integrable function indexed by the nuisance parameters
+\( (\Lambda_1, \Lambda_2, \Gamma) \), which we denote by
+\( \psi(\blank ; \Lambda_1, \Lambda_2, \Gamma) \). The name is
+justified because any regular asymptotically linear estimator that has
+\( \psi \) as its influence function is asymptotically efficient,
+meaning that it has smallest asymptotic variance among all regular
+asymptotically linear estimators.
 
@@
 \end{equation}
-where \( \empmeas \) is the empirical measure of a sample \(\{O_i\}_{i=1}^n\).
-Under a suitable Donsker class regularity condition we can make the following
-asymptotic expansion of the one-step estimator
+where \( \empmeas \) is the empirical measure of a data set
+\(\{O_i\}_{i=1}^n\).  Under a Donsker class condition
+% ANDERS: Donsker class condition on what set?
+we have the
+following asymptotic expansion of the one-step estimator
 \citep{pfanzagl1985contributions,van2003unified,fisher2021visually,kennedy2022semiparametric,rytgaard2022continuous},
@@
 \end{equation*}
-where the remainder term is typically such that
+where the remainder term has the form
 \begin{equation}
@@
 for some suitable norm \( \|\blank \| \), for instance the
-\( \mathcal{L}_{P}^2 \)-norm. Hence when equation~(\ref{eq:4}) holds and the
-nuisance parameters $\Lambda_1$, $\Lambda_2$, and $\Gamma$ are consistently
-estimated at rate \( \smallO_P{(n^{-1/4})} \), then
+\( \mathcal{L}_{P}^2 \)-norm. When equation~(\ref{eq:4}) holds and the
+nuisance parameters $\Lambda_1$, $\Lambda_2$, and $\Gamma$ are
+consistently estimated at rate \( \smallO_P{(n^{-1/4})} \), then
 \begin{equation}
@@
 
-\subsection{Average treatment effect on the cumulative incidence}
+\subsection{Average treatment effect on the absolute risk of an event}
 \label{sec:cause-spec-aver}
 
-In the following we demonstrate how the state learner and the general estimation
-strategy outlined above can be used to construct an estimator of the
-cause-specific average treatment effect. We assume that the covariate vector
-\( X \in \R^p \) can be separated into a binary treatment indicator \( A \) and
-a vector of potential confounders, \( W \in \mathcal{W}\subset \R^{p-1} \). We
-abuse notation slightly by writing
-\( \Lambda_j(t \mid x) = \Lambda_j(t \mid a, w) \) and
-\( S(t \mid x) = S(t \mid a, w) \) when \( x=(a, w) \). We use $\mu$ to denote
-the marginal distribution of \( W \) and $\pi$ to denote the conditional
-probability of treatment,
+In the following we demonstrate how the state learner and the general
+estimation strategy outlined above can be used to construct an
+estimator of the cause-specific average treatment effect. We assume
+that the covariate vector \( X \in \R^p \) contains a binary treatment
+indicator \( A \) and a vector of potential confounders,
+\( W \in \mathcal{W}\subset \R^{p-1} \).
+% ANDERS: why is this abuse of notation?
+% We abuse notation slightly by writing
+% \( \Lambda_j(t \mid x) = \Lambda_j(t \mid a, w) \) and
+% \( S(t \mid x) = S(t \mid a, w) \) when \( x=(a, w) \).
+We use $\mu$
+to denote the marginal distribution of \( W \) and $\pi$ to denote the
+conditional probability of treatment,
 \begin{equation*}
@@
 \end{equation*}
-We assume throughout that $\pi$ is uniformly bounded away from \( 0 \) and
-\( 1 \) on \( \mathcal{W} \). As both \( A \) and \( W \) are fully observed for
-all individuals we can use an existing super learner to estimate $\pi$
-\citep{Polley_Ledell_Kennedy_Laan_2023_Superlearn}, and we denote this estimator
-by $\hat{\pi}_n$. We use the empirical measure of \( \{W_1, \dots, W_n\} \) to
-estimate $\mu$, and denote this estimator by $\hat{\mu}_n$. As parameter of
-interest we consider the standardised difference in the cumulative incidence due
-to cause 1 at time $\tau$, which can be expressed as
+We assume throughout that $\pi$ is uniformly bounded away from \( 0 \)
+and \( 1 \) on \( \mathcal{W} \) and that both \( A \) and \( W \) are
+fully observed for all individuals. We use an super learner to
+estimate $\pi$ \citep{Polley_Ledell_Kennedy_Laan_2023_Superlearn}, and
+we denote this estimator by $\hat{\pi}_n$. We use the empirical
+measure of \( \{W_1, \dots, W_n\} \) to estimate $\mu$, and denote
+this estimator by $\hat{\mu}_n$. As parameter of interest we consider
+the standardised difference in the absolute risk of an event with
+cause 1 at time $\tau$:
 \begin{equation*}
@@
 Under a set of
+% ANDERS: better to give these assumptions a name: positivity, ... and refer to Helene's paper?
 additional assumptions, \( \theta_{{\tau}} \) can be given the causal
@@
 on the set \( (\Lambda_1, \Lambda_2, \Gamma, \pi) \) of nuisance parameters.
-Define the quantities
+We define
 \begin{align*}
@@
 \end{align*}
-The efficient influence function can then be written as
+The efficient influence function can now be written as
 \citep{van2003unified,jewell2007non,rytgaard2022targeted},

### 2024-04-29 tex/jrss-version/state-learner-jrss-style.tex
response to TAG's comments

@@
 
-Medical decision making is traditionally based on the discrepancy
-between intuitive jugdment and the law of chance
-\citep{redelmeier1995probability}. We are concerned with the
-statistical models that provide probabilistic predictions about the
-medical future of a patient. We have two different applications of
-these models in mind. One is as a stand-alone tool where the model
-provides information for a freshly diagnosed patient. The other is as
-input to an algorithm which estimates a causal target parameter such
-as an average treatment effect. The models are learned on data from
-previous patients. A super learner is a machine learning algorithm
-that combines a finite set of learners into a meta learner by
-estimating prediction performance in hold-out samples using a
-pre-specified loss function \citep{van2007super}. When the aim is to
-make a prediction model, super learners combine strong learners, such
-as Cox regression models and random survival forests
-\citep{gerds2021medical}.
+Medical decision making is traditionally based on the discrepancy between
+intuitive judgment and the law of chance \citep{redelmeier1995probability}. We
+are concerned with the statistical models that provide probabilistic predictions
+about the medical future of a patient. We have two different applications of
+these models in mind.
+% THOMAS: I think we only really have one in mind in this article. For
+% stand-alone prediction, I think some further things should be
+% addressed/discussed, which I had thought could be for another article. But
+% maybe we can but it in here.
+One is as a stand-alone tool where the model provides
+information for a freshly diagnosed patient. The other is as input to an
+algorithm which estimates a causal target parameter such as an average treatment
+effect. The models are learned on data from previous patients. A super learner
+is a machine learning algorithm that combines a finite set of learners into a
+meta learner by estimating prediction performance in hold-out samples using a
+pre-specified loss function \citep{van2007super}. When the aim is to make a
+prediction model, super learners combine strong learners, such as Cox regression
+models and random survival forests \citep{gerds2021medical}.
 % For targeted learning of a low dimensional
@@
 event time
-distribution. % ANDERS: mangler afrundning efter sidste saetning. eller maaske kan saetningen flyttes laenger ned?
+distribution. 
+% ANDERS: mangler afrundning efter sidste saetning. eller maaske kan saetningen
+% flyttes laenger ned?
+% 
+% THOMAS: Tror du har skrevet den her bid. Har forsøgt at afrunde.
+It is not clear, however, whether such a procedure converges towards an element
+minimizing the average loss.
 
@@
 learner is based on the observable
-quantities. % ANDERS: hvad mener vi med observable quantities? distribution of the observed random variables?
+variables. % ANDERS: hvad mener vi med observable quantities? distribution of
+% the observed random variables?
+% 
+% THOMAS: Changed to 'variables'. It is important that the loss is based on the
+% observed variables directly, not just their distribution.
 The state learner can be applied to all types of survival models,
@@
 cross-validated loss
-\citep{van2011targeted}. % ANDERS: er der bedre referencer fra Mark paper/bog for discrete super learner?
+\citep{van2007super}.
+% ANDERS: er der bedre referencer fra Mark paper/bog for discrete super learner?
+% 
+% THOMAS: Yes, har taget Hubbard+laan super learner article i stedet.
 In the presence of competing risks, our algorithm uses separate
-libraries to model the hazard rates of the competing risks and the
+libraries to model the cumulative hazard functions of the competing risks and the
 censoring distribution. We show that the oracle selector of the state
@@
 event based on covariates in the presence of competing risks
-\citep{gerds2021medical}. The other application is in targeted
+\citep{gerds2021medical}.
+% THOMAS: This is not the output directly, but can easily be obtained.
+The other application is in targeted
 learning where conditional event probabilities occur as
@@
 For ease of notation we assume throughout that \( \Lambda_j(\blank \mid x) \) is
-continuous for all \( x \) and \( j \). %ANDERS: need a note here which says that this is not a limitation?
-We denote by \(S\) the conditional
-event-free survival function:
+continuous for all \( x \) and \( j \). 
+% ANDERS: need a note here which says that this is not a limitation?
+% 
+% THOMAS: I think it is already there in 'For ease of notation...', but I have
+% added the next sentence. We could also just work with the product integral,
+% would that be better?
+This is not a limitation as everything caries over directly to the general case
+and is solely imposed to simplify the formulas. We denote by \(S\) the
+conditional event-free survival function:
 \begin{equation}
@@
 % ANDERS: I prefer to have the \tau explicit
+% 
+% THOMAS: OK
 Let \( \mathcal{M}_{\tau}\) denote the space of all conditional cumulative hazard
@@
 \(\Gamma\in\mathcal M\) to denote the conditional cumulative hazard
-function for censoring. % ANDERS: can we not allow Gamma to be discrete?
+function for censoring.
+% ANDERS: can we not allow Gamma to be discrete?
+% 
+% THOMAS: Yes, no problem, it is only for notation convenience. 
 For notational convenience, we assume that \( \Gamma(\blank \mid x) \) is
@@
 % ANDERS: H is missing in the model?
+% 
+% THOMAS: H is part of Q (see equation between equation 1 and 2).
 Hence, we may write
@@
 log-likelihood loss.
-% ANDERS: would be good to have a formula for a the negative partial log-likelihood loss showing that the result is infitity
+% ANDERS: would be good to have a formula for a the negative partial
+% log-likelihood loss showing that the result is infitity
+%
+% THOMAS: I would argue that this should be clear from the previous sentence and
+% the definition of the log-likelihood function: If a discrete model assigns a
+% probability of zero to an event, then the log-likelihood of that event is
+% log(0) = -infinity.
+%
+% Such a formula was part of one of the earlier versions where I had put more
+% focus on the problem than on the solution. If we add this, we should work with
+% the product integral notation.
 
@@
 drawback is that the library is now restricted to include only these
-models. % ANDERS: can the library really include multiple proportional hazards models? only one of them can be correct ...
-% also, would be good to merge this paragraph with the previous somehow
+models.
+% ANDERS: can the library really include multiple proportional hazards models?
+% only one of them can be correct ... also, would be good to merge this
+% paragraph with the previous somehow
+%
+% THOMAS: Yes, it can include multiple PH models (if I understand you
+% correctly). A major point is that there is no proof that this works, and when
+% you start to think about it, it is not completely obvious whether it should
+% work or not. I think it works and am trying to prove it.
+%
+% I disagree about merging this section with the previous one, I see them as
+% describing two completely different approaches and issues.
 
@@
 The main idea of the state learner is to jointly use learners of
-\( \Lambda_1 \), \( \Lambda_2 \), and \( \Gamma \), and the relations
-in equation~(\ref{eq:parametrizeP}), to learn a feature of the
-observed data distribution \( P \). The discrete state learner ranks a
-tuple of learners of \( (\Lambda_1, \Lambda_2, \Gamma) \) based on how
-well they jointly model the observed data. %ANDERS: would be good to add one sentence about how to obtain a prediction model from such a ranking.
-To formally introduce the
-state learner, we define the multi-state process
+\( \Lambda_1 \), \( \Lambda_2 \), and \( \Gamma \), and the relations in
+equation~(\ref{eq:parametrizeP}), to learn a feature of the observed data
+distribution \( P \). The discrete state learner ranks a tuple of learners of
+\( (\Lambda_1, \Lambda_2, \Gamma) \) based on how well they jointly model the
+observed
+data.
+% ANDERS: would be good to add one sentence about how to obtain a prediction
+% model from such a ranking.
+%
+% THOMAS: I think this depends on whether or not we want prediction to be one of
+% the tasks addressed in this article. I have focused on something else. We
+% could add something like:
+Risk predictions for causes one and two can then be obtained from
+\( \Lambda_1 \) and $\Lambda_2$ \citep{benichou1990estimates}. To formally
+introduce the state learner, we define the multi-state process
 \begin{equation*}
@@
 \end{figure}
-% ANDERS: Figuren er lavet med prodlim, ik? du kan angive nummer for boksene (0 = initial, 1 = cause 1, 2= cause 2 og jeg ville foretraekke 3 i stedet for -1 for censored)
+% ANDERS: Figuren er lavet med prodlim, ik? du kan angive nummer for boksene (0
+% = initial, 1 = cause 1, 2= cause 2 og jeg ville foretraekke 3 i stedet for -1
+% for censored)
+%
+% THOMAS: Jeg foretrækker -1 frem for 3, så vil gerne høre dine argumenter for
+% 3.
 The conditional distribution of \( \eta(t) \) given \( X \) is determined by
@@
 equation~(\ref{eq:F-def}).
-% ANDERS: the next sentence requires conditional independent censoring and restriction to [0,\tau], right?
-Because each quadruple
+% ANDERS: the next sentence requires conditional independent censoring and
+% restriction to [0,\tau], right?
+%
+% THOMAS: Yes, added now.
+Under conditional independent censoring each quadruple
 $(\Lambda_{1}, \Lambda_{2}, \Gamma, H)$ characterizes a distribution
-\(P\in\mathcal P\) which in turn determines \( (F, H) \),
+\(P\in\mathcal P\), c.f.\ equation~\eqref{eq:parametrizeP}, which in turn
+determines \( (F, H) \),
 % ANDERS: and we should refer to equations (2) and (6)
-a learner for \( F \) can be constructed from learners of \( \Lambda_1 \),
-\( \Lambda_2 \), and $\Gamma$ as follows:
+%
+% THOMAS: Yes, added above -- OK?
+and so a learner for \( F \) can be constructed from learners of
+\( \Lambda_1 \), \( \Lambda_2 \), and $\Gamma$ as follows:
 \begin{equation}\label{eq:transition}
@@
 
-In this section we establish theoretical guarantees for the state learner.
+In this section we establish theoretical guarantees for the state learner. We
+show that the state learner is consistent if its library contains a consistent
+learner, and we establish a finite sample inequality for the excess risk of the
+state learner compared to the oracle. Finally we show that a certain second
+order structure is preserved when the state learner is used for targeted
+learning.
+
+\subsection{Consistency}
+\label{sec:consistency}
+
 % We let \( \E_P \) denote
@@
 % collection of all conditional state-occupation probability functions.
-Proposition~\ref{prop:stric-prop} can be derived from the fact that
-the integrated Brier score (also called the continuous ranked
-probability score) is a strictly proper scoring rule
-\citep{gneiting2007strictly}. This implies that if we minimise the
-average loss of the integrated Brier score, we recover the parameters
-of the data-generating distribution. Specifically, this implies that
-the oracle of a state learner is consistent if the library of learners
-contains at least one learner that is consistent for estimation of
-\( F \). Recall that the function \(F\) implicitly depends on the data
-generating probability measure \(P\in\mathcal P\) but that this was
-suppressed in the notation. We now make this dependence explicit by
-writing \(F_0\) for the function which is obtained by substituting a
-specific \(P_0\in\mathcal{P}\) for \(P\) in equation
-\eqref{eq:transition}.
+Proposition~\ref{prop:stric-prop} can be derived from the fact that the
+integrated Brier score (also called the continuous ranked probability score) is
+a strictly proper scoring rule \citep{gneiting2007strictly}. This implies that
+if we minimise the average loss of the integrated Brier score, we recover the
+parameters of the data-generating distribution. Specifically, this implies that
+the oracle of a state learner is consistent if the library of learners contains
+at least one learner that is consistent for estimation of \( F \). Recall that
+the function \(F\) implicitly depends on the data generating probability measure
+\(P\in\mathcal P\) but that this was suppressed in the notation. We now make
+this dependence explicit by writing \(F_0\) for the function which is obtained
+by substituting a specific \(P_0\in\mathcal{P}\) for \(P\) in equation
+\eqref{eq:transition}. In the following we let
+\( \mathcal{H} = \{F_P : P \in \mathcal{P}\} \) where \( F_P \) is defined as in
+equation~(\ref{eq:F-def}) using the measure \( P \in \mathcal{P} \).
 
@@
   \begin{equation*}
-    F^* = \argmin_{F} P_0{[\bar{B}_\tau(F, \blank)]},
+    F^* = \argmin_{F \in \mathcal{H}} P_0{[\bar{B}_\tau(F, \blank)]}.
   % F^* = \argmin_{(\Lambda_1,\Lambda_2,\Gamma) \in \mathcal{M}_{\tau}\otimes\mathcal{M}_{\tau}\otimes\mathcal{M}_{\tau}} P_0{[\bar B_\tau(\phi_{(\Lambda_1,\Lambda_2,\Gamma)}, \blank)]},
   \end{equation*}
-  %ANDERS: would be good to define the set over which F is minimized with a symbol rather than only words
-where the minimum is taken over all \( F \), such that \( F \) is a conditional
-state occupation probability function for some measure \( P \) as defined in equation~(\ref{eq:F-def}).
+  % ANDERS: would be good to define the set over which F is minimized with a
+  % symbol rather than only words
+  %
+  % THOMAS: How about now? \mathcal{H} is now defined just before this proposition.
 Then \( F^*(t,j,\blank) = F_0(t,j,\blank) \) \( H \)-almost surely for
@@
 
-We now establish a finite sample oracle result for the state learner.
-Our Corollary~\ref{cor:oracle-prop} is in essence a special case of Theorem 2.3
-in \cite{vaart2006oracle}. % ANDERS: is Theo 2.3 correct?
-We assume
-that we split the data into equally sized folds, and for simplicity of
-presentation we take \( n \) to be such that
-\( |\data_n^{-k}| = n/K \) with \( K \) fixed. We will allow the
-number of learners to grow with \( n \) and write
+\subsection{Oracle inequalities}
+\label{sec:finite-sample-oracle}
+
+We establish a finite sample oracle result for the state learner. Our
+Corollary~\ref{cor:oracle-prop} is in essence a special case of Theorem 2.3 in
+\cite{vaart2006oracle}.
+% ANDERS: is Theo 2.3 correct?
+%
+% THOMAS: Yes. What do you mean?
+We assume that we split the data into equally sized folds, and for simplicity of
+presentation we take \( n \) to be such that \( |\data_n^{-k}| = n/K \) with
+\( K \) fixed. We will allow the number of learners to grow with \( n \) and
+write
 \( \mathcal{F}_n=\mathcal{F}(\mathcal{A}_{1,n}, \mathcal{A}_{2,n},
-\mathcal{B}_n)\) as short-hand notation and to emphasise the
-dependence on \( n \).
+\mathcal{B}_n)\) as short-hand notation and to emphasise the dependence on
+\( n \).
 % The discrete super $\hat{\phi}_n$ and the
@@
 % \( B \).
-In the following we let \( \| \blank \|_{P_0} \) denote the norm
+In the following we let the space \( \mathcal{H} \) be equipped with the norm
+\( \| \blank \|_{P_0} \) defined as
 % ANDERS: norm on which space?
+%
+% THOMAS: Fixed.
 \begin{equation}
@@
 \end{equation}
+
 \begin{corollary}
@@
 
-% ANDERS: would be good to separate 2-3 subsection in this section. here is a vague proposal. above could be under \subsection{Oracle property of the state learner}
-\subsection{Transience of double-robustness}
+% ANDERS: would be good to separate 2-3 subsection in this section. here is a
+% vague proposal. above could be under \subsection{Oracle property of the state
+% learner}
+%
+% THOMAS: OK, I have added a suggestion for two sections more.
+
+% \subsection{Transience of double-robustness}
+\subsection{Transience of second order remainder structure}
+% THOMAS: It is not clear that double-robustness is preserved, but a second
+% order product structure is, so I have changed the title. 
+% 
 % In Section~\ref{sec:targeted-learning} we demonstrate how the state learner can
@@
 using equation~(\ref{eq:7}). The key feature of a targeted estimator
-% ANDERS: at this point we have not defined "targeted estimator" does fisher 2021 define this? Bickel does not ...
-is that it
-is asymptotically equivalent to a sum of i.i.d.\ random variables plus a second
-order remainder term \citep{bickel1993efficient,fisher2021visually}. For the
-setting with competing risks, the remainder term is dominated by terms
-of the form
+% ANDERS: at this point we have not defined "targeted estimator" does fisher
+% 2021 define this? Bickel does not ...
+%
+% THOMAS: I would think that Fisher+Kennedy does define it. But my solution to
+% this problem was to reference Section 7, which you have removed.
+% Alternatively, this could be part of Section 7, but I am quite sure you once
+% moved it from there to here. What do you suggest?
+is that it is asymptotically equivalent to a sum of i.i.d.\ random variables
+plus a second order remainder term
+\citep{bickel1993efficient,fisher2021visually}. For the setting with competing
+risks, the remainder term is dominated by terms of the form
 \begin{equation}
@@
 % semiparametric efficiency theory. The reader may want to skip this ...
-Under the assumption of
-conditional independent censoring and positivity, $\theta$ is
-identifiable from \( \mathcal{P} \) which means that there exists an
+%
+% THOMAS: Hmm, I think this is already stated below ('Using tools from
+% semi-parametric efficiency theory...'). But we could add something like the
+% following? Depending on the framing of the paper (with focus exclusively on
+% state learner for targeted learning or also on prediction?), I don't think the
+% reader can skip this section.
+To discuss the role of the state learner for targeted learning we briefly review
+some results from semiparametric efficiency theory. Extensive reviews and
+introductions are available elsewhere
+\cite[e.g.,][]{pfanzagl1985contributions,bickel1993efficient,van2003unified,tsiatis2007semiparametric,kennedy2016semiparametric}.
+Under the assumption of conditional independent censoring and positivity,
+$\theta$ is identifiable from \( \mathcal{P} \) which means that there exists an
 operator \( \Psi \colon \mathcal{P} \rightarrow \Theta \) such that
-\( \theta(Q) = \Psi(P_{Q, \Gamma}) \) for all
-$\Gamma \in \mathcal{M}_{\tau}$. By equation~(\ref{eq:parametrizeP})
-this implies that we may write
+\( \theta(Q) = \Psi(P_{Q, \Gamma}) \) for all $\Gamma \in \mathcal{M}_{\tau}$.
+By equation~(\ref{eq:parametrizeP}) this implies that we may write
 \begin{equation*}
@@
 % ANDERS: is sub-optimal = too slow?
+%
+% THOMAS: Hmm, we could use 'too slow', but it is then unclear why it is too
+% slow -- too slow for what? Writing 'sub-optimal' suggests (too me at least),
+% that there exists a rate that is better -- i.e., this rate it not the optimal
+% one for this problem, but an optimal one exists.
 rate. Using tools from semi-parametric efficiency theory, it is
@@
 % ANDERS: Donsker class condition on what set?
+%
+% THOMAS: Things are quite informal here so I don't see the point in making this
+% precise.
 we have the
@@
 % ANDERS: why is this abuse of notation?
+%
+% THOMAS: Because we should write \Lambda_j(t \mid (a, w)), because \Lambda as
+% originally defined only takes two arguments and not three.
+%
 % We abuse notation slightly by writing
@@
 Under a set of
-% ANDERS: better to give these assumptions a name: positivity, ... and refer to Helene's paper?
+% ANDERS: better to give these assumptions a name: positivity, ... and refer to
+% Helene's paper?
+%
+% THOMAS: Alternatively, we could just call it a 'standardized risk difference'
+% and then not discuss counterfactuals, but say that it can be given a causal
+% interpretation and reference some papers? I would like to not discuss
+% assumptions from causality as it is not very important for the state learner.
 additional assumptions, \( \theta_{{\tau}} \) can be given the causal

### 2024-04-29 figures/figure-multi-state-process.R

@@ -3,9 +3,9 @@
 ## Author: Thomas Alexander Gerds
 ## Created: Jul 27 2023 (10:04) 
 ## Version: 
-## Last-Updated: Oct  9 2023 (18:22) 
+## Last-Updated: Apr 29 2024 (10:30) 
 ##           By: Anders Munch
-##     Update #: 2
+##     Update #: 8
 #----------------------------------------------------------------------
 ## 
 ### Commentary: 
@@ -15,11 +15,15 @@
 ## 
 ### Code:
 library(prodlim)
+library(here)
 nTrans <- 3
 stateLabels = c("Initial","Cause 1", "Cause 2", "Censored")
 crHist <- Hist(time = 1:nTrans, event = list(from = rep("1", nTrans), to = stateLabels[-1]))
-pdf("~/research/SuperVision/Anders/survival-loss/statelearner/figure-multi-state-process.pdf")
-plot(crHist,stateLabels = c("Initial","Cause 1", "Cause 2", "Censored"),arrowLabels = FALSE)
+try(setwd("~/research/SuperVision/Anders/survival-loss/statelearner/figures"))
+try(setwd(here("figures")))
+pdf("figure-multi-state-process.pdf")
+plot(crHist,stateLabels = stateLabels,arrowLabels = FALSE,
+     tagBoxes = FALSE) ## Make true and/or change to -1?
 dev.off()
 
 ######################################################################

### 2024-05-02 tex/jrss-version/state-learner-jrss-style.tex
smoothing + more comments

--- 2024-04-29 tex/jrss-version/state-learner-jrss-style.tex
@@
 Medical decision making is traditionally based on the discrepancy
-between intuitive jugdment and the law of chance
+between intuitive judgment and the law of chance
 \citep{redelmeier1995probability}. We are concerned with the
@@
 censoring weighted loss has been studied for decades
-\citep{graf1999assessment, hothorn2006survival, gerds2006consistent}. A disadvantage of inverse probability of
+\citep{graf1999assessment, hothorn2006survival,
+  gerds2006consistent}. A disadvantage of inverse probability of
 censoring weighted loss estimators is that they require a
-pre-specified model for the censoring
-distribution. In \cite{westling2021inference} this problem is tackled by
-iterating between super learning of the censoring distribution and the
-event time
-distribution. % ANDERS: mangler afrundning efter sidste saetning. eller maaske kan saetningen flyttes laenger ned?
+pre-specified model for the censoring distribution. Recently
+\cite{westling2021inference} proposed a procedure which iterates
+between super learning of the censoring distribution and super
+learning of the event time distribution. In this article we pursue a
+different route.
 
@@
 distribution. The loss function which is used to define the state
-learner is based on the observable
-quantities. % ANDERS: hvad mener vi med observable quantities? distribution of the observed random variables?
+learner is based on the observed variables, that is the minimum and
+the order of the censoring time and the event
+time.
+
 The state learner can be applied to all types of survival models,
@@
 library of learners by picking the one that minimises the
-cross-validated loss
-\citep{van2011targeted}. % ANDERS: er der bedre referencer fra Mark paper/bog for discrete super learner?
-In the presence of competing risks, our algorithm uses separate
-libraries to model the hazard rates of the competing risks and the
-censoring distribution. We show that the oracle selector of the state
-learner is consistent if all libraries contain a consistent learner
-and prove a finite sample oracle inequality. The result of the state
-learner algorithm is a model which predicts the probability of an
-event based on covariates in the presence of competing risks
+cross-validated loss \citep{van2007super}. In the presence of
+competing risks, our algorithm uses separate libraries to model the
+cumulative hazard functions of the competing risks and the censoring
+distribution. We show that the oracle selector of the state learner is
+consistent if all libraries contain a consistent learner and prove a
+finite sample oracle inequality. A result of the state learner
+algorithm is a model which predicts the probability of an event based
+on covariates in the presence of competing risks
 \citep{gerds2021medical}. The other application is in targeted
@@
 \end{equation*}
-For ease of notation we assume throughout that \( \Lambda_j(\blank \mid x) \) is
-continuous for all \( x \) and \( j \). %ANDERS: need a note here which says that this is not a limitation?
-We denote by \(S\) the conditional
-event-free survival function:
+For ease of presentation we assume throughout that
+\( \Lambda_j(\blank \mid x) \) is continuous for all \( x \) and
+\( j \). This is not a limitation as everything caries over directly
+to the general case. We denote by \(S\) the conditional event-free
+survival function:
 \begin{equation}
@@
 \end{equation}
-% ANDERS: I prefer to have the \tau explicit
 Let \( \mathcal{M}_{\tau}\) denote the space of all conditional cumulative hazard
@@
 \(\Gamma\in\mathcal M\) to denote the conditional cumulative hazard
-function for censoring. % ANDERS: can we not allow Gamma to be discrete?
-For notational convenience, we assume that \( \Gamma(\blank \mid x) \) is
-continuous for all \( x \), and let
+function for censoring. For ease of presentation we now also assume
+that \( \Gamma(\blank \mid x) \) is continuous for all \( x \). We let
 \(G(t \mid x)=\exp\left\{-\Gamma(t \mid x)\right\}\) denote the
-survival function of the conditional censoring distribution. The distribution 
-is characterized by
+survival function of the conditional censoring distribution. The
+distribution is characterized by
 \begin{equation}\label{eq:parametrizeP}
@@
 \end{equation}
-% ANDERS: H is missing in the model?
 Hence, we may write
@@
 \mathcal{G} \} \) for some \( \mathcal{G} \subset \mathcal{M}_{\tau} \). We
-also have
+also have \(H\)-almost everywhere
 \begin{equation*}
@@
 \citep[e.g.,][]{li2016regularized,yao2017deep,lee2018deephit,katzman2018deepsurv,gensheimer2019scalable,lee2021boosted,kvamme2021continuous}.
-It seems however that this loss function is not perfectly suited for
-super learning, because it cannot be applied to important survival
-learners: The Kaplan-Meier estimator, the random survival forest, and
-the semi-parametric Cox regression model provide cumulative hazard
-functions that are piece-wise constant in the time argument, and hence
-assign zero probability to event times not observed in the training
-data. This implies that when event times are observed in continuous
-time, any of these learners will almost surely have infinite loss in
-an independent test set according to the negative partial
-log-likelihood loss.
-% ANDERS: would be good to have a formula for a the negative partial log-likelihood loss showing that the result is infitity
+This loss function is however not perfectly suited for super
+learning. The reason is that it cannot be applied to important
+survival learners: The Kaplan-Meier estimator, the random survival
+forest, and the semi-parametric Cox regression model provide
+cumulative hazard functions that are piece-wise constant in the time
+argument. Suppose these methods are applied to a learning data set and
+the aim is to evaluate the partial log-likelihood loss in an
+independent test data set. Suppose further that the test data set
+contains event times that were not also seen in the learning data set,
+as it would naturally happen when the events are observed in
+continuous time. The problem is that these methods assign zero
+probability to the test set event times that were not also in the
+training data set, and hence (the logarithm of zero is infitity) the
+partial likelihood loss in the test set would be infitity.
 
 When a proportional hazards model is assumed, the baseline hazard
-function can be profiled out of the likelihood to give a new partial
-log-likelihood loss \citep{cox1972regression}, which has been
-suggested as a loss function for super learning
-\citep{golmakani2020super,verweij1993cross}. While this allows the
-library of learners to include Cox' proportional hazard models, the
-drawback is that the library is now restricted to include only these
-models. % ANDERS: can the library really include multiple proportional hazards models? only one of them can be correct ...
-% also, would be good to merge this paragraph with the previous somehow
+function can be profiled out of the likelihood \citep{cox1972regression}.
+The cross-validated partial log-likelihood loss  \citep{verweij1993cross} has therefore been
+suggested as a loss function for super learning which however
+restricts the library of learners to include only Cox proportional
+hazards models \citep{golmakani2020super}.
 
@@
 The main idea of the state learner is to jointly use learners of
-\( \Lambda_1 \), \( \Lambda_2 \), and \( \Gamma \), and the relations
-in equation~(\ref{eq:parametrizeP}), to learn a feature of the
-observed data distribution \( P \). The discrete state learner ranks a
-tuple of learners of \( (\Lambda_1, \Lambda_2, \Gamma) \) based on how
-well they jointly model the observed data. %ANDERS: would be good to add one sentence about how to obtain a prediction model from such a ranking.
-To formally introduce the
-state learner, we define the multi-state process
+\( \Lambda_1 \), \( \Lambda_2 \), and \( \Gamma \), and the relations in
+equation~(\ref{eq:parametrizeP}), to learn a feature of the observed data
+distribution \( P \). The discrete state learner ranks a tuple of learners of
+\( (\Lambda_1, \Lambda_2, \Gamma) \) based on how well they jointly model the
+observed
+data.
+Risk predictions can then be obtained from
+\( \Lambda_1 \) and $\Lambda_2$ \citep{benichou1990estimates, ozenne2017riskregression}. To formally
+introduce the state learner, we define the multi-state process
 \begin{equation*}
@@
 \end{figure}
-% ANDERS: Figuren er lavet med prodlim, ik? du kan angive nummer for boksene (0 = initial, 1 = cause 1, 2= cause 2 og jeg ville foretraekke 3 i stedet for -1 for censored)
 The conditional distribution of \( \eta(t) \) given \( X \) is determined by
@@
 equation~(\ref{eq:F-def}).
-% ANDERS: the next sentence requires conditional independent censoring and restriction to [0,\tau], right?
-Because each quadruple
+Under conditional independent censoring each quadruple
 $(\Lambda_{1}, \Lambda_{2}, \Gamma, H)$ characterizes a distribution
-\(P\in\mathcal P\) which in turn determines \( (F, H) \),
-% ANDERS: and we should refer to equations (2) and (6) 
-a learner for \( F \) can be constructed from learners of \( \Lambda_1 \),
-\( \Lambda_2 \), and $\Gamma$ as follows:
+\(P\in\mathcal P\), c.f.\ equation~\eqref{eq:parametrizeP}, which in turn
+determines \( (F, H) \),
+and so a learner for \( F \) can be constructed from learners of
+\( \Lambda_1 \), \( \Lambda_2 \), and $\Gamma$ as follows:
 \begin{equation}\label{eq:transition}
@@
 % collection of all conditional state-occupation probability functions.
-Proposition~\ref{prop:stric-prop} can be derived from the fact that
-the integrated Brier score (also called the continuous ranked
-probability score) is a strictly proper scoring rule
-\citep{gneiting2007strictly}. This implies that if we minimise the
-average loss of the integrated Brier score, we recover the parameters
-of the data-generating distribution. Specifically, this implies that
-the oracle of a state learner is consistent if the library of learners
-contains at least one learner that is consistent for estimation of
-\( F \). Recall that the function \(F\) implicitly depends on the data
-generating probability measure \(P\in\mathcal P\) but that this was
-suppressed in the notation. We now make this dependence explicit by
-writing \(F_0\) for the function which is obtained by substituting a
-specific \(P_0\in\mathcal{P}\) for \(P\) in equation
-\eqref{eq:transition}.
+Proposition~\ref{prop:stric-prop} can be derived from the fact that the
+integrated Brier score (also called the continuous ranked probability score) is
+a strictly proper scoring rule \citep{gneiting2007strictly}. This implies that
+if we minimise the average loss of the integrated Brier score, we recover the
+parameters of the data-generating distribution. Specifically, this implies that
+the oracle of a state learner is consistent if the library of learners contains
+at least one learner that is consistent for estimation of \( F \). Recall that
+the function \(F\) implicitly depends on the data generating probability measure
+\(P\in\mathcal P\) but that this was suppressed in the notation. We now make
+this dependence explicit by writing \(F_0\) for the function which is obtained
+by substituting a specific \(P_0\in\mathcal{P}\) for \(P\) in equation
+\eqref{eq:transition}. In the following we let
+\( \mathcal{H}_{\mathcal{P}} = \{F_P : P \in \mathcal{P}\} \) where \( F_P \) is defined as in
+equation~(\ref{eq:F-def}) using the measure \( P \in \mathcal{P} \).
 
@@
   \begin{equation*}
-    F^* = \argmin_{F} P_0{[\bar{B}_\tau(F, \blank)]},
+    F^* = \argmin_{F \in \mathcal{H}_{\mathcal{P}}} P_0{[\bar{B}_\tau(F, \blank)]}.
   % F^* = \argmin_{(\Lambda_1,\Lambda_2,\Gamma) \in \mathcal{M}_{\tau}\otimes\mathcal{M}_{\tau}\otimes\mathcal{M}_{\tau}} P_0{[\bar B_\tau(\phi_{(\Lambda_1,\Lambda_2,\Gamma)}, \blank)]},
   \end{equation*}
-  %ANDERS: would be good to define the set over which F is minimized with a symbol rather than only words
-where the minimum is taken over all \( F \), such that \( F \) is a conditional
-state occupation probability function for some measure \( P \) as defined in equation~(\ref{eq:F-def}).
 Then \( F^*(t,j,\blank) = F_0(t,j,\blank) \) \( H \)-almost surely for
@@
 
-We now establish a finite sample oracle result for the state learner.
-Our Corollary~\ref{cor:oracle-prop} is in essence a special case of Theorem 2.3
-in \cite{vaart2006oracle}. % ANDERS: is Theo 2.3 correct?
-We assume
-that we split the data into equally sized folds, and for simplicity of
-presentation we take \( n \) to be such that
-\( |\data_n^{-k}| = n/K \) with \( K \) fixed. We will allow the
-number of learners to grow with \( n \) and write
+\subsection{Oracle inequalities}
+\label{sec:finite-sample-oracle}
+
+We establish a finite sample oracle result for the state learner. Our
+Corollary~\ref{cor:oracle-prop} is in essence a special case of
+Theorem 2.3 in \cite{vaart2006oracle}.  We assume that we split the
+data into equally sized folds, and for simplicity of presentation we
+take \( n \) to be such that \( |\data_n^{-k}| = n/K \) with \( K \)
+fixed. We will allow the number of learners to grow with \( n \) and
+write
 \( \mathcal{F}_n=\mathcal{F}(\mathcal{A}_{1,n}, \mathcal{A}_{2,n},
@@
 % \( B \).
-In the following we let \( \| \blank \|_{P_0} \) denote the norm
-% ANDERS: norm on which space? 
+In the following we let the space \( \mathcal{H} \) be equipped with the norm
+\( \| \blank \|_{P_0} \) defined as
 \begin{equation}
@@
 
-% ANDERS: would be good to separate 2-3 subsection in this section. here is a vague proposal. above could be under \subsection{Oracle property of the state learner}
-\subsection{Transience of double-robustness}
-% In Section~\ref{sec:targeted-learning} we demonstrate how the state learner can
-% be used for targeted learning.
-A targeted estimator is typically obtained from
+\subsection{Transience of the second order remainder structure}
+
+% ANDERS / THOMAS
+We consider 
+anestimator which is obtained from
 estimators of the nuisance parameters $\Lambda_1$, $\Lambda_2$, and $\Gamma$. By
@@
 using equation~(\ref{eq:7}). The key feature of a targeted estimator
-% ANDERS: at this point we have not defined "targeted estimator" does fisher 2021 define this? Bickel does not ...
-is that it
-is asymptotically equivalent to a sum of i.i.d.\ random variables plus a second
-order remainder term \citep{bickel1993efficient,fisher2021visually}. For the
-setting with competing risks, the remainder term is dominated by terms
-of the form
+is that it is asymptotically equivalent to a sum of i.i.d.\ random variables
+plus a second order remainder term
+
+\citep{bickel1993efficient,fisher2021visually}. For the setting with competing
+risks, the remainder term is dominated by terms of the form
 \begin{equation}
@@
 treatment effect and \( \Theta = [-1,1] \).
-% ANDERS: need a sentence a la: Before we discuss the role of the
-% state learner for targeted learning we review some results from
-% semiparametric efficiency theory. The reader may want to skip this ...
-Under the assumption of
-conditional independent censoring and positivity, $\theta$ is
-identifiable from \( \mathcal{P} \) which means that there exists an
+To discuss the role of the state learner for targeted learning we briefly review
+some results from semiparametric efficiency theory. Extensive reviews and
+introductions are available elsewhere
+\cite[e.g.,][]{pfanzagl1985contributions,bickel1993efficient,van2003unified,tsiatis2007semiparametric,kennedy2016semiparametric}.
+Under the assumption of conditional independent censoring and positivity,
+$\theta$ is identifiable from \( \mathcal{P} \) which means that there exists an
 operator \( \Psi \colon \mathcal{P} \rightarrow \Theta \) such that
-\( \theta(Q) = \Psi(P_{Q, \Gamma}) \) for all
-$\Gamma \in \mathcal{M}_{\tau}$. By equation~(\ref{eq:parametrizeP})
-this implies that we may write
+\( \theta(Q) = \Psi(P_{Q, \Gamma}) \) for all $\Gamma \in \mathcal{M}_{\tau}$.
+By equation~(\ref{eq:parametrizeP}) this implies that we may write
 \begin{equation*}
@@
 asymptotic bias that vanishes at a too slow 
-% ANDERS: is sub-optimal = too slow?
 rate. Using tools from semi-parametric efficiency theory, it is
@@
 where \( \empmeas \) is the empirical measure of a data set
-\(\{O_i\}_{i=1}^n\).  Under a Donsker class condition
-% ANDERS: Donsker class condition on what set?
+\(\{O_i\}_{i=1}^n\).
+Under suitable regularity conditions
 we have the
@@
 \( W \in \mathcal{W}\subset \R^{p-1} \).
-% ANDERS: why is this abuse of notation?
-% We abuse notation slightly by writing
-% \( \Lambda_j(t \mid x) = \Lambda_j(t \mid a, w) \) and
-% \( S(t \mid x) = S(t \mid a, w) \) when \( x=(a, w) \).
 We use $\mu$
@@
 \end{equation*}
-Under a set of
-% ANDERS: better to give these assumptions a name: positivity, ... and refer to Helene's paper?
-additional assumptions, \( \theta_{{\tau}} \) can be given the causal
-interpretation
+Under the usual assumptions for causal inference (consistency,
+positivity, no unmeasured confounding) \( \theta_{{\tau}} \) can be
+given the causal interpretation
 \begin{equation*}
@@
 
-As a proof of concept we conduct two small simulation studies in a setting
-without competing risks. We generate data according to a distribution motivated
-from a real dataset in which censoring depends on the baseline covariates. We
+In this section we report results from two settings of our simulation
+studies, both settings are without competing risks. In the next
+section, we consider a setting with competing risks.
+% ANDERS: kan du skrive her hvad forskellen er mellem og formaalet er
+% med de to simulationsstudierscenarier?
+% i den ene sammenligner vi med IPCW i den anden med westling, ik?
+% Maaske en figur (flowchart) eller tabel kunne hjaelpe laeseren at faa overblik.
+% Maaske separation i to subsections kunne hjaelpe yderliger
+
+We generate data according to a distribution motivated from a real
+dataset in which censoring depends on the baseline covariates. We
 simulate data based on a prostate cancer study described in
-\citep{kattan2000pretreatment}. The outcome of interest was the time to tumor
-recurrence, and five baseline covariates were used to predict outcome:
-prostate-specific antigen (PSA, ng/mL), Gleason score sum (GSS, values between 6
-and 10), radiation dose (RD), hormone therapy (HT, yes/no) and clinical stage
-(CS, six values). The study was designed such that a patient's radiation dose
-depended on when the patient entered the study \citep{gerds2013estimating}. This
-in turn implied that the time of censoring depended on the radiation dose. The
-data were re-analyzed in \citep{gerds2013estimating} where a sensitivity
-analysis was conducted based on simulated data. We use the same simulation
-setup, where event and censoring times are generated according to parametric
-Cox-Weibull models estimated from the original data, and the covariates are
-generated according to either marginal Gaussian normal or binomial distributions
-estimated from the original data
-\citep[c.f.,][Section~4.6]{gerds2013estimating}. We refer to this simulation
-setting as ``dependent censoring''. We also considered a simulation setting
-where data were generated in the same way, except that censoring was generated
-completely independently. We refer to this simulation setting as ``independent
-censoring''.
+\citep{kattan2000pretreatment}. The outcome of interest was the time
+to tumor recurrence, and five baseline covariates were used to predict
+outcome: prostate-specific antigen (PSA, ng/mL), Gleason score sum
+(GSS, values between 6 and 10), radiation dose (RD), hormone therapy
+(HT, yes/no) and clinical stage (CS, six values). The study was
+designed such that a patient's radiation dose depended on when the
+patient entered the study \citep{gerds2013estimating}. This in turn
+implied that the time of censoring depended on the radiation dose. The
+data were re-analyzed in \citep{gerds2013estimating} where a
+sensitivity analysis was conducted based on simulated data. We use the
+same simulation setup, where event and censoring times are generated
+according to parametric Cox-Weibull models estimated from the original
+data, and the covariates are generated according to either marginal
+Gaussian normal or binomial distributions estimated from the original
+data \citep[c.f.,][Section~4.6]{gerds2013estimating}. We refer to this
+simulation setting as ``dependent censoring''. We also considered a
+simulation setting where data were generated in the same way, except
+that censoring was generated completely independently. We refer to
+this simulation setting as ``independent censoring''.
 
@@
 
-We first compare the state learner to two IPCW based super learners: The first
-uses a Cox model with main effects to estimate the censoring probabilities
-(IPCW(Cox)), while the second uses the Kaplan-Meier estimator to estimate the
-censoring probabilities (IPCW(KM)). The Cox model for the censoring distribution
-is correctly specified in both simulation settings while the Kaplan Meier
-estimator only estimates the censoring model correctly in the simulation setting
-where censoring is independent. Both IPCW super learners are fitted using the
-\texttt{R}-package \texttt{riskRegression} \citep{Gerds_Ohlendorff_Ozenne_2023}.
-Each super learner provides a learner for the cumulative hazard function for the
-outcome of interest, and from this a risk prediction model can be obtained
-(c.f., equation~(\ref{eq:surv-def}) with $\Lambda_2 = 0$). We measure
-performance of each super learner by calculating the index of prediction
-accuracy (IPA) \citep{kattan2018index} at a fixed time horizon (36 months) for
-the risk prediction model provided by the super learner. The IPA is 1 minus the
-ratio between the model's Brier score and the null model's Brier score, where
-the null model is the model that does not use any covariate information. The IPA
-is approximated using a large (\( n = 20,000 \)) independent data set of
-uncensored data. As a benchmark we calculate the performance of the risk
-prediction model chosen by the oracle selector. The oracle selector uses the
-large data set of uncensored event times to select the learner with the highest
-IPA.
-
-The results are shown in Figure~\ref{fig:ipcw-fail}. We see that when censoring
-is not independent, using the Kaplan-Meier estimator to estimate the censoring
-probabilities provides a risk prediction model with an IPA that is lower than
-the risk prediction model provided by the state learner. The performance of the
-risk prediction model selected by the state learner is similar to the risk
-prediction model selected by the IPCW(Cox) super learner which a priori uses a
-correctly specified model for the censoring distribution. Both these risk
-prediction models are close to the performance of the oracle, except for small
-sample sizes.
+We compare the state learner to two IPCW based super learners: The
+first superlearner, called IPCW(Cox), uses a Cox model with main
+effects to estimate the censoring probabilities, while the second
+super learner, called IPCW(KM), uses the Kaplan-Meier estimator to
+estimate the censoring probabilities. The Cox model for the censoring
+distribution is correctly specified in both simulation settings while
+the Kaplan Meier estimator only estimates the censoring model
+correctly in the simulation setting where censoring is
+independent. Both IPCW super learners are fitted using the
+\texttt{R}-package \texttt{riskRegression}
+\citep{Gerds_Ohlendorff_Ozenne_2023}.
+% ANDERS: det er uklar for mig hvordan
+% de super learners valger/kombinerer de learners der ingaar i library: er det Brier score? over time?
+Each super learner provides a learner for the cumulative hazard
+function for the outcome of interest. From the cumulative hazard
+function a risk prediction model can be obtained (c.f.,
+equation~(\ref{eq:surv-def}) with $\Lambda_2 = 0$). We measure the
+performance of each super learner by calculating the index of
+prediction accuracy (IPA) \citep{kattan2018index} at a fixed time
+horizon (36 months) for the risk prediction model provided by the
+super learner. The IPA is 1 minus the ratio between the model's Brier
+score and the null model's Brier score, where the null model is the
+model that does not use any covariate information. The IPA is
+approximated using a large (\( n = 20,000 \)) independent data set of
+uncensored data. As a benchmark we calculate the performance of the
+risk prediction model chosen by the oracle selector, which uses the
+large data set of uncensored event times to select the learner with
+the highest IPA.
+
+The results are shown in Figure~\ref{fig:ipcw-fail}. We see that in
+the scenario where censoring depends on the covariates, using the
+Kaplan-Meier estimator to estimate the censoring probabilities
+provides a risk prediction model with an IPA that is lower than the
+risk prediction model provided by the state learner. The performance
+of the risk prediction model selected by the state learner is similar
+to the risk prediction model selected by the IPCW(Cox) super learner
+which a priori uses a correctly specified model for the censoring
+distribution. Both these risk prediction models are close to the
+performance of the oracle, except for small sample sizes.
 
@@
   \caption[]{For the risk prediction models provided by each of the super
-    learners, the IPA is plotted against sample size. The results are based on
-    1000 repetitions and the error bars are used to quantify the Monte Carlo
-    uncertainty.}
+    learners, the IPA is plotted against sample size. The results are averages across 
+    1000 simulated data sets and the error bars are used to quantify the Monte Carlo
+    uncertainty.
+    % ANDERS: er det korrekt at det er averages across 1000 data sets? 
+  }
   \label{fig:ipcw-fail}
@@
 We next compare the state learner to the super learner \texttt{survSL}
-\citep{westling2021inference}, as this is the only existing super learner we
-know which works without a pre-specified censoring model. As both the state
-learner and \texttt{survSL} also provides a prediction model for the probability
-of being censored, we compare the performance of both the outcome and the
-censoring modeling using the IPA.
-
-The results are shown in Figure~\ref{fig:zelefski}. We see that the prediction
-models selected for both censoring and outcome have similar or higher API
-compared to the prediction models selected by \texttt{survSL} for most sample
-sizes.
+\citep{westling2021inference}. This is another super learner which
+like the state learner works without a pre-specified censoring
+model. Note that both the state learner and \texttt{survSL} provide a
+prediction model for the event time outcome and also for the
+probability of being censored. Hence, we compare the performance of
+these methods with respect to both the outcome and the censoring
+distribution. Again we use the IPA to quantify the predictive
+performance.
+
+The results are shown in Figure~\ref{fig:zelefski}. We see that for
+most sample sizes, the state learner selected prediction models for
+both censoring and outcome which have similar or higher IPA compared
+to the prediction models selected by \texttt{survSL}.
+% ANDERS: it is a bit lazy to write "for most sample sizes" :) it is
+% slightly annoying that the y-axis scale changes in the 4
+% figures. would that be better with Brier instead of IPA?
 
@@
     censoring probability provided by each of the super learners, the IPA is
-    plotted against sample size. The results are based on 1000 repetitions and
+    plotted against sample size. The results are averages across 1000 repetitions and
     the error bars are used to quantify the Monte Carlo uncertainty.}
+  % ANDERS: would be good to know how how many trainings set individuals were right censored before
+  % the evaluation time point of the IPA. maybe of interest how you chose the evaluation time point.
   \label{fig:zelefski}
@@
 
-\section{Example with prostate cancer study}
+\section{Prostate cancer study}
 \label{sec:real-data-appl}
 
-The prostate cancer data analyzed by \cite{kattan2000pretreatment}, which we
-introduced in Section~\ref{sec:numer-exper}, include a competing event in the
-form of death without tumor recurrence. For illustration, we fit the state
-learner to the original data set consisting of 1,042 patients. We consider death
-without tumor recurrence and recurrence of tumor as two competing events of
-interest. We include the fives learner \texttt{KM}, \texttt{Cox strata CS},
-\texttt{Lasso}, \texttt{Elastic}, and \texttt{RF} which are described in
-Table~\ref{tab:zel-library}. We use the same library of learners to learn
-\( \Lambda_1 \), \( \Lambda_2 \), and $\Gamma$. In this case, \( \Lambda_1 \)
-denotes the cause-specific cumulative hazard function of tumor recurrence, and
-\( \Lambda_2 \) denotes the cause-specific cumulative hazard function of death
-without tumor recurrence.
+% ANDERS: har brug for en gentle intro saetning som fortaeller hvad der kommer: in this section we re-analyse the ...
+%         Det skal ogsaa vaere klart i starten af sektion af der kommer noget om average treatment effect.
+The prostate cancer data analyzed by \cite{kattan2000pretreatment},
+which we introduced in Section~\ref{sec:numer-exper}, include a
+competing event in the form of death without tumor recurrence. For
+illustration, we fit the state learner to the original data set
+consisting of 1,042 patients. We consider death without tumor
+recurrence and recurrence of tumor as competing risks. We include the
+fives learner \texttt{KM}, \texttt{Cox strata CS}, \texttt{Lasso},
+\texttt{Elastic}, and \texttt{RF} which are described in
+Table~\ref{tab:zel-library}. We use the same library of learners to
+learn \( \Lambda_1 \), \( \Lambda_2 \), and $\Gamma$. In this case,
+\( \Lambda_1 \) denotes the cause-specific cumulative hazard function
+of tumor recurrence, and \( \Lambda_2 \) denotes the cause-specific
+cumulative hazard function of death without tumor recurrence.
 
@@
 
-We have proposed a new super learner that can be used with right-censored data
-and competing events. Compared to existing methods, the advantage of the state
-learner is that it does not depend on a pre-specified estimator of the censoring
-distribution, but selects one automatically based on the provided library
-\( \mathcal{B} \). Furthermore, the state learner neither requires evaluation of
-the density of \( \Lambda_j \) nor assumes a particular semi-parametric
-structure for $\Lambda_j$. In the remainder of this section we point to some
-limitations of our proposal and discuss avenues for further research.
-
-A major advantage of the state learner is that performance of each combination
-of learners is measured in terms of observable quantities. This means that no
-additional nuisance parameters need to be estimated to evaluate the loss. The
-drawback with this approach is that we are rarely interested in features of the
-observed data distribution when the data are right-censored. The finite sample
-oracle inequality in Corollary~\ref{cor:oracle-prop} concerns the function
-\( F \), which is a feature of \( P \in \mathcal{P} \), while what we are
-typically interested in is \( \Lambda_j \) or \( S \), which are features of
-\( Q \in \mathcal{Q} \). We emphasise that while the state learner provides us
-with estimates of \( \Lambda_j \) and $\Gamma$ based on libraries
-\( \mathcal{A}_j \) and \( \mathcal{B} \), performance are not assessed directly
-for these parameters, but only jointly for estimation of the parameter \( F \).
-For settings without a competing risk, our numerical studies suggest that
-measuring performance with respect to estimation of \( F \) also leads to good
-performance for estimation of \( S \). Further research on this topic, both
-numerical and theoretical, is warranted.
+We have proposed a new super learner that can be used with
+right-censored data and competing events. Compared to existing
+IPCW-based methods, the advantage of the state learner is that it does
+not depend on a pre-specified estimator of the censoring distribution,
+but selects one automatically based on a library of learners for the
+censoring distribution. Furthermore, the state learner neither
+requires evaluation of the density of \( \Lambda_j \) nor assumes a
+% ANDERS: "density of \Lambda_j" er uklar hvordan er denne density defineret?
+(semi-)parametric form for $\Lambda_j$. In the remainder of this
+section we point to some limitations of our proposal and discuss
+avenues for further research.
+
+A major advantage of the state learner is that the performance of each
+combination of learners can be estimated without (semi)parametric
+assumptions or additional nuisance parameters. The drawback with this
+approach is that we are rarely interested in features of the observed
+data distribution when the data are right-censored: we are only
+interested in the uncensored event time distribution. The finite
+sample oracle inequality in Corollary~\ref{cor:oracle-prop} concerns
+the function \( F \), which is a feature of \( P \in \mathcal{P} \),
+while what we are typically interested in is \( \Lambda_j \) or
+\( S \), which are features of \( Q \in \mathcal{Q} \). We emphasise
+that while the state learner provides us with estimates of
+\( \Lambda_j \) and $\Gamma$ based on libraries \( \mathcal{A}_j \)
+and \( \mathcal{B} \), the performance of these learners is not
+assessed directly for their respective target parameters, but only
+indirectly via the performance of \( F \).
+% ANDERS: er du enig med min omformulering af den sidste saetning?
+For settings without competing risks, our numerical studies suggest
+that measuring the performance of \( F \) also leads to good
+performance for estimation of \( S \). However, further research on
+this issue, both numerical and theoretical, is warranted.
+% ANDERS: ved ikke helt hvad du taenker man kan goer i fremtiden men synes egentig at den sidste saetning kunne godt slettes. :)
 
@@
 Our proposed super learner can be implemented with a broad library of learners
-using existing software, for instance the \texttt{R}-package
-\texttt{riskRegression} \citep{Gerds_Ohlendorff_Ozenne_2023}. Furthermore, while
+and using existing software.
+% For instance the \texttt{R}-package
+% \texttt{riskRegression} \citep{Gerds_Ohlendorff_Ozenne_2023} and the \texttt{R}-package by westling
+Furthermore, while
 the library \( \mathcal{F}(\mathcal{A}_1,\mathcal{A}_2,\mathcal{B}) \) consists
-of \( |\mathcal{A}_1||\mathcal{A}_2||\mathcal{B}| \) learners, as long as we
-have sufficient memory we need only fit
-\( |\mathcal{A}_1| +|\mathcal{A}_2| + |\mathcal{B}| \) learners in each fold. To
+of \( |\mathcal{A}_1||\mathcal{A}_2||\mathcal{B}| \) many learners, we only need to fit
+\( |\mathcal{A}_1| +|\mathcal{A}_2| + |\mathcal{B}| \) many learners in each fold. To
 evaluate the performance of each learner we need to perform
-\( |\mathcal{A}_1||\mathcal{A}_2||\mathcal{B}| \) operations to calculate the
+\( |\mathcal{A}_1||\mathcal{A}_2||\mathcal{B}| \) many operations to calculate the
 integrated Brier score in each hold-out sample, one for each combination of the
 fitted models, but these operations are often negligible compared to fitting the
-models. Hence the state learner is essentially no more computationally demanding
+models. Hence the state learner is essentially not more computationally demanding
 than any procedure that uses super learning to learn $\Lambda_1$, $\Lambda_2$,
@@
 \( F \) directly.
+% ANDERS: how would one estimate F? using HAL? or RF? but such a learner would not provide risk prediction models or would it? if yes you could say it
 

### 2024-05-03 tex/jrss-version/state-learner-jrss-style.tex
clean up section 6.3

@@
 between super learning of the censoring distribution and super
-learning of the event time distribution. In this article we pursue a
+learning of the event time distribution. We pursue a
 different route.
@@
 probability to the test set event times that were not also in the
-training data set, and hence (the logarithm of zero is infitity) the
-partial likelihood loss in the test set would be infitity.
+training data set, and hence (the logarithm of zero is infinity) the
+partial log-likelihood loss in the test set would be infinity.
 
@@
   \label{prop:stric-prop}
-  For \(P_0\in\mathcal{P}\) define
-  \begin{equation*}
-    F^* = \argmin_{F \in \mathcal{H}_{\mathcal{P}}} P_0{[\bar{B}_\tau(F, \blank)]}.
+  If \(P_0\in\mathcal{P}\) then
+  \begin{equation*}
+    F_0 = \argmin_{F \in \mathcal{H}_{\mathcal{P}}} P_0{[\bar{B}_\tau(F, \blank)]},
   % F^* = \argmin_{(\Lambda_1,\Lambda_2,\Gamma) \in \mathcal{M}_{\tau}\otimes\mathcal{M}_{\tau}\otimes\mathcal{M}_{\tau}} P_0{[\bar B_\tau(\phi_{(\Lambda_1,\Lambda_2,\Gamma)}, \blank)]},
   \end{equation*}
-Then \( F^*(t,j,\blank) = F_0(t,j,\blank) \) \( H \)-almost surely for
-any \( j\in \{-1,0,1,2\} \) and almost any \( t \in [0, \tau]\).
+  \( H \)-almost surely for any \( j\in \{-1,0,1,2\} \) and almost any
+  \( t \in [0, \tau]\).
 \end{proposition}
@@
 Corollary~\ref{cor:oracle-prop} is in essence a special case of
-Theorem 2.3 in \cite{vaart2006oracle}.  We assume that we split the
+Theorem 2.3 in \citep{vaart2006oracle}.  We assume that we split the
 data into equally sized folds, and for simplicity of presentation we
@@
 % \( B \).
-In the following we let the space \( \mathcal{H} \) be equipped with the norm
+In the following we let the space \( \mathcal{H}_{\mathcal{P}} \) be equipped with the norm
 \( \| \blank \|_{P_0} \) defined as
@@
 % ANDERS / THOMAS
-We consider 
-anestimator which is obtained from
-estimators of the nuisance parameters $\Lambda_1$, $\Lambda_2$, and $\Gamma$. By
-equations~(\ref{eq:lambdaj}) and~(\ref{eq:gamma}) and the definition of \( F \),
-we have
+We here consider the situation where $\Lambda_1$, $\Lambda_2$, and $\Gamma$ are
+nuisance parameters and an the estimator of interest is obtained from estimators
+of these nuisance parameters. An example is an estimator of the cumulative
+incidence curve, which can be obtained from estimators of $\Lambda_1$ and
+$\Lambda_2$.
+% THOMAS: Jeg ville så også tilføje: We give another example in Section 7.
+By equations~(\ref{eq:lambdaj}) and~(\ref{eq:gamma}) and the definition of
+\( F \), we have
 \begin{equation}
@@
 \end{equation}
-and thus a targeted estimator can also be obtained from an estimator of $F$
-using equation~(\ref{eq:7}). The key feature of a targeted estimator
-is that it is asymptotically equivalent to a sum of i.i.d.\ random variables
-plus a second order remainder term
-
-\citep{bickel1993efficient,fisher2021visually}. For the setting with competing
+and thus an estimator based on \( \hat{\Lambda}_{1,n} \),
+\( \hat{\Lambda}_{2,n} \), and \( \hat{\Gamma}_{n} \) can also be obtained from
+an estimator of $F$ using equation~(\ref{eq:7}). A so-called targeted estimator
+has the key feature that it is asymptotically equivalent to a sum of i.i.d.\
+random variables plus a second order remainder term
+\citep{van2011targeted,hines2022demystifying}. For the setting with competing
 risks, the remainder term is dominated by terms of the form
@@
 \( \smallO_P{(n^{-1/2})}\). Proposition~\ref{prop:dr-structure} states that if
-equation~(\ref{eq:dr-term}) holds for the a targeted estimator based on
-estimators $\hat{\Lambda}_{1,n}$, $\hat{\Lambda}_{2,n}$, and $\hat{\Gamma}_{n}$,
-then a similar product structure holds for a targeted estimator based on
+equation~(\ref{eq:dr-term}) holds for a targeted estimator based on estimators
+$\hat{\Lambda}_{1,n}$, $\hat{\Lambda}_{2,n}$, and $\hat{\Gamma}_{n}$, then a
+similar product structure holds for a targeted estimator based on
 \( \hat{F}_n \). We state the result for the special case that
 \(\hat{M}_{1,n}= \Gamma-\hat{\Gamma}_n \) and
-\(\hat{M}_{2,n} =\Lambda_1-\hat{\Lambda}_{1,n} \), but similar results holds for
+\(\hat{M}_{2,n} =\Lambda_1-\hat{\Lambda}_{1,n} \), but similar results hold for
 any combinations of \( \Gamma-\hat{\Gamma}_n\),

### 2024-05-03 tex/jrss-version/state-learner-jrss-style.tex
changed potential outcome notation

@@
 and \( 1 \) on \( \mathcal{W} \) and that both \( A \) and \( W \) are
-fully observed for all individuals. We use an super learner to
+fully observed for all individuals. We use a super learner to
 estimate $\pi$ \citep{Polley_Ledell_Kennedy_Laan_2023_Superlearn}, and
@@
   \theta_{{\tau}}(Q) =
-  P{(T^{A=1} \leq \tau, D^{A=1}=1)}-
-  P{(T^{A=0} \leq \tau, D^{A=0}=1)},
-\end{equation*}
-where \( (T^A, D^A) \) denotes potential outcomes \citep{hernanRobinsWhatIf}. In
-this case, the interpretation of $\theta_{\tau}$ is the difference in the
-average risk of cause \( 1 \) occurring before time \( \tau \) in the population
-if everyone had been given treatment (\( A=1 \)) compared to if no one had been
-given treatment \( (A=0) \).
+  P{(T^{1} \leq \tau, D^{1}=1)}-
+  P{(T^{0} \leq \tau, D^{0}=1)},
+\end{equation*}
+where \( (T^a, D^a) \), \( a \in \{0,1\} \), denotes potential outcomes
+\citep{hernanRobinsWhatIf}. In this case, the interpretation of $\theta_{\tau}$
+is the difference in the average risk of cause \( 1 \) occurring before time
+\( \tau \) in the population if everyone had been given treatment (\( A=1 \))
+compared to if no one had been given treatment \( (A=0) \).
 

### 2024-05-03 tex/jrss-version/state-learner-jrss-style.tex
answered TAG's comments for sections 8-10

@@
 \end{equation}
-
 Equations~(\ref{eq:1}) and~(\ref{eq:5}) allow us to construct a one-step
@@
 
-In this section we report results from two settings of our simulation
-studies, both settings are without competing risks. In the next
-section, we consider a setting with competing risks.
+In this section we report results from a simulation study without competing
+risks. In the next section, we consider a setting with competing risks. Here we
+first compare the state learner to an IPCW based discrete super learner that
+uses either the Kaplan-Meier estimator or a Cox model to estimate the censoring
+probability. We next compare the state learner to the super learner proposed by
+\cite{westling2021inference}.
 % ANDERS: kan du skrive her hvad forskellen er mellem og formaalet er
@@
 % Maaske separation i to subsections kunne hjaelpe yderliger
+%
+% THOMAS: Jeg ved ikke, om jeg ville sige at der er to simulationstudier, der er
+% bare to ting vi undersøge med det samme studie. Data er genereret på samme
+% måde til at undersøge de to spørgsmål. Har opdateret introen til afsnittet, er
+% det bedre?
 
@@
 % ANDERS: det er uklar for mig hvordan
-% de super learners valger/kombinerer de learners der ingaar i library: er det Brier score? over time?
+% de super learners valger/kombinerer de learners der ingaar i library: er det
+% Brier score? over time?
+% 
+% THOMAS: De er diskrete super learner baseret på Score function med summary
+% measure 'ibs', så det er integrated brier score.
 Each super learner provides a learner for the cumulative hazard
@@
     % ANDERS: er det korrekt at det er averages across 1000 data sets?
+    %
+    % THOMAS: Ja. ...?
   }
@@
 % figures. would that be better with Brier instead of IPA?
+%
+% THOMAS: No, it is worse with Brier score I seem to recall. I have fixed to
+% y-axis horizontally now, it works well for the outcome model but poorly for
+% the censoring. Let me know what you like the best.
 
@@
     the error bars are used to quantify the Monte Carlo uncertainty.}
-  % ANDERS: would be good to know how how many trainings set individuals were right censored before
-  % the evaluation time point of the IPA. maybe of interest how you chose the evaluation time point.
+  % ANDERS: would be good to know how how many trainings set individuals were
+  % right censored before the evaluation time point of the IPA. maybe of
+  % interest how you chose the evaluation time point.
+  %
+  % THOMAS: Good point, but I think that the time of evaluation did not matter
+  % too much. Of course, if you pick a very early time point where nothing has
+  % happened, there is almost no difference in performance.
   \label{fig:zelefski}
@@
 
-% ANDERS: har brug for en gentle intro saetning som fortaeller hvad der kommer: in this section we re-analyse the ...
-%         Det skal ogsaa vaere klart i starten af sektion af der kommer noget om average treatment effect.
-The prostate cancer data analyzed by \cite{kattan2000pretreatment},
-which we introduced in Section~\ref{sec:numer-exper}, include a
-competing event in the form of death without tumor recurrence. For
-illustration, we fit the state learner to the original data set
-consisting of 1,042 patients. We consider death without tumor
-recurrence and recurrence of tumor as competing risks. We include the
-fives learner \texttt{KM}, \texttt{Cox strata CS}, \texttt{Lasso},
-\texttt{Elastic}, and \texttt{RF} which are described in
-Table~\ref{tab:zel-library}. We use the same library of learners to
-learn \( \Lambda_1 \), \( \Lambda_2 \), and $\Gamma$. In this case,
-\( \Lambda_1 \) denotes the cause-specific cumulative hazard function
-of tumor recurrence, and \( \Lambda_2 \) denotes the cause-specific
-cumulative hazard function of death without tumor recurrence.
+% ANDERS: har brug for en gentle intro saetning som fortaeller hvad der kommer:
+% in this section we re-analyse the ... Det skal ogsaa vaere klart i starten af
+% sektion af der kommer noget om average treatment effect.
+%
+% THOMAS: Yes, klart. Fungerer det her:
+In this section we re-analyze the prostate cancer data analyzed by
+\cite{kattan2000pretreatment}, which we introduced in
+Section~\ref{sec:numer-exper}. The data consists of 1,042 patients and includes
+a competing event in the form of death without tumor recurrence. We use the
+state learner to rank learners of the cause-specific cumulative hazard
+functions. We then use the highest ranked combination of learners and the
+estimation strategy outlined in Section~\ref{sec:targeted-learning} to estimate
+the average treatment effect of hormone therapy on death and tumor recurrence.
+
+We consider death without tumor recurrence and recurrence of tumor as competing
+risks. We include the fives learner \texttt{KM}, \texttt{Cox strata CS},
+\texttt{Lasso}, \texttt{Elastic}, and \texttt{RF} which are described in
+Table~\ref{tab:zel-library}. We use the same library of learners to learn
+\( \Lambda_1 \), \( \Lambda_2 \), and $\Gamma$. In this case, \( \Lambda_1 \)
+denotes the cause-specific cumulative hazard function of tumor recurrence, and
+\( \Lambda_2 \) denotes the cause-specific cumulative hazard function of death
+without tumor recurrence.
 
@@
 
-As an illustration, we use the learners of the three cumulative hazard functions
+We use the learners of the three cumulative hazard functions
 selected by the state learner and the estimator defined in
@@
 censoring distribution. Furthermore, the state learner neither
-requires evaluation of the density of \( \Lambda_j \) nor assumes a
+requires that \( \Lambda_j \) can be written as an integral with respect to
+Lebesgue measure nor assumes a
 % ANDERS: "density of \Lambda_j" er uklar hvordan er denne density defineret?
+%
+% THOMAS: Better now?
 (semi-)parametric form for $\Lambda_j$. In the remainder of this
@@
 % ANDERS: er du enig med min omformulering af den sidste saetning?
+%
+% THOMAS: Yes, super godt. 
 For settings without competing risks, our numerical studies suggest
@@
 this issue, both numerical and theoretical, is warranted.
-% ANDERS: ved ikke helt hvad du taenker man kan goer i fremtiden men synes egentig at den sidste saetning kunne godt slettes. :)
+% ANDERS: ved ikke helt hvad du taenker man kan goer i fremtiden men synes
+% egentig at den sidste saetning kunne godt slettes. :)
+%
+% THOMAS: Forstår ikke helt din kommentar. Hvilken sætning, den ovenover din
+% kommentar eller den nedenunder? :)
 
@@
 \( F \) directly.
-% ANDERS: how would one estimate F? using HAL? or RF? but such a learner would not provide risk prediction models or would it? if yes you could say it
+% ANDERS: how would one estimate F? using HAL? or RF? but such a learner would
+% not provide risk prediction models or would it? if yes you could say it
+%
+% THOMAS: I don't think that estimators for F exists -- that is what Vi is
+% working on, right? Given F we can always obtain \Lambda_1 and \Lambda_2
+% (equation 8 or eq:7 in tex) and then we can get predictions from these.
 

### 2024-05-08 tex/jrss-version/state-learner-jrss-style.tex
progress

@@
 
-% ANDERS / THOMAS
-We here consider the situation where $\Lambda_1$, $\Lambda_2$, and $\Gamma$ are
-nuisance parameters and an the estimator of interest is obtained from estimators
-of these nuisance parameters. An example is an estimator of the cumulative
-incidence curve, which can be obtained from estimators of $\Lambda_1$ and
-$\Lambda_2$.
-% THOMAS: Jeg ville så også tilføje: We give another example in Section 7.
+In this section we demonstrate a theoretical property of the state
+learner which is useful for targeted learning (c.f., Section
+\ref{sec:targeted-learning}). Specifically we consider an estimator of
+a target parameter which is obtained by substituting the state learner
+estimates of the nuisance parameters $\Lambda_1$, $\Lambda_2$, and
+$\Gamma$. An example is an estimator of the cumulative incidence
+curve, which can be obtained from estimators of $\Lambda_1$ and
+$\Lambda_2$. Another example is provided in Section \ref{sec:targeted-learning}.
 By equations~(\ref{eq:lambdaj}) and~(\ref{eq:gamma}) and the definition of
@@
 
-In this section we report results from a simulation study without competing
-risks. In the next section, we consider a setting with competing risks. Here we
-first compare the state learner to an IPCW based discrete super learner that
-uses either the Kaplan-Meier estimator or a Cox model to estimate the censoring
-probability. We next compare the state learner to the super learner proposed by
-\cite{westling2021inference}.
-% ANDERS: kan du skrive her hvad forskellen er mellem og formaalet er
-% med de to simulationsstudierscenarier?
-% i den ene sammenligner vi med IPCW i den anden med westling, ik?
-% Maaske en figur (flowchart) eller tabel kunne hjaelpe laeseren at faa overblik.
-% Maaske separation i to subsections kunne hjaelpe yderliger
-%
-% THOMAS: Jeg ved ikke, om jeg ville sige at der er to simulationstudier, der er
-% bare to ting vi undersøge med det samme studie. Data er genereret på samme
-% måde til at undersøge de to spørgsmål. Har opdateret introen til afsnittet, er
-% det bedre?
-
-We generate data according to a distribution motivated from a real
-dataset in which censoring depends on the baseline covariates. We
-simulate data based on a prostate cancer study described in
+In this section we report results from a simulation study. In the
+first part we consider estimation of the conditional survival
+function. Here we compare the state learner to two IPCW based discrete
+super learners that use either the Kaplan-Meier estimator or a Cox
+model to estimate the censoring probability
+\cite{gonzalez2021stacked}.  In the second part we compare the state
+learner to the super learner proposed by \cite{westling2021inference}.
+
+In both parts we use the same data generating mechanism. We generate
+data according to a distribution motivated from a real dataset in
+which censoring depends on the baseline covariates. We simulate data
+based on a prostate cancer study described in
 \citep{kattan2000pretreatment}. The outcome of interest was the time
@@
 \citep{Gerds_Ohlendorff_Ozenne_2023}.
-% ANDERS: det er uklar for mig hvordan
-% de super learners valger/kombinerer de learners der ingaar i library: er det
-% Brier score? over time?
-% 
-% THOMAS: De er diskrete super learner baseret på Score function med summary
-% measure 'ibs', så det er integrated brier score.
+% ANDERS: describe that you use a single evaluation time point and say how many censored, event happen before it.
 Each super learner provides a learner for the cumulative hazard
@@
     uncertainty.
-    % ANDERS: er det korrekt at det er averages across 1000 data sets?
-    %
-    % THOMAS: Ja. ...?
   }

### 2024-05-10 tex/jrss-version/state-learner-jrss-style.tex
add event and censoring probabiltiy

@@
 
-In this section we report results from a simulation study. In the
-first part we consider estimation of the conditional survival
-function. Here we compare the state learner to two IPCW based discrete
-super learners that use either the Kaplan-Meier estimator or a Cox
-model to estimate the censoring probability
-\cite{gonzalez2021stacked}.  In the second part we compare the state
-learner to the super learner proposed by \cite{westling2021inference}.
+In this section we report results from a simulation study. In the first part we
+consider estimation of the conditional survival function. Here we compare the
+state learner to two IPCW based discrete super learners that use either the
+Kaplan-Meier estimator or a Cox model to estimate the censoring probability
+\citep{gonzalez2021stacked}. In the second part we compare the state learner to
+the super learner proposed by \cite{westling2021inference}.
 
@@
 \citep{Gerds_Ohlendorff_Ozenne_2023}.
-% ANDERS: describe that you use a single evaluation time point and say how many censored, event happen before it.
-Each super learner provides a learner for the cumulative hazard
-function for the outcome of interest. From the cumulative hazard
-function a risk prediction model can be obtained (c.f.,
-equation~(\ref{eq:surv-def}) with $\Lambda_2 = 0$). We measure the
-performance of each super learner by calculating the index of
-prediction accuracy (IPA) \citep{kattan2018index} at a fixed time
-horizon (36 months) for the risk prediction model provided by the
-super learner. The IPA is 1 minus the ratio between the model's Brier
-score and the null model's Brier score, where the null model is the
-model that does not use any covariate information. The IPA is
-approximated using a large (\( n = 20,000 \)) independent data set of
-uncensored data. As a benchmark we calculate the performance of the
-risk prediction model chosen by the oracle selector, which uses the
-large data set of uncensored event times to select the learner with
-the highest IPA.
+% ANDERS: describe that you use a single evaluation time point and say how many
+% censored, event happen before it.
+%
+% THOMAS: Done below now, OK?
+%
+% | time | sim_setting | true_events | true_cens | at_risk |
+% |------+-------------+-------------+-----------+---------|
+% |   36 | original    |      24.619 |    61.853 |  25.774 |
+% |   36 | indep_cens  |      24.674 |    38.740 |  46.141 |
+%
+The IPCW super learners use the integrated Brier score up to a fixed time
+horizon (36 months). The marginal risk of the event before this time horizon is
+25\%. Under the ``dependent censoring'', setting the marginal censoring
+probability before this time horizon is 62\%; under the ``independent
+censoring'' setting, the marginal censoring probability before this time horizon
+is 39\%.
+
+Each super learner provides a learner for the cumulative hazard function for the
+outcome of interest. From the cumulative hazard function a risk prediction model
+can be obtained (c.f., equation~(\ref{eq:surv-def}) with $\Lambda_2 = 0$). We
+measure the performance of each super learner by calculating the index of
+prediction accuracy (IPA) \citep{kattan2018index} at a fixed time horizon (36
+months) for the risk prediction model provided by the super learner. The IPA is
+1 minus the ratio between the model's Brier score and the null model's Brier
+score, where the null model is the model that does not use any covariate
+information. The IPA is approximated using a large (\( n = 20,000 \))
+independent data set of uncensored data. As a benchmark we calculate the
+performance of the risk prediction model chosen by the oracle selector, which
+uses the large data set of uncensored event times to select the learner with the
+highest IPA.
 

### 2024-05-10 experiments/_targets.R

@@ -3,9 +3,9 @@
 ## Author: Anders Munch
 ## Created: Nov 14 2023 (09:52) 
 ## Version: 
-## Last-Updated: Jan  5 2024 (14:13) 
+## Last-Updated: May 10 2024 (10:03) 
 ##           By: Anders Munch
-##     Update #: 40
+##     Update #: 44
 #----------------------------------------------------------------------
 ## 
 ### Commentary: 
@@ -16,6 +16,9 @@
 ### Code:
 library(targets)
 library(data.table)
+library(here)
+tar_source(here("R-code/functions"))
+tar_source(here("experiments/zel-sim2-1-n-events-mc-fun.R"))
 
 combine_results <- function(job_name, results_dir){
     fls = list.files(results_dir)
@@ -33,6 +36,7 @@ list(
     tar_target(zel_all0_results, combine_results("zel-all-settings0", results_dir = results_dir)),
     tar_target(zel_all2_results, combine_results("zel-all-settings2", results_dir = results_dir)),
     tar_target(zel_sim2_1, combine_results("zel-sim2-1", results_dir = results_dir)),
+    tar_target(zel_sim2_1_n_events, zel_sim2_1_n_events_mc_fun()),
     tar_target(ipcw_fail_sim, {
         run0 = combine_results("ipcw-fail-sim", results_dir = results_dir)
         state_more = combine_results("ipcw-fail-sim-more-state-learners", results_dir = results_dir)

### 2024-05-10 experiments/zel-sim2-1-n-events-mc-fun.R

@@ -1,0 +1,18 @@
+zel_sim2_1_n_events_mc_fun <- function(){
+    ## Write sim setup here
+    eval_times = seq(6, 36, 6)
+    sim_sets = list(original = simZelefsky_wrapper,
+                    indep_cens = simZelefsky_indep_cens_wrapper)
+    out = do.call(rbind, lapply(seq_along(sim_sets), function(pp){
+        sim_set_name = names(sim_sets)[pp]
+        mc_sim = sim_sets[[pp]](n = 100000)
+        do.call(rbind, lapply(eval_times, function(tt){
+            data.table(sim_setting = sim_set_name,
+                       time = tt,
+                       true_events = mc_sim[, sum(true_time<tt)],
+                       true_cens = mc_sim[, sum(cens_time<tt)],
+                       at_risk = mc_sim[, sum(time>tt)])[]
+        }))
+    }))
+    return(out[])
+}

### 2024-05-13 tex/jrss-version/state-learner-jrss-style.tex
last comments

@@
 \citep{Gerds_Ohlendorff_Ozenne_2023}.
-% ANDERS: describe that you use a single evaluation time point and say how many
-% censored, event happen before it.
-%
-% THOMAS: Done below now, OK?
 %
@@
 horizon (36 months). The marginal risk of the event before this time horizon is
-25\%. Under the ``dependent censoring'', setting the marginal censoring
-probability before this time horizon is 62\%; under the ``independent
-censoring'' setting, the marginal censoring probability before this time horizon
-is 39\%.
+\(\sim 24.6\)\%. Under the ``dependent censoring'' setting the marginal censoring
+probability before the time horizon is \(\sim 61.9\)\%. Under the ``independent
+censoring'' setting the marginal censoring probability before this time horizon
+is \(\sim 38.7\)\%.
 
@@
 to the prediction models selected by \texttt{survSL}.
-% ANDERS: it is a bit lazy to write "for most sample sizes" :) it is
-% slightly annoying that the y-axis scale changes in the 4
-% figures. would that be better with Brier instead of IPA?
-%
-% THOMAS: No, it is worse with Brier score I seem to recall. I have fixed to
-% y-axis horizontally now, it works well for the outcome model but poorly for
-% the censoring. Let me know what you like the best.
-
+% ANDERS: Divide Fig 3 into two figures. keep y-axis fix for Outcome model panels but use different y-axes for  Censoring model.
 \begin{figure}
@@
   \caption[]{For the risk prediction models of both the outcome and the
-    censoring probability provided by each of the super learners, the IPA is
+    censoring probability provided by each of the super learners, the IPA at the fixed time horizon is
     plotted against sample size. The results are averages across 1000 repetitions and
     the error bars are used to quantify the Monte Carlo uncertainty.}
-  % ANDERS: would be good to know how how many trainings set individuals were
-  % right censored before the evaluation time point of the IPA. maybe of
-  % interest how you chose the evaluation time point.
-  %
-  % THOMAS: Good point, but I think that the time of evaluation did not matter
-  % too much. Of course, if you pick a very early time point where nothing has
-  % happened, there is almost no difference in performance.
+  % ANDERS: add marginal censoring probability to panel titles
   \label{fig:zelefski}
@@
 % THOMAS: Yes, klart. Fungerer det her:
-In this section we re-analyze the prostate cancer data analyzed by
-\cite{kattan2000pretreatment}, which we introduced in
-Section~\ref{sec:numer-exper}. The data consists of 1,042 patients and includes
-a competing event in the form of death without tumor recurrence. We use the
-state learner to rank learners of the cause-specific cumulative hazard
-functions. We then use the highest ranked combination of learners and the
-estimation strategy outlined in Section~\ref{sec:targeted-learning} to estimate
-the average treatment effect of hormone therapy on death and tumor recurrence.
-
-We consider death without tumor recurrence and recurrence of tumor as competing
-risks. We include the fives learner \texttt{KM}, \texttt{Cox strata CS},
-\texttt{Lasso}, \texttt{Elastic}, and \texttt{RF} which are described in
-Table~\ref{tab:zel-library}. We use the same library of learners to learn
-\( \Lambda_1 \), \( \Lambda_2 \), and $\Gamma$. In this case, \( \Lambda_1 \)
-denotes the cause-specific cumulative hazard function of tumor recurrence, and
-\( \Lambda_2 \) denotes the cause-specific cumulative hazard function of death
-without tumor recurrence.
+In this section we use the prostate cancer data of
+\cite{kattan2000pretreatment} to illustrate the use of the state
+learner in the presence of competing risks. We have introduced the
+data in Section~\ref{sec:numer-exper}. The data consists of 1,042
+patients who are followed from start of followup until tumor
+recurrence, death without tumor recurrence or end of followup
+(censored) whatever came first. For the sole purpose of illustration,
+we estimate the average treatment effect of hormone therapy on death
+and tumor recurrence. To do this we adapt the estimation strategy of
+Section~\ref{sec:targeted-learning} as follows.  We use the state
+learner to rank libraries of learners for the cause-specific
+cumulative hazard functions of tumor recurrence, death without tumor
+recurrence, and censoring.  The libraries of learners include the five
+learners described in Table~\ref{tab:zel-library}.  We use the same
+set of learners to learn the cumulative hazard function of tumor
+recurrence \( \Lambda_1 \), the cumulative hazard function of death
+without tumor recurrence \( \Lambda_2 \), and the cumulative hazard
+function of the conditional censoring distribution $\Gamma$. 
+We then use the highest ranked combination
+of learners and apply formula \ref{eq:one-step-comp-ate}.
 
@@
 censoring distribution. Furthermore, the state learner neither
-requires that \( \Lambda_j \) can be written as an integral with respect to
-Lebesgue measure nor assumes a
-% ANDERS: "density of \Lambda_j" er uklar hvordan er denne density defineret?
-%
-% THOMAS: Better now?
-(semi-)parametric form for $\Lambda_j$. In the remainder of this
-section we point to some limitations of our proposal and discuss
-avenues for further research.
+requires that \( \Lambda_j \) can be written as an integral with
+respect to Lebesgue measure, nor assumes a (semi-)parametric form for
+$\Lambda_j$. In the remainder of this section we point to some
+limitations of our proposal and discuss avenues for further research.
 
@@
 indirectly via the performance of \( F \).
-% ANDERS: er du enig med min omformulering af den sidste saetning?
-%
-% THOMAS: Yes, super godt. 
 For settings without competing risks, our numerical studies suggest
 that measuring the performance of \( F \) also leads to good
-performance for estimation of \( S \). However, further research on
-this issue, both numerical and theoretical, is warranted.
-% ANDERS: ved ikke helt hvad du taenker man kan goer i fremtiden men synes
-% egentig at den sidste saetning kunne godt slettes. :)
-%
-% THOMAS: Forstår ikke helt din kommentar. Hvilken sætning, den ovenover din
-% kommentar eller den nedenunder? :)
-
+performance for estimation of \( S \).
 % In the context of targeted learning, a drawback of the state learner is that the
@@
 \( \mathcal{F} \) from libraries for learning \( \Lambda_1 \), $\Lambda_2$, and
-$\Gamma$, it would also be of interest to consider learners that estimate
+$\Gamma$, it could also be of interest to consider learners that estimate
 \( F \) directly.

### 2024-05-14 tex/jrss-version/old/state-learner-jrss-style.tex
moved to updated template

--- 2024-05-13 tex/jrss-version/state-learner-jrss-style.tex
@@
 
-The results are shown in Figure~\ref{fig:zelefski}. We see that for
-most sample sizes, the state learner selected prediction models for
-both censoring and outcome which have similar or higher IPA compared
-to the prediction models selected by \texttt{survSL}.
-% ANDERS: Divide Fig 3 into two figures. keep y-axis fix for Outcome model panels but use different y-axes for  Censoring model.
+The results are shown in Figures~\ref{fig:zelefski-out}
+and~\ref{fig:zelefski-cens}. We see that for most sample sizes, the state
+learner selected prediction models for both censoring and outcome which have
+similar or higher IPA compared to the prediction models selected by
+\texttt{survSL}.
 \begin{figure}
   \centering %
-  \includeFigCond[width=1\linewidth]{experiment-fig-sl-survSL.pdf}
-  \caption[]{For the risk prediction models of both the outcome and the
-    censoring probability provided by each of the super learners, the IPA at the fixed time horizon is
-    plotted against sample size. The results are averages across 1000 repetitions and
+  \includeFigCond[width=1\linewidth]{experiment-fig-sl-survSL-out.pdf}
+  \caption[]{For the risk prediction models of the outcome provided by each
+    of the super learners, the IPA at the fixed time horizon is plotted against
+    sample size. The results are averages across 1000 repetitions and the error
+    bars are used to quantify the Monte Carlo uncertainty.}
+  \label{fig:zelefski-out}
+\end{figure}
+
+\begin{figure}
+  \centering %
+  \includeFigCond[width=.7\linewidth]{experiment-fig-sl-survSL-cens.pdf}
+  \caption[]{For the risk prediction models of the censoring model provided by
+    each of the super learners, the IPA at the fixed time horizon is plotted
+    against sample size. The results are averages across 1000 repetitions and
     the error bars are used to quantify the Monte Carlo uncertainty.}
-  % ANDERS: add marginal censoring probability to panel titles
-  \label{fig:zelefski}
+  \label{fig:zelefski-cens}
 \end{figure}
@@
 
-% ANDERS: har brug for en gentle intro saetning som fortaeller hvad der kommer:
-% in this section we re-analyse the ... Det skal ogsaa vaere klart i starten af
-% sektion af der kommer noget om average treatment effect.
-%
-% THOMAS: Yes, klart. Fungerer det her:
 In this section we use the prostate cancer data of
@@
 \( F \) directly.
-% ANDERS: how would one estimate F? using HAL? or RF? but such a learner would
-% not provide risk prediction models or would it? if yes you could say it
-%
-% THOMAS: I don't think that estimators for F exists -- that is what Vi is
-% working on, right? Given F we can always obtain \Lambda_1 and \Lambda_2
-% (equation 8 or eq:7 in tex) and then we can get predictions from these.
 

### 2024-05-14 tex/jrss-version/state-learner-jrss-style.tex
uk spelling

@@
-\documentclass{statsoc}
-\usepackage[a4paper]{geometry}
-
-\usepackage{natbib, dsfont, enumitem,
-  amssymb,soul,xcolor,amsmath,graphicx,verbatim,pgfplots,tikz,prodint,booktabs}
-
-\usepackage[tablesonly,heads,nolists]{endfloat}
-
-%% Theorem environment
-% \theoremstyle{plain} % plain italic style
-\newtheorem{theorem}{Theorem}
-% \numberwithin{theorem}{section}
+%%
+%% Copyright 2022 OXFORD UNIVERSITY PRESS
+%%
+%% This file is part of the 'oup-authoring-template Bundle'.
+%% ---------------------------------------------
+%%
+%% It may be distributed under the conditions of the LaTeX Project Public
+%% License, either version 1.2 of this license or (at your option) any
+%% later version.  The latest version of this license is in
+%%    http://www.latex-project.org/lppl.txt
+%% and version 1.2 or later is part of all distributions of LaTeX
+%% version 1999/12/01 or later.
+%%
+%% The list of all files belonging to the 'oup-authoring-template Bundle' is
+%% given in the file `manifest.txt'.
+%%
+%% Template article for OXFORD UNIVERSITY PRESS's document class `oup-authoring-template'
+%% with bibliographic references
+%%
+
+%%%CONTEMPORARY%%%
+% \documentclass[unnumsec,webpdf,contemporary,large]{oup-authoring-template}%
+\documentclass[unnumsec,webpdf,contemporary,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
+%\documentclass[unnumsec,webpdf,contemporary,medium]{oup-authoring-template}
+%\documentclass[unnumsec,webpdf,contemporary,small]{oup-authoring-template}
+
+%%%MODERN%%%
+%\documentclass[unnumsec,webpdf,modern,large]{oup-authoring-template}
+% \documentclass[unnumsec,webpdf,modern,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
+%\documentclass[unnumsec,webpdf,modern,medium]{oup-authoring-template}
+%\documentclass[unnumsec,webpdf,modern,small]{oup-authoring-template}
+
+%%%TRADITIONAL%%%
+%\documentclass[unnumsec,webpdf,traditional,large]{oup-authoring-template}
+% \documentclass[unnumsec,webpdf,traditional,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
+%\documentclass[unnumsec,namedate,webpdf,traditional,medium]{oup-authoring-template}
+%\documentclass[namedate,webpdf,traditional,small]{oup-authoring-template}
+
+\onecolumn % for one column layouts
+
+%\usepackage{showframe}
+
+% \graphicspath{{Fig/}}
+
+% line numbers
+%\usepackage[mathlines, switch]{lineno}
+%\usepackage[right]{lineno}
+
+\theoremstyle{thmstyleone}%\
+\newtheorem{theorem}{Theorem}%  meant for continuous numbers
+%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
+%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
+% \newtheorem{proposition}[theorem]{Proposition}%
+% to get separate numbers for theorem and proposition etc.:
+\newtheorem{proposition}{Proposition}
 \newtheorem{lemma}{Lemma}
-\newtheorem{proposition}{Proposition}
 \newtheorem{corollary}{Corollary}
+\theoremstyle{thmstyletwo}%
+\newtheorem{example}{Example}%
+\newtheorem{remark}{Remark}%
+\theoremstyle{thmstylethree}%
+\newtheorem{definition}{Definition}
+
+%% OLD
+
+% \usepackage{natbib, dsfont,
+% amssymb,soul,xcolor,amsmath,graphicx,verbatim,pgfplots,tikz,prodint,booktabs}
+
+\usepackage{dsfont, booktabs}
+
+% \usepackage[tablesonly,heads,nolists]{endfloat}
 
@@
 
-\bibliographystyle{rss}
+% \bibliographystyle{rss}
 
@@
 
-%% TODO: Update figures with name 'figure-'.
-%% Write macro for including figure or not.
-
-% \newcommand{\includeFigCond}[2]{
-%   \ifx\nofig\undefined %
-%     \includegraphics[#1]{#2} %
-%   \else %
-%     \texttt{#2} %
-%   \fi %
-% }
 \newcommand{\includeFigCond}[2][]{
@@
 
-\title[The state learner]{The state learner \\ {-- a super learner for right-censored data}}
-\author{Anders Munch}
-\address{Section of Biostatistics, University of Copenhagen, Copenhagen, Denmark}
-\email{a.munch@sund.ku.dk}
-\author[Munch {\it et al.}]{Thomas A.\ Gerds}
-\address{Section of Biostatistics, University of Copenhagen, Copenhagen, Denmark}
-
 \begin{document}
 
-\begin{abstract}
-  In survival analysis, prediction models are needed as stand-alone
-  tools and in applications of causal inference to estimate nuisance
-  parameters. The super learner is a machine learning algorithm which
-  combines a library of prediction models into a meta learner based on
-  cross-validated loss. In right censored data, loss needs to be estimated somehow.
-  % based on a model for the censoring distribution.
-  % Unfortunately, the
-  % commonly used partial likelihood loss is not suited for super learning, and
-  % inverse probability of censoring weighted loss functions require a
-  % pre-specified estimator of the censoring distribution. To relax this,
-  We introduce the state learner, a new super learner for survival analysis, which
-  simultaneously evaluates libraries
-  of prediction models for the event of interest and the censoring
-  distribution. The state learner can
-be applied to all types of survival models, works in the presence
-of competing risks, and does not require a single pre-specified
-estimator of the conditional censoring distribution.
-  We establish an oracle inequality for the state learner and
+%% NEW
+
+\journaltitle{Journal of the Royal Statistical Society Series B: Statistical
+  Methodology} %
+\DOI{DOI HERE} %
+\copyrightyear{xxxx} %
+\pubyear{yyyy}
+\access{Advance Access Publication Date: Day Month Year}
+\appnotes{Paper}
+
+\firstpage{1}
+
+%\subtitle{Subject Section}
+
+\title[The state learner]{The state learner -- a super learner for
+  right-censored data}
+
+\author[1,$\ast$]{Anders Munch}
+\author[1]{Thomas A.\ Gerds}
+
+\authormark{Munch et al.}
+
+\address[1]{\orgdiv{Section of Biostatistics}, \orgname{University of
+    Copenhagen}, \orgaddress{\street{Østerfarigmagsgade 5}, \postcode{2100}, \state{Copenhagen}, \country{Denmark}}}
+
+\corresp[$\ast$]{Corresponding author. \href{email:email-id.com}{a.munch@sund.ku.dk}}
+
+% \received{Date}{0}{Year}
+% \revised{Date}{0}{Year}
+% \accepted{Date}{0}{Year}
+
+\abstract{In survival analysis, prediction models are needed as stand-alone
+  tools and in applications of causal inference to estimate nuisance parameters.
+  The super learner is a machine learning algorithm which combines a library of
+  prediction models into a meta learner based on cross-validated loss. In right
+  censored data, loss needs to be estimated somehow. We introduce the state
+  learner, a new super learner for survival analysis, which simultaneously
+  evaluates libraries of prediction models for the event of interest and the
+  censoring distribution. The state learner can be applied to all types of
+  survival models, works in the presence of competing risks, and does not
+  require a single pre-specified estimator of the conditional censoring
+  distribution. We establish an oracle inequality for the state learner and
   investigate its performance through numerical experiments. We illustrate how
@@
   setting without having to pre-specify models for neither the cause-specific
-  hazards of interest nor the censoring distribution.
-\end{abstract}
+  hazards of interest nor the censoring distribution.}
 
@@
   right-censored data, super learner}
+
+\maketitle
 
@@
 survival function of the conditional censoring distribution. The
-distribution is characterized by
+distribution is characterised by
 \begin{equation}\label{eq:parametrizeP}
@@
   \centering %
-  \includeFigCond[width=.5\textwidth]{figure-multi-state-process.pdf}
+  \includeFigCond[width=.3\textwidth]{figure-multi-state-process.pdf}
   \caption{Illustration of the multi-state process \(\eta\) used by
@@
 Under conditional independent censoring each quadruple
-$(\Lambda_{1}, \Lambda_{2}, \Gamma, H)$ characterizes a distribution
+$(\Lambda_{1}, \Lambda_{2}, \Gamma, H)$ characterises a distribution
 \(P\in\mathcal P\), c.f.\ equation~\eqref{eq:parametrizeP}, which in turn
@@
   &= \{ \phi_{a_1,a_2, b} : a_1 \in \mathcal{A}_1, a_2 \in \mathcal{A}_2, b \in \mathcal{B}\},
-    \intertext{where in correspondance with  the relations in equation \eqref{eq:transition},} 
+    \intertext{where in correspondence with  the relations in equation \eqref{eq:transition},} 
     \phi_{a_1,a_2, b}(\data_n)(t,1,x) &= \int_0^t e^{\{-a_1(\data_n)(s \mid x)-a_2(\data_n)(s \mid x) - b(\data_n)(s \mid x)\} }  a_1(\data_n)(\diff s \mid x),\\
@@
   \bigO(n^{-\alpha}) \), for some \( \alpha\leq 1 \).
-  \begin{enumerate}[label=(\roman*), topsep=0pt]
+  \begin{enumerate}
   \item If $\alpha=1$ then
@@
   \centering %
-  \includeFigCond[width=1\linewidth]{experiment-fig-sl-ipcw.pdf}
+  \includeFigCond[width=.8\linewidth]{experiment-fig-sl-ipcw.pdf}
   \caption[]{For the risk prediction models provided by each of the super
@@
 
-The results are shown in Figure~\ref{fig:zelefski}. We see that for
-most sample sizes, the state learner selected prediction models for
-both censoring and outcome which have similar or higher IPA compared
-to the prediction models selected by \texttt{survSL}.
-% ANDERS: Divide Fig 3 into two figures. keep y-axis fix for Outcome model panels but use different y-axes for  Censoring model.
+The results are shown in Figures~\ref{fig:zelefski-out}
+and~\ref{fig:zelefski-cens}. We see that for most sample sizes, the state
+learner selected prediction models for both censoring and outcome which have
+similar or higher IPA compared to the prediction models selected by
+\texttt{survSL}.
 \begin{figure}
   \centering %
-  \includeFigCond[width=1\linewidth]{experiment-fig-sl-survSL.pdf}
-  \caption[]{For the risk prediction models of both the outcome and the
-    censoring probability provided by each of the super learners, the IPA at the fixed time horizon is
-    plotted against sample size. The results are averages across 1000 repetitions and
+  \includeFigCond[width=.8\linewidth]{experiment-fig-sl-survSL-out.pdf}
+  \caption[]{For the risk prediction models of the outcome provided by each
+    of the super learners, the IPA at the fixed time horizon is plotted against
+    sample size. The results are averages across 1000 repetitions and the error
+    bars are used to quantify the Monte Carlo uncertainty.}
+  \label{fig:zelefski-out}
+\end{figure}
+
+\begin{figure}
+  \centering %
+  \includeFigCond[width=.5\linewidth]{experiment-fig-sl-survSL-cens.pdf}
+  \caption[]{For the risk prediction models of the censoring model provided by
+    each of the super learners, the IPA at the fixed time horizon is plotted
+    against sample size. The results are averages across 1000 repetitions and
     the error bars are used to quantify the Monte Carlo uncertainty.}
-  % ANDERS: add marginal censoring probability to panel titles
-  \label{fig:zelefski}
+  \label{fig:zelefski-cens}
 \end{figure}
@@
   \centering %
-  \includeFigCond[width=1\linewidth]{zelefski-real-data.pdf}
+  \includeFigCond[width=.9\linewidth]{zelefski-real-data.pdf}
   \caption[]{The results of applying the 125 combinations of learners to the
@@
   \centering%
-  \includeFigCond[width=1\linewidth]{zelefsky-data-target-par.pdf}
+  \includeFigCond[width=.8\linewidth]{zelefsky-data-target-par.pdf}
   \caption[]{Estimates of the average treatment effect of hormone therapy on the
@@
 
-\appendix
+\begin{appendices}
 
@@
 
+\end{appendices}
+
+\bibliographystyle{abbrvnat}
 \bibliography{bib.bib}

### 2024-05-14 tex/jrss-version/state-learner-jrss-style.tex
separate figures

@@
 % \documentclass[unnumsec,webpdf,contemporary,large]{oup-authoring-template}%
-\documentclass[unnumsec,webpdf,contemporary,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
+\documentclass[numsec,webpdf,contemporary,medium,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
 %\documentclass[unnumsec,webpdf,contemporary,medium]{oup-authoring-template}
@@
 
-% \usepackage[tablesonly,heads,nolists]{endfloat}
+\usepackage[tablesonly,heads,nolists]{endfloat}
 
@@
 %% Toggle with or without figures
-% \def\nofig{1}
+\def\nofig{1}
 

### 2024-05-14 tex/jrss-version/state-learner-jrss-style.tex
added number to multi state diagram

@@
 \copyrightyear{xxxx} %
-\pubyear{yyyy}
-\access{Advance Access Publication Date: Day Month Year}
+\pubyear{yyyy} \access{Advance Access Publication Date: Day Month Year}
 \appnotes{Paper}
@@
 
-\corresp[$\ast$]{Corresponding author. \href{email:email-id.com}{a.munch@sund.ku.dk}}
+\corresp[$\ast$]{Address for correspondence: \href{email:email-id.com}{a.munch@sund.ku.dk}}
 
@@
 % \citep{van2011targeted, benkeser2016highly,van2017generally}.
-
-While the general idea of combining strong learners based on
-cross-validation data stems from earlier work
-\citep{wolpert1992stacked,breiman1996stacked}, the name super learner
-is justified by an oracle inequality
-\citep{van2003unicv,vaart2006oracle}: The super learner is guaranteed
-to perform almost as well as the model which minimises the expected
-performance, i.e., the model we would select if we could evaluate the
-prediction performance in an infinite hold-out sample. The choice of
+While the general idea of combining strong learners based on cross-validation
+data stems from earlier work \citep{wolpert1992stacked,breiman1996stacked}, the
+name super learner is justified by an oracle inequality
+\citep{van2003unicv,vaart2006oracle}: The super learner is guaranteed to perform
+almost as well as the model which minimises the expected performance, i.e., the
+model we would select if we could evaluate the prediction performance in an
+infinite hold-out sample.
+
+The choice of
 the loss function for super learning in survival analysis deserves a
@@
 weighted loss
-\citep{polley2011-sl-cens,keles2004asymptotically,golmakani2020super,westling2021inference}.
+\citep{polley2011-sl-cens,keles2004asymptotically,golmakani2020super,gonzalez2021stacked,westling2021inference}.
 The use of the partial log-likelihood loss, however, restricts the
@@
 In this article we define the state learner, a new super learner for
-right-censored data, which simultaneously estimates the expected loss
-of learners of the event time distribution and the censoring
-distribution. The loss function which is used to define the state
-learner is based on the observed variables, that is the minimum and
-the order of the censoring time and the event
-time.
+right-censored data, which simultaneously estimates the expected loss of
+learners of the event time distribution and the censoring distribution. The loss
+function which is used to define the state learner is based on the observed
+variables, that is, the minimum and the order of the censoring time and the
+event time.
 
 The state learner can be applied to all types of survival models,
-works in the presence of competing risks, and does not require a
+it can handle competing risks, and it does not require a
 single pre-specified estimator of the conditional censoring
@@
   \begin{split}
-    Q(\diff t,j,\diff x)&= \left\{S(t- \mid x)\Lambda_1(\diff t \mid x)H(\diff x)\right\}^{1_{\{j=1\}}}\\
-                        &\quad\left\{S(t- \mid x)\Lambda_2(\diff t \mid x)H(\diff x)\right\}^{1_{\{j=2\}}},
+    Q(\diff t,j,\diff x)=& \left\{S(t- \mid x)\Lambda_1(\diff t \mid x)H(\diff x)\right\}^{\1{\{j=1\}}}\\
+                         &  \left\{S(t- \mid x)\Lambda_2(\diff t \mid x)H(\diff x)\right\}^{\1{\{j=2\}}},
   \end{split}
@@
 survival function of the conditional censoring distribution. The
-distribution is characterised by
+distribution \( P \) is characterised by
 \begin{equation}\label{eq:parametrizeP}
   \begin{split}
-    P(\diff t, j, \diff x) &= \left\{G(t- \mid x)S(t- \mid x)\Lambda_1(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=1\}}}}\\
-                           &\quad\left\{G(t- \mid x)S(t- \mid x)\Lambda_2(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=2\}}}}\\
-                           &\quad\left\{G(t- \mid x)S(t- \mid x)\Gamma(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=0\}}}}\\
-                           &=\left\{G(t- \mid x)Q(\diff t,j,\diff x)\right\}^{\1{{\{j\ne 0\}}}}\\    
-                           &\quad\left\{G(t- \mid x)S(t- \mid x)\Gamma(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=0\}}}}.
+    P(\diff t, j, \diff x) =& \left\{G(t- \mid x)S(t- \mid x)\Lambda_1(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=1\}}}}\\
+                            & \left\{G(t- \mid x)S(t- \mid x)\Lambda_2(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=2\}}}}\\
+                            & \left\{G(t- \mid x)S(t- \mid x)\Gamma(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=0\}}}}\\
+    = & \left\{G(t- \mid x)Q(\diff t,j,\diff x)\right\}^{\1{{\{j\ne 0\}}}}\\    
+                            & \left\{G(t- \mid x)S(t- \mid x)\Gamma(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=0\}}}}.
   \end{split}
@@
 \end{equation*}
-At time \(t\), we observe that each individual is in one of four
-mutually exclusive states (Figure \ref{fig:multi-state-process}).
+At time \(t\), we observe that each individual is in one of four mutually
+exclusive states (Figure \ref{fig:multi-state-process}).
+
 \begin{figure}[h]
   \centering %
-  \includeFigCond[width=.3\textwidth]{figure-multi-state-process.pdf}
+  \includeFigCond[width=.5\textwidth]{figure-multi-state-process.pdf}
   \caption{Illustration of the multi-state process \(\eta\) used by
@@
 \end{figure}
-The conditional distribution of \( \eta(t) \) given \( X \) is determined by
-the function
+
+The conditional
+distribution of \( \eta(t) \) given \( X \) is determined by the function
 \begin{equation}
@@
 at least one learner that is consistent for estimation of \( F \). Recall that
-the function \(F\) implicitly depends on the data generating probability measure
+the function \(F\) implicitly depends on the data-generating probability measure
 \(P\in\mathcal P\) but that this was suppressed in the notation. We now make
@@
 
-In both parts we use the same data generating mechanism. We generate
+In both parts we use the same data-generating mechanism. We generate
 data according to a distribution motivated from a real dataset in

### 2024-05-14 figures/figure-multi-state-process.R

@@ -3,9 +3,9 @@
 ## Author: Thomas Alexander Gerds
 ## Created: Jul 27 2023 (10:04) 
 ## Version: 
-## Last-Updated: Apr 29 2024 (10:30) 
+## Last-Updated: May 14 2024 (12:53) 
 ##           By: Anders Munch
-##     Update #: 8
+##     Update #: 9
 #----------------------------------------------------------------------
 ## 
 ### Commentary: 
@@ -21,6 +21,11 @@ stateLabels = c("Initial","Cause 1", "Cause 2", "Censored")
 crHist <- Hist(time = 1:nTrans, event = list(from = rep("1", nTrans), to = stateLabels[-1]))
 try(setwd("~/research/SuperVision/Anders/survival-loss/statelearner/figures"))
 try(setwd(here("figures")))
+
+plot(crHist,stateLabels = stateLabels,arrowLabels = FALSE,
+     tagBoxes = c(0,1,2,-1),
+     box.width) ## Make true and/or change to -1?
+
 pdf("figure-multi-state-process.pdf")
 plot(crHist,stateLabels = stateLabels,arrowLabels = FALSE,
      tagBoxes = FALSE) ## Make true and/or change to -1?

### 2024-05-14 tex/jrss-version/state-learner-jrss-style.tex
better naming of figure files

@@
                                       - \phi_{a_1,a_2, b}(\data_n)(t,2,x)
-  \\
-  & \qquad
     -\phi_{a_1,a_2, b}(\data_n)(t,-1,x)
@@
       \E_{P_0}{\left[ \Vert \tilde{\phi}_n(\data_n^{-k}) - F_0 \Vert_{P_0}^2 \right]}
-    \\
-    & \quad
+    % \\
+    % & \quad
       + (1+ \delta) 16   K \tau
@@
 following asymptotic expansion of the one-step estimator
-\citep{pfanzagl1985contributions,van2003unified,fisher2021visually,kennedy2022semiparametric,rytgaard2022continuous},
+\citep{pfanzagl1985contributions,van2003unified,fisher2021visually,kennedy2022semiparametric},
 \begin{equation*}
@@
 We compare the state learner to two IPCW based super learners: The
-first superlearner, called IPCW(Cox), uses a Cox model with main
+first super learner, called IPCW(Cox), uses a Cox model with main
 effects to estimate the censoring probabilities, while the second
@@
 horizon (36 months). The marginal risk of the event before this time horizon is
-\(\sim 24.6\)\%. Under the ``dependent censoring'' setting the marginal censoring
-probability before the time horizon is \(\sim 61.9\)\%. Under the ``independent
-censoring'' setting the marginal censoring probability before this time horizon
-is \(\sim 38.7\)\%.
+\(\approx 24.6\)\%. Under the ``dependent censoring'' setting the marginal
+censoring probability before the time horizon is \(\approx 61.9\)\%. Under the
+``independent censoring'' setting the marginal censoring probability before this
+time horizon is \( \approx 38.7 \)\%.
 
@@
 We then use the highest ranked combination
-of learners and apply formula \ref{eq:one-step-comp-ate}.
+of learners and apply formula~\eqref{eq:one-step-comp-ate}.
 
@@
   \centering %
-  \includeFigCond[width=.9\linewidth]{zelefski-real-data.pdf}
+  \includeFigCond[width=.9\linewidth]{real-data-state-learner.pdf}
   \caption[]{The results of applying the 125 combinations of learners to the
@@
   \centering
-  \input{./zel-tab.tex}
+  \input{./real-data-tab.tex}
 \end{table}
@@
   \centering%
-  \includeFigCond[width=.8\linewidth]{zelefsky-data-target-par.pdf}
+  \includeFigCond[width=.8\linewidth]{real-data-target.pdf}
   \caption[]{Estimates of the average treatment effect of hormone therapy on the
@@
 
-\begin{proof}[Proof of Proposition~\ref{prop:stric-prop}]
+\begin{proof}[Proof of Proposition~\ref{prop:stric-prop}.]
   The result follows from Lemma~\ref{lemma:norm}.
@@
 
-In the following, let $\Theta$ denote the function space consisting of all
-conditional state occupation probability functions for some measure \( P \) as
-defined in equation~(\ref{eq:F-def}).
-
-\begin{proof}[Proof of Corollary~\ref{cor:oracle-prop}]
+Recall that $\mathcal{H}_{\mathcal{P}}$ denote the function space consisting of
+all conditional state occupation probability functions for some measure
+\( P \in \mathcal{P} \).
+
+\begin{proof}[Proof of Corollary~\ref{cor:oracle-prop}.]
   First note that minimising the loss \( \bar{B}_{\tau} \) is equivalent to
@@
   identical. By Lemma~\ref{lemma:norm}, \( R_0(F) \geq 0 \) for any
-  \( F \in \Theta \), and so using Theorem 2.3 from \citep{vaart2006oracle} with
+  \( F \in \mathcal{H}_{\mathcal{P}} \), and so using Theorem 2.3 from \citep{vaart2006oracle} with
   \( p=1 \), we have that for all \( \delta >0 \),
@@
   & \quad + (1+\delta) \frac{16 K}{n}
-    \log(1 + |\mathcal{F}_n|)\sup_{F \in \Theta}
+    \log(1 + |\mathcal{F}_n|)\sup_{F \in \mathcal{H}_{\mathcal{P}}}
     \left\{
@@
 \end{align*}
-where for each \( F \in \Theta \), \( (M(F), v(F)) \) is some Bernstein pair for
+where for each \( F \in \mathcal{H}_{\mathcal{P}} \), \( (M(F), v(F)) \) is some Bernstein pair for
 the function \(o \mapsto \bar{B}_{\tau,0}(F, o) \). As
 \( \bar{B}_{\tau,0}(F, \blank) \) is uniformly bounded by \( \tau \) for any
-\( F \in \Theta \), it follows from section 8.1 in \citep{vaart2006oracle} that
+\( F \in \mathcal{H}_{\mathcal{P}} \), it follows from section 8.1 in \citep{vaart2006oracle} that
 \( (\tau, 1.5 P_0{[\bar{B}_{\tau,0}(F, \blank)^2]}) \) is a Bernstein pair for
@@
 \begin{equation*}
-  \sup_{F \in \Theta}
+  \sup_{F \in \mathcal{H}_{\mathcal{P}}}
   \left\{
@@
 
-\begin{proof}[Proof of Corollary~\ref{cor:asymp-cons}]
+\begin{proof}[Proof of Corollary~\ref{cor:asymp-cons}.]
   By definition of the oracle and Lemma~\ref{lemma:norm},
@@
 
-\begin{proof}[Proof of Proposition~\ref{prop:dr-structure}]
+\begin{proof}[Proof of Proposition~\ref{prop:dr-structure}.]
   For notational convenience we suppress \( X \) in the following. The final

### 2024-05-16 tex/jrss-version/state-learner-jrss-style.tex
first round of typos
last minute korrektions

@@
 
-\abstract{In survival analysis, prediction models are needed as stand-alone
-  tools and in applications of causal inference to estimate nuisance parameters.
-  The super learner is a machine learning algorithm which combines a library of
-  prediction models into a meta learner based on cross-validated loss. In right
-  censored data, loss needs to be estimated somehow. We introduce the state
-  learner, a new super learner for survival analysis, which simultaneously
-  evaluates libraries of prediction models for the event of interest and the
-  censoring distribution. The state learner can be applied to all types of
-  survival models, works in the presence of competing risks, and does not
-  require a single pre-specified estimator of the conditional censoring
-  distribution. We establish an oracle inequality for the state learner and
-  investigate its performance through numerical experiments. We illustrate how
-  the state learner allows us to estimate causal effects in a competing risks
-  setting without having to pre-specify models for neither the cause-specific
-  hazards of interest nor the censoring distribution.}
+\abstract{In survival analysis, prediction models are needed as
+  stand-alone tools and in applications of causal inference to
+  estimate nuisance parameters.  The super learner is a machine
+  learning algorithm which combines a library of prediction models
+  into a meta learner based on cross-validated loss. In right censored
+  data, the choice of the loss function and the estimation of the
+  expected loss need careful consideration. We introduce the state
+  learner, a new super learner for survival analysis, which
+  simultaneously evaluates libraries of prediction models for the
+  event of interest and the censoring distribution. The state learner
+  can be applied to all types of survival models, works in the
+  presence of competing risks, and does not require a single
+  pre-specified estimator of the conditional censoring
+  distribution. We establish an oracle inequality for the state
+  learner and investigate its performance through numerical
+  experiments. We illustrate the application of the state learner with
+  prostate cancer data, as a stand-alone prediction tool, and, for
+  causal inference, as a way to estimate the nuissance parameter models of
+  a smooth statistical functional.
+  % We illustrate how the state learner allows us to
+  % estimate causal effects in a competing risks setting without having
+  % to pre-specify models for neither the cause-specific hazards of
+  % interest nor the censoring distribution.
+}
 
@@
 these models in mind. One is as a stand-alone tool where the model
-provides information for a freshly diagnosed patient. The other is as
-input to an algorithm which estimates a causal target parameter such
-as an average treatment effect. The models are learned on data from
-previous patients. A super learner is a machine learning algorithm
-that combines a finite set of learners into a meta learner by
-estimating prediction performance in hold-out samples using a
-pre-specified loss function \citep{van2007super}. When the aim is to
-make a prediction model, super learners combine strong learners, such
-as Cox regression models and random survival forests
-\citep{gerds2021medical}.
+provides predictions about the future, such as the predicted risk of
+adverse outcomes for a newly diagnosed patient. The other is as input
+to an algorithm which estimates a causal target parameter such as an
+average treatment effect and depends on estimates of nuisance
+parameters such as the conditional probability of being uncensored.
+An example of the latter is the targeted minimum loss based estimator
+\citep{rytgaard2022targeted}. In both applications, the models have to
+be learned based on available data from previous patients.
+
+A super learner is a machine learning algorithm that combines a finite
+set of learners into a meta learner by estimating prediction
+performance in hold-out samples using a pre-specified loss function
+\citep{van2007super}. When the aim is to make a prediction model,
+super learners combine strong learners, such as Cox regression models
+and random survival forests \citep[][Section 8.4]{gerds2021medical}.
 % For targeted learning of a low dimensional
@@
 While the general idea of combining strong learners based on cross-validation
-data stems from earlier work \citep{wolpert1992stacked,breiman1996stacked}, the
+stems from earlier work \citep{wolpert1992stacked,breiman1996stacked}, the
 name super learner is justified by an oracle inequality
-\citep{van2003unicv,vaart2006oracle}: The super learner is guaranteed to perform
-almost as well as the model which minimises the expected performance, i.e., the
-model we would select if we could evaluate the prediction performance in an
-infinite hold-out sample.
-
-The choice of
-the loss function for super learning in survival analysis deserves a
-closer look. Existing super learner algorithms for right-censored data
-use partial log-likelihood loss or inverse probability of censoring
-weighted loss
+\citep{van2003unicv,vaart2006oracle}.
+% The super learner is guaranteed to perform
+% almost as well as the model which minimises the expected performance, i.e., the
+% model we would select if we could evaluate the prediction performance in an
+% infinite hold-out sample.
+The work presented in this article is motivated by the choice of the
+loss function for super learning in survival analysis. Existing super
+learner algorithms for right-censored data use partial log-likelihood
+loss or inverse probability of censoring weighted loss
 \citep{polley2011-sl-cens,keles2004asymptotically,golmakani2020super,gonzalez2021stacked,westling2021inference}.
-The use of the partial log-likelihood loss, however, restricts the
-class of learners and excludes for example simple Kaplan-Meier based
-learners and also more complex random survival forest algorithms. For
-this reason \cite{golmakani2020super} restrict their learners to Cox
-proportional hazard models. A lesser known fact is that a super
-learner constructed with the negative partial log-likelihood loss
-implicitly depends on the censoring distribution
+Unfortunately, the use of the partial log-likelihood loss restricts
+the class of learners and excludes for example simple Kaplan-Meier
+based learners and also more complex random survival forest
+algorithms. For this reason \cite{golmakani2020super} restrict their
+learners to Cox proportional hazard models. A lesser known fact is
+that a super learner constructed with the negative partial
+log-likelihood loss implicitly depends on the censoring distribution
 \citep{hjort1992inference,whitney2019comment}. Inverse probability of
@@
 between super learning of the censoring distribution and super
-learning of the event time distribution. We pursue a
+learning of the event time distribution. In this article, we pursue a
 different route.
 
-In this article we define the state learner, a new super learner for
-right-censored data, which simultaneously estimates the expected loss of
-learners of the event time distribution and the censoring distribution. The loss
-function which is used to define the state learner is based on the observed
-variables, that is, the minimum and the order of the censoring time and the
-event time.
-
-The state learner can be applied to all types of survival models,
-it can handle competing risks, and it does not require a
-single pre-specified estimator of the conditional censoring
-distribution.  To analyze the theoretical properties of the state
-learner we focus on the discrete super learner which combines the
-library of learners by picking the one that minimises the
-cross-validated loss \citep{van2007super}. In the presence of
-competing risks, our algorithm uses separate libraries to model the
-cumulative hazard functions of the competing risks and the censoring
-distribution. We show that the oracle selector of the state learner is
-consistent if all libraries contain a consistent learner and prove a
-finite sample oracle inequality. A result of the state learner
-algorithm is a model which predicts the probability of an event based
-on covariates in the presence of competing risks
-\citep{gerds2021medical}. The other application is in targeted
-learning where conditional event probabilities occur as
-high-dimensional nuisance parameters which need to be estimated at a
-certain rate \citep{van2011targeted, rytgaard2021estimation,
-  rytgaard2022targeted}. We show for a targeted estimator, which uses
-the state learner to estimate nuisance parameters, that a second order
-product structure holds for the asymptotic bias term.
-
-The article is organised as follows. We introduce our notation and
-framework in Section~\ref{sec:framework}. In
-Section~\ref{sec:super-learning} we define general super learning
-for right-censored data, and in Section~\ref{sec:relat-liter-exist}
-we review existing methods.  Section~\ref{sec:super-learner-simple}
-introduces our novel proposal, the state learner, and
-Section~\ref{sec:theor-results-prop} provides theoretical guarantees
-for the state learner. In Section~\ref{sec:targeted-learning} we
-discuss the use of the state learner in the context of targeted
-learning. We report results of our numerical experiments in
-Section~\ref{sec:numer-exper} and analyze a prostate cancer data set
-in Section~\ref{sec:real-data-appl}.  Section~\ref{sec:discussion}
+We define the state learner, a new super learner for right-censored
+data, which simultaneously estimates the expected loss of learners of
+the event time distribution and the censoring distribution. The loss
+function underlying the state learner operates on the coarsened
+data. In the right censored survival setting the coarsened data
+consist of the minimum and the order of the censoring time and the
+event time as well as the baseline covariates. The state learner can
+include a broad class of survival models as learners, it can handle
+competing risks, and it does not require a single pre-specified
+estimator of the conditional censoring distribution. To analyze the
+theoretical properties of the state learner we focus on the discrete
+super learner which combines the library of learners by picking the
+one that minimises the cross-validated loss \citep{van2007super}. In
+the presence of competing risks, our algorithm uses separate libraries
+of learners for the cumulative hazard functions for each of the
+competing risks and for the censoring distribution. We show that the
+oracle selector of the state learner is consistent if all libraries
+contain a consistent learner and prove a finite sample oracle
+inequality. The result of the state learner algorithm could be a
+medical risk prediction model \citep{gerds2021medical} which predicts
+the probability of an event based on covariates in the presence of
+competing risks. The other application is in targeted learning where
+conditional event probabilities occur as high-dimensional nuisance
+parameters which need to be estimated at a certain rate
+\citep{van2011targeted, rytgaard2021estimation, rytgaard2022targeted}.
+For the asymptotic bias term of targeted estimator, which uses the
+state learner to estimate nuisance parameters, we show that a second
+order product structure holds. We illustrate both applications of the
+state learner with prostate cancer data
+\citep{kattan2000pretreatment}.
+
+\subsection{Existing methods}
+\label{sec:relat-liter-exist}
+
+Machine learning based on right-censored data commonly uses the
+negative partial log-likelihood as a loss function
+\citep[e.g.,][]{li2016regularized,yao2017deep,lee2018deephit,katzman2018deepsurv,gensheimer2019scalable,lee2021boosted,kvamme2021continuous}.
+This loss function is however not perfectly suited for super learning
+because it cannot be applied to prominent survival learners. For
+example, the Kaplan-Meier estimator, the random survival forest, and
+the semi-parametric Cox regression model provide cumulative hazard
+functions that are piece-wise constant in the time argument. Suppose
+these methods were applied to a learning data set and the aim is to
+evaluate the partial log-likelihood loss in an independent test data
+set. Suppose further that the test data set contains event times that
+were not also seen in the learning data set, as it would naturally
+happen when the events are observed in continuous time. The problem is
+that these methods assign zero probability to all the event times in
+the test set which were not also contained in the training data set.
+Hence the partial log-likelihood loss in the test set would be
+infinity (the logarithm of zero is infinity).
+
+When a proportional hazards model is assumed, the baseline hazard
+function can be profiled out of the likelihood \citep{cox1972regression}.
+The cross-validated partial log-likelihood loss  \citep{verweij1993cross} has therefore been
+suggested as a loss function for super learning which however
+restricts the library of learners to include only Cox proportional
+hazards models \citep{golmakani2020super}.
+
+Another approach for super learning with right-censored data is to use an
+inverse probability of censoring weighted (IPCW) loss function
+\citep{graf1999assessment,van2003unicv,molinaro2004tree,keles2004asymptotically,hothorn2006survival,gerds2006consistent,gonzalez2021stacked}.
+An IPCW loss function is attractive because the associated risk does not depend
+on the censoring distribution but describes a feature of the population of
+interested governed by the measure \( Q \in \mathcal{Q} \). Similar results can
+be obtained using censoring unbiased transformations
+\citep{fan1996local,steingrimsson2019censoring} or pseudo-values
+\citep{andersen2003generalised,mogensen2013random,sachs2019ensemble}. All these
+methods rely on an estimator of the censoring distribution, and their drawback
+is that this estimator has to be pre-specified. When the data-generating
+mechanism is complex and not well-understood, pre-specification of the censoring
+distribution is a challenge.
+
+An attempt to avoid a pre-specified censoring model was proposed
+independently by \cite{han2021inverse} and
+\cite{westling2021inference}. In both articles, the authors suggest to
+iterate between learning of \( \Lambda \) and learning of $\Gamma$
+using IPCW loss functions. However, no general theoretical guarantees
+seem to exist for this procedure, and it has not yet been extended to
+the situation with competing risks.
+
+\subsection{Organization of article}
+
+We introduce our notation and framework in
+Section~\ref{sec:framework}. In Section~\ref{sec:super-learning} we
+define general super learning for right-censored data.
+% , and in
+% Section~\ref{sec:relat-liter-exist} we review existing methods.
+Section~\ref{sec:super-learner-simple} introduces the state learner,
+and Section~\ref{sec:theor-results-prop} provides theoretical
+guarantees. In Section~\ref{sec:targeted-learning} we discuss the use
+of the state learner in the context of targeted learning. We report
+results of our numerical experiments in Section~\ref{sec:numer-exper}
+and analyze a prostate cancer data set in
+Section~\ref{sec:real-data-appl}.  Section~\ref{sec:discussion}
 contains a discussion of the merits and limitations of our
@@
 
-In a competing risk framework \citep{andersen2012statistical}, let \( T\) be a
-time to event variable, \(D\in\{1,2\}\) the cause of the event, and
-$X \in \mathcal{X}$ a vector of baseline covariates taking values in a bounded
-subset \( \mathcal{X} \subset \R^p \), \( p\in\N \). Let $\tau< \infty$ be the
-maximal length of follow-up. We use \( \mathcal{Q} \) to denote the collection
-of all probability measures on \( [0,\tau] \times \{1,2\}\times \mathcal{X} \)
-such that \( (T, D, X) \sim Q \) for some unknown \( Q \in \mathcal{Q} \). For
-\(j\in\{1,2\}\), the cause-specific conditional cumulative hazard functions are
-defined by
-\( \Lambda_{j} \colon [0, \tau] \times \mathcal{X} \rightarrow \R_+ \) such that
+In a competing risk framework \citep{andersen2012statistical}, let
+\( T\) be a time to event variable, \(D\in\{1,2\}\) the cause of the
+event, and $X \in \mathcal{X}$ a vector of baseline covariates taking
+values in a bounded subset \( \mathcal{X} \subset \R^p \),
+\( p\in\N \). Let $\tau< \infty$ be the prediction horizon. We use
+\( \mathcal{Q} \) to denote the collection of all probability measures
+on \( [0,\tau] \times \{1,2\}\times \mathcal{X} \) such that
+\( (T, D, X) \sim Q \) for some unknown \( Q \in \mathcal{Q} \). For
+\(j\in\{1,2\}\), the cause-specific conditional cumulative hazard
+functions are defined by
+\( \Lambda_{j} \colon [0, \tau] \times \mathcal{X} \rightarrow \R_+ \)
+such that
 \begin{equation*}
@@
 \end{equation*}
-For ease of presentation we assume throughout that
-\( \Lambda_j(\blank \mid x) \) is continuous for all \( x \) and
-\( j \). This is not a limitation as everything caries over directly
+For ease of presentation we assume throughout that the map
+\( t\mapsto \Lambda_j(t \mid x) \) is continuous for all \( x \) and
+\( j \). This is not a limitation: All arguments carry over directly
 to the general case. We denote by \(S\) the conditional event-free
@@
 
-We consider the right-censored setting in which we observe data
+We consider the right-censored setting in which we observe the coarsened data
 \(O = (\tilde{T},\tilde D, X)\), where $\tilde T = \min(T,C)$ for a
@@
 \citep[c.f.,][]{begun1983information,gill1997coarsening}. We use
-\(\Gamma\in\mathcal M\) to denote the conditional cumulative hazard
+\(\Gamma\in\mathcal{M}_{\tau}\) to denote the conditional cumulative hazard
 function for censoring. For ease of presentation we now also assume
 that \( \Gamma(\blank \mid x) \) is continuous for all \( x \). We let
-\(G(t \mid x)=\exp\left\{-\Gamma(t \mid x)\right\}\) denote the
+\((t,x)\mapsto G(t \mid x)=\exp\left\{-\Gamma(t \mid x)\right\}\) denote the
 survival function of the conditional censoring distribution. The
@@
 Thus, we can consider $\Lambda_j$ and \(\Gamma\) as operators which map from
-\( \mathcal{P} \) to \(\mathcal M\).
+\( \mathcal{P} \) to \(\mathcal M_{\tau}\).
 
@@
 discrete super learner and the oracle learner, we consider estimation of the
-function-valued parameter \(\Psi:\mathcal P\to\mathcal M\), given by
+function-valued parameter \(\Psi:\mathcal P\to\mathcal M_{\tau}\), given by
 \(\Psi(P)=\Lambda_j\). This parameter is identified via equation
@@
 As input to the super learner we need a data set
-\( \data_n=\{O_i\}_{i=1}^n \) of i.i.d.\ observations from \( P \in \mathcal{P} \) and a finite collection of candidate
+\( \data_n=\{O_i\}_{i=1}^n \) of i.i.d.\ observations from \( P \in \mathcal{P} \) and a collection of candidate
 learners $\mathcal{A}$. Each learner \(a \in \mathcal{A}\) is a map
@@
 The oracle learner is defined as the learner that minimises the
-average loss according to the data-generating distribution \( P \),
+expected loss under the data-generating distribution \( P \),
 i.e.,
@@
 
-\section{Existing methods}
-\label{sec:relat-liter-exist}
-
-Machine learning based on right-censored data commonly use the
-negative partial log-likelihood as loss function
-\citep[e.g.,][]{li2016regularized,yao2017deep,lee2018deephit,katzman2018deepsurv,gensheimer2019scalable,lee2021boosted,kvamme2021continuous}.
-This loss function is however not perfectly suited for super
-learning. The reason is that it cannot be applied to important
-survival learners: The Kaplan-Meier estimator, the random survival
-forest, and the semi-parametric Cox regression model provide
-cumulative hazard functions that are piece-wise constant in the time
-argument. Suppose these methods are applied to a learning data set and
-the aim is to evaluate the partial log-likelihood loss in an
-independent test data set. Suppose further that the test data set
-contains event times that were not also seen in the learning data set,
-as it would naturally happen when the events are observed in
-continuous time. The problem is that these methods assign zero
-probability to the test set event times that were not also in the
-training data set, and hence (the logarithm of zero is infinity) the
-partial log-likelihood loss in the test set would be infinity.
-
-When a proportional hazards model is assumed, the baseline hazard
-function can be profiled out of the likelihood \citep{cox1972regression}.
-The cross-validated partial log-likelihood loss  \citep{verweij1993cross} has therefore been
-suggested as a loss function for super learning which however
-restricts the library of learners to include only Cox proportional
-hazards models \citep{golmakani2020super}.
-
-Another approach for super learning with right-censored data is to use an
-inverse probability of censoring weighted (IPCW) loss function
-\citep{graf1999assessment,van2003unicv,molinaro2004tree,keles2004asymptotically,hothorn2006survival,gerds2006consistent,gonzalez2021stacked}.
-An IPCW loss function is attractive because the associated risk does not depend
-on the censoring distribution but describes a feature of the population of
-interested governed by the measure \( Q \in \mathcal{Q} \). Similar results can
-be obtained using censoring unbiased transformations
-\citep{fan1996local,steingrimsson2019censoring} or pseudo-values
-\citep{andersen2003generalised,mogensen2013random,sachs2019ensemble}. All these
-methods rely on an estimator of the censoring distribution, and their drawback
-is that this estimator has to be pre-specified. When the data-generating
-mechanism is complex and not well-understood, pre-specification of the censoring
-distribution is a challenge.
-
-An attempt to avoid a pre-specified censoring model was proposed
-independently by \cite{han2021inverse} and
-\cite{westling2021inference}. In both articles, the authors suggest to
-iterate between learning of \( \Lambda \) and learning of $\Gamma$
-using IPCW loss functions. However, no general theoretical guarantees
-seem to exist for this procedure, and it has not yet been extended to
-the situation with competing risks.
-
 \section{The state learner}
@@
 % they depend on a pre-specified estimator of the censoring distribution.
-The main idea of the state learner is to jointly use learners of
-\( \Lambda_1 \), \( \Lambda_2 \), and \( \Gamma \), and the relations in
-equation~(\ref{eq:parametrizeP}), to learn a feature of the observed data
-distribution \( P \). The discrete state learner ranks a tuple of learners of
-\( (\Lambda_1, \Lambda_2, \Gamma) \) based on how well they jointly model the
-observed
-data.
-Risk predictions can then be obtained from
-\( \Lambda_1 \) and $\Lambda_2$ \citep{benichou1990estimates, ozenne2017riskregression}. To formally
-introduce the state learner, we define the multi-state process
+The main idea of the state learner is to jointly use learners for
+\( \Lambda_1 \), \( \Lambda_2 \), and \( \Gamma \), and the relations
+in equation~(\ref{eq:parametrizeP}), to learn a feature of the
+observed data distribution \( P \). The discrete state learner ranks a
+tuple of learners for the tuple of the cumulative hazard functions
+\( (\Lambda_1, \Lambda_2, \Gamma) \) based on how well they jointly
+model the observed data.  Risk predictions can then be obtained by
+combining \( \Lambda_1 \) and $\Lambda_2$ from the highest ranked
+tupel using a well-known formula \citep{benichou1990estimates,
+  ozenne2017riskregression}. To formally introduce the state learner,
+we define the multi-state process
 \begin{equation*}
@@
 
-The conditional
-distribution of \( \eta(t) \) given \( X \) is determined by the function
+The conditional distribution of the process \( \eta(t) \) given
+baseline covariates \( X \) is determined by the function
 \begin{equation}
@@
 \(P\in\mathcal P\), c.f.\ equation~\eqref{eq:parametrizeP}, which in turn
-determines \( (F, H) \),
-and so a learner for \( F \) can be constructed from learners of
+determines \( (F, H) \).
+Hence, a learner for \( F \) can be constructed from learners for
 \( \Lambda_1 \), \( \Lambda_2 \), and $\Gamma$ as follows:
@@
 \end{equation}
-The state learner requires three libraries of learners,
+The state learner requires three libraries of learners (c.f., Section \ref{sec:super-learning}),
 \(\mathcal{A}_1\), \( \mathcal{A}_2 \), and \( \mathcal{B} \), where
-\(\mathcal{A}_1\) and \( \mathcal{A}_2\) contain learners of the
+\(\mathcal{A}_1\) and \( \mathcal{A}_2\) contain learners for the
 conditional cause-specific cumulative hazard functions \(\Lambda_1\)
 and \( \Lambda_2\), respectively, and \(\mathcal{B}\) contains
-learners of the conditional cumulative hazard function of the
+learners for the conditional cumulative hazard function of the
 censoring distribution. % We further
@@
 
-We are interested in a parameter
-\( \theta \colon \mathcal{Q} \rightarrow \Theta \) that expresses a
-property of the uncensored population governed by the measure
-\( Q \in \mathcal{Q} \). The parameter space $\Theta$ can be a subset
-of \(\R^d\) or a subset of a function space, for example a subset of
-\(\mathcal{M}_{\tau}\). In subsection~\ref{sec:cause-spec-aver} we
+In this section, we consider a suitably smooth operator
+\( \theta \colon \mathcal{Q} \rightarrow \Theta \) which represents a
+target parameter of interest. The parameter space $\Theta$ can be a
+subset of \(\R^d\) or a subset of a function space, for example a
+subset of \(\mathcal{M}_{\tau}\) as in Section
+\ref{sec:super-learning}. In subsection~\ref{sec:cause-spec-aver} we
 discuss an example from causal inference where $\theta$ is the average
-treatment effect and \( \Theta = [-1,1] \).
-To discuss the role of the state learner for targeted learning we briefly review
-some results from semiparametric efficiency theory. Extensive reviews and
+treatment effect and \( \Theta = [-1,1] \).  To discuss the role of
+the state learner for targeted learning we briefly review some results
+from semiparametric efficiency theory. Extensive reviews and
 introductions are available elsewhere
 \cite[e.g.,][]{pfanzagl1985contributions,bickel1993efficient,van2003unified,tsiatis2007semiparametric,kennedy2016semiparametric}.
-Under the assumption of conditional independent censoring and positivity,
-$\theta$ is identifiable from \( \mathcal{P} \) which means that there exists an
-operator \( \Psi \colon \mathcal{P} \rightarrow \Theta \) such that
-\( \theta(Q) = \Psi(P_{Q, \Gamma}) \) for all $\Gamma \in \mathcal{M}_{\tau}$.
-By equation~(\ref{eq:parametrizeP}) this implies that we may write
+Under the assumption of conditional independent censoring and
+positivity, $\theta$ is identifiable from \( \mathcal{P} \) which
+means that there exists an operator
+\( \Psi \colon \mathcal{P} \rightarrow \Theta \) such that
+\( \theta(Q) = \Psi(P_{Q, \Gamma}) \) for all
+$\Gamma \in \mathcal{M}_{\tau}$.  By equation~(\ref{eq:parametrizeP})
+this implies that we may write
 \begin{equation*}
@@
 for some operator \( \tilde{\Psi}^0 \). The state learner provides a
-ranking of tuples
+ranking of all tuples
 \( (a_1, a_2, b) \in \mathcal{A}_1 \times \mathcal{A}_2 \times
-\mathcal{B} \), and we use \( \hat{a}_{1,n} \), \( \hat{a}_{2,n} \),
-and \( \hat{b}_n \) to denote the learners corresponding to the
-discrete state learner \( \hat{\phi}_n \). Letting \( H(\data_n) \)
-denote the empirical measure of \( \{X_1, \dots, X_n\} \), we obtain a
-simple plug-in estimator of $\theta$:
+\mathcal{B} \). We use \( \hat{a}_{1,n} \), \( \hat{a}_{2,n} \), and
+\( \hat{b}_n \) to denote the learners corresponding to the discrete
+state learner \( \hat{\phi}_n \), i.e., the tupel with the highest
+rank. Letting \( H_n(\data_n) \) denote the empirical measure of
+\( \{X_1, \dots, X_n\} \), we obtain a plug-in estimator of $\theta$:
 \begin{equation}
@@
   \hat{\Psi}^0(\data_n) =
-  \tilde{\Psi}^0(\hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n), H(\data_n)). 
+  \tilde{\Psi}^0(\hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n), H_n(\data_n)). 
 \end{equation}
 
-The asymptotic distribution of \( \hat{\Psi}_{\tau, j}^0 \) is
-difficult to analyze due to the cross-validated model selection
+The asymptotic distribution of \( \hat{\Psi}^0 \) is
+difficult to analyze due to the cross-validated model selection step
 involved in the estimation of the nuisance parameters $\Lambda_1$ and
-$\Lambda_2$. In addition, the estimator will typically have an
-asymptotic bias that vanishes at a too slow 
-rate. Using tools from semi-parametric efficiency theory, it is
-possible to construct a so-called targeted or debiased estimator with
-smaller asymptotic bias and a asymptotic distribution which we know
-how to estimate
+$\Lambda_2$.
+% In addition, the estimator will typically have an
+% asymptotic bias that vanishes at a too slow 
+% rate.
+Using tools from semi-parametric efficiency theory, it is possible to
+construct a so-called targeted or debiased estimator with an
+asymptotic distribution which we know how to estimate
 \citep{bickel1993efficient,van2011targeted,chernozhukov2018double}. A
 targeted estimator is based on the efficient influence function for
-the parameter $\tilde{\Psi}^0$ and relies on an estimator of $\Gamma$
-in addition to estimators of \( \Lambda_1 \) and $\Lambda_2$. The
+the parameter $\tilde{\Psi}^0$ and relies on estimators of the
+nuisance parameters $\Gamma$, \( \Lambda_1 \) and $\Lambda_2$. The
 efficient influence function is a \( P \)-zero mean and square
-integrable function indexed by the nuisance parameters
-\( (\Lambda_1, \Lambda_2, \Gamma) \), which we denote by
+integrable function which we denote by
 \( \psi(\blank ; \Lambda_1, \Lambda_2, \Gamma) \). The name is
@@
 meaning that it has smallest asymptotic variance among all regular
-asymptotically linear estimators.
+asymptotically linear estimators \citep{bickel1993efficient}.
 
@@
     \tilde{\Psi}^0(\hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
-    H(\data_n))
+    H_n(\data_n))
     + \empmeas{[\psi(\blank; \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
@@
 distribution and
-\( \hat{\sigma}(\data_n)^2 = \empmeas{ \left[ \psi(\blank;
+\begin{equation*}
+  \hat{\sigma}(\data_n)^2 = \empmeas{ \left[ \psi(\blank;
       \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n), \hat{b}_n(\data_n))^2
-  \right]} \).
+    \right]}.
+\end{equation*}
 
@@
 
-In the following we demonstrate how the state learner and the general
-estimation strategy outlined above can be used to construct an
-estimator of the cause-specific average treatment effect. We assume
-that the covariate vector \( X \in \R^p \) contains a binary treatment
-indicator \( A \) and a vector of potential confounders,
-\( W \in \mathcal{W}\subset \R^{p-1} \).
-We use $\mu$
-to denote the marginal distribution of \( W \) and $\pi$ to denote the
-conditional probability of treatment,
+In the following we detail how the state learner can be used to
+construct a targeted estimator of the cause-specific average treatment
+effect \citep{rytgaard2022targeted}. We assume that the covariate
+vector \( X \in \R^p \) contains a binary treatment indicator \( A \)
+and a vector of potential confounders,
+\( W \in \mathcal{W}\subset \R^{p-1} \).  We use $\mu$ to denote the
+marginal distribution of \( W \) and $\pi$ to denote the conditional
+probability of treatment,
 \begin{equation*}
@@
 \end{equation*}
-where \( (T^a, D^a) \), \( a \in \{0,1\} \), denotes potential outcomes
+where \( (T^a, D^a) \), \( a \in \{0,1\} \), denote potential outcomes
 \citep{hernanRobinsWhatIf}. In this case, the interpretation of $\theta_{\tau}$
@@
 which censoring depends on the baseline covariates. We simulate data
-based on a prostate cancer study described in
-\citep{kattan2000pretreatment}. The outcome of interest was the time
-to tumor recurrence, and five baseline covariates were used to predict
+based on the prostate cancer study of
+\cite{kattan2000pretreatment}. The outcome of interest is the time to
+tumor recurrence, and five baseline covariates are used to predict
 outcome: prostate-specific antigen (PSA, ng/mL), Gleason score sum
@@
 patient entered the study \citep{gerds2013estimating}. This in turn
-implied that the time of censoring depended on the radiation dose. The
+implies that the time of censoring depends on the radiation dose. The
 data were re-analyzed in \citep{gerds2013estimating} where a
-sensitivity analysis was conducted based on simulated data. We use the
+sensitivity analysis was conducted based on simulated data. Here we use the
 same simulation setup, where event and censoring times are generated
@@
 
-For all super learners we use a library consisting of three learners: The
-Kaplan-Meier estimator \citep{kaplan1958nonparametric,Gerds_2019prodlim}, a Cox
-model with main effects \citep{cox1972regression, survival-package}, and a
-random survival forest \citep{ishwaran2008random,rfsrc-paclage}. We use the same
-library to learn the outcome and the censoring mechanism. To be precise, note
-that given a learner for the cumulative hazard function of the outcome event, we
-can typically use the same method to construct a learner of the cumulative
-hazard function of the censoring distribution. This works by training the
-learner on the data set \( \data_n^c \), where
+% ANDERS: hedder det ikke Nelson-Aalen i stedet for Kaplan-Meier?
+%         kan du skrive klart hvad outcome er i denne foerste del? er det combined endpoint som man kalder for tumor-free survival?  
+For all super learners we use a library consisting of three learners:
+The Kaplan-Meier estimator
+\citep{kaplan1958nonparametric,Gerds_2019prodlim}, a Cox model with
+main effects \citep{cox1972regression, survival-package}, and a random
+survival forest \citep{ishwaran2008random,rfsrc-paclage}. We use the
+same library to learn the outcome distribution and the censoring
+distribution. Note that the three learners in our library of learners
+can be used to learn the cumulative hazard functions of the 
+outcome and the censoring distribution. This works by
+training the learner on the data set \( \data_n^c \), where
 \( \data_n^c = \{O_i^c\}_{i=1}^n \) with
-\( O_i^c = (\tilde{T}_i, 1-\Delta, X_i) \). When we say that we use a learner
-for the cumulative hazard function of the outcome to learn the cumulative hazard
-function of the censoring time, we mean that the learner is trained on
-\( \data_n^c \).
+\( O_i^c = (\tilde{T}_i, 1-\Delta, X_i) \). When we say that we use a
+learner for the cumulative hazard function of the outcome to learn the
+cumulative hazard function of the censoring time, we mean that the
+learner is trained on \( \data_n^c \).
 
@@
 
-Each super learner provides a learner for the cumulative hazard function for the
-outcome of interest. From the cumulative hazard function a risk prediction model
-can be obtained (c.f., equation~(\ref{eq:surv-def}) with $\Lambda_2 = 0$). We
-measure the performance of each super learner by calculating the index of
-prediction accuracy (IPA) \citep{kattan2018index} at a fixed time horizon (36
-months) for the risk prediction model provided by the super learner. The IPA is
-1 minus the ratio between the model's Brier score and the null model's Brier
-score, where the null model is the model that does not use any covariate
-information. The IPA is approximated using a large (\( n = 20,000 \))
-independent data set of uncensored data. As a benchmark we calculate the
-performance of the risk prediction model chosen by the oracle selector, which
-uses the large data set of uncensored event times to select the learner with the
-highest IPA.
+Each super learner provides a learner for the cumulative hazard
+function for the outcome of interest. From the cumulative hazard
+function a risk prediction model can be obtained (c.f.,
+equation~(\ref{eq:surv-def}) with $\Lambda_2 = 0$). We measure the
+performance of each super learner by calculating the index of
+prediction accuracy (IPA) \citep{kattan2018index} at a fixed time
+horizon (36 months) for the risk prediction model provided by the
+super learner. The IPA is 1 minus the ratio between the model's Brier
+score and the null model's Brier score, where the null model is the
+model that does not use any covariate information. The IPA is
+approximated using a large (\( n = 20,000 \)) independent data set of
+uncensored data. As a benchmark we calculate the performance of the
+risk prediction model chosen by the oracle selector, which uses the
+large data set of uncensored event times to select the learner with
+the highest IPA.
 
@@
 
-We use the learners of the three cumulative hazard functions
-selected by the state learner and the estimator defined in
-Section~\ref{sec:cause-spec-aver}, equation~(\ref{eq:one-step-comp-ate}), to
-estimate the average treatment effect of hormone therapy on risk of tumor
-recurrence and death. The propensity score is estimated with a lasso model that
-includes all levels of interaction. The results are shown in
-Figure~\ref{fig:zelefski-real-target} for 6 month intervals after baseline with
-point wise 95\% confidence intervals. We see that hormone therapy appears to
-decrease the risk of tumor recurrence and increase the risk of death without
-tumor recurrence, but that none of the estimated effects are statistically
-significant from zero.
+We use the learners of the three cumulative hazard functions selected
+by the state learner and the estimator defined in
+Section~\ref{sec:cause-spec-aver},
+equation~(\ref{eq:one-step-comp-ate}), to estimate the average
+treatment effect of hormone therapy on risk of tumor recurrence and
+death. The propensity score is estimated with a lasso model that
+includes all levels of interaction.
+% ANDERS: er du sikker paa at det er ALL levels of interaction, ogsaa higher order?
+%         der mangler en reference for lasso model
+The results are shown in Figure~\ref{fig:zelefski-real-target} for 6
+month intervals after baseline with point wise 95\% confidence
+intervals. We see that hormone therapy decreases the risk of tumor
+recurrence and increases the risk of death without tumor recurrence,
+but that none of the estimated effects are statistically significant.
 
@@
 
-We have proposed a new super learner that can be used with
+The state learner is a new super learner that can be used with
 right-censored data and competing events. Compared to existing
@@
 censoring distribution. Furthermore, the state learner neither
-requires that \( \Lambda_j \) can be written as an integral with
-respect to Lebesgue measure, nor assumes a (semi-)parametric form for
-$\Lambda_j$. In the remainder of this section we point to some
-limitations of our proposal and discuss avenues for further research.
+requires that the cause-specific cumulative hazard functions
+\( \Lambda_j \) can be written as integrals with respect to Lebesgue
+measure, nor does it assume a (semi-)parametric formula. In the
+remainder of this section we discuss the limitations of our proposal
+and avenues for further research.
 
 A major advantage of the state learner is that the performance of each
-combination of learners can be estimated without (semi)parametric
-assumptions or additional nuisance parameters. The drawback with this
-approach is that we are rarely interested in features of the observed
-data distribution when the data are right-censored: we are only
-interested in the uncensored event time distribution. The finite
-sample oracle inequality in Corollary~\ref{cor:oracle-prop} concerns
-the function \( F \), which is a feature of \( P \in \mathcal{P} \),
-while what we are typically interested in is \( \Lambda_j \) or
-\( S \), which are features of \( Q \in \mathcal{Q} \). We emphasise
-that while the state learner provides us with estimates of
-\( \Lambda_j \) and $\Gamma$ based on libraries \( \mathcal{A}_j \)
-and \( \mathcal{B} \), the performance of these learners is not
-assessed directly for their respective target parameters, but only
-indirectly via the performance of \( F \).
-For settings without competing risks, our numerical studies suggest
-that measuring the performance of \( F \) also leads to good
-performance for estimation of \( S \).
+combination of learners can be estimated without additional nuisance
+parameters. A potential drawback of our approach is that we are
+evaluating the loss of the learners on the level of the observed data
+distribution while the target of the analysis is either the event time
+distribution, or the censoring distribution, or both.
+% ANDERS: er du enig i min omformulering af forgaaende saetning?
+Specifically, the finite sample oracle inequality in
+Corollary~\ref{cor:oracle-prop} concerns the function \( F \), which
+is a feature of \( P \in \mathcal{P} \), while what we are typically
+interested in is \( \Lambda_j \) or \( S \), which are features of
+\( Q \in \mathcal{Q} \). We emphasise that while the state learner
+provides us with estimates of \( \Lambda_j \) and $\Gamma$ based on
+libraries \( \mathcal{A}_j \) and \( \mathcal{B} \), the performance
+of these learners is not assessed directly for their respective target
+parameters, but only indirectly via the performance of \( F \).  For
+settings without competing risks, our numerical studies suggest that
+measuring the performance of \( F \) also leads to good performance
+for estimation of \( S \).
+% ANDERS: den forgaaende paragraf gentager det samme 2-3 gang. kan du forkorte en gang?
+
 % In the context of targeted learning, a drawback of the state learner is that the
@@
 is more involved, but we expect that a good approximation can be achieved by
-discretisation. In the future, we intend to investigate the performance of the
-state learner when using a broader library of learners in more comprehensive
-simulation studies.
+discretisation.
 

### 2024-05-17 tex/jrss-version/state-learner-jrss-style.tex
separate figures from article
fixed hanging notation

@@
 on the censoring distribution but describes a feature of the population of
-interested governed by the measure \( Q \in \mathcal{Q} \). Similar results can
-be obtained using censoring unbiased transformations
-\citep{fan1996local,steingrimsson2019censoring} or pseudo-values
+interested. Similar results can be obtained using censoring unbiased
+transformations \citep{fan1996local,steingrimsson2019censoring} or pseudo-values
 \citep{andersen2003generalised,mogensen2013random,sachs2019ensemble}. All these
@@
 
-An attempt to avoid a pre-specified censoring model was proposed
-independently by \cite{han2021inverse} and
-\cite{westling2021inference}. In both articles, the authors suggest to
-iterate between learning of \( \Lambda \) and learning of $\Gamma$
-using IPCW loss functions. However, no general theoretical guarantees
-seem to exist for this procedure, and it has not yet been extended to
+An attempt to avoid a pre-specified censoring model was proposed independently
+by \cite{han2021inverse} and \cite{westling2021inference}. In both articles, the
+authors suggest to iterate between learning of the outcome model and learning of
+the censoring model using IPCW loss functions. However, no general theoretical
+guarantees seem to exist for this procedure, and it has not yet been extended to
 the situation with competing risks.

### 2024-05-17 tex/jrss-version/state-learner-jrss-style.tex
fixed Delta vs tilde H confusion
fixed t vs tau

@@
 The main idea of the state learner is to jointly use learners for
-\( \Lambda_1 \), \( \Lambda_2 \), and \( \Gamma \), and the relations
-in equation~(\ref{eq:parametrizeP}), to learn a feature of the
-observed data distribution \( P \). The discrete state learner ranks a
-tuple of learners for the tuple of the cumulative hazard functions
-\( (\Lambda_1, \Lambda_2, \Gamma) \) based on how well they jointly
-model the observed data.  Risk predictions can then be obtained by
-combining \( \Lambda_1 \) and $\Lambda_2$ from the highest ranked
-tupel using a well-known formula \citep{benichou1990estimates,
-  ozenne2017riskregression}. To formally introduce the state learner,
-we define the multi-state process
+\( \Lambda_1 \), \( \Lambda_2 \), and \( \Gamma \), and the relations in
+equation~(\ref{eq:parametrizeP}), to learn a feature of the observed data
+distribution \( P \). The discrete state learner ranks a tuple of learners for
+the tuple of the cumulative hazard functions
+\( (\Lambda_1, \Lambda_2, \Gamma) \) based on how well they jointly model the
+observed data. Risk predictions can then be obtained by combining
+\( \Lambda_1 \) and $\Lambda_2$ from the highest ranked tuple using a well-known
+formula \citep{benichou1990estimates, ozenne2017riskregression}. To formally
+introduce the state learner, we define the multi-state process
 \begin{equation*}
@@
   F(t, 1, x)
-  & = P(\tilde{T} \leq t, \Delta=1 \mid X=x)
+  & = P(\tilde{T} \leq t, \tilde{D}=1 \mid X=x)
     = \int_0^t e^{\{-\Lambda_{1}(s \mid x)-\Lambda_{2}(s \mid x) - \Gamma(s \mid x)\} }  \Lambda_{1}(\diff s \mid x),
@@
   F(t, 2, x)
-  & = P(\tilde{T} \leq t, \Delta=2 \mid X=x)
+  & = P(\tilde{T} \leq t, \tilde{D}=2 \mid X=x)
     = \int_0^t e^{\{-\Lambda_{1}(s \mid x)-\Lambda_{2}(s \mid x) - \Gamma(s \mid x)\} }  \Lambda_{2}(\diff s \mid x),
@@
   & =
-    P(\tilde{T} \leq t, \Delta=0 \mid X=x)
+    P(\tilde{T} \leq t, \tilde{D}=0 \mid X=x)
     = \int_0^t e^{\{-\Lambda_{1}(s \mid x)-\Lambda_{2}(s \mid x) - \Gamma(s \mid x)\} }  \Gamma(\diff s \mid x),
@@
 \end{equation*}
-for some operator \( \tilde{\Psi}^0 \). The state learner provides a
-ranking of all tuples
-\( (a_1, a_2, b) \in \mathcal{A}_1 \times \mathcal{A}_2 \times
-\mathcal{B} \). We use \( \hat{a}_{1,n} \), \( \hat{a}_{2,n} \), and
-\( \hat{b}_n \) to denote the learners corresponding to the discrete
-state learner \( \hat{\phi}_n \), i.e., the tupel with the highest
-rank. Letting \( H_n(\data_n) \) denote the empirical measure of
-\( \{X_1, \dots, X_n\} \), we obtain a plug-in estimator of $\theta$:
+for some operator \( \tilde{\Psi}^0 \). The state learner provides a ranking of
+all tuples
+\( (a_1, a_2, b) \in \mathcal{A}_1 \times \mathcal{A}_2 \times \mathcal{B} \).
+We use \( \hat{a}_{1,n} \), \( \hat{a}_{2,n} \), and \( \hat{b}_n \) to denote
+the learners corresponding to the discrete state learner \( \hat{\phi}_n \),
+i.e., the tuple with the highest rank. Letting \( H_n(\data_n) \) denote the
+empirical measure of \( \{X_1, \dots, X_n\} \), we obtain a plug-in estimator of
+$\theta$:
 \begin{equation}
@@
 Using equation~(\ref{eq:surv-def}) we may write
-\( \theta_{{\tau}}(Q) = \tilde{\Psi}_{t}^0(\Lambda_1, \Lambda_2, \mu) \), where
+\( \theta_{{\tau}}(Q) = \tilde{\Psi}_{\tau}^0(\Lambda_1, \Lambda_2, \mu) \), where
 \begin{equation}
@@
   \begin{split}
-  \tilde{\Psi}_{t}^0(\Lambda_1, \Lambda_2, \mu) & =
+  \tilde{\Psi}_{\tau}^0(\Lambda_1, \Lambda_2, \mu) & =
   \int_{\mathcal{W}} 
@@
   - \psi_{\tau}^0(O; \Lambda_1, \Lambda_2, \Gamma, \pi)
-  -\tilde{\Psi}_{t}^0(\Lambda_1, \Lambda_2, \mu),
+  -\tilde{\Psi}_{\tau}^0(\Lambda_1, \Lambda_2, \mu),
 \end{equation*}
@@
   \begin{split}
-    \hat{\Psi}_{t,\text{OS}}(\data_n)
+    \hat{\Psi}_{\tau,\text{OS}}(\data_n)
     = &
-        \tilde{\Psi}_{t}^0(\hat{a}_1(\data_n), \hat{a}_2(\data_n),
+        \tilde{\Psi}_{\tau}^0(\hat{a}_1(\data_n), \hat{a}_2(\data_n),
         \hat{\mu}_n(\data_n))
@@
 \end{lemma}
-\begin{proof}
+\begin{proof}[Proof]
   For any \( t \in [0, \tau] \) and \( j\in \{-1,0,1,2\} \) we have
@@
 
-\begin{proof}[Proof of Proposition~\ref{prop:stric-prop}.]
+\begin{proof}[Proof of Proposition~\ref{prop:stric-prop}]
   The result follows from Lemma~\ref{lemma:norm}.
@@
 
-\begin{proof}[Proof of Corollary~\ref{cor:oracle-prop}.]
+\begin{proof}[Proof of Corollary~\ref{cor:oracle-prop}]
   First note that minimising the loss \( \bar{B}_{\tau} \) is equivalent to
@@
 
-\begin{proof}[Proof of Corollary~\ref{cor:asymp-cons}.]
+\begin{proof}[Proof of Corollary~\ref{cor:asymp-cons}]
   By definition of the oracle and Lemma~\ref{lemma:norm},
@@
 
-\begin{proof}[Proof of Proposition~\ref{prop:dr-structure}.]
+\begin{proof}[Proof of Proposition~\ref{prop:dr-structure}]
   For notational convenience we suppress \( X \) in the following. The final

### 2024-05-17 tex/jrss-version/state-learner-jrss-style.tex
final read through

@@
 In both parts we use the same data-generating mechanism. We generate
-data according to a distribution motivated from a real dataset in
+data according to a distribution motivated from a real data set in
 which censoring depends on the baseline covariates. We simulate data
@@
 The results are shown in Figure~\ref{fig:zelefski-real-target} for 6
-month intervals after baseline with point wise 95\% confidence
+month intervals after baseline with pointwise 95\% confidence
 intervals. We see that hormone therapy decreases the risk of tumor
@@
     analyzed by \cite{kattan2000pretreatment}. The estimates are based on the
-    estimator defined in equation~(\ref{eq:one-step-comp-ate}) with point wise
+    estimator defined in equation~(\ref{eq:one-step-comp-ate}) with pointwise
     95\% confidence intervals calculated as described in

### 2024-05-21 tex/jrss-version/state-learner-jrss-style.tex
item labels
exp fixed
fig and tabs in main text again

@@
 
-\usepackage[tablesonly,heads,nolists]{endfloat}
+% \usepackage[tablesonly,heads,nolists]{endfloat}
+
+%% change enumerate labels
+\renewcommand{\labelenumi}{\normalfont (\alph{enumi})}
 
@@
 
-\title[The state learner]{The state learner -- a super learner for
-  right-censored data}
+\title[The state learner]{The state learner -- a super learner for right-censored data}
 
@@
   prostate cancer data, as a stand-alone prediction tool, and, for
-  causal inference, as a way to estimate the nuissance parameter models of
+  causal inference, as a way to estimate the nuisance parameter models of
   a smooth statistical functional.
@@
   & = P(\tilde{T} \leq t, \tilde{D}=1 \mid X=x)
-    = \int_0^t e^{\{-\Lambda_{1}(s \mid x)-\Lambda_{2}(s \mid x) - \Gamma(s \mid x)\} }  \Lambda_{1}(\diff s \mid x),
+    = \int_0^t \exp{{\{-\Lambda_{1}(s \mid x)-\Lambda_{2}(s \mid x) - \Gamma(s \mid x)\} }}  \Lambda_{1}(\diff s \mid x),
   \\
@@
   & = P(\tilde{T} \leq t, \tilde{D}=2 \mid X=x)
-    = \int_0^t e^{\{-\Lambda_{1}(s \mid x)-\Lambda_{2}(s \mid x) - \Gamma(s \mid x)\} }  \Lambda_{2}(\diff s \mid x),
+    = \int_0^t \exp{\{-\Lambda_{1}(s \mid x)-\Lambda_{2}(s \mid x) - \Gamma(s \mid x)\} }  \Lambda_{2}(\diff s \mid x),
   \\
@@
     P(\tilde{T} \leq t, \tilde{D}=0 \mid X=x)
-    = \int_0^t e^{\{-\Lambda_{1}(s \mid x)-\Lambda_{2}(s \mid x) - \Gamma(s \mid x)\} }  \Gamma(\diff s \mid x),
+    = \int_0^t \exp{\{-\Lambda_{1}(s \mid x)-\Lambda_{2}(s \mid x) - \Gamma(s \mid x)\} }  \Gamma(\diff s \mid x),
   \\
@@
     \intertext{where in correspondence with  the relations in equation \eqref{eq:transition},} 
-    \phi_{a_1,a_2, b}(\data_n)(t,1,x) &= \int_0^t e^{\{-a_1(\data_n)(s \mid x)-a_2(\data_n)(s \mid x) - b(\data_n)(s \mid x)\} }  a_1(\data_n)(\diff s \mid x),\\
-  \phi_{a_1,a_2, b}(\data_n)(t,2,x) &= \int_0^t e^{\{-a_1(\data_n)(s \mid x)-a_2(\data_n)(s \mid x) - b(\data_n)(s \mid x)\} }  a_2(\data_n)(\diff s \mid x),\\
-  \phi_{a_1,a_2, b}(\data_n)(t,-1,x) &= \int_0^t e^{\{-a_1(\data_n)(s \mid x)-a_2(\data_n)(s \mid x) - b(\data_n)(s \mid x)\} }  b(\data_n)(\diff s \mid x),\\
+    \phi_{a_1,a_2, b}(\data_n)(t,1,x) &= \int_0^t \exp{\{-a_1(\data_n)(s \mid x)-a_2(\data_n)(s \mid x) - b(\data_n)(s \mid x)\} }  a_1(\data_n)(\diff s \mid x),\\
+  \phi_{a_1,a_2, b}(\data_n)(t,2,x) &= \int_0^t \exp{\{-a_1(\data_n)(s \mid x)-a_2(\data_n)(s \mid x) - b(\data_n)(s \mid x)\} }  a_2(\data_n)(\diff s \mid x),\\
+  \phi_{a_1,a_2, b}(\data_n)(t,-1,x) &= \int_0^t \exp{\{-a_1(\data_n)(s \mid x)-a_2(\data_n)(s \mid x) - b(\data_n)(s \mid x)\} }  b(\data_n)(\diff s \mid x),\\
   \phi_{a_1,a_2, b}(\data_n)(t,0,x) &= 1
@@
   \bigO(n^{-\alpha}) \), for some \( \alpha\leq 1 \).
-  \begin{enumerate}
+  \begin{enumerate}[1.]
   \item If $\alpha=1$ then
@@
   \begin{split}
-  \tilde{\Psi}_{\tau}^0(\Lambda_1, \Lambda_2, \mu) & =
+    \tilde{\Psi}_{\tau}^0(\Lambda_1, \Lambda_2, \mu)
+    & =
       \int_{\mathcal{W}} 
       \int_0^{\tau}
-  e^{-\Lambda_1(s- \mid w, 1)-\Lambda_2(s- \mid w, 1)}  \Lambda_1(\diff s \mid
+      \exp{\{{-\Lambda_1(s- \mid w, 1)-\Lambda_2(s- \mid w, 1)}\}}  \Lambda_1(\diff s \mid
       w, 1)
@@
       \int_0^{\tau}
-  e^{-\Lambda_1(s- \mid w, 0)-\Lambda_2(s- \mid w, 0)}  \Lambda_1(\diff s \mid w, 0)
+      \exp{\{{-\Lambda_1(s- \mid w, 0)-\Lambda_2(s- \mid w, 0)}\}}  \Lambda_1(\diff s \mid w, 0)
       \mu(\diff w).
@@
   & = \int_0^{t}
-    e^{-\Lambda_1(s- \mid W, A)-\Lambda_2(s- \mid W, A)}  \Lambda_1(\diff s \mid
+    \exp{\{ {-\Lambda_1(s- \mid W, A)-\Lambda_2(s- \mid W, A)} \}}  \Lambda_1(\diff s \mid
     W, A),
@@
       \omega_a(A,W; \pi)
-      \int_0^{\tau} e^{\Gamma(t- \mid A, W)}   
+      \int_0^{\tau} \exp{\{{\Gamma(t- \mid A, W)}\}}   
       M_1(\diff t \mid A, W; \Lambda_1)
@@
       \int_0^{\tau}
-      e^{[\Gamma+\Lambda_1 + \Lambda_2](t- \mid A, W)}
+      \exp{\{{[\Gamma+\Lambda_1 + \Lambda_2](t- \mid A, W)}\}}
       M(\diff t \mid A, W; \Lambda_1, \Lambda_2)
@@
       g(t, A, W; \Lambda_1, \Lambda_2)
-      e^{[\Gamma+\Lambda_1 + \Lambda_2](t- \mid A, W)}
+      \exp{\{{[\Gamma+\Lambda_1 + \Lambda_2](t- \mid A, W)}\}}
       M(\diff t \mid A, W; \Lambda_1, \Lambda_2)

### 2024-05-21 tex/draft-version/state-learner-submit-version0.tex
renaming

--- 2023-11-21 tex/state-learner-pub-version.tex
@@
 \input{./latex-settings/title-compact.tex}
+\definecolor{linkcolor}{rgb}{0, 0, 0.54}
 
@@
 
-%% Specific commands
 \newcommand{\data}{\ensuremath{\mathcal{D}}}
 
+\title{The state learner \\ {-- a super learner for right-censored data}}
+
+\author{Anders Munch and Thomas A. Gerds}
+
 \begin{document}
-
-% \articletype{ARTICLE TEMPLATE}% Specify the article type or omit as appropriate
-
-\title{The state learner \\ {-- a super learner for right-censored data}}
-
-\author{Anders Munch and Thomas A. Gerds
-  % \name{Anders Munch\textsuperscript{a}\thanks{CONTACT: Anders Munch. Email: a.munch@sund.ku.dk}
-  %   and Thomas A. Gerds\textsuperscript{a}}
-  % \affil{\textsuperscript{a} Section of Biostatistics, Department of Public Health, University of Copenhagen}
-}
-
 \maketitle
@@
   hazards of interest nor the censoring distribution.
-  % In survival analysis, prediction models are needed as stand-alone tools and in
-  % applications of causal inference to estimate nuisance parameters. The super
-  % learner is a machine learning algorithm which combines a library of prediction
-  % models into a meta learner based on cross-validated loss. Unfortunately, the
-  % commonly used partial likelihood loss is not suited for super learning, and
-  % inverse probability of censoring weighted loss functions require a
-  % pre-specified estimator of the censoring distribution. To relax this, we
-  % introduce the state learner, a new super learner for survival analysis, which
-  % evaluates the loss based on the observed data simultaneously for two libraries
-  % of prediction models, one for the event time distribution and one for the
-  % censoring distribution. We establish an oracle inequality for the state
-  % learner and investigate its performance through numerical experiments.
 \end{abstract}
@@
 
+\section{Existing methods}
+\label{sec:relat-liter-exist}
+
+Machine learning based on right-censored data commonly use the negative partial
+log-likelihood as loss function
+\citep[e.g.,][]{li2016regularized,yao2017deep,lee2018deephit,katzman2018deepsurv,gensheimer2019scalable,lee2021boosted,kvamme2021continuous}.
+However, this loss function is unsuited for super learning, because many
+canonical survival learners (e.g., the Kaplan-Meier estimator, random survival
+forest, and semi-parametric Cox models) provide cumulative hazard functions that
+are piece-wise constant in the time argument, and hence assign zero probability
+to event times not observed in the training data. This implies that when data
+are observed in continuous time, any of these learners will almost surely have
+infinite loss in any independent hold-out sample according to the negative
+partial log-likelihood loss.
+
+When a proportional hazards model is assumed, the baseline hazard function can
+be profiled out of the likelihood to give a new partial log-likelihood loss
+\citep{cox1972regression}, which has been suggested as a loss function for super
+learning \citep{golmakani2020super,verweij1993cross}. While this allows the
+library of learners to include Cox' proportional hazard models, the drawback is
+that the library is now restricted to include only these models.
+
+Another approach for super learning with right-censored data is to use an
+inverse probability of censoring weighted (IPCW) loss function
+\citep{graf1999assessment,van2003unicv,molinaro2004tree,keles2004asymptotically,hothorn2006survival,gerds2006consistent,gonzalez2021stacked}.
+An IPCW loss function is attractive because the associated risk does not depend
+on the censoring distribution but describes a feature of the population of
+interested governed by the measure \( Q \in \mathcal{Q} \). Similar results can
+be obtained using censoring unbiased transformations
+\citep{fan1996local,steingrimsson2019censoring} or pseudo-values
+\citep{andersen2003generalised,mogensen2013random,sachs2019ensemble}. All these
+methods rely on an estimator of the censoring distribution, and their drawback
+is that this estimator has to be pre-specified. When the data-generating
+mechanism is complex and not well-understood, pre-specification of the censoring
+distribution is a challenge.
+
+As far as we are aware, the only existing attempt at avoiding the need to
+pre-specify a censoring model is a recent proposal suggested independently by
+\cite{han2021inverse} and \cite{westling2021inference}. The authors do not
+consider competing risks but suggest to iterate between learning \( \Lambda \)
+and $\Gamma$ using IPCW loss functions and select the final learner when the
+iterative procedure has converged.
+% Using the estimator \( \hat{b}^{(0)}(\data_n^c) \) of $\Gamma$ based on some
+% initial learner \( \hat{b}^{(0)} \in \mathcal{B} \), we calculate the
+% estimated IPCW risk for each leaner in \( \mathcal{A} \). Picking
+% \( a^{(1)} \) as the discrete super learner according to this IPCW loss
+% function, we then estimate weight
+To the best of our knowledege, no general theoretical guarantees exist for this
+procedure. It seems unclear whether this iterative procedure will always be able
+to find a correctly specified model in a given library.
+
+% In a non-survival targeted learning setting \cite{sun2022selective}, based on
+% ideas presented by \cite{robins2007comment}, proposed to use as a model
+% selection criteria the so-called `doubly robustness' property that some
+% targeted estimators enjoy. In many special cases the remainder term
+% \( \mathrm{Rem}{(\hat{\Lambda}_{n}, \hat{\Gamma}_n, P)} \) defined in
+% equation~(\ref{eq:ral}) has the property that
+% \begin{equation}
+%   \label{eq:dr-property}
+%   \mathrm{Rem}{(\Lambda^*, \Gamma_0, P_0)}
+%   = \mathrm{Rem}{(\Lambda_0, \Gamma^*, P_0)} = 0,
+%   \quad \text{for any }
+%   \Lambda^*, \Gamma^* \in \mathcal{M}.
+% \end{equation}
+% The property in equation~(\ref{eq:dr-property}) is known as `doubly
+% robustness' because it ensures that
+% $\tilde{\Psi}(\hat{\Lambda}_{n}, \hat{\Gamma}_n, H_n)$ defined in
+% equation~(\ref{eq:ral}) is consistent if just one of $\Lambda$ or $\Gamma$ is
+% estimated consistently
+% \citep{van2003unified,bang2005doubly,kang2007demystifying}. For
+% $a \in \mathcal{A}$ and \( b \in \mathcal{B} \), define
+% \( \hat{\Psi}^k_n(a, b) = \tilde{\Psi}(a(\data_n^k), b(\data_n^k), \empmeas^k) \)
+% where \( \empmeas^{k} \) is the empirical measure of \( \data_n^{k} \).
+% \cite{sun2022selective} suggest to select $a \in \mathcal{A}$ as the minimizer
+% of the estimated `fluctuation pseudo-risk',
+% \begin{equation*}
+%   \hat R_{\mathrm{f}}(a) =
+%   \max_{b_1, b_2 \in \mathcal{B}}
+%   \frac{1}{K}\sum_{k=1}^{K}
+%   \left\{
+%     \hat{\Psi}^k_n(a, b_1)
+%     - \hat{\Psi}^k_n(a, b_2)
+%   \right\}^2.
+% \end{equation*}
+% The idea is that the doubly robustness property
+% (equation~(\ref{eq:dr-property})) implies that \(\hat{R}_{\mathrm{f}}(a)\) will
+% be zero (asymptotically) if the learner \(a\) correctly estimates $\Lambda$. The
+% authors establish finite sample results guarantying that the selected learner
+% \(a\) will be robust against changing the learner of the other nuisance
+% parameter across $\mathcal{B}$. It seems unclear to what extend this robustness
+% property also guarantees consistency of the estimator of the target parameters
+% when we use the fluctuation risk \(\hat{R}_{\mathrm{f}}\) to select the learners
+% of the nuisance parameters. It would be interesting to explore the performance
+% of this procedure in a survival setting and compare it to targeted estimators
+% obtained using the state learner as outlined in
+% Section~\ref{sec:targeted-learning}.
+
 \section{The state learner}
@@
     \left\{
-      (\Lambda_1-\hat{\Lambda}_{1,n})^2
+      \|\Lambda_1-\hat{\Lambda}_{1,n}\|^2
       +
-      (\Lambda_2-\hat{\Lambda}_{2,n})^2
+      \|\Lambda_2-\hat{\Lambda}_{2,n}\|^2
       +
-      (\Gamma-\hat{\Gamma}_{n})^2
+      \|\Gamma-\hat{\Gamma}_{n}\|^2
     \right\}
-  }.
+  },
 \end{equation}
-Hence, under a suitable Donsker class regularity condition
-\citep{bickel1993efficient,kennedy2016semiparametric}, when
+for some suitable norm \( \|\blank \| \), for instance the
+\( \mathcal{L}_{P}^2 \)-norm. Hence, under a suitable Donsker class regularity
+condition \citep{bickel1993efficient,kennedy2016semiparametric}, when
 equation~(\ref{eq:4}) holds and the nuisance parameters $\Lambda_1$,
@@
 
-\subsection{Cause-specific average treatment effect}
+\subsection{Average treatment effect on the cumulative incidence}
 \label{sec:cause-spec-aver}
 
-We now demonstrate how the state learner and the general estimation strategy
-outlined above can be used to estimate a concrete parameter of interest in the
-competing risk setting. We assume that the covariate vector \( X \in \R^p \) can
-be separated into a binary treatment indicator \( A \) and a vector of potential
-confounders, \( W \in \mathcal{W}\subset \R^{p-1} \). We abuse notation slightly
-by writing \( \Lambda_j(t \mid x) = \Lambda_j(t \mid a, w) \) and
+In the following we demonstrate how the state learner and the general estimation
+strategy outlined above can be used to construct an estimator of the
+cause-specific average treatment effect. We assume that the covariate vector
+\( X \in \R^p \) can be separated into a binary treatment indicator \( A \) and
+a vector of potential confounders, \( W \in \mathcal{W}\subset \R^{p-1} \). We
+abuse notation slightly by writing
+\( \Lambda_j(t \mid x) = \Lambda_j(t \mid a, w) \) and
 \( S(t \mid x) = S(t \mid a, w) \) when \( x=(a, w) \). We use $\mu$ to denote
@@
         - \empmeas{[\psi_{\tau}^0(\blank; \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
-        \hat{b}_n(\data_n), \hat{\pi}_n(\data_n))]}
+        \hat{b}_n(\data_n), \hat{\pi}_n(\data_n))]}.
   \end{split}
@@
 
-%% IPCW with potentially misspecified censoring model.
-%%
-%% Applying SL, concrete, and Westling on classical data set. (Where to put
-%% this?)
-
-We compare the state learner with two other discrete super learners and an
-oracle selector in a simulation study without a competing event. The two other
-super learners are based on inverse probability of censoring weighted (IPCW)
-Brier scores \citep{graf1999assessment,gerds2006consistent}, and we refer to
-these as IPCW super learners. These super learners depend on an estimator of the
-censoring distribution, and we consider IPCW super learners that use either the
-Kaplan-Meier estimator (IPCW(KM)) or a correctly specified Cox model (IPCW(Cox))
-to estimate the censoring distribution. Both IPCW super learners are fitted
-using the \texttt{R}-package \texttt{riskRegression}
-\citep{Gerds_Ohlendorff_Ozenne_2023}. Each discrete super learner provides a
-learner for the cumulative hazard function for the outcome of interest, and from
-this a risk prediction model can be obtained. We measure performance of each
-super learner in terms of the Brier score of the provided risk prediction model
-at a specific time horizon. The Brier score is approximated using a large
-(\( n = 20,000 \)) independent data set of uncensored data. The oracle selector
-uses the large data set of uncensored event times to select the learner with the
-lowest expected Brier score. The expected Brier score of the oracle selector
-serves as a lower benchmark value. For all super learners we split the data into
-five folds for training and testing.
-
-Note that given a learner for the cumulative hazard function of the
-outcome event, we can typically use the same method to construct a
-learner of the cumulative hazard function of the censoring
-distribution. This would typically work by training the learner on the
-data set \( \data_n^c \), where \( \data_n^c = \{O_i^c\}_{i=1}^n \)
-with \( O_i^c = (\tilde{T}_i, 1-\Delta, X_i) \). When we say that we
-use a learner for the cumulative hazard function of the outcome to
-learn the cumulative hazard function of the censoring time, we mean
-that the learner is trained on \( \data_n^c \).
-
-\paragraph{Scenario 1}
-
-We first generate a simple dataset to demonstrate that an IPCW super learner can
-perform poorly when the censoring model is misspecified. We start by generating a
-binomial baseline covariate \( A \) with success probability \( 30\% \). We then
-generate outcome and censoring variables according to a Cox-Weibull distribution
-with hazard rates of approximately \( 0.5 \) and \( 2.5 \), respectively. We use
-a library \( \mathcal{A} \) consisting of two learners, the (marginal)
-Kaplan-Meier estimator and the Kaplan-Meier estimator stratified on the binary
-covariate. For the state learner we use the same learners to construct the
-library \( \mathcal{B} \).
-
-The results are shown in Figure~\ref{fig:ipcw-fail}. We see that the IPCW super
-learner based on a misspecified censoring model (IPCW(KM)) has a larger Brier
-score than the other super learners and that its performance does not improve
-much with sample size. % When the censoring model is misspecified, the
-% IPCW(KM) super learner fails to identify the learner corresponding to the
-% data-generating mechanism, even though this learner is included in the
-% library.
-The performance of the state learner is comparable to or slightly better than
-the IPCW super learner based on a correctly pre-specified censoring model. The
-performance of both the IPCW(Cox) super learner and the state learner are close
-to that of the oracle for most sample sizes.
-
-\begin{figure}
-  \centerline{\includegraphics[width=1\linewidth]{../figures/ipcw-fail.pdf}}
-  \caption[]{Results for scenario~1 of the simulation study. For the learner
-    selected by each of the four discrete super learners, the Brier score
-    calculated in a large independent data set without censoring is plotted
-    against sample size. The results are based on 200 repetitions.}
-  \label{fig:ipcw-fail}
-\end{figure}
-
-\paragraph{Scenario 2}
-
-We next generate data according to a more complex distribution motivated from a
-real dataset in which censoring depends on the baseline covariates. We simulate
-data based on a prostate cancer study described in
+As a proof of concept we conduct two small simulation studies in a setting
+without competing risks. We generate data according to a distribution motivated
+from a real dataset in which censoring depends on the baseline covariates. We
+simulate data based on a prostate cancer study described in
 \citep{kattan2000pretreatment}. The outcome of interest was the time to tumor
@@
 estimated from the original data
-\citep[c.f.,][Section~4.6]{gerds2013estimating}. We use the library consisting
-of the nine learners described in Table~\ref{tab:zel-library}. For the state
-learner we use the same library to learn the censoring distribution.
+\citep[c.f.,][Section~4.6]{gerds2013estimating}. We refer to this simulation
+setting as ``dependent censoring''. We also considered a simulation setting
+where data were generated in the same way, except that censoring was generated
+completely independently. We refer to this simulation setting as ``independent
+censoring''.
+
+For all super learners we use a library consisting of three learners: The
+Kaplan-Meier estimator \citep{kaplan1958nonparametric,Gerds_2019prodlim}, a Cox
+model with main effects \citep{cox1972regression, survival-package}, and a
+random survival forest \citep{ishwaran2008random,rfsrc-paclage}. We use the same
+library to learn the outcome and the censoring mechanism. To be precise, note
+that given a learner for the cumulative hazard function of the outcome event, we
+can typically use the same method to construct a learner of the cumulative
+hazard function of the censoring distribution. This would typically work by
+training the learner on the data set \( \data_n^c \), where
+\( \data_n^c = \{O_i^c\}_{i=1}^n \) with
+\( O_i^c = (\tilde{T}_i, 1-\Delta, X_i) \). When we say that we use a learner
+for the cumulative hazard function of the outcome to learn the cumulative hazard
+function of the censoring time, we mean that the learner is trained on
+\( \data_n^c \).
+
+We first compare the state learner to two IPCW based super learners: The first
+uses a Cox model with main effects to estimate the censoring probabilities
+(IPCW(Cox)), while the second uses the Kaplan-Meier estimator to estimate the
+censoring probabilities (IPCW(KM)). The Cox model for the censoring distribution
+is correctly specified in both simulation settings while the Kaplan Meier
+estimator only estimates the censoring model correctly in the simulation setting
+where censoring is independent. Both IPCW super learners are fitted using the
+\texttt{R}-package \texttt{riskRegression} \citep{Gerds_Ohlendorff_Ozenne_2023}.
+Each super learner provides a learner for the cumulative hazard function for the
+outcome of interest, and from this a risk prediction model can be obtained. We
+measure performance of each super learner by calculating the index of prediction
+accuracy (IPA) \citep{kattan2018index} at a fixed time horizon (36 months) for
+the risk prediction model provided by the super learner. The IPA is 1 minus the
+ratio between the model's Brier score and the null model's Brier score, where
+the null model is the model that does not use any covariate information. The IPA
+is approximated using a large (\( n = 20,000 \)) independent data set of
+uncensored data. As a benchmark we calculate the performance of the risk
+prediction model chosen by the oracle selector. The oracle selector uses the
+large data set of uncensored event times to select the learner with the highest
+IPA.
+
+The results are shown in Figure~\ref{fig:ipcw-fail}. We see that when censoring
+is not independent, using the Kaplan-Meier estimator to estimate the censoring
+probabilities provides a risk prediction model with an IPA that is lower than
+the risk prediction model provided by the state learner. The performance of the
+risk prediction model selected by the state learner is similar to the risk
+prediction model selected by the IPCW(Cox) super learner which a priori uses a
+correctly specified model for the censoring distribution. Both these risk
+prediction models are close to the performance of the oracle, except for small
+sample sizes.
+
+\begin{figure}
+  \centerline{\includegraphics[width=1\linewidth]{../figures/experiment-fig-sl-ipcw.pdf}}
+  \caption[]{For the risk prediction models provided by each of the super
+    learners, the IPA is plotted against sample size. The results are based on
+    1000 repetitions and the error bars are used to quantify the Monte Carlo
+    uncertainty.}
+  \label{fig:ipcw-fail}
+\end{figure}
+
+We next compare the state learner to the super learner \texttt{survSL}
+\citep{westling2021inference}, as this is the only existing super learner we
+know which works without a pre-specified censoring model. As both the state
+learner and \texttt{survSL} also provides a prediction model for the probability
+of being censored, we compare the performance of both the outcome and the
+censoring modeling using the IPA.
+
+The results are shown in Figure~\ref{fig:zelefski}. We see that the prediction
+models selected for both censoring and outcome have similar or higher API
+compared to the prediction models selected by \texttt{survSL} for most sample
+sizes.
+
+\begin{figure}
+  \centerline{\includegraphics[width=1\linewidth]{../figures/experiment-fig-sl-survSL.pdf}}
+  \caption[]{For the risk prediction models of both the outcome and the
+    censoring probability provided by each of the super learners, the IPA is
+    plotted against sample size. The results are based on 1000 repetitions and
+    the error bars are used to quantify the Monte Carlo uncertainty.}
+  \label{fig:zelefski}
+\end{figure}
+
+\section{Example with prostate cancer study}
+\label{sec:real-data-appl}
+
+The prostate cancer data analyzed by \cite{kattan2000pretreatment}, which we
+introduced in Section~\ref{sec:numer-exper}, include a competing event in the
+form of death without tumor recurrence. For illustration, we fit the state
+learner to the original data set consisting of 1,042 patients. We consider death
+without tumor recurrence and recurrence of tumor as two competing events of
+interest. We include the fives learner \texttt{KM}, \texttt{Cox strata CS},
+\texttt{Lasso}, \texttt{Elastic}, and \texttt{RF} which are described in
+Table~\ref{tab:zel-library}. We use the same library of learners to learn
+\( \Lambda_1 \), \( \Lambda_2 \), and $\Gamma$. In this case, \( \Lambda_1 \)
+denotes the cause-specific cumulative hazard function of tumor recurrence, and
+\( \Lambda_2 \) denotes the cause-specific cumulative hazard function of death
+without tumor recurrence.
 
@@
            & \( \texttt{Cox strata CS} \)  & Cox model stratified on CS  \\
-           & \( \texttt{Cox strata HT} \)  & Cox model stratified on HT \\
-           & \( \texttt{Cox spline} \)  & PSA and RD modeled with splines \\ 
+           % & \( \texttt{Cox strata HT} \)  & Cox model stratified on HT \\
+           % & \( \texttt{Cox spline} \)  & PSA and RD modeled with splines \\ 
     Penalized Cox & \texttt{Lasso} & Cox model with \( L_1 \)-norm penalty   \\
-           & \texttt{Ridge} & Cox model with \( L_2 \)-norm penalty \\
+           % & \texttt{Ridge} & Cox model with \( L_2 \)-norm penalty \\
            & \texttt{Elastic} & Cox model with \( L_1 \)- and \( L_2 \)-norm penalty \\
-    Random forest & \texttt{RF} & Random forest with 50 trees and default settings \\
+    Random forest & \texttt{RF} & Random forest with default settings \\
     \bottomrule 
   \end{tabular}
-  \caption[]{Overview of the nine learners used in scenario~2 of the simulation
-    study. The Kaplan-Meier estimator was fitted using the package
-    \texttt{prodlim} \citep{Gerds_2019prodlim}. All Cox models included all five
-    covariates in the model and were fitted using the package \texttt{survival}
-    \citep{survival-package}. All penalized Cox models included all five
-    covariates as linear predictors and were fitted using the package
-    \texttt{glmnet} \citep{glmnet-cox,glmnet-glm}. The random forest was fitted
-    with the package \texttt{randomForestSRC} \citep{rfsrc-paclage}.}
+  \caption[]{Overview of the five learners used by the state learner when
+    applied to the prostate cancer data set. The Kaplan-Meier estimator was
+    fitted using the package \texttt{prodlim} \citep{Gerds_2019prodlim}. All Cox
+    models included all five covariates in the model and were fitted using the
+    package \texttt{survival} \citep{survival-package}. All penalized Cox models
+    included all five covariates as linear predictors and were fitted using the
+    package \texttt{glmnet} \citep{glmnet-cox,glmnet-glm}. The random forest was
+    fitted with the package \texttt{randomForestSRC} \citep{rfsrc-paclage}.}
   \label{tab:zel-library}
 \end{table}
-
-The results are shown in Figure~\ref{fig:zelefski}. We see that the Brier score
-of the IPCW super learner based on a misspecified censoring model (IPCW(KM))
-decreased with sample size, but is larger than that of the other super learners
-for all sample sizes. The state learner and the IPCW super learner based on a
-correctly pre-specified censoring model demonstrate similar performance for all
-sample sizes. The performance of both the IPCW(Cox) super learner and the state
-learner approaches the benchmark provided by the oracle selector for large sample
-sizes.
-\begin{figure}
-  \centerline{\includegraphics[width=1\linewidth]{../figures/zelefski-sim.pdf}}
-  \caption[]{Results for scenario~2 of the simulation study. For the learner
-    selected by each of the four discrete super learners, the Brier score
-    calculated in a large independent data set without censoring is plotted
-    against sample size. The results are based on 200 repetitions.}
-  \label{fig:zelefski}
-\end{figure}
-
-\section{Case study: Bystander efforts in out-of-hospital cardiac arrest}
-\label{sec:case-study:-byst}
-
-\section{Real data application}
-\label{sec:real-data-appl}
-
-The original prostate cancer data analyzed by \cite{kattan2000pretreatment},
-which we introduced in Section~\ref{sec:numer-exper}, include a competing event
-in the form of death without tumor recurrence. To illustrate our method we fit
-the state learner to the original data set consisting of 1,042 patients. We
-consider death without tumor recurrence and recurrence of tumor as two competing
-events of interest. We include the fives learner \texttt{KM}, \texttt{Cox strata
-  CS}, \texttt{Lasso}, \texttt{Elastic}, and \texttt{RF} which are described in
-Table~\ref{tab:zel-library}. We use the same library of learners to learn
-\( \Lambda_1 \), \( \Lambda_2 \), and $\Gamma$. In this case, \( \Lambda_1 \)
-denotes the cause-specific cumulative hazard function of tumor recurrence, and
-\( \Lambda_2 \) denotes the cause-specific cumulative hazard function of death
-without tumor recurrence.
 
@@
 models, and we repeat training and evaluation five times with different splits.
-The integrated Brier score for all learners are shown in
-Figure~\ref{fig:zelefski-real}, and the top 10 combinations of learners are
-displayed in Table~\ref{tab:zelefski-real}. We see that the prediction
-performance is mostly affected by the choice of learner for the censoring
-distribution. Several combinations of learners give similar performance as
-measured by the integrated Brier score, as long as a random forest is used to
-model the censoring distribution.
+The integrated Brier score (defined in Section~\ref{sec:super-learner-simple})
+for all learners are shown in Figure~\ref{fig:zelefski-real}, and the top 10
+combinations of learners are displayed in Table~\ref{tab:zelefski-real}. We see
+that the prediction performance is mostly affected by the choice of learner for
+the censoring distribution. Several combinations of learners give similar
+performance as measured by the integrated Brier score, as long as a random
+forest is used to model the censoring distribution.
 
@@
     learners of \( \Lambda_1 \), \( \Lambda_2 \), and $\Gamma$ as `Tumor
-    learner', `Death leaner', and `Censoring learner', respectively.}
+    learner', `Death learner', and `Censoring learner', respectively.}
   \label{tab:zelefski-real}
 \end{table}
+
+As an illustration, we use the learners of the three cumulative hazard functions
+selected by the state learner and the estimator defined in
+Section~\ref{sec:cause-spec-aver}, equation~(\ref{eq:one-step-comp-ate}), to
+estimate the average treatment effect of hormone therapy on risk of tumor
+recurrence and death. The results are shown in
+Figure~\ref{fig:zelefski-real-target} for 6 month intervals after baseline with
+point wise 95\% confidence intervals. We see that hormone therapy appears to
+decrease the risk of tumor recurrence and increase the risk of death without
+tumor recurrence, but that not of the estimated effects are significant.
+
+\begin{figure}
+  \centerline{\includegraphics[width=1\linewidth]{../figures/zelefsky-data-target-par.pdf}}
+  \caption[]{Estimates of the average treatment effect of hormone therapy on the
+    risk of tumor recurrence and death obtained from the prostate cancer study
+    analyzed by \cite{kattan2000pretreatment}. The estimates are based on the
+    estimator defined in equation~(\ref{eq:one-step-comp-ate}) with point wise
+    95\% confidence intervals calculated as described in
+    Section~\ref{sec:targeted-learning}. The estimates of the nuisance
+    parameters are provided by the state learner.}
+  \label{fig:zelefski-real-target}
+\end{figure}
+
 \section{Discussion}
@@
 
-\fxnote*{mention this?}{[Double robustness maybe lacking]}
+% \fxnote*{mention this?}{[Double robustness maybe lacking]}
 

### 2024-05-21 sandbox/zelefsky-study/_targets.R

@@ -1,176 +1,0 @@
-### _targets.R --- 
-#----------------------------------------------------------------------
-## Author: Thomas Alexander Gerds
-## Created: Apr 27 2023 (11:06) 
-## Version: 
-## Last-Updated: Mar 13 2024 (09:48) 
-##           By: Anders Munch
-##     Update #: 177
-#----------------------------------------------------------------------
-## 
-### Commentary: 
-## 
-### Change Log:
-#----------------------------------------------------------------------
-## 
-### Code:
-library(here)
-try(setwd(here("zelefsky-study")))
-try(setwd("~/research/SuperVision/Anders/survival-loss/statelearner/zelefsky-study/"))
-library(targets)
-library(tarchetypes)
-library(parallel)
-tar_option_set(packages = c("glmnet","lava","foreach","data.table","prodlim","survival","riskRegression","ranger"))
-tar_source("functions")
-list(
-    tar_target(test_data,draw_data(484)),
-    tar_target(test_fit,{
-        outcome_models <- list(cox0 = coxph(Surv(time,event = event)~1, data=test_data, x = TRUE, y = TRUE),
-                               cox1 = coxph(Surv(time,event = event)~X1+X2, data=test_data, x = TRUE, y = TRUE),
-                               cox2 = coxph(Surv(time,event = event)~X6+X7, data=test_data, x = TRUE, y = TRUE),
-                               cox_dgm = coxph(Surv(time,event = event)~ X1 + X2 + X5 + X6 + X7 + X8 + X11 + X12 + quad.X3 + quad.X4 + quad.X5 + quad.X6,data = test_data,x = TRUE,y = TRUE),
-                               rf = ranger(Surv(time, event = event) ~ X1+X2+X3+X4+X5+X6+X7+X8+X9+X10+X11+X12, data = test_data,num.tree = 50)
-                               ## rf_try = ranger(Surv(time, event = event) ~ X1+X2+X6+X7, mtry = 4, data = test_data),
-                               ## rf_deep = ranger(Surv(time, event = event) ~ X1+X2+X6+X7, mtry = 4, min.node.size = 1, data = test_data)
-                               )
-        cens_models <- list(cox1 = coxph(Surv(time,event = cens_status)~X1+X2, data=test_data, x = TRUE, y = TRUE),
-                            cox2 = coxph(Surv(time,event = cens_status)~X6+X7, data=test_data, x = TRUE, y = TRUE),
[…]

### 2024-05-21 sandbox/zelefsky-study/functions/ipcw-fail-sim.R

@@ -1,94 +1,0 @@
-### ipcw-fail-sim.R --- 
-#----------------------------------------------------------------------
-## Author: Anders Munch
-## Created: Jul 23 2023 (10:29) 
-## Version: 
-## Last-Updated: Aug  3 2023 (08:52) 
-##           By: Anders Munch
-##     Update #: 76
-#----------------------------------------------------------------------
-## 
-### Commentary: 
-## 
-### Change Log:
-#----------------------------------------------------------------------
-## 
-### Code:
-library(data.table)
-library(prodlim)
-library(riskRegression)
-library(lava)
-library(survival)
-
-sim_data <- function(n){
-    m <- lvm()
-    formula_event <- ~f(A,-.8)
-    formula_cens <- ~f(A,.9)
-    event_scale <- 1/500
-    cens_scale <- 1/500
-    lava::distribution(m,~A) <- lava::binomial.lvm(p = .3)
-    lava::distribution(m,~censtime) <- lava::coxWeibull.lvm(scale=event_scale)
-    lava::distribution(m,~eventtime) <- lava::coxWeibull.lvm(scale=cens_scale)
-    m <- lava::eventTime(m, time ~ min(eventtime = 1, censtime = 0), "event")
-    lava::regression(m) <- stats::update(formula_event, "eventtime~.")
-    lava::regression(m) <- stats::update(formula_cens, "censtime~.")
-    out <- setDT(sim(m,n))
-    out[,dummy := 1]
-    return(out[])
-}
-
-sim_ipcw_fail <- function(ns, eval_time = 20, B = 1, split.method = "cv5"){
-    do.call(rbind, lapply(ns, function(nn){
-        sim_dt = sim_data(n = nn)
-        mms = list(dgm = coxph(Surv(time, event)~strata(A), data = sim_dt, x = TRUE, y = TRUE),
-                   km = coxph(Surv(time, event)~1, data = sim_dt, x = TRUE, y = TRUE))
-        forms = list(km = Hist(time,event)~1,
-                     dgm = Hist(time,event)~strat(A))
-        ipcw_briers = lapply(seq_along(forms), function(ii){
-            try_message <- try({
-                ## What we want to happen
-                xx = Score(mms,data = sim_dt,formula = forms[[ii]], metrics = "brier",se.fit = FALSE,times = eval_time,contrasts = FALSE,null.model = FALSE, B = B, split.method = split.method, progress.bar = NULL)$Brier$score
[…]

### 2024-05-21 sandbox/zelefsky-study/functions/run_simulation.R

@@ -1,51 +1,0 @@
-### run_simulation.R --- 
-#----------------------------------------------------------------------
-## Author: Thomas Alexander Gerds
-## Created: Jul 18 2023 (10:49) 
-## Version: 
-## Last-Updated: Jul 18 2023 (15:22) 
-##           By: Thomas Alexander Gerds
-##     Update #: 10
-#----------------------------------------------------------------------
-## 
-### Commentary: 
-## 
-### Change Log:
-#----------------------------------------------------------------------
-## 
-### Code:
-run_simulation <- function(n,nlarge = 100000,time_horizon){
-    learn_data = draw_data(n = n)
-    outcome_models <- list(cox0 = coxph(Surv(time,event = event)~1, data=learn_data, x = TRUE, y = TRUE),
-                           cox1 = coxph(Surv(time,event = event)~X1+X2, data=learn_data, x = TRUE, y = TRUE),
-                           cox2 = coxph(Surv(time,event = event)~X6+X7, data=learn_data, x = TRUE, y = TRUE),
-                           ## cox_dgm = coxph(Surv(time,event = event)~ X1 + X2 + X5 + X6 + X7 + X8 + X11 + X12 + quad.X3 + quad.X4 + quad.X5 + quad.X6,data = learn_data,x = TRUE,y = TRUE),
-                           rf = ranger(Surv(time, event = event) ~ X1+X2+X3+X4+X5+X6+X7+X8+X9+X10+X11+X12, data = learn_data,num.tree = 50))
-    cens_models <- list(cox0 = coxph(Surv(time,event = event)~1, data=learn_data, x = TRUE, y = TRUE),
-                        cox1 = coxph(Surv(time,event = cens_status)~X1+X2, data=learn_data, x = TRUE, y = TRUE),
-                        cox2 = coxph(Surv(time,event = cens_status)~X6+X7, data=learn_data, x = TRUE, y = TRUE),
-                        ## cox_dgm = coxph(Surv(time,event = cens_status)~ X1 + X2 + X5 + X6 + X7 + X8 + X11 + X12 + quad.X3 + quad.X4 + quad.X5 + quad.X6,data = learn_data,x = TRUE,y = TRUE),
-                        rf = ranger(Surv(time, event = cens_status) ~ X1+X2+X6+X7, data = learn_data,num.tree = 50))
-    sl = statelearner(learners = list(state = outcome_models,censoring = cens_models),data = learn_data,B = 4,times = time_horizon)
[…]

### 2024-05-21 sandbox/zelefsky-study/functions/simZelefsky.R

@@ -1,69 +1,0 @@
-### zelefsky-empirical-simulate-data.R ---
-#----------------------------------------------------------------------
-## Author: Thomas Alexander Gerds
-## Created: Nov 9 2018 (9:49)
-## Version:
-#----------------------------------------------------------------------
-##
-### Commentary: Function simulation data similar to the zelefsky data
-##
-### Change Log:
-#----------------------------------------------------------------------
-##
-### Code:
-simZelefsky <- function(n,censoring = TRUE,simulation_input, ...){
-    m <- lvm()
-    input.surv = simulation_input$surv
-    input.cens = simulation_input$cens
-    zelefsky = simulation_input$zelefsky
-    distribution(m,~time.recur) <- coxWeibull.lvm(scale=exp(-input.surv$coef["(Intercept)"]/input.surv$scale),shape=1/input.surv$scale)
-    distribution(m,~time.cens) <- coxWeibull.lvm(scale=exp(-input.cens$coef["(Intercept)"]/input.cens$scale),shape=(1/input.cens$scale))
-    ## put code to generate covariates: psa, stage, ggtot, dose100, hormones
-    ## alike the real data distributions, maybe even with
-    ## regression statements between them to make them correlated
-
-    distribution(m,~logPSA) <- normal.lvm(mean=mean(log(zelefsky$psa)),sd=sd(log(zelefsky$psa)))
-    m <- categorical(m,~stage, K=6, p=c(table(zelefsky$stage)[-6]/nrow(zelefsky)), labels=c("T1c","T2a","T2b","T2c","T3ab","T3c"))
-    transform(m,stageT2a~stage) <- function(x){1*(x=="T2a")}
-    transform(m,stageT2b~stage) <- function(x){1*(x=="T2b")}
-    transform(m,stageT2c~stage) <- function(x){1*(x=="T2c")}
-    transform(m,stageT3ab~stage) <- function(x){1*(x=="T3ab")}
-    transform(m,stageT3c~stage) <- function(x){1*(x=="T3c")}
-    distribution(m,~ggtot) <- normal.lvm(mean=mean(zelefsky$ggtot),sd=sd(zelefsky$ggtot))
-    distribution(m,~sDose) <- normal.lvm(mean=mean(zelefsky$sDose),sd=sd(zelefsky$sDose))    
[…]

### 2024-05-21 sandbox/zelefsky-study/functions/statelearner-v2.R

@@ -1,263 +1,0 @@
-### statelearner-v2.R --- 
-#----------------------------------------------------------------------
-## Author: Anders Munch
-## Created: Aug 18 2023 (10:26) 
-## Version: 
-## Last-Updated: Mar 13 2024 (09:14) 
-##           By: Anders Munch
-##     Update #: 365
-#----------------------------------------------------------------------
-## 
-### Commentary: 
-## 
-### Change Log:
-#----------------------------------------------------------------------
-## 
-### Code:
-
-
-## Set data on canonical form
-form_data <- function(data, max_time = Inf, time = "time", status = "status", cause_codes = c("1" = 1, "2" = 2, "c" = 0), vars = NULL){
-    if(!is.null(vars))
-        wd = copy(data)[, c(time, status, vars), with = FALSE]
-    else
-        wd = copy(data)
-    setnames(wd, old = c(time, status), new = c("time", "status"))
-    ## Make cause of interest = 1, everything else = 2
-    wd[, cause1 := 1+1*(status != cause_codes[["1"]] & time <= max_time)]
-    wd[, censor := 1+1*(status != cause_codes[["c"]] & time <= max_time)]
-    if(wd[, length(unique(status))>2])
-        wd[, cause2 := 1+1*(status != cause_codes[["2"]] & time <= max_time)]
-    wd[, status := NULL]
-    return(wd[])
-}
-## Fit model on canonical data form
-fit_cause_model <- function(model, data, cause, x_form = NULL, ...){
-    cause_names = c("cause1", "cause2", "censor")
-    cause_names = cause_names[cause_names %in% names(data)]
-    stopifnot(cause %in% cause_names)
-    wd = copy(data)[, -cause_names[cause_names != cause], with = FALSE]
-    setnames(wd, old = cause, new = "event")
-    ## not so elegant
-    if(is.null(x_form))
-        form = Surv(time, event) ~ .
-    else
-        form = update(x_form, Surv(time, event) ~ .)
-    if(model == "cox"){
-        wd[, event := 1*(event == 1)] ## To make this cause of interest
-        out = coxph(form, data = wd,x = TRUE,y = TRUE, ...)
-    }
-    if(model == "GLMnet"){
-        wd[, event := 1*(event == 1)] ## To make this cause of interest
-        out = GLMnet(form, data = wd, ...)
-    }
-    if(model == "rfsrc"){
-        out = rfsrc(form, data = wd, ...)
-    }
[…]

### 2024-05-21 sandbox/zelefsky-study/functions/statelearner.R

@@ -1,137 +1,0 @@
-### statelearner.R --- 
-#----------------------------------------------------------------------
-## Author: Thomas Alexander Gerds
-## Created: Jul 11 2023 (13:29) 
-## Version: 
-## Last-Updated: Aug 21 2023 (10:12) 
-##           By: Anders Munch
-##     Update #: 89
-#----------------------------------------------------------------------
-## 
-### Commentary: 
-## 
-### Change Log:
-#----------------------------------------------------------------------
-## 
-### Code:
-statelearner <- function(learners,
-                         data,
-                         times,
-                         integrate = TRUE,
-                         split.method = "cv5",
-                         collapse = TRUE,
-                         B=1,
-                         verbose = FALSE,
-                         time_grid_length = 100){
-    requireNamespace(package = "data.table")
-    ## NB: Expecting that time variable is called time
-    partition = riskRegression::getSplitMethod(split.method=split.method,B=B,N=data[, .N])
-    wd = data.table::copy(data)
-    setDT(wd)
-    wd[, event_c := 1*(!event)]
-    stopifnot(length(times)>0)
-    if (length(times) == 1)
-        time_grid = seq(0, times, length.out = time_grid_length)
-    else
-        time_grid = times
-    ## fix names
-    if(is.null(names(learners$state)))
-        names(learners$state) = paste0("out_", 1:length(learners$state))
-    if(is.null(names(learners$censoring)))
-        names(learners$censoring) = paste0("cens_", 1:length(learners$censoring))
-    cv_fits = do.call(rbind, lapply(1:partition$B, function(bb){
-        if(verbose) message("--- Running on split ", bb, " out of ", B, " ---")
-        out0 = do.call(rbind, lapply(1:partition$k, function(fold_k){
-            if(verbose) message("------ Running on split ", fold_k, " out of ", partition$k)
-            train = wd[partition$index(bb)!=fold_k]
-            test = wd[partition$index(bb)==fold_k]
-            ## Construct time grid to evaluate hazard function
-            eval_times = sort(unique(c(0, train[, sort(unique(time))])))
-            eval_times = eval_times[eval_times <= max(time_grid)]
-            ## Construct list of cumulative hazard predictions
-            ## This might give memory problems...
-            ## Solving this partly by only calculating all CHs for censoring only
-            ## The other calculation happens in outer loop below
[…]

### 2024-05-21 sandbox/zelefsky-study/ipcw-fail-example.R

@@ -1,72 +1,0 @@
-### ipcw-fail-example.R --- 
-#----------------------------------------------------------------------
-## Author: Anders Munch
-## Created: Jul 21 2023 (16:27) 
-## Version: 
-## Last-Updated: Jul 23 2023 (11:46) 
-##           By: Anders Munch
-##     Update #: 34
-#----------------------------------------------------------------------
-## 
-### Commentary: 
-## 
-### Change Log:
-#----------------------------------------------------------------------
-## 
-### Code:
-library(data.table)
-library(prodlim)
-library(riskRegression)
-library(targets)
-library(lava)
-library(data.table)
-library(riskRegression)
-library(survival)
-library(ranger)
-
-try(setwd("~/research/SuperVision/Anders/survival-loss/statelearner/empirical-study/"))
-try(setwd("~/Documents/phd/survival-loss-function/statelearner/empirical-study/"))
-tar_source("functions")
-
-## Trying with binary
-sim2 <- function(n){
-    m <- lvm()
-    formula_event <- ~f(A,-.8)
-    formula_cens <- ~f(A,.9)
-    event_scale <- 1/500
-    cens_scale <- 1/500
-    lava::distribution(m,~A) <- lava::binomial.lvm(p = .3)
-    lava::distribution(m,~censtime) <- lava::coxWeibull.lvm(scale=event_scale)
-    lava::distribution(m,~eventtime) <- lava::coxWeibull.lvm(scale=cens_scale)
-    m <- lava::eventTime(m, time ~ min(eventtime = 1, censtime = 0), "event")
-    lava::regression(m) <- stats::update(formula_event, "eventtime~.")
-    lava::regression(m) <- stats::update(formula_cens, "censtime~.")
-    out <- setDT(sim(m,n))
-    out[,dummy := 1]
-    return(out[])
-}
-
-eval_time <- 20
-dlarge <- sim2(1000)
-par(mfrow = c(1,2))
-plot(prodlim(Hist(time,event)~A, data = dlarge), xlim = c(0,eval_time), type = "risk")
-plot(prodlim(Hist(time,event)~A, data = dlarge, reverse = TRUE), xlim = c(0,eval_time))
-plot(prodlim(Hist(time,event)~1, data = dlarge, reverse = TRUE), xlim = c(0,eval_time), add = TRUE, col = "red")
-par(mfrow = c(1,1))
-mms <- list(dgm = coxph(Surv(time, event)~strata(A), data = dlarge, x = TRUE, y = TRUE),
[…]

### 2024-05-21 sandbox/zelefsky-study/run.R

@@ -1,200 +1,0 @@
-### run.R --- 
-#----------------------------------------------------------------------
-## Author: Thomas Alexander Gerds
-## Created: Jul 11 2023 (14:07) 
-## Version: 
-## Last-Updated: Jul 23 2023 (19:48) 
-##           By: Anders Munch
-##     Update #: 173
-#----------------------------------------------------------------------
-## 
-### Commentary: 
-## 
-### Change Log:
-#----------------------------------------------------------------------
-## 
-### Code:
-## library(here)
-try(setwd("~/research/SuperVision/Anders/survival-loss/statelearner/empirical-study/"))
-try(setwd("/home/amnudn/Documents/phd/survival-loss-function/statelearner/empirical-study"))
-library(targets)
-library(lava)
-library(data.table)
-library(riskRegression)
-library(survival)
-library(ranger)
-library(ggplot2)
-tar_make()
-
-## IPCW fail example:
-tar_load(ipcw_fail)
-ipcw_fail[loss_type == "sl-brier", learner := "statelearner"]
-ipcw_fail[loss_type == "brier", learner := paste0("IPCW(", cens_model , ")")]
-winners <- rbind(ipcw_fail[!is.na(loss), .SD[min(loss) == loss, .(winner = out_model[1], oracle_loss = oracle_loss[1])], .(learner, n, n_sim)],
-                 ipcw_fail[learner == "IPCW(dgm)", .SD[min(oracle_loss) == oracle_loss, .(winner = out_model[1], oracle_loss = oracle_loss[1])], .(learner, n, n_sim)][, learner := "oracle"])
-winners_summ <- winners[, .(correct = mean(winner == "dgm")), .(learner,n)]
-
-## Compare oracle risk
-ggplot(winners[,.(oracle_risk = 100*mean(oracle_loss)), .(learner, n)], aes(x = n, y = oracle_risk, col = learner)) +
-    geom_line() + geom_point() + 
-    theme_bw()
-
-## See how many times we get things right
-ggplot(winners_summ, aes(x = n, y = correct, col = learner)) +
-    geom_line() + geom_point() + 
-    theme_bw() + 
-    ylim(c(0,1))
-
-
-## Simulate data based on zelefsky data
-tar_load(sim_zel_learners)
[…]

### 2024-05-21 sandbox/zelefsky-study/sandbox/yet_another.R

@@ -1,50 +1,0 @@
-library(riskRegression)
-library(targets)
-library(survival)
-library(lava)
-library(data.table)
-library(prodlim)
-try(setwd("~/research/SuperVision/Anders/survival-loss/statelearner/empirical-study/"))
-tar_source("functions")
-m <- lvm()
-# B discriminates only when A == 1
-formula_event <- ~f(A,0)+f(B,0)+f(AB,-1.5)+f(AC,-1.3)
-formula_cens <- ~f(A,0)+f(B,2)+f(C,0)
-event_scale <- 1/1000
-cens_scale <- 1/1000
-lava::distribution(m,~A) <- lava::binomial.lvm(p = .4)
-lava::distribution(m,~B) <- lava::normal.lvm()
-lava::distribution(m,~C) <- lava::normal.lvm()
-transform(m,AB~A+B) <- prod
-transform(m,AC~A+C) <- prod
-lava::distribution(m,~censtime) <- lava::coxWeibull.lvm(scale=event_scale)
-lava::distribution(m,~eventtime) <- lava::coxWeibull.lvm(scale=cens_scale)
-m <- lava::eventTime(m, time ~ min(eventtime = 1, censtime = 0), "event")
-lava::regression(m) <- stats::update(formula_event, "eventtime~.")
-lava::regression(m) <- stats::update(formula_cens, "censtime~.")
-d <- setDT(sim(m,5000))
-d[,table(event)]
-dlarge <- setDT(sim(m,5000))
-dlarge[,dummy := 1]
-## pdf("~/tmp/test.pdf",width = 12,height = 6)
-## pdf("~/tmp/test.pdf",width = 10,height = 10)
-## par(mfrow = c(2,2))
-## plot(prodlim(Hist(time,event)~A+B,data = d),xlim = c(0,20),newdata = expand.grid(A =c(0,1),B =c(0,1)),type = "risk",plot.main = "Event probability",confint = FALSE,legend.cex = 0.5)
-## plot(prodlim(Hist(time,event)~A+C,data = d),xlim = c(0,20),newdata = expand.grid(A = c(0,1),C = c(0,1)),type = "risk",plot.main = "Event probability",confint = FALSE,legend.cex = 0.5)
[…]

### 2024-05-27 tex/journal-version/manuscript.tex
renaming
reformat manuscript

--- 2024-05-21 tex/jrss-version/state-learner-jrss-style.tex
@@
-%%
-%% Copyright 2022 OXFORD UNIVERSITY PRESS
-%%
-%% This file is part of the 'oup-authoring-template Bundle'.
-%% ---------------------------------------------
-%%
-%% It may be distributed under the conditions of the LaTeX Project Public
-%% License, either version 1.2 of this license or (at your option) any
-%% later version.  The latest version of this license is in
-%%    http://www.latex-project.org/lppl.txt
-%% and version 1.2 or later is part of all distributions of LaTeX
-%% version 1999/12/01 or later.
-%%
-%% The list of all files belonging to the 'oup-authoring-template Bundle' is
-%% given in the file `manifest.txt'.
-%%
-%% Template article for OXFORD UNIVERSITY PRESS's document class `oup-authoring-template'
-%% with bibliographic references
-%%
-
-%%%CONTEMPORARY%%%
-% \documentclass[unnumsec,webpdf,contemporary,large]{oup-authoring-template}%
-\documentclass[numsec,webpdf,contemporary,medium,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
-%\documentclass[unnumsec,webpdf,contemporary,medium]{oup-authoring-template}
-%\documentclass[unnumsec,webpdf,contemporary,small]{oup-authoring-template}
-
-%%%MODERN%%%
-%\documentclass[unnumsec,webpdf,modern,large]{oup-authoring-template}
-% \documentclass[unnumsec,webpdf,modern,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
-%\documentclass[unnumsec,webpdf,modern,medium]{oup-authoring-template}
-%\documentclass[unnumsec,webpdf,modern,small]{oup-authoring-template}
-
-%%%TRADITIONAL%%%
-%\documentclass[unnumsec,webpdf,traditional,large]{oup-authoring-template}
-% \documentclass[unnumsec,webpdf,traditional,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
-%\documentclass[unnumsec,namedate,webpdf,traditional,medium]{oup-authoring-template}
-%\documentclass[namedate,webpdf,traditional,small]{oup-authoring-template}
-
-\onecolumn % for one column layouts
-
-%\usepackage{showframe}
-
-% \graphicspath{{Fig/}}
-
-% line numbers
-%\usepackage[mathlines, switch]{lineno}
-%\usepackage[right]{lineno}
+\documentclass[12pt]{article}
+\usepackage[a4paper,left=2.8cm, right=2.8cm, top=3cm, bottom=3cm]{geometry}
+
+%% Packages
+\usepackage{amsthm, dsfont, amssymb,soul,xcolor,
+  % amsfonts,
+  enumitem,
+  amsmath,graphicx,verbatim,pgfplots,tikz,prodint,booktabs}
+
+\usepackage[authoryear]{natbib}
+\bibliographystyle{abbrvnat}
+
+% % Handling new lines
+% \setlength{\parskip}{1em}
+% \setlength{\parindent}{0em}
+
+% % Handling space after sections
+% \usepackage{titlesec}
+% \titlespacing*{\section}{0em}{2em}{0em}
+% \titlespacing*{\subsection}{0em}{2em}{0em}
+% \titlespacing*{\subsubsection}{0em}{2em}{0em}
+
+% % No spacing after in start of list
+% \setlist[itemize]{topsep=0pt}
+% \setlist[enumerate]{topsep=0pt}
+
+\definecolor{linkcolor}{rgb}{0, 0, 0.54}
+\usepackage[colorlinks,allcolors=linkcolor]{hyperref}
 
@@
 
-%% OLD
-
-% \usepackage{natbib, dsfont,
-% amssymb,soul,xcolor,amsmath,graphicx,verbatim,pgfplots,tikz,prodint,booktabs}
-
-\usepackage{dsfont, booktabs}
-
 % \usepackage[tablesonly,heads,nolists]{endfloat}
@@
 %% change enumerate labels
-\renewcommand{\labelenumi}{\normalfont (\alph{enumi})}
+% \renewcommand{\labelenumi}{\normalfont (\alph{enumi})}
+
+%% Font size etc
+\linespread{1.65}
+
+\usepackage{lineno}
+\linenumbers
 
@@
 
-% \bibliographystyle{rss}
-
 %% Toggle with or without figures
@@
 
+\RequirePackage{graphicx}
+% \RequirePackage{caption}
+% \RequirePackage{array}
+% \RequirePackage{color}
+% \RequirePackage{xcolor}
+% \RequirePackage{amssymb}
+% \RequirePackage{flushend}
+% \RequirePackage{stfloats}
+% \RequirePackage[figuresright]{rotating}
+% \RequirePackage{chngpage}
+% \RequirePackage{totcount}
+% \RequirePackage{fix-cm}
 \newcommand{\includeFigCond}[2][]{
@@
 
+%% Title
+\title{The state learner -- a super learner for right-censored data}
+
+\usepackage{authblk}
+\author[1]{Anders Munch}
+\author[1]{Thomas~A.~Gerds}
+\affil[1]{Section of Biostatistics, University of Copenhagen}
+
 \begin{document}
 
-%% NEW
-
-\journaltitle{Journal of the Royal Statistical Society Series B: Statistical
-  Methodology} %
-\DOI{DOI HERE} %
-\copyrightyear{xxxx} %
-\pubyear{yyyy} \access{Advance Access Publication Date: Day Month Year}
-\appnotes{Paper}
-
-\firstpage{1}
-
-%\subtitle{Subject Section}
-
-\title[The state learner]{The state learner -- a super learner for right-censored data}
-
-\author[1,$\ast$]{Anders Munch}
-\author[1]{Thomas A.\ Gerds}
-
-\authormark{Munch et al.}
-
-\address[1]{\orgdiv{Section of Biostatistics}, \orgname{University of
-    Copenhagen}, \orgaddress{\street{Østerfarigmagsgade 5}, \postcode{2100}, \state{Copenhagen}, \country{Denmark}}}
-
-\corresp[$\ast$]{Address for correspondence: \href{email:email-id.com}{a.munch@sund.ku.dk}}
+% \authormark{Munch et al.}
+
+% \address[1]{\orgdiv{Section of Biostatistics}, \orgname{University of
+%     Copenhagen}, \orgaddress{\street{Østerfarigmagsgade 5}, \postcode{2100}, \state{Copenhagen}, \country{Denmark}}}
+
+% \corresp[$\ast$]{Address for correspondence: \href{email:email-id.com}{a.munch@sund.ku.dk}}
 
@@
 
-\abstract{In survival analysis, prediction models are needed as
-  stand-alone tools and in applications of causal inference to
-  estimate nuisance parameters. The super learner is a machine
-  learning algorithm which combines a library of prediction models
-  into a meta learner based on cross-validated loss. In right censored
-  data, the choice of the loss function and the estimation of the
-  expected loss need careful consideration. We introduce the state
-  learner, a new super learner for survival analysis, which
-  simultaneously evaluates libraries of prediction models for the
-  event of interest and the censoring distribution. The state learner
-  can be applied to all types of survival models, works in the
-  presence of competing risks, and does not require a single
-  pre-specified estimator of the conditional censoring
-  distribution. We establish an oracle inequality for the state
-  learner and investigate its performance through numerical
-  experiments. We illustrate the application of the state learner with
-  prostate cancer data, as a stand-alone prediction tool, and, for
-  causal inference, as a way to estimate the nuisance parameter models of
-  a smooth statistical functional.
-  % We illustrate how the state learner allows us to
-  % estimate causal effects in a competing risks setting without having
-  % to pre-specify models for neither the cause-specific hazards of
-  % interest nor the censoring distribution.
-}
-
-\keywords{Competing risks, cross-validation, loss based estimation,
-  right-censored data, super learner}
+% \keywords{Competing risks, cross-validation, loss based estimation,
+%   right-censored data, super learner}
 
 \maketitle
+
+\begin{abstract}
+  In survival analysis, prediction models are needed as stand-alone tools and in
+  applications of causal inference to estimate nuisance parameters. The super
+  learner is a machine learning algorithm which combines a library of prediction
+  models into a meta learner based on cross-validated loss. In right censored
+  data, the choice of the loss function and the estimation of the expected loss
+  need careful consideration. We introduce the state learner, a new super
+  learner for survival analysis, which simultaneously evaluates libraries of
+  prediction models for the event of interest and the censoring distribution.
+  The state learner can be applied to all types of survival models, works in the
+  presence of competing risks, and does not require a single pre-specified
+  estimator of the conditional censoring distribution. We establish an oracle
+  inequality for the state learner and investigate its performance through
+  numerical experiments. We illustrate the application of the state learner with
+  prostate cancer data, as a stand-alone prediction tool, and, for causal
+  inference, as a way to estimate the nuisance parameter models of a smooth
+  statistical functional.
+\end{abstract}
 
@@
 and~\ref{sec:state-learner-with} contain proofs.
+An implementation of the state learner is available at
+\url{https://github.com/amnudn/statelearner} along with codes for reproducing
+our numerical experiments.
 
@@
 \end{equation}
-The function \( F \) describes the conditional state occupation
-probabilities of the multi-state process \(\eta\). We construct a
-super learner for \( F \). The target of this super learner is the
-function-valued parameter $\Psi(P) = F$ which is identified through
-equation~(\ref{eq:F-def}).
-Under conditional independent censoring each quadruple
-$(\Lambda_{1}, \Lambda_{2}, \Gamma, H)$ characterises a distribution
-\(P\in\mathcal P\), c.f.\ equation~\eqref{eq:parametrizeP}, which in turn
-determines \( (F, H) \).
-Hence, a learner for \( F \) can be constructed from learners for
-\( \Lambda_1 \), \( \Lambda_2 \), and $\Gamma$ as follows:
+The function \( F \) describes the conditional state occupation probabilities of
+the multi-state process \(\eta\). We construct a super learner for \( F \). The
+target of this super learner is the function-valued parameter $\Psi(P) = F$
+which is identified through equation~(\ref{eq:F-def}). Under conditional
+independent censoring each quadruple $(\Lambda_{1}, \Lambda_{2}, \Gamma, H)$
+characterises a distribution \(P\in\mathcal P\), c.f.\
+equation~\eqref{eq:parametrizeP}, which in turn determines \( (F, H) \). Hence,
+a learner for \( F \) can be constructed from learners for \( \Lambda_1 \),
+\( \Lambda_2 \), and $\Gamma$ as follows:
 \begin{equation}\label{eq:transition}
   \begin{split}
-  F(t, 1, x)
-  & = P(\tilde{T} \leq t, \tilde{D}=1 \mid X=x)
-    = \int_0^t \exp{{\{-\Lambda_{1}(s \mid x)-\Lambda_{2}(s \mid x) - \Gamma(s \mid x)\} }}  \Lambda_{1}(\diff s \mid x),
-  \\
-  F(t, 2, x)
-  & = P(\tilde{T} \leq t, \tilde{D}=2 \mid X=x)
-    = \int_0^t \exp{\{-\Lambda_{1}(s \mid x)-\Lambda_{2}(s \mid x) - \Gamma(s \mid x)\} }  \Lambda_{2}(\diff s \mid x),
-  \\
-  F(t, -1, x)
-  & =
-    P(\tilde{T} \leq t, \tilde{D}=0 \mid X=x)
-    = \int_0^t \exp{\{-\Lambda_{1}(s \mid x)-\Lambda_{2}(s \mid x) - \Gamma(s \mid x)\} }  \Gamma(\diff s \mid x),
-  \\
     F(t, 0, x)
@@
       = P(\tilde{T} > t \mid X= x)
-    = 1- F(t, 1, x) - F(t, 2, x)- F(t, -1, x).
+      =
+      \exp{{\{-\Lambda_{1}(t \mid x)-\Lambda_{2}(t \mid x) - \Gamma(t \mid x)\}
+      }},
+    \\
+    F(t, 1, x)
+    &
+      = P(\tilde{T} \leq t, \tilde{D}=1 \mid X=x)
+      = \int_0^t F(s-, 0, x)  \Lambda_{1}(\diff s \mid x),
+    \\
+    F(t, 2, x)
+    &
+      = P(\tilde{T} \leq t, \tilde{D}=2 \mid X=x)
+      = \int_0^t  F(s-, 0, x)  \Lambda_{2}(\diff s \mid x),
+    \\
+    F(t, -1, x)
+    &
+      = P(\tilde{T} \leq t, \tilde{D}=0 \mid X=x)
+      = \int_0^t F(s-, 0, x)  \Gamma(\diff s \mid x).
   \end{split}
 \end{equation}
+% \begin{equation}\label{eq:transition}
+%   \begin{split}
+%   F(t, 1, x)
+%     &
+%       % = P(\tilde{T} \leq t, \tilde{D}=1 \mid X=x)
+%     = \int_0^t \exp{{\{-\Lambda_{1}(s \mid x)-\Lambda_{2}(s \mid x) - \Gamma(s \mid x)\} }}  \Lambda_{1}(\diff s \mid x),
+%   \\
+%   F(t, 2, x)
+%     &
+%       % = P(\tilde{T} \leq t, \tilde{D}=2 \mid X=x)
+%       = \int_0^t \exp{\{-\Lambda_{1}(s \mid x)-\Lambda_{2}(s \mid x) - \Gamma(s \mid x)\} }  \Lambda_{2}(\diff s \mid x),
+%     \\
+%     F(t, -1, x)
+%     &
+%       % = P(\tilde{T} \leq t, \tilde{D}=0 \mid X=x)
+%       = \int_0^t \exp{\{-\Lambda_{1}(s \mid x)-\Lambda_{2}(s \mid x) - \Gamma(s \mid x)\} }  \Gamma(\diff s \mid x),
+%     \\
+%     F(t, 0, x)
+%     &
+%       = P(\tilde{T} > t \mid X= x)
+%       = 1- F(t, 1, x) - F(t, 2, x)- F(t, -1, x).
+%   \end{split}
+% \end{equation}
 The state learner requires three libraries of learners (c.f., Section \ref{sec:super-learning}),
@@
   &= \{ \phi_{a_1,a_2, b} : a_1 \in \mathcal{A}_1, a_2 \in \mathcal{A}_2, b \in \mathcal{B}\},
-    \intertext{where in correspondence with  the relations in equation \eqref{eq:transition},} 
-    \phi_{a_1,a_2, b}(\data_n)(t,1,x) &= \int_0^t \exp{\{-a_1(\data_n)(s \mid x)-a_2(\data_n)(s \mid x) - b(\data_n)(s \mid x)\} }  a_1(\data_n)(\diff s \mid x),\\
-  \phi_{a_1,a_2, b}(\data_n)(t,2,x) &= \int_0^t \exp{\{-a_1(\data_n)(s \mid x)-a_2(\data_n)(s \mid x) - b(\data_n)(s \mid x)\} }  a_2(\data_n)(\diff s \mid x),\\
-  \phi_{a_1,a_2, b}(\data_n)(t,-1,x) &= \int_0^t \exp{\{-a_1(\data_n)(s \mid x)-a_2(\data_n)(s \mid x) - b(\data_n)(s \mid x)\} }  b(\data_n)(\diff s \mid x),\\
-  \phi_{a_1,a_2, b}(\data_n)(t,0,x) &= 1
-                                      - \phi_{a_1,a_2, b}(\data_n)(t,1,x)
-                                      - \phi_{a_1,a_2, b}(\data_n)(t,2,x)
-    -\phi_{a_1,a_2, b}(\data_n)(t,-1,x)
+    \intertext{where in correspondence with  the relations in equation
+    \eqref{eq:transition},}
+    \phi_{a_1,a_2, b}(\data_n)(t,0,x)
+  &= \exp{\{-a_1(\data_n)(s \mid x)-a_2(\data_n)(s \mid x) - b(\data_n)(s \mid
+    x)\} },
+  \\
+  \phi_{a_1,a_2, b}(\data_n)(t,1,x)
+  &= \int_0^t
+    \phi_{a_1,a_2, b}(\data_n)(s-,0,x)  a_1(\data_n)(\diff s \mid x),
+  \\
+  \phi_{a_1,a_2, b}(\data_n)(t,2,x)
+  &= \int_0^t\phi_{a_1,a_2, b}(\data_n)(s-,0,x)  a_2(\data_n)(\diff s \mid x),
+  \\
+  \phi_{a_1,a_2, b}(\data_n)(t,-1,x)
+  &= \int_0^t \phi_{a_1,a_2, b}(\data_n)(s-,0,x)  b(\data_n)(\diff s \mid x).
 \end{align*}
+% \begin{align*}
+%   \mathcal{F}(\mathcal{A}_1, \mathcal{A}_2, \mathcal{B})
+%   &= \{ \phi_{a_1,a_2, b} : a_1 \in \mathcal{A}_1, a_2 \in \mathcal{A}_2, b \in \mathcal{B}\},
+%     \intertext{where in correspondence with  the relations in equation \eqref{eq:transition},} 
+%     \phi_{a_1,a_2, b}(\data_n)(t,1,x) &= \int_0^t \exp{\{-a_1(\data_n)(s \mid x)-a_2(\data_n)(s \mid x) - b(\data_n)(s \mid x)\} }  a_1(\data_n)(\diff s \mid x),\\
+%   \phi_{a_1,a_2, b}(\data_n)(t,2,x) &= \int_0^t \exp{\{-a_1(\data_n)(s \mid x)-a_2(\data_n)(s \mid x) - b(\data_n)(s \mid x)\} }  a_2(\data_n)(\diff s \mid x),\\
+%   \phi_{a_1,a_2, b}(\data_n)(t,-1,x) &= \int_0^t \exp{\{-a_1(\data_n)(s \mid x)-a_2(\data_n)(s \mid x) - b(\data_n)(s \mid x)\} }  b(\data_n)(\diff s \mid x),\\
+%   \phi_{a_1,a_2, b}(\data_n)(t,0,x) &= 1
+%                                       - \phi_{a_1,a_2, b}(\data_n)(t,1,x)
+%                                       - \phi_{a_1,a_2, b}(\data_n)(t,2,x)
+%     -\phi_{a_1,a_2, b}(\data_n)(t,-1,x)
+% \end{align*}
 To evaluate how well a function \( F \) predicts the observed
@@
       \E_{P_0}{\left[ \Vert \tilde{\phi}_n(\data_n^{-k}) - F_0 \Vert_{P_0}^2 \right]}
-    % \\
-    % & \quad
+    \\
+    & \quad
       + (1+ \delta) 16   K \tau
@@
   \label{cor:asymp-cons}
-  Assume that \( |\mathcal{F}_n| = \bigO(n^q)\), for some \( q \in \N \)
-  and that there exists a sequence \( \phi_n \in \mathcal{F}_n \),
-  \( n \in \N \), such that
+  Assume that \( |\mathcal{F}_n| = \bigO(n^q)\), for some \( q \in \N \) and
+  that there exists a sequence \( \phi_n \in \mathcal{F}_n \), \( n \in \N \),
+  such that
   \( \E_{P_0}{\left[ \Vert \phi_n(\data_n^{-k}) - F_{0} \Vert_{P_0}^2 \right]} =
   \bigO(n^{-\alpha}) \), for some \( \alpha\leq 1 \).
-  \begin{enumerate}[1.]
+  \begin{enumerate}[label=(\alph*)]
   \item If $\alpha=1$ then
@@
       g(\tau, A, W; \Lambda_1, \Lambda_2)
+    \\
+    & \quad \quad \times
       \int_0^{\tau}
@@
       g(t, A, W; \Lambda_1, \Lambda_2)
+    \\
+    & \quad \quad \times
       \exp{\{{[\Gamma+\Lambda_1 + \Lambda_2](t- \mid A, W)}\}}
@@
 
-\begin{appendices}
+\appendix
 
@@
 \begin{align*}
-  \E_{P_0}{\left[ R_0(\hat{\phi}_n(\data_n^{-k})) \right]}
-  \leq
-  &(1+2\delta)\E_{P_0}{\left[ R_0(\tilde{\phi}_n(\data_n^{-k})) \right]}
-  \\
-  & \quad + (1+\delta) \frac{16 K}{n}
+  & \E_{P_0}{\left[ R_0(\hat{\phi}_n(\data_n^{-k})) \right]}
+  \\
+  &  \quad \leq
+    (1+2\delta)\E_{P_0}{\left[ R_0(\tilde{\phi}_n(\data_n^{-k})) \right]}
+  \\
+  & \qquad + (1+\delta) \frac{16 K}{n}
     \log(1 + |\mathcal{F}_n|)\sup_{F \in \mathcal{H}_{\mathcal{P}}}
@@
 
-\end{appendices}
-
-\bibliographystyle{abbrvnat}
 \bibliography{bib.bib}

### 2024-05-27 tex/journal-version/manuscript.tex
nicely shortened intro

@@
 
-Medical decision making is traditionally based on the discrepancy
-between intuitive judgment and the law of chance
-\citep{redelmeier1995probability}. We are concerned with the
-statistical models that provide probabilistic predictions about the
-medical future of a patient. We have two different applications of
-these models in mind. One is as a stand-alone tool where the model
-provides predictions about the future, such as the predicted risk of
-adverse outcomes for a newly diagnosed patient. The other is as input
-to an algorithm which estimates a causal target parameter such as an
-average treatment effect and depends on estimates of nuisance
-parameters such as the conditional probability of being uncensored.
-An example of the latter is the targeted minimum loss based estimator
-\citep{rytgaard2022targeted}. In both applications, the models have to
-be learned based on available data from previous patients.
-
+% Medical decision making is traditionally based on the discrepancy
+% between intuitive judgment and the law of chance
+% \citep{redelmeier1995probability}.
+% We are concerned with the statistical models that provide
+% probabilistic predictions about the medical future of a patient. We
+% have two different applications of these models in mind. One is as a
+% stand-alone tool which predicts the likelihood of medical outcomes.
+% The other is as nuisance parameter estimate
+% input to an algorithm which estimates a causal target
+% parameter such as an average treatment effect and depends on estimates
+% of nuisance parameters such as the conditional probability of being
+% uncensored.  An example of the latter is the targeted minimum loss
+% based estimator \citep{rytgaard2022targeted}. In both applications,
+% the models have to be learned based on available data from previous
+% patients.
 A super learner is a machine learning algorithm that combines a finite
@@
 % infinite hold-out sample.
-The work presented in this article is motivated by the choice of the
-loss function for super learning in survival analysis. Existing super
-learner algorithms for right-censored data use partial log-likelihood
-loss or inverse probability of censoring weighted loss
-\citep{polley2011-sl-cens,keles2004asymptotically,golmakani2020super,gonzalez2021stacked,westling2021inference}.
-Unfortunately, the use of the partial log-likelihood loss restricts
-the class of learners and excludes for example simple Kaplan-Meier
-based learners and also more complex random survival forest
-algorithms. For this reason \cite{golmakani2020super} restrict their
-learners to Cox proportional hazard models. A lesser known fact is
-that a super learner constructed with the negative partial
-log-likelihood loss implicitly depends on the censoring distribution
-\citep{hjort1992inference,whitney2019comment}. Inverse probability of
-censoring weighted loss has been studied for decades
-\citep{graf1999assessment, hothorn2006survival,
-  gerds2006consistent}. A disadvantage of inverse probability of
-censoring weighted loss estimators is that they require a
-pre-specified model for the censoring distribution. Recently
-\cite{westling2021inference} proposed a procedure which iterates
-between super learning of the censoring distribution and super
-learning of the event time distribution. In this article, we pursue a
-different route.
 
@@
 contain a consistent learner and prove a finite sample oracle
-inequality. The result of the state learner algorithm could be a
-medical risk prediction model \citep{gerds2021medical} which predicts
-the probability of an event based on covariates in the presence of
+inequality.
+
+Machine learning based on right-censored data commonly uses the
+partial log-likelihood as a loss function
+\citep[e.g.,][]{li2016regularized,yao2017deep,lee2018deephit,katzman2018deepsurv,gensheimer2019scalable,lee2021boosted,kvamme2021continuous}.
+However, this loss function does not work well with data splitting
+(cross-validation) because the partial log-likelihood loss assigns an
+infinite value when applied to a test set time point which does not
+occur in the learning set as soon as the learner predicts piece-wise
+constant cumulative hazard functions. This is the case for prominent
+survival learners including the Kaplan-Meier estimator, the random
+survival forest, and the semi-parametric Cox regression model.  When a
+proportional hazards model is assumed, the baseline hazard function
+can be profiled out of the likelihood \citep{cox1972regression}.  The
+cross-validated partial log-likelihood loss \citep{verweij1993cross}
+has therefore been suggested as a loss function for super learning
+which however restricts the library of learners to include only Cox
+proportional hazards models \citep{golmakani2020super}.
+% A lesser
+% known fact is that a super learner constructed with the partial
+% log-likelihood loss implicitly depends on the censoring distribution
+% \citep{hjort1992inference,whitney2019comment}.
+
+Alternative approaches for super learning with right-censored data use
+an inverse probability of censoring weighted (IPCW) loss function
+\citep{graf1999assessment,van2003unicv,molinaro2004tree,keles2004asymptotically,hothorn2006survival,gerds2006consistent,gonzalez2021stacked},
+% An IPCW loss function is attractive because the associated risk does
+% not depend on the censoring distribution but describes a feature of
+% the population of interested. Similar results can be obtained using
+censoring unbiased transformations
+\citep{fan1996local,steingrimsson2019censoring}, or pseudo-values
+\citep{andersen2003generalised,mogensen2013random,sachs2019ensemble}.
+All these methods rely on an estimator of the censoring distribution,
+and their drawback is that this estimator has to be pre-specified.
+% When the data-generating mechanism is complex and not
+% well-understood, pre-specification of the censoring distribution is a
+% challenge.
+% Inverse probability of censoring weighted loss has been studied for
+% decades \citep{graf1999assessment, hothorn2006survival,
+  % gerds2006consistent}. A disadvantage of inverse probability of
+% censoring weighted loss estimators is that they require a
+% pre-specified model for the censoring distribution.
+% Recently \cite{westling2021inference} proposed a procedure which
+% iterates between super learning of the censoring distribution and
+% super learning of the event time distribution.
+An approach which avoids a pre-specified censoring model was proposed
+independently by \cite{han2021inverse} and
+\cite{westling2021inference}. In both articles, the authors suggest to
+iterate between learning of the outcome model and learning of the
+censoring model using IPCW loss functions. However, no general
+theoretical guarantees seem to exist for this procedure, and it has
+not yet been extended to the situation with competing risks.
+
+The result of the state learner algorithm could be a medical risk
+prediction model \citep{gerds2021medical} which predicts the
+probability of an event based on covariates in the presence of
 competing risks. The other application is in targeted learning where
@@
 state learner with prostate cancer data
-\citep{kattan2000pretreatment}.
-
-\subsection{Existing methods}
-\label{sec:relat-liter-exist}
-
-Machine learning based on right-censored data commonly uses the
-negative partial log-likelihood as a loss function
-\citep[e.g.,][]{li2016regularized,yao2017deep,lee2018deephit,katzman2018deepsurv,gensheimer2019scalable,lee2021boosted,kvamme2021continuous}.
-This loss function is however not perfectly suited for super learning
-because it cannot be applied to prominent survival learners. For
-example, the Kaplan-Meier estimator, the random survival forest, and
-the semi-parametric Cox regression model provide cumulative hazard
-functions that are piece-wise constant in the time argument. Suppose
-these methods were applied to a learning data set and the aim is to
-evaluate the partial log-likelihood loss in an independent test data
-set. Suppose further that the test data set contains event times that
-were not also seen in the learning data set, as it would naturally
-happen when the events are observed in continuous time. The problem is
-that these methods assign zero probability to all the event times in
-the test set which were not also contained in the training data set.
-Hence the partial log-likelihood loss in the test set would be
-infinity (the logarithm of zero is infinity).
-
-When a proportional hazards model is assumed, the baseline hazard
-function can be profiled out of the likelihood \citep{cox1972regression}.
-The cross-validated partial log-likelihood loss  \citep{verweij1993cross} has therefore been
-suggested as a loss function for super learning which however
-restricts the library of learners to include only Cox proportional
-hazards models \citep{golmakani2020super}.
-
-Another approach for super learning with right-censored data is to use an
-inverse probability of censoring weighted (IPCW) loss function
-\citep{graf1999assessment,van2003unicv,molinaro2004tree,keles2004asymptotically,hothorn2006survival,gerds2006consistent,gonzalez2021stacked}.
-An IPCW loss function is attractive because the associated risk does not depend
-on the censoring distribution but describes a feature of the population of
-interested. Similar results can be obtained using censoring unbiased
-transformations \citep{fan1996local,steingrimsson2019censoring} or pseudo-values
-\citep{andersen2003generalised,mogensen2013random,sachs2019ensemble}. All these
-methods rely on an estimator of the censoring distribution, and their drawback
-is that this estimator has to be pre-specified. When the data-generating
-mechanism is complex and not well-understood, pre-specification of the censoring
-distribution is a challenge.
-
-An attempt to avoid a pre-specified censoring model was proposed independently
-by \cite{han2021inverse} and \cite{westling2021inference}. In both articles, the
-authors suggest to iterate between learning of the outcome model and learning of
-the censoring model using IPCW loss functions. However, no general theoretical
-guarantees seem to exist for this procedure, and it has not yet been extended to
-the situation with competing risks.
-
-\subsection{Organization of article}
-
-We introduce our notation and framework in
-Section~\ref{sec:framework}. In Section~\ref{sec:super-learning} we
-define general super learning for right-censored data.
-% , and in
-% Section~\ref{sec:relat-liter-exist} we review existing methods.
-Section~\ref{sec:super-learner-simple} introduces the state learner,
-and Section~\ref{sec:theor-results-prop} provides theoretical
-guarantees. In Section~\ref{sec:targeted-learning} we discuss the use
-of the state learner in the context of targeted learning. We report
-results of our numerical experiments in Section~\ref{sec:numer-exper}
-and analyze a prostate cancer data set in
-Section~\ref{sec:real-data-appl}.  Section~\ref{sec:discussion}
+\citep{kattan2000pretreatment}. We introduce our notation and
+framework in Section~\ref{sec:framework}. In
+Section~\ref{sec:super-learning} we define general super learning for
+right-censored data. Section~\ref{sec:super-learner-simple} introduces
+the state learner, and Section~\ref{sec:theor-results-prop} provides
+theoretical guarantees. In Section~\ref{sec:targeted-learning} we
+discuss the use of the state learner in the context of targeted
+learning. We report results of our numerical experiments in
+Section~\ref{sec:numer-exper} and analyze a prostate cancer data set
+in Section~\ref{sec:real-data-appl}.  Section~\ref{sec:discussion}
 contains a discussion of the merits and limitations of our
 proposal. Appendices~\ref{sec:proof-proposition}
-and~\ref{sec:state-learner-with} contain proofs.
-An implementation of the state learner is available at
-\url{https://github.com/amnudn/statelearner} along with codes for reproducing
-our numerical experiments.
+and~\ref{sec:state-learner-with} contain proofs.  An implementation of
+the state learner is available at
+\url{https://github.com/amnudn/statelearner} along with codes for
+reproducing our numerical experiments.
 
@@
 At time \(t\), we observe that each individual is in one of four mutually
-exclusive states (Figure \ref{fig:multi-state-process}).
-
-\begin{figure}[h]
-  \centering %
-  \includeFigCond[width=.5\textwidth]{figure-multi-state-process.pdf}
-  \caption{Illustration of the multi-state process \(\eta\) used by
-    the state learner. Note that `censored' is a state, hence the
-    process is always observed at any time.}
-  \label{fig:multi-state-process}
-\end{figure}
-
+exclusive states. % (Figure \ref{fig:multi-state-process}).
+% \begin{figure}[h]
+  % \centering %
+  % \includeFigCond[width=.5\textwidth]{figure-multi-state-process.pdf}
+  % \caption{Illustration of the multi-state process \(\eta\) used by
+    % the state learner. Note that `censored' is a state, hence the
+    % process is always observed at any time.}
+  % \label{fig:multi-state-process}
+% \end{figure}
 The conditional distribution of the process \( \eta(t) \) given
@@
 cumulative hazard functions of tumor recurrence, death without tumor
-recurrence, and censoring.  The libraries of learners include the five
-learners described in Table~\ref{tab:zel-library}.  We use the same
-set of learners to learn the cumulative hazard function of tumor
-recurrence \( \Lambda_1 \), the cumulative hazard function of death
-without tumor recurrence \( \Lambda_2 \), and the cumulative hazard
-function of the conditional censoring distribution $\Gamma$. 
-We then use the highest ranked combination
-of learners and apply formula~\eqref{eq:one-step-comp-ate}.
-
-\begin{table}
-  \caption{\label{tab:zel-library}Overview of the five learners used by the
-    state learner when applied to the prostate cancer data set. The Kaplan-Meier
-    estimator was fitted using the package \texttt{prodlim}
-    \citep{Gerds_2019prodlim}. All Cox models included all five covariates in
-    the model and were fitted using the package \texttt{survival}
-    \citep{survival-package}. All penalised Cox models included all five
-    covariates as linear predictors and were fitted using the package
-    \texttt{glmnet} \citep{glmnet-cox,glmnet-glm}. The random forest was fitted
-    with the package \texttt{randomForestSRC} \citep{rfsrc-paclage}.}
-  \centering
-  \begin{tabular}{lll}
-    \toprule
-    Family & Model & Description \\
-    \midrule
-    Marginal & \texttt{KM} & The Kaplan-Meier estimator \\ 
-    Cox & \texttt{Cox} & All five covariates included with additive effects \\
-           & \( \texttt{Cox strata CS} \)  & Cox model stratified on CS  \\
-           % & \( \texttt{Cox strata HT} \)  & Cox model stratified on HT \\
-           % & \( \texttt{Cox spline} \)  & PSA and RD modeled with splines \\ 
-    Penalised Cox & \texttt{Lasso} & Cox model with \( L_1 \)-norm penalty   \\
-           % & \texttt{Ridge} & Cox model with \( L_2 \)-norm penalty \\
-           & \texttt{Elastic} & Cox model with \( L_1 \)- and \( L_2 \)-norm penalty \\
-    Random forest & \texttt{RF} & Random forest with default settings \\
-    \bottomrule 
-  \end{tabular}
-\end{table}
-
-This gives a library consisting of \( 5^3 = 125 \) learners for the conditional
-state occupation probability function \( F \) defined in
-equation~(\ref{eq:F-def}). We use five folds for training and testing the
-models, and we repeat training and evaluation five times with different splits.
-The integrated Brier score (defined in Section~\ref{sec:super-learner-simple})
-for all learners are shown in Figure~\ref{fig:zelefski-real}, and the top 10
-combinations of learners are displayed in Table~\ref{tab:zelefski-real}. We see
-that the prediction performance is mostly affected by the choice of learner for
-the censoring distribution. Several combinations of learners give similar
-performance as measured by the integrated Brier score, as long as a random
-forest is used to model the censoring distribution.
+recurrence, and censoring.  The libraries of learners each include
+five learners: a Nelson-Aalen, three Cox regression models
+(unpenalized, Lasso, Elastic net) each including additive effects of
+the 5 covariates (Section~\ref{sec:numer-exper}), and a random
+survival forest. We use the same set of learners to learn the
+cumulative hazard function of tumor recurrence \( \Lambda_1 \), the
+cumulative hazard function of death without tumor recurrence
+\( \Lambda_2 \), and the cumulative hazard function of the conditional
+censoring distribution $\Gamma$.  We then use the highest ranked
+combination of learners and apply
+formula~\eqref{eq:one-step-comp-ate}.
+
+% \begin{table}
+  % \caption{\label{tab:zel-library}Overview of the five learners used by the
+    % state learner when applied to the prostate cancer data set. The Kaplan-Meier
+    % estimator was fitted using the package \texttt{prodlim}
+    % \citep{Gerds_2019prodlim}. All Cox models included all five covariates in
+    % the model and were fitted using the package \texttt{survival}
+    % \citep{survival-package}. All penalised Cox models included all five
+    % covariates as linear predictors and were fitted using the package
+    % \texttt{glmnet} \citep{glmnet-cox,glmnet-glm}. The random forest was fitted
+    % with the package \texttt{randomForestSRC} \citep{rfsrc-paclage}.}
+  % \centering
+  % \begin{tabular}{lll}
+    % \toprule
+    % Family & Model & Description \\
+    % \midrule
+    % Marginal & \texttt{KM} & The Kaplan-Meier estimator \\ 
+    % Cox & \texttt{Cox} & All five covariates included with additive effects \\
+           % & \( \texttt{Cox strata CS} \)  & Cox model stratified on CS  \\
+    % Penalised Cox & \texttt{Lasso} & Cox model with \( L_1 \)-norm penalty   \\
+           % & \texttt{Elastic} & Cox model with \( L_1 \)- and \( L_2 \)-norm penalty \\
+    % Random forest & \texttt{RF} & Random forest with default settings \\
+    % \bottomrule 
+  % \end{tabular}
+% \end{table}
+
+This gives a library consisting of \( 5^3 = 125 \) learners for the
+conditional state occupation probability function \( F \) defined in
+equation~(\ref{eq:F-def}). We use five folds for training and testing
+the models, and we repeat training and evaluation five times with
+different splits.  The integrated Brier score (defined in
+Section~\ref{sec:super-learner-simple}) for all learners are shown in
+Figure~\ref{fig:zelefski-real}. We see that the prediction
+performance is mostly affected by the choice of learner for the
+censoring distribution. Several combinations of learners give similar
+performance as measured by the integrated Brier score, as long as a
+random forest is used to model the censoring distribution.
 
@@
     learners of \( \Lambda_1 \), \( \Lambda_2 \), and $\Gamma$ as `Tumor
-    learner', `Death learner', and `Censoring learner', respectively.}
+    learner', `Mortality learner', and `Censoring learner', respectively.}
   \label{fig:zelefski-real}
@@
 
-\begin{table}
-  \caption{\label{tab:zelefski-real}The 10 best performing models in terms of integrated Brier score. The
-    reported standard errors are based on five repetitions using different
-    splits. The models are described in Table~\ref{tab:zel-library}. We refer to
-    learners of \( \Lambda_1 \), \( \Lambda_2 \), and $\Gamma$ as `Tumor
-    learner', `Death learner', and `Censoring learner', respectively.}
-  \centering
-  \input{./real-data-tab.tex}
-\end{table}
+% \begin{table}
+  % \caption{\label{tab:zelefski-real}The 10 best performing models in terms of integrated Brier score. The
+    % reported standard errors are based on five repetitions using different
+    % splits. The models are described in Table~\ref{tab:zel-library}. We refer to
+    % learners of \( \Lambda_1 \), \( \Lambda_2 \), and $\Gamma$ as `Tumor
+    % learner', `Mortality learner', and `Censoring learner', respectively.}
+  % \centering
+  % \input{./real-data-tab.tex}
+% \end{table}
 

### 2024-05-27 tex/journal-version/manuscript.tex
reformatting

@@
 %% Packages
-\usepackage{amsthm, dsfont, amssymb,soul,xcolor,
-  % amsfonts,
-  enumitem,
-  amsmath,graphicx,verbatim,pgfplots,tikz,prodint,booktabs}
+\usepackage{amsthm, dsfont, amssymb,soul,xcolor, enumitem, amsmath,verbatim}
 
@@
 
-% % Handling new lines
-% \setlength{\parskip}{1em}
-% \setlength{\parindent}{0em}
-
-% % Handling space after sections
-% \usepackage{titlesec}
-% \titlespacing*{\section}{0em}{2em}{0em}
-% \titlespacing*{\subsection}{0em}{2em}{0em}
-% \titlespacing*{\subsubsection}{0em}{2em}{0em}
-
-% % No spacing after in start of list
-% \setlist[itemize]{topsep=0pt}
-% \setlist[enumerate]{topsep=0pt}
-
 \definecolor{linkcolor}{rgb}{0, 0, 0.54}
@@
 \newtheorem{theorem}{Theorem}%  meant for continuous numbers
-%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
-%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
-% \newtheorem{proposition}[theorem]{Proposition}%
-% to get separate numbers for theorem and proposition etc.:
 \newtheorem{proposition}{Proposition}
@@
 
-% \usepackage[tablesonly,heads,nolists]{endfloat}
-
-%% change enumerate labels
-% \renewcommand{\labelenumi}{\normalfont (\alph{enumi})}
-
-%% Font size etc
+%% Line spacing
 \linespread{1.65}
 
-\usepackage{lineno}
-\linenumbers
+% \usepackage{lineno}
+% \linenumbers
 
@@
 \RequirePackage{graphicx}
-% \RequirePackage{caption}
-% \RequirePackage{array}
-% \RequirePackage{color}
-% \RequirePackage{xcolor}
-% \RequirePackage{amssymb}
-% \RequirePackage{flushend}
-% \RequirePackage{stfloats}
-% \RequirePackage[figuresright]{rotating}
-% \RequirePackage{chngpage}
-% \RequirePackage{totcount}
-% \RequirePackage{fix-cm}
+\usepackage[font=small]{caption}
+
 \newcommand{\includeFigCond}[2][]{
@@
 \usepackage{ifthen} %% Empirical process with default argument
-% \newcommand{\G}[1][]{%
-%    \ifthenelse{ \equal{#1}{} }
-%       {\ensuremath{\mathbb{G}_n}}
-%       {\ensuremath{\mathbb{G}_{#1}}}
-% }
-% New version:
 \newcommand{\G}[2][n]{
@@
 \newcommand{\empmeas}{\ensuremath{\mathbb{P}_n}} % empirical measure
-
 \newcommand{\data}{\ensuremath{\mathcal{D}}}
@@
 %% Title
-\title{The state learner -- a super learner for right-censored data}
+\title{\Large The state learner -- a super learner for right-censored data}
+\date{\small \today}
 
 \usepackage{authblk}
-\author[1]{Anders Munch}
-\author[1]{Thomas~A.~Gerds}
-\affil[1]{Section of Biostatistics, University of Copenhagen}
+\author[1,*]{\normalsize Anders Munch}
+\author[1]{\normalsize Thomas~A.~Gerds}
+\affil[1]{\normalsize Section of Biostatistics, University of Copenhagen}
+\affil[*]{\normalsize Address for correspondence: \href{email:email-id.com}{a.munch@sund.ku.dk}}
+
+\usepackage{fancyhdr}
+\fancyhf{}
+\fancyhead[l]{\textit{Munch et al.}}
+\fancyhead[r]{\textit{The state learner}}
+\fancyfoot[c]{\thepage}
+\renewcommand\headrulewidth{0pt}
+\pagestyle{fancy}
 
 \begin{document}
-
-% \authormark{Munch et al.}
-
-% \address[1]{\orgdiv{Section of Biostatistics}, \orgname{University of
-%     Copenhagen}, \orgaddress{\street{Østerfarigmagsgade 5}, \postcode{2100}, \state{Copenhagen}, \country{Denmark}}}
-
-% \corresp[$\ast$]{Address for correspondence: \href{email:email-id.com}{a.munch@sund.ku.dk}}
-
-% \received{Date}{0}{Year}
-% \revised{Date}{0}{Year}
-% \accepted{Date}{0}{Year}
-
-% \keywords{Competing risks, cross-validation, loss based estimation,
-%   right-censored data, super learner}
 
@@
   learner is a machine learning algorithm which combines a library of prediction
-  models into a meta learner based on cross-validated loss. In right censored
+  models into a meta learner based on cross-validated loss. In right-censored
   data, the choice of the loss function and the estimation of the expected loss
@@
 \end{abstract}
+
+% \keywords{Competing risks, cross-validation, loss based estimation,
+% right-censored data, super learner}
+
+\textit{{\small \textbf{Keywords:} Competing risks, cross-validation, loss based
+    estimation, right-censored data, super learner}}
+
+\pagebreak
 
@@
 function underlying the state learner operates on the coarsened
-data. In the right censored survival setting the coarsened data
+data. In the right-censored survival setting the coarsened data
 consist of the minimum and the order of the censoring time and the
@@
 competing risks, and it does not require a single pre-specified
-estimator of the conditional censoring distribution. To analyze the
+estimator of the conditional censoring distribution. To analyse the
 theoretical properties of the state learner we focus on the discrete
@@
 
-The result of the state learner algorithm could be a medical risk
-prediction model \citep{gerds2021medical} which predicts the
-probability of an event based on covariates in the presence of
-competing risks. The other application is in targeted learning where
-conditional event probabilities occur as high-dimensional nuisance
-parameters which need to be estimated at a certain rate
-\citep{van2011targeted, rytgaard2021estimation, rytgaard2022targeted}.
-For the asymptotic bias term of targeted estimator, which uses the
-state learner to estimate nuisance parameters, we show that a second
-order product structure holds. We illustrate both applications of the
-state learner with prostate cancer data
-\citep{kattan2000pretreatment}. We introduce our notation and
-framework in Section~\ref{sec:framework}. In
-Section~\ref{sec:super-learning} we define general super learning for
-right-censored data. Section~\ref{sec:super-learner-simple} introduces
-the state learner, and Section~\ref{sec:theor-results-prop} provides
-theoretical guarantees. In Section~\ref{sec:targeted-learning} we
-discuss the use of the state learner in the context of targeted
-learning. We report results of our numerical experiments in
-Section~\ref{sec:numer-exper} and analyze a prostate cancer data set
-in Section~\ref{sec:real-data-appl}.  Section~\ref{sec:discussion}
-contains a discussion of the merits and limitations of our
-proposal. Appendices~\ref{sec:proof-proposition}
-and~\ref{sec:state-learner-with} contain proofs.  An implementation of
-the state learner is available at
-\url{https://github.com/amnudn/statelearner} along with codes for
-reproducing our numerical experiments.
+The state learner algorithm can output a medical risk prediction model
+\citep{gerds2021medical} which predicts the probability of an event based on
+covariates in the presence of competing risks. The other application is in
+targeted learning where conditional event probabilities occur as
+high-dimensional nuisance parameters which need to be estimated at a certain
+rate \citep{van2011targeted, rytgaard2021estimation, rytgaard2022targeted}. For
+the asymptotic bias term of targeted estimator, which uses the state learner to
+estimate nuisance parameters, we show that a second order product structure
+holds. We illustrate both applications of the state learner with prostate cancer
+data \citep{kattan2000pretreatment}.
+
+We introduce our notation and framework in
+Section~\ref{sec:framework}. In Section~\ref{sec:super-learning} we define
+general super learning for right-censored data.
+Section~\ref{sec:super-learner-simple} introduces the state learner, and
+Section~\ref{sec:theor-results-prop} provides theoretical guarantees. In
+Section~\ref{sec:targeted-learning} we discuss the use of the state learner in
+the context of targeted learning. We report results of our numerical experiments
+in Section~\ref{sec:numer-exper} and analyse a prostate cancer data set in
+Section~\ref{sec:real-data-appl}. Section~\ref{sec:discussion} contains a
+discussion of the merits and limitations of our proposal.
+Appendices~\ref{sec:proof-proposition} and~\ref{sec:state-learner-with} contain
+proofs. An implementation of the state learner is available at
+\url{https://github.com/amnudn/statelearner} along with codes for reproducing
+our numerical experiments.
 
@@
 At time \(t\), we observe that each individual is in one of four mutually
-exclusive states. % (Figure \ref{fig:multi-state-process}).
+exclusive states: \( 0 \), \( 1 \), \( 2 \), or \( -1 \).
+% (Figure \ref{fig:multi-state-process}).
 % \begin{figure}[h]
@@
 The asymptotic distribution of \( \hat{\Psi}^0 \) is
-difficult to analyze due to the cross-validated model selection step
+difficult to analyse due to the cross-validated model selection step
 involved in the estimation of the nuisance parameters $\Lambda_1$ and
@@
   \mathrm{Rem}{(\hat{\Lambda}_{1,n},\hat{\Lambda}_{2,n},  \hat{\Gamma}_n, P)}
-  = \mathcal{O}_P{
+  = \bigO_P{
     \left\{
@@
 
-In this section we report results from a simulation study. In the first part we
-consider estimation of the conditional survival function. Here we compare the
-state learner to two IPCW based discrete super learners that use either the
+In this section we report results from a simulation study where we consider
+estimation of the conditional survival function. In the first part, we compare
+the state learner to two IPCW based discrete super learners that use either the
 Kaplan-Meier estimator or a Cox model to estimate the censoring probability
@@
 
-In both parts we use the same data-generating mechanism. We generate
-data according to a distribution motivated from a real data set in
-which censoring depends on the baseline covariates. We simulate data
-based on the prostate cancer study of
-\cite{kattan2000pretreatment}. The outcome of interest is the time to
-tumor recurrence, and five baseline covariates are used to predict
-outcome: prostate-specific antigen (PSA, ng/mL), Gleason score sum
-(GSS, values between 6 and 10), radiation dose (RD), hormone therapy
-(HT, yes/no) and clinical stage (CS, six values). The study was
-designed such that a patient's radiation dose depended on when the
-patient entered the study \citep{gerds2013estimating}. This in turn
-implies that the time of censoring depends on the radiation dose. The
-data were re-analyzed in \citep{gerds2013estimating} where a
-sensitivity analysis was conducted based on simulated data. Here we use the
-same simulation setup, where event and censoring times are generated
-according to parametric Cox-Weibull models estimated from the original
-data, and the covariates are generated according to either marginal
-Gaussian normal or binomial distributions estimated from the original
-data \citep[c.f.,][Section~4.6]{gerds2013estimating}. We refer to this
-simulation setting as ``dependent censoring''. We also considered a
-simulation setting where data were generated in the same way, except
-that censoring was generated completely independently. We refer to
-this simulation setting as ``independent censoring''.
+In both parts we use the same data-generating mechanism. We generate data
+according to a distribution motivated from a real data set in which censoring
+depends on the baseline covariates. We simulate data based on the prostate
+cancer study of \cite{kattan2000pretreatment}. The outcome of interest is the
+time to tumor recurrence, and five baseline covariates are used to predict
+outcome: prostate-specific antigen (PSA, ng/mL), Gleason score sum (GSS, values
+between 6 and 10), radiation dose (RD), hormone therapy (HT, yes/no) and
+clinical stage (CS, six values). The study was designed such that a patient's
+radiation dose depended on when the patient entered the study
+\citep{gerds2013estimating}. This in turn implies that the time of censoring
+depends on the radiation dose. The data were re-analysed in
+\citep{gerds2013estimating} where a sensitivity analysis was conducted based on
+simulated data. Here we use the same simulation setup, where event and censoring
+times are generated according to parametric Cox-Weibull models estimated from
+the original data, and the covariates are generated according to either marginal
+Gaussian normal or binomial distributions estimated from the original data
+\citep[c.f.,][Section~4.6]{gerds2013estimating}. We refer to this simulation
+setting as `dependent censoring'. We also considered a simulation setting where
+data were generated in the same way, except that censoring was generated
+completely independently. We refer to this simulation setting as `independent
+censoring'.
 
@@
 can be used to learn the cumulative hazard functions of the 
-outcome and the censoring distribution. This works by
+outcome and the censoring distribution. The latter works by
 training the learner on the data set \( \data_n^c \), where
@@
 horizon (36 months). The marginal risk of the event before this time horizon is
-\(\approx 24.6\)\%. Under the ``dependent censoring'' setting the marginal
+\(\approx 24.6\)\%. Under the `dependent censoring' setting the marginal
 censoring probability before the time horizon is \(\approx 61.9\)\%. Under the
-``independent censoring'' setting the marginal censoring probability before this
+`independent censoring' setting the marginal censoring probability before this
 time horizon is \( \approx 38.7 \)\%.
@@
   \centering %
-  \includeFigCond[width=.8\linewidth]{experiment-fig-sl-ipcw.pdf}
+  \includeFigCond[width=1\linewidth]{experiment-fig-sl-ipcw.pdf}
   \caption[]{For the risk prediction models provided by each of the super
@@
   \centering %
-  \includeFigCond[width=.8\linewidth]{experiment-fig-sl-survSL-out.pdf}
+  \includeFigCond[width=1\linewidth]{experiment-fig-sl-survSL-out.pdf}
   \caption[]{For the risk prediction models of the outcome provided by each
@@
   \centering %
-  \includeFigCond[width=.5\linewidth]{experiment-fig-sl-survSL-cens.pdf}
+  \includeFigCond[width=.7\linewidth]{experiment-fig-sl-survSL-cens.pdf}
   \caption[]{For the risk prediction models of the censoring model provided by
@@
 
-% \begin{table}
-  % \caption{\label{tab:zel-library}Overview of the five learners used by the
-    % state learner when applied to the prostate cancer data set. The Kaplan-Meier
-    % estimator was fitted using the package \texttt{prodlim}
-    % \citep{Gerds_2019prodlim}. All Cox models included all five covariates in
-    % the model and were fitted using the package \texttt{survival}
-    % \citep{survival-package}. All penalised Cox models included all five
-    % covariates as linear predictors and were fitted using the package
-    % \texttt{glmnet} \citep{glmnet-cox,glmnet-glm}. The random forest was fitted
-    % with the package \texttt{randomForestSRC} \citep{rfsrc-paclage}.}
-  % \centering
-  % \begin{tabular}{lll}
-    % \toprule
-    % Family & Model & Description \\
-    % \midrule
-    % Marginal & \texttt{KM} & The Kaplan-Meier estimator \\ 
-    % Cox & \texttt{Cox} & All five covariates included with additive effects \\
-           % & \( \texttt{Cox strata CS} \)  & Cox model stratified on CS  \\
-    % Penalised Cox & \texttt{Lasso} & Cox model with \( L_1 \)-norm penalty   \\
-           % & \texttt{Elastic} & Cox model with \( L_1 \)- and \( L_2 \)-norm penalty \\
-    % Random forest & \texttt{RF} & Random forest with default settings \\
-    % \bottomrule 
-  % \end{tabular}
-% \end{table}
-
 This gives a library consisting of \( 5^3 = 125 \) learners for the
@@
   \centering %
-  \includeFigCond[width=.9\linewidth]{real-data-state-learner.pdf}
+  \includeFigCond[width=1\linewidth]{real-data-state-learner.pdf}
   \caption[]{The results of applying the 125 combinations of learners to the
-    prostate cancer data set. The learners are \texttt{KM} (KM), \texttt{Cox
-      strata CS} (strata), \texttt{Lasso} (lasso), \texttt{Elastic} (elastic),
-    and \texttt{RF} (RF) as described Table~\ref{tab:zel-library}. The error
-    bars are based on five repetitions using different splits. We refer to
-    learners of \( \Lambda_1 \), \( \Lambda_2 \), and $\Gamma$ as `Tumor
-    learner', `Mortality learner', and `Censoring learner', respectively.}
+    prostate cancer data set. The error bars are based on five repetitions using
+    different splits. We refer to learners of \( \Lambda_1 \), \( \Lambda_2 \),
+    and $\Gamma$ as `Tumor learner', `Mortality learner', and `Censoring
+    learner', respectively.}
   \label{fig:zelefski-real}
 \end{figure}
-
-% \begin{table}
-  % \caption{\label{tab:zelefski-real}The 10 best performing models in terms of integrated Brier score. The
-    % reported standard errors are based on five repetitions using different
-    % splits. The models are described in Table~\ref{tab:zel-library}. We refer to
-    % learners of \( \Lambda_1 \), \( \Lambda_2 \), and $\Gamma$ as `Tumor
-    % learner', `Mortality learner', and `Censoring learner', respectively.}
-  % \centering
-  % \input{./real-data-tab.tex}
-% \end{table}
 
@@
   \centering%
-  \includeFigCond[width=.8\linewidth]{real-data-target.pdf}
+  \includeFigCond[width=.9\linewidth]{real-data-target.pdf}
   \caption[]{Estimates of the average treatment effect of hormone therapy on the
     risk of tumor recurrence and death obtained from the prostate cancer study
-    analyzed by \cite{kattan2000pretreatment}. The estimates are based on the
+    analysed by \cite{kattan2000pretreatment}. The estimates are based on the
     estimator defined in equation~(\ref{eq:one-step-comp-ate}) with pointwise
@@
   averaging. We use the relations from equation~(\ref{eq:7}) to write
+  \begingroup %
+  \allowdisplaybreaks
     \begin{align*}
@@
     \end{align*}
+    \endgroup %
     Consider the first term on the right hand side. Defining

### 2024-05-27 tex/journal-version/manuscript.tex
readme update

@@
 recurrence, and censoring.  The libraries of learners each include
-five learners: a Nelson-Aalen, three Cox regression models
+five learners: the Nelson-Aalen estimator, three Cox regression models
 (unpenalized, Lasso, Elastic net) each including additive effects of

### 2024-05-28 tex/arxiv-version/manuscript.tex
arxiv link

--- 2024-05-27 tex/journal-version/manuscript.tex
@@
-\documentclass[12pt]{article}
-\usepackage[a4paper,left=2.8cm, right=2.8cm, top=3cm, bottom=3cm]{geometry}
+\documentclass{article}
+\usepackage[a4paper,margin=3cm]{geometry}
 
@@
 
-\theoremstyle{thmstyleone}%\
+\theoremstyle{plain}%\
 \newtheorem{theorem}{Theorem}%  meant for continuous numbers
@@
 \newtheorem{corollary}{Corollary}
-\theoremstyle{thmstyletwo}%
-\newtheorem{example}{Example}%
-\newtheorem{remark}{Remark}%
-\theoremstyle{thmstylethree}%
-\newtheorem{definition}{Definition}
-
-%% Line spacing
-\linespread{1.65}
-
-% \usepackage{lineno}
-% \linenumbers
-
-%% Only for comments:
-\usepackage[author=]{fixme}
-\fxusetheme{color}
-\definecolor{fxtarget}{rgb}{.5,.5,.5}
-\definecolor{fxnote}{rgb}{.5,.5,.5}
-\fxsetup{status=draft}
+
+% Handling new lines
+\setlength{\parskip}{1em}
+\setlength{\parindent}{0em}
+
+% Handling space after sections
+\usepackage{titlesec}
+\titlespacing*{\section}{0em}{2em}{0em}
+\titlespacing*{\subsection}{0em}{2em}{0em}
+\titlespacing*{\subsubsection}{0em}{2em}{0em}
+
+% No spacing after in start of list
+\setlist[itemize]{topsep=0pt}
+\setlist[enumerate]{topsep=0pt}
 
@@
 %% Title
-\title{\Large The state learner -- a super learner for right-censored data}
-\date{\small \today}
+\title{The state learner -- a super learner for right-censored data}
+\date{\today}
 
 \usepackage{authblk}
-\author[1,*]{\normalsize Anders Munch}
-\author[1]{\normalsize Thomas~A.~Gerds}
-\affil[1]{\normalsize Section of Biostatistics, University of Copenhagen}
-\affil[*]{\normalsize Address for correspondence: \href{email:email-id.com}{a.munch@sund.ku.dk}}
+\author[1,*]{Anders Munch}
+\author[1]{Thomas~A.~Gerds}
+\affil[1]{Section of Biostatistics, University of Copenhagen}
 
@@
 
-% \keywords{Competing risks, cross-validation, loss based estimation,
-% right-censored data, super learner}
-
 \textit{{\small \textbf{Keywords:} Competing risks, cross-validation, loss based
@@
 
-\pagebreak
-
 \section{Introduction}
@@
 
-% Medical decision making is traditionally based on the discrepancy
-% between intuitive judgment and the law of chance
-% \citep{redelmeier1995probability}.
-% We are concerned with the statistical models that provide
-% probabilistic predictions about the medical future of a patient. We
-% have two different applications of these models in mind. One is as a
-% stand-alone tool which predicts the likelihood of medical outcomes.
-% The other is as nuisance parameter estimate
-% input to an algorithm which estimates a causal target
-% parameter such as an average treatment effect and depends on estimates
-% of nuisance parameters such as the conditional probability of being
-% uncensored.  An example of the latter is the targeted minimum loss
-% based estimator \citep{rytgaard2022targeted}. In both applications,
-% the models have to be learned based on available data from previous
-% patients.
 A super learner is a machine learning algorithm that combines a finite
@@
 and random survival forests \citep[][Section 8.4]{gerds2021medical}.
-% For targeted learning of a low dimensional
-% target parameter in presence of a high-dimensional nuisance parameter,
-% a super learner will usually also include the highly adaptive lasso
-% \citep{van2011targeted, benkeser2016highly,van2017generally}.
 While the general idea of combining strong learners based on cross-validation
@@
 \citep{van2003unicv,vaart2006oracle}.
-% The super learner is guaranteed to perform
-% almost as well as the model which minimises the expected performance, i.e., the
-% model we would select if we could evaluate the prediction performance in an
-% infinite hold-out sample.
 
@@
 proportional hazards models \citep{golmakani2020super}.
-% A lesser
-% known fact is that a super learner constructed with the partial
-% log-likelihood loss implicitly depends on the censoring distribution
-% \citep{hjort1992inference,whitney2019comment}.
-
-Alternative approaches for super learning with right-censored data use
-an inverse probability of censoring weighted (IPCW) loss function
+
+Alternative approaches for super learning with right-censored data use an
+inverse probability of censoring weighted (IPCW) loss function
 \citep{graf1999assessment,van2003unicv,molinaro2004tree,keles2004asymptotically,hothorn2006survival,gerds2006consistent,gonzalez2021stacked},
-% An IPCW loss function is attractive because the associated risk does
-% not depend on the censoring distribution but describes a feature of
-% the population of interested. Similar results can be obtained using
 censoring unbiased transformations
 \citep{fan1996local,steingrimsson2019censoring}, or pseudo-values
-\citep{andersen2003generalised,mogensen2013random,sachs2019ensemble}.
-All these methods rely on an estimator of the censoring distribution,
-and their drawback is that this estimator has to be pre-specified.
-% When the data-generating mechanism is complex and not
-% well-understood, pre-specification of the censoring distribution is a
-% challenge.
-% Inverse probability of censoring weighted loss has been studied for
-% decades \citep{graf1999assessment, hothorn2006survival,
-  % gerds2006consistent}. A disadvantage of inverse probability of
-% censoring weighted loss estimators is that they require a
-% pre-specified model for the censoring distribution.
-% Recently \cite{westling2021inference} proposed a procedure which
-% iterates between super learning of the censoring distribution and
-% super learning of the event time distribution.
-An approach which avoids a pre-specified censoring model was proposed
-independently by \cite{han2021inverse} and
-\cite{westling2021inference}. In both articles, the authors suggest to
-iterate between learning of the outcome model and learning of the
-censoring model using IPCW loss functions. However, no general
-theoretical guarantees seem to exist for this procedure, and it has
-not yet been extended to the situation with competing risks.
+\citep{andersen2003generalised,mogensen2013random,sachs2019ensemble}. All these
+methods rely on an estimator of the censoring distribution, and their drawback
+is that this estimator has to be pre-specified. An approach which avoids a
+pre-specified censoring model was proposed independently by
+\cite{han2021inverse} and \cite{westling2021inference}. In both articles, the
+authors suggest to iterate between learning of the outcome model and learning of
+the censoring model using IPCW loss functions. However, no general theoretical
+guarantees seem to exist for this procedure, and it has not yet been extended to
+the situation with competing risks.
 
@@
 \begin{equation*}
-  % \label{eq:cum-haz}
   \Lambda_{j}(t \mid x) = \int_0^t\frac{  Q(T \in \diff s, D=j \mid X=x )}{Q(T \geq s \mid X=x )}.
@@
 $\Lambda_{j}$.
-% Formally, the domain of $a$ depends on \(n\) but we suppress this in the notation.
 In what follows, we use the short-hand notation
@@
 \begin{equation*}
-  % \label{eq:cv-risk-est}
   \hat{R}_n(a; L) =
   \frac{1}{K}\sum_{k=1}^{K}
-  % \empmeas^k{[L {(a{ (\data_n^{-k})} , \blank) }]},
   \frac{1}{| \data_n^{k} |}\sum_{O_i \in \data_n^{k}}
@@
 \end{equation*}
-% where \( \empmeas^{k} \) is the empirical measure of \( \data_n^{k} \). 
-The subset \(\data_n^{-k}\) is referred to as the \(k\)'th training
-sample, while \(\data_n^{k}\) is referred to as the \(k\)'th test or
-hold-out sample.
+The subset \(\data_n^{-k}\) is referred to as the \(k\)'th training sample,
+while \(\data_n^{k}\) is referred to as the \(k\)'th test or hold-out sample.
 The discrete super learner is defined as
@@
 \end{equation*}
-% The final estimator of \(\Psi(P)=\Lambda_j\) is then the selected
-% learner applied to the full data set, i.e., \(\hat{a}_n(\data_n)\).
 The oracle learner is defined as the learner that minimises the
@@
 
-% The problem with most existing super learners for right-censored data is that
-% they depend on a pre-specified estimator of the censoring distribution.
 The main idea of the state learner is to jointly use learners for
@@
 exclusive states: \( 0 \), \( 1 \), \( 2 \), or \( -1 \).
-% (Figure \ref{fig:multi-state-process}).
-% \begin{figure}[h]
-  % \centering %
-  % \includeFigCond[width=.5\textwidth]{figure-multi-state-process.pdf}
-  % \caption{Illustration of the multi-state process \(\eta\) used by
-    % the state learner. Note that `censored' is a state, hence the
-    % process is always observed at any time.}
-  % \label{fig:multi-state-process}
-% \end{figure}
 The conditional distribution of the process \( \eta(t) \) given
@@
 \end{equation}
-% \begin{equation}\label{eq:transition}
-%   \begin{split}
-%   F(t, 1, x)
-%     &
-%       % = P(\tilde{T} \leq t, \tilde{D}=1 \mid X=x)
-%     = \int_0^t \exp{{\{-\Lambda_{1}(s \mid x)-\Lambda_{2}(s \mid x) - \Gamma(s \mid x)\} }}  \Lambda_{1}(\diff s \mid x),
-%   \\
-%   F(t, 2, x)
-%     &
-%       % = P(\tilde{T} \leq t, \tilde{D}=2 \mid X=x)
-%       = \int_0^t \exp{\{-\Lambda_{1}(s \mid x)-\Lambda_{2}(s \mid x) - \Gamma(s \mid x)\} }  \Lambda_{2}(\diff s \mid x),
-%     \\
-%     F(t, -1, x)
-%     &
-%       % = P(\tilde{T} \leq t, \tilde{D}=0 \mid X=x)
-%       = \int_0^t \exp{\{-\Lambda_{1}(s \mid x)-\Lambda_{2}(s \mid x) - \Gamma(s \mid x)\} }  \Gamma(\diff s \mid x),
-%     \\
-%     F(t, 0, x)
-%     &
-%       = P(\tilde{T} > t \mid X= x)
-%       = 1- F(t, 1, x) - F(t, 2, x)- F(t, -1, x).
-%   \end{split}
-% \end{equation}
 The state learner requires three libraries of learners (c.f., Section \ref{sec:super-learning}),
@@
 learners for the conditional cumulative hazard function of the
-censoring distribution. % We further
-% define \(\mathcal{H}\) as the set of all probability distributions on
-% \( \mathcal{X} \) and
-% \begin{equation*}
-%   \mathcal{H}_n=\left\{h\colon\mathcal{O}^n\longrightarrow\mathcal{H} \midd
-%     h(\data_n)=H_n = \frac 1 n \sum_{i=1}^n \delta_{X_i} \right\}
-% \end{equation*}
-% as the library of learners which consists of a single learner, the empirical
-% distribution function.
+censoring distribution. 
 Based on the Cartesian product of
@@
 \end{align*}
-% \begin{align*}
-%   \mathcal{F}(\mathcal{A}_1, \mathcal{A}_2, \mathcal{B})
-%   &= \{ \phi_{a_1,a_2, b} : a_1 \in \mathcal{A}_1, a_2 \in \mathcal{A}_2, b \in \mathcal{B}\},
-%     \intertext{where in correspondence with  the relations in equation \eqref{eq:transition},} 
-%     \phi_{a_1,a_2, b}(\data_n)(t,1,x) &= \int_0^t \exp{\{-a_1(\data_n)(s \mid x)-a_2(\data_n)(s \mid x) - b(\data_n)(s \mid x)\} }  a_1(\data_n)(\diff s \mid x),\\
-%   \phi_{a_1,a_2, b}(\data_n)(t,2,x) &= \int_0^t \exp{\{-a_1(\data_n)(s \mid x)-a_2(\data_n)(s \mid x) - b(\data_n)(s \mid x)\} }  a_2(\data_n)(\diff s \mid x),\\
-%   \phi_{a_1,a_2, b}(\data_n)(t,-1,x) &= \int_0^t \exp{\{-a_1(\data_n)(s \mid x)-a_2(\data_n)(s \mid x) - b(\data_n)(s \mid x)\} }  b(\data_n)(\diff s \mid x),\\
-%   \phi_{a_1,a_2, b}(\data_n)(t,0,x) &= 1
-%                                       - \phi_{a_1,a_2, b}(\data_n)(t,1,x)
-%                                       - \phi_{a_1,a_2, b}(\data_n)(t,2,x)
-%     -\phi_{a_1,a_2, b}(\data_n)(t,-1,x)
-% \end{align*}
 To evaluate how well a function \( F \) predicts the observed
@@
 
-% We let \( \E_P \) denote
-% expectation under \( P \) and use \( \Theta \) to denote the
-% collection of all conditional state-occupation probability functions.
 Proposition~\ref{prop:stric-prop} can be derived from the fact that the
@@
     F_0 = \argmin_{F \in \mathcal{H}_{\mathcal{P}}} P_0{[\bar{B}_\tau(F, \blank)]},
-  % F^* = \argmin_{(\Lambda_1,\Lambda_2,\Gamma) \in \mathcal{M}_{\tau}\otimes\mathcal{M}_{\tau}\otimes\mathcal{M}_{\tau}} P_0{[\bar B_\tau(\phi_{(\Lambda_1,\Lambda_2,\Gamma)}, \blank)]},
   \end{equation*}
@@
 dependence on \( n \).
-% The discrete super $\hat{\phi}_n$ and the
-% oracle $\tilde{\phi}_n$ are defined as in Section~\ref{sec:framework}
-% but now with respect to the library \( \mathcal{F}_n \) and the loss
-% \( B \).
 In the following we let the space \( \mathcal{H}_{\mathcal{P}} \) be equipped with the norm
@@
 $\Lambda_2$.
-% In addition, the estimator will typically have an
-% asymptotic bias that vanishes at a too slow 
-% rate.
 Using tools from semi-parametric efficiency theory, it is possible to
@@
 
-% ANDERS: hedder det ikke Nelson-Aalen i stedet for Kaplan-Meier?
-%         kan du skrive klart hvad outcome er i denne foerste del? er det combined endpoint som man kalder for tumor-free survival?  
-For all super learners we use a library consisting of three learners:
-The Kaplan-Meier estimator
-\citep{kaplan1958nonparametric,Gerds_2019prodlim}, a Cox model with
-main effects \citep{cox1972regression, survival-package}, and a random
-survival forest \citep{ishwaran2008random,rfsrc-paclage}. We use the
-same library to learn the outcome distribution and the censoring
-distribution. Note that the three learners in our library of learners
-can be used to learn the cumulative hazard functions of the 
-outcome and the censoring distribution. The latter works by
-training the learner on the data set \( \data_n^c \), where
+For all super learners we use a library consisting of three learners: The
+Kaplan-Meier estimator \citep{kaplan1958nonparametric,Gerds_2019prodlim}, a Cox
+model with main effects \citep{cox1972regression, survival-package}, and a
+random survival forest \citep{ishwaran2008random,rfsrc-paclage}. We use the same
+library to learn the outcome distribution and the censoring distribution. Note
+that the three learners in our library of learners can be used to learn the
+cumulative hazard functions of the outcome and the censoring distribution. The
+latter works by training the learner on the data set \( \data_n^c \), where
 \( \data_n^c = \{O_i^c\}_{i=1}^n \) with
-\( O_i^c = (\tilde{T}_i, 1-\Delta, X_i) \). When we say that we use a
-learner for the cumulative hazard function of the outcome to learn the
-cumulative hazard function of the censoring time, we mean that the
-learner is trained on \( \data_n^c \).
-
-We compare the state learner to two IPCW based super learners: The
-first super learner, called IPCW(Cox), uses a Cox model with main
-effects to estimate the censoring probabilities, while the second
-super learner, called IPCW(KM), uses the Kaplan-Meier estimator to
-estimate the censoring probabilities. The Cox model for the censoring
-distribution is correctly specified in both simulation settings while
-the Kaplan Meier estimator only estimates the censoring model
-correctly in the simulation setting where censoring is
-independent. Both IPCW super learners are fitted using the
-\texttt{R}-package \texttt{riskRegression}
-\citep{Gerds_Ohlendorff_Ozenne_2023}.
-%
-% | time | sim_setting | true_events | true_cens | at_risk |
-% |------+-------------+-------------+-----------+---------|
-% |   36 | original    |      24.619 |    61.853 |  25.774 |
-% |   36 | indep_cens  |      24.674 |    38.740 |  46.141 |
-%
-The IPCW super learners use the integrated Brier score up to a fixed time
-horizon (36 months). The marginal risk of the event before this time horizon is
-\(\approx 24.6\)\%. Under the `dependent censoring' setting the marginal
-censoring probability before the time horizon is \(\approx 61.9\)\%. Under the
-`independent censoring' setting the marginal censoring probability before this
-time horizon is \( \approx 38.7 \)\%.
+\( O_i^c = (\tilde{T}_i, 1-\Delta, X_i) \). When we say that we use a learner
+for the cumulative hazard function of the outcome to learn the cumulative hazard
+function of the censoring time, we mean that the learner is trained on
+\( \data_n^c \).
+
+We compare the state learner to two IPCW based super learners: The first super
+learner, called IPCW(Cox), uses a Cox model with main effects to estimate the
+censoring probabilities, while the second super learner, called IPCW(KM), uses
+the Kaplan-Meier estimator to estimate the censoring probabilities. The Cox
+model for the censoring distribution is correctly specified in both simulation
+settings while the Kaplan Meier estimator only estimates the censoring model
+correctly in the simulation setting where censoring is independent. Both IPCW
+super learners are fitted using the \texttt{R}-package \texttt{riskRegression}
+\citep{Gerds_Ohlendorff_Ozenne_2023}. The IPCW super learners use the integrated
+Brier score up to a fixed time horizon (36 months). The marginal risk of the
+event before this time horizon is \(\approx 24.6\)\%. Under the `dependent
+censoring' setting the marginal censoring probability before the time horizon is
+\(\approx 61.9\)\%. Under the `independent censoring' setting the marginal
+censoring probability before this time horizon is \( \approx 38.7 \)\%.
 
@@
 
-We use the learners of the three cumulative hazard functions selected
-by the state learner and the estimator defined in
-Section~\ref{sec:cause-spec-aver},
-equation~(\ref{eq:one-step-comp-ate}), to estimate the average
-treatment effect of hormone therapy on risk of tumor recurrence and
-death. The propensity score is estimated with a lasso model that
-includes all levels of interaction.
-% ANDERS: er du sikker paa at det er ALL levels of interaction, ogsaa higher order?
-%         der mangler en reference for lasso model
-The results are shown in Figure~\ref{fig:zelefski-real-target} for 6
-month intervals after baseline with pointwise 95\% confidence
-intervals. We see that hormone therapy decreases the risk of tumor
-recurrence and increases the risk of death without tumor recurrence,
-but that none of the estimated effects are statistically significant.
+We use the learners of the three cumulative hazard functions selected by the
+state learner and the estimator defined in Section~\ref{sec:cause-spec-aver},
+equation~(\ref{eq:one-step-comp-ate}), to estimate the average treatment effect
+of hormone therapy on risk of tumor recurrence and death. The propensity score
+is estimated with a lasso model that includes all levels of interaction. The
+results are shown in Figure~\ref{fig:zelefski-real-target} for 6 month intervals
+after baseline with pointwise 95\% confidence intervals. We see that hormone
+therapy decreases the risk of tumor recurrence and increases the risk of death
+without tumor recurrence, but that none of the estimated effects are
+statistically significant.
 
@@
   \centering%
-  \includeFigCond[width=.9\linewidth]{real-data-target.pdf}
+  \includeFigCond[width=1\linewidth]{real-data-target.pdf}
   \caption[]{Estimates of the average treatment effect of hormone therapy on the
@@
 A major advantage of the state learner is that the performance of each
-combination of learners can be estimated without additional nuisance
-parameters. A potential drawback of our approach is that we are
-evaluating the loss of the learners on the level of the observed data
-distribution while the target of the analysis is either the event time
-distribution, or the censoring distribution, or both.
-% ANDERS: er du enig i min omformulering af forgaaende saetning?
-Specifically, the finite sample oracle inequality in
-Corollary~\ref{cor:oracle-prop} concerns the function \( F \), which
-is a feature of \( P \in \mathcal{P} \), while what we are typically
-interested in is \( \Lambda_j \) or \( S \), which are features of
-\( Q \in \mathcal{Q} \). We emphasise that while the state learner
-provides us with estimates of \( \Lambda_j \) and $\Gamma$ based on
-libraries \( \mathcal{A}_j \) and \( \mathcal{B} \), the performance
-of these learners is not assessed directly for their respective target
-parameters, but only indirectly via the performance of \( F \).  For
-settings without competing risks, our numerical studies suggest that
-measuring the performance of \( F \) also leads to good performance
-for estimation of \( S \).
-% ANDERS: den forgaaende paragraf gentager det samme 2-3 gang. kan du forkorte en gang?
-
-% In the context of targeted learning, a drawback of the state learner is that the
-% doubly robustness property defined in equation~(\ref{eq:dr-property}) seem to be
-% lost because we only use a single nuisance parameter estimator. As defined here,
-% however, the state learner is build using libraries for the conditional
-% cause-specific cumulative hazard functions, so some doubly robustness might be
-% preserved.
+combination of learners can be estimated without additional nuisance parameters.
+A potential drawback of our approach is that we are evaluating the loss of the
+learners on the level of the observed data distribution while the target of the
+analysis is either the event time distribution, or the censoring distribution,
+or both. Specifically, the finite sample oracle inequality in
+Corollary~\ref{cor:oracle-prop} concerns the function \( F \), which is a
+feature of \( P \in \mathcal{P} \), while what we are typically interested in is
+\( \Lambda_j \) or \( S \), which are features of \( Q \in \mathcal{Q} \). We
+emphasise that while the state learner provides us with estimates of
+\( \Lambda_j \) and $\Gamma$ based on libraries \( \mathcal{A}_j \) and
+\( \mathcal{B} \), the performance of these learners is not assessed directly
+for their respective target parameters, but only indirectly via the performance
+of \( F \). For settings without competing risks, our numerical studies suggest
+that measuring the performance of \( F \) also leads to good performance for
+estimation of \( S \).
 
 Our proposed super learner can be implemented with a broad library of learners
-and using existing software.
-% For instance the \texttt{R}-package
-% \texttt{riskRegression} \citep{Gerds_Ohlendorff_Ozenne_2023} and the \texttt{R}-package by westling
-Furthermore, while
-the library \( \mathcal{F}(\mathcal{A}_1,\mathcal{A}_2,\mathcal{B}) \) consists
-of \( |\mathcal{A}_1||\mathcal{A}_2||\mathcal{B}| \) many learners, we only need to fit
-\( |\mathcal{A}_1| +|\mathcal{A}_2| + |\mathcal{B}| \) many learners in each fold. To
-evaluate the performance of each learner we need to perform
-\( |\mathcal{A}_1||\mathcal{A}_2||\mathcal{B}| \) many operations to calculate the
-integrated Brier score in each hold-out sample, one for each combination of the
-fitted models, but these operations are often negligible compared to fitting the
-models. Hence the state learner is essentially not more computationally demanding
-than any procedure that uses super learning to learn $\Lambda_1$, $\Lambda_2$,
-and $\Gamma$ separately. While our proposal is based on constructing the library
-\( \mathcal{F} \) from libraries for learning \( \Lambda_1 \), $\Lambda_2$, and
-$\Gamma$, it could also be of interest to consider learners that estimate
-\( F \) directly.
+and using existing software. Furthermore, while the library
+\( \mathcal{F}(\mathcal{A}_1,\mathcal{A}_2,\mathcal{B}) \) consists of
+\( |\mathcal{A}_1||\mathcal{A}_2||\mathcal{B}| \) many learners, we only need to
+fit \( |\mathcal{A}_1| +|\mathcal{A}_2| + |\mathcal{B}| \) many learners in each
+fold. To evaluate the performance of each learner we need to perform
+\( |\mathcal{A}_1||\mathcal{A}_2||\mathcal{B}| \) many operations to calculate
+the integrated Brier score in each hold-out sample, one for each combination of
+the fitted models, but these operations are often negligible compared to fitting
+the models. Hence the state learner is essentially not more computationally
+demanding than any procedure that uses super learning to learn $\Lambda_1$,
+$\Lambda_2$, and $\Gamma$ separately. While our proposal is based on
+constructing the library \( \mathcal{F} \) from libraries for learning
+\( \Lambda_1 \), $\Lambda_2$, and $\Gamma$, it could also be of interest to
+consider learners that estimate \( F \) directly.
 
@@
 \end{document}
+
+\typeout{get arXiv to do 4 passes: Label(s) may have changed. Rerun}

### 2024-12-19 R-code/functions/statelearner.R
restructure only one version

@@ -3,9 +3,9 @@
 ## Author: Anders Munch
 ## Created: Aug 18 2023 (10:26) 
 ## Version: 
-## Last-Updated: May 21 2024 (17:28) 
+## Last-Updated: Jul 16 2024 (16:16) 
 ##           By: Anders Munch
-##     Update #: 413
+##     Update #: 418
 #----------------------------------------------------------------------
 ## 
 ### Commentary: 
@@ -159,6 +159,158 @@ statelearner <- function(learners,
                                                     list_ch_censor[[base_dt[, censor]]])
                     event_counts = list(cause1_event, censor_event)
                 }
+                loss = sum(as.numeric(sapply(1:length(abs_risks), function(ll){
+                    risk_pred = abs_risks[[ll]][, sindex(jump.times = eval_times,eval.times = time_grid)]
+                    obs_event = event_counts[[ll]]
+                    ## Check this ok!
+                    brier = (risk_pred-obs_event)^2
+                    if (integrate)
+                        loss0 = mean(rowSums(brier*diff(time_grid)[1]))
+                    else
+                        loss0 = mean(brier[, dim(brier)[2]])
+                    return(loss0)
+                })))
+                out = copy(base_dt)
+                out[, loss := loss]
+                return(out)                
+            }))
+            time_brier_calc = Sys.time() - time_brier_calc
+            return(list(brier_scores = brier_scores,
+                        time_fit =  time_fit,
+                        time_brier_calc = time_brier_calc))
+        })
+        out0 = rewrap(raw_out0, c(rbind, sum, sum))
+        if(comp_event_present)
+            out0$brier_scores = out0$brier_scores[, .(loss=mean(loss)), .(cause1, cause2, censor)]
+        else
+            out0$brier_scores = out0$brier_scores[, .(loss=mean(loss)), .(cause1, censor)]
+        out0$brier_scores[, b:=bb]
+        return(out0)
+    })
+    rw_list_cv_fits = rewrap(list_cv_fits, c(rbind, sum, sum))
+    cv_fits = rw_list_cv_fits$brier_scores
+    if(verbose){
+        all_time_fit = rw_list_cv_fits$time_fit
+        all_time_brier_calc = rw_list_cv_fits$time_brier_calc
+        message("Spend ", round(as.numeric(all_time_fit)/60, digits = 2), " minutes fitting models\n",
[…]

### 2024-12-19 manuscript/manuscript.tex
restructure only one version
edit intro, focus on other approaches

--- 2024-05-27 tex/journal-version/manuscript.tex
@@
 %% Line spacing
-\linespread{1.65}
+% \linespread{1.65}
 
@@
 
-Machine learning based on right-censored data commonly uses the
-partial log-likelihood as a loss function
+The state learner algorithm can output a medical risk prediction model
+\citep{gerds2021medical} which predicts the probability of an event based on
+covariates in the presence of competing risks. Another important application,
+which motivated our contribution, is in targeted learning where conditional
+event probabilities occur as high-dimensional nuisance parameters which need to
+be estimated at a certain rate \citep{van2011targeted, rytgaard2021estimation,
+  rytgaard2022targeted}. The state learner is particularly well-suited for this
+application as it simultaneously provides estimates for both the outcome and the
+censoring, both of which are needed for targeted learning. For the asymptotic
+bias term of a targeted estimator, which uses the state learner to estimate
+nuisance parameters, we show that a second order product structure holds. We
+illustrate both applications of the state learner with prostate cancer data
+\citep{kattan2000pretreatment}.
+
+There are many existing super learners for right-censored data. We briefly
+review them here and contrast our suggestion. Machine learning based on
+right-censored data commonly uses the partial log-likelihood as a loss function
 \citep[e.g.,][]{li2016regularized,yao2017deep,lee2018deephit,katzman2018deepsurv,gensheimer2019scalable,lee2021boosted,kvamme2021continuous}.
-However, this loss function does not work well with data splitting
-(cross-validation) because the partial log-likelihood loss assigns an
-infinite value when applied to a test set time point which does not
-occur in the learning set as soon as the learner predicts piece-wise
-constant cumulative hazard functions. This is the case for prominent
-survival learners including the Kaplan-Meier estimator, the random
-survival forest, and the semi-parametric Cox regression model.  When a
-proportional hazards model is assumed, the baseline hazard function
-can be profiled out of the likelihood \citep{cox1972regression}.  The
-cross-validated partial log-likelihood loss \citep{verweij1993cross}
-has therefore been suggested as a loss function for super learning
-which however restricts the library of learners to include only Cox
-proportional hazards models \citep{golmakani2020super}.
+This loss function is also suggested for super learning with right-censored data
+by \cite{polley2011-sl-cens}, where data is assumed to be observed in discrete
+time. However, the partial log-likelihood loss does not work well with data
+splitting (cross-validation) in continuous time because this loss assigns an
+infinite value when applied to a test set time point which does not occur in the
+learning set as soon as the learner predicts piece-wise constant cumulative
+hazard functions. This is the case for prominent survival learners including the
+Kaplan-Meier estimator, the random survival forest, and the semi-parametric Cox
+regression model, which hence cannot be included in the library of the super
+learner proposed by \cite{polley2011-sl-cens}.
+
+When a proportional hazards model is assumed, the baseline hazard function can
+be profiled out of the likelihood \citep{cox1972regression}. The cross-validated
+partial log-likelihood loss \citep{verweij1993cross} has therefore been
+suggested as a loss function for super learning \citep{golmakani2020super}. This
+choice of loss function restricts the library of learners to include only Cox
+proportional hazards models, and hence excludes many learners such as, e.g.,
+random survival models, additive hazards models and accelerated failure time
+models.
 % A lesser
@@
 
-Alternative approaches for super learning with right-censored data use
-an inverse probability of censoring weighted (IPCW) loss function
+Other attempts at super learning with right-censored data use an inverse
+probability of censoring weighted (IPCW) loss function
 \citep{graf1999assessment,van2003unicv,molinaro2004tree,keles2004asymptotically,hothorn2006survival,gerds2006consistent,gonzalez2021stacked},
@@
 % super learning of the event time distribution.
-An approach which avoids a pre-specified censoring model was proposed
-independently by \cite{han2021inverse} and
-\cite{westling2021inference}. In both articles, the authors suggest to
-iterate between learning of the outcome model and learning of the
-censoring model using IPCW loss functions. However, no general
-theoretical guarantees seem to exist for this procedure, and it has
-not yet been extended to the situation with competing risks.
-
-The state learner algorithm can output a medical risk prediction model
-\citep{gerds2021medical} which predicts the probability of an event based on
-covariates in the presence of competing risks. The other application is in
-targeted learning where conditional event probabilities occur as
-high-dimensional nuisance parameters which need to be estimated at a certain
-rate \citep{van2011targeted, rytgaard2021estimation, rytgaard2022targeted}. For
-the asymptotic bias term of targeted estimator, which uses the state learner to
-estimate nuisance parameters, we show that a second order product structure
-holds. We illustrate both applications of the state learner with prostate cancer
-data \citep{kattan2000pretreatment}.
+
+A recent approach which avoids a pre-specified censoring model was proposed
+independently by \cite{han2021inverse} and \cite{westling2021inference}. In both
+articles, the authors suggest to iterate between learning of the outcome model
+and learning of the censoring model using IPCW loss functions. However, no
+general theoretical guarantees seem to exist for this procedure. For instance,
+the procedure is not guaranteed to converge to the true data-generating
+mechanism \citep{munch2024thesis}.
+
+Contrary to the super learners considered by \cite{polley2011-sl-cens} and
+\cite{golmakani2020super}, the state learner poses no restrictions on the type
+of learners that can be included in the library. Contrary to IPCW, censoring
+unbiased transformations, or pseudo-value based super learners, the state
+learner does not require a pre-specified censoring distribution, but constructs
+a super learner for both the outcome and the censoring distribution. Contrary to
+the proposals by \cite{han2021inverse} and \cite{westling2021inference}, the
+state learner is backed by theoretical guarantees.
 

### 2024-12-20 manuscript/manuscript.tex
rewordings in the introduction

@@
 data, which simultaneously estimates the expected loss of learners of
-the event time distribution and the censoring distribution. The loss
-function underlying the state learner operates on the coarsened
-data. In the right-censored survival setting the coarsened data
-consist of the minimum and the order of the censoring time and the
-event time as well as the baseline covariates. The state learner can
-include a broad class of survival models as learners, it can handle
-competing risks, and it does not require a single pre-specified
+the event time distribution and the censoring distribution. Compared
+to the super learners of \cite{polley2011-sl-cens} and
+\cite{golmakani2020super}, the state learner proposed here poses no
+restrictions on the type of learners that can be included in the
+library. Compared to super learners that rely on inverse probability
+of censoring weighting, censoring unbiased transformations, or
+pseudo-value based super learners, the state learner does not require
+a pre-specified censoring distribution. Compared to the super learners
+by \cite{han2021inverse} and \cite{westling2021inference}, the state
+learner has theoretical guarantees which these methods do not
+have.
+% Moreover, state learner seems to be the first super learner for
+% right censored data which learns the outcome and the censoring
+% distribution simulaneously.
+
+The loss function underlying the state learner operates on the
+coarsened data. In the right-censored survival setting the coarsened
+data consist of the minimum and the order of the censoring time and
+the event time as well as the baseline covariates. The state learner
+can include a broad class of survival models as learners, it can
+handle competing risks, and it does not require a single pre-specified
 estimator of the conditional censoring distribution. To analyse the
@@
 
-There are many existing super learners for right-censored data. We briefly
-review them here and contrast our suggestion. Machine learning based on
-right-censored data commonly uses the partial log-likelihood as a loss function
+There are many existing super learners for right-censored data. We
+briefly review them here. Machine learning based on right-censored
+data commonly uses the partial log-likelihood as a loss function
 \citep[e.g.,][]{li2016regularized,yao2017deep,lee2018deephit,katzman2018deepsurv,gensheimer2019scalable,lee2021boosted,kvamme2021continuous}.
-This loss function is also suggested for super learning with right-censored data
-by \cite{polley2011-sl-cens}, where data is assumed to be observed in discrete
-time. However, the partial log-likelihood loss does not work well with data
-splitting (cross-validation) in continuous time because this loss assigns an
-infinite value when applied to a test set time point which does not occur in the
-learning set as soon as the learner predicts piece-wise constant cumulative
-hazard functions. This is the case for prominent survival learners including the
-Kaplan-Meier estimator, the random survival forest, and the semi-parametric Cox
-regression model, which hence cannot be included in the library of the super
-learner proposed by \cite{polley2011-sl-cens}.
-
-When a proportional hazards model is assumed, the baseline hazard function can
-be profiled out of the likelihood \citep{cox1972regression}. The cross-validated
-partial log-likelihood loss \citep{verweij1993cross} has therefore been
-suggested as a loss function for super learning \citep{golmakani2020super}. This
-choice of loss function restricts the library of learners to include only Cox
-proportional hazards models, and hence excludes many learners such as, e.g.,
-random survival models, additive hazards models and accelerated failure time
-models.
-% A lesser
-% known fact is that a super learner constructed with the partial
-% log-likelihood loss implicitly depends on the censoring distribution
-% \citep{hjort1992inference,whitney2019comment}.
-
-Other attempts at super learning with right-censored data use an inverse
-probability of censoring weighted (IPCW) loss function
+This loss function is also suggested for super learning with
+right-censored data by \cite{polley2011-sl-cens}, where data is
+assumed to be observed in discrete time. However, the partial
+log-likelihood loss does not work well with data splitting
+(cross-validation) in continuous time. The reason is that the partial
+log-likelihood loss assigns an infinite value for any learner that
+predicts piece-wise constant cumulative hazard functions when there
+are event times in the test set which do not occur in the learning
+set. This problem occurs with prominent survival learners including
+the Kaplan-Meier estimator, the random survival forest, and the
+semi-parametric Cox regression model and these learners cannot be
+included in the library of the super learner proposed by
+\cite{polley2011-sl-cens}. When a proportional hazards model is
+assumed, the baseline hazard function can be profiled out of the
+likelihood \citep{cox1972regression}. The cross-validated partial
+log-likelihood loss \citep{verweij1993cross} has therefore been
+suggested as a loss function for super learning
+\citep{golmakani2020super}. This choice of loss function restricts the
+library of learners to include only Cox proportional hazards models,
+and hence excludes many learners such as, e.g., random survival
+models, additive hazards models and accelerated failure time models.
+Other proposals for super learning with right-censored data use an
+inverse probability of censoring weighted (IPCW) loss function
 \citep{graf1999assessment,van2003unicv,molinaro2004tree,keles2004asymptotically,hothorn2006survival,gerds2006consistent,gonzalez2021stacked},
-% An IPCW loss function is attractive because the associated risk does
-% not depend on the censoring distribution but describes a feature of
-% the population of interested. Similar results can be obtained using
 censoring unbiased transformations
@@
 All these methods rely on an estimator of the censoring distribution,
-and their drawback is that this estimator has to be pre-specified.
-% When the data-generating mechanism is complex and not
-% well-understood, pre-specification of the censoring distribution is a
-% challenge.
-% Inverse probability of censoring weighted loss has been studied for
-% decades \citep{graf1999assessment, hothorn2006survival,
-  % gerds2006consistent}. A disadvantage of inverse probability of
-% censoring weighted loss estimators is that they require a
-% pre-specified model for the censoring distribution.
-% Recently \cite{westling2021inference} proposed a procedure which
-% iterates between super learning of the censoring distribution and
-% super learning of the event time distribution.
-
-A recent approach which avoids a pre-specified censoring model was proposed
-independently by \cite{han2021inverse} and \cite{westling2021inference}. In both
-articles, the authors suggest to iterate between learning of the outcome model
-and learning of the censoring model using IPCW loss functions. However, no
-general theoretical guarantees seem to exist for this procedure. For instance,
-the procedure is not guaranteed to converge to the true data-generating
+and their drawback is that this estimator has to be pre-specified. An
+approach which avoids a pre-specified censoring model was proposed
+independently by \cite{han2021inverse} and
+\cite{westling2021inference}. In both articles, the authors suggest to
+iterate between learning of the outcome model and learning of the
+censoring model using IPCW loss functions. However, no general
+theoretical guarantees seem to exist for this procedure and the
+procedure is not guaranteed to converge to the true data-generating
 mechanism \citep{munch2024thesis}.
-
-Contrary to the super learners considered by \cite{polley2011-sl-cens} and
-\cite{golmakani2020super}, the state learner poses no restrictions on the type
-of learners that can be included in the library. Contrary to IPCW, censoring
-unbiased transformations, or pseudo-value based super learners, the state
-learner does not require a pre-specified censoring distribution, but constructs
-a super learner for both the outcome and the censoring distribution. Contrary to
-the proposals by \cite{han2021inverse} and \cite{westling2021inference}, the
-state learner is backed by theoretical guarantees.
 

### 2024-12-20 manuscript/manuscript.tex
adapted Thomas changes

@@
 
-There are many existing super learners for right-censored data. We
-briefly review them here. Machine learning based on right-censored
-data commonly uses the partial log-likelihood as a loss function
+There are many existing super learners for right-censored data. We briefly
+review them here. Machine learning based on right-censored data commonly uses
+the partial log-likelihood as a loss function
 \citep[e.g.,][]{li2016regularized,yao2017deep,lee2018deephit,katzman2018deepsurv,gensheimer2019scalable,lee2021boosted,kvamme2021continuous}.
-This loss function is also suggested for super learning with
-right-censored data by \cite{polley2011-sl-cens}, where data is
-assumed to be observed in discrete time. However, the partial
-log-likelihood loss does not work well with data splitting
-(cross-validation) in continuous time. The reason is that the partial
-log-likelihood loss assigns an infinite value for any learner that
-predicts piece-wise constant cumulative hazard functions when there
-are event times in the test set which do not occur in the learning
-set. This problem occurs with prominent survival learners including
-the Kaplan-Meier estimator, the random survival forest, and the
-semi-parametric Cox regression model and these learners cannot be
-included in the library of the super learner proposed by
-\cite{polley2011-sl-cens}. When a proportional hazards model is
-assumed, the baseline hazard function can be profiled out of the
-likelihood \citep{cox1972regression}. The cross-validated partial
-log-likelihood loss \citep{verweij1993cross} has therefore been
-suggested as a loss function for super learning
-\citep{golmakani2020super}. This choice of loss function restricts the
-library of learners to include only Cox proportional hazards models,
-and hence excludes many learners such as, e.g., random survival
-models, additive hazards models and accelerated failure time models.
-Other proposals for super learning with right-censored data use an
-inverse probability of censoring weighted (IPCW) loss function
+This loss function is also suggested for super learning with right-censored data
+by \cite{polley2011-sl-cens}, where data is assumed to be observed in discrete
+time. However, the partial log-likelihood loss does not work well with data
+splitting (cross-validation) in continuous time. The reason is that the partial
+log-likelihood loss assigns an infinite value for any learner that predicts
+piece-wise constant cumulative hazard functions when there are event times in
+the test set which do not occur in the learning set. This problem occurs with
+prominent survival learners including the Kaplan-Meier estimator, the random
+survival forest, and the semi-parametric Cox regression model and these learners
+cannot be included in the library of the super learner proposed by
+\cite{polley2011-sl-cens}. When a proportional hazards model is assumed, the
+baseline hazard function can be profiled out of the likelihood
+\citep{cox1972regression}. The cross-validated partial log-likelihood loss
+\citep{verweij1993cross} has therefore been suggested as a loss function for
+super learning by \cite{golmakani2020super}. This choice of loss function
+restricts the library of learners to include only Cox proportional hazards
+models, and hence excludes many learners such as, e.g., random survival models,
+additive hazards models, and accelerated failure time models. Other proposals
+for super learning with right-censored data use an inverse probability of
+censoring weighted (IPCW) loss function
 \citep{graf1999assessment,van2003unicv,molinaro2004tree,keles2004asymptotically,hothorn2006survival,gerds2006consistent,gonzalez2021stacked},
@@
 \citep{fan1996local,steingrimsson2019censoring}, or pseudo-values
-\citep{andersen2003generalised,mogensen2013random,sachs2019ensemble}.
-All these methods rely on an estimator of the censoring distribution,
-and their drawback is that this estimator has to be pre-specified. An
-approach which avoids a pre-specified censoring model was proposed
-independently by \cite{han2021inverse} and
-\cite{westling2021inference}. In both articles, the authors suggest to
-iterate between learning of the outcome model and learning of the
-censoring model using IPCW loss functions. However, no general
-theoretical guarantees seem to exist for this procedure and the
-procedure is not guaranteed to converge to the true data-generating
+\citep{andersen2003generalised,mogensen2013random,sachs2019ensemble}. All these
+methods rely on an estimator of the censoring distribution, and their drawback
+is that this estimator has to be pre-specified. An approach which avoids a
+pre-specified censoring model was proposed independently by
+\cite{han2021inverse} and \cite{westling2021inference}. In both articles, the
+authors suggest to iterate between learning of the outcome model and learning of
+the censoring model using IPCW loss functions. However, it seems that this
+procedure is in general not guaranteed to converge to the true data-generating
 mechanism \citep{munch2024thesis}.

### 2024-12-21 manuscript/manuscript.tex
small changes

@@
 
-We define the state learner, a new super learner for right-censored
-data, which simultaneously estimates the expected loss of learners of
-the event time distribution and the censoring distribution. Compared
-to the super learners of \cite{polley2011-sl-cens} and
-\cite{golmakani2020super}, the state learner proposed here poses no
-restrictions on the type of learners that can be included in the
-library. Compared to super learners that rely on inverse probability
-of censoring weighting, censoring unbiased transformations, or
-pseudo-value based super learners, the state learner does not require
-a pre-specified censoring distribution. Compared to the super learners
-by \cite{han2021inverse} and \cite{westling2021inference}, the state
-learner has theoretical guarantees which these methods do not
-have.
+We define the state learner, a new super learner for right-censored data, which
+simultaneously estimates the expected loss of learners of the event time
+distribution and the censoring distribution. Compared to the super learners of
+\cite{polley2011-sl-cens} and \cite{golmakani2020super}, the state learner
+proposed here poses no restrictions on the type of learners that can be included
+in the library. Compared to super learners that rely on inverse probability of
+censoring weighting, censoring unbiased transformations, or pseudo-values, the
+state learner does not require a pre-specified censoring distribution. Compared
+to the super learners by \cite{han2021inverse} and \cite{westling2021inference},
+the state learner has theoretical guarantees which these methods do not have.
 % Moreover, state learner seems to be the first super learner for
@@
 
-The loss function underlying the state learner operates on the
-coarsened data. In the right-censored survival setting the coarsened
-data consist of the minimum and the order of the censoring time and
-the event time as well as the baseline covariates. The state learner
-can include a broad class of survival models as learners, it can
-handle competing risks, and it does not require a single pre-specified
-estimator of the conditional censoring distribution. To analyse the
-theoretical properties of the state learner we focus on the discrete
-super learner which combines the library of learners by picking the
-one that minimises the cross-validated loss \citep{van2007super}. In
-the presence of competing risks, our algorithm uses separate libraries
-of learners for the cumulative hazard functions for each of the
-competing risks and for the censoring distribution. We show that the
-oracle selector of the state learner is consistent if all libraries
-contain a consistent learner and prove a finite sample oracle
-inequality.
+The loss function underlying the state learner operates on coarsened data. In
+the right-censored survival setting the coarsened data consist of the minimum
+and the order of the censoring time and the event time as well as the baseline
+covariates. The state learner works by simultaneously assessing how well
+learners of the event time distribution and the censoring distribution predict
+the states of the coarsened data. To analyse the theoretical properties of the
+state learner we focus on the discrete super learner which combines the library
+of learners by picking the one that minimises the cross-validated loss
+\citep{van2007super}. In the presence of competing risks, our algorithm uses
+separate libraries of learners for the cumulative hazard functions for each of
+the competing risks and for the censoring distribution. We show that the oracle
+selector of the state learner is consistent if all libraries contain a
+consistent learner and prove a finite sample oracle inequality.
 
@@
 restricts the library of learners to include only Cox proportional hazards
-models, and hence excludes many learners such as, e.g., random survival models,
+models, and hence excludes many learners such as, e.g., random survival forests,
 additive hazards models, and accelerated failure time models. Other proposals
@@
 procedure is in general not guaranteed to converge to the true data-generating
-mechanism \citep{munch2024thesis}.
+mechanism \citep[][Appendix~A.4]{munch2024thesis}.
 

### 2025-01-13 manuscript/manuscript-wo-appendix.tex
biostats version

--- 2024-12-21 manuscript/manuscript.tex
@@
-\documentclass[12pt]{article}
+\documentclass[11pt]{article}
 \usepackage[a4paper,left=2.8cm, right=2.8cm, top=3cm, bottom=3cm]{geometry}
@@
 %% Line spacing
-% \linespread{1.65}
+\linespread{2}
 
@@
 % \linenumbers
+
+\usepackage{endfloat}
 
@@
 
-We introduce our notation and framework in
-Section~\ref{sec:framework}. In Section~\ref{sec:super-learning} we define
-general super learning for right-censored data.
-Section~\ref{sec:super-learner-simple} introduces the state learner, and
-Section~\ref{sec:theor-results-prop} provides theoretical guarantees. In
-Section~\ref{sec:targeted-learning} we discuss the use of the state learner in
-the context of targeted learning. We report results of our numerical experiments
-in Section~\ref{sec:numer-exper} and analyse a prostate cancer data set in
-Section~\ref{sec:real-data-appl}. Section~\ref{sec:discussion} contains a
-discussion of the merits and limitations of our proposal.
-Appendices~\ref{sec:proof-proposition} and~\ref{sec:state-learner-with} contain
-proofs. An implementation of the state learner is available at
+We introduce our notation and framework in Section~\ref{sec:framework}. In
+Section~\ref{sec:super-learning} we define general super learning for
+right-censored data. Section~\ref{sec:super-learner-simple} introduces the state
+learner, and Section~\ref{sec:theor-results-prop} provides theoretical
+guarantees. In Section~\ref{sec:targeted-learning} we discuss the use of the
+state learner in the context of targeted learning. We report results of our
+numerical experiments in Section~\ref{sec:numer-exper} and analyse a prostate
+cancer data set in Section~\ref{sec:real-data-appl}.
+Section~\ref{sec:discussion} contains a discussion of the merits and limitations
+of our proposal. Proof are given in the supplementary material. An
+implementation of the state learner is available at
 \url{https://github.com/amnudn/statelearner} along with codes for reproducing
@@
 \begin{proof}
-  See Appendix~\ref{sec:proof-proposition}.
+  See Section~S1 of the Supplementary material.
 \end{proof}
@@
 \begin{proof}
-  See Appendix~\ref{sec:proof-proposition}.
+  See Section~S2 of the Supplementary material.
 \end{proof}
@@
   \begin{align*}
-    & P_0{\left[
+    & P{\left[
       \int_0^{\tau} w(s, \blank)
       \left\{
-      \Gamma_0(s,\blank) -\hat{\Gamma}_n(s,\blank)
+      \Gamma(s,\blank) -\hat{\Gamma}_n(s,\blank)
       \right\}
-      [\Lambda_0-\hat{\Lambda}_n]
+      [\Lambda_1-\hat{\Lambda}_{1,n}]
       (\diff s, \blank)
@@
      & =
-      P_0{\left[
-      \int_0^{\tau} \int_0^{s} w^a_n(s,u,\blank) [F_0 - \hat{F}_n](u-, 0, \blank)[F_0 - \hat{F}_n](s-, 0, \blank) F_0(\diff u, 2, \blank ) F_0 ( \diff s, 1, \blank)
+      P{\left[
+      \int_0^{\tau} \int_0^{s} w^a_n(s,u,\blank) [F - \hat{F}_n](u-, 0, \blank)[F - \hat{F}_n](s-, 0, \blank) F(\diff u, -1, \blank ) F ( \diff s, 1, \blank)
       \right]}
@@
     & \quad +
-      P_0{\left[
-      \int_0^{\tau} \int_0^{s} w^b_n(s,u,\blank) [F_0 - \hat{F}_n](u-, 0, \blank)
-      F_0(\diff u, 2, \blank ) [F_0 - \hat{F}_n](\diff s, 1, \blank)
+      P{\left[
+      \int_0^{\tau} \int_0^{s} w^b_n(s,u,\blank) [F - \hat{F}_n](u-, 0, \blank)
+      F(\diff u, -1, \blank ) [F - \hat{F}_n](\diff s, 1, \blank)
       \right]}
@@
     & \quad +
-      P_0{\left[
-      \int_0^{\tau} \int_0^{s} w^c_n(s,u,\blank) [F_0 - \hat{F}_n](\diff u, 2, \blank)
-      [F_0 - \hat{F}_n](s-, 0, \blank)
-      F_0(\diff s, 1, \blank ) 
+      P{\left[
+      \int_0^{\tau} \int_0^{s} w^c_n(s,u,\blank) [F - \hat{F}_n](\diff u, -1, \blank)
+      [F - \hat{F}_n](s-, 0, \blank)
+      F(\diff s, 1, \blank ) 
       \right]}
@@
     & \quad +
-      P_0{\left[
-      \int_0^{\tau} \int_0^{s} w^d_n(s,u,\blank) [F_0 - \hat{F}_n](\diff u, 2, \blank)
-      [F_0 - \hat{F}_n](\diff s, 1, \blank)
+      P{\left[
+      \int_0^{\tau} \int_0^{s} w^d_n(s,u,\blank) [F - \hat{F}_n](\diff u, -1, \blank)
+      [F - \hat{F}_n](\diff s, 1, \blank)
       \right]}.
@@
 \begin{proof}
-  See Appendix~\ref{sec:state-learner-with}.
+  See Section~S3 of the Supplementary material.
 \end{proof}
@@
 
-\appendix
-
-\section{Theoretical guarantees for the state learner}
-\label{sec:proof-proposition}
-
-In this section we provide proofs of the results stated in
-Section~\ref{sec:theor-results-prop}.
-
-Define
-\( \bar{B}_{\tau,0}(F, o) = \bar{B}_{\tau}(F, o) - \bar{B}_{\tau}(F_0, o) \) and
-\( R_{0}(F) = P_0{[\bar{B}_{\tau,0}(F, \blank)]} \).
-\begin{lemma}
-  \label{lemma:norm}
-  \( R_{0}(F) = \Vert F - F_0 \Vert_{P_0}^2 \), where \( \Vert \blank \Vert_{P_0}\) is defined
-  in equation~(\ref{eq:norm}).
-\end{lemma}
-\begin{proof}[Proof]
-  For any \( t \in [0, \tau] \) and \( j\in \{-1,0,1,2\} \) we have
-  \begin{align*}
-    & \E_{P_0}{\left[ (F(t, j, X) - \1{\{\eta(t) = j \}})^2 \right]}
-    \\
-    & =    \E_{P_0}{\left[ (F(t, j, X) - F_0(t, j, X) + F_0(t, j, X) - \1{\{\eta(t) = j
-      \}})^2 \right]}
-    \\
-    & =    \E_{P_0}{\left[ (F(t, j, X) - F_0(t, j, X))^2\right]}
-      + \E_{P_0}{\left[ (F_0(t, j, X) - \1{\{\eta(t) = j \}})^2\right]}
-    \\
-    & \quad
-      + 2\E_{P_0}{\left[ (F(t, j, X) - F_0(t, j, X))(F_0(t, j, X) - \1{\{\eta(t) = j
-      \}})\right]}
-    \\
-    & =    \E_{P_0}{\left[ (F(t, j, X) - F_0(t, j, X))^2\right]}
-      + \E_{P_0}{\left[ (F_0(t, j, X) - \1{\{\eta(t) = j \}})^2\right]},
-  \end{align*}
-  where the last equality follows from the tower property. Hence, using Fubini,
-  we have
-  \begin{equation*}
-    P{[\bar{B}_{\tau}(F, \blank)]}
-    = \Vert F - F_0 \Vert_{P_0}^2 + P_0{[\bar{B}_{\tau}(F_0, \blank)]}.
-  \end{equation*}
-\end{proof}
-
-\begin{proof}[Proof of Proposition~\ref{prop:stric-prop}]
-  The result follows from Lemma~\ref{lemma:norm}.
-\end{proof}
-
-Recall that $\mathcal{H}_{\mathcal{P}}$ denote the function space consisting of
-all conditional state occupation probability functions for some measure
-\( P \in \mathcal{P} \).
-
-\begin{proof}[Proof of Corollary~\ref{cor:oracle-prop}]
-  First note that minimising the loss \( \bar{B}_{\tau} \) is equivalent to
-  minimising the loss \( \bar{B}_{\tau,0} \), so the discrete super learner and
-  oracle according to \( \bar{B}_{\tau} \) and \( \bar{B}_{\tau,0} \) are
-  identical. By Lemma~\ref{lemma:norm}, \( R_0(F) \geq 0 \) for any
-  \( F \in \mathcal{H}_{\mathcal{P}} \), and so using Theorem 2.3 from \citep{vaart2006oracle} with
-  \( p=1 \), we have that for all \( \delta >0 \),
-\begin{align*}
-  & \E_{P_0}{\left[ R_0(\hat{\phi}_n(\data_n^{-k})) \right]}
-  \\
-  &  \quad \leq
-    (1+2\delta)\E_{P_0}{\left[ R_0(\tilde{\phi}_n(\data_n^{-k})) \right]}
-  \\
-  & \qquad + (1+\delta) \frac{16 K}{n}
-    \log(1 + |\mathcal{F}_n|)\sup_{F \in \mathcal{H}_{\mathcal{P}}}
-    \left\{
-    M(F) + \frac{v(F)}{R_0(F)}
-    \left(
-    \frac{1}{\delta} + 1
-    \right)
-    \right\}
-\end{align*}
-where for each \( F \in \mathcal{H}_{\mathcal{P}} \), \( (M(F), v(F)) \) is some Bernstein pair for
-the function \(o \mapsto \bar{B}_{\tau,0}(F, o) \). As
-\( \bar{B}_{\tau,0}(F, \blank) \) is uniformly bounded by \( \tau \) for any
-\( F \in \mathcal{H}_{\mathcal{P}} \), it follows from section 8.1 in \citep{vaart2006oracle} that
-\( (\tau, 1.5 P_0{[\bar{B}_{\tau,0}(F, \blank)^2]}) \) is a Bernstein pair for
-\( \bar{B}_{\tau,0}(F, \blank) \). Now, for any \( a,b,c \in \R \) we have
-\begin{align*}
-  (a-c)^2 - (b-c)^2
-  & = (a-b+b-c)^2 - (b-c)^2
-  \\
-  & = (a-b)^2 + (b-c)^2 +2(b-c)(a-b) - (b-c)^2
-  \\
-  & = (a-b)
-    \left\{
-    (a-b) +  2(b-c)
-    \right\}
-  \\
-  & = (a-b)
-    \left\{
-     a + b -2c
-    \right\},
-\end{align*}
-so using this with \( a=F(t, j, x) \), \( b=F_0(t, j, x) \), and
-\( c = \1{\{\eta(t) = j\}} \), we have by Jensen's inequality
-\begin{align*}
-  & P_0{[\bar{B}_{\tau,0}(F, \blank)^2]}
-  \\
-  & \leq
-    2\tau\E_{P_0}{\left[
-    \sum_{j=-1}^{2} \int_0^{\tau}
-    \left\{
-    \left(
-    F(t, j, X) - \1{\{\eta(t) = j\}}
-    \right)^2
-    -
-    \left(
-    F_0(t, j, X) - \1{\{\eta(t) = j\}}
-    \right)^2
-    \right\}^2
-    \diff t 
-    \right]}
-  \\
-  & =2\tau
-    \E_{P_0}\Bigg[
-    \sum_{j=-1}^{2} \int_0^{\tau}
-    \left(
-    F(t, j, X) - F_0(t, j, X)
-    \right)^2
-  \\
-  & \quad \quad \quad\quad \quad \quad \times
-    \left\{
-    F(t, j, X) +  F_0(t, j, X)-2 \1{\{\eta(t) = j\}}
-    \right\}^2
-    \diff t 
-    \Bigg]
-  \\
-  & \leq
-    8\tau \E_{P_0}{\left[
-    \sum_{j=-1}^{2} \int_0^{\tau}
-    \left(
-    F(t, j, X) - F_0(t, j, X)
-    \right)^2
-    \diff t 
-    \right]}.
-  \\
-  & =
-    8\tau \Vert F - F_0 \Vert_{P_0}^2.
-\end{align*}
-Thus when \( v(F) = 1.5 P_0{[\bar{B}_{\tau,0}(F, \blank)^2]} \) we have by
-Lemma~\ref{lemma:norm}
-\begin{equation*}
-  \frac{v(F)}{R_0(F)}
-  = 1.5 \frac{P_0{[\bar{B}_{\tau,0}(F, \blank)^2]}}{P_0{[\bar{B}_{\tau,0}(F, \blank)]}}
-  \leq 12 \tau,
-\end{equation*}
-and so using the Bernstein pairs \( (\tau, 1.5 P_0{[\bar{B}_{\tau,0}(F, \blank)^2]}) \) we have
-\begin{equation*}
-  \sup_{F \in \mathcal{H}_{\mathcal{P}}}
-  \left\{
-    M(F) + \frac{v(F)}{R_0(F)}
-    \left(
-      \frac{1}{\delta} + 1
-    \right)
-  \right\}
-  \leq \tau
-  \left(
-    13 + \frac{12}{\delta}
-  \right),
-\end{equation*}
-For all $\delta>0$ we thus have
-\begin{align*}
-  \E_{P_0}{\left[ R_0(\hat{\phi}_n(\data_n^{-k})) \right]}
-  \leq
-  &(1+2\delta)\E_{P_0}{\left[ R_0(\tilde{\phi}_n(\data_n^{-k})) \right]}
-  \\
-  & \quad
-    + (1+\delta)\log(1 + |\mathcal{F}_n|) \tau \frac{16 K}{n}
-    \left(
-    13 + \frac{12}{\delta}
-    \right),
-\end{align*}
-and then the final result follows from Lemma~\ref{lemma:norm}.
-\end{proof}
-
-\begin{proof}[Proof of Corollary~\ref{cor:asymp-cons}]
-  By definition of the oracle and Lemma~\ref{lemma:norm},
-  \begin{equation*}
-    \E_{P_0}{\left[ \Vert \tilde{\phi}_n(\data_n^{-k}) - F_0 \Vert_{P_0}^2
-      \right]} \leq \E_{P_0}{\left[ \Vert \phi_n(\data_n^{-k}) - F_0 \Vert_{P_0}^2
-      \right]}  
-  \end{equation*}
-  for all \( n \in \N \). The results then follows from
-  Corollary~\ref{cor:oracle-prop}.
-\end{proof}
-
-\section{The state learner with targeted learning}
-\label{sec:state-learner-with}
-
-In this section show that a product structure is preserved when an estimator
-$\bar\Psi(\hat{F}_n, \hat{H}_n)$ is used instead of
-$\tilde\Psi(\hat{\Lambda}_n, \hat{\Gamma}_n, \hat{H}_n)$.
-
-\begin{proof}[Proof of Proposition~\ref{prop:dr-structure}]
-  For notational convenience we suppress \( X \) in the following. The final
-  result can be obtained by adding the argument \( X \) to all functions and
-  averaging. We use the relations from equation~(\ref{eq:7}) to write
-  \begingroup %
-  \allowdisplaybreaks
-    \begin{align*}
-      & \int_0^{\tau} w(s) 
-        \left\{
-        \Gamma(s) - \hat{\Gamma}_n(s)
-        \right\}
-        [\Lambda - \hat{\Lambda}_n](\diff s)
-      \\
-      & =
-        \int_0^{\tau} w(s) 
-        \left\{
-        \int_0^s \frac{F(\diff u, 2)}{F(u-, 0)} -
-        \int_0^s \frac{\hat{F}_n(\diff u, 2)}{\hat{F}_n(u-, 0)}  -
-        \right\}
-        \left[
-        \frac{F(\diff s, 1)}{F(s-, 0)}
-        - \frac{\hat{F}_n(\diff s, 1)}{\hat{F}_n(s-, 0)}
-        \right]
-      \\
-      & =
-        \int_0^{\tau} w(s) 
-        \Bigg\{
-        \int_0^s 
-        \left(
-        \frac{1}{F(u-, 0)} -  \frac{1}{\hat{F}_n(u-, 0)}
-        \right) F(\diff u, 2)
-      \\
-      & \qquad\qquad \qquad
-        +
-        \int_0^s \frac{1}{\hat{F}_n(u-, 0)} 
-        \left[
-        F(\diff u, 2) - \hat{F}_n(\diff u, 2)
-        \right]
-        \Bigg\}
-      \\
-      & \qquad\qquad \times
-        \left[
-        \left(
-        \frac{1}{F(s-, 0)} -
-        \frac{1}{\hat{F}_n(s-, 0)}
-        \right)F(\diff s, 1)
-        + \frac{1}{\hat{F}_n(s-, 0)}
-        \left(
-        F(\diff s, 1) -
-        \hat{F}_n(\diff s, 1)
-        \right)
-        \right]
-      \\
-      &
-        = \int_0^{\tau} 
-        \int_0^s
-        w(s) 
-        \left(
-        \frac{1}{F(u-, 0)} -  \frac{1}{\hat{F}_n(u-, 0)}
-        \right) 
-        \left(
-        \frac{1}{F(s-, 0)} -
-        \frac{1}{\hat{F}_n(s-, 0)}
-        \right)F(\diff u, 2)F(\diff s, 1)
-      \\
-      & \quad +
-        \int_0^{\tau}
-        \int_0^s
-        w(s) 
-        \left(
-        \frac{1}{F(u-, 0)} -  \frac{1}{\hat{F}_n(u-, 0)}
-        \right) \frac{F(\diff u, 2) }{\hat{F}_n(u-,0)}
-        \left(
-        F(\diff s, 1) -
-        \hat{F}_n(\diff s, 1)
-        \right)
-      \\
-      & \quad +
-        \int_0^{\tau} 
-        \int_0^s      
-        \frac{w(s) }{\hat{F}_n(u-, 0)} 
-        \left[
-        F(\diff u, 2) - \hat{F}_n(\diff u, 2)
-        \right]
-        \left(
-        \frac{1}{F(s-, 0)} -
-        \frac{1}{\hat{F}_n(s-, 0)}
-        \right)F(\diff s, 1)
-      \\
-      & \quad +
-        \int_0^{\tau} 
-        \int_0^s      
-        \frac{w(s) }{\hat{F}_n(u-, 0)} 
-        \left[
-        F(\diff u, 2) - \hat{F}_n(\diff u, 2)
-        \right]
-        \frac{1}{\hat{F}_n(s-, 0)}
-        \left(
-        F(\diff s, 1) -
-        \hat{F}_n(\diff s, 1)
-        \right).
-    \end{align*}
-    \endgroup %
-    Consider the first term on the right hand side. Defining
-  \begin{equation*}
-    w_n^*(t)  = 
-    \left(
-      F(t-, 0)
-      - \hat{F}_n(t-, 0)
-    \right)
-    \left(
-      \frac{1}{F(t-, 0)}
-      - \frac{1}{\hat{F}_n(t-, 0)}
-    \right),
-  \end{equation*}
-  we can write
-  \begin{align*}
-    & \int_0^{\tau} 
-      \int_0^s
-      w(s) 
-      \left(
-      \frac{1}{F(u-, 0)} -  \frac{1}{\hat{F}_n(u-, 0)}
-      \right)      
-      \left(
-      \frac{1}{F(s-, 0)} -
-      \frac{1}{\hat{F}_n(s-, 0)}
-      \right)F(\diff u, 2)F(\diff s, 1)
-    \\
-    & =
-      \int_0^{\tau} 
-      \int_0^s
-      w(s)
-      w_n^*(u) 
-      \left(
-      F(u-, 0) - \hat{F}_n(u-, 0)
-      \right)
-    \\
-    & \qquad \qquad \quad
-      \times
-      w_n^*(s) 
-      \left(
-      F(s-, 0) - \hat{F}_n(s-, 0)
-      \right)       
-      F(\diff u, 2)F(\diff s, 1)
-    \\
-    & =
-      \int_0^{\tau} 
-      \int_0^s
-      w_n^a(s,u)
-      \left(
-      F(u-, 0) - \hat{F}_n(u-, 0)
-      \right)
-      \left(
-      F(s-, 0) - \hat{F}_n(s-, 0)
-      \right)       
-      F(\diff u, 2)F(\diff s, 1),
-  \end{align*}
-  where we have defined \( w_n^a(s,u) = w(s)w^*_n(s)w^*_n(u) \). By assumption,
-  \( w_n^a(s,u) \) is uniformly bounded. The same approach can be applied to the
-  three remaining terms which gives the result.
-\end{proof}
+\section{Competing interests}
+No competing interest is declared.
 

### 2025-01-13 manuscript/oup-authoring-template.tex
biostats version

%%
%% Copyright 2022 OXFORD UNIVERSITY PRESS
%%
%% This file is part of the 'oup-authoring-template Bundle'.
%% ---------------------------------------------
%%
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%%
%% The list of all files belonging to the 'oup-authoring-template Bundle' is
%% given in the file `manifest.txt'.
%%
%% Template article for OXFORD UNIVERSITY PRESS's document class `oup-authoring-template'
%% with bibliographic references
%%

%%%CONTEMPORARY%%%
% \documentclass[unnumsec,webpdf,contemporary,large]{oup-authoring-template}%
%\documentclass[unnumsec,webpdf,contemporary,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
%\documentclass[unnumsec,webpdf,contemporary,medium]{oup-authoring-template}
%\documentclass[unnumsec,webpdf,contemporary,small]{oup-authoring-template}

%%%MODERN%%%
%\documentclass[unnumsec,webpdf,modern,large]{oup-authoring-template}
%\documentclass[unnumsec,webpdf,modern,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
%\documentclass[unnumsec,webpdf,modern,medium]{oup-authoring-template}
%\documentclass[unnumsec,webpdf,modern,small]{oup-authoring-template}

%%%TRADITIONAL%%%
%\documentclass[unnumsec,webpdf,traditional,large]{oup-authoring-template}
% \documentclass[unnumsec,webpdf,traditional,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
\documentclass[unnumsec,namedate,webpdf,traditional,medium]{oup-authoring-template}
%\documentclass[namedate,webpdf,traditional,small]{oup-authoring-template}

\onecolumn % for one column layouts

%\usepackage{showframe}

\graphicspath{{Fig/}}

% line numbers
\linespread{1.5}
% \usepackage[mathlines, switch]{lineno}
% \usepackage[right]{lineno}

\theoremstyle{thmstyleone}%
\newtheorem{theorem}{Theorem}%  meant for continuous numbers
%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
\newtheorem{proposition}[theorem]{Proposition}%
\newtheorem{lemma}{Lemma}
%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.
\theoremstyle{thmstyletwo}%
\newtheorem{example}{Example}%
\newtheorem{remark}{Remark}%
\theoremstyle{thmstylethree}%
\newtheorem{definition}{Definition}

% New operators and commands
\DeclareMathOperator{\E}{\mathbb{E}} % expectation
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\C}{\mathbb{C}}
\renewcommand{\S}{\mathbb{S}}
\newcommand{\blank}{\makebox[1ex]{\textbf{$\cdot$}}}
\newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
\def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
\renewcommand{\phi}{\varphi}
\renewcommand{\epsilon}{\varepsilon}
\newcommand*\diff{\mathop{}\!\mathrm{d}}
\newcommand{\weakly}{\rightsquigarrow}
\newcommand\smallO{\textit{o}}
\newcommand\bigO{\textit{O}}
\newcommand{\midd}{\; \middle|\;}
\newcommand{\1}{\mathds{1}}
\usepackage{ifthen} %% Empirical process with default argument
\newcommand{\G}[2][n]{
{\ensuremath{\mathbb{G}_{#1}}{\left[#2\right]}}
}
\DeclareMathOperator*{\argmin}{\arg\!\min}
\DeclareMathOperator*{\argmax}{\arg\!\max}
\newcommand{\empmeas}{\ensuremath{\mathbb{P}_n}} % empirical measure
\newcommand{\data}{\ensuremath{\mathcal{D}}}

\begin{document}

\journaltitle{Submitted to Biostatistics}
\DOI{xxxx}
\copyrightyear{xxxx}
\pubyear{xxxx}
\access{xxx}
\appnotes{Paper}

\firstpage{1}

%\subtitle{Subject Section}

\title[Short Article Title]{Article Title}

\author[1,$\ast$]{Anders Munch}
\author[1]{Thomas~A.~Gerds}

\authormark{Munch and Gerds}

\address[1]{\orgdiv{Section of Biostatistics}, \orgname{University of
    Copenhagen}, \orgaddress{\street{Østerfarigmagsgade 5}, \postcode{1357}, \state{Copenhagen}, \country{Denmark}}}

\corresp[$\ast$]{Corresponding author. \href{email:a.munch@sund.ku.dk}{a.munch@sund.ku.dk}}

\received{Date}{0}{Year}
\revised{Date}{0}{Year}
\accepted{Date}{0}{Year}

%\editor{Associate Editor: Name}

%\abstract{
%\textbf{Motivation:} .\\
%\textbf{Results:} .\\
%\textbf{Availability:} .\\
%\textbf{Contact:} \href{name@email.com}{name@email.com}\\
%\textbf{Supplementary information:} Supplementary data are available at \textit{Journal Name}
%online.}

\abstract{ In survival analysis, prediction models are needed as stand-alone
  tools and in applications of causal inference to estimate nuisance parameters.
  The super learner is a machine learning algorithm which combines a library of
  prediction models into a meta learner based on cross-validated loss. In
  right-censored data, the choice of the loss function and the estimation of the
  expected loss need careful consideration. We introduce the state learner, a
  new super learner for survival analysis, which simultaneously evaluates
  libraries of prediction models for the event of interest and the censoring
  distribution. The state learner can be applied to all types of survival
  models, works in the presence of competing risks, and does not require a
  single pre-specified estimator of the conditional censoring distribution. We
  establish an oracle inequality for the state learner and investigate its
  performance through numerical experiments. We illustrate the application of
  the state learner with prostate cancer data, as a stand-alone prediction tool,
  and, for causal inference, as a way to estimate the nuisance parameter models
  of a smooth statistical functional. }

\keywords{Competing risks, cross-validation, loss based estimation,
  right-censored data, super learner}

% \boxedtext{
% \begin{itemize}
% \item Key boxed text here.
% \item Key boxed text here.
% \item Key boxed text here.
% \end{itemize}}

\maketitle

\section{Introduction}
The introduction introduces the context and summarizes the manuscript. It is importantly to clearly state the contributions
of this piece of work. Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

This is an example of a new parapgraph with a numbered footnote\footnote{\url{https://data.gov.uk/}} and a second footnote marker.\footnote{Example of footnote text.}

\section{This is an example for first level head - section head}\label{sec2}

Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum (refer Section~\ref{sec5}).

\subsection{This is an example for second level head - subsection head}\label{subsec1}

Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

\subsubsection{This is an example for third level head - subsubsection head}\label{subsubsec1}

Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do
eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. %Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

\paragraph{This is an example for fourth level head - paragraph head}

Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

\section{This is an example for first level head}\label{sec3}

\subsection{This is an example for second level head - subsection head}\label{subsec2}

\subsubsection{This is an example for third level head - subsubsection head}\label{subsubsec2}

Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

\paragraph{This is an example for fourth level head - paragraph head}

Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

\section{Equations}\label{sec4}

Equations in \LaTeX{} can either be inline or set as display equations. For
inline equations use the \verb+$...$+ commands. Eg: the equation
$H\psi = E \psi$ is written via the command \verb+$H \psi = E \psi$+.

For display equations (with auto generated equation numbers)
one can use the equation or eqnarray environments:
\begin{equation}
\|\tilde{X}(k)\|^2 \leq\frac{\sum\limits_{i=1}^{p}\left\|\tilde{Y}_i(k)\right\|^2+\sum\limits_{j=1}^{q}\left\|\tilde{Z}_j(k)\right\|^2 }{p+q},\label{eq1}
\end{equation}
where,
\begin{align}
D_\mu &=  \partial_\mu - ig \frac{\lambda^a}{2} A^a_\mu \nonumber \\
F^a_{\mu\nu} &= \partial_\mu A^a_\nu - \partial_\nu A^a_\mu + g f^{abc} A^b_\mu A^a_\nu.\label{eq2}
\end{align}
Notice the use of \verb+\nonumber+ in the align environment at the end
of each line, except the last, so as not to produce equation numbers on
lines where no equation numbers are required. The \verb+\label{}+ command
should only be used at the last line of an align environment where
\verb+\nonumber+ is not used.
\begin{equation}
Y_\infty = \left( \frac{m}{\textrm{GeV}} \right)^{-3}
    \left[ 1 + \frac{3 \ln(m/\textrm{GeV})}{15}
    + \frac{\ln(c_2/5)}{15} \right].
\end{equation}
The class file also supports the use of \verb+\mathbb{}+, \verb+\mathscr{}+ and
\verb+\mathcal{}+ commands. As such \verb+\mathbb{R}+, \verb+\mathscr{R}+
and \verb+\mathcal{R}+ produces $\mathbb{R}$, $\mathscr{R}$ and $\mathcal{R}$
respectively (refer Subsubsection~\ref{subsubsec3}).

Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do
eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum. Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do
eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum. 
Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do
eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.

\section{Tables}\label{sec5}

Tables can be inserted via the normal table and tabular environment. To put
footnotes inside tables one has to Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum. use the additional ``tablenotes" environment
enclosing the tabular environment. The footnote appears just below the table
itself (refer Tables~\ref{tab1} and \ref{tab2}).

\begin{verbatim}
\begin{table}[t]
\begin{center}
\begin{minipage}{<width>}
\caption{<table-caption>\label{<table-label>}}%
\begin{tabular}{@{}llll@{}}
\toprule
column 1 & column 2 & column 3 & column 4\\
\midrule
row 1 & data 1 & data 2          & data 3 \\
row 2 & data 4 & data 5$^{1}$ & data 6 \\
row 3 & data 7 & data 8      & data 9$^{2}$\\
\botrule
\end{tabular}
\begin{tablenotes}%
\item Source: Example for source.
\item[$^{1}$] Example for a 1st table footnote.
\item[$^{2}$] Example for a 2nd table footnote.
\end{tablenotes}
\end{minipage}
\end{center}
\end{table}
\end{verbatim}

Lengthy tables which do not fit within textwidth should be set as rotated tables. For this, we need to use \verb+\begin{sidewaystable}...+ \verb+\end{sidewaystable}+ instead of\break \verb+\begin{table}...+ \verb+\end{table}+ environment.

\begin{table}[!t]
\caption{Caption text\label{tab1}}%
\begin{tabular*}{\columnwidth}{@{\extracolsep\fill}llll@{\extracolsep\fill}}
\toprule
column 1 & column 2  & column 3 & column 4\\
\midrule
row 1    & data 1   & data 2  & data 3  \\
row 2    & data 4   & data 5$^{1}$  & data 6  \\
row 3    & data 7   & data 8  & data 9$^{2}$  \\
\botrule
\end{tabular*}
\begin{tablenotes}%
\item Source: This is an example of table footnote this is an example of table footnote this is an example of table footnote this is an example of~table footnote this is an example of table footnote
\item[$^{1}$] Example for a first table footnote.
\item[$^{2}$] Example for a second table footnote.
\end{tablenotes}
\end{table}

\begin{table*}[t]
\caption{Example of a lengthy table which is set to full textwidth.\label{tab2}}
\tabcolsep=0pt%%
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lcccccc@{\extracolsep{\fill}}}
\toprule%
& \multicolumn{3}{@{}c@{}}{Element 1$^{1}$} & \multicolumn{3}{@{}c@{}}{Element 2$^{2}$} \\
\cline{2-4}\cline{5-7}%
Project & Energy & $\sigma_{calc}$ & $\sigma_{expt}$ & Energy & $\sigma_{calc}$ & $\sigma_{expt}$ \\
\midrule
Element 3  & 990 A & 1168 & $1547\pm12$ & 780 A & 1166 & $1239\pm100$\\
Element 4  & 500 A & 961  & $922\pm10$  & 900 A & 1268 & $1092\pm40$\\
\botrule
\end{tabular*}
\begin{tablenotes}%
\item Note: This is an example of table footnote this is an example of table footnote this is an example of table footnote this is an example of~table footnote this is an example of table footnote
\item[$^{1}$] Example for a first table footnote.
\item[$^{2}$] Example for a second table footnote.\vspace*{6pt}
\end{tablenotes}
\end{table*}

\begin{sidewaystable}%[!p]
\caption{Tables which are too long to fit, should be written using the ``sidewaystable" environment as shown here\label{tab3}}
\tabcolsep=0pt%
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lcccccc@{\extracolsep{\fill}}}
\toprule%
& \multicolumn{3}{@{}c@{}}{Element 1$^{1}$}& \multicolumn{3}{@{}c@{}}{Element$^{2}$} \\
\cline{2-4}\cline{5-7}%
Projectile & Energy     & $\sigma_{calc}$ & $\sigma_{expt}$ & Energy & $\sigma_{calc}$ & $\sigma_{expt}$ \\
\midrule
Element 3 & 990 A & 1168 & $1547\pm12$ & 780 A & 1166 & $1239\pm100$ \\
Element 4 & 500 A & 961  & $922\pm10$  & 900 A & 1268 & $1092\pm40$ \\
\botrule
\end{tabular*}
\begin{tablenotes}%
\item Note: This is an example of a table footnote this is an example of a table footnote this is an example of a table footnote this is an example of a table footnote this is an example of a table footnote
\item[$^{1}$] This is an example of a table footnote
\end{tablenotes}
\end{sidewaystable}

\section{Figures}\label{sec6}

As per display \LaTeX\ standards one has to use eps images for \verb+latex+ compilation and \verb+pdf/jpg/png+ images for
\verb+pdflatex+ compilation. This is one of the major differences between \verb+latex+
and \verb+pdflatex+. The images should be single-page documents. The command for inserting images
for \verb+latex+ and \verb+pdflatex+ can be generalized. The package used to insert images in \verb+latex/pdflatex+ is the
graphicx package. Figures can be inserted via the normal figure environment as shown in the below example:

\begin{figure}[!t]%
\centering
{\color{black!20}\rule{213pt}{37pt}}
\caption{This is a widefig. This is an example of a long caption this is an example of a long caption  this is an example of a long caption this is an example of a long caption}\label{fig1}
\end{figure}

\begin{figure*}[!t]%
\centering
{\color{black!20}\rule{438pt}{74pt}}
\caption{This is a widefig. This is an example of a long caption this is an example of a long caption  this is an example of a long caption this is an example of a long caption}\label{fig2}
\end{figure*}

\begin{verbatim}
\begin{figure}[t]
        \centering\includegraphics{<eps-file>}
        \caption{<figure-caption>}
        \label{<figure-label>}
\end{figure}
\end{verbatim}

Test text here.

For sample purposes, we have included the width of images in the
optional argument of \verb+\includegraphics+ tag. Please ignore this.
Lengthy figures which do not fit within textwidth should be set in rotated mode. For rotated figures, we need to use \verb+\begin{sidewaysfigure}+ \verb+...+ \verb+\end{sidewaysfigure}+ instead of the \verb+\begin{figure}+ \verb+...+ \verb+\end{figure}+ environment.

\begin{sidewaysfigure}%
\centering
{\color{black!20}\rule{610pt}{102pt}}
\caption{This is an example for a sideways figure. This is an example of a long caption this is an example of a long caption  this is an example of a long caption this is an example of a long caption}\label{fig3}
\end{sidewaysfigure}

\section{Algorithms, Program codes and Listings}\label{sec7}

Packages \verb+algorithm+, \verb+algorithmicx+ and \verb+algpseudocode+ are used for setting algorithms in latex.
For this, one has to use the below format:

\begin{verbatim}
\begin{algorithm}
\caption{<alg-caption>}\label{<alg-label>}
\begin{algorithmic}[1]
. . .
\end{algorithmic}
\end{algorithm}
\end{verbatim}

You may need to refer to the above-listed package documentations for more details before setting an \verb+algorithm+ environment.
To set program codes, one has to use the \verb+program+ package. We need to use the \verb+\begin{program}+ \verb+...+
\verb+\end{program}+ environment to set program codes.

\begin{algorithm}[!t]
\caption{Calculate $y = x^n$}\label{algo1}
\begin{algorithmic}[1]
\Require $n \geq 0 \vee x \neq 0$
\Ensure $y = x^n$
\State $y \Leftarrow 1$
\If{$n < 0$}
        \State $X \Leftarrow 1 / x$
        \State $N \Leftarrow -n$
\Else
        \State $X \Leftarrow x$
        \State $N \Leftarrow n$
\EndIf
\While{$N \neq 0$}
        \If{$N$ is even}
            \State $X \Leftarrow X \times X$
            \State $N \Leftarrow N / 2$
        \Else[$N$ is odd]
            \State $y \Leftarrow y \times X$
            \State $N \Leftarrow N - 1$
        \EndIf
\EndWhile
\end{algorithmic}
\end{algorithm}

Similarly, for \verb+listings+, one has to use the \verb+listings+ package. The \verb+\begin{lstlisting}+ \verb+...+ \verb+\end{lstlisting}+ environment is used to set environments similar to the \verb+verbatim+ environment. Refer to the \verb+lstlisting+ package documentation for more details on this.

\begin{minipage}{\hsize}%
\lstset{language=Pascal}% Set your language (you can change the language for each code-block optionally)
\begin{lstlisting}[frame=single,framexleftmargin=-1pt,framexrightmargin=-17pt,framesep=12pt,linewidth=0.98\textwidth]
for i:=maxint to 0 do
begin
{ do nothing }
end;
Write('Case insensitive ');
Write('Pascal keywords.');
\end{lstlisting}
\end{minipage}

\section{Cross referencing}\label{sec8}

Environments such as figure, table, equation, and align can have a label
declared via the \verb+\label{#label}+ command. For figures and table
environments one should use the \verb+\label{}+ command inside or just
below the \verb+\caption{}+ command.  One can then use the
\verb+\ref{#label}+ command to cross-reference them. As an example, consider
the label declared for Figure \ref{fig1} which is
\verb+\label{fig1}+. To cross-reference it, use the command
\verb+ Figure \ref{fig1}+, for which it comes up as
``Figure~\ref{fig1}".

\subsection{Details on reference citations}\label{subsec3}

With standard numerical .bst files, only numerical citations are possible.
With an author-year .bst file, both numerical and author-year citations are possible.

If author-year citations are selected, \verb+\bibitem+ must have one of the following forms:

{\footnotesize%
\begin{verbatim}
\bibitem[Jones et al.(1990)]{key}...
\bibitem[Jones et al.(1990)Jones,
                Baker, and Williams]{key}...
\bibitem[Jones et al., 1990]{key}...
\bibitem[\protect\citeauthoryear{Jones,
                Baker, and Williams}
                {Jones et al.}{1990}]{key}...
\bibitem[\protect\citeauthoryear{Jones et al.}
                {1990}]{key}...
\bibitem[\protect\astroncite{Jones et al.}
                {1990}]{key}...
\bibitem[\protect\citename{Jones et al., }
                1990]{key}...
\harvarditem[Jones et al.]{Jones, Baker, and
                Williams}{1990}{key}...
\end{verbatim}}

This is either to be made up manually, or to be generated by an
appropriate .bst file with BibTeX. Then,

{%
\begin{verbatim}
                    Author-year mode
                        || Numerical mode
\citet{key} ==>>  Jones et al. (1990)
                        || Jones et al. [21]
\citep{key} ==>> (Jones et al., 1990) || [21]
\end{verbatim}}

\noindent
Multiple citations as normal:

{%
\begin{verbatim}
\citep{key1,key2} ==> (Jones et al., 1990;
                         Smith, 1989)||[21,24]
        or (Jones et al., 1990, 1991)||[21,24]
        or (Jones et al., 1990a,b)   ||[21,24]
\end{verbatim}}

\noindent
\verb+\cite{key}+ is the equivalent of \verb+\citet{key}+ in author-year mode
and  of \verb+\citep{key}+ in numerical mode. Full author lists may be forced with
\verb+\citet*+ or \verb+\citep*+, e.g.

{%
\begin{verbatim}
\citep*{key} ==>> (Jones, Baker, and Mark, 1990)
\end{verbatim}}

\noindent
Optional notes as:

{%
\begin{verbatim}
\citep[chap. 2]{key}     ==>>
        (Jones et al., 1990, chap. 2)
\citep[e.g.,][]{key}     ==>>
        (e.g., Jones et al., 1990)
\citep[see][pg. 34]{key} ==>>
        (see Jones et al., 1990, pg. 34)
\end{verbatim}}

\noindent
(Note: in standard LaTeX, only one note is allowed, after the ref.
Here, one note is like the standard, two make pre- and post-notes.)

{%
\begin{verbatim}
\citealt{key}   ==>> Jones et al. 1990
\citealt*{key}  ==>> Jones, Baker, and
                        Williams 1990
\citealp{key}   ==>> Jones et al., 1990
\citealp*{key}  ==>> Jones, Baker, and
                        Williams, 1990
\end{verbatim}}

\noindent
Additional citation possibilities (both author-year and numerical modes):

{%
\begin{verbatim}
\citeauthor{key}       ==>> Jones et al.
\citeauthor*{key}      ==>> Jones, Baker, and
                                Williams
\citeyear{key}         ==>> 1990
\citeyearpar{key}      ==>> (1990)
\citetext{priv. comm.} ==>> (priv. comm.)
\citenum{key}          ==>> 11 [non-superscripted]
\end{verbatim}}

\noindent
Note: full author lists depend on whether the bib style supports them;
if not, the abbreviated list is printed even when full is requested.

\noindent
For names like della Robbia at the start of a sentence, use

{%
\begin{verbatim}
\Citet{dRob98}      ==>> Della Robbia (1998)
\Citep{dRob98}      ==>> (Della Robbia, 1998)
\Citeauthor{dRob98} ==>> Della Robbia
\end{verbatim}}

\noindent
The following is an example for \verb+\cite{...}+: \cite{rahman2019centroidb}. Another example for \verb+\citep{...}+: \citep{bahdanau2014neural,imboden2018cardiorespiratory,motiian2017unified,murphy2012machine,ji20123d}.
Sample cites here \cite{krizhevsky2012imagenet,horvath2018dna} and \cite{pyrkov2018quantitative}, \cite{wang2018face}, \cite{lecun2015deep,zhang2018fine,ravi2016deep}.

\section{Lists}\label{sec9}

List in \LaTeX{} can be of three types: numbered, bulleted and unnumbered. The ``enumerate'' environment produces a numbered list, the 
``itemize'' environment produces a bulleted list and the ``unlist''
environment produces an unnumbered list.
In each environment, a new entry is added via the \verb+\item+ command.
\begin{enumerate}[1.]
\item This is the 1st item

\item Enumerate creates numbered lists, itemize creates bulleted lists and
unnumerate creates unnumbered lists.
\begin{enumerate}[(a)]
\item Second level numbered list. Enumerate creates numbered lists, itemize creates bulleted lists and
description creates unnumbered lists.

\item Second level numbered list. Enumerate creates numbered lists, itemize creates bulleted lists and
description creates unnumbered lists.
\begin{enumerate}[(ii)]
\item Third level numbered list. Enumerate creates numbered lists, itemize creates bulleted lists and
description creates unnumbered lists.

\item Third level numbered list. Enumerate creates numbered lists, itemize creates bulleted lists and
description creates unnumbered lists.
\end{enumerate}

\item Second level numbered list. Enumerate creates numbered lists, itemize creates bulleted lists and
description creates unnumbered lists.

\item Second level numbered list. Enumerate creates numbered lists, itemize creates bulleted lists and
description creates unnumbered lists.
\end{enumerate}

\item Enumerate creates numbered lists, itemize creates bulleted lists and
description creates unnumbered lists.

\item Numbered lists continue.
\end{enumerate}
Lists in \LaTeX{} can be of three types: enumerate, itemize and description.
In each environment, a new entry is added via the \verb+\item+ command.
\begin{itemize}
\item First level bulleted list. This is the 1st item

\item First level bulleted list. Itemize creates bulleted lists and description creates unnumbered lists.
\begin{itemize}
\item Second level dashed list. Itemize creates bulleted lists and description creates unnumbered lists.

\item Second level dashed list. Itemize creates bulleted lists and description creates unnumbered lists.

\item Second level dashed list. Itemize creates bulleted lists and description creates unnumbered lists.
\end{itemize}

\item First level bulleted list. Itemize creates bulleted lists and description creates unnumbered lists.

\item First level bulleted list. Bullet lists continue.
\end{itemize}

\noindent
Example for unnumbered list items:

\begin{unlist}
\item Sample unnumberd list text. Sample unnumberd list text. Sample unnumberd list text. Sample unnumberd list text. Sample unnumberd list text.

\item Sample unnumberd list text. Sample unnumberd list text. Sample unnumberd list text.

\item sample unnumberd list text. Sample unnumberd list text. Sample unnumberd list text. Sample unnumberd list text. Sample unnumberd list text. Sample unnumberd list text. Sample unnumberd list text.
\end{unlist}

\section{Examples for theorem-like environments}\label{sec10}

For theorem-like environments, we require the \verb+amsthm+ package. There are three types of predefined theorem styles - \verb+thmstyleone+, \verb+thmstyletwo+ and \verb+thmstylethree+   (check your journal's instructions page in case a specific style is required).

\medskip
\noindent\begin{tabular}{|l|p{13pc}|}
\hline
\verb+thmstyleone+ & Numbered, theorem head in bold font and theorem text in italic style \\\hline
\verb+thmstyletwo+ & Numbered, theorem head in roman font and theorem text in italic style \\\hline
\verb+thmstylethree+ & Numbered, theorem head in bold font and theorem text in roman style \\\hline
\end{tabular}

\begin{theorem}[Theorem subhead]\label{thm1}
Example theorem text. Example theorem text. Example theorem text. Example theorem text. Example theorem text.
Example theorem text. Example theorem text. Example theorem text. Example theorem text. Example theorem text.
Example theorem text.
\end{theorem}

Quisque ullamcorper placerat ipsum. Cras nibh. Morbi vel justo vitae lacus tincidunt ultrices. Lorem ipsum dolor sit
amet, consectetuer adipiscing elit. In hac habitasse platea dictumst. Integer tempus convallis augue.

\begin{proposition}
Example proposition text. Example proposition text. Example proposition text. Example proposition text. Example proposition text.
Example proposition text. Example proposition text. Example proposition text. Example proposition text. Example proposition text.
\end{proposition}

Nulla malesuada porttitor diam. Donec felis erat, congue non, volutpat at, tincidunt tristique, libero. Vivamus
viverra fermentum felis. Donec nonummy pellentesque ante.

\begin{example}
Phasellus adipiscing semper elit. Proin fermentum massa
ac quam. Sed diam turpis, molestie vitae, placerat a, molestie nec, leo. Maecenas lacinia. Nam ipsum ligula, eleifend
at, accumsan nec, suscipit a, ipsum. Morbi blandit ligula feugiat magna. Nunc eleifend consequat lorem.
\end{example}

Nulla malesuada porttitor diam. Donec felis erat, congue non, volutpat at, tincidunt tristique, libero. Vivamus
viverra fermentum felis. Donec nonummy pellentesque ante.

\begin{remark}
Phasellus adipiscing semper elit. Proin fermentum massa
ac quam. Sed diam turpis, molestie vitae, placerat a, molestie nec, leo. Maecenas lacinia. Nam ipsum ligula, eleifend
at, accumsan nec, suscipit a, ipsum. Morbi blandit ligula feugiat magna. Nunc eleifend consequat lorem.
\end{remark}

Quisque ullamcorper placerat ipsum. Cras nibh. Morbi vel justo vitae lacus tincidunt ultrices. Lorem ipsum dolor sit
amet, consectetuer adipiscing elit. In hac habitasse platea dictumst.

\begin{definition}[Definition sub head]
Example definition text. Example definition text. Example definition text. Example definition text. Example definition text. Example definition text. Example definition text. Example definition text.
\end{definition}

Apart from the above styles, we have the \verb+\begin{proof}+ \verb+...+ \verb+\end{proof}+ environment - with the proof head in italic style and the body text in roman font with an open square at the end of each proof environment.

\begin{proof}Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text.
\end{proof}

Nam dui ligula, fringilla a, euismod sodales, sollicitudin vel, wisi. Morbi auctor lorem non justo. Nam lacus libero,
pretium at, lobortis vitae, ultricies et, tellus. Donec aliquet, tortor sed accumsan bibendum, erat ligula aliquet magna,
vitae ornare odio metus a mi.

\begin{proof}[Proof of Theorem~{\upshape\ref{thm1}}]
Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text.
\end{proof}

\noindent
For a quote environment, one has to use\newline \verb+\begin{quote}...\end{quote}+
\begin{quote}
Quoted text example. Aliquam porttitor quam a lacus. Praesent vel arcu ut tortor cursus volutpat. In vitae pede quis diam bibendum placerat. Fusce elementum
convallis neque. Sed dolor orci, scelerisque ac, dapibus nec, ultricies ut, mi. Duis nec dui quis leo sagittis commodo.
\end{quote}
Donec congue. Maecenas urna mi, suscipit in, placerat ut, vestibulum ut, massa. Fusce ultrices nulla et nisl (refer Figure~\ref{fig3}). Pellentesque habitant morbi tristique senectus et netus et malesuada fames ac turpis egestas. Etiam ligula arcu,
elementum a, venenatis quis, sollicitudin sed, metus. Donec nunc pede, tincidunt in, venenatis vitae, faucibus vel (refer Table~\ref{tab3}).

\section{Conclusion}

Some Conclusions here.

%%%%%%%%%%%%%%

\begin{appendices}

\section{Section title of first appendix}\label{sec11}

Nam dui ligula, fringilla a, euismod sodales, sollicitudin vel, wisi. Morbi auctor lorem non justo. Nam lacus libero,
pretium at, lobortis vitae, ultricies et, tellus. Donec aliquet, tortor sed accumsan bibendum, erat ligula aliquet magna,
vitae ornare odio metus a mi. Morbi ac orci et nisl hendrerit mollis. Suspendisse ut massa. Cras nec ante. Pellentesque
a nulla. Cum sociis natoque penatibus et magnis dis parturient montes, nascetur ridiculus mus. Aliquam tincidunt
urna. Nulla ullamcorper vestibulum turpis. Pellentesque cursus luctus mauris.

\subsection{Subsection title of first appendix}\label{subsec4}

Nam dui ligula, fringilla a, euismod sodales, sollicitudin vel, wisi. Morbi auctor lorem non justo. Nam lacus libero,
pretium at, lobortis vitae, ultricies et, tellus. Donec aliquet, tortor sed accumsan bibendum, erat ligula aliquet magna,
vitae ornare odio metus a mi. Morbi ac orci et nisl hendrerit mollis. Suspendisse ut massa. Cras nec ante. Pellentesque
a nulla. Cum sociis natoque penatibus et magnis dis parturient montes, nascetur ridiculus mus. Aliquam tincidunt
urna. Nulla ullamcorper vestibulum turpis. Pellentesque cursus luctus mauris.

\subsubsection{Subsubsection title of first appendix}\label{subsubsec3}

Example for an unnumbered figure:

\begin{figure}[!h]
\centering
{\color{black!20}\rule{85pt}{92pt}}
\end{figure}

Fusce mauris. Vestibulum luctus nibh at lectus. Sed bibendum, nulla a faucibus semper, leo velit ultricies tellus, ac
venenatis arcu wisi vel nisl. Vestibulum diam. Aliquam pellentesque, augue quis sagittis posuere, turpis lacus congue
quam, in hendrerit risus eros eget felis.

\section{Section title of second appendix}\label{sec12}%

Fusce mauris. Vestibulum luctus nibh at lectus. Sed bibendum, nulla a faucibus semper, leo velit ultricies tellus, ac
venenatis arcu wisi vel nisl. Vestibulum diam. Aliquam pellentesque, augue quis sagittis posuere, turpis lacus congue
quam, in hendrerit risus eros eget felis. Maecenas eget erat in sapien mattis porttitor. Vestibulum porttitor. Nulla
facilisi. Sed a turpis eu lacus commodo facilisis. Morbi fringilla, wisi in dignissim interdum, justo lectus sagittis dui, et
vehicula libero dui cursus dui. Mauris tempor ligula sed lacus. Duis cursus enim ut augue. Cras ac magna. Cras nulla.

\begin{figure}[b]
\centering
{\color{black!20}\rule{217pt}{120pt}}
\caption{This is an example for appendix figure\label{fig4}}
\end{figure}

\begin{table}[t]%
\begin{center}
\begin{minipage}{.52\columnwidth}
\caption{This is an example of Appendix table showing food requirements of army, navy and airforce\label{tab4}}%
\begin{tabular}{@{}lcc@{}}%
\toprule
col1 head & col2 head & col3 head \\
\midrule
col1 text & col2 text & col3 text \\
col1 text & col2 text & col3 text \\
col1 text & col2 text & col3 text \\
\botrule
\end{tabular}
\end{minipage}
\end{center}
\end{table}

\subsection{Subsection title of second appendix}\label{subsec5}

Sed commodo posuere pede. Mauris ut est. Ut quis purus. Sed ac odio. Sed vehicula hendrerit sem. Duis non odio.
Morbi ut dui. Sed accumsan risus eget odio. In hac habitasse platea dictumst. Pellentesque non elit. Fusce sed justo
eu urna porta tincidunt. Mauris felis odio, sollicitudin sed, volutpat a, ornare ac, erat. Morbi quis dolor. Donec
pellentesque, erat ac sagittis semper, nunc dui lobortis purus, quis congue purus metus ultricies tellus. Proin et quam.
Class aptent taciti sociosqu ad litora torquent per conubia nostra, per inceptos hymenaeos. Praesent sapien turpis,
fermentum vel, eleifend faucibus, vehicula eu, lacus.

Sed commodo posuere pede. Mauris ut est. Ut quis purus. Sed ac odio. Sed vehicula hendrerit sem. Duis non odio.
Morbi ut dui. Sed accumsan risus eget odio. In hac habitasse platea dictumst. Pellentesque non elit. Fusce sed justo
eu urna porta tincidunt. Mauris felis odio, sollicitudin sed, volutpat a, ornare ac, erat. Morbi quis dolor. Donec
pellentesque, erat ac sagittis semper, nunc dui lobortis purus, quis congue purus metus ultricies tellus. Proin et quam.
Class aptent taciti sociosqu ad litora torquent per conubia nostra, per inceptos hymenaeos. Praesent sapien turpis,
fermentum vel, eleifend faucibus, vehicula eu, lacus.

\subsubsection{Subsubsection title of second appendix}\label{subsubsec4}

Lorem ipsum dolor sit amet, consectetuer adipiscing elit. Ut purus elit, vestibulum ut, placerat ac, adipiscing vitae,
felis. Curabitur dictum gravida mauris. Nam arcu libero, nonummy eget, consectetuer id, vulputate a, magna. Donec
vehicula augue eu neque.

Example for an equation inside the appendix:
\begin{align}
 p &= \frac{\gamma^{2} - (n_{C} -1)H}{(n_{C} - 1) + H - 2\gamma}, \label{1eq:hybobo:pfromgH} \\
 \theta &= \frac{(\gamma - H)^{2}(\gamma - n_{C} -1)^{2}}{(n_{C} - 1 + H - 2\gamma)^{2}} \label{2eq:hybobo:tfromgH}\; .
\end{align}

\section{Example of another appendix section}\label{sec13}%

Nam dui ligula, fringilla a, euismod sodales, sollicitudin vel, wisi. Morbi auctor lorem non justo. Nam lacus libero,
pretium at, lobortis vitae, ultricies et, tellus. Donec aliquet, tortor sed accumsan bibendum, erat ligula aliquet magna,
vitae ornare odio metus a mi. Morbi ac orci et nisl hendrerit mollis. Suspendisse ut massa. Cras nec ante. Pellentesque
a nulla. Cum sociis natoque penatibus et magnis dis parturient montes, nascetur ridiculus mus. Aliquam tincidunt
urna. Nulla ullamcorper vestibulum turpis. Pellentesque cursus luctus mauris
\begin{equation}
\mathcal{L} = i \bar{\psi} \gamma^\mu D_\mu \psi
    - \frac{1}{4} F_{\mu\nu}^a F^{a\mu\nu} - m \bar{\psi} \psi.
\label{eq26}
\end{equation}

Nulla malesuada porttitor diam. Donec felis erat, congue non, volutpat at, tincidunt tristique, libero. Vivamus
viverra fermentum felis. Donec nonummy pellentesque ante. Phasellus adipiscing semper elit. Proin fermentum massa
ac quam. Sed diam turpis, molestie vitae, placerat a, molestie nec, leo. Maecenas lacinia. Nam ipsum ligula, eleifend
at, accumsan nec, suscipit a, ipsum. Morbi blandit ligula feugiat magna. Nunc eleifend consequat lorem. Sed lacinia
nulla vitae enim. Pellentesque tincidunt purus vel magna. Integer non enim. Praesent euismod nunc eu purus. Donec
bibendum quam in tellus. Nullam cursus pulvinar lectus. Donec et mi. Nam vulputate metus eu enim. Vestibulum
pellentesque felis eu massa.

Nulla malesuada porttitor diam. Donec felis erat, congue non, volutpat at, tincidunt tristique, libero. Vivamus
viverra fermentum felis. Donec nonummy pellentesque ante. Phasellus adipiscing semper elit. Proin fermentum massa
ac quam. Sed diam turpis, molestie vitae, placerat a, molestie nec, leo. Maecenas lacinia. Nam ipsum ligula, eleifend
at, accumsan nec, suscipit a, ipsum. Morbi blandit ligula feugiat magna. Nunc eleifend consequat lorem. Sed lacinia
nulla vitae enim. Pellentesque tincidunt purus vel magna. Integer non enim. Praesent euismod nunc eu purus. Donec
bibendum quam in tellus. Nullam cursus pulvinar lectus. Donec et mi. Nam vulputate metus eu enim. Vestibulum
pellentesque felis eu massa.

Nulla malesuada porttitor diam. Donec felis erat, congue non, volutpat at, tincidunt tristique, libero. Vivamus
viverra fermentum felis. Donec nonummy pellentesque ante. Phasellus adipiscing semper elit. Proin fermentum massa
ac quam. Sed diam turpis, molestie vitae, placerat a, molestie nec, leo. Maecenas lacinia. Nam ipsum ligula, eleifend
at, accumsan nec, suscipit a, ipsum. Morbi blandit ligula feugiat magna. Nunc eleifend consequat lorem. Sed lacinia
nulla vitae enim. Pellentesque tincidunt purus vel magna. Integer non enim. Praesent euismod nunc eu purus. Donec
bibendum quam in tellus. Nullam cursus pulvinar lectus. Donec et mi. Nam vulputate metus eu enim. Vestibulum
pellentesque felis eu massa.

Nulla malesuada porttitor diam. Donec felis erat, congue non, volutpat at, tincidunt tristique, libero. Vivamus
viverra fermentum felis. Donec nonummy pellentesque ante. Phasellus adipiscing semper elit. Proin fermentum massa
ac quam. Sed diam turpis, molestie vitae, placerat a, molestie nec, leo. Maecenas lacinia. Nam ipsum ligula, eleifend
at, accumsan nec, suscipit a, ipsum. Morbi blandit ligula feugiat magna. Nunc eleifend consequat lorem. Sed lacinia
nulla vitae enim. Pellentesque tincidunt purus vel magna. Integer non enim. Praesent euismod nunc eu purus. Donec
bibendum quam in tellus. Nullam cursus pulvinar lectus. Donec et mi. Nam vulputate metus eu enim. Vestibulum
pellentesque felis eu massa. Donec
bibendum quam in tellus. Nullam cursus pulvinar lectus. Donec et mi. Nam vulputate metus eu enim. Vestibulum
pellentesque felis eu massa.

%% Example for unnumbered table inside appendix
\begin{table}
\begin{center}
\begin{minipage}{.52\columnwidth}
\caption{}{%
\begin{tabular}{lcc}%
\toprule
col1 head & col2 head & col3 head \\
\midrule
col1 text & col2 text & col3 text \\
col1 text & col2 text & col3 text \\
col1 text & col2 text & col3 text \\
\botrule
\end{tabular}}{}
\end{minipage}
\end{center}
\end{table}

\end{appendices}

\section{Competing interests}
No competing interest is declared.

\section{Author contributions statement}

Must include all authors, identified by initials, for example:
S.R. and D.A. conceived the experiment(s),  S.R. conducted the experiment(s), S.R. and D.A. analysed the results.  S.R. and D.A. wrote and reviewed the manuscript.

\section{Acknowledgments}
The authors thank the anonymous reviewers for their valuable suggestions. This work is supported in part by funds from the National Science Foundation (NSF: \# 1636933 and \# 1920920).

%\bibliographystyle{plain}
%\bibliography{reference}

\begin{thebibliography}{10}

\bibitem{bahdanau2014neural}
Dzmitry Bahdanau, Kyunghyun Cho, and Yoshua Bengio.
\newblock Neural machine translation by jointly learning to align and
  translate.
\newblock {\em arXiv preprint arXiv:1409.0473}, 2014.

\bibitem{horvath2018dna}
Steve Horvath and Kenneth Raj.
\newblock Dna methylation-based biomarkers and the epigenetic clock theory of
  ageing.
\newblock {\em Nature Reviews Genetics}, 19(6):371, 2018.

\bibitem{imboden2018cardiorespiratory}
Mary~T Imboden, Matthew~P Harber, Mitchell~H Whaley, W~Holmes Finch, Derron~L
  Bishop, and Leonard~A Kaminsky.
\newblock Cardiorespiratory fitness and mortality in healthy men and women.
\newblock {\em Journal of the American College of Cardiology},
  72(19):2283--2292, 2018.

\bibitem{ji20123d}
Shuiwang Ji, Wei Xu, Ming Yang, and Kai Yu.
\newblock 3d convolutional neural networks for human action recognition.
\newblock {\em IEEE Transactions on Pattern Analysis and Machine Intelligence},
  35(1):221--231, 2012.

\bibitem{krizhevsky2012imagenet}
Alex Krizhevsky, Ilya Sutskever, and Geoffrey~E Hinton.
\newblock Imagenet classification with deep convolutional neural networks.
\newblock In {\em Advances in Neural Information Processing Systems}, pages
  1097--1105, 2012.

\bibitem{lecun2015deep}
Yann LeCun, Yoshua Bengio, and Geoffrey Hinton.
\newblock Deep learning.
\newblock {\em Nature}, 521(7553):436, 2015.

\bibitem{motiian2017unified}
Saeid Motiian, Marco Piccirilli, Donald~A Adjeroh, and Gianfranco Doretto.
\newblock Unified deep supervised domain adaptation and generalization.
\newblock In {\em Proceedings of the IEEE International Conference on Computer
  Vision}, pages 5715--5725, 2017.

\bibitem{murphy2012machine}
Kevin~P Murphy.
\newblock {\em Machine learning: A probabilistic perspective}.
\newblock MIT press, 2012.

\bibitem{american2013acsm}
American~College of~Sports~Medicine et~al.
\newblock {\em ACSM's guidelines for exercise testing and prescription}.
\newblock Lippincott Williams \& Wilkins, 2013.

\bibitem{pyrkov2018quantitative}
Timothy~V Pyrkov, Evgeny Getmantsev, Boris Zhurov, Konstantin Avchaciov,
  Mikhail Pyatnitskiy, Leonid Menshikov, Kristina Khodova, Andrei~V Gudkov, and
  Peter~O Fedichev.
\newblock Quantitative characterization of biological age and frailty based on
  locomotor activity records.
\newblock {\em Aging (Albany NY)}, 10(10):2973, 2018.

\bibitem{rahman2019centroidb}
Syed~Ashiqur Rahman and Donald Adjeroh.
\newblock Centroid of age neighborhoods: A generalized approach to estimate
  biological age.
\newblock In {\em 2019 IEEE EMBS International Conference on Biomedical \&
  Health Informatics (BHI)}, pages 1--4. IEEE, 2019.

\bibitem{ravi2016deep}
Daniele Rav{\`\i}, Charence Wong, Fani Deligianni, Melissa Berthelot, Javier
  Andreu-Perez, Benny Lo, and Guang-Zhong Yang.
\newblock Deep learning for health informatics.
\newblock {\em IEEE {J}ournal of {B}iomedical and {H}ealth {I}nformatics},
  21(1):4--21, 2016.

\bibitem{wang2018face}
Zongwei Wang, Xu~Tang, Weixin Luo, and Shenghua Gao.
\newblock Face aging with identity-preserved conditional generative adversarial
  networks.
\newblock In {\em Proceedings of the IEEE Conference on Computer Vision and
  Pattern Recognition}, pages 7939--7947, 2018.

\bibitem{zhang2018fine}
Ke~Zhang, Na~Liu, Xingfang Yuan, Xinyao Guo, Ce~Gao, and Zhenbing Zhao.
\newblock Fine-grained age estimation in the wild with attention {LSTM}
  networks.
\newblock {\em arXiv preprint arXiv:1805.10445}, 2018.

\end{thebibliography}

%USE THE BELOW OPTIONS IN CASE YOU NEED AUTHOR YEAR FORMAT.
%\bibliographystyle{abbrvnat}
%\bibliography{reference}

%% sample for biography with author's image
\begin{biography}{{\color{black!20}\rule{77pt}{77pt}}}{\author{Author Name.} This is sample author biography text. The values provided in the optional argument are meant for sample purposes. There is no need to include the width and height of an image in the optional argument for live articles. This is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text.}
\end{biography}

%% sample for biography without author's image
\begin{biography}{}{\author{Author Name.} This is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text.}
\end{biography}

\end{document}

### 2025-01-13 manuscript/suppl.tex
biostats version

--- 2024-12-21 manuscript/manuscript.tex
@@
 %% Title
-\title{\Large The state learner -- a super learner for right-censored data}
+\title{\Large Supplementary material: The state learner -- a super learner for right-censored data}
 \date{\small \today}
@@
 
+\renewcommand{\thesection}{S\arabic{section}}
+
 \begin{document}
@@
 
-\begin{abstract}
-  In survival analysis, prediction models are needed as stand-alone tools and in
-  applications of causal inference to estimate nuisance parameters. The super
-  learner is a machine learning algorithm which combines a library of prediction
-  models into a meta learner based on cross-validated loss. In right-censored
-  data, the choice of the loss function and the estimation of the expected loss
-  need careful consideration. We introduce the state learner, a new super
-  learner for survival analysis, which simultaneously evaluates libraries of
-  prediction models for the event of interest and the censoring distribution.
-  The state learner can be applied to all types of survival models, works in the
-  presence of competing risks, and does not require a single pre-specified
-  estimator of the conditional censoring distribution. We establish an oracle
-  inequality for the state learner and investigate its performance through
-  numerical experiments. We illustrate the application of the state learner with
-  prostate cancer data, as a stand-alone prediction tool, and, for causal
-  inference, as a way to estimate the nuisance parameter models of a smooth
-  statistical functional.
-\end{abstract}
+This document contains proofs of the results stated in the main paper.
+Section~\ref{sec:consistency} contains a proof of the consistency result from
+Section~5.1 of the main paper; Section~\ref{sec:oracle-inequalities} contains
+proofs of the oracle inequalities from Section~5.2 of the main paper; and
+Section~\ref{sec:state-learner-with} demonstrates transience of the second order
+remainder structure stated in Section~5.3 of the main paper.
 
-% \keywords{Competing risks, cross-validation, loss based estimation,
-% right-censored data, super learner}
-
-\textit{{\small \textbf{Keywords:} Competing risks, cross-validation, loss based
-    estimation, right-censored data, super learner}}
-
-\pagebreak
-
-\section{Introduction}
-\label{sec:introduction}
-
-% Medical decision making is traditionally based on the discrepancy
-% between intuitive judgment and the law of chance
-% \citep{redelmeier1995probability}.
-% We are concerned with the statistical models that provide
-% probabilistic predictions about the medical future of a patient. We
-% have two different applications of these models in mind. One is as a
-% stand-alone tool which predicts the likelihood of medical outcomes.
-% The other is as nuisance parameter estimate
-% input to an algorithm which estimates a causal target
-% parameter such as an average treatment effect and depends on estimates
-% of nuisance parameters such as the conditional probability of being
-% uncensored.  An example of the latter is the targeted minimum loss
-% based estimator \citep{rytgaard2022targeted}. In both applications,
-% the models have to be learned based on available data from previous
-% patients.
-A super learner is a machine learning algorithm that combines a finite
-set of learners into a meta learner by estimating prediction
-performance in hold-out samples using a pre-specified loss function
-\citep{van2007super}. When the aim is to make a prediction model,
-super learners combine strong learners, such as Cox regression models
-and random survival forests \citep[][Section 8.4]{gerds2021medical}.
-% For targeted learning of a low dimensional
-% target parameter in presence of a high-dimensional nuisance parameter,
-% a super learner will usually also include the highly adaptive lasso
-% \citep{van2011targeted, benkeser2016highly,van2017generally}.
-While the general idea of combining strong learners based on cross-validation
-stems from earlier work \citep{wolpert1992stacked,breiman1996stacked}, the
-name super learner is justified by an oracle inequality
-\citep{van2003unicv,vaart2006oracle}.
-% The super learner is guaranteed to perform
-% almost as well as the model which minimises the expected performance, i.e., the
-% model we would select if we could evaluate the prediction performance in an
-% infinite hold-out sample.
-
-We define the state learner, a new super learner for right-censored data, which
-simultaneously estimates the expected loss of learners of the event time
-distribution and the censoring distribution. Compared to the super learners of
-\cite{polley2011-sl-cens} and \cite{golmakani2020super}, the state learner
-proposed here poses no restrictions on the type of learners that can be included
-in the library. Compared to super learners that rely on inverse probability of
-censoring weighting, censoring unbiased transformations, or pseudo-values, the
-state learner does not require a pre-specified censoring distribution. Compared
-to the super learners by \cite{han2021inverse} and \cite{westling2021inference},
-the state learner has theoretical guarantees which these methods do not have.
-% Moreover, state learner seems to be the first super learner for
-% right censored data which learns the outcome and the censoring
-% distribution simulaneously.
-
-The loss function underlying the state learner operates on coarsened data. In
-the right-censored survival setting the coarsened data consist of the minimum
-and the order of the censoring time and the event time as well as the baseline
-covariates. The state learner works by simultaneously assessing how well
-learners of the event time distribution and the censoring distribution predict
-the states of the coarsened data. To analyse the theoretical properties of the
-state learner we focus on the discrete super learner which combines the library
-of learners by picking the one that minimises the cross-validated loss
-\citep{van2007super}. In the presence of competing risks, our algorithm uses
-separate libraries of learners for the cumulative hazard functions for each of
-the competing risks and for the censoring distribution. We show that the oracle
-selector of the state learner is consistent if all libraries contain a
-consistent learner and prove a finite sample oracle inequality.
-
-The state learner algorithm can output a medical risk prediction model
-\citep{gerds2021medical} which predicts the probability of an event based on
-covariates in the presence of competing risks. Another important application,
-which motivated our contribution, is in targeted learning where conditional
-event probabilities occur as high-dimensional nuisance parameters which need to
-be estimated at a certain rate \citep{van2011targeted, rytgaard2021estimation,
-  rytgaard2022targeted}. The state learner is particularly well-suited for this
-application as it simultaneously provides estimates for both the outcome and the
-censoring, both of which are needed for targeted learning. For the asymptotic
-bias term of a targeted estimator, which uses the state learner to estimate
-nuisance parameters, we show that a second order product structure holds. We
-illustrate both applications of the state learner with prostate cancer data
-\citep{kattan2000pretreatment}.
-
-There are many existing super learners for right-censored data. We briefly
-review them here. Machine learning based on right-censored data commonly uses
-the partial log-likelihood as a loss function
-\citep[e.g.,][]{li2016regularized,yao2017deep,lee2018deephit,katzman2018deepsurv,gensheimer2019scalable,lee2021boosted,kvamme2021continuous}.
-This loss function is also suggested for super learning with right-censored data
-by \cite{polley2011-sl-cens}, where data is assumed to be observed in discrete
-time. However, the partial log-likelihood loss does not work well with data
-splitting (cross-validation) in continuous time. The reason is that the partial
-log-likelihood loss assigns an infinite value for any learner that predicts
-piece-wise constant cumulative hazard functions when there are event times in
-the test set which do not occur in the learning set. This problem occurs with
-prominent survival learners including the Kaplan-Meier estimator, the random
-survival forest, and the semi-parametric Cox regression model and these learners
-cannot be included in the library of the super learner proposed by
-\cite{polley2011-sl-cens}. When a proportional hazards model is assumed, the
-baseline hazard function can be profiled out of the likelihood
-\citep{cox1972regression}. The cross-validated partial log-likelihood loss
-\citep{verweij1993cross} has therefore been suggested as a loss function for
-super learning by \cite{golmakani2020super}. This choice of loss function
-restricts the library of learners to include only Cox proportional hazards
-models, and hence excludes many learners such as, e.g., random survival forests,
-additive hazards models, and accelerated failure time models. Other proposals
-for super learning with right-censored data use an inverse probability of
-censoring weighted (IPCW) loss function
-\citep{graf1999assessment,van2003unicv,molinaro2004tree,keles2004asymptotically,hothorn2006survival,gerds2006consistent,gonzalez2021stacked},
-censoring unbiased transformations
-\citep{fan1996local,steingrimsson2019censoring}, or pseudo-values
-\citep{andersen2003generalised,mogensen2013random,sachs2019ensemble}. All these
-methods rely on an estimator of the censoring distribution, and their drawback
-is that this estimator has to be pre-specified. An approach which avoids a
-pre-specified censoring model was proposed independently by
-\cite{han2021inverse} and \cite{westling2021inference}. In both articles, the
-authors suggest to iterate between learning of the outcome model and learning of
-the censoring model using IPCW loss functions. However, it seems that this
-procedure is in general not guaranteed to converge to the true data-generating
-mechanism \citep[][Appendix~A.4]{munch2024thesis}.
-
-We introduce our notation and framework in
-Section~\ref{sec:framework}. In Section~\ref{sec:super-learning} we define
-general super learning for right-censored data.
-Section~\ref{sec:super-learner-simple} introduces the state learner, and
-Section~\ref{sec:theor-results-prop} provides theoretical guarantees. In
-Section~\ref{sec:targeted-learning} we discuss the use of the state learner in
-the context of targeted learning. We report results of our numerical experiments
-in Section~\ref{sec:numer-exper} and analyse a prostate cancer data set in
-Section~\ref{sec:real-data-appl}. Section~\ref{sec:discussion} contains a
-discussion of the merits and limitations of our proposal.
-Appendices~\ref{sec:proof-proposition} and~\ref{sec:state-learner-with} contain
-proofs. An implementation of the state learner is available at
-\url{https://github.com/amnudn/statelearner} along with codes for reproducing
-our numerical experiments.
-
-\section{Notation and framework}
-\label{sec:framework}
-
-In a competing risk framework \citep{andersen2012statistical}, let
-\( T\) be a time to event variable, \(D\in\{1,2\}\) the cause of the
-event, and $X \in \mathcal{X}$ a vector of baseline covariates taking
-values in a bounded subset \( \mathcal{X} \subset \R^p \),
-\( p\in\N \). Let $\tau< \infty$ be the prediction horizon. We use
-\( \mathcal{Q} \) to denote the collection of all probability measures
-on \( [0,\tau] \times \{1,2\}\times \mathcal{X} \) such that
-\( (T, D, X) \sim Q \) for some unknown \( Q \in \mathcal{Q} \). For
-\(j\in\{1,2\}\), the cause-specific conditional cumulative hazard
-functions are defined by
-\( \Lambda_{j} \colon [0, \tau] \times \mathcal{X} \rightarrow \R_+ \)
-such that
-\begin{equation*}
-  % \label{eq:cum-haz}
-  \Lambda_{j}(t \mid x) = \int_0^t\frac{  Q(T \in \diff s, D=j \mid X=x )}{Q(T \geq s \mid X=x )}.
-\end{equation*}
-For ease of presentation we assume throughout that the map
-\( t\mapsto \Lambda_j(t \mid x) \) is continuous for all \( x \) and
-\( j \). This is not a limitation: All arguments carry over directly
-to the general case. We denote by \(S\) the conditional event-free
-survival function:
-\begin{equation}
-  \label{eq:surv-def}
-  S(t \mid x)=\exp\left\{-\Lambda_{1}(t \mid x)-\Lambda_{2}(t \mid x)\right\}.
-\end{equation}
-Let \( \mathcal{M}_{\tau}\) denote the space of all conditional cumulative hazard
-functions on \( [0,\tau] \times\mathcal{X}\). Any distribution
-\( Q \in \mathcal{Q} \) can be characterised by
-\begin{equation*}
-  \label{eq:parametrizeQ}
-  \begin{split}
-    Q(\diff t,j,\diff x)=& \left\{S(t- \mid x)\Lambda_1(\diff t \mid x)H(\diff x)\right\}^{\1{\{j=1\}}}\\
-                         &  \left\{S(t- \mid x)\Lambda_2(\diff t \mid x)H(\diff x)\right\}^{\1{\{j=2\}}},
-  \end{split}
-\end{equation*}
-where \(\Lambda_{j} \in \mathcal{M}_{\tau}\) for \(j=1,2\) and \(H\) is the marginal
-distribution of the covariates.
-
-We consider the right-censored setting in which we observe the coarsened data
-\(O = (\tilde{T},\tilde D, X)\), where $\tilde T = \min(T,C)$ for a
-right-censoring time \(C\), $\Delta = \1{\{T \leq C\}}$, and
-\(\tilde D=\Delta D\). Let \(\mathcal{P}\) denote a set of probability
-measures on the sample space
-\(\mathcal{O} = [0, \tau] \times \{0, 1, 2\} \times \mathcal{X}\) such
-that \(O \sim P \) for some unknown \(P\in \mathcal{P}\). We assume
-that the event times and the censoring times are conditionally
-independent given covariates, \( T \independent C \mid X \). This
-implies that any distribution \( P \in \mathcal{P} \) is characterised
-by a distribution \( Q \in \mathcal{Q} \) and a conditional cumulative
-hazard function for \( C \) given \( X \)
-\citep[c.f.,][]{begun1983information,gill1997coarsening}. We use
-\(\Gamma\in\mathcal{M}_{\tau}\) to denote the conditional cumulative hazard
-function for censoring. For ease of presentation we now also assume
-that \( \Gamma(\blank \mid x) \) is continuous for all \( x \). We let
-\((t,x)\mapsto G(t \mid x)=\exp\left\{-\Gamma(t \mid x)\right\}\) denote the
-survival function of the conditional censoring distribution. The
-distribution \( P \) is characterised by
-\begin{equation}\label{eq:parametrizeP}
-  \begin{split}
-    P(\diff t, j, \diff x) =& \left\{G(t- \mid x)S(t- \mid x)\Lambda_1(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=1\}}}}\\
-                            & \left\{G(t- \mid x)S(t- \mid x)\Lambda_2(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=2\}}}}\\
-                            & \left\{G(t- \mid x)S(t- \mid x)\Gamma(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=0\}}}}\\
-    = & \left\{G(t- \mid x)Q(\diff t,j,\diff x)\right\}^{\1{{\{j\ne 0\}}}}\\    
-                            & \left\{G(t- \mid x)S(t- \mid x)\Gamma(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=0\}}}}.
-  \end{split}
-\end{equation}
-Hence, we may write
-\( \mathcal{P} = \{ P_{Q, \Gamma} : Q \in \mathcal{Q}, \Gamma \in
-\mathcal{G} \} \) for some \( \mathcal{G} \subset \mathcal{M}_{\tau} \). We
-also have \(H\)-almost everywhere
-\begin{equation*}
-P(\tilde T>t \mid X=x) = S(t \mid x)G(t \mid x) = \exp\left\{-\Lambda_{1}(t \mid x)-\Lambda_{2}(t \mid x)-\Gamma(t \mid x) \right\}.
-\end{equation*}
-We further assume that there exists \(\kappa<\infty\) such that
-\(\Lambda_{j}(\tau- \mid x)<\kappa \), for \(j\in\{1,2\}\), and
-\(\Gamma(\tau- \mid x)<\kappa\) for almost all \(x\in\mathcal X\). Note that this
-implies that \(G(\tau- \mid x)\) is bounded away from zero for almost all \(x\in\mathcal X\).
-Under these assumptions, the conditional cumulative hazard functions
-\(\Lambda_{j}\) and \(\Gamma\) can be identified from \(P\) by
-\begin{align}
-  \Lambda_{j}(t \mid x) &= \int_0^t\frac{  P(\tilde T \in \diff s, \tilde D=j \mid X=x )}{P(\tilde T \geq s \mid X=x )}, \label{eq:lambdaj}\\
-  \Gamma(t \mid x) &= \int_0^t\frac{  P(\tilde T \in \diff s, \tilde D=0 \mid X=x )}{P(\tilde T \geq s \mid X=x )}\label{eq:gamma}.
-\end{align}
-Thus, we can consider $\Lambda_j$ and \(\Gamma\) as operators which map from
-\( \mathcal{P} \) to \(\mathcal M_{\tau}\).
-
-\section{The concept of super learning}
-\label{sec:super-learning}
-
-In survival analysis, a super learner can be used to estimate a
-parameter $\Psi$ which can be identified from the observed data
-distribution \(P\in\mathcal P\). In this section, to introduce the
-discrete super learner and the oracle learner, we consider estimation of the
-function-valued parameter \(\Psi:\mathcal P\to\mathcal M_{\tau}\), given by
-\(\Psi(P)=\Lambda_j\). This parameter is identified via equation
-\eqref{eq:lambdaj} on \([0,\tau]\).
-
-As input to the super learner we need a data set
-\( \data_n=\{O_i\}_{i=1}^n \) of i.i.d.\ observations from \( P \in \mathcal{P} \) and a collection of candidate
-learners $\mathcal{A}$. Each learner \(a \in \mathcal{A}\) is a map
-\( a \colon \mathcal{O}^n \rightarrow \mathcal{M}_{\tau}\) which takes a data
-set as input and returns an estimate $a(\data_n) \in \mathcal{M}_{\tau}$ of
-$\Lambda_{j}$.
-% Formally, the domain of $a$ depends on \(n\) but we suppress this in the notation.
-In what follows, we use the short-hand notation
-\(P[f] = \int f(o) P(\diff o) \). A super learner evaluates the
-performance of \(a \in \mathcal{A}\) with a loss function
-\(L\colon \mathcal{M}_{\tau} \times \mathcal{O} \rightarrow \R_+\) and
-estimates the expected loss \(P[L(a(\data_n), \blank)]\) using
-cross-validation. Specifically, the expected loss of $a\in\mathcal A$
-is estimated by splitting the data set $\data_n$ into $K$ disjoint
-approximately equally sized subsets
-\(\data_n^1, \data_n^2, \dots, \data_n^K \) and then calculating the
-cross-validated loss
-\begin{equation*}
-  % \label{eq:cv-risk-est}
-  \hat{R}_n(a; L) =
-  \frac{1}{K}\sum_{k=1}^{K}
-  % \empmeas^k{[L {(a{ (\data_n^{-k})} , \blank) }]},
-  \frac{1}{| \data_n^{k} |}\sum_{O_i \in \data_n^{k}}
-  L
-  {
-    \left(
-      a{ (\data_n^{-k})}
-      , O_i
-    \right)
-  },
-  \quad \text{with} \quad
-  \data_n^{-k} = \data_n \setminus \data_n^{k}.
-\end{equation*}
-% where \( \empmeas^{k} \) is the empirical measure of \( \data_n^{k} \). 
-The subset \(\data_n^{-k}\) is referred to as the \(k\)'th training
-sample, while \(\data_n^{k}\) is referred to as the \(k\)'th test or
-hold-out sample.
-The discrete super learner is defined as
-\begin{equation*}
-\hat{a}_n = \argmin_{a\in\mathcal A}\hat{R}_n(a; L).
-\end{equation*}
-% The final estimator of \(\Psi(P)=\Lambda_j\) is then the selected
-% learner applied to the full data set, i.e., \(\hat{a}_n(\data_n)\).
-The oracle learner is defined as the learner that minimises the
-expected loss under the data-generating distribution \( P \),
-i.e.,
-\begin{equation*}
-  \tilde{a}_n =
-  \argmin_{a \in \mathcal{A}}
-  \tilde{R}_n(a; L),
-  \quad \text{with} \quad 
-  \tilde{R}_n(a; L)=
-  \frac{1}{K}\sum_{k=1}^{K} 
-  P{
-    \left[
-      L
-      {
-        \left(
-          a{ (\data_n^{-k})}
-          , \blank
-        \right)
-      }
-    \right]}
-  .
-\end{equation*}
-Note that both the discrete super learner and the oracle learner
-depend on the library of learners and on the number of folds \(K\),
-and that the oracle learner is a function of the data and the unknown
-data-generating distribution. However, these dependencies are
-suppressed in the notation.
-
-\section{The state learner}
-\label{sec:super-learner-simple}
-
-% The problem with most existing super learners for right-censored data is that
-% they depend on a pre-specified estimator of the censoring distribution.
-The main idea of the state learner is to jointly use learners for
-\( \Lambda_1 \), \( \Lambda_2 \), and \( \Gamma \), and the relations in
-equation~(\ref{eq:parametrizeP}), to learn a feature of the observed data
-distribution \( P \). The discrete state learner ranks a tuple of learners for
-the tuple of the cumulative hazard functions
-\( (\Lambda_1, \Lambda_2, \Gamma) \) based on how well they jointly model the
-observed data. Risk predictions can then be obtained by combining
-\( \Lambda_1 \) and $\Lambda_2$ from the highest ranked tuple using a well-known
-formula \citep{benichou1990estimates, ozenne2017riskregression}. To formally
-introduce the state learner, we define the multi-state process
-\begin{equation*}
-  \eta(t) = \1\{\tilde{T} \leq t, \tilde D=1\} + 2\,\1\{\tilde{T} \leq t, \tilde
-  D=2\} - \1\{\tilde{T} \leq t, \tilde D=0\},
-  \quad \text{for} \quad t \in [0, \tau].
-\end{equation*}
-At time \(t\), we observe that each individual is in one of four mutually
-exclusive states: \( 0 \), \( 1 \), \( 2 \), or \( -1 \).
-% (Figure \ref{fig:multi-state-process}).
-% \begin{figure}[h]
-  % \centering %
-  % \includeFigCond[width=.5\textwidth]{figure-multi-state-process.pdf}
-  % \caption{Illustration of the multi-state process \(\eta\) used by
-    % the state learner. Note that `censored' is a state, hence the
-    % process is always observed at any time.}
-  % \label{fig:multi-state-process}
-% \end{figure}
-The conditional distribution of the process \( \eta(t) \) given
-baseline covariates \( X \) is determined by the function
-\begin{equation}
-  \label{eq:F-def}
-  F(t, k, x) = P(\eta(t) = k \mid X=x),
-  \quad \text{for all} \quad
-  t \in [0,\tau],
-  k \in \{-1,0,1,2\},
-  x \in \mathcal{X}.
-\end{equation}
-The function \( F \) describes the conditional state occupation probabilities of
-the multi-state process \(\eta\). We construct a super learner for \( F \). The
-target of this super learner is the function-valued parameter $\Psi(P) = F$
-which is identified through equation~(\ref{eq:F-def}). Under conditional
-independent censoring each quadruple $(\Lambda_{1}, \Lambda_{2}, \Gamma, H)$
-characterises a distribution \(P\in\mathcal P\), c.f.\
-equation~\eqref{eq:parametrizeP}, which in turn determines \( (F, H) \). Hence,
-a learner for \( F \) can be constructed from learners for \( \Lambda_1 \),
-\( \Lambda_2 \), and $\Gamma$ as follows:
-\begin{equation}\label{eq:transition}
-  \begin{split}
-    F(t, 0, x)
-    &
-      = P(\tilde{T} > t \mid X= x)
-      =
-      \exp{{\{-\Lambda_{1}(t \mid x)-\Lambda_{2}(t \mid x) - \Gamma(t \mid x)\}
-      }},
-    \\
-    F(t, 1, x)
-    &
-      = P(\tilde{T} \leq t, \tilde{D}=1 \mid X=x)
-      = \int_0^t F(s-, 0, x)  \Lambda_{1}(\diff s \mid x),
-    \\
-    F(t, 2, x)
-    &
-      = P(\tilde{T} \leq t, \tilde{D}=2 \mid X=x)
-      = \int_0^t  F(s-, 0, x)  \Lambda_{2}(\diff s \mid x),
-    \\
-    F(t, -1, x)
-    &
-      = P(\tilde{T} \leq t, \tilde{D}=0 \mid X=x)
-      = \int_0^t F(s-, 0, x)  \Gamma(\diff s \mid x).
-  \end{split}
-\end{equation}
-% \begin{equation}\label{eq:transition}
-%   \begin{split}
-%   F(t, 1, x)
-%     &
-%       % = P(\tilde{T} \leq t, \tilde{D}=1 \mid X=x)
-%     = \int_0^t \exp{{\{-\Lambda_{1}(s \mid x)-\Lambda_{2}(s \mid x) - \Gamma(s \mid x)\} }}  \Lambda_{1}(\diff s \mid x),
-%   \\
-%   F(t, 2, x)
-%     &
-%       % = P(\tilde{T} \leq t, \tilde{D}=2 \mid X=x)
-%       = \int_0^t \exp{\{-\Lambda_{1}(s \mid x)-\Lambda_{2}(s \mid x) - \Gamma(s \mid x)\} }  \Lambda_{2}(\diff s \mid x),
-%     \\
-%     F(t, -1, x)
-%     &
-%       % = P(\tilde{T} \leq t, \tilde{D}=0 \mid X=x)
-%       = \int_0^t \exp{\{-\Lambda_{1}(s \mid x)-\Lambda_{2}(s \mid x) - \Gamma(s \mid x)\} }  \Gamma(\diff s \mid x),
-%     \\
-%     F(t, 0, x)
-%     &
-%       = P(\tilde{T} > t \mid X= x)
-%       = 1- F(t, 1, x) - F(t, 2, x)- F(t, -1, x).
-%   \end{split}
-% \end{equation}
-The state learner requires three libraries of learners (c.f., Section \ref{sec:super-learning}),
-\(\mathcal{A}_1\), \( \mathcal{A}_2 \), and \( \mathcal{B} \), where
-\(\mathcal{A}_1\) and \( \mathcal{A}_2\) contain learners for the
-conditional cause-specific cumulative hazard functions \(\Lambda_1\)
-and \( \Lambda_2\), respectively, and \(\mathcal{B}\) contains
-learners for the conditional cumulative hazard function of the
-censoring distribution. % We further
-% define \(\mathcal{H}\) as the set of all probability distributions on
-% \( \mathcal{X} \) and
-% \begin{equation*}
-%   \mathcal{H}_n=\left\{h\colon\mathcal{O}^n\longrightarrow\mathcal{H} \midd
-%     h(\data_n)=H_n = \frac 1 n \sum_{i=1}^n \delta_{X_i} \right\}
-% \end{equation*}
-% as the library of learners which consists of a single learner, the empirical
-% distribution function.
-Based on the Cartesian product of
-libraries of learners for \((\Lambda_1,\Lambda_2,\Gamma)\) we construct a library
-$\mathcal{F}$ of learners
-for \( F \):
-\begin{align*}
-  \mathcal{F}(\mathcal{A}_1, \mathcal{A}_2, \mathcal{B})
-  &= \{ \phi_{a_1,a_2, b} : a_1 \in \mathcal{A}_1, a_2 \in \mathcal{A}_2, b \in \mathcal{B}\},
-    \intertext{where in correspondence with  the relations in equation
-    \eqref{eq:transition},}
-    \phi_{a_1,a_2, b}(\data_n)(t,0,x)
-  &= \exp{\{-a_1(\data_n)(s \mid x)-a_2(\data_n)(s \mid x) - b(\data_n)(s \mid
-    x)\} },
-  \\
-  \phi_{a_1,a_2, b}(\data_n)(t,1,x)
-  &= \int_0^t
-    \phi_{a_1,a_2, b}(\data_n)(s-,0,x)  a_1(\data_n)(\diff s \mid x),
-  \\
-  \phi_{a_1,a_2, b}(\data_n)(t,2,x)
-  &= \int_0^t\phi_{a_1,a_2, b}(\data_n)(s-,0,x)  a_2(\data_n)(\diff s \mid x),
-  \\
-  \phi_{a_1,a_2, b}(\data_n)(t,-1,x)
-  &= \int_0^t \phi_{a_1,a_2, b}(\data_n)(s-,0,x)  b(\data_n)(\diff s \mid x).
-\end{align*}
-% \begin{align*}
-%   \mathcal{F}(\mathcal{A}_1, \mathcal{A}_2, \mathcal{B})
-%   &= \{ \phi_{a_1,a_2, b} : a_1 \in \mathcal{A}_1, a_2 \in \mathcal{A}_2, b \in \mathcal{B}\},
-%     \intertext{where in correspondence with  the relations in equation \eqref{eq:transition},} 
-%     \phi_{a_1,a_2, b}(\data_n)(t,1,x) &= \int_0^t \exp{\{-a_1(\data_n)(s \mid x)-a_2(\data_n)(s \mid x) - b(\data_n)(s \mid x)\} }  a_1(\data_n)(\diff s \mid x),\\
-%   \phi_{a_1,a_2, b}(\data_n)(t,2,x) &= \int_0^t \exp{\{-a_1(\data_n)(s \mid x)-a_2(\data_n)(s \mid x) - b(\data_n)(s \mid x)\} }  a_2(\data_n)(\diff s \mid x),\\
-%   \phi_{a_1,a_2, b}(\data_n)(t,-1,x) &= \int_0^t \exp{\{-a_1(\data_n)(s \mid x)-a_2(\data_n)(s \mid x) - b(\data_n)(s \mid x)\} }  b(\data_n)(\diff s \mid x),\\
-%   \phi_{a_1,a_2, b}(\data_n)(t,0,x) &= 1
-%                                       - \phi_{a_1,a_2, b}(\data_n)(t,1,x)
-%                                       - \phi_{a_1,a_2, b}(\data_n)(t,2,x)
-%     -\phi_{a_1,a_2, b}(\data_n)(t,-1,x)
-% \end{align*}
-To evaluate how well a function \( F \) predicts the observed
-multi-state process we use the integrated Brier score
-\( \bar B_\tau( F,O) = \int_0^{\tau} B_t(F,O) \diff t \), where \( B_t \) is the
-Brier score \citep{brier1950verification} at time \( t \in [0, \tau] \),
+We start by recalling some notation. We use \( F \) to denote a function that
+describes the conditional state occupation probabilities of the observed data,
+as defined in equation~(5) of the main paper. We denote the integrated Brier
+score as \( \bar B_\tau( F,O) = \int_0^{\tau} B_t(F,O) \diff t \), where
+\( B_t \) is defined
 \begin{equation*}
@@
 \end{equation*}
-Based on a split of a data set \(\data_n\) into $K$ disjoint
-approximately equally sized subsets (see Section \ref{sec:super-learning}), each learner
-\( \phi_{a_1, a_2, b} \) in the library
-\( \mathcal{F}(\mathcal{A}_1, \mathcal{A}_2, \mathcal{B}) \) is
-evaluated using the cross-validated loss,
-\begin{equation*}
-  \hat{R}_{n}(\phi_{a_1,a_2,b} ; \bar{B}_{\tau}) =
-  \frac{1}{K}\sum_{k=1}^{K}
-  \frac{1}{| \data_n^{k} |}\sum_{O_i \in \data_n^{k}}
-  \bar B_\tau
-  {
-    \left(
-      \phi_{a_1,a_2,b}{ (\data_n^{-k})}
-      , O_i
-    \right)
-  },
-\end{equation*}
-and the discrete state learner is given by
-\begin{align*}\label{eq:discrete-state-learner}
-  \hat{\phi}_n
-  &=  \argmin_{(a_1,a_2,b)\in \mathcal{A}_1\times\mathcal{A}_2\times\mathcal{B}}
-    \hat{R}_{n}(\phi_{a_1,a_2,b} ; \bar{B}_{\tau}).
-\end{align*}
-
-\section{Theoretical results for the state learner}
-\label{sec:theor-results-prop}
-
-In this section we establish theoretical guarantees for the state learner. We
-show that the state learner is consistent if its library contains a consistent
-learner, and we establish a finite sample inequality for the excess risk of the
-state learner compared to the oracle. Finally we show that a certain second
-order structure is preserved when the state learner is used for targeted
-learning.
-
-\subsection{Consistency}
-\label{sec:consistency}
-
-% We let \( \E_P \) denote
-% expectation under \( P \) and use \( \Theta \) to denote the
-% collection of all conditional state-occupation probability functions.
-Proposition~\ref{prop:stric-prop} can be derived from the fact that the
-integrated Brier score (also called the continuous ranked probability score) is
-a strictly proper scoring rule \citep{gneiting2007strictly}. This implies that
-if we minimise the average loss of the integrated Brier score, we recover the
-parameters of the data-generating distribution. Specifically, this implies that
-the oracle of a state learner is consistent if the library of learners contains
-at least one learner that is consistent for estimation of \( F \). Recall that
-the function \(F\) implicitly depends on the data-generating probability measure
-\(P\in\mathcal P\) but that this was suppressed in the notation. We now make
-this dependence explicit by writing \(F_0\) for the function which is obtained
-by substituting a specific \(P_0\in\mathcal{P}\) for \(P\) in equation
-\eqref{eq:transition}. In the following we let
-\( \mathcal{H}_{\mathcal{P}} = \{F_P : P \in \mathcal{P}\} \) where \( F_P \) is defined as in
-equation~(\ref{eq:F-def}) using the measure \( P \in \mathcal{P} \).
-
-\begin{proposition}
-  \label{prop:stric-prop}
-  If \(P_0\in\mathcal{P}\) then
-  \begin{equation*}
-    F_0 = \argmin_{F \in \mathcal{H}_{\mathcal{P}}} P_0{[\bar{B}_\tau(F, \blank)]},
-  % F^* = \argmin_{(\Lambda_1,\Lambda_2,\Gamma) \in \mathcal{M}_{\tau}\otimes\mathcal{M}_{\tau}\otimes\mathcal{M}_{\tau}} P_0{[\bar B_\tau(\phi_{(\Lambda_1,\Lambda_2,\Gamma)}, \blank)]},
-  \end{equation*}
-  \( H \)-almost surely for any \( j\in \{-1,0,1,2\} \) and almost any
-  \( t \in [0, \tau]\).
-\end{proposition}
-\begin{proof}
-  See Appendix~\ref{sec:proof-proposition}.
-\end{proof}
-
-\subsection{Oracle inequalities}
-\label{sec:finite-sample-oracle}
-
-We establish a finite sample oracle result for the state learner. Our
-Corollary~\ref{cor:oracle-prop} is in essence a special case of
-Theorem 2.3 in \citep{vaart2006oracle}.  We assume that we split the
-data into equally sized folds, and for simplicity of presentation we
-take \( n \) to be such that \( |\data_n^{-k}| = n/K \) with \( K \)
-fixed. We will allow the number of learners to grow with \( n \) and
-write
-\( \mathcal{F}_n=\mathcal{F}(\mathcal{A}_{1,n}, \mathcal{A}_{2,n},
-\mathcal{B}_n)\) as short-hand notation and to emphasise the
-dependence on \( n \).
-% The discrete super $\hat{\phi}_n$ and the
-% oracle $\tilde{\phi}_n$ are defined as in Section~\ref{sec:framework}
-% but now with respect to the library \( \mathcal{F}_n \) and the loss
-% \( B \).
-In the following we let the space \( \mathcal{H}_{\mathcal{P}} \) be equipped with the norm
-\( \| \blank \|_{P_0} \) defined as
+Finally, recall that we let $\mathcal{H}_{\mathcal{P}}$ denote the function
+space consisting of all conditional state occupation probability functions for
+some measure \( P \in \mathcal{P} \), and that we equipped this space with the
+norm
 \begin{equation}
@@
     \sum_{j=-1}^{2}\int_{\mathcal{X}} \int_0^{\tau} F(t, j, x)^2 \diff t H_0( \diff x)
-  \right\}^{1/2}.
+  \right\}^{1/2},
 \end{equation}
+for some fixed measure \( P_0 \in \mathcal{P} \). We use \( F_0 \) to denote the
+conditional state occupation probability function associated with the measure
+\( P_0 \).
 
-\begin{corollary}
-  \label{cor:oracle-prop}
-  For all \(P_0\in\mathcal{P}\), \( n \in \N \), \( k \in \{1, \dots, K\} \),
-  and $\delta>0$,
-  \begin{align*}
-    \E_{P_0}{\left[ \Vert \hat{\phi}_n(\data_n^{-k}) - F_0 \Vert_{P_0}^2 \right]}
-    & \leq (1 + 2\delta)
-      \E_{P_0}{\left[ \Vert \tilde{\phi}_n(\data_n^{-k}) - F_0 \Vert_{P_0}^2 \right]}
-    \\
-    & \quad
-      + (1+ \delta) 16   K \tau
-      \left(
-      13 + \frac{12}{\delta}
-      \right)
-      \frac{\log(1 + |\mathcal{F}_n|)}{n}.
-  \end{align*}
-\end{corollary}
-\begin{proof}
-  See Appendix~\ref{sec:proof-proposition}.
-\end{proof}
-
-Corollary~\ref{cor:oracle-prop} has the following asymptotic consequences.
-
-\begin{corollary}
-  \label{cor:asymp-cons}
-  Assume that \( |\mathcal{F}_n| = \bigO(n^q)\), for some \( q \in \N \) and
-  that there exists a sequence \( \phi_n \in \mathcal{F}_n \), \( n \in \N \),
-  such that
-  \( \E_{P_0}{\left[ \Vert \phi_n(\data_n^{-k}) - F_{0} \Vert_{P_0}^2 \right]} =
-  \bigO(n^{-\alpha}) \), for some \( \alpha\leq 1 \).
-  \begin{enumerate}[label=(\alph*)]
-  \item If $\alpha=1$ then
-    \( \E_{P_0}{\left[ \Vert \hat{\phi}_n(\data_n^{-k}) - F_0 \Vert_{P_0}^2
-      \right]} = \bigO(\log(n)n^{-1}) \).
-  \item If $\alpha<1$ then
-    \( \E_{P_0}{\left[ \Vert \hat{\phi}_n(\data_n^{-k}) - F_0 \Vert_{P_0}^2 \right]} =
-    \bigO(n^{-\alpha}) \).
-  \end{enumerate}
-\end{corollary}
-\begin{proof}
-  See Appendix~\ref{sec:proof-proposition}.
-\end{proof}
-
-\subsection{Transience of the second order remainder structure}
-
-In this section we demonstrate a theoretical property of the state
-learner which is useful for targeted learning (c.f., Section
-\ref{sec:targeted-learning}). Specifically we consider an estimator of
-a target parameter which is obtained by substituting the state learner
-estimates of the nuisance parameters $\Lambda_1$, $\Lambda_2$, and
-$\Gamma$. An example is an estimator of the cumulative incidence
-curve, which can be obtained from estimators of $\Lambda_1$ and
-$\Lambda_2$. Another example is provided in Section \ref{sec:targeted-learning}.
-By equations~(\ref{eq:lambdaj}) and~(\ref{eq:gamma}) and the definition of
-\( F \), we have
-\begin{equation}
-  \label{eq:7}
-  \Gamma(t , x) 
-  = \int_0^t  \frac{F(\diff s, -1, x )}{F(s-, 0, x )},
-  \quad \text{and} \quad
-  \Lambda_j(t , x) 
-  = \int_0^t  \frac{F(\diff s, j, x )}{F(s-, 0, x )},
-  \quad j \in \{1,2\},
-\end{equation}
-and thus an estimator based on \( \hat{\Lambda}_{1,n} \),
-\( \hat{\Lambda}_{2,n} \), and \( \hat{\Gamma}_{n} \) can also be obtained from
-an estimator of $F$ using equation~(\ref{eq:7}). A so-called targeted estimator
-has the key feature that it is asymptotically equivalent to a sum of i.i.d.\
-random variables plus a second order remainder term
-\citep{van2011targeted,hines2022demystifying}. For the setting with competing
-risks, the remainder term is dominated by terms of the form
-\begin{equation}
-  \label{eq:dr-term}
-  P{\left[
-      \int_0^{\tau} w_n(s, \blank)
-      \hat{M}_{1,n}(s \mid  \blank)
-      \hat{M}_{2,n}(\diff s \mid  \blank)
-    \right]},
-\end{equation}
-where \( (\hat{M}_{1,n}, \hat{M}_{2,n}) \) is any of the nine combinations of
-\( \hat{M}_{1,n} \in \{[\Gamma -\hat{\Gamma}_n], [\Lambda_1
--\hat{\Lambda}_{1,n}], [\Lambda_2 -\hat{\Lambda}_{2,n}]\} \) and
-\( \hat{M}_{2,n} \in \{[\Gamma -\hat{\Gamma}_n], [\Lambda_1
--\hat{\Lambda}_{1,n}], [\Lambda_2 -\hat{\Lambda}_{2,n}]\} \), and \( w_n \) is
-some data-dependent function with domain \([0,\tau]\times\mathcal X \)
-\citep{van2003unified}. In particular, a targeted estimator will be
-asymptotically linear if the `products' of the estimation errors
-\( \hat{M}_{1,n} \) and \( \hat{M}_{2,n} \) in equation~(\ref{eq:dr-term}) are
-\( \smallO_P{(n^{-1/2})}\). Proposition~\ref{prop:dr-structure} states that if
-equation~(\ref{eq:dr-term}) holds for a targeted estimator based on estimators
-$\hat{\Lambda}_{1,n}$, $\hat{\Lambda}_{2,n}$, and $\hat{\Gamma}_{n}$, then a
-similar product structure holds for a targeted estimator based on
-\( \hat{F}_n \). We state the result for the special case that
-\(\hat{M}_{1,n}= \Gamma-\hat{\Gamma}_n \) and
-\(\hat{M}_{2,n} =\Lambda_1-\hat{\Lambda}_{1,n} \), but similar results hold for
-any combinations of \( \Gamma-\hat{\Gamma}_n\),
-\( \Lambda_1-\hat{\Lambda}_{1,n} \), and \( \Lambda_2-\hat{\Lambda}_{2,n} \).
-\begin{proposition}
-  \label{prop:dr-structure}
-  Assume that \( w(s,x)\leq c \), \( F(s, 0, x) \geq 1/c \) and
-  \( \hat{F}_n(s, 0, x) \geq 1/c \) for some \( c>0 \) for all
-  \( s \in [0, \tau] \) and \( x \in \mathcal{X} \). Then there are real-valued
-  uniformly bounded functions \( w^a_n \), \( w^b_n \), \( w^c_n \), and
-  \( w^d_n \) with domain \( [0,\tau]^2 \times \mathcal{X} \) such that
-  \begin{align*}
-    & P_0{\left[
-      \int_0^{\tau} w(s, \blank)
-      \left\{
-      \Gamma_0(s,\blank) -\hat{\Gamma}_n(s,\blank)
-      \right\}
-      [\Lambda_0-\hat{\Lambda}_n]
-      (\diff s, \blank)
-      \right]}
-    \\
-    & =
-      P_0{\left[
-      \int_0^{\tau} \int_0^{s} w^a_n(s,u,\blank) [F_0 - \hat{F}_n](u-, 0, \blank)[F_0 - \hat{F}_n](s-, 0, \blank) F_0(\diff u, 2, \blank ) F_0 ( \diff s, 1, \blank)
-      \right]}
-    \\
-    & \quad +
-      P_0{\left[
-      \int_0^{\tau} \int_0^{s} w^b_n(s,u,\blank) [F_0 - \hat{F}_n](u-, 0, \blank)
-      F_0(\diff u, 2, \blank ) [F_0 - \hat{F}_n](\diff s, 1, \blank)
-      \right]}
-    \\
-    & \quad +
-      P_0{\left[
-      \int_0^{\tau} \int_0^{s} w^c_n(s,u,\blank) [F_0 - \hat{F}_n](\diff u, 2, \blank)
-      [F_0 - \hat{F}_n](s-, 0, \blank)
-      F_0(\diff s, 1, \blank ) 
-      \right]}
-    \\
-    & \quad +
-      P_0{\left[
-      \int_0^{\tau} \int_0^{s} w^d_n(s,u,\blank) [F_0 - \hat{F}_n](\diff u, 2, \blank)
-      [F_0 - \hat{F}_n](\diff s, 1, \blank)
-      \right]}.
-  \end{align*}
-\end{proposition}
-\begin{proof}
-  See Appendix~\ref{sec:state-learner-with}.
-\end{proof}
-
-\section{Targeted learning}
-\label{sec:targeted-learning}
-
-In this section, we consider a suitably smooth operator
-\( \theta \colon \mathcal{Q} \rightarrow \Theta \) which represents a
-target parameter of interest. The parameter space $\Theta$ can be a
-subset of \(\R^d\) or a subset of a function space, for example a
-subset of \(\mathcal{M}_{\tau}\) as in Section
-\ref{sec:super-learning}. In subsection~\ref{sec:cause-spec-aver} we
-discuss an example from causal inference where $\theta$ is the average
-treatment effect and \( \Theta = [-1,1] \).  To discuss the role of
-the state learner for targeted learning we briefly review some results
-from semiparametric efficiency theory. Extensive reviews and
-introductions are available elsewhere
-\cite[e.g.,][]{pfanzagl1985contributions,bickel1993efficient,van2003unified,tsiatis2007semiparametric,kennedy2016semiparametric}.
-Under the assumption of conditional independent censoring and
-positivity, $\theta$ is identifiable from \( \mathcal{P} \) which
-means that there exists an operator
-\( \Psi \colon \mathcal{P} \rightarrow \Theta \) such that
-\( \theta(Q) = \Psi(P_{Q, \Gamma}) \) for all
-$\Gamma \in \mathcal{M}_{\tau}$.  By equation~(\ref{eq:parametrizeP})
-this implies that we may write
-\begin{equation*}
-  \theta(Q) = \Psi(P) = \tilde{\Psi}^0(\Lambda_1, \Lambda_2, H),
-\end{equation*}
-for some operator \( \tilde{\Psi}^0 \). The state learner provides a ranking of
-all tuples
-\( (a_1, a_2, b) \in \mathcal{A}_1 \times \mathcal{A}_2 \times \mathcal{B} \).
-We use \( \hat{a}_{1,n} \), \( \hat{a}_{2,n} \), and \( \hat{b}_n \) to denote
-the learners corresponding to the discrete state learner \( \hat{\phi}_n \),
-i.e., the tuple with the highest rank. Letting \( H_n(\data_n) \) denote the
-empirical measure of \( \{X_1, \dots, X_n\} \), we obtain a plug-in estimator of
-$\theta$:
-\begin{equation}
-  \label{eq:2}
-  \hat{\Psi}^0(\data_n) =
-  \tilde{\Psi}^0(\hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n), H_n(\data_n)). 
-\end{equation}
-
-The asymptotic distribution of \( \hat{\Psi}^0 \) is
-difficult to analyse due to the cross-validated model selection step
-involved in the estimation of the nuisance parameters $\Lambda_1$ and
-$\Lambda_2$.
-% In addition, the estimator will typically have an
-% asymptotic bias that vanishes at a too slow 
-% rate.
-Using tools from semi-parametric efficiency theory, it is possible to
-construct a so-called targeted or debiased estimator with an
-asymptotic distribution which we know how to estimate
-\citep{bickel1993efficient,van2011targeted,chernozhukov2018double}. A
-targeted estimator is based on the efficient influence function for
-the parameter $\tilde{\Psi}^0$ and relies on estimators of the
-nuisance parameters $\Gamma$, \( \Lambda_1 \) and $\Lambda_2$. The
-efficient influence function is a \( P \)-zero mean and square
-integrable function which we denote by
-\( \psi(\blank ; \Lambda_1, \Lambda_2, \Gamma) \). The name is
-justified because any regular asymptotically linear estimator that has
-\( \psi \) as its influence function is asymptotically efficient,
-meaning that it has smallest asymptotic variance among all regular
-asymptotically linear estimators \citep{bickel1993efficient}.
-
-An example of a targeted estimator is the one-step estimator, defined as
-\begin{equation}
-    \label{eq:one-step-def}
-    \hat{\Psi}_{\text{OS}}(\data_n)
-    =
-    \tilde{\Psi}^0(\hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
-    H_n(\data_n))
-    + \empmeas{[\psi(\blank; \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
-      \hat{b}_n(\data_n) )]},
-\end{equation}
-where \( \empmeas \) is the empirical measure of a data set
-\(\{O_i\}_{i=1}^n\).
-Under suitable regularity conditions
-we have the
-following asymptotic expansion of the one-step estimator
-\citep{pfanzagl1985contributions,van2003unified,fisher2021visually,kennedy2022semiparametric},
-\begin{equation*}
-  \hat{\Psi}_{\text{OS}}(\data_n)- \Psi(P)
-  =  \empmeas{[\psi(\blank ; \Lambda_1, \Lambda_2, \Gamma)]}
-  +\mathrm{Rem}{(\hat{\Lambda}_{1,n},\hat{\Lambda}_{2,n},  \hat{\Gamma}_n, P)} + \smallO_{P}(n^{-1/2}),
-\end{equation*}
-where the remainder term has the form
-\begin{equation}
-  \label{eq:4}
-  \mathrm{Rem}{(\hat{\Lambda}_{1,n},\hat{\Lambda}_{2,n},  \hat{\Gamma}_n, P)}
-  = \bigO_P{
-    \left\{
-      \|\Lambda_1-\hat{\Lambda}_{1,n}\|^2
-      +
-      \|\Lambda_2-\hat{\Lambda}_{2,n}\|^2
-      +
-      \|\Gamma-\hat{\Gamma}_{n}\|^2
-    \right\}
-  },
-\end{equation}
-for some suitable norm \( \|\blank \| \), for instance the
-\( \mathcal{L}_{P}^2 \)-norm. When equation~(\ref{eq:4}) holds and the
-nuisance parameters $\Lambda_1$, $\Lambda_2$, and $\Gamma$ are
-consistently estimated at rate \( \smallO_P{(n^{-1/4})} \), then
-\begin{equation}
-  \label{eq:3}
-  \sqrt{n}(\hat{\Psi}_{\text{OS}}(\data_n)- \Psi(P)) \weakly \mathcal{N}(0,
-  P{[\psi(\blank; \Lambda_1, \Lambda_2, \Gamma)^2]}),
-\end{equation}
-where we use \( \weakly \) to denote weak convergence \citep{van2000asymptotic}.
-In particular, equation~(\ref{eq:3}) and Slutsky's lemma imply that we can
-obtain asymptotically valid \((1-\alpha)\cdot100\%\) confidence intervals by
-calculating
-\begin{equation*}
-  \left[
-    \hat{\Psi}_{\text{OS}}(\data_n) - q_{\alpha/2} \hat{\sigma}(\data_n) ,
-    \;
-    \hat{\Psi}_{\text{OS}}(\data_n) + q_{\alpha/2} \hat{\sigma}(\data_n)
-  \right],
-\end{equation*}
-where \( q_{\alpha} \) is the \( (1-\alpha) \)-quantile of the standard normal
-distribution and
-\begin{equation*}
-  \hat{\sigma}(\data_n)^2 = \empmeas{ \left[ \psi(\blank;
-      \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n), \hat{b}_n(\data_n))^2
-    \right]}.
-\end{equation*}
-
-\subsection{Average treatment effect on the absolute risk of an event}
-\label{sec:cause-spec-aver}
-
-In the following we detail how the state learner can be used to
-construct a targeted estimator of the cause-specific average treatment
-effect \citep{rytgaard2022targeted}. We assume that the covariate
-vector \( X \in \R^p \) contains a binary treatment indicator \( A \)
-and a vector of potential confounders,
-\( W \in \mathcal{W}\subset \R^{p-1} \).  We use $\mu$ to denote the
-marginal distribution of \( W \) and $\pi$ to denote the conditional
-probability of treatment,
-\begin{equation*}
-  \pi(w) = P(A=1 \mid W=w).
-\end{equation*}
-We assume throughout that $\pi$ is uniformly bounded away from \( 0 \)
-and \( 1 \) on \( \mathcal{W} \) and that both \( A \) and \( W \) are
-fully observed for all individuals. We use a super learner to
-estimate $\pi$ \citep{Polley_Ledell_Kennedy_Laan_2023_Superlearn}, and
-we denote this estimator by $\hat{\pi}_n$. We use the empirical
-measure of \( \{W_1, \dots, W_n\} \) to estimate $\mu$, and denote
-this estimator by $\hat{\mu}_n$. As parameter of interest we consider
-the standardised difference in the absolute risk of an event with
-cause 1 at time $\tau$:
-\begin{equation*}
-  \theta_{{\tau}}(Q) = \int_{\mathcal{W}} 
-  \left\{
-    \int_0^{\tau}
-    S(s- \mid w, 1)  \Lambda_1(\diff s \mid w, 1)
-    -
-    \int_0^{\tau}
-    S(s- \mid w, 0)  \Lambda_1(\diff s \mid w, 0)
-  \right\}
-  \mu(\diff w).
-\end{equation*}
-Under the usual assumptions for causal inference (consistency,
-positivity, no unmeasured confounding) \( \theta_{{\tau}} \) can be
-given the causal interpretation
-\begin{equation*}
-  \theta_{{\tau}}(Q) =
-  P{(T^{1} \leq \tau, D^{1}=1)}-
-  P{(T^{0} \leq \tau, D^{0}=1)},
-\end{equation*}
-where \( (T^a, D^a) \), \( a \in \{0,1\} \), denote potential outcomes
-\citep{hernanRobinsWhatIf}. In this case, the interpretation of $\theta_{\tau}$
-is the difference in the average risk of cause \( 1 \) occurring before time
-\( \tau \) in the population if everyone had been given treatment (\( A=1 \))
-compared to if no one had been given treatment \( (A=0) \).
-
-Using equation~(\ref{eq:surv-def}) we may write
-\( \theta_{{\tau}}(Q) = \tilde{\Psi}_{\tau}^0(\Lambda_1, \Lambda_2, \mu) \), where
-\begin{equation}
-  \label{eq:1}    
-  \begin{split}
-    \tilde{\Psi}_{\tau}^0(\Lambda_1, \Lambda_2, \mu)
-    & =
-      \int_{\mathcal{W}} 
-      \int_0^{\tau}
-      \exp{\{{-\Lambda_1(s- \mid w, 1)-\Lambda_2(s- \mid w, 1)}\}}  \Lambda_1(\diff s \mid
-      w, 1)
-      \mu(\diff w)
-    \\
-    &  \quad
-      -\int_{\mathcal{W}} 
-      \int_0^{\tau}
-      \exp{\{{-\Lambda_1(s- \mid w, 0)-\Lambda_2(s- \mid w, 0)}\}}  \Lambda_1(\diff s \mid w, 0)
-      \mu(\diff w).
-  \end{split}
-\end{equation}
-The efficient influence function for the parameter $\tilde{\Psi}_{\tau}$ depends
-on the set \( (\Lambda_1, \Lambda_2, \Gamma, \pi) \) of nuisance parameters.
-We define
-\begin{align*}
-  \omega_a(A,W; \pi)
-  &=  \frac{% (-1)^{a+1}
-    \1{\{A=a\}}}{\pi(W)^{a}(1-\pi(W))^{1-a}},
-  \\
-  g(t, A, W; \Lambda_1, \Lambda_2)
-  & = \int_0^{t}
-    \exp{\{ {-\Lambda_1(s- \mid W, A)-\Lambda_2(s- \mid W, A)} \}}  \Lambda_1(\diff s \mid
-    W, A),
-  \\  
-  M_j(\diff t \mid A, W;  \Lambda_j  )
-  & = N_j(\diff t) -
-    \1{\{\tilde{T} \geq t\}} \Lambda_j(\diff t \mid W, A),
-    \quad j \in \{1,2\},
-    \intertext{and}
-    M(\diff t \mid A, W;  \Lambda_1, \Lambda_2  )
-  & = M_1(\diff t \mid A, W;  \Lambda_1  ) +
-    M_2(\diff t \mid A, W;  \Lambda_2  ).
-\end{align*}
-The efficient influence function can now be written as
-\citep{van2003unified,jewell2007non,rytgaard2022targeted},
-\begin{equation*}
-  \psi_{\tau}(O; \Lambda_1, \Lambda_2, \Gamma, \pi)
-  = \psi_{\tau}^1(O; \Lambda_1, \Lambda_2, \Gamma, \pi)
-  - \psi_{\tau}^0(O; \Lambda_1, \Lambda_2, \Gamma, \pi)
-  -\tilde{\Psi}_{\tau}^0(\Lambda_1, \Lambda_2, \mu),
-\end{equation*}
-where
-\begin{equation}
-  \label{eq:5}
-  \begin{split}
-    & \psi_{\tau}^a(O; \Lambda_1, \Lambda_2, \Gamma, \pi)
-    \\
-    & =
-      \omega_a(A,W; \pi)
-      \int_0^{\tau} \exp{\{{\Gamma(t- \mid A, W)}\}}   
-      M_1(\diff t \mid A, W; \Lambda_1)
-    \\
-    & \quad
-      -
-      \omega_a(A,W; \pi)
-      g(\tau, A, W; \Lambda_1, \Lambda_2)
-    \\
-    & \quad \quad \times
-      \int_0^{\tau}
-      \exp{\{{[\Gamma+\Lambda_1 + \Lambda_2](t- \mid A, W)}\}}
-      M(\diff t \mid A, W; \Lambda_1, \Lambda_2)
-    \\
-    & \quad
-      +
-      \omega_a(A,W; \pi)      
-      \int_0^{\tau}
-      g(t, A, W; \Lambda_1, \Lambda_2)
-    \\
-    & \quad \quad \times
-      \exp{\{{[\Gamma+\Lambda_1 + \Lambda_2](t- \mid A, W)}\}}
-      M(\diff t \mid A, W; \Lambda_1, \Lambda_2)
-    \\
-    & \quad + g(\tau, a, W; \Lambda_1, \Lambda_2).
-  \end{split}
-\end{equation}
-Equations~(\ref{eq:1}) and~(\ref{eq:5}) allow us to construct a one-step
-estimator by using the definition given in equation~(\ref{eq:one-step-def}),
-which gives the estimator
-\begin{equation}
-  \label{eq:one-step-comp-ate}
-  \begin{split}
-    \hat{\Psi}_{\tau,\text{OS}}(\data_n)
-    = &
-        \tilde{\Psi}_{\tau}^0(\hat{a}_1(\data_n), \hat{a}_2(\data_n),
-        \hat{\mu}_n(\data_n))
-    \\
-      &
-        +
-        \empmeas{[\psi_{\tau}(\blank; \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
-        \hat{b}_n(\data_n), \hat{\pi}_n(\data_n))]}
-    \\
-    = &
-        \empmeas{[\psi_{\tau}^1(\blank; \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
-        \hat{b}_n(\data_n), \hat{\pi}_n(\data_n))]}
-    \\
-      &
-        - \empmeas{[\psi_{\tau}^0(\blank; \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
-        \hat{b}_n(\data_n), \hat{\pi}_n(\data_n))]}.
-  \end{split}
-\end{equation}
-
-\section{Numerical experiments}
-\label{sec:numer-exper}
-
-In this section we report results from a simulation study where we consider
-estimation of the conditional survival function. In the first part, we compare
-the state learner to two IPCW based discrete super learners that use either the
-Kaplan-Meier estimator or a Cox model to estimate the censoring probability
-\citep{gonzalez2021stacked}. In the second part we compare the state learner to
-the super learner proposed by \cite{westling2021inference}.
-
-In both parts we use the same data-generating mechanism. We generate data
-according to a distribution motivated from a real data set in which censoring
-depends on the baseline covariates. We simulate data based on the prostate
-cancer study of \cite{kattan2000pretreatment}. The outcome of interest is the
-time to tumor recurrence, and five baseline covariates are used to predict
-outcome: prostate-specific antigen (PSA, ng/mL), Gleason score sum (GSS, values
-between 6 and 10), radiation dose (RD), hormone therapy (HT, yes/no) and
-clinical stage (CS, six values). The study was designed such that a patient's
-radiation dose depended on when the patient entered the study
-\citep{gerds2013estimating}. This in turn implies that the time of censoring
-depends on the radiation dose. The data were re-analysed in
-\citep{gerds2013estimating} where a sensitivity analysis was conducted based on
-simulated data. Here we use the same simulation setup, where event and censoring
-times are generated according to parametric Cox-Weibull models estimated from
-the original data, and the covariates are generated according to either marginal
-Gaussian normal or binomial distributions estimated from the original data
-\citep[c.f.,][Section~4.6]{gerds2013estimating}. We refer to this simulation
-setting as `dependent censoring'. We also considered a simulation setting where
-data were generated in the same way, except that censoring was generated
-completely independently. We refer to this simulation setting as `independent
-censoring'.
-
-% ANDERS: hedder det ikke Nelson-Aalen i stedet for Kaplan-Meier?
-%         kan du skrive klart hvad outcome er i denne foerste del? er det combined endpoint som man kalder for tumor-free survival?  
-For all super learners we use a library consisting of three learners:
-The Kaplan-Meier estimator
-\citep{kaplan1958nonparametric,Gerds_2019prodlim}, a Cox model with
-main effects \citep{cox1972regression, survival-package}, and a random
-survival forest \citep{ishwaran2008random,rfsrc-paclage}. We use the
-same library to learn the outcome distribution and the censoring
-distribution. Note that the three learners in our library of learners
-can be used to learn the cumulative hazard functions of the 
-outcome and the censoring distribution. The latter works by
-training the learner on the data set \( \data_n^c \), where
-\( \data_n^c = \{O_i^c\}_{i=1}^n \) with
-\( O_i^c = (\tilde{T}_i, 1-\Delta, X_i) \). When we say that we use a
-learner for the cumulative hazard function of the outcome to learn the
-cumulative hazard function of the censoring time, we mean that the
-learner is trained on \( \data_n^c \).
-
-We compare the state learner to two IPCW based super learners: The
-first super learner, called IPCW(Cox), uses a Cox model with main
-effects to estimate the censoring probabilities, while the second
-super learner, called IPCW(KM), uses the Kaplan-Meier estimator to
-estimate the censoring probabilities. The Cox model for the censoring
-distribution is correctly specified in both simulation settings while
-the Kaplan Meier estimator only estimates the censoring model
-correctly in the simulation setting where censoring is
-independent. Both IPCW super learners are fitted using the
-\texttt{R}-package \texttt{riskRegression}
-\citep{Gerds_Ohlendorff_Ozenne_2023}.
-%
-% | time | sim_setting | true_events | true_cens | at_risk |
-% |------+-------------+-------------+-----------+---------|
-% |   36 | original    |      24.619 |    61.853 |  25.774 |
-% |   36 | indep_cens  |      24.674 |    38.740 |  46.141 |
-%
-The IPCW super learners use the integrated Brier score up to a fixed time
-horizon (36 months). The marginal risk of the event before this time horizon is
-\(\approx 24.6\)\%. Under the `dependent censoring' setting the marginal
-censoring probability before the time horizon is \(\approx 61.9\)\%. Under the
-`independent censoring' setting the marginal censoring probability before this
-time horizon is \( \approx 38.7 \)\%.
-
-Each super learner provides a learner for the cumulative hazard
-function for the outcome of interest. From the cumulative hazard
-function a risk prediction model can be obtained (c.f.,
-equation~(\ref{eq:surv-def}) with $\Lambda_2 = 0$). We measure the
-performance of each super learner by calculating the index of
-prediction accuracy (IPA) \citep{kattan2018index} at a fixed time
-horizon (36 months) for the risk prediction model provided by the
-super learner. The IPA is 1 minus the ratio between the model's Brier
-score and the null model's Brier score, where the null model is the
-model that does not use any covariate information. The IPA is
-approximated using a large (\( n = 20,000 \)) independent data set of
-uncensored data. As a benchmark we calculate the performance of the
-risk prediction model chosen by the oracle selector, which uses the
-large data set of uncensored event times to select the learner with
-the highest IPA.
-
-The results are shown in Figure~\ref{fig:ipcw-fail}. We see that in
-the scenario where censoring depends on the covariates, using the
-Kaplan-Meier estimator to estimate the censoring probabilities
-provides a risk prediction model with an IPA that is lower than the
-risk prediction model provided by the state learner. The performance
-of the risk prediction model selected by the state learner is similar
-to the risk prediction model selected by the IPCW(Cox) super learner
-which a priori uses a correctly specified model for the censoring
-distribution. Both these risk prediction models are close to the
-performance of the oracle, except for small sample sizes.
-
-\begin{figure}
-  \centering %
-  \includeFigCond[width=1\linewidth]{experiment-fig-sl-ipcw.pdf}
-  \caption[]{For the risk prediction models provided by each of the super
-    learners, the IPA is plotted against sample size. The results are averages across 
-    1000 simulated data sets and the error bars are used to quantify the Monte Carlo
-    uncertainty.
-  }
-  \label{fig:ipcw-fail}
-\end{figure}
-
-We next compare the state learner to the super learner \texttt{survSL}
-\citep{westling2021inference}. This is another super learner which
-like the state learner works without a pre-specified censoring
-model. Note that both the state learner and \texttt{survSL} provide a
-prediction model for the event time outcome and also for the
-probability of being censored. Hence, we compare the performance of
-these methods with respect to both the outcome and the censoring
-distribution. Again we use the IPA to quantify the predictive
-performance.
-
-The results are shown in Figures~\ref{fig:zelefski-out}
-and~\ref{fig:zelefski-cens}. We see that for most sample sizes, the state
-learner selected prediction models for both censoring and outcome which have
-similar or higher IPA compared to the prediction models selected by
-\texttt{survSL}.
-\begin{figure}
-  \centering %
-  \includeFigCond[width=1\linewidth]{experiment-fig-sl-survSL-out.pdf}
-  \caption[]{For the risk prediction models of the outcome provided by each
-    of the super learners, the IPA at the fixed time horizon is plotted against
-    sample size. The results are averages across 1000 repetitions and the error
-    bars are used to quantify the Monte Carlo uncertainty.}
-  \label{fig:zelefski-out}
-\end{figure}
-
-\begin{figure}
-  \centering %
-  \includeFigCond[width=.7\linewidth]{experiment-fig-sl-survSL-cens.pdf}
-  \caption[]{For the risk prediction models of the censoring model provided by
-    each of the super learners, the IPA at the fixed time horizon is plotted
-    against sample size. The results are averages across 1000 repetitions and
-    the error bars are used to quantify the Monte Carlo uncertainty.}
-  \label{fig:zelefski-cens}
-\end{figure}
-
-\section{Prostate cancer study}
-\label{sec:real-data-appl}
-
-In this section we use the prostate cancer data of
-\cite{kattan2000pretreatment} to illustrate the use of the state
-learner in the presence of competing risks. We have introduced the
-data in Section~\ref{sec:numer-exper}. The data consists of 1,042
-patients who are followed from start of followup until tumor
-recurrence, death without tumor recurrence or end of followup
-(censored) whatever came first. For the sole purpose of illustration,
-we estimate the average treatment effect of hormone therapy on death
-and tumor recurrence. To do this we adapt the estimation strategy of
-Section~\ref{sec:targeted-learning} as follows. We use the state
-learner to rank libraries of learners for the cause-specific
-cumulative hazard functions of tumor recurrence, death without tumor
-recurrence, and censoring.  The libraries of learners each include
-five learners: the Nelson-Aalen estimator, three Cox regression models
-(unpenalized, Lasso, Elastic net) each including additive effects of
-the 5 covariates (Section~\ref{sec:numer-exper}), and a random
-survival forest. We use the same set of learners to learn the
-cumulative hazard function of tumor recurrence \( \Lambda_1 \), the
-cumulative hazard function of death without tumor recurrence
-\( \Lambda_2 \), and the cumulative hazard function of the conditional
-censoring distribution $\Gamma$.  We then use the highest ranked
-combination of learners and apply
-formula~\eqref{eq:one-step-comp-ate}.
-
-This gives a library consisting of \( 5^3 = 125 \) learners for the
-conditional state occupation probability function \( F \) defined in
-equation~(\ref{eq:F-def}). We use five folds for training and testing
-the models, and we repeat training and evaluation five times with
-different splits.  The integrated Brier score (defined in
-Section~\ref{sec:super-learner-simple}) for all learners are shown in
-Figure~\ref{fig:zelefski-real}. We see that the prediction
-performance is mostly affected by the choice of learner for the
-censoring distribution. Several combinations of learners give similar
-performance as measured by the integrated Brier score, as long as a
-random forest is used to model the censoring distribution.
-
-\begin{figure}
-  \centering %
-  \includeFigCond[width=1\linewidth]{real-data-state-learner.pdf}
-  \caption[]{The results of applying the 125 combinations of learners to the
-    prostate cancer data set. The error bars are based on five repetitions using
-    different splits. We refer to learners of \( \Lambda_1 \), \( \Lambda_2 \),
-    and $\Gamma$ as `Tumor learner', `Mortality learner', and `Censoring
-    learner', respectively.}
-  \label{fig:zelefski-real}
-\end{figure}
-
-We use the learners of the three cumulative hazard functions selected
-by the state learner and the estimator defined in
-Section~\ref{sec:cause-spec-aver},
-equation~(\ref{eq:one-step-comp-ate}), to estimate the average
-treatment effect of hormone therapy on risk of tumor recurrence and
-death. The propensity score is estimated with a lasso model that
-includes all levels of interaction.
-% ANDERS: er du sikker paa at det er ALL levels of interaction, ogsaa higher order?
-%         der mangler en reference for lasso model
-The results are shown in Figure~\ref{fig:zelefski-real-target} for 6
-month intervals after baseline with pointwise 95\% confidence
-intervals. We see that hormone therapy decreases the risk of tumor
-recurrence and increases the risk of death without tumor recurrence,
-but that none of the estimated effects are statistically significant.
-
-\begin{figure}
-  \centering%
-  \includeFigCond[width=1\linewidth]{real-data-target.pdf}
-  \caption[]{Estimates of the average treatment effect of hormone therapy on the
-    risk of tumor recurrence and death obtained from the prostate cancer study
-    analysed by \cite{kattan2000pretreatment}. The estimates are based on the
-    estimator defined in equation~(\ref{eq:one-step-comp-ate}) with pointwise
-    95\% confidence intervals calculated as described in
-    Section~\ref{sec:targeted-learning}. The estimates of the nuisance
-    parameters are provided by the state learner.}
-  \label{fig:zelefski-real-target}
-\end{figure}
-
-\section{Discussion}
-\label{sec:discussion}
-
-The state learner is a new super learner that can be used with
-right-censored data and competing events. Compared to existing
-IPCW-based methods, the advantage of the state learner is that it does
-not depend on a pre-specified estimator of the censoring distribution,
-but selects one automatically based on a library of learners for the
-censoring distribution. Furthermore, the state learner neither
-requires that the cause-specific cumulative hazard functions
-\( \Lambda_j \) can be written as integrals with respect to Lebesgue
-measure, nor does it assume a (semi-)parametric formula. In the
-remainder of this section we discuss the limitations of our proposal
-and avenues for further research.
-
-A major advantage of the state learner is that the performance of each
-combination of learners can be estimated without additional nuisance
-parameters. A potential drawback of our approach is that we are
-evaluating the loss of the learners on the level of the observed data
-distribution while the target of the analysis is either the event time
-distribution, or the censoring distribution, or both.
-% ANDERS: er du enig i min omformulering af forgaaende saetning?
-Specifically, the finite sample oracle inequality in
-Corollary~\ref{cor:oracle-prop} concerns the function \( F \), which
-is a feature of \( P \in \mathcal{P} \), while what we are typically
-interested in is \( \Lambda_j \) or \( S \), which are features of
-\( Q \in \mathcal{Q} \). We emphasise that while the state learner
-provides us with estimates of \( \Lambda_j \) and $\Gamma$ based on
-libraries \( \mathcal{A}_j \) and \( \mathcal{B} \), the performance
-of these learners is not assessed directly for their respective target
-parameters, but only indirectly via the performance of \( F \).  For
-settings without competing risks, our numerical studies suggest that
-measuring the performance of \( F \) also leads to good performance
-for estimation of \( S \).
-% ANDERS: den forgaaende paragraf gentager det samme 2-3 gang. kan du forkorte en gang?
-
-% In the context of targeted learning, a drawback of the state learner is that the
-% doubly robustness property defined in equation~(\ref{eq:dr-property}) seem to be
-% lost because we only use a single nuisance parameter estimator. As defined here,
-% however, the state learner is build using libraries for the conditional
-% cause-specific cumulative hazard functions, so some doubly robustness might be
-% preserved.
-
-Our proposed super learner can be implemented with a broad library of learners
-and using existing software.
-% For instance the \texttt{R}-package
-% \texttt{riskRegression} \citep{Gerds_Ohlendorff_Ozenne_2023} and the \texttt{R}-package by westling
-Furthermore, while
-the library \( \mathcal{F}(\mathcal{A}_1,\mathcal{A}_2,\mathcal{B}) \) consists
-of \( |\mathcal{A}_1||\mathcal{A}_2||\mathcal{B}| \) many learners, we only need to fit
-\( |\mathcal{A}_1| +|\mathcal{A}_2| + |\mathcal{B}| \) many learners in each fold. To
-evaluate the performance of each learner we need to perform
-\( |\mathcal{A}_1||\mathcal{A}_2||\mathcal{B}| \) many operations to calculate the
-integrated Brier score in each hold-out sample, one for each combination of the
-fitted models, but these operations are often negligible compared to fitting the
-models. Hence the state learner is essentially not more computationally demanding
-than any procedure that uses super learning to learn $\Lambda_1$, $\Lambda_2$,
-and $\Gamma$ separately. While our proposal is based on constructing the library
-\( \mathcal{F} \) from libraries for learning \( \Lambda_1 \), $\Lambda_2$, and
-$\Gamma$, it could also be of interest to consider learners that estimate
-\( F \) directly.
-
-In our numerical studies, we only considered learners of $\Lambda_j$ and
-$\Gamma$ that provide cumulative hazard functions which are piece-wise constant
-in the time argument. This simplifies the calculation of \( F \) as the
-integrals in equation~(\ref{eq:transition}) reduce to sums. When $\Lambda_j$ or
-\( \Gamma \) are absolutely continuous in the time argument, calculating \( F \)
-is more involved, but we expect that a good approximation can be achieved by
-discretisation.
-
-\appendix
-
-\section{Theoretical guarantees for the state learner}
-\label{sec:proof-proposition}
-
-In this section we provide proofs of the results stated in
-Section~\ref{sec:theor-results-prop}.
+\section{Consistency}
+\label{sec:consistency}
 
@@
 
-\begin{proof}[Proof of Proposition~\ref{prop:stric-prop}]
+\begin{proof}[Proof of Proposition~1 of the main paper]
   The result follows from Lemma~\ref{lemma:norm}.
@@
 
-Recall that $\mathcal{H}_{\mathcal{P}}$ denote the function space consisting of
-all conditional state occupation probability functions for some measure
-\( P \in \mathcal{P} \).
+\section{Oracle inequalities}
+\label{sec:oracle-inequalities}
 
-\begin{proof}[Proof of Corollary~\ref{cor:oracle-prop}]
+Recall that we use \( \mathcal{F}_n \) to denote a library of learners for the
+function \( F \), and that \( \hat{\phi} \) and \( \tilde{\phi} \) denotes,
+respectively, the discrete super learner and the oracle learner for the library
+\( \mathcal{F}_n \), c.f., Section~4 of the main paper.
+
+\begin{proof}[Proof of Corollary~1 of the main paper]
   First note that minimising the loss \( \bar{B}_{\tau} \) is equivalent to
@@
 
-\begin{proof}[Proof of Corollary~\ref{cor:asymp-cons}]
+\begin{proof}[Proof of Corollary~2 of the main paper]
   By definition of the oracle and Lemma~\ref{lemma:norm},
@@
   for all \( n \in \N \). The results then follows from
-  Corollary~\ref{cor:oracle-prop}.
+  Corollary~1 of the main paper.
 \end{proof}
 
-\section{The state learner with targeted learning}
+\section{Transience of the second order remainder structure}
 \label{sec:state-learner-with}
 
-In this section show that a product structure is preserved when an estimator
-$\bar\Psi(\hat{F}_n, \hat{H}_n)$ is used instead of
-$\tilde\Psi(\hat{\Lambda}_n, \hat{\Gamma}_n, \hat{H}_n)$.
+Recall that we let $\Lambda_1$ denote the conditional cumulative hazard function
+for one of the event times of interest and $\Gamma$ the conditional cumulative
+hazard function for censoring, c.f., Section~2 of the main paper.
 
-\begin{proof}[Proof of Proposition~\ref{prop:dr-structure}]
+\begin{proof}[Proof of Proposition~2 of the main paper]
   For notational convenience we suppress \( X \) in the following. The final
   result can be obtained by adding the argument \( X \) to all functions and
-  averaging. We use the relations from equation~(\ref{eq:7}) to write
+  averaging. We use the relations from equation~(8) in the main paper to write
   \begingroup %
@@
         \right\}
-        [\Lambda - \hat{\Lambda}_n](\diff s)
+        [\Lambda_1 - \hat{\Lambda}_{1,n}](\diff s)
       \\
@@
         \left\{
-        \int_0^s \frac{F(\diff u, 2)}{F(u-, 0)} -
-        \int_0^s \frac{\hat{F}_n(\diff u, 2)}{\hat{F}_n(u-, 0)}  -
+        \int_0^s \frac{F(\diff u, -1)}{F(u-, 0)} -
+        \int_0^s \frac{\hat{F}_n(\diff u, -1)}{\hat{F}_n(u-, 0)}  -
         \right\}
@@
         \frac{1}{F(u-, 0)} -  \frac{1}{\hat{F}_n(u-, 0)}
-        \right) F(\diff u, 2)
+        \right) F(\diff u, -1)
       \\
@@
         \left[
-        F(\diff u, 2) - \hat{F}_n(\diff u, 2)
+        F(\diff u, -1) - \hat{F}_n(\diff u, -1)
         \right]
@@
         \frac{1}{\hat{F}_n(s-, 0)}
-        \right)F(\diff u, 2)F(\diff s, 1)
+        \right)F(\diff u, -1)F(\diff s, 1)
       \\
@@
         \frac{1}{F(u-, 0)} -  \frac{1}{\hat{F}_n(u-, 0)}
-        \right) \frac{F(\diff u, 2) }{\hat{F}_n(u-,0)}
+        \right) \frac{F(\diff u, -1) }{\hat{F}_n(u-,0)}
         \left(
@@
     \endgroup %
-    Consider the first term on the right hand side. Defining
+    Consider the first term on the right hand side. By the mean value theorem,
     \begin{equation*}
-    w_n^*(t)  = 
-    \left(
+      \frac{1}{F(t-, 0)}
+      - \frac{1}{\hat{F}_n(t-, 0)}
+      = \frac{-1}{\tilde{r}_n(t)^2}
+      \left[
         F(t-, 0)
         - \hat{F}_n(t-, 0)
-    \right)
-    \left(
-      \frac{1}{F(t-, 0)}
-      - \frac{1}{\hat{F}_n(t-, 0)}
-    \right),
+      \right],
     \end{equation*}
-  we can write
+    where \( \tilde{r}_n(t) \) is some value between \( \hat{F}(t-, 0) \) and
+    \( \hat{F}_n(t-, 0) \). Letting \( w_n^*(t) = -\tilde{r}_n(t)^{-2} \), we
+    may write
   \begin{align*}
@@
       \frac{1}{\hat{F}_n(s-, 0)}
-      \right)F(\diff u, 2)F(\diff s, 1)
+      \right)F(\diff u, -1)F(\diff s, 1)
     \\
@@
       \right)       
-      F(\diff u, 2)F(\diff s, 1)
+      F(\diff u, -1)F(\diff s, 1)
     \\
@@
       \right)       
-      F(\diff u, 2)F(\diff s, 1),
+      F(\diff u, -1)F(\diff s, 1),
   \end{align*}
-  where we have defined \( w_n^a(s,u) = w(s)w^*_n(s)w^*_n(u) \). By assumption,
-  \( w_n^a(s,u) \) is uniformly bounded. The same approach can be applied to the
-  three remaining terms which gives the result.
+  where we have defined \( w_n^a(s,u) = w(s)w^*_n(s)w^*_n(u) \). By the
+  assumption that \( F(\blank,0) \) and \( F_n(\blank,0) \) are uniformly
+  bounded away from zero on \( [0,\tau] \), it follows that \( w_n^* \) is
+  uniformly bounded, and hence \( w_n^a(s,u) \) is also uniformly bounded,
+  because \( w \) was assumed uniformly bounded. The same approach can be
+  applied to the three remaining terms which gives the result.
 \end{proof}

### 2025-01-13 manuscript/manuscript.tex
updated manuscript and appendix to late fixes

@@
 
-We introduce our notation and framework in
-Section~\ref{sec:framework}. In Section~\ref{sec:super-learning} we define
-general super learning for right-censored data.
-Section~\ref{sec:super-learner-simple} introduces the state learner, and
-Section~\ref{sec:theor-results-prop} provides theoretical guarantees. In
-Section~\ref{sec:targeted-learning} we discuss the use of the state learner in
-the context of targeted learning. We report results of our numerical experiments
-in Section~\ref{sec:numer-exper} and analyse a prostate cancer data set in
-Section~\ref{sec:real-data-appl}. Section~\ref{sec:discussion} contains a
-discussion of the merits and limitations of our proposal.
-Appendices~\ref{sec:proof-proposition} and~\ref{sec:state-learner-with} contain
-proofs. An implementation of the state learner is available at
-\url{https://github.com/amnudn/statelearner} along with codes for reproducing
-our numerical experiments.
+We introduce our notation and framework in Section~\ref{sec:framework}. In
+Section~\ref{sec:super-learning} we define general super learning for
+right-censored data. Section~\ref{sec:super-learner-simple} introduces the state
+learner, and Section~\ref{sec:theor-results-prop} provides theoretical
+guarantees. In Section~\ref{sec:targeted-learning} we discuss the use of the
+state learner in the context of targeted learning. We report results of our
+numerical experiments in Section~\ref{sec:numer-exper} and analyse a prostate
+cancer data set in Section~\ref{sec:real-data-appl}.
+Section~\ref{sec:discussion} contains a discussion of the merits and limitations
+of our proposal. Proofs are given in the Appendix. An implementation of the
+state learner is available at \url{https://github.com/amnudn/statelearner} along
+with codes for reproducing our numerical experiments.
 
@@
 \begin{proof}
-  See Appendix~\ref{sec:proof-proposition}.
+  See Appendix~\ref{sec:consistency-proof}.
 \end{proof}
@@
 \begin{proof}
-  See Appendix~\ref{sec:proof-proposition}.
+  See Appendix~\ref{sec:oracle-inequalities}.
 \end{proof}
@@
 \subsection{Transience of the second order remainder structure}
+\label{sec:trans-second-order}
 
@@
 
+\section{Competing interests}
+No competing interest is declared.
+
+\bibliography{bib.bib}
+
 \appendix
 
-\section{Theoretical guarantees for the state learner}
-\label{sec:proof-proposition}
-
-In this section we provide proofs of the results stated in
-Section~\ref{sec:theor-results-prop}.
+\section{Proofs}
+\label{sec:proofs}
+
+This section contains proofs of the results stated in the paper.
+Section~\ref{sec:consistency-proof} contains a proof of the consistency result
+from Section~\ref{sec:consistency}; Section~\ref{sec:oracle-inequalities}
+contains proofs of the oracle inequalities from
+Section~\ref{sec:finite-sample-oracle}; and
+Section~\ref{sec:state-learner-with} demonstrates transience of the second
+order remainder structure stated in Section~\ref{sec:trans-second-order}.
+
+% We start by recalling some notation. We use \( F \) to denote a function that
+% describes the conditional state occupation probabilities of the observed data,
+% as defined in equation~(5) of the main paper. We denote the integrated Brier
+% score as \( \bar B_\tau( F,O) = \int_0^{\tau} B_t(F,O) \diff t \), where
+% \( B_t \) is defined
+% \begin{equation*}
+%   B_t(F,O) = \sum_{j=-1}^{2}
+%   \left(
+%       F(t,j,X) - \1{\{\eta(t)=j\}}
+%   \right)^2.
+% \end{equation*}
+% Finally, recall that we let $\mathcal{H}_{\mathcal{P}}$ denote the function
+% space consisting of all conditional state occupation probability functions for
+% some measure \( P \in \mathcal{P} \), and that we equipped this space with the
+% norm
+% \begin{equation}
+%   \label{eq:norm}
+%   \| F \|_{P_0} = 
+%   \left\{
+%     \sum_{j=-1}^{2}\int_{\mathcal{X}} \int_0^{\tau} F(t, j, x)^2 \diff t H_0( \diff x)
+%   \right\}^{1/2},
+% \end{equation}
+% for some fixed measure \( P_0 \in \mathcal{P} \). We use \( F_0 \) to denote the
+% conditional state occupation probability function associated with the measure
+% \( P_0 \).
+
+\subsection{Consistency}
+\label{sec:consistency-proof}
 
@@
 \( \bar{B}_{\tau,0}(F, o) = \bar{B}_{\tau}(F, o) - \bar{B}_{\tau}(F_0, o) \) and
-\( R_{0}(F) = P_0{[\bar{B}_{\tau,0}(F, \blank)]} \).
+\( R_{0}(F) = P_0{[\bar{B}_{\tau,0}(F, \blank)]} \), where the integrated Brier
+score \( \bar{B}_{\tau} \) was defined in
+Section~\ref{sec:super-learner-simple}.
 \begin{lemma}
@@
 
-Recall that $\mathcal{H}_{\mathcal{P}}$ denote the function space consisting of
-all conditional state occupation probability functions for some measure
-\( P \in \mathcal{P} \).
+\subsection{Oracle inequalities}
+\label{sec:oracle-inequalities}
+
+Recall that we use \( \mathcal{F}_n \) to denote a library of learners for the
+function \( F \), and that \( \hat{\phi} \) and \( \tilde{\phi} \) denotes,
+respectively, the discrete super learner and the oracle learner for the library
+\( \mathcal{F}_n \), c.f., Section~\ref{sec:super-learner-simple}.
 
@@
   identical. By Lemma~\ref{lemma:norm}, \( R_0(F) \geq 0 \) for any
-  \( F \in \mathcal{H}_{\mathcal{P}} \), and so using Theorem 2.3 from \citep{vaart2006oracle} with
-  \( p=1 \), we have that for all \( \delta >0 \),
+  \( F \in \mathcal{H}_{\mathcal{P}} \), and so using Theorem 2.3 from
+  \citep{vaart2006oracle} with \( p=1 \), we have that for all \( \delta >0 \),
 \begin{align*}
@@
 
-\section{The state learner with targeted learning}
+\subsection{Transience of the second order remainder structure}
 \label{sec:state-learner-with}
 
-In this section show that a product structure is preserved when an estimator
-$\bar\Psi(\hat{F}_n, \hat{H}_n)$ is used instead of
-$\tilde\Psi(\hat{\Lambda}_n, \hat{\Gamma}_n, \hat{H}_n)$.
+Recall that we let $\Lambda_1$ denote the conditional cumulative hazard function
+for one of the event times of interest and $\Gamma$ the conditional cumulative
+hazard function for censoring, c.f., Section~\ref{sec:framework}.
 
@@
   result can be obtained by adding the argument \( X \) to all functions and
-  averaging. We use the relations from equation~(\ref{eq:7}) to write
+  averaging. We use the relations from equation~\eqref{eq:7} to write
   \begingroup %
@@
     \endgroup %
-    Consider the first term on the right hand side. Defining
-  \begin{equation*}
-    w_n^*(t)  = 
-    \left(
+    Consider the first term on the right hand side. By the mean value theorem,
+    \begin{equation*}
+      \frac{1}{F(t-, 0)}
+      - \frac{1}{\hat{F}_n(t-, 0)}
+      = \frac{-1}{\tilde{r}_n(t)^2}
+      \left[
         F(t-, 0)
         - \hat{F}_n(t-, 0)
-    \right)
-    \left(
-      \frac{1}{F(t-, 0)}
-      - \frac{1}{\hat{F}_n(t-, 0)}
-    \right),
-  \end{equation*}
-  we can write
+      \right],
+    \end{equation*}
+    where \( \tilde{r}_n(t) \) is some value between \( \hat{F}(t-, 0) \) and
+    \( \hat{F}_n(t-, 0) \). Letting \( w_n^*(t) = -\tilde{r}_n(t)^{-2} \), we
+    may write
   \begin{align*}
@@
       \right)       
-      F(\diff u, 2)F(\diff s, 1),
+      F(\diff u, -1)F(\diff s, 1),
   \end{align*}
-  where we have defined \( w_n^a(s,u) = w(s)w^*_n(s)w^*_n(u) \). By assumption,
-  \( w_n^a(s,u) \) is uniformly bounded. The same approach can be applied to the
-  three remaining terms which gives the result.
+  where we have defined \( w_n^a(s,u) = w(s)w^*_n(s)w^*_n(u) \). By the
+  assumption that \( F(\blank,0) \) and \( F_n(\blank,0) \) are uniformly
+  bounded away from zero on \( [0,\tau] \), it follows that \( w_n^* \) is
+  uniformly bounded, and hence \( w_n^a(s,u) \) is also uniformly bounded,
+  because \( w \) was assumed uniformly bounded. The same approach can be
+  applied to the three remaining terms which gives the result.
 \end{proof}
-
-\bibliography{bib.bib}
-
 \end{document}

### 2025-01-13 manuscript/manuscript-lda.tex
manuscript template for lda

--- 2025-01-13 manuscript/manuscript.tex
@@
-\documentclass[11pt]{article}
-\usepackage[a4paper,left=2.8cm, right=2.8cm, top=3cm, bottom=3cm]{geometry}
-
-%% Packages
-\usepackage{amsthm, dsfont, amssymb,soul,xcolor, enumitem, amsmath,verbatim}
-
+%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% This is a general template file for the LaTeX package SVJour3
+% for Springer journals.          Springer Heidelberg 2010/09/16
+%
+% Copy it to a new file with a new name and use it as the basis
+% for your article. Delete % signs as needed.
+%
+% This template includes a few options for different layouts and
+% content for various journals. Please consult a previous issue of
+% your journal as needed.
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% First comes an example EPS file -- just ignore it and
+% proceed on the \documentclass line
+% your LaTeX will extract the file if required
+\begin{filecontents*}{example.eps}
+%!PS-Adobe-3.0 EPSF-3.0
+%%BoundingBox: 19 19 221 221
+%%CreationDate: Mon Sep 29 1997
+%%Creator: programmed by hand (JK)
+%%EndComments
+gsave
+newpath
+  20 20 moveto
+  20 220 lineto
+  220 220 lineto
+  220 20 lineto
+closepath
+2 setlinewidth
+gsave
+  .4 setgray fill
+grestore
+stroke
+grestore
+\end{filecontents*}
+%
+\RequirePackage{fix-cm}
+%
+%\documentclass{svjour3}                     % onecolumn (standard format)
+%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
+\documentclass[smallextended]{svjour3}       % onecolumn (second format)
+%\documentclass[twocolumn]{svjour3}          % twocolumn
+%
+\smartqed  % flush right qed marks, e.g. at end of proof
+%
+\usepackage{graphicx}
+%
+% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
+%
+% insert here the call for the packages your document requires
+\usepackage{dsfont, amssymb,soul,xcolor, enumitem, amsmath,verbatim}
 \usepackage[authoryear]{natbib}
-\bibliographystyle{abbrvnat}
-
 \definecolor{linkcolor}{rgb}{0, 0, 0.54}
 \usepackage[colorlinks,allcolors=linkcolor]{hyperref}
-
-\theoremstyle{thmstyleone}%\
-\newtheorem{theorem}{Theorem}%  meant for continuous numbers
-\newtheorem{proposition}{Proposition}
-\newtheorem{lemma}{Lemma}
-\newtheorem{corollary}{Corollary}
-\theoremstyle{thmstyletwo}%
-\newtheorem{example}{Example}%
-\newtheorem{remark}{Remark}%
-\theoremstyle{thmstylethree}%
-\newtheorem{definition}{Definition}
-
-%% Line spacing
-\linespread{2}
-
-% \usepackage{lineno}
-% \linenumbers
-
-\usepackage{endfloat}
-
-%% Only for comments:
-\usepackage[author=]{fixme}
-\fxusetheme{color}
-\definecolor{fxtarget}{rgb}{.5,.5,.5}
-\definecolor{fxnote}{rgb}{.5,.5,.5}
-\fxsetup{status=draft}
-
-%% Toggle with or without figures
-% \def\nofig{1}
-
-\RequirePackage{graphicx}
-\usepackage[font=small]{caption}
-
-\newcommand{\includeFigCond}[2][]{
-  \ifx\nofig\undefined %
-    \includegraphics[#1]{#2} %
-  \else %
-    \texttt{#2} %
-  \fi %
-}
-
-% New operators and commands
+%
+% please place your own definitions here and don't use \def but
 \DeclareMathOperator{\E}{\mathbb{E}} % expectation
@@
 \newcommand{\data}{\ensuremath{\mathcal{D}}}
-
-%% Title
-\title{\Large The state learner -- a super learner for right-censored data}
-\date{\small \today}
-
-\usepackage{authblk}
-\author[1,*]{\normalsize Anders Munch}
-\author[1]{\normalsize Thomas~A.~Gerds}
-\affil[1]{\normalsize Section of Biostatistics, University of Copenhagen}
-\affil[*]{\normalsize Address for correspondence: \href{email:email-id.com}{a.munch@sund.ku.dk}}
-
-\usepackage{fancyhdr}
-\fancyhf{}
-\fancyhead[l]{\textit{Munch et al.}}
-\fancyhead[r]{\textit{The state learner}}
-\fancyfoot[c]{\thepage}
-\renewcommand\headrulewidth{0pt}
-\pagestyle{fancy}
-
+%
+% Insert the name of "your journal" with
+\journalname{Lifetime Data Analysis}
+%
 \begin{document}
+
+\title{The state learner -- a super learner for right-censored data
+  % \thanks{Grants or other notes about the article that should go on the front
+  % page should be placed here. General acknowledgments should be placed at the
+  % end of the article.}
+}
+% \subtitle{Do you have a subtitle?\\ If so, write it here}
+
+\titlerunning{The state learner}        % if too long for running head
+
+\author{Anders Munch         \and
+        Thomas~A.~Gerds %etc.
+}
+
+%\authorrunning{Short form of author list} % if too long for running head
+
+\institute{Anders Munch \at
+              Section of Biostatistics, University of Copenhagen \\
+              % Tel.: +123-45-678910\\
+              % Fax: +123-45-678910\\
+              \email{a.munch@sund.ku.dk}           %  \\
+%             \emph{Present address:} of F. Author  %  if needed
+              \and %
+              Thomas~A.~Gerds %
+              \at Section of Biostatistics, University of Copenhagen }
+
+\date{Received: date / Accepted: date}
+% The correct dates will be entered by the editor
 
@@
   statistical functional.
+
+  \keywords{Competing risks, cross-validation, loss based estimation,
+    right-censored data, super learner}
+% \PACS{PACS code1 \and PACS code2 \and more}
+% \subclass{MSC code1 \and MSC code2 \and more}
 \end{abstract}
-
-% \keywords{Competing risks, cross-validation, loss based estimation,
-% right-censored data, super learner}
-
-\textit{{\small \textbf{Keywords:} Competing risks, cross-validation, loss based
-    estimation, right-censored data, super learner}}
-
-\pagebreak
 
@@
 
-% ANDERS: hedder det ikke Nelson-Aalen i stedet for Kaplan-Meier?
-%         kan du skrive klart hvad outcome er i denne foerste del? er det combined endpoint som man kalder for tumor-free survival?  
 For all super learners we use a library consisting of three learners:
@@
   \centering %
-  \includeFigCond[width=1\linewidth]{experiment-fig-sl-ipcw.pdf}
+  \includegraphics[width=1\linewidth]{experiment-fig-sl-ipcw.pdf}
   \caption[]{For the risk prediction models provided by each of the super
@@
   \centering %
-  \includeFigCond[width=1\linewidth]{experiment-fig-sl-survSL-out.pdf}
+  \includegraphics[width=1\linewidth]{experiment-fig-sl-survSL-out.pdf}
   \caption[]{For the risk prediction models of the outcome provided by each
@@
   \centering %
-  \includeFigCond[width=.7\linewidth]{experiment-fig-sl-survSL-cens.pdf}
+  \includegraphics[width=.7\linewidth]{experiment-fig-sl-survSL-cens.pdf}
   \caption[]{For the risk prediction models of the censoring model provided by
@@
   \centering %
-  \includeFigCond[width=1\linewidth]{real-data-state-learner.pdf}
+  \includegraphics[width=1\linewidth]{real-data-state-learner.pdf}
   \caption[]{The results of applying the 125 combinations of learners to the
@@
 includes all levels of interaction.
-% ANDERS: er du sikker paa at det er ALL levels of interaction, ogsaa higher order?
-%         der mangler en reference for lasso model
 The results are shown in Figure~\ref{fig:zelefski-real-target} for 6
@@
   \centering%
-  \includeFigCond[width=1\linewidth]{real-data-target.pdf}
+  \includegraphics[width=1\linewidth]{real-data-target.pdf}
   \caption[]{Estimates of the average treatment effect of hormone therapy on the
@@
 distribution, or the censoring distribution, or both.
-% ANDERS: er du enig i min omformulering af forgaaende saetning?
 Specifically, the finite sample oracle inequality in
@@
 for estimation of \( S \).
-% ANDERS: den forgaaende paragraf gentager det samme 2-3 gang. kan du forkorte en gang?
-
-% In the context of targeted learning, a drawback of the state learner is that the
-% doubly robustness property defined in equation~(\ref{eq:dr-property}) seem to be
-% lost because we only use a single nuisance parameter estimator. As defined here,
-% however, the state learner is build using libraries for the conditional
-% cause-specific cumulative hazard functions, so some doubly robustness might be
-% preserved.
 
@@
 
-\section{Competing interests}
-No competing interest is declared.
-
-\bibliography{bib.bib}
+\section*{Conflict of interest}
+
+The authors declare that they have no conflict of interest.
+
+% BibTeX users please use one of
+% \bibliographystyle{spbasic}      % basic style, author-year citations
+%\bibliographystyle{spmpsci}      % mathematics and physical sciences
+%\bibliographystyle{spphys}       % APS-like style for physics
+
+\bibliographystyle{abbrvnat}
+\bibliography{bib.bib}   % name your BibTeX data base
 
@@
 \end{proof}
+
 \end{document}
+% end of file template.tex

### 2025-01-13 manuscript/manuscript-lda.tex
split some equations

@@
 \institute{Anders Munch \at
-              Section of Biostatistics, University of Copenhagen \\
+              Section of Biostatistics, University of Copenhagen, Denmark \\
               % Tel.: +123-45-678910\\
@@
               Thomas~A.~Gerds %
-              \at Section of Biostatistics, University of Copenhagen }
+              \at Section of Biostatistics, University of Copenhagen, Denmark}
 
@@
 Section~\ref{sec:discussion} contains a discussion of the merits and limitations
-of our proposal. Proofs are given in the Appendix. An implementation of the
-state learner is available at \url{https://github.com/amnudn/statelearner} along
-with codes for reproducing our numerical experiments.
+of our proposal. Proofs are given in Appendix~\ref{sec:proofs}. An
+implementation of the state learner is available at
+\url{https://github.com/amnudn/statelearner} along with code for reproducing our
+numerical experiments.
 
@@
   F(t, k, x) = P(\eta(t) = k \mid X=x),
-  \quad \text{for all} \quad
-  t \in [0,\tau],
-  k \in \{-1,0,1,2\},
-  x \in \mathcal{X}.
 \end{equation}
+for all \( t \in [0,\tau] \), \( k \in \{-1,0,1,2\} \), and \( x \in \mathcal{X} \).
 The function \( F \) describes the conditional state occupation probabilities of
@@
       = P(\tilde{T} > t \mid X= x)
-      =
+      \\
+      & =
       \exp{{\{-\Lambda_{1}(t \mid x)-\Lambda_{2}(t \mid x) - \Gamma(t \mid x)\}
@@
      & =
-      P{\left[
-      \int_0^{\tau} \int_0^{s} w^a_n(s,u,\blank) [F - \hat{F}_n](u-, 0, \blank)[F - \hat{F}_n](s-, 0, \blank) F(\diff u, -1, \blank ) F ( \diff s, 1, \blank)
-      \right]}
+      P{\big[
+      \int_0^{\tau} \int_0^{s} w^a_n(s,u,\blank) [F - \hat{F}_n](u-, 0,
+       \blank)[F - \hat{F}_n](s-, 0, \blank)}
+    \\
+    & \qquad \quad \times F(\diff u, -1, \blank ) F ( \diff s, 1, \blank)
+      \big]
     \\
@@
     \label{eq:one-step-def}
+    \begin{split}
       \hat{\Psi}_{\text{OS}}(\data_n)
-    =
+      = &
           \tilde{\Psi}^0(\hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
           H_n(\data_n))
-    + \empmeas{[\psi(\blank; \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
+      \\
+        & + \empmeas{[\psi(\blank; \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
           \hat{b}_n(\data_n) )]},
+    \end{split}
 \end{equation}
@@
 cause 1 at time $\tau$:
-\begin{equation*}
-  \theta_{{\tau}}(Q) = \int_{\mathcal{W}} 
+\begin{align*}
+  & \theta_{{\tau}}(Q)
+  \\
+  & = \int_{\mathcal{W}} 
     \left\{
@@
     \mu(\diff w).
-\end{equation*}
+\end{align*}
 Under the usual assumptions for causal inference (consistency,
@@
   \begin{split}
-    \tilde{\Psi}_{\tau}^0(\Lambda_1, \Lambda_2, \mu)
+    & \tilde{\Psi}_{\tau}^0(\Lambda_1, \Lambda_2, \mu)
+    \\
     & =
@@
 
-% We start by recalling some notation. We use \( F \) to denote a function that
-% describes the conditional state occupation probabilities of the observed data,
-% as defined in equation~(5) of the main paper. We denote the integrated Brier
-% score as \( \bar B_\tau( F,O) = \int_0^{\tau} B_t(F,O) \diff t \), where
-% \( B_t \) is defined
-% \begin{equation*}
-%   B_t(F,O) = \sum_{j=-1}^{2}
-%   \left(
-%       F(t,j,X) - \1{\{\eta(t)=j\}}
-%   \right)^2.
-% \end{equation*}
-% Finally, recall that we let $\mathcal{H}_{\mathcal{P}}$ denote the function
-% space consisting of all conditional state occupation probability functions for
-% some measure \( P \in \mathcal{P} \), and that we equipped this space with the
-% norm
-% \begin{equation}
-%   \label{eq:norm}
-%   \| F \|_{P_0} = 
-%   \left\{
-%     \sum_{j=-1}^{2}\int_{\mathcal{X}} \int_0^{\tau} F(t, j, x)^2 \diff t H_0( \diff x)
-%   \right\}^{1/2},
-% \end{equation}
-% for some fixed measure \( P_0 \in \mathcal{P} \). We use \( F_0 \) to denote the
-% conditional state occupation probability function associated with the measure
-% \( P_0 \).
-
 \subsection{Consistency}

### 2025-05-21 manuscript/manuscript-one.tex
divide into two manuscripts and basic template

--- 2025-01-13 manuscript/manuscript-lda.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
-%
-% This is a general template file for the LaTeX package SVJour3
-% for Springer journals.          Springer Heidelberg 2010/09/16
-%
-% Copy it to a new file with a new name and use it as the basis
-% for your article. Delete % signs as needed.
-%
-% This template includes a few options for different layouts and
-% content for various journals. Please consult a previous issue of
-% your journal as needed.
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-% First comes an example EPS file -- just ignore it and
-% proceed on the \documentclass line
-% your LaTeX will extract the file if required
-\begin{filecontents*}{example.eps}
-%!PS-Adobe-3.0 EPSF-3.0
-%%BoundingBox: 19 19 221 221
-%%CreationDate: Mon Sep 29 1997
-%%Creator: programmed by hand (JK)
-%%EndComments
-gsave
-newpath
-  20 20 moveto
-  20 220 lineto
-  220 220 lineto
-  220 20 lineto
-closepath
-2 setlinewidth
-gsave
-  .4 setgray fill
-grestore
-stroke
-grestore
-\end{filecontents*}
-%
-\RequirePackage{fix-cm}
-%
-%\documentclass{svjour3}                     % onecolumn (standard format)
-%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
-\documentclass[smallextended]{svjour3}       % onecolumn (second format)
-%\documentclass[twocolumn]{svjour3}          % twocolumn
-%
-\smartqed  % flush right qed marks, e.g. at end of proof
-%
-\usepackage{graphicx}
-%
-% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
-%
-% insert here the call for the packages your document requires
-\usepackage{dsfont, amssymb,soul,xcolor, enumitem, amsmath,verbatim}
-\usepackage[authoryear]{natbib}
+\documentclass[a4paper,danish]{article}
+\usepackage[margin=3.5cm]{geometry}
+
+\usepackage[utf8]{inputenc}
+\usepackage[T1]{fontenc}
+\usepackage{natbib, dsfont, enumitem,
+  amssymb,soul,xcolor,amsmath,amsthm,graphicx,subcaption,verbatim,pgfplots,tikz,prodint}
+\usetikzlibrary{calc,patterns,angles,quotes,automata,
+  positioning,arrows,shapes}
+
 \definecolor{linkcolor}{rgb}{0, 0, 0.54}
 \usepackage[colorlinks,allcolors=linkcolor]{hyperref}
-%
-% please place your own definitions here and don't use \def but
+
+% Ref and bibliography style 
+\bibliographystyle{abbrvnat}
+
+% Handling new lines
+\setlength{\parskip}{1em}
+\setlength{\parindent}{0em}
+
+% Handling space after sections
+\usepackage{titlesec}
+\titlespacing*{\section}{0em}{2em}{0em}
+\titlespacing*{\subsection}{0em}{2em}{0em}
+\titlespacing*{\subsubsection}{0em}{2em}{0em}
+
+% No spacing after in start of list
+\setlist[itemize]{topsep=0pt}
+\setlist[enumerate]{topsep=0pt}
+
+% Theorem environments 
+\theoremstyle{plain} % plain italic style
+\newtheorem{theorem}{Theorem}
+\numberwithin{theorem}{section}
+\newtheorem*{theorem*}{Theorem}
+\newtheorem{conjecture}{Conjecture}
+\newtheorem{lemma}[theorem]{Lemma}
+\newtheorem{proposition}[theorem]{Proposition}
+\newtheorem{corollary}[theorem]{Corollary}
+
+\theoremstyle{definition} % non-italic
+\newtheorem{definition}[theorem]{Definition}
+\newtheorem{assumption}[theorem]{Assumption}
+\newtheorem*{assumption*}{Assumption}
+
+\theoremstyle{remark}
+\newtheorem{remark}[theorem]{Remark}
+\newtheorem{example}[theorem]{Example}
+\newtheorem{note}[theorem]{Note}
+
+% Notation
 \DeclareMathOperator{\E}{\mathbb{E}} % expectation
@@
 \newcommand{\data}{\ensuremath{\mathcal{D}}}
-%
-% Insert the name of "your journal" with
-\journalname{Lifetime Data Analysis}
-%
+
+\title{The state learner -- a super learner for right-censored
+  data}
+\author{Anders Munch \& Thomas A. Gerds}
+\date{\today}
+
 \begin{document}
-
-\title{The state learner -- a super learner for right-censored data
-  % \thanks{Grants or other notes about the article that should go on the front
-  % page should be placed here. General acknowledgments should be placed at the
-  % end of the article.}
-}
-% \subtitle{Do you have a subtitle?\\ If so, write it here}
-
-\titlerunning{The state learner}        % if too long for running head
-
-\author{Anders Munch         \and
-        Thomas~A.~Gerds %etc.
-}
-
-%\authorrunning{Short form of author list} % if too long for running head
-
-\institute{Anders Munch \at
-              Section of Biostatistics, University of Copenhagen, Denmark \\
-              % Tel.: +123-45-678910\\
-              % Fax: +123-45-678910\\
-              \email{a.munch@sund.ku.dk}           %  \\
-%             \emph{Present address:} of F. Author  %  if needed
-              \and %
-              Thomas~A.~Gerds %
-              \at Section of Biostatistics, University of Copenhagen, Denmark}
-
-\date{Received: date / Accepted: date}
-% The correct dates will be entered by the editor
-
 \maketitle
 
+% %%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
+% %
+% % This is a general template file for the LaTeX package SVJour3
+% % for Springer journals.          Springer Heidelberg 2010/09/16
+% %
+% % Copy it to a new file with a new name and use it as the basis
+% % for your article. Delete % signs as needed.
+% %
+% % This template includes a few options for different layouts and
+% % content for various journals. Please consult a previous issue of
+% % your journal as needed.
+% %
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% %
+% % First comes an example EPS file -- just ignore it and
+% % proceed on the \documentclass line
+% % your LaTeX will extract the file if required
+% \begin{filecontents*}{example.eps}
+% %!PS-Adobe-3.0 EPSF-3.0
+% %%BoundingBox: 19 19 221 221
+% %%CreationDate: Mon Sep 29 1997
+% %%Creator: programmed by hand (JK)
+% %%EndComments
+% gsave
+% newpath
+%   20 20 moveto
+%   20 220 lineto
+%   220 220 lineto
+%   220 20 lineto
+% closepath
+% 2 setlinewidth
+% gsave
+%   .4 setgray fill
+% grestore
+% stroke
+% grestore
+% \end{filecontents*}
+% %
+% \RequirePackage{fix-cm}
+% %
+% %\documentclass{svjour3}                     % onecolumn (standard format)
+% %\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
+% \documentclass[smallextended]{svjour3}       % onecolumn (second format)
+% %\documentclass[twocolumn]{svjour3}          % twocolumn
+% %
+% \smartqed  % flush right qed marks, e.g. at end of proof
+% %
+% \usepackage{graphicx}
+% %
+% % \usepackage{mathptmx}      % use Times fonts if available on your TeX system
+% %
+% % insert here the call for the packages your document requires
+% \usepackage{dsfont, amssymb,soul,xcolor, enumitem, amsmath,verbatim}
+% \usepackage[authoryear]{natbib}
+% \definecolor{linkcolor}{rgb}{0, 0, 0.54}
+% \usepackage[colorlinks,allcolors=linkcolor]{hyperref}
+% %
+% % please place your own definitions here and don't use \def but
+% \DeclareMathOperator{\E}{\mathbb{E}} % expectation
+% \newcommand{\Z}{\mathbb{Z}}
+% \newcommand{\R}{\mathbb{R}}
+% \newcommand{\N}{\mathbb{N}}
+% \newcommand{\C}{\mathbb{C}}
+% \renewcommand{\S}{\mathbb{S}}
+% \newcommand{\blank}{\makebox[1ex]{\textbf{$\cdot$}}}
+% \newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
+% \def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
+% \renewcommand{\phi}{\varphi}
+% \renewcommand{\epsilon}{\varepsilon}
+% \newcommand*\diff{\mathop{}\!\mathrm{d}}
+% \newcommand{\weakly}{\rightsquigarrow}
+% \newcommand\smallO{\textit{o}}
+% \newcommand\bigO{\textit{O}}
+% \newcommand{\midd}{\; \middle|\;}
+% \newcommand{\1}{\mathds{1}}
+% \usepackage{ifthen} %% Empirical process with default argument
+% \newcommand{\G}[2][n]{
+% {\ensuremath{\mathbb{G}_{#1}}{\left[#2\right]}}
+% }
+% \DeclareMathOperator*{\argmin}{\arg\!\min}
+% \DeclareMathOperator*{\argmax}{\arg\!\max}
+% \newcommand{\empmeas}{\ensuremath{\mathbb{P}_n}} % empirical measure
+% \newcommand{\data}{\ensuremath{\mathcal{D}}}
+
+% \begin{document}
+
+% \title{The state learner -- a super learner for right-censored
+%   data}
+
+% \author{Anders Munch \& Thomas A. Gerds}
+
+\maketitle
+
 \begin{abstract}
-  In survival analysis, prediction models are needed as stand-alone tools and in
-  applications of causal inference to estimate nuisance parameters. The super
-  learner is a machine learning algorithm which combines a library of prediction
-  models into a meta learner based on cross-validated loss. In right-censored
-  data, the choice of the loss function and the estimation of the expected loss
-  need careful consideration. We introduce the state learner, a new super
-  learner for survival analysis, which simultaneously evaluates libraries of
-  prediction models for the event of interest and the censoring distribution.
-  The state learner can be applied to all types of survival models, works in the
-  presence of competing risks, and does not require a single pre-specified
-  estimator of the conditional censoring distribution. We establish an oracle
-  inequality for the state learner and investigate its performance through
-  numerical experiments. We illustrate the application of the state learner with
-  prostate cancer data, as a stand-alone prediction tool, and, for causal
-  inference, as a way to estimate the nuisance parameter models of a smooth
-  statistical functional.
-
-  \keywords{Competing risks, cross-validation, loss based estimation,
-    right-censored data, super learner}
+  In survival analysis, prediction models are needed as
+  stand-alone tools and in applications of causal inference to
+  estimate nuisance parameters. The super learner is a machine
+  learning algorithm which combines a library of prediction
+  models into a meta learner based on cross-validated loss. In
+  right-censored data, the choice of the loss function and the
+  estimation of the expected loss need careful consideration.
+  We introduce the state learner, a new super learner for
+  survival analysis, which simultaneously evaluates libraries
+  of prediction models for the event of interest and the
+  censoring distribution. The state learner places no
+  restrictions on the algorithms that can be included in the
+  library, works in the presence of competing risks, and does
+  not require a single pre-specified estimator of the
+  conditional censoring distribution. We establish an oracle
+  inequality for the state learner and investigate its
+  performance through numerical experiments. We illustrate the
+  application of the state learner with prostate cancer data.
+
+  % \keywords{Competing risks, cross-validation, loss based estimation,
+  %   right-censored data, super learner}
 % \PACS{PACS code1 \and PACS code2 \and more}
@@
 
-% Medical decision making is traditionally based on the discrepancy
-% between intuitive judgment and the law of chance
-% \citep{redelmeier1995probability}.
-% We are concerned with the statistical models that provide
-% probabilistic predictions about the medical future of a patient. We
-% have two different applications of these models in mind. One is as a
-% stand-alone tool which predicts the likelihood of medical outcomes.
-% The other is as nuisance parameter estimate
-% input to an algorithm which estimates a causal target
-% parameter such as an average treatment effect and depends on estimates
-% of nuisance parameters such as the conditional probability of being
-% uncensored.  An example of the latter is the targeted minimum loss
-% based estimator \citep{rytgaard2022targeted}. In both applications,
-% the models have to be learned based on available data from previous
-% patients.
-A super learner is a machine learning algorithm that combines a finite
-set of learners into a meta learner by estimating prediction
-performance in hold-out samples using a pre-specified loss function
-\citep{van2007super}. When the aim is to make a prediction model,
-super learners combine strong learners, such as Cox regression models
-and random survival forests \citep[][Section 8.4]{gerds2021medical}.
-% For targeted learning of a low dimensional
-% target parameter in presence of a high-dimensional nuisance parameter,
-% a super learner will usually also include the highly adaptive lasso
-% \citep{van2011targeted, benkeser2016highly,van2017generally}.
-While the general idea of combining strong learners based on cross-validation
-stems from earlier work \citep{wolpert1992stacked,breiman1996stacked}, the
-name super learner is justified by an oracle inequality
+[More motivation]
+
+A super learner is a machine learning algorithm that combines
+a finite set of learners into a meta learner by estimating
+prediction performance in hold-out samples using a
+pre-specified loss function \citep{van2007super}. When the aim
+is to make a prediction model, super learners combine strong
+learners, such as Cox regression models and random survival
+forests \citep[][Section 8.4]{gerds2021medical}. While the
+general idea of combining strong learners based on
+cross-validation stems from earlier work
+\citep{wolpert1992stacked,breiman1996stacked}, the name super
+learner is justified by an oracle inequality
 \citep{van2003unicv,vaart2006oracle}.
-% The super learner is guaranteed to perform
-% almost as well as the model which minimises the expected performance, i.e., the
-% model we would select if we could evaluate the prediction performance in an
-% infinite hold-out sample.
-
-We define the state learner, a new super learner for right-censored data, which
-simultaneously estimates the expected loss of learners of the event time
-distribution and the censoring distribution. Compared to the super learners of
-\cite{polley2011-sl-cens} and \cite{golmakani2020super}, the state learner
-proposed here poses no restrictions on the type of learners that can be included
-in the library. Compared to super learners that rely on inverse probability of
-censoring weighting, censoring unbiased transformations, or pseudo-values, the
-state learner does not require a pre-specified censoring distribution. Compared
-to the super learners by \cite{han2021inverse} and \cite{westling2021inference},
-the state learner has theoretical guarantees which these methods do not have.
-% Moreover, state learner seems to be the first super learner for
-% right censored data which learns the outcome and the censoring
-% distribution simulaneously.
-
-The loss function underlying the state learner operates on coarsened data. In
-the right-censored survival setting the coarsened data consist of the minimum
-and the order of the censoring time and the event time as well as the baseline
-covariates. The state learner works by simultaneously assessing how well
-learners of the event time distribution and the censoring distribution predict
-the states of the coarsened data. To analyse the theoretical properties of the
-state learner we focus on the discrete super learner which combines the library
-of learners by picking the one that minimises the cross-validated loss
-\citep{van2007super}. In the presence of competing risks, our algorithm uses
-separate libraries of learners for the cumulative hazard functions for each of
-the competing risks and for the censoring distribution. We show that the oracle
-selector of the state learner is consistent if all libraries contain a
-consistent learner and prove a finite sample oracle inequality.
-
-The state learner algorithm can output a medical risk prediction model
-\citep{gerds2021medical} which predicts the probability of an event based on
-covariates in the presence of competing risks. Another important application,
-which motivated our contribution, is in targeted learning where conditional
-event probabilities occur as high-dimensional nuisance parameters which need to
-be estimated at a certain rate \citep{van2011targeted, rytgaard2021estimation,
-  rytgaard2022targeted}. The state learner is particularly well-suited for this
-application as it simultaneously provides estimates for both the outcome and the
-censoring, both of which are needed for targeted learning. For the asymptotic
-bias term of a targeted estimator, which uses the state learner to estimate
-nuisance parameters, we show that a second order product structure holds. We
-illustrate both applications of the state learner with prostate cancer data
+
+We define the state learner, a new super learner for
+right-censored data, which simultaneously estimates the
+expected loss of learners of the event time distribution and
+the censoring distribution. Compared to the super learners of
+\cite{polley2011-sl-cens} and \cite{golmakani2020super}, the
+state learner proposed here poses no restrictions on the type
+of learners that can be included in the library. Compared to
+super learners that rely on inverse probability of censoring
+weighting, censoring unbiased transformations, or
+pseudo-values, the state learner does not require a
+pre-specified censoring distribution. Compared to the super
+learners by \cite{han2021inverse} and
+\cite{westling2021inference}, the state learner has
+theoretical guarantees which these methods do not have.
+
+The loss function underlying the state learner operates on
+coarsened data. In the right-censored survival setting the
+coarsened data consist of the minimum and the order of the
+censoring time and the event time as well as the baseline
+covariates. The state learner works by simultaneously
+assessing how well learners of the event time distribution and
+the censoring distribution predict the states of the coarsened
+data. To analyse the theoretical properties of the state
+learner we focus on the discrete super learner which combines
+the library of learners by picking the one that minimises the
+cross-validated loss \citep{van2007super}. In the presence of
+competing risks, our algorithm uses separate libraries of
+learners for the cumulative hazard functions for each of the
+competing risks and for the censoring distribution. We show
+that the oracle selector of the state learner is consistent if
+all libraries contain a consistent learner and prove a finite
+sample oracle inequality.
+
+The state learner algorithm can output a medical risk
+prediction model \citep{gerds2021medical} which predicts the
+probability of an event based on covariates in the presence of
+competing risks. Another important application, which
+motivated our contribution, is in targeted learning where
+conditional event probabilities occur as high-dimensional
+nuisance parameters which need to be estimated at a certain
+rate \citep{van2011targeted, rytgaard2021estimation,
+  rytgaard2022targeted}. The state learner is particularly
+well-suited for this application as it simultaneously provides
+estimates for both the outcome and the censoring, both of
+which are needed for targeted learning. For the asymptotic
+bias term of a targeted estimator, which uses the state
+learner to estimate nuisance parameters, we show that a second
+order product structure holds. We illustrate both applications
+of the state learner with prostate cancer data
 \citep{kattan2000pretreatment}.
 
-There are many existing super learners for right-censored data. We briefly
-review them here. Machine learning based on right-censored data commonly uses
-the partial log-likelihood as a loss function
+There are many existing super learners for right-censored
+data. We briefly review them here. Machine learning based on
+right-censored data commonly uses the partial log-likelihood
+as a loss function
 \citep[e.g.,][]{li2016regularized,yao2017deep,lee2018deephit,katzman2018deepsurv,gensheimer2019scalable,lee2021boosted,kvamme2021continuous}.
-This loss function is also suggested for super learning with right-censored data
-by \cite{polley2011-sl-cens}, where data is assumed to be observed in discrete
-time. However, the partial log-likelihood loss does not work well with data
-splitting (cross-validation) in continuous time. The reason is that the partial
-log-likelihood loss assigns an infinite value for any learner that predicts
-piece-wise constant cumulative hazard functions when there are event times in
-the test set which do not occur in the learning set. This problem occurs with
-prominent survival learners including the Kaplan-Meier estimator, the random
-survival forest, and the semi-parametric Cox regression model and these learners
-cannot be included in the library of the super learner proposed by
-\cite{polley2011-sl-cens}. When a proportional hazards model is assumed, the
-baseline hazard function can be profiled out of the likelihood
-\citep{cox1972regression}. The cross-validated partial log-likelihood loss
-\citep{verweij1993cross} has therefore been suggested as a loss function for
-super learning by \cite{golmakani2020super}. This choice of loss function
-restricts the library of learners to include only Cox proportional hazards
-models, and hence excludes many learners such as, e.g., random survival forests,
-additive hazards models, and accelerated failure time models. Other proposals
-for super learning with right-censored data use an inverse probability of
-censoring weighted (IPCW) loss function
+This loss function is also suggested for super learning with
+right-censored data by \cite{polley2011-sl-cens}, where data
+is assumed to be observed in discrete time. However, the
+partial log-likelihood loss does not work well with data
+splitting (cross-validation) in continuous time. The reason is
+that the partial log-likelihood loss assigns an infinite value
+for any learner that predicts piece-wise constant cumulative
+hazard functions when there are event times in the test set
+which do not occur in the learning set. This problem occurs
+with prominent survival learners including the Kaplan-Meier
+estimator, the random survival forest, and the semi-parametric
+Cox regression model and these learners cannot be included in
+the library of the super learner proposed by
+\cite{polley2011-sl-cens}. When a proportional hazards model
+is assumed, the baseline hazard function can be profiled out
+of the likelihood \citep{cox1972regression}. The
+cross-validated partial log-likelihood loss
+\citep{verweij1993cross} has therefore been suggested as a
+loss function for super learning by \cite{golmakani2020super}.
+This choice of loss function restricts the library of learners
+to include only Cox proportional hazards models, and hence
+excludes many learners such as, e.g., random survival forests,
+additive hazards models, and accelerated failure time models.
+Other proposals for super learning with right-censored data
+use an inverse probability of censoring weighted (IPCW) loss
+function
 \citep{graf1999assessment,van2003unicv,molinaro2004tree,keles2004asymptotically,hothorn2006survival,gerds2006consistent,gonzalez2021stacked},
 censoring unbiased transformations
-\citep{fan1996local,steingrimsson2019censoring}, or pseudo-values
-\citep{andersen2003generalised,mogensen2013random,sachs2019ensemble}. All these
-methods rely on an estimator of the censoring distribution, and their drawback
-is that this estimator has to be pre-specified. An approach which avoids a
-pre-specified censoring model was proposed independently by
-\cite{han2021inverse} and \cite{westling2021inference}. In both articles, the
-authors suggest to iterate between learning of the outcome model and learning of
-the censoring model using IPCW loss functions. However, it seems that this
-procedure is in general not guaranteed to converge to the true data-generating
-mechanism \citep[][Appendix~A.4]{munch2024thesis}.
-
-We introduce our notation and framework in Section~\ref{sec:framework}. In
-Section~\ref{sec:super-learning} we define general super learning for
-right-censored data. Section~\ref{sec:super-learner-simple} introduces the state
-learner, and Section~\ref{sec:theor-results-prop} provides theoretical
-guarantees. In Section~\ref{sec:targeted-learning} we discuss the use of the
-state learner in the context of targeted learning. We report results of our
-numerical experiments in Section~\ref{sec:numer-exper} and analyse a prostate
-cancer data set in Section~\ref{sec:real-data-appl}.
-Section~\ref{sec:discussion} contains a discussion of the merits and limitations
-of our proposal. Proofs are given in Appendix~\ref{sec:proofs}. An
-implementation of the state learner is available at
-\url{https://github.com/amnudn/statelearner} along with code for reproducing our
-numerical experiments.
+\citep{fan1996local,steingrimsson2019censoring}, or
+pseudo-values
+\citep{andersen2003generalised,mogensen2013random,sachs2019ensemble}.
+All these methods rely on an estimator of the censoring
+distribution, and their drawback is that this estimator has to
+be pre-specified. An approach which avoids a pre-specified
+censoring model was proposed independently by
+\cite{han2021inverse} and \cite{westling2021inference}. In
+both articles, the authors suggest to iterate between learning
+of the outcome model and learning of the censoring model using
+IPCW loss functions. However, it seems that this procedure is
+in general not guaranteed to converge to the true
+data-generating mechanism
+\citep[][Appendix~A.4]{munch2024thesis}.
+
+We introduce our notation and framework in
+Section~\ref{sec:framework}. In
+Section~\ref{sec:super-learning} we define general super
+learning for right-censored data.
+Section~\ref{sec:super-learner-simple} introduces the state
+learner, and Section~\ref{sec:theor-results-prop} provides
+theoretical guarantees. We report results of our numerical
+experiments in Section~\ref{sec:numer-exper} and analyse a
+prostate cancer data set in Section~\ref{sec:real-data-appl}.
+Section~\ref{sec:discussion} contains a discussion of the
+merits and limitations of our proposal. Proofs are given in
+Appendix~\ref{sec:proofs}. An implementation of the state
+learner is available at
+\url{https://github.com/amnudn/statelearner} along with code
+for reproducing our numerical experiments.
 
@@
 
-In survival analysis, a super learner can be used to estimate a
-parameter $\Psi$ which can be identified from the observed data
-distribution \(P\in\mathcal P\). In this section, to introduce the
-discrete super learner and the oracle learner, we consider estimation of the
-function-valued parameter \(\Psi:\mathcal P\to\mathcal M_{\tau}\), given by
-\(\Psi(P)=\Lambda_j\). This parameter is identified via equation
-\eqref{eq:lambdaj} on \([0,\tau]\).
-
-As input to the super learner we need a data set
-\( \data_n=\{O_i\}_{i=1}^n \) of i.i.d.\ observations from \( P \in \mathcal{P} \) and a collection of candidate
-learners $\mathcal{A}$. Each learner \(a \in \mathcal{A}\) is a map
-\( a \colon \mathcal{O}^n \rightarrow \mathcal{M}_{\tau}\) which takes a data
-set as input and returns an estimate $a(\data_n) \in \mathcal{M}_{\tau}$ of
-$\Lambda_{j}$.
+In survival analysis, a super learner can be used to estimate
+a parameter $\theta \in \Theta$ which can be identified from
+the observed data distribution \(P\in\mathcal P\). A super
+learner typically estimates a function, and thus the parameter
+space $\Theta$ is most often a class of functions. For
+example, $\theta$ could be a conditional survival function or
+a cumulative hazard function. In
+Section~\ref{sec:super-learner-simple} we consider a specific
+choice of parameter as target for our super learner, but in
+this section we describe super learning for a general
+parameter $\theta$ with values in some parameter space
+$\Theta$.
+
+As input to a super learner we need a data set
+\( \data_n=\{O_i\}_{i=1}^n \) of i.i.d.\ observations from
+\( P \in \mathcal{P} \) and a collection of candidate learners
+$\mathcal{A}$. Each learner \(a \in \mathcal{A}\) is a map
+\( a \colon \mathcal{O}^n \rightarrow \Theta \) which takes a
+data set as input and returns an estimate
+$a(\data_n) \in \Theta$ of $\theta$.
 % Formally, the domain of $a$ depends on \(n\) but we suppress this in the notation.
@@
 performance of \(a \in \mathcal{A}\) with a loss function
-\(L\colon \mathcal{M}_{\tau} \times \mathcal{O} \rightarrow \R_+\) and
+\(L\colon \Theta \times \mathcal{O} \rightarrow \R_+\) and
 estimates the expected loss \(P[L(a(\data_n), \blank)]\) using
@@
 
-The main idea of the state learner is to jointly use learners for
-\( \Lambda_1 \), \( \Lambda_2 \), and \( \Gamma \), and the relations in
-equation~(\ref{eq:parametrizeP}), to learn a feature of the observed data
-distribution \( P \). The discrete state learner ranks a tuple of learners for
-the tuple of the cumulative hazard functions
-\( (\Lambda_1, \Lambda_2, \Gamma) \) based on how well they jointly model the
-observed data. Risk predictions can then be obtained by combining
-\( \Lambda_1 \) and $\Lambda_2$ from the highest ranked tuple using a well-known
-formula \citep{benichou1990estimates, ozenne2017riskregression}. To formally
-introduce the state learner, we define the multi-state process
+The main idea of the state learner is to jointly use learners
+for \( \Lambda_1 \), \( \Lambda_2 \), and \( \Gamma \), and
+the relations in equation~(\ref{eq:parametrizeP}), to learn a
+feature of the observed data distribution \( P \). The
+discrete state learner ranks a tuple of learners for the tuple
+of the cumulative hazard functions
+\( (\Lambda_1, \Lambda_2, \Gamma) \) based on how well they
+jointly model the observed data. To formally introduce the
+state learner, we define the multi-state process
 \begin{equation*}
@@
 \end{equation}
-for all \( t \in [0,\tau] \), \( k \in \{-1,0,1,2\} \), and \( x \in \mathcal{X} \).
-The function \( F \) describes the conditional state occupation probabilities of
-the multi-state process \(\eta\). We construct a super learner for \( F \). The
-target of this super learner is the function-valued parameter $\Psi(P) = F$
-which is identified through equation~(\ref{eq:F-def}). Under conditional
-independent censoring each quadruple $(\Lambda_{1}, \Lambda_{2}, \Gamma, H)$
-characterises a distribution \(P\in\mathcal P\), c.f.\
-equation~\eqref{eq:parametrizeP}, which in turn determines \( (F, H) \). Hence,
-a learner for \( F \) can be constructed from learners for \( \Lambda_1 \),
-\( \Lambda_2 \), and $\Gamma$ as follows:
+for all \( t \in [0,\tau] \), \( k \in \{-1,0,1,2\} \), and
+\( x \in \mathcal{X} \). The function \( F \) describes the
+conditional state occupation probabilities of the multi-state
+process \(\eta\). We construct a super learner for \( F \).
+The target of this super learner is the function-valued
+parameter $\theta(P) = F$ which is identified through
+equation~(\ref{eq:F-def}). Under conditional independent
+censoring each quadruple
+$(\Lambda_{1}, \Lambda_{2}, \Gamma, H)$ characterises a
+distribution \(P\in\mathcal P\), c.f.\
+equation~\eqref{eq:parametrizeP}, which in turn determines
+\( (F, H) \). Hence, a learner for \( F \) can be constructed
+from learners for \( \Lambda_1 \), \( \Lambda_2 \), and
+$\Gamma$ as follows:
 \begin{equation}\label{eq:transition}
@@
 
+\section{Use cases for the state learner}
+\label{sec:use-cases-state}
+
+The parameter \( F \) is not of much interest...
+
+\subsection{Survival and risk predictions}
+\label{sec:surv-risk-pred}
+
+...
+
+Risk predictions can then be obtained by combining
+\( \Lambda_1 \) and $\Lambda_2$ from the highest ranked tuple
+using a well-known formula \citep{benichou1990estimates,
+  ozenne2017riskregression}.
+
+\subsection{Targeted and debiased machine learning}
+\label{sec:targeted-learning}
+
+... TL + SL = strong... The state learner is particularly well
+suited for targeted and debiased machine learning as it
+immediately provides estimates of the nuisance parameters
+pertaining to the outcome and the censoring models. While it
+was our original motivation for developing the state learning,
+we leave the study of the state learner in the context of
+targeted and debiased machine learning for a future paper.
+
 \section{Theoretical results for the state learner}
@@
 
-In this section we establish theoretical guarantees for the state learner. We
-show that the state learner is consistent if its library contains a consistent
-learner, and we establish a finite sample inequality for the excess risk of the
-state learner compared to the oracle. Finally we show that a certain second
-order structure is preserved when the state learner is used for targeted
-learning.
+In this section we establish theoretical guarantees for the
+state learner. We show that the state learner is consistent if
+its library contains a consistent learner, and we establish a
+finite sample inequality for the excess risk of the state
+learner compared to the oracle.
 
@@
 
-\subsection{Transience of the second order remainder structure}
-\label{sec:trans-second-order}
-
-In this section we demonstrate a theoretical property of the state
-learner which is useful for targeted learning (c.f., Section
-\ref{sec:targeted-learning}). Specifically we consider an estimator of
-a target parameter which is obtained by substituting the state learner
-estimates of the nuisance parameters $\Lambda_1$, $\Lambda_2$, and
-$\Gamma$. An example is an estimator of the cumulative incidence
-curve, which can be obtained from estimators of $\Lambda_1$ and
-$\Lambda_2$. Another example is provided in Section \ref{sec:targeted-learning}.
-By equations~(\ref{eq:lambdaj}) and~(\ref{eq:gamma}) and the definition of
-\( F \), we have
-\begin{equation}
-  \label{eq:7}
-  \Gamma(t , x) 
-  = \int_0^t  \frac{F(\diff s, -1, x )}{F(s-, 0, x )},
-  \quad \text{and} \quad
-  \Lambda_j(t , x) 
-  = \int_0^t  \frac{F(\diff s, j, x )}{F(s-, 0, x )},
-  \quad j \in \{1,2\},
-\end{equation}
-and thus an estimator based on \( \hat{\Lambda}_{1,n} \),
-\( \hat{\Lambda}_{2,n} \), and \( \hat{\Gamma}_{n} \) can also be obtained from
-an estimator of $F$ using equation~(\ref{eq:7}). A so-called targeted estimator
-has the key feature that it is asymptotically equivalent to a sum of i.i.d.\
-random variables plus a second order remainder term
-\citep{van2011targeted,hines2022demystifying}. For the setting with competing
-risks, the remainder term is dominated by terms of the form
-\begin{equation}
-  \label{eq:dr-term}
-  P{\left[
-      \int_0^{\tau} w_n(s, \blank)
-      \hat{M}_{1,n}(s \mid  \blank)
-      \hat{M}_{2,n}(\diff s \mid  \blank)
-    \right]},
-\end{equation}
-where \( (\hat{M}_{1,n}, \hat{M}_{2,n}) \) is any of the nine combinations of
-\( \hat{M}_{1,n} \in \{[\Gamma -\hat{\Gamma}_n], [\Lambda_1
--\hat{\Lambda}_{1,n}], [\Lambda_2 -\hat{\Lambda}_{2,n}]\} \) and
-\( \hat{M}_{2,n} \in \{[\Gamma -\hat{\Gamma}_n], [\Lambda_1
--\hat{\Lambda}_{1,n}], [\Lambda_2 -\hat{\Lambda}_{2,n}]\} \), and \( w_n \) is
-some data-dependent function with domain \([0,\tau]\times\mathcal X \)
-\citep{van2003unified}. In particular, a targeted estimator will be
-asymptotically linear if the `products' of the estimation errors
-\( \hat{M}_{1,n} \) and \( \hat{M}_{2,n} \) in equation~(\ref{eq:dr-term}) are
-\( \smallO_P{(n^{-1/2})}\). Proposition~\ref{prop:dr-structure} states that if
-equation~(\ref{eq:dr-term}) holds for a targeted estimator based on estimators
-$\hat{\Lambda}_{1,n}$, $\hat{\Lambda}_{2,n}$, and $\hat{\Gamma}_{n}$, then a
-similar product structure holds for a targeted estimator based on
-\( \hat{F}_n \). We state the result for the special case that
-\(\hat{M}_{1,n}= \Gamma-\hat{\Gamma}_n \) and
-\(\hat{M}_{2,n} =\Lambda_1-\hat{\Lambda}_{1,n} \), but similar results hold for
-any combinations of \( \Gamma-\hat{\Gamma}_n\),
-\( \Lambda_1-\hat{\Lambda}_{1,n} \), and \( \Lambda_2-\hat{\Lambda}_{2,n} \).
-\begin{proposition}
-  \label{prop:dr-structure}
-  Assume that \( w(s,x)\leq c \), \( F(s, 0, x) \geq 1/c \) and
-  \( \hat{F}_n(s, 0, x) \geq 1/c \) for some \( c>0 \) for all
-  \( s \in [0, \tau] \) and \( x \in \mathcal{X} \). Then there are real-valued
-  uniformly bounded functions \( w^a_n \), \( w^b_n \), \( w^c_n \), and
-  \( w^d_n \) with domain \( [0,\tau]^2 \times \mathcal{X} \) such that
-  \begin{align*}
-    & P{\left[
-      \int_0^{\tau} w(s, \blank)
-      \left\{
-      \Gamma(s,\blank) -\hat{\Gamma}_n(s,\blank)
-      \right\}
-      [\Lambda_1-\hat{\Lambda}_{1,n}]
-      (\diff s, \blank)
-      \right]}
-    \\
-     & =
-      P{\big[
-      \int_0^{\tau} \int_0^{s} w^a_n(s,u,\blank) [F - \hat{F}_n](u-, 0,
-       \blank)[F - \hat{F}_n](s-, 0, \blank)}
-    \\
-    & \qquad \quad \times F(\diff u, -1, \blank ) F ( \diff s, 1, \blank)
-      \big]
-    \\
-    & \quad +
-      P{\left[
-      \int_0^{\tau} \int_0^{s} w^b_n(s,u,\blank) [F - \hat{F}_n](u-, 0, \blank)
-      F(\diff u, -1, \blank ) [F - \hat{F}_n](\diff s, 1, \blank)
-      \right]}
-    \\
-    & \quad +
-      P{\left[
-      \int_0^{\tau} \int_0^{s} w^c_n(s,u,\blank) [F - \hat{F}_n](\diff u, -1, \blank)
-      [F - \hat{F}_n](s-, 0, \blank)
-      F(\diff s, 1, \blank ) 
-      \right]}
-    \\
-    & \quad +
-      P{\left[
-      \int_0^{\tau} \int_0^{s} w^d_n(s,u,\blank) [F - \hat{F}_n](\diff u, -1, \blank)
-      [F - \hat{F}_n](\diff s, 1, \blank)
-      \right]}.
-  \end{align*}
-\end{proposition}
-\begin{proof}
-  See Appendix~\ref{sec:state-learner-with}.
-\end{proof}
-
-\section{Targeted learning}
-\label{sec:targeted-learning}
-
-In this section, we consider a suitably smooth operator
-\( \theta \colon \mathcal{Q} \rightarrow \Theta \) which represents a
-target parameter of interest. The parameter space $\Theta$ can be a
-subset of \(\R^d\) or a subset of a function space, for example a
-subset of \(\mathcal{M}_{\tau}\) as in Section
-\ref{sec:super-learning}. In subsection~\ref{sec:cause-spec-aver} we
-discuss an example from causal inference where $\theta$ is the average
-treatment effect and \( \Theta = [-1,1] \).  To discuss the role of
-the state learner for targeted learning we briefly review some results
-from semiparametric efficiency theory. Extensive reviews and
-introductions are available elsewhere
-\cite[e.g.,][]{pfanzagl1985contributions,bickel1993efficient,van2003unified,tsiatis2007semiparametric,kennedy2016semiparametric}.
-Under the assumption of conditional independent censoring and
-positivity, $\theta$ is identifiable from \( \mathcal{P} \) which
-means that there exists an operator
-\( \Psi \colon \mathcal{P} \rightarrow \Theta \) such that
-\( \theta(Q) = \Psi(P_{Q, \Gamma}) \) for all
-$\Gamma \in \mathcal{M}_{\tau}$.  By equation~(\ref{eq:parametrizeP})
-this implies that we may write
-\begin{equation*}
-  \theta(Q) = \Psi(P) = \tilde{\Psi}^0(\Lambda_1, \Lambda_2, H),
-\end{equation*}
-for some operator \( \tilde{\Psi}^0 \). The state learner provides a ranking of
-all tuples
-\( (a_1, a_2, b) \in \mathcal{A}_1 \times \mathcal{A}_2 \times \mathcal{B} \).
-We use \( \hat{a}_{1,n} \), \( \hat{a}_{2,n} \), and \( \hat{b}_n \) to denote
-the learners corresponding to the discrete state learner \( \hat{\phi}_n \),
-i.e., the tuple with the highest rank. Letting \( H_n(\data_n) \) denote the
-empirical measure of \( \{X_1, \dots, X_n\} \), we obtain a plug-in estimator of
-$\theta$:
-\begin{equation}
-  \label{eq:2}
-  \hat{\Psi}^0(\data_n) =
-  \tilde{\Psi}^0(\hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n), H_n(\data_n)). 
-\end{equation}
-
-The asymptotic distribution of \( \hat{\Psi}^0 \) is
-difficult to analyse due to the cross-validated model selection step
-involved in the estimation of the nuisance parameters $\Lambda_1$ and
-$\Lambda_2$.
-% In addition, the estimator will typically have an
-% asymptotic bias that vanishes at a too slow 
-% rate.
-Using tools from semi-parametric efficiency theory, it is possible to
-construct a so-called targeted or debiased estimator with an
-asymptotic distribution which we know how to estimate
-\citep{bickel1993efficient,van2011targeted,chernozhukov2018double}. A
-targeted estimator is based on the efficient influence function for
-the parameter $\tilde{\Psi}^0$ and relies on estimators of the
-nuisance parameters $\Gamma$, \( \Lambda_1 \) and $\Lambda_2$. The
-efficient influence function is a \( P \)-zero mean and square
-integrable function which we denote by
-\( \psi(\blank ; \Lambda_1, \Lambda_2, \Gamma) \). The name is
-justified because any regular asymptotically linear estimator that has
-\( \psi \) as its influence function is asymptotically efficient,
-meaning that it has smallest asymptotic variance among all regular
-asymptotically linear estimators \citep{bickel1993efficient}.
-
-An example of a targeted estimator is the one-step estimator, defined as
-\begin{equation}
-    \label{eq:one-step-def}
-    \begin{split}
-      \hat{\Psi}_{\text{OS}}(\data_n)
-      = &
-          \tilde{\Psi}^0(\hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
-          H_n(\data_n))
-      \\
-        & + \empmeas{[\psi(\blank; \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
-          \hat{b}_n(\data_n) )]},
-    \end{split}
-\end{equation}
-where \( \empmeas \) is the empirical measure of a data set
-\(\{O_i\}_{i=1}^n\).
-Under suitable regularity conditions
-we have the
-following asymptotic expansion of the one-step estimator
-\citep{pfanzagl1985contributions,van2003unified,fisher2021visually,kennedy2022semiparametric},
-\begin{equation*}
-  \hat{\Psi}_{\text{OS}}(\data_n)- \Psi(P)
-  =  \empmeas{[\psi(\blank ; \Lambda_1, \Lambda_2, \Gamma)]}
-  +\mathrm{Rem}{(\hat{\Lambda}_{1,n},\hat{\Lambda}_{2,n},  \hat{\Gamma}_n, P)} + \smallO_{P}(n^{-1/2}),
-\end{equation*}
-where the remainder term has the form
-\begin{equation}
-  \label{eq:4}
-  \mathrm{Rem}{(\hat{\Lambda}_{1,n},\hat{\Lambda}_{2,n},  \hat{\Gamma}_n, P)}
-  = \bigO_P{
-    \left\{
-      \|\Lambda_1-\hat{\Lambda}_{1,n}\|^2
-      +
-      \|\Lambda_2-\hat{\Lambda}_{2,n}\|^2
-      +
-      \|\Gamma-\hat{\Gamma}_{n}\|^2
-    \right\}
-  },
-\end{equation}
-for some suitable norm \( \|\blank \| \), for instance the
-\( \mathcal{L}_{P}^2 \)-norm. When equation~(\ref{eq:4}) holds and the
-nuisance parameters $\Lambda_1$, $\Lambda_2$, and $\Gamma$ are
-consistently estimated at rate \( \smallO_P{(n^{-1/4})} \), then
-\begin{equation}
-  \label{eq:3}
-  \sqrt{n}(\hat{\Psi}_{\text{OS}}(\data_n)- \Psi(P)) \weakly \mathcal{N}(0,
-  P{[\psi(\blank; \Lambda_1, \Lambda_2, \Gamma)^2]}),
-\end{equation}
-where we use \( \weakly \) to denote weak convergence \citep{van2000asymptotic}.
-In particular, equation~(\ref{eq:3}) and Slutsky's lemma imply that we can
-obtain asymptotically valid \((1-\alpha)\cdot100\%\) confidence intervals by
-calculating
-\begin{equation*}
-  \left[
-    \hat{\Psi}_{\text{OS}}(\data_n) - q_{\alpha/2} \hat{\sigma}(\data_n) ,
-    \;
-    \hat{\Psi}_{\text{OS}}(\data_n) + q_{\alpha/2} \hat{\sigma}(\data_n)
-  \right],
-\end{equation*}
-where \( q_{\alpha} \) is the \( (1-\alpha) \)-quantile of the standard normal
-distribution and
-\begin{equation*}
-  \hat{\sigma}(\data_n)^2 = \empmeas{ \left[ \psi(\blank;
-      \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n), \hat{b}_n(\data_n))^2
-    \right]}.
-\end{equation*}
-
-\subsection{Average treatment effect on the absolute risk of an event}
-\label{sec:cause-spec-aver}
-
-In the following we detail how the state learner can be used to
-construct a targeted estimator of the cause-specific average treatment
-effect \citep{rytgaard2022targeted}. We assume that the covariate
-vector \( X \in \R^p \) contains a binary treatment indicator \( A \)
-and a vector of potential confounders,
-\( W \in \mathcal{W}\subset \R^{p-1} \).  We use $\mu$ to denote the
-marginal distribution of \( W \) and $\pi$ to denote the conditional
-probability of treatment,
-\begin{equation*}
-  \pi(w) = P(A=1 \mid W=w).
-\end{equation*}
-We assume throughout that $\pi$ is uniformly bounded away from \( 0 \)
-and \( 1 \) on \( \mathcal{W} \) and that both \( A \) and \( W \) are
-fully observed for all individuals. We use a super learner to
-estimate $\pi$ \citep{Polley_Ledell_Kennedy_Laan_2023_Superlearn}, and
-we denote this estimator by $\hat{\pi}_n$. We use the empirical
-measure of \( \{W_1, \dots, W_n\} \) to estimate $\mu$, and denote
-this estimator by $\hat{\mu}_n$. As parameter of interest we consider
-the standardised difference in the absolute risk of an event with
-cause 1 at time $\tau$:
-\begin{align*}
-  & \theta_{{\tau}}(Q)
-  \\
-  & = \int_{\mathcal{W}} 
-    \left\{
-    \int_0^{\tau}
-    S(s- \mid w, 1)  \Lambda_1(\diff s \mid w, 1)
-    -
-    \int_0^{\tau}
-    S(s- \mid w, 0)  \Lambda_1(\diff s \mid w, 0)
-    \right\}
-    \mu(\diff w).
-\end{align*}
-Under the usual assumptions for causal inference (consistency,
-positivity, no unmeasured confounding) \( \theta_{{\tau}} \) can be
-given the causal interpretation
-\begin{equation*}
-  \theta_{{\tau}}(Q) =
-  P{(T^{1} \leq \tau, D^{1}=1)}-
-  P{(T^{0} \leq \tau, D^{0}=1)},
-\end{equation*}
-where \( (T^a, D^a) \), \( a \in \{0,1\} \), denote potential outcomes
-\citep{hernanRobinsWhatIf}. In this case, the interpretation of $\theta_{\tau}$
-is the difference in the average risk of cause \( 1 \) occurring before time
-\( \tau \) in the population if everyone had been given treatment (\( A=1 \))
-compared to if no one had been given treatment \( (A=0) \).
-
-Using equation~(\ref{eq:surv-def}) we may write
-\( \theta_{{\tau}}(Q) = \tilde{\Psi}_{\tau}^0(\Lambda_1, \Lambda_2, \mu) \), where
-\begin{equation}
-  \label{eq:1}    
-  \begin{split}
-    & \tilde{\Psi}_{\tau}^0(\Lambda_1, \Lambda_2, \mu)
-    \\
-    & =
-      \int_{\mathcal{W}} 
-      \int_0^{\tau}
-      \exp{\{{-\Lambda_1(s- \mid w, 1)-\Lambda_2(s- \mid w, 1)}\}}  \Lambda_1(\diff s \mid
-      w, 1)
-      \mu(\diff w)
-    \\
-    &  \quad
-      -\int_{\mathcal{W}} 
-      \int_0^{\tau}
-      \exp{\{{-\Lambda_1(s- \mid w, 0)-\Lambda_2(s- \mid w, 0)}\}}  \Lambda_1(\diff s \mid w, 0)
-      \mu(\diff w).
-  \end{split}
-\end{equation}
-The efficient influence function for the parameter $\tilde{\Psi}_{\tau}$ depends
-on the set \( (\Lambda_1, \Lambda_2, \Gamma, \pi) \) of nuisance parameters.
-We define
-\begin{align*}
-  \omega_a(A,W; \pi)
-  &=  \frac{% (-1)^{a+1}
-    \1{\{A=a\}}}{\pi(W)^{a}(1-\pi(W))^{1-a}},
-  \\
-  g(t, A, W; \Lambda_1, \Lambda_2)
-  & = \int_0^{t}
-    \exp{\{ {-\Lambda_1(s- \mid W, A)-\Lambda_2(s- \mid W, A)} \}}  \Lambda_1(\diff s \mid
-    W, A),
-  \\  
-  M_j(\diff t \mid A, W;  \Lambda_j  )
-  & = N_j(\diff t) -
-    \1{\{\tilde{T} \geq t\}} \Lambda_j(\diff t \mid W, A),
-    \quad j \in \{1,2\},
-    \intertext{and}
-    M(\diff t \mid A, W;  \Lambda_1, \Lambda_2  )
-  & = M_1(\diff t \mid A, W;  \Lambda_1  ) +
-    M_2(\diff t \mid A, W;  \Lambda_2  ).
-\end{align*}
-The efficient influence function can now be written as
-\citep{van2003unified,jewell2007non,rytgaard2022targeted},
-\begin{equation*}
-  \psi_{\tau}(O; \Lambda_1, \Lambda_2, \Gamma, \pi)
-  = \psi_{\tau}^1(O; \Lambda_1, \Lambda_2, \Gamma, \pi)
-  - \psi_{\tau}^0(O; \Lambda_1, \Lambda_2, \Gamma, \pi)
-  -\tilde{\Psi}_{\tau}^0(\Lambda_1, \Lambda_2, \mu),
-\end{equation*}
-where
-\begin{equation}
-  \label{eq:5}
-  \begin{split}
-    & \psi_{\tau}^a(O; \Lambda_1, \Lambda_2, \Gamma, \pi)
-    \\
-    & =
-      \omega_a(A,W; \pi)
-      \int_0^{\tau} \exp{\{{\Gamma(t- \mid A, W)}\}}   
-      M_1(\diff t \mid A, W; \Lambda_1)
-    \\
-    & \quad
-      -
-      \omega_a(A,W; \pi)
-      g(\tau, A, W; \Lambda_1, \Lambda_2)
-    \\
-    & \quad \quad \times
-      \int_0^{\tau}
-      \exp{\{{[\Gamma+\Lambda_1 + \Lambda_2](t- \mid A, W)}\}}
-      M(\diff t \mid A, W; \Lambda_1, \Lambda_2)
-    \\
-    & \quad
-      +
-      \omega_a(A,W; \pi)      
-      \int_0^{\tau}
-      g(t, A, W; \Lambda_1, \Lambda_2)
-    \\
-    & \quad \quad \times
-      \exp{\{{[\Gamma+\Lambda_1 + \Lambda_2](t- \mid A, W)}\}}
-      M(\diff t \mid A, W; \Lambda_1, \Lambda_2)
-    \\
-    & \quad + g(\tau, a, W; \Lambda_1, \Lambda_2).
-  \end{split}
-\end{equation}
-Equations~(\ref{eq:1}) and~(\ref{eq:5}) allow us to construct a one-step
-estimator by using the definition given in equation~(\ref{eq:one-step-def}),
-which gives the estimator
-\begin{equation}
-  \label{eq:one-step-comp-ate}
-  \begin{split}
-    \hat{\Psi}_{\tau,\text{OS}}(\data_n)
-    = &
-        \tilde{\Psi}_{\tau}^0(\hat{a}_1(\data_n), \hat{a}_2(\data_n),
-        \hat{\mu}_n(\data_n))
-    \\
-      &
-        +
-        \empmeas{[\psi_{\tau}(\blank; \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
-        \hat{b}_n(\data_n), \hat{\pi}_n(\data_n))]}
-    \\
-    = &
-        \empmeas{[\psi_{\tau}^1(\blank; \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
-        \hat{b}_n(\data_n), \hat{\pi}_n(\data_n))]}
-    \\
-      &
-        - \empmeas{[\psi_{\tau}^0(\blank; \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
-        \hat{b}_n(\data_n), \hat{\pi}_n(\data_n))]}.
-  \end{split}
-\end{equation}
-
 \section{Numerical experiments}
@@
 recurrence, death without tumor recurrence or end of followup
-(censored) whatever came first. For the sole purpose of illustration,
-we estimate the average treatment effect of hormone therapy on death
-and tumor recurrence. To do this we adapt the estimation strategy of
-Section~\ref{sec:targeted-learning} as follows. We use the state
+(censored) whatever came first. ... We use the state
 learner to rank libraries of learners for the cause-specific
@@
 \( \Lambda_2 \), and the cumulative hazard function of the conditional
-censoring distribution $\Gamma$.  We then use the highest ranked
-combination of learners and apply
-formula~\eqref{eq:one-step-comp-ate}.
+censoring distribution $\Gamma$.
 
@@
   \label{fig:zelefski-real}
-\end{figure}
-
-We use the learners of the three cumulative hazard functions selected
-by the state learner and the estimator defined in
-Section~\ref{sec:cause-spec-aver},
-equation~(\ref{eq:one-step-comp-ate}), to estimate the average
-treatment effect of hormone therapy on risk of tumor recurrence and
-death. The propensity score is estimated with a lasso model that
-includes all levels of interaction.
-The results are shown in Figure~\ref{fig:zelefski-real-target} for 6
-month intervals after baseline with pointwise 95\% confidence
-intervals. We see that hormone therapy decreases the risk of tumor
-recurrence and increases the risk of death without tumor recurrence,
-but that none of the estimated effects are statistically significant.
-
-\begin{figure}
-  \centering%
-  \includegraphics[width=1\linewidth]{real-data-target.pdf}
-  \caption[]{Estimates of the average treatment effect of hormone therapy on the
-    risk of tumor recurrence and death obtained from the prostate cancer study
-    analysed by \cite{kattan2000pretreatment}. The estimates are based on the
-    estimator defined in equation~(\ref{eq:one-step-comp-ate}) with pointwise
-    95\% confidence intervals calculated as described in
-    Section~\ref{sec:targeted-learning}. The estimates of the nuisance
-    parameters are provided by the state learner.}
-  \label{fig:zelefski-real-target}
 \end{figure}
@@
 
-This section contains proofs of the results stated in the paper.
-Section~\ref{sec:consistency-proof} contains a proof of the consistency result
-from Section~\ref{sec:consistency}; Section~\ref{sec:oracle-inequalities}
-contains proofs of the oracle inequalities from
-Section~\ref{sec:finite-sample-oracle}; and
-Section~\ref{sec:state-learner-with} demonstrates transience of the second
-order remainder structure stated in Section~\ref{sec:trans-second-order}.
+This section contains proofs of the results stated in the
+paper. Section~\ref{sec:consistency-proof} contains a proof of
+the consistency result from Section~\ref{sec:consistency}, and
+Section~\ref{sec:oracle-inequalities} contains proofs of the
+oracle inequalities from
+Section~\ref{sec:finite-sample-oracle}.
 
@@
 
-\subsection{Transience of the second order remainder structure}
-\label{sec:state-learner-with}
-
-Recall that we let $\Lambda_1$ denote the conditional cumulative hazard function
-for one of the event times of interest and $\Gamma$ the conditional cumulative
-hazard function for censoring, c.f., Section~\ref{sec:framework}.
-
-\begin{proof}[Proof of Proposition~\ref{prop:dr-structure}]
-  For notational convenience we suppress \( X \) in the following. The final
-  result can be obtained by adding the argument \( X \) to all functions and
-  averaging. We use the relations from equation~\eqref{eq:7} to write
-  \begingroup %
-  \allowdisplaybreaks
-    \begin{align*}
-      & \int_0^{\tau} w(s) 
-        \left\{
-        \Gamma(s) - \hat{\Gamma}_n(s)
-        \right\}
-        [\Lambda_1 - \hat{\Lambda}_{1,n}](\diff s)
-      \\
-      & =
-        \int_0^{\tau} w(s) 
-        \left\{
-        \int_0^s \frac{F(\diff u, -1)}{F(u-, 0)} -
-        \int_0^s \frac{\hat{F}_n(\diff u, -1)}{\hat{F}_n(u-, 0)}  -
-        \right\}
-        \left[
-        \frac{F(\diff s, 1)}{F(s-, 0)}
-        - \frac{\hat{F}_n(\diff s, 1)}{\hat{F}_n(s-, 0)}
-        \right]
-      \\
-      & =
-        \int_0^{\tau} w(s) 
-        \Bigg\{
-        \int_0^s 
-        \left(
-        \frac{1}{F(u-, 0)} -  \frac{1}{\hat{F}_n(u-, 0)}
-        \right) F(\diff u, -1)
-      \\
-      & \qquad\qquad \qquad
-        +
-        \int_0^s \frac{1}{\hat{F}_n(u-, 0)} 
-        \left[
-        F(\diff u, -1) - \hat{F}_n(\diff u, -1)
-        \right]
-        \Bigg\}
-      \\
-      & \qquad\qquad \times
-        \left[
-        \left(
-        \frac{1}{F(s-, 0)} -
-        \frac{1}{\hat{F}_n(s-, 0)}
-        \right)F(\diff s, 1)
-        + \frac{1}{\hat{F}_n(s-, 0)}
-        \left(
-        F(\diff s, 1) -
-        \hat{F}_n(\diff s, 1)
-        \right)
-        \right]
-      \\
-      &
-        = \int_0^{\tau} 
-        \int_0^s
-        w(s) 
-        \left(
-        \frac{1}{F(u-, 0)} -  \frac{1}{\hat{F}_n(u-, 0)}
-        \right) 
-        \left(
-        \frac{1}{F(s-, 0)} -
-        \frac{1}{\hat{F}_n(s-, 0)}
-        \right)F(\diff u, -1)F(\diff s, 1)
-      \\
-      & \quad +
-        \int_0^{\tau}
-        \int_0^s
-        w(s) 
-        \left(
-        \frac{1}{F(u-, 0)} -  \frac{1}{\hat{F}_n(u-, 0)}
-        \right) \frac{F(\diff u, -1) }{\hat{F}_n(u-,0)}
-        \left(
-        F(\diff s, 1) -
-        \hat{F}_n(\diff s, 1)
-        \right)
-      \\
-      & \quad +
-        \int_0^{\tau} 
-        \int_0^s      
-        \frac{w(s) }{\hat{F}_n(u-, 0)} 
-        \left[
-        F(\diff u, -1) - \hat{F}_n(\diff u, -1)
-        \right]
-        \left(
-        \frac{1}{F(s-, 0)} -
-        \frac{1}{\hat{F}_n(s-, 0)}
-        \right)F(\diff s, 1)
-      \\
-      & \quad +
-        \int_0^{\tau} 
-        \int_0^s      
-        \frac{w(s) }{\hat{F}_n(u-, 0)} 
-        \left[
-        F(\diff u, -1) - \hat{F}_n(\diff u, -1)
-        \right]
-        \frac{1}{\hat{F}_n(s-, 0)}
-        \left(
-        F(\diff s, 1) -
-        \hat{F}_n(\diff s, 1)
-        \right).
-    \end{align*}
-    \endgroup %
-    Consider the first term on the right hand side. By the mean value theorem,
-    \begin{equation*}
-      \frac{1}{F(t-, 0)}
-      - \frac{1}{\hat{F}_n(t-, 0)}
-      = \frac{-1}{\tilde{r}_n(t)^2}
-      \left[
-        F(t-, 0)
-        - \hat{F}_n(t-, 0)
-      \right],
-    \end{equation*}
-    where \( \tilde{r}_n(t) \) is some value between \( \hat{F}(t-, 0) \) and
-    \( \hat{F}_n(t-, 0) \). Letting \( w_n^*(t) = -\tilde{r}_n(t)^{-2} \), we
-    may write
-  \begin{align*}
-    & \int_0^{\tau} 
-      \int_0^s
-      w(s) 
-      \left(
-      \frac{1}{F(u-, 0)} -  \frac{1}{\hat{F}_n(u-, 0)}
-      \right)      
-      \left(
-      \frac{1}{F(s-, 0)} -
-      \frac{1}{\hat{F}_n(s-, 0)}
-      \right)F(\diff u, -1)F(\diff s, 1)
-    \\
-    & =
-      \int_0^{\tau} 
-      \int_0^s
-      w(s)
-      w_n^*(u) 
-      \left(
-      F(u-, 0) - \hat{F}_n(u-, 0)
-      \right)
-    \\
-    & \qquad \qquad \quad
-      \times
-      w_n^*(s) 
-      \left(
-      F(s-, 0) - \hat{F}_n(s-, 0)
-      \right)       
-      F(\diff u, -1)F(\diff s, 1)
-    \\
-    & =
-      \int_0^{\tau} 
-      \int_0^s
-      w_n^a(s,u)
-      \left(
-      F(u-, 0) - \hat{F}_n(u-, 0)
-      \right)
-      \left(
-      F(s-, 0) - \hat{F}_n(s-, 0)
-      \right)       
-      F(\diff u, -1)F(\diff s, 1),
-  \end{align*}
-  where we have defined \( w_n^a(s,u) = w(s)w^*_n(s)w^*_n(u) \). By the
-  assumption that \( F(\blank,0) \) and \( F_n(\blank,0) \) are uniformly
-  bounded away from zero on \( [0,\tau] \), it follows that \( w_n^* \) is
-  uniformly bounded, and hence \( w_n^a(s,u) \) is also uniformly bounded,
-  because \( w \) was assumed uniformly bounded. The same approach can be
-  applied to the three remaining terms which gives the result.
-\end{proof}
-
 \end{document}

### 2025-05-21 manuscript/manuscript-two.tex
divide into two manuscripts and basic template

--- 2025-01-13 manuscript/manuscript-lda.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
-%
-% This is a general template file for the LaTeX package SVJour3
-% for Springer journals.          Springer Heidelberg 2010/09/16
-%
-% Copy it to a new file with a new name and use it as the basis
-% for your article. Delete % signs as needed.
-%
-% This template includes a few options for different layouts and
-% content for various journals. Please consult a previous issue of
-% your journal as needed.
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-% First comes an example EPS file -- just ignore it and
-% proceed on the \documentclass line
-% your LaTeX will extract the file if required
-\begin{filecontents*}{example.eps}
-%!PS-Adobe-3.0 EPSF-3.0
-%%BoundingBox: 19 19 221 221
-%%CreationDate: Mon Sep 29 1997
-%%Creator: programmed by hand (JK)
-%%EndComments
-gsave
-newpath
-  20 20 moveto
-  20 220 lineto
-  220 220 lineto
-  220 20 lineto
-closepath
-2 setlinewidth
-gsave
-  .4 setgray fill
-grestore
-stroke
-grestore
-\end{filecontents*}
-%
-\RequirePackage{fix-cm}
-%
-%\documentclass{svjour3}                     % onecolumn (standard format)
-%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
-\documentclass[smallextended]{svjour3}       % onecolumn (second format)
-%\documentclass[twocolumn]{svjour3}          % twocolumn
-%
-\smartqed  % flush right qed marks, e.g. at end of proof
-%
-\usepackage{graphicx}
-%
-% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
-%
-% insert here the call for the packages your document requires
-\usepackage{dsfont, amssymb,soul,xcolor, enumitem, amsmath,verbatim}
-\usepackage[authoryear]{natbib}
-\definecolor{linkcolor}{rgb}{0, 0, 0.54}
-\usepackage[colorlinks,allcolors=linkcolor]{hyperref}
-%
-% please place your own definitions here and don't use \def but
-\DeclareMathOperator{\E}{\mathbb{E}} % expectation
-\newcommand{\Z}{\mathbb{Z}}
-\newcommand{\R}{\mathbb{R}}
-\newcommand{\N}{\mathbb{N}}
-\newcommand{\C}{\mathbb{C}}
-\renewcommand{\S}{\mathbb{S}}
-\newcommand{\blank}{\makebox[1ex]{\textbf{$\cdot$}}}
-\newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
-\def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
-\renewcommand{\phi}{\varphi}
-\renewcommand{\epsilon}{\varepsilon}
-\newcommand*\diff{\mathop{}\!\mathrm{d}}
-\newcommand{\weakly}{\rightsquigarrow}
-\newcommand\smallO{\textit{o}}
-\newcommand\bigO{\textit{O}}
-\newcommand{\midd}{\; \middle|\;}
-\newcommand{\1}{\mathds{1}}
-\usepackage{ifthen} %% Empirical process with default argument
-\newcommand{\G}[2][n]{
-{\ensuremath{\mathbb{G}_{#1}}{\left[#2\right]}}
-}
-\DeclareMathOperator*{\argmin}{\arg\!\min}
-\DeclareMathOperator*{\argmax}{\arg\!\max}
-\newcommand{\empmeas}{\ensuremath{\mathbb{P}_n}} % empirical measure
-\newcommand{\data}{\ensuremath{\mathcal{D}}}
-%
-% Insert the name of "your journal" with
-\journalname{Lifetime Data Analysis}
-%
-\begin{document}
-
-\title{The state learner -- a super learner for right-censored data
-  % \thanks{Grants or other notes about the article that should go on the front
-  % page should be placed here. General acknowledgments should be placed at the
-  % end of the article.}
-}
-% \subtitle{Do you have a subtitle?\\ If so, write it here}
-
-\titlerunning{The state learner}        % if too long for running head
-
-\author{Anders Munch         \and
-        Thomas~A.~Gerds %etc.
-}
-
-%\authorrunning{Short form of author list} % if too long for running head
-
-\institute{Anders Munch \at
-              Section of Biostatistics, University of Copenhagen, Denmark \\
-              % Tel.: +123-45-678910\\
-              % Fax: +123-45-678910\\
-              \email{a.munch@sund.ku.dk}           %  \\
-%             \emph{Present address:} of F. Author  %  if needed
-              \and %
-              Thomas~A.~Gerds %
-              \at Section of Biostatistics, University of Copenhagen, Denmark}
-
-\date{Received: date / Accepted: date}
-% The correct dates will be entered by the editor
-
-\maketitle
-
-\begin{abstract}
-  In survival analysis, prediction models are needed as stand-alone tools and in
-  applications of causal inference to estimate nuisance parameters. The super
-  learner is a machine learning algorithm which combines a library of prediction
-  models into a meta learner based on cross-validated loss. In right-censored
-  data, the choice of the loss function and the estimation of the expected loss
-  need careful consideration. We introduce the state learner, a new super
-  learner for survival analysis, which simultaneously evaluates libraries of
-  prediction models for the event of interest and the censoring distribution.
-  The state learner can be applied to all types of survival models, works in the
-  presence of competing risks, and does not require a single pre-specified
-  estimator of the conditional censoring distribution. We establish an oracle
-  inequality for the state learner and investigate its performance through
-  numerical experiments. We illustrate the application of the state learner with
-  prostate cancer data, as a stand-alone prediction tool, and, for causal
-  inference, as a way to estimate the nuisance parameter models of a smooth
-  statistical functional.
-
-  \keywords{Competing risks, cross-validation, loss based estimation,
-    right-censored data, super learner}
-% \PACS{PACS code1 \and PACS code2 \and more}
-% \subclass{MSC code1 \and MSC code2 \and more}
-\end{abstract}
-
-\section{Introduction}
-\label{sec:introduction}
-
-% Medical decision making is traditionally based on the discrepancy
-% between intuitive judgment and the law of chance
-% \citep{redelmeier1995probability}.
-% We are concerned with the statistical models that provide
-% probabilistic predictions about the medical future of a patient. We
-% have two different applications of these models in mind. One is as a
-% stand-alone tool which predicts the likelihood of medical outcomes.
-% The other is as nuisance parameter estimate
-% input to an algorithm which estimates a causal target
-% parameter such as an average treatment effect and depends on estimates
-% of nuisance parameters such as the conditional probability of being
-% uncensored.  An example of the latter is the targeted minimum loss
-% based estimator \citep{rytgaard2022targeted}. In both applications,
-% the models have to be learned based on available data from previous
-% patients.
-A super learner is a machine learning algorithm that combines a finite
-set of learners into a meta learner by estimating prediction
-performance in hold-out samples using a pre-specified loss function
-\citep{van2007super}. When the aim is to make a prediction model,
-super learners combine strong learners, such as Cox regression models
-and random survival forests \citep[][Section 8.4]{gerds2021medical}.
-% For targeted learning of a low dimensional
-% target parameter in presence of a high-dimensional nuisance parameter,
-% a super learner will usually also include the highly adaptive lasso
-% \citep{van2011targeted, benkeser2016highly,van2017generally}.
-While the general idea of combining strong learners based on cross-validation
-stems from earlier work \citep{wolpert1992stacked,breiman1996stacked}, the
-name super learner is justified by an oracle inequality
-\citep{van2003unicv,vaart2006oracle}.
-% The super learner is guaranteed to perform
-% almost as well as the model which minimises the expected performance, i.e., the
-% model we would select if we could evaluate the prediction performance in an
-% infinite hold-out sample.
-
-We define the state learner, a new super learner for right-censored data, which
-simultaneously estimates the expected loss of learners of the event time
-distribution and the censoring distribution. Compared to the super learners of
-\cite{polley2011-sl-cens} and \cite{golmakani2020super}, the state learner
-proposed here poses no restrictions on the type of learners that can be included
-in the library. Compared to super learners that rely on inverse probability of
-censoring weighting, censoring unbiased transformations, or pseudo-values, the
-state learner does not require a pre-specified censoring distribution. Compared
-to the super learners by \cite{han2021inverse} and \cite{westling2021inference},
-the state learner has theoretical guarantees which these methods do not have.
-% Moreover, state learner seems to be the first super learner for
-% right censored data which learns the outcome and the censoring
-% distribution simulaneously.
-
-The loss function underlying the state learner operates on coarsened data. In
-the right-censored survival setting the coarsened data consist of the minimum
-and the order of the censoring time and the event time as well as the baseline
-covariates. The state learner works by simultaneously assessing how well
-learners of the event time distribution and the censoring distribution predict
-the states of the coarsened data. To analyse the theoretical properties of the
-state learner we focus on the discrete super learner which combines the library
-of learners by picking the one that minimises the cross-validated loss
-\citep{van2007super}. In the presence of competing risks, our algorithm uses
-separate libraries of learners for the cumulative hazard functions for each of
-the competing risks and for the censoring distribution. We show that the oracle
-selector of the state learner is consistent if all libraries contain a
-consistent learner and prove a finite sample oracle inequality.
-
-The state learner algorithm can output a medical risk prediction model
-\citep{gerds2021medical} which predicts the probability of an event based on
-covariates in the presence of competing risks. Another important application,
-which motivated our contribution, is in targeted learning where conditional
-event probabilities occur as high-dimensional nuisance parameters which need to
-be estimated at a certain rate \citep{van2011targeted, rytgaard2021estimation,
-  rytgaard2022targeted}. The state learner is particularly well-suited for this
-application as it simultaneously provides estimates for both the outcome and the
-censoring, both of which are needed for targeted learning. For the asymptotic
-bias term of a targeted estimator, which uses the state learner to estimate
-nuisance parameters, we show that a second order product structure holds. We
-illustrate both applications of the state learner with prostate cancer data
-\citep{kattan2000pretreatment}.
-
-There are many existing super learners for right-censored data. We briefly
-review them here. Machine learning based on right-censored data commonly uses
-the partial log-likelihood as a loss function
-\citep[e.g.,][]{li2016regularized,yao2017deep,lee2018deephit,katzman2018deepsurv,gensheimer2019scalable,lee2021boosted,kvamme2021continuous}.
-This loss function is also suggested for super learning with right-censored data
-by \cite{polley2011-sl-cens}, where data is assumed to be observed in discrete
-time. However, the partial log-likelihood loss does not work well with data
-splitting (cross-validation) in continuous time. The reason is that the partial
-log-likelihood loss assigns an infinite value for any learner that predicts
-piece-wise constant cumulative hazard functions when there are event times in
-the test set which do not occur in the learning set. This problem occurs with
-prominent survival learners including the Kaplan-Meier estimator, the random
-survival forest, and the semi-parametric Cox regression model and these learners
-cannot be included in the library of the super learner proposed by
-\cite{polley2011-sl-cens}. When a proportional hazards model is assumed, the
-baseline hazard function can be profiled out of the likelihood
-\citep{cox1972regression}. The cross-validated partial log-likelihood loss
-\citep{verweij1993cross} has therefore been suggested as a loss function for
-super learning by \cite{golmakani2020super}. This choice of loss function
-restricts the library of learners to include only Cox proportional hazards
-models, and hence excludes many learners such as, e.g., random survival forests,
-additive hazards models, and accelerated failure time models. Other proposals
-for super learning with right-censored data use an inverse probability of
-censoring weighted (IPCW) loss function
-\citep{graf1999assessment,van2003unicv,molinaro2004tree,keles2004asymptotically,hothorn2006survival,gerds2006consistent,gonzalez2021stacked},
-censoring unbiased transformations
-\citep{fan1996local,steingrimsson2019censoring}, or pseudo-values
-\citep{andersen2003generalised,mogensen2013random,sachs2019ensemble}. All these
-methods rely on an estimator of the censoring distribution, and their drawback
-is that this estimator has to be pre-specified. An approach which avoids a
-pre-specified censoring model was proposed independently by
-\cite{han2021inverse} and \cite{westling2021inference}. In both articles, the
-authors suggest to iterate between learning of the outcome model and learning of
-the censoring model using IPCW loss functions. However, it seems that this
-procedure is in general not guaranteed to converge to the true data-generating
-mechanism \citep[][Appendix~A.4]{munch2024thesis}.
-
-We introduce our notation and framework in Section~\ref{sec:framework}. In
-Section~\ref{sec:super-learning} we define general super learning for
-right-censored data. Section~\ref{sec:super-learner-simple} introduces the state
-learner, and Section~\ref{sec:theor-results-prop} provides theoretical
-guarantees. In Section~\ref{sec:targeted-learning} we discuss the use of the
-state learner in the context of targeted learning. We report results of our
-numerical experiments in Section~\ref{sec:numer-exper} and analyse a prostate
-cancer data set in Section~\ref{sec:real-data-appl}.
-Section~\ref{sec:discussion} contains a discussion of the merits and limitations
-of our proposal. Proofs are given in Appendix~\ref{sec:proofs}. An
-implementation of the state learner is available at
-\url{https://github.com/amnudn/statelearner} along with code for reproducing our
-numerical experiments.
-
-\section{Notation and framework}
-\label{sec:framework}
-
-In a competing risk framework \citep{andersen2012statistical}, let
-\( T\) be a time to event variable, \(D\in\{1,2\}\) the cause of the
-event, and $X \in \mathcal{X}$ a vector of baseline covariates taking
-values in a bounded subset \( \mathcal{X} \subset \R^p \),
-\( p\in\N \). Let $\tau< \infty$ be the prediction horizon. We use
-\( \mathcal{Q} \) to denote the collection of all probability measures
-on \( [0,\tau] \times \{1,2\}\times \mathcal{X} \) such that
-\( (T, D, X) \sim Q \) for some unknown \( Q \in \mathcal{Q} \). For
-\(j\in\{1,2\}\), the cause-specific conditional cumulative hazard
-functions are defined by
-\( \Lambda_{j} \colon [0, \tau] \times \mathcal{X} \rightarrow \R_+ \)
-such that
-\begin{equation*}
-  % \label{eq:cum-haz}
-  \Lambda_{j}(t \mid x) = \int_0^t\frac{  Q(T \in \diff s, D=j \mid X=x )}{Q(T \geq s \mid X=x )}.
-\end{equation*}
-For ease of presentation we assume throughout that the map
-\( t\mapsto \Lambda_j(t \mid x) \) is continuous for all \( x \) and
-\( j \). This is not a limitation: All arguments carry over directly
-to the general case. We denote by \(S\) the conditional event-free
-survival function:
-\begin{equation}
-  \label{eq:surv-def}
-  S(t \mid x)=\exp\left\{-\Lambda_{1}(t \mid x)-\Lambda_{2}(t \mid x)\right\}.
-\end{equation}
-Let \( \mathcal{M}_{\tau}\) denote the space of all conditional cumulative hazard
-functions on \( [0,\tau] \times\mathcal{X}\). Any distribution
-\( Q \in \mathcal{Q} \) can be characterised by
-\begin{equation*}
-  \label{eq:parametrizeQ}
-  \begin{split}
-    Q(\diff t,j,\diff x)=& \left\{S(t- \mid x)\Lambda_1(\diff t \mid x)H(\diff x)\right\}^{\1{\{j=1\}}}\\
-                         &  \left\{S(t- \mid x)\Lambda_2(\diff t \mid x)H(\diff x)\right\}^{\1{\{j=2\}}},
-  \end{split}
-\end{equation*}
-where \(\Lambda_{j} \in \mathcal{M}_{\tau}\) for \(j=1,2\) and \(H\) is the marginal
-distribution of the covariates.
-
-We consider the right-censored setting in which we observe the coarsened data
-\(O = (\tilde{T},\tilde D, X)\), where $\tilde T = \min(T,C)$ for a
-right-censoring time \(C\), $\Delta = \1{\{T \leq C\}}$, and
-\(\tilde D=\Delta D\). Let \(\mathcal{P}\) denote a set of probability
-measures on the sample space
-\(\mathcal{O} = [0, \tau] \times \{0, 1, 2\} \times \mathcal{X}\) such
-that \(O \sim P \) for some unknown \(P\in \mathcal{P}\). We assume
-that the event times and the censoring times are conditionally
-independent given covariates, \( T \independent C \mid X \). This
-implies that any distribution \( P \in \mathcal{P} \) is characterised
-by a distribution \( Q \in \mathcal{Q} \) and a conditional cumulative
-hazard function for \( C \) given \( X \)
-\citep[c.f.,][]{begun1983information,gill1997coarsening}. We use
-\(\Gamma\in\mathcal{M}_{\tau}\) to denote the conditional cumulative hazard
-function for censoring. For ease of presentation we now also assume
-that \( \Gamma(\blank \mid x) \) is continuous for all \( x \). We let
-\((t,x)\mapsto G(t \mid x)=\exp\left\{-\Gamma(t \mid x)\right\}\) denote the
-survival function of the conditional censoring distribution. The
-distribution \( P \) is characterised by
-\begin{equation}\label{eq:parametrizeP}
-  \begin{split}
-    P(\diff t, j, \diff x) =& \left\{G(t- \mid x)S(t- \mid x)\Lambda_1(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=1\}}}}\\
-                            & \left\{G(t- \mid x)S(t- \mid x)\Lambda_2(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=2\}}}}\\
-                            & \left\{G(t- \mid x)S(t- \mid x)\Gamma(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=0\}}}}\\
-    = & \left\{G(t- \mid x)Q(\diff t,j,\diff x)\right\}^{\1{{\{j\ne 0\}}}}\\    
-                            & \left\{G(t- \mid x)S(t- \mid x)\Gamma(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=0\}}}}.
-  \end{split}
-\end{equation}
-Hence, we may write
-\( \mathcal{P} = \{ P_{Q, \Gamma} : Q \in \mathcal{Q}, \Gamma \in
-\mathcal{G} \} \) for some \( \mathcal{G} \subset \mathcal{M}_{\tau} \). We
-also have \(H\)-almost everywhere
-\begin{equation*}
-P(\tilde T>t \mid X=x) = S(t \mid x)G(t \mid x) = \exp\left\{-\Lambda_{1}(t \mid x)-\Lambda_{2}(t \mid x)-\Gamma(t \mid x) \right\}.
-\end{equation*}
-We further assume that there exists \(\kappa<\infty\) such that
-\(\Lambda_{j}(\tau- \mid x)<\kappa \), for \(j\in\{1,2\}\), and
-\(\Gamma(\tau- \mid x)<\kappa\) for almost all \(x\in\mathcal X\). Note that this
-implies that \(G(\tau- \mid x)\) is bounded away from zero for almost all \(x\in\mathcal X\).
-Under these assumptions, the conditional cumulative hazard functions
-\(\Lambda_{j}\) and \(\Gamma\) can be identified from \(P\) by
-\begin{align}
-  \Lambda_{j}(t \mid x) &= \int_0^t\frac{  P(\tilde T \in \diff s, \tilde D=j \mid X=x )}{P(\tilde T \geq s \mid X=x )}, \label{eq:lambdaj}\\
-  \Gamma(t \mid x) &= \int_0^t\frac{  P(\tilde T \in \diff s, \tilde D=0 \mid X=x )}{P(\tilde T \geq s \mid X=x )}\label{eq:gamma}.
-\end{align}
-Thus, we can consider $\Lambda_j$ and \(\Gamma\) as operators which map from
-\( \mathcal{P} \) to \(\mathcal M_{\tau}\).
-
-\section{The concept of super learning}
-\label{sec:super-learning}
-
-In survival analysis, a super learner can be used to estimate a
-parameter $\Psi$ which can be identified from the observed data
-distribution \(P\in\mathcal P\). In this section, to introduce the
-discrete super learner and the oracle learner, we consider estimation of the
-function-valued parameter \(\Psi:\mathcal P\to\mathcal M_{\tau}\), given by
-\(\Psi(P)=\Lambda_j\). This parameter is identified via equation
-\eqref{eq:lambdaj} on \([0,\tau]\).
-
-As input to the super learner we need a data set
-\( \data_n=\{O_i\}_{i=1}^n \) of i.i.d.\ observations from \( P \in \mathcal{P} \) and a collection of candidate
-learners $\mathcal{A}$. Each learner \(a \in \mathcal{A}\) is a map
-\( a \colon \mathcal{O}^n \rightarrow \mathcal{M}_{\tau}\) which takes a data
-set as input and returns an estimate $a(\data_n) \in \mathcal{M}_{\tau}$ of
-$\Lambda_{j}$.
-% Formally, the domain of $a$ depends on \(n\) but we suppress this in the notation.
-In what follows, we use the short-hand notation
-\(P[f] = \int f(o) P(\diff o) \). A super learner evaluates the
-performance of \(a \in \mathcal{A}\) with a loss function
-\(L\colon \mathcal{M}_{\tau} \times \mathcal{O} \rightarrow \R_+\) and
-estimates the expected loss \(P[L(a(\data_n), \blank)]\) using
-cross-validation. Specifically, the expected loss of $a\in\mathcal A$
-is estimated by splitting the data set $\data_n$ into $K$ disjoint
-approximately equally sized subsets
-\(\data_n^1, \data_n^2, \dots, \data_n^K \) and then calculating the
-cross-validated loss
-\begin{equation*}
-  % \label{eq:cv-risk-est}
-  \hat{R}_n(a; L) =
-  \frac{1}{K}\sum_{k=1}^{K}
-  % \empmeas^k{[L {(a{ (\data_n^{-k})} , \blank) }]},
-  \frac{1}{| \data_n^{k} |}\sum_{O_i \in \data_n^{k}}
-  L
-  {
-    \left(
-      a{ (\data_n^{-k})}
-      , O_i
-    \right)
-  },
-  \quad \text{with} \quad
-  \data_n^{-k} = \data_n \setminus \data_n^{k}.
-\end{equation*}
-% where \( \empmeas^{k} \) is the empirical measure of \( \data_n^{k} \). 
-The subset \(\data_n^{-k}\) is referred to as the \(k\)'th training
-sample, while \(\data_n^{k}\) is referred to as the \(k\)'th test or
-hold-out sample.
-The discrete super learner is defined as
-\begin{equation*}
-\hat{a}_n = \argmin_{a\in\mathcal A}\hat{R}_n(a; L).
-\end{equation*}
-% The final estimator of \(\Psi(P)=\Lambda_j\) is then the selected
-% learner applied to the full data set, i.e., \(\hat{a}_n(\data_n)\).
-The oracle learner is defined as the learner that minimises the
-expected loss under the data-generating distribution \( P \),
-i.e.,
-\begin{equation*}
-  \tilde{a}_n =
-  \argmin_{a \in \mathcal{A}}
-  \tilde{R}_n(a; L),
-  \quad \text{with} \quad 
-  \tilde{R}_n(a; L)=
-  \frac{1}{K}\sum_{k=1}^{K} 
-  P{
-    \left[
-      L
-      {
-        \left(
-          a{ (\data_n^{-k})}
-          , \blank
-        \right)
-      }
-    \right]}
-  .
-\end{equation*}
-Note that both the discrete super learner and the oracle learner
-depend on the library of learners and on the number of folds \(K\),
-and that the oracle learner is a function of the data and the unknown
-data-generating distribution. However, these dependencies are
-suppressed in the notation.
-
-\section{The state learner}
-\label{sec:super-learner-simple}
-
-The main idea of the state learner is to jointly use learners for
-\( \Lambda_1 \), \( \Lambda_2 \), and \( \Gamma \), and the relations in
-equation~(\ref{eq:parametrizeP}), to learn a feature of the observed data
-distribution \( P \). The discrete state learner ranks a tuple of learners for
-the tuple of the cumulative hazard functions
-\( (\Lambda_1, \Lambda_2, \Gamma) \) based on how well they jointly model the
-observed data. Risk predictions can then be obtained by combining
-\( \Lambda_1 \) and $\Lambda_2$ from the highest ranked tuple using a well-known
-formula \citep{benichou1990estimates, ozenne2017riskregression}. To formally
-introduce the state learner, we define the multi-state process
-\begin{equation*}
-  \eta(t) = \1\{\tilde{T} \leq t, \tilde D=1\} + 2\,\1\{\tilde{T} \leq t, \tilde
-  D=2\} - \1\{\tilde{T} \leq t, \tilde D=0\},
-  \quad \text{for} \quad t \in [0, \tau].
-\end{equation*}
-At time \(t\), we observe that each individual is in one of four mutually
-exclusive states: \( 0 \), \( 1 \), \( 2 \), or \( -1 \).
-The conditional distribution of the process \( \eta(t) \) given
-baseline covariates \( X \) is determined by the function
-\begin{equation}
-  \label{eq:F-def}
-  F(t, k, x) = P(\eta(t) = k \mid X=x),
-\end{equation}
-for all \( t \in [0,\tau] \), \( k \in \{-1,0,1,2\} \), and \( x \in \mathcal{X} \).
-The function \( F \) describes the conditional state occupation probabilities of
-the multi-state process \(\eta\). We construct a super learner for \( F \). The
-target of this super learner is the function-valued parameter $\Psi(P) = F$
-which is identified through equation~(\ref{eq:F-def}). Under conditional
-independent censoring each quadruple $(\Lambda_{1}, \Lambda_{2}, \Gamma, H)$
-characterises a distribution \(P\in\mathcal P\), c.f.\
-equation~\eqref{eq:parametrizeP}, which in turn determines \( (F, H) \). Hence,
-a learner for \( F \) can be constructed from learners for \( \Lambda_1 \),
-\( \Lambda_2 \), and $\Gamma$ as follows:
-\begin{equation}\label{eq:transition}
-  \begin{split}
-    F(t, 0, x)
-    &
-      = P(\tilde{T} > t \mid X= x)
-      \\
-      & =
-      \exp{{\{-\Lambda_{1}(t \mid x)-\Lambda_{2}(t \mid x) - \Gamma(t \mid x)\}
-      }},
-    \\
-    F(t, 1, x)
-    &
-      = P(\tilde{T} \leq t, \tilde{D}=1 \mid X=x)
-      = \int_0^t F(s-, 0, x)  \Lambda_{1}(\diff s \mid x),
-    \\
-    F(t, 2, x)
-    &
-      = P(\tilde{T} \leq t, \tilde{D}=2 \mid X=x)
-      = \int_0^t  F(s-, 0, x)  \Lambda_{2}(\diff s \mid x),
-    \\
-    F(t, -1, x)
-    &
-      = P(\tilde{T} \leq t, \tilde{D}=0 \mid X=x)
-      = \int_0^t F(s-, 0, x)  \Gamma(\diff s \mid x).
-  \end{split}
-\end{equation}
-The state learner requires three libraries of learners (c.f., Section \ref{sec:super-learning}),
-\(\mathcal{A}_1\), \( \mathcal{A}_2 \), and \( \mathcal{B} \), where
-\(\mathcal{A}_1\) and \( \mathcal{A}_2\) contain learners for the
-conditional cause-specific cumulative hazard functions \(\Lambda_1\)
-and \( \Lambda_2\), respectively, and \(\mathcal{B}\) contains
-learners for the conditional cumulative hazard function of the
-censoring distribution. % We further
-% define \(\mathcal{H}\) as the set of all probability distributions on
-% \( \mathcal{X} \) and
-% \begin{equation*}
-%   \mathcal{H}_n=\left\{h\colon\mathcal{O}^n\longrightarrow\mathcal{H} \midd
-%     h(\data_n)=H_n = \frac 1 n \sum_{i=1}^n \delta_{X_i} \right\}
-% \end{equation*}
-% as the library of learners which consists of a single learner, the empirical
-% distribution function.
-Based on the Cartesian product of
-libraries of learners for \((\Lambda_1,\Lambda_2,\Gamma)\) we construct a library
-$\mathcal{F}$ of learners
-for \( F \):
-\begin{align*}
-  \mathcal{F}(\mathcal{A}_1, \mathcal{A}_2, \mathcal{B})
-  &= \{ \phi_{a_1,a_2, b} : a_1 \in \mathcal{A}_1, a_2 \in \mathcal{A}_2, b \in \mathcal{B}\},
-    \intertext{where in correspondence with  the relations in equation
-    \eqref{eq:transition},}
-    \phi_{a_1,a_2, b}(\data_n)(t,0,x)
-  &= \exp{\{-a_1(\data_n)(s \mid x)-a_2(\data_n)(s \mid x) - b(\data_n)(s \mid
-    x)\} },
-  \\
-  \phi_{a_1,a_2, b}(\data_n)(t,1,x)
-  &= \int_0^t
-    \phi_{a_1,a_2, b}(\data_n)(s-,0,x)  a_1(\data_n)(\diff s \mid x),
-  \\
-  \phi_{a_1,a_2, b}(\data_n)(t,2,x)
-  &= \int_0^t\phi_{a_1,a_2, b}(\data_n)(s-,0,x)  a_2(\data_n)(\diff s \mid x),
-  \\
-  \phi_{a_1,a_2, b}(\data_n)(t,-1,x)
-  &= \int_0^t \phi_{a_1,a_2, b}(\data_n)(s-,0,x)  b(\data_n)(\diff s \mid x).
-\end{align*}
-To evaluate how well a function \( F \) predicts the observed
-multi-state process we use the integrated Brier score
-\( \bar B_\tau( F,O) = \int_0^{\tau} B_t(F,O) \diff t \), where \( B_t \) is the
-Brier score \citep{brier1950verification} at time \( t \in [0, \tau] \),
-\begin{equation*}
-  B_t(F,O) = \sum_{j=-1}^{2}
-  \left(
-      F(t,j,X) - \1{\{\eta(t)=j\}}
-  \right)^2.
-\end{equation*}
-Based on a split of a data set \(\data_n\) into $K$ disjoint
-approximately equally sized subsets (see Section \ref{sec:super-learning}), each learner
-\( \phi_{a_1, a_2, b} \) in the library
-\( \mathcal{F}(\mathcal{A}_1, \mathcal{A}_2, \mathcal{B}) \) is
-evaluated using the cross-validated loss,
-\begin{equation*}
-  \hat{R}_{n}(\phi_{a_1,a_2,b} ; \bar{B}_{\tau}) =
-  \frac{1}{K}\sum_{k=1}^{K}
-  \frac{1}{| \data_n^{k} |}\sum_{O_i \in \data_n^{k}}
-  \bar B_\tau
-  {
-    \left(
-      \phi_{a_1,a_2,b}{ (\data_n^{-k})}
-      , O_i
-    \right)
-  },
-\end{equation*}
-and the discrete state learner is given by
-\begin{align*}\label{eq:discrete-state-learner}
-  \hat{\phi}_n
-  &=  \argmin_{(a_1,a_2,b)\in \mathcal{A}_1\times\mathcal{A}_2\times\mathcal{B}}
-    \hat{R}_{n}(\phi_{a_1,a_2,b} ; \bar{B}_{\tau}).
-\end{align*}
-
-\section{Theoretical results for the state learner}
-\label{sec:theor-results-prop}
-
-In this section we establish theoretical guarantees for the state learner. We
-show that the state learner is consistent if its library contains a consistent
-learner, and we establish a finite sample inequality for the excess risk of the
-state learner compared to the oracle. Finally we show that a certain second
-order structure is preserved when the state learner is used for targeted
-learning.
-
-\subsection{Consistency}
-\label{sec:consistency}
-
-% We let \( \E_P \) denote
-% expectation under \( P \) and use \( \Theta \) to denote the
-% collection of all conditional state-occupation probability functions.
-Proposition~\ref{prop:stric-prop} can be derived from the fact that the
-integrated Brier score (also called the continuous ranked probability score) is
-a strictly proper scoring rule \citep{gneiting2007strictly}. This implies that
-if we minimise the average loss of the integrated Brier score, we recover the
-parameters of the data-generating distribution. Specifically, this implies that
-the oracle of a state learner is consistent if the library of learners contains
-at least one learner that is consistent for estimation of \( F \). Recall that
-the function \(F\) implicitly depends on the data-generating probability measure
-\(P\in\mathcal P\) but that this was suppressed in the notation. We now make
-this dependence explicit by writing \(F_0\) for the function which is obtained
-by substituting a specific \(P_0\in\mathcal{P}\) for \(P\) in equation
-\eqref{eq:transition}. In the following we let
-\( \mathcal{H}_{\mathcal{P}} = \{F_P : P \in \mathcal{P}\} \) where \( F_P \) is defined as in
-equation~(\ref{eq:F-def}) using the measure \( P \in \mathcal{P} \).
-
-\begin{proposition}
-  \label{prop:stric-prop}
-  If \(P_0\in\mathcal{P}\) then
-  \begin{equation*}
-    F_0 = \argmin_{F \in \mathcal{H}_{\mathcal{P}}} P_0{[\bar{B}_\tau(F, \blank)]},
-  % F^* = \argmin_{(\Lambda_1,\Lambda_2,\Gamma) \in \mathcal{M}_{\tau}\otimes\mathcal{M}_{\tau}\otimes\mathcal{M}_{\tau}} P_0{[\bar B_\tau(\phi_{(\Lambda_1,\Lambda_2,\Gamma)}, \blank)]},
-  \end{equation*}
-  \( H \)-almost surely for any \( j\in \{-1,0,1,2\} \) and almost any
-  \( t \in [0, \tau]\).
-\end{proposition}
-\begin{proof}
-  See Appendix~\ref{sec:consistency-proof}.
-\end{proof}
-
-\subsection{Oracle inequalities}
-\label{sec:finite-sample-oracle}
-
-We establish a finite sample oracle result for the state learner. Our
-Corollary~\ref{cor:oracle-prop} is in essence a special case of
-Theorem 2.3 in \citep{vaart2006oracle}.  We assume that we split the
-data into equally sized folds, and for simplicity of presentation we
-take \( n \) to be such that \( |\data_n^{-k}| = n/K \) with \( K \)
-fixed. We will allow the number of learners to grow with \( n \) and
-write
-\( \mathcal{F}_n=\mathcal{F}(\mathcal{A}_{1,n}, \mathcal{A}_{2,n},
-\mathcal{B}_n)\) as short-hand notation and to emphasise the
-dependence on \( n \).
-% The discrete super $\hat{\phi}_n$ and the
-% oracle $\tilde{\phi}_n$ are defined as in Section~\ref{sec:framework}
-% but now with respect to the library \( \mathcal{F}_n \) and the loss
-% \( B \).
-In the following we let the space \( \mathcal{H}_{\mathcal{P}} \) be equipped with the norm
-\( \| \blank \|_{P_0} \) defined as
-\begin{equation}
-  \label{eq:norm}
-  \| F \|_{P_0} = 
-  \left\{
-    \sum_{j=-1}^{2}\int_{\mathcal{X}} \int_0^{\tau} F(t, j, x)^2 \diff t H_0( \diff x)
-  \right\}^{1/2}.
-\end{equation}
-
-\begin{corollary}
-  \label{cor:oracle-prop}
-  For all \(P_0\in\mathcal{P}\), \( n \in \N \), \( k \in \{1, \dots, K\} \),
-  and $\delta>0$,
-  \begin{align*}
-    \E_{P_0}{\left[ \Vert \hat{\phi}_n(\data_n^{-k}) - F_0 \Vert_{P_0}^2 \right]}
-    & \leq (1 + 2\delta)
-      \E_{P_0}{\left[ \Vert \tilde{\phi}_n(\data_n^{-k}) - F_0 \Vert_{P_0}^2 \right]}
-    \\
-    & \quad
-      + (1+ \delta) 16   K \tau
-      \left(
-      13 + \frac{12}{\delta}
-      \right)
-      \frac{\log(1 + |\mathcal{F}_n|)}{n}.
-  \end{align*}
-\end{corollary}
-\begin{proof}
-  See Appendix~\ref{sec:oracle-inequalities}.
-\end{proof}
-
-Corollary~\ref{cor:oracle-prop} has the following asymptotic consequences.
-
-\begin{corollary}
-  \label{cor:asymp-cons}
-  Assume that \( |\mathcal{F}_n| = \bigO(n^q)\), for some \( q \in \N \) and
-  that there exists a sequence \( \phi_n \in \mathcal{F}_n \), \( n \in \N \),
-  such that
-  \( \E_{P_0}{\left[ \Vert \phi_n(\data_n^{-k}) - F_{0} \Vert_{P_0}^2 \right]} =
-  \bigO(n^{-\alpha}) \), for some \( \alpha\leq 1 \).
-  \begin{enumerate}[label=(\alph*)]
-  \item If $\alpha=1$ then
-    \( \E_{P_0}{\left[ \Vert \hat{\phi}_n(\data_n^{-k}) - F_0 \Vert_{P_0}^2
-      \right]} = \bigO(\log(n)n^{-1}) \).
-  \item If $\alpha<1$ then
-    \( \E_{P_0}{\left[ \Vert \hat{\phi}_n(\data_n^{-k}) - F_0 \Vert_{P_0}^2 \right]} =
-    \bigO(n^{-\alpha}) \).
-  \end{enumerate}
-\end{corollary}
-\begin{proof}
-  See Appendix~\ref{sec:oracle-inequalities}.
-\end{proof}
-
 \subsection{Transience of the second order remainder structure}
@@
 
-\section{Numerical experiments}
-\label{sec:numer-exper}
-
-In this section we report results from a simulation study where we consider
-estimation of the conditional survival function. In the first part, we compare
-the state learner to two IPCW based discrete super learners that use either the
-Kaplan-Meier estimator or a Cox model to estimate the censoring probability
-\citep{gonzalez2021stacked}. In the second part we compare the state learner to
-the super learner proposed by \cite{westling2021inference}.
-
-In both parts we use the same data-generating mechanism. We generate data
-according to a distribution motivated from a real data set in which censoring
-depends on the baseline covariates. We simulate data based on the prostate
-cancer study of \cite{kattan2000pretreatment}. The outcome of interest is the
-time to tumor recurrence, and five baseline covariates are used to predict
-outcome: prostate-specific antigen (PSA, ng/mL), Gleason score sum (GSS, values
-between 6 and 10), radiation dose (RD), hormone therapy (HT, yes/no) and
-clinical stage (CS, six values). The study was designed such that a patient's
-radiation dose depended on when the patient entered the study
-\citep{gerds2013estimating}. This in turn implies that the time of censoring
-depends on the radiation dose. The data were re-analysed in
-\citep{gerds2013estimating} where a sensitivity analysis was conducted based on
-simulated data. Here we use the same simulation setup, where event and censoring
-times are generated according to parametric Cox-Weibull models estimated from
-the original data, and the covariates are generated according to either marginal
-Gaussian normal or binomial distributions estimated from the original data
-\citep[c.f.,][Section~4.6]{gerds2013estimating}. We refer to this simulation
-setting as `dependent censoring'. We also considered a simulation setting where
-data were generated in the same way, except that censoring was generated
-completely independently. We refer to this simulation setting as `independent
-censoring'.
-
-For all super learners we use a library consisting of three learners:
-The Kaplan-Meier estimator
-\citep{kaplan1958nonparametric,Gerds_2019prodlim}, a Cox model with
-main effects \citep{cox1972regression, survival-package}, and a random
-survival forest \citep{ishwaran2008random,rfsrc-paclage}. We use the
-same library to learn the outcome distribution and the censoring
-distribution. Note that the three learners in our library of learners
-can be used to learn the cumulative hazard functions of the 
-outcome and the censoring distribution. The latter works by
-training the learner on the data set \( \data_n^c \), where
-\( \data_n^c = \{O_i^c\}_{i=1}^n \) with
-\( O_i^c = (\tilde{T}_i, 1-\Delta, X_i) \). When we say that we use a
-learner for the cumulative hazard function of the outcome to learn the
-cumulative hazard function of the censoring time, we mean that the
-learner is trained on \( \data_n^c \).
-
-We compare the state learner to two IPCW based super learners: The
-first super learner, called IPCW(Cox), uses a Cox model with main
-effects to estimate the censoring probabilities, while the second
-super learner, called IPCW(KM), uses the Kaplan-Meier estimator to
-estimate the censoring probabilities. The Cox model for the censoring
-distribution is correctly specified in both simulation settings while
-the Kaplan Meier estimator only estimates the censoring model
-correctly in the simulation setting where censoring is
-independent. Both IPCW super learners are fitted using the
-\texttt{R}-package \texttt{riskRegression}
-\citep{Gerds_Ohlendorff_Ozenne_2023}.
-%
-% | time | sim_setting | true_events | true_cens | at_risk |
-% |------+-------------+-------------+-----------+---------|
-% |   36 | original    |      24.619 |    61.853 |  25.774 |
-% |   36 | indep_cens  |      24.674 |    38.740 |  46.141 |
-%
-The IPCW super learners use the integrated Brier score up to a fixed time
-horizon (36 months). The marginal risk of the event before this time horizon is
-\(\approx 24.6\)\%. Under the `dependent censoring' setting the marginal
-censoring probability before the time horizon is \(\approx 61.9\)\%. Under the
-`independent censoring' setting the marginal censoring probability before this
-time horizon is \( \approx 38.7 \)\%.
-
-Each super learner provides a learner for the cumulative hazard
-function for the outcome of interest. From the cumulative hazard
-function a risk prediction model can be obtained (c.f.,
-equation~(\ref{eq:surv-def}) with $\Lambda_2 = 0$). We measure the
-performance of each super learner by calculating the index of
-prediction accuracy (IPA) \citep{kattan2018index} at a fixed time
-horizon (36 months) for the risk prediction model provided by the
-super learner. The IPA is 1 minus the ratio between the model's Brier
-score and the null model's Brier score, where the null model is the
-model that does not use any covariate information. The IPA is
-approximated using a large (\( n = 20,000 \)) independent data set of
-uncensored data. As a benchmark we calculate the performance of the
-risk prediction model chosen by the oracle selector, which uses the
-large data set of uncensored event times to select the learner with
-the highest IPA.
-
-The results are shown in Figure~\ref{fig:ipcw-fail}. We see that in
-the scenario where censoring depends on the covariates, using the
-Kaplan-Meier estimator to estimate the censoring probabilities
-provides a risk prediction model with an IPA that is lower than the
-risk prediction model provided by the state learner. The performance
-of the risk prediction model selected by the state learner is similar
-to the risk prediction model selected by the IPCW(Cox) super learner
-which a priori uses a correctly specified model for the censoring
-distribution. Both these risk prediction models are close to the
-performance of the oracle, except for small sample sizes.
-
-\begin{figure}
-  \centering %
-  \includegraphics[width=1\linewidth]{experiment-fig-sl-ipcw.pdf}
-  \caption[]{For the risk prediction models provided by each of the super
-    learners, the IPA is plotted against sample size. The results are averages across 
-    1000 simulated data sets and the error bars are used to quantify the Monte Carlo
-    uncertainty.
-  }
-  \label{fig:ipcw-fail}
-\end{figure}
-
-We next compare the state learner to the super learner \texttt{survSL}
-\citep{westling2021inference}. This is another super learner which
-like the state learner works without a pre-specified censoring
-model. Note that both the state learner and \texttt{survSL} provide a
-prediction model for the event time outcome and also for the
-probability of being censored. Hence, we compare the performance of
-these methods with respect to both the outcome and the censoring
-distribution. Again we use the IPA to quantify the predictive
-performance.
-
-The results are shown in Figures~\ref{fig:zelefski-out}
-and~\ref{fig:zelefski-cens}. We see that for most sample sizes, the state
-learner selected prediction models for both censoring and outcome which have
-similar or higher IPA compared to the prediction models selected by
-\texttt{survSL}.
-\begin{figure}
-  \centering %
-  \includegraphics[width=1\linewidth]{experiment-fig-sl-survSL-out.pdf}
-  \caption[]{For the risk prediction models of the outcome provided by each
-    of the super learners, the IPA at the fixed time horizon is plotted against
-    sample size. The results are averages across 1000 repetitions and the error
-    bars are used to quantify the Monte Carlo uncertainty.}
-  \label{fig:zelefski-out}
-\end{figure}
-
-\begin{figure}
-  \centering %
-  \includegraphics[width=.7\linewidth]{experiment-fig-sl-survSL-cens.pdf}
-  \caption[]{For the risk prediction models of the censoring model provided by
-    each of the super learners, the IPA at the fixed time horizon is plotted
-    against sample size. The results are averages across 1000 repetitions and
-    the error bars are used to quantify the Monte Carlo uncertainty.}
-  \label{fig:zelefski-cens}
-\end{figure}
-
-\section{Prostate cancer study}
-\label{sec:real-data-appl}
-
-In this section we use the prostate cancer data of
-\cite{kattan2000pretreatment} to illustrate the use of the state
-learner in the presence of competing risks. We have introduced the
-data in Section~\ref{sec:numer-exper}. The data consists of 1,042
-patients who are followed from start of followup until tumor
-recurrence, death without tumor recurrence or end of followup
-(censored) whatever came first. For the sole purpose of illustration,
-we estimate the average treatment effect of hormone therapy on death
-and tumor recurrence. To do this we adapt the estimation strategy of
-Section~\ref{sec:targeted-learning} as follows. We use the state
-learner to rank libraries of learners for the cause-specific
-cumulative hazard functions of tumor recurrence, death without tumor
-recurrence, and censoring.  The libraries of learners each include
-five learners: the Nelson-Aalen estimator, three Cox regression models
-(unpenalized, Lasso, Elastic net) each including additive effects of
-the 5 covariates (Section~\ref{sec:numer-exper}), and a random
-survival forest. We use the same set of learners to learn the
-cumulative hazard function of tumor recurrence \( \Lambda_1 \), the
-cumulative hazard function of death without tumor recurrence
-\( \Lambda_2 \), and the cumulative hazard function of the conditional
-censoring distribution $\Gamma$.  We then use the highest ranked
-combination of learners and apply
-formula~\eqref{eq:one-step-comp-ate}.
-
-This gives a library consisting of \( 5^3 = 125 \) learners for the
-conditional state occupation probability function \( F \) defined in
-equation~(\ref{eq:F-def}). We use five folds for training and testing
-the models, and we repeat training and evaluation five times with
-different splits.  The integrated Brier score (defined in
-Section~\ref{sec:super-learner-simple}) for all learners are shown in
-Figure~\ref{fig:zelefski-real}. We see that the prediction
-performance is mostly affected by the choice of learner for the
-censoring distribution. Several combinations of learners give similar
-performance as measured by the integrated Brier score, as long as a
-random forest is used to model the censoring distribution.
-
-\begin{figure}
-  \centering %
-  \includegraphics[width=1\linewidth]{real-data-state-learner.pdf}
-  \caption[]{The results of applying the 125 combinations of learners to the
-    prostate cancer data set. The error bars are based on five repetitions using
-    different splits. We refer to learners of \( \Lambda_1 \), \( \Lambda_2 \),
-    and $\Gamma$ as `Tumor learner', `Mortality learner', and `Censoring
-    learner', respectively.}
-  \label{fig:zelefski-real}
-\end{figure}
-
 We use the learners of the three cumulative hazard functions selected
@@
 
-\section{Discussion}
-\label{sec:discussion}
-
-The state learner is a new super learner that can be used with
-right-censored data and competing events. Compared to existing
-IPCW-based methods, the advantage of the state learner is that it does
-not depend on a pre-specified estimator of the censoring distribution,
-but selects one automatically based on a library of learners for the
-censoring distribution. Furthermore, the state learner neither
-requires that the cause-specific cumulative hazard functions
-\( \Lambda_j \) can be written as integrals with respect to Lebesgue
-measure, nor does it assume a (semi-)parametric formula. In the
-remainder of this section we discuss the limitations of our proposal
-and avenues for further research.
-
-A major advantage of the state learner is that the performance of each
-combination of learners can be estimated without additional nuisance
-parameters. A potential drawback of our approach is that we are
-evaluating the loss of the learners on the level of the observed data
-distribution while the target of the analysis is either the event time
-distribution, or the censoring distribution, or both.
-Specifically, the finite sample oracle inequality in
-Corollary~\ref{cor:oracle-prop} concerns the function \( F \), which
-is a feature of \( P \in \mathcal{P} \), while what we are typically
-interested in is \( \Lambda_j \) or \( S \), which are features of
-\( Q \in \mathcal{Q} \). We emphasise that while the state learner
-provides us with estimates of \( \Lambda_j \) and $\Gamma$ based on
-libraries \( \mathcal{A}_j \) and \( \mathcal{B} \), the performance
-of these learners is not assessed directly for their respective target
-parameters, but only indirectly via the performance of \( F \).  For
-settings without competing risks, our numerical studies suggest that
-measuring the performance of \( F \) also leads to good performance
-for estimation of \( S \).
-
-Our proposed super learner can be implemented with a broad library of learners
-and using existing software.
-% For instance the \texttt{R}-package
-% \texttt{riskRegression} \citep{Gerds_Ohlendorff_Ozenne_2023} and the \texttt{R}-package by westling
-Furthermore, while
-the library \( \mathcal{F}(\mathcal{A}_1,\mathcal{A}_2,\mathcal{B}) \) consists
-of \( |\mathcal{A}_1||\mathcal{A}_2||\mathcal{B}| \) many learners, we only need to fit
-\( |\mathcal{A}_1| +|\mathcal{A}_2| + |\mathcal{B}| \) many learners in each fold. To
-evaluate the performance of each learner we need to perform
-\( |\mathcal{A}_1||\mathcal{A}_2||\mathcal{B}| \) many operations to calculate the
-integrated Brier score in each hold-out sample, one for each combination of the
-fitted models, but these operations are often negligible compared to fitting the
-models. Hence the state learner is essentially not more computationally demanding
-than any procedure that uses super learning to learn $\Lambda_1$, $\Lambda_2$,
-and $\Gamma$ separately. While our proposal is based on constructing the library
-\( \mathcal{F} \) from libraries for learning \( \Lambda_1 \), $\Lambda_2$, and
-$\Gamma$, it could also be of interest to consider learners that estimate
-\( F \) directly.
-
-In our numerical studies, we only considered learners of $\Lambda_j$ and
-$\Gamma$ that provide cumulative hazard functions which are piece-wise constant
-in the time argument. This simplifies the calculation of \( F \) as the
-integrals in equation~(\ref{eq:transition}) reduce to sums. When $\Lambda_j$ or
-\( \Gamma \) are absolutely continuous in the time argument, calculating \( F \)
-is more involved, but we expect that a good approximation can be achieved by
-discretisation.
-
-\section*{Conflict of interest}
-
-The authors declare that they have no conflict of interest.
-
-% BibTeX users please use one of
-% \bibliographystyle{spbasic}      % basic style, author-year citations
-%\bibliographystyle{spmpsci}      % mathematics and physical sciences
-%\bibliographystyle{spphys}       % APS-like style for physics
-
-\bibliographystyle{abbrvnat}
-\bibliography{bib.bib}   % name your BibTeX data base
-
 \appendix
-
-\section{Proofs}
-\label{sec:proofs}
-
-This section contains proofs of the results stated in the paper.
-Section~\ref{sec:consistency-proof} contains a proof of the consistency result
-from Section~\ref{sec:consistency}; Section~\ref{sec:oracle-inequalities}
-contains proofs of the oracle inequalities from
-Section~\ref{sec:finite-sample-oracle}; and
-Section~\ref{sec:state-learner-with} demonstrates transience of the second
-order remainder structure stated in Section~\ref{sec:trans-second-order}.
-
-\subsection{Consistency}
-\label{sec:consistency-proof}
-
-Define
-\( \bar{B}_{\tau,0}(F, o) = \bar{B}_{\tau}(F, o) - \bar{B}_{\tau}(F_0, o) \) and
-\( R_{0}(F) = P_0{[\bar{B}_{\tau,0}(F, \blank)]} \), where the integrated Brier
-score \( \bar{B}_{\tau} \) was defined in
-Section~\ref{sec:super-learner-simple}.
-\begin{lemma}
-  \label{lemma:norm}
-  \( R_{0}(F) = \Vert F - F_0 \Vert_{P_0}^2 \), where \( \Vert \blank \Vert_{P_0}\) is defined
-  in equation~(\ref{eq:norm}).
-\end{lemma}
-\begin{proof}[Proof]
-  For any \( t \in [0, \tau] \) and \( j\in \{-1,0,1,2\} \) we have
-  \begin{align*}
-    & \E_{P_0}{\left[ (F(t, j, X) - \1{\{\eta(t) = j \}})^2 \right]}
-    \\
-    & =    \E_{P_0}{\left[ (F(t, j, X) - F_0(t, j, X) + F_0(t, j, X) - \1{\{\eta(t) = j
-      \}})^2 \right]}
-    \\
-    & =    \E_{P_0}{\left[ (F(t, j, X) - F_0(t, j, X))^2\right]}
-      + \E_{P_0}{\left[ (F_0(t, j, X) - \1{\{\eta(t) = j \}})^2\right]}
-    \\
-    & \quad
-      + 2\E_{P_0}{\left[ (F(t, j, X) - F_0(t, j, X))(F_0(t, j, X) - \1{\{\eta(t) = j
-      \}})\right]}
-    \\
-    & =    \E_{P_0}{\left[ (F(t, j, X) - F_0(t, j, X))^2\right]}
-      + \E_{P_0}{\left[ (F_0(t, j, X) - \1{\{\eta(t) = j \}})^2\right]},
-  \end{align*}
-  where the last equality follows from the tower property. Hence, using Fubini,
-  we have
-  \begin{equation*}
-    P{[\bar{B}_{\tau}(F, \blank)]}
-    = \Vert F - F_0 \Vert_{P_0}^2 + P_0{[\bar{B}_{\tau}(F_0, \blank)]}.
-  \end{equation*}
-\end{proof}
-
-\begin{proof}[Proof of Proposition~\ref{prop:stric-prop}]
-  The result follows from Lemma~\ref{lemma:norm}.
-\end{proof}
-
-\subsection{Oracle inequalities}
-\label{sec:oracle-inequalities}
-
-Recall that we use \( \mathcal{F}_n \) to denote a library of learners for the
-function \( F \), and that \( \hat{\phi} \) and \( \tilde{\phi} \) denotes,
-respectively, the discrete super learner and the oracle learner for the library
-\( \mathcal{F}_n \), c.f., Section~\ref{sec:super-learner-simple}.
-
-\begin{proof}[Proof of Corollary~\ref{cor:oracle-prop}]
-  First note that minimising the loss \( \bar{B}_{\tau} \) is equivalent to
-  minimising the loss \( \bar{B}_{\tau,0} \), so the discrete super learner and
-  oracle according to \( \bar{B}_{\tau} \) and \( \bar{B}_{\tau,0} \) are
-  identical. By Lemma~\ref{lemma:norm}, \( R_0(F) \geq 0 \) for any
-  \( F \in \mathcal{H}_{\mathcal{P}} \), and so using Theorem 2.3 from
-  \citep{vaart2006oracle} with \( p=1 \), we have that for all \( \delta >0 \),
-\begin{align*}
-  & \E_{P_0}{\left[ R_0(\hat{\phi}_n(\data_n^{-k})) \right]}
-  \\
-  &  \quad \leq
-    (1+2\delta)\E_{P_0}{\left[ R_0(\tilde{\phi}_n(\data_n^{-k})) \right]}
-  \\
-  & \qquad + (1+\delta) \frac{16 K}{n}
-    \log(1 + |\mathcal{F}_n|)\sup_{F \in \mathcal{H}_{\mathcal{P}}}
-    \left\{
-    M(F) + \frac{v(F)}{R_0(F)}
-    \left(
-    \frac{1}{\delta} + 1
-    \right)
-    \right\}
-\end{align*}
-where for each \( F \in \mathcal{H}_{\mathcal{P}} \), \( (M(F), v(F)) \) is some Bernstein pair for
-the function \(o \mapsto \bar{B}_{\tau,0}(F, o) \). As
-\( \bar{B}_{\tau,0}(F, \blank) \) is uniformly bounded by \( \tau \) for any
-\( F \in \mathcal{H}_{\mathcal{P}} \), it follows from section 8.1 in \citep{vaart2006oracle} that
-\( (\tau, 1.5 P_0{[\bar{B}_{\tau,0}(F, \blank)^2]}) \) is a Bernstein pair for
-\( \bar{B}_{\tau,0}(F, \blank) \). Now, for any \( a,b,c \in \R \) we have
-\begin{align*}
-  (a-c)^2 - (b-c)^2
-  & = (a-b+b-c)^2 - (b-c)^2
-  \\
-  & = (a-b)^2 + (b-c)^2 +2(b-c)(a-b) - (b-c)^2
-  \\
-  & = (a-b)
-    \left\{
-    (a-b) +  2(b-c)
-    \right\}
-  \\
-  & = (a-b)
-    \left\{
-     a + b -2c
-    \right\},
-\end{align*}
-so using this with \( a=F(t, j, x) \), \( b=F_0(t, j, x) \), and
-\( c = \1{\{\eta(t) = j\}} \), we have by Jensen's inequality
-\begin{align*}
-  & P_0{[\bar{B}_{\tau,0}(F, \blank)^2]}
-  \\
-  & \leq
-    2\tau\E_{P_0}{\left[
-    \sum_{j=-1}^{2} \int_0^{\tau}
-    \left\{
-    \left(
-    F(t, j, X) - \1{\{\eta(t) = j\}}
-    \right)^2
-    -
-    \left(
-    F_0(t, j, X) - \1{\{\eta(t) = j\}}
-    \right)^2
-    \right\}^2
-    \diff t 
-    \right]}
-  \\
-  & =2\tau
-    \E_{P_0}\Bigg[
-    \sum_{j=-1}^{2} \int_0^{\tau}
-    \left(
-    F(t, j, X) - F_0(t, j, X)
-    \right)^2
-  \\
-  & \quad \quad \quad\quad \quad \quad \times
-    \left\{
-    F(t, j, X) +  F_0(t, j, X)-2 \1{\{\eta(t) = j\}}
-    \right\}^2
-    \diff t 
-    \Bigg]
-  \\
-  & \leq
-    8\tau \E_{P_0}{\left[
-    \sum_{j=-1}^{2} \int_0^{\tau}
-    \left(
-    F(t, j, X) - F_0(t, j, X)
-    \right)^2
-    \diff t 
-    \right]}.
-  \\
-  & =
-    8\tau \Vert F - F_0 \Vert_{P_0}^2.
-\end{align*}
-Thus when \( v(F) = 1.5 P_0{[\bar{B}_{\tau,0}(F, \blank)^2]} \) we have by
-Lemma~\ref{lemma:norm}
-\begin{equation*}
-  \frac{v(F)}{R_0(F)}
-  = 1.5 \frac{P_0{[\bar{B}_{\tau,0}(F, \blank)^2]}}{P_0{[\bar{B}_{\tau,0}(F, \blank)]}}
-  \leq 12 \tau,
-\end{equation*}
-and so using the Bernstein pairs \( (\tau, 1.5 P_0{[\bar{B}_{\tau,0}(F, \blank)^2]}) \) we have
-\begin{equation*}
-  \sup_{F \in \mathcal{H}_{\mathcal{P}}}
-  \left\{
-    M(F) + \frac{v(F)}{R_0(F)}
-    \left(
-      \frac{1}{\delta} + 1
-    \right)
-  \right\}
-  \leq \tau
-  \left(
-    13 + \frac{12}{\delta}
-  \right),
-\end{equation*}
-For all $\delta>0$ we thus have
-\begin{align*}
-  \E_{P_0}{\left[ R_0(\hat{\phi}_n(\data_n^{-k})) \right]}
-  \leq
-  &(1+2\delta)\E_{P_0}{\left[ R_0(\tilde{\phi}_n(\data_n^{-k})) \right]}
-  \\
-  & \quad
-    + (1+\delta)\log(1 + |\mathcal{F}_n|) \tau \frac{16 K}{n}
-    \left(
-    13 + \frac{12}{\delta}
-    \right),
-\end{align*}
-and then the final result follows from Lemma~\ref{lemma:norm}.
-\end{proof}
-
-\begin{proof}[Proof of Corollary~\ref{cor:asymp-cons}]
-  By definition of the oracle and Lemma~\ref{lemma:norm},
-  \begin{equation*}
-    \E_{P_0}{\left[ \Vert \tilde{\phi}_n(\data_n^{-k}) - F_0 \Vert_{P_0}^2
-      \right]} \leq \E_{P_0}{\left[ \Vert \phi_n(\data_n^{-k}) - F_0 \Vert_{P_0}^2
-      \right]}  
-  \end{equation*}
-  for all \( n \in \N \). The results then follows from
-  Corollary~\ref{cor:oracle-prop}.
-\end{proof}
 
@@
 \end{proof}
-
-\end{document}
-% end of file template.tex

### 2025-05-21 manuscript/manuscript-one.tex
use cases section draft

@@
 
-% %%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
-% %
-% % This is a general template file for the LaTeX package SVJour3
-% % for Springer journals.          Springer Heidelberg 2010/09/16
-% %
-% % Copy it to a new file with a new name and use it as the basis
-% % for your article. Delete % signs as needed.
-% %
-% % This template includes a few options for different layouts and
-% % content for various journals. Please consult a previous issue of
-% % your journal as needed.
-% %
-% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% %
-% % First comes an example EPS file -- just ignore it and
-% % proceed on the \documentclass line
-% % your LaTeX will extract the file if required
-% \begin{filecontents*}{example.eps}
-% %!PS-Adobe-3.0 EPSF-3.0
-% %%BoundingBox: 19 19 221 221
-% %%CreationDate: Mon Sep 29 1997
-% %%Creator: programmed by hand (JK)
-% %%EndComments
-% gsave
-% newpath
-%   20 20 moveto
-%   20 220 lineto
-%   220 220 lineto
-%   220 20 lineto
-% closepath
-% 2 setlinewidth
-% gsave
-%   .4 setgray fill
-% grestore
-% stroke
-% grestore
-% \end{filecontents*}
-% %
-% \RequirePackage{fix-cm}
-% %
-% %\documentclass{svjour3}                     % onecolumn (standard format)
-% %\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
-% \documentclass[smallextended]{svjour3}       % onecolumn (second format)
-% %\documentclass[twocolumn]{svjour3}          % twocolumn
-% %
-% \smartqed  % flush right qed marks, e.g. at end of proof
-% %
-% \usepackage{graphicx}
-% %
-% % \usepackage{mathptmx}      % use Times fonts if available on your TeX system
-% %
-% % insert here the call for the packages your document requires
-% \usepackage{dsfont, amssymb,soul,xcolor, enumitem, amsmath,verbatim}
-% \usepackage[authoryear]{natbib}
-% \definecolor{linkcolor}{rgb}{0, 0, 0.54}
-% \usepackage[colorlinks,allcolors=linkcolor]{hyperref}
-% %
-% % please place your own definitions here and don't use \def but
-% \DeclareMathOperator{\E}{\mathbb{E}} % expectation
-% \newcommand{\Z}{\mathbb{Z}}
-% \newcommand{\R}{\mathbb{R}}
-% \newcommand{\N}{\mathbb{N}}
-% \newcommand{\C}{\mathbb{C}}
-% \renewcommand{\S}{\mathbb{S}}
-% \newcommand{\blank}{\makebox[1ex]{\textbf{$\cdot$}}}
-% \newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
-% \def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
-% \renewcommand{\phi}{\varphi}
-% \renewcommand{\epsilon}{\varepsilon}
-% \newcommand*\diff{\mathop{}\!\mathrm{d}}
-% \newcommand{\weakly}{\rightsquigarrow}
-% \newcommand\smallO{\textit{o}}
-% \newcommand\bigO{\textit{O}}
-% \newcommand{\midd}{\; \middle|\;}
-% \newcommand{\1}{\mathds{1}}
-% \usepackage{ifthen} %% Empirical process with default argument
-% \newcommand{\G}[2][n]{
-% {\ensuremath{\mathbb{G}_{#1}}{\left[#2\right]}}
-% }
-% \DeclareMathOperator*{\argmin}{\arg\!\min}
-% \DeclareMathOperator*{\argmax}{\arg\!\max}
-% \newcommand{\empmeas}{\ensuremath{\mathbb{P}_n}} % empirical measure
-% \newcommand{\data}{\ensuremath{\mathcal{D}}}
-
-% \begin{document}
-
-% \title{The state learner -- a super learner for right-censored
-%   data}
-
-% \author{Anders Munch \& Thomas A. Gerds}
-
-\maketitle
-
 \begin{abstract}
@@
   application of the state learner with prostate cancer data.
-
-  % \keywords{Competing risks, cross-validation, loss based estimation,
-  %   right-censored data, super learner}
-% \PACS{PACS code1 \and PACS code2 \and more}
-% \subclass{MSC code1 \and MSC code2 \and more}
 \end{abstract}
+
+\textit{\textbf{Keywords:} Competing risks, cross-validation,
+  loss based estimation, right-censored data, super learner}
 
@@
 
-\section{The concept of super learning}
+\section{Super learning}
 \label{sec:super-learning}
@@
 equation~(\ref{eq:F-def}). Under conditional independent
-censoring each quadruple
-$(\Lambda_{1}, \Lambda_{2}, \Gamma, H)$ characterises a
-distribution \(P\in\mathcal P\), c.f.\
+censoring each tuple $(\Lambda_{1}, \Lambda_{2}, \Gamma, H)$
+characterises a distribution \(P\in\mathcal P\), c.f.\
 equation~\eqref{eq:parametrizeP}, which in turn determines
@@
       = P(\tilde{T} > t \mid X= x)
-      \\
-      & =
+      =
       \exp{{\{-\Lambda_{1}(t \mid x)-\Lambda_{2}(t \mid x) - \Gamma(t \mid x)\}
@@
 \end{equation}
-The state learner requires three libraries of learners (c.f., Section \ref{sec:super-learning}),
-\(\mathcal{A}_1\), \( \mathcal{A}_2 \), and \( \mathcal{B} \), where
-\(\mathcal{A}_1\) and \( \mathcal{A}_2\) contain learners for the
-conditional cause-specific cumulative hazard functions \(\Lambda_1\)
-and \( \Lambda_2\), respectively, and \(\mathcal{B}\) contains
-learners for the conditional cumulative hazard function of the
-censoring distribution. % We further
+The state learner requires three libraries of learners:
+\(\mathcal{A}_1\), \( \mathcal{A}_2 \), and \( \mathcal{B} \),
+where \(\mathcal{A}_1\) and \( \mathcal{A}_2\) contain
+learners for the conditional cause-specific cumulative hazard
+functions \(\Lambda_1\) and \( \Lambda_2\), respectively, and
+\(\mathcal{B}\) contains learners for the conditional
+cumulative hazard function of the censoring
+distribution. % We further
 % define \(\mathcal{H}\) as the set of all probability distributions on
@@
 Based on a split of a data set \(\data_n\) into $K$ disjoint
-approximately equally sized subsets (see Section \ref{sec:super-learning}), each learner
+approximately equally sized subsets (c.f., Section
+\ref{sec:super-learning}), each learner
 \( \phi_{a_1, a_2, b} \) in the library
-\( \mathcal{F}(\mathcal{A}_1, \mathcal{A}_2, \mathcal{B}) \) is
-evaluated using the cross-validated loss,
+\( \mathcal{F}(\mathcal{A}_1, \mathcal{A}_2, \mathcal{B}) \)
+is evaluated using the cross-validated loss,
 \begin{equation*}
@@
 
-The parameter \( F \) is not of much interest...
+The state learner estimates the parameter \( F \) which is a
+feature of the observed right-censored data distribution
+\( P \). In particular, \( F \) depends on the censoring
+distribution and is typically not of direct interest in
+itself. In the following, we describe how an estimate of
+\( F \), as provided by the state learner, can be used to
+estimate more relevant parameters of interest, such as
+survival probabilities.
 
@@
 
-...
-
-Risk predictions can then be obtained by combining
-\( \Lambda_1 \) and $\Lambda_2$ from the highest ranked tuple
-using a well-known formula \citep{benichou1990estimates,
-  ozenne2017riskregression}.
+We here provide explicit formulas for obtaining event-free
+survival probabilities and risk predictions given the
+state-occupation function \( F \) defined in
+equation~\eqref{eq:F-def}. These formula holds under the
+assumption of conditional independent censoring and
+positivity, as introduced in Section~\ref{sec:framework}.
+
+By equations~(\ref{eq:lambdaj}) and~(\ref{eq:gamma}) and the
+definition of \( F \), we have
+\begin{equation}
+  \label{eq:7}
+  % \Gamma(t , x) = \int_0^t \frac{F(\diff s, -1, x )}{F(s-,
+  % 0, x )}, \quad \text{and} \quad
+  \Lambda_j(t , x) 
+  = \int_0^t  \frac{F(\diff s, j, x )}{F(s-, 0, x )},
+  \quad j \in \{1,2\}.
+\end{equation}
+Equation~\eqref{eq:surv-def} provides a formula for obtaining
+the event-free survival probabilities from the cause-specific
+cumulative hazard functions. Similarly, cause-specific risk
+predictions can be obtained from \( \Lambda_1 \) and
+$\Lambda_2$ using the formula
+\citep[e.g.,][]{benichou1990estimates,
+  ozenne2017riskregression},
+\begin{equation}
+  \label{eq:cs-risk-def}
+  Q(T \leq t, D = j \mid X=x) =
+  \int_0^t \exp\left\{-\Lambda_{1}(u \mid x)-\Lambda_{2}(u
+    \mid x)\right\}  \Lambda_j(\diff u \mid x),
+  \quad j \in \{1,2\}.
+\end{equation}
+Hence, given the state learner's estimate of \( F \) we can
+use equation~\eqref{eq:7} to obtain estimates of the
+cause-specific cumulative hazard functions $\Lambda_j$, which
+can in turn be used to obtain estimates of the event-free
+survival probabilites and cause-specific risks through
+equations~\eqref{eq:surv-def} and~\eqref{eq:cs-risk-def}.
+
+In Section~\ref{sec:super-learner-simple} we suggested to
+implement the state learner by building a library using
+learners of the cause-specific cumulative hazard functions,
+$\Lambda_j$, and the cumulative hazard function for censoring,
+$\Gamma$. With this implementation we can directly input the
+highest ranked tuple of cause-specific hazard functions
+$(\Lambda_1, \Lambda_2)$ provided by the state learner as
+input to equations~\eqref{eq:surv-def}
+and~\eqref{eq:cs-risk-def}.
 
@@
 
-% BibTeX users please use one of
-% \bibliographystyle{spbasic}      % basic style, author-year citations
-%\bibliographystyle{spmpsci}      % mathematics and physical sciences
-%\bibliographystyle{spphys}       % APS-like style for physics
-
-\bibliographystyle{abbrvnat}
-\bibliography{bib.bib}   % name your BibTeX data base
+\bibliography{bib.bib}
 

### 2025-05-28 manuscript/manuscript-one.tex
rewritten intro

@@
 \usepackage[colorlinks,allcolors=linkcolor]{hyperref}
+
+\usepackage[author=]{fixme}
+\fxusetheme{color}
+\definecolor{fxtarget}{rgb}{.5,.5,.5}
+\definecolor{fxnote}{rgb}{.5,.5,.5}
+
+\fxsetup{status=draft}
 
@@
 
-[More motivation]
-
-A super learner is a machine learning algorithm that combines
-a finite set of learners into a meta learner by estimating
-prediction performance in hold-out samples using a
-pre-specified loss function \citep{van2007super}. When the aim
-is to make a prediction model, super learners combine strong
-learners, such as Cox regression models and random survival
-forests \citep[][Section 8.4]{gerds2021medical}. While the
-general idea of combining strong learners based on
-cross-validation stems from earlier work
-\citep{wolpert1992stacked,breiman1996stacked}, the name super
-learner is justified by an oracle inequality
-\citep{van2003unicv,vaart2006oracle}.
-
-We define the state learner, a new super learner for
-right-censored data, which simultaneously estimates the
-expected loss of learners of the event time distribution and
-the censoring distribution. Compared to the super learners of
-\cite{polley2011-sl-cens} and \cite{golmakani2020super}, the
-state learner proposed here poses no restrictions on the type
-of learners that can be included in the library. Compared to
-super learners that rely on inverse probability of censoring
-weighting, censoring unbiased transformations, or
-pseudo-values, the state learner does not require a
-pre-specified censoring distribution. Compared to the super
-learners by \cite{han2021inverse} and
-\cite{westling2021inference}, the state learner has
-theoretical guarantees which these methods do not have.
-
-The loss function underlying the state learner operates on
-coarsened data. In the right-censored survival setting the
-coarsened data consist of the minimum and the order of the
-censoring time and the event time as well as the baseline
-covariates. The state learner works by simultaneously
-assessing how well learners of the event time distribution and
-the censoring distribution predict the states of the coarsened
-data. To analyse the theoretical properties of the state
-learner we focus on the discrete super learner which combines
-the library of learners by picking the one that minimises the
-cross-validated loss \citep{van2007super}. In the presence of
-competing risks, our algorithm uses separate libraries of
-learners for the cumulative hazard functions for each of the
-competing risks and for the censoring distribution. We show
-that the oracle selector of the state learner is consistent if
-all libraries contain a consistent learner and prove a finite
-sample oracle inequality.
-
-The state learner algorithm can output a medical risk
-prediction model \citep{gerds2021medical} which predicts the
-probability of an event based on covariates in the presence of
-competing risks. Another important application, which
-motivated our contribution, is in targeted learning where
-conditional event probabilities occur as high-dimensional
-nuisance parameters which need to be estimated at a certain
-rate \citep{van2011targeted, rytgaard2021estimation,
-  rytgaard2022targeted}. The state learner is particularly
-well-suited for this application as it simultaneously provides
-estimates for both the outcome and the censoring, both of
-which are needed for targeted learning. For the asymptotic
-bias term of a targeted estimator, which uses the state
-learner to estimate nuisance parameters, we show that a second
-order product structure holds. We illustrate both applications
-of the state learner with prostate cancer data
-\citep{kattan2000pretreatment}.
+Accurate risk prediction from time-to-event data is a central
+topic in many areas of medical research and public health.
+Super learning \citep{van2007super}, also known as ensemble
+learning or stacked regression
+\citep{wolpert1992stacked,breiman1996stacked}, provides a
+powerful approach by combining multiple candidate prediction
+models. In the context of survival analysis, a super learner
+can for instance combine Cox regression models and random
+survival forests \citep[][Section 8.4]{gerds2021medical}.
+
+In this paper we propose the state learner, a new super
+learner designed to handle the specific challenges of ensemble
+learning with right-censored data. The state learner estimates
+the performance of a library of candidate models for both the
+event-time and censoring distributions simultaneously. This
+joint estimation framework avoids the need for pre-specifying
+a censoring model, which is a limitation of many existing
+methods. Our approach also allows for full flexibility in the
+choice of models in the library, contrasting with earlier
+proposals that restrict the library to specific model classes
+\citep{polley2011-sl-cens,golmakani2020super}. Furthermore,
+unlike iterative approaches such as those in
+\cite{han2021inverse} and \cite{westling2021inference}, we can
+provide theoretical guarantees for the performance of the
+state learner.
+
+The state learner is based on the simple idea of considering
+right-censored competing risks data as an artificial
+multi-state system where censoring is included as a state of
+its own. Candidate models are then assessed based on how well
+they predict the state occupations of this system across time,
+based on baseline covariates. We demonstrate how to construct
+a library for the state learner using existing survival
+models, and how to obtain risk predictions from the resulting
+ensemble. To analyse the theoretical properties of the state
+learner we focus on the discrete super learner which pick the
+model in the library with best estimated performance
+\citep{van2007super}. We show that the discrete state learner
+is consistent under mild conditions and prove a finite-sample
+oracle inequality.
+
+\subsection{Related work}
+\label{sec:related-work}
 
 There are many existing super learners for right-censored
-data. We briefly review them here. Machine learning based on
-right-censored data commonly uses the partial log-likelihood
-as a loss function
+data. Machine learning based on right-censored data commonly
+uses the partial log-likelihood as a loss function
 \citep[e.g.,][]{li2016regularized,yao2017deep,lee2018deephit,katzman2018deepsurv,gensheimer2019scalable,lee2021boosted,kvamme2021continuous}.
@@
 additive hazards models, and accelerated failure time models.
-Other proposals for super learning with right-censored data
-use an inverse probability of censoring weighted (IPCW) loss
-function
+
+Alternative approaches for super learning with right-censored
+data use an inverse probability of censoring weighted (IPCW)
+loss function
 \citep{graf1999assessment,van2003unicv,molinaro2004tree,keles2004asymptotically,hothorn2006survival,gerds2006consistent,gonzalez2021stacked},
@@
 distribution, and their drawback is that this estimator has to
-be pre-specified. An approach which avoids a pre-specified
-censoring model was proposed independently by
-\cite{han2021inverse} and \cite{westling2021inference}. In
-both articles, the authors suggest to iterate between learning
-of the outcome model and learning of the censoring model using
-IPCW loss functions. However, it seems that this procedure is
-in general not guaranteed to converge to the true
-data-generating mechanism
+be pre-specified.
+
+More recent work by \cite{han2021inverse} and
+\cite{westling2021inference} circumvents the need to
+pre-specify a censoring model by iterating
+between estimation of the outcome and censoring models. However, it seems that this procedure is in general not
+guaranteed to converge to the true data-generating mechanism
 \citep[][Appendix~A.4]{munch2024thesis}.
 
+\subsection{Outline of the paper}
+\label{sec:outline-paper}
+
 We introduce our notation and framework in
-Section~\ref{sec:framework}. In
-Section~\ref{sec:super-learning} we define general super
-learning for right-censored data.
-Section~\ref{sec:super-learner-simple} introduces the state
-learner, and Section~\ref{sec:theor-results-prop} provides
-theoretical guarantees. We report results of our numerical
-experiments in Section~\ref{sec:numer-exper} and analyse a
-prostate cancer data set in Section~\ref{sec:real-data-appl}.
-Section~\ref{sec:discussion} contains a discussion of the
-merits and limitations of our proposal. Proofs are given in
-Appendix~\ref{sec:proofs}. An implementation of the state
-learner is available at
-\url{https://github.com/amnudn/statelearner} along with code
-for reproducing our numerical experiments.
+Section~\ref{sec:framework}. Section~\ref{sec:super-learning}
+defines general super learning for right-censored data.
+Section~\ref{sec:super-learner-simple} presents the state
+learner. In Section~\ref{sec:use-cases-state}, we demonstrate
+how the state learner can be used to obtain risk predictions
+and how it can be applied in the context of causal inference.
+Section~\ref{sec:theor-results-prop} provides theoretical
+guarantees. Section~\ref{sec:numer-exper} reports the results
+of numerical experiments, and Section~\ref{sec:real-data-appl}
+illustrates the method on prostate cancer data. We conclude
+with a discussion in Section~\ref{sec:discussion}. Proofs are
+included in Appendix~\ref{sec:proofs}. Code and an
+implementation of the state learner are available at
+\url{https://github.com/amnudn/statelearner}.
+
+% A super learner is a machine learning algorithm that combines
+% a finite set of learners into a meta learner by estimating
+% prediction performance in hold-out samples using a
+% pre-specified loss function \citep{van2007super}. When the aim
+% is to make a prediction model, super learners combine strong
+% learners, such as Cox regression models and random survival
+% forests \citep[][Section 8.4]{gerds2021medical}. While the
+% general idea of combining strong learners based on
+% cross-validation stems from earlier work
+% \citep{wolpert1992stacked,breiman1996stacked}, the name super
+% learner is justified by an oracle inequality
+% \citep{van2003unicv,vaart2006oracle}.
+
+% We define the state learner, a new super learner for
+% right-censored data, which simultaneously estimates the
+% expected loss of learners of the event time distribution and
+% the censoring distribution. Compared to the super learners of
+% \cite{polley2011-sl-cens} and \cite{golmakani2020super}, the
+% state learner proposed here poses no restrictions on the type
+% of learners that can be included in the library. Compared to
+% super learners that rely on inverse probability of censoring
+% weighting, censoring unbiased transformations, or
+% pseudo-values, the state learner does not require a
+% pre-specified censoring distribution. Compared to the super
+% learners by \cite{han2021inverse} and
+% \cite{westling2021inference}, the state learner has
+% theoretical guarantees which these methods do not have.
+
+% The loss function underlying the state learner operates on
+% coarsened data. In the right-censored survival setting the
+% coarsened data consist of the minimum and the order of the
+% censoring time and the event time as well as the baseline
+% covariates. The state learner works by simultaneously
+% assessing how well learners of the event time distribution and
+% the censoring distribution predict the states of the coarsened
+% data. To analyse the theoretical properties of the state
+% learner we focus on the discrete super learner which combines
+% the library of learners by picking the one that minimises the
+% cross-validated loss \citep{van2007super}. In the presence of
+% competing risks, our algorithm uses separate libraries of
+% learners for the cumulative hazard functions for each of the
+% competing risks and for the censoring distribution. We show
+% that the oracle selector of the state learner is consistent if
+% all libraries contain a consistent learner and prove a finite
+% sample oracle inequality.
+
+% The state learner algorithm can output a medical risk
+% prediction model \citep{gerds2021medical} which predicts the
+% probability of an event based on covariates in the presence of
+% competing risks. Another important application, which
+% motivated our contribution, is in targeted learning where
+% conditional event probabilities occur as high-dimensional
+% nuisance parameters which need to be estimated at a certain
+% rate \citep{van2011targeted, rytgaard2021estimation,
+%   rytgaard2022targeted}. The state learner is particularly
+% well-suited for this application as it simultaneously provides
+% estimates for both the outcome and the censoring, both of
+% which are needed for targeted learning. For the asymptotic
+% bias term of a targeted estimator, which uses the state
+% learner to estimate nuisance parameters, we show that a second
+% order product structure holds. We illustrate both applications
+% of the state learner with prostate cancer data
+% \citep{kattan2000pretreatment}.
 
@@
 \( \data_n=\{O_i\}_{i=1}^n \) of i.i.d.\ observations from
-\( P \in \mathcal{P} \) and a collection of candidate learners
-$\mathcal{A}$. Each learner \(a \in \mathcal{A}\) is a map
-\( a \colon \mathcal{O}^n \rightarrow \Theta \) which takes a
-data set as input and returns an estimate
+\( P \in \mathcal{P} \) and a collection of candidate models
+or learners $\mathcal{A}$. Each learner \(a \in \mathcal{A}\)
+is a map \( a \colon \mathcal{O}^n \rightarrow \Theta \) which
+takes a data set as input and returns an estimate
 $a(\data_n) \in \Theta$ of $\theta$.
@@
 \( F \), as provided by the state learner, can be used to
-estimate more relevant parameters of interest, such as
+estimate more relevant parameters of interest, such as, e.g.,
 survival probabilities.
@@
 
-We here provide explicit formulas for obtaining event-free
-survival probabilities and risk predictions given the
-state-occupation function \( F \) defined in
-equation~\eqref{eq:F-def}. These formula holds under the
+Event-free survival probabilities and risk predictions can be
+obtained from the state-occupation function \( F \) under the
 assumption of conditional independent censoring and
-positivity, as introduced in Section~\ref{sec:framework}.
-
-By equations~(\ref{eq:lambdaj}) and~(\ref{eq:gamma}) and the
+positivity, as introduced in Section~\ref{sec:framework}. By
+equations~(\ref{eq:lambdaj}) and~(\ref{eq:gamma}) and the
 definition of \( F \), we have
@@
 Equation~\eqref{eq:surv-def} provides a formula for obtaining
-the event-free survival probabilities from the cause-specific
+event-free survival probabilities from the cause-specific
 cumulative hazard functions. Similarly, cause-specific risk
@@
 
-... TL + SL = strong... The state learner is particularly well
-suited for targeted and debiased machine learning as it
-immediately provides estimates of the nuisance parameters
-pertaining to the outcome and the censoring models. While it
-was our original motivation for developing the state learning,
-we leave the study of the state learner in the context of
+The framework of \textit{targeted} or \textit{debiased}
+machine learning is a general methodology for combining
+data-adaptive modeling, such as machine and super learning,
+with asymptotically valid statistical inference for
+interpretable target parameters \citep{van2006targeted,
+  chernozhukov2018double}. \fxnote*{add more references?}{This
+  framework has been applied extensively for causal inference
+  based on observational data
+  \citep[e.g.,][]{van2011targeted,kennedy2016semiparametric,rytgaard2022targeted}.}
+A targeted or debiased estimator is constructed by first
+estimation high-dimensional nuisance parameter using machine
+learning, and then employing a debiasing or targeting step
+using tools from semi-parametric efficiency theory
+\citep{pfanzagl1985contributions,bickel1993efficient,van2003unified,tsiatis2007semiparametric,kennedy2022semiparametric}.
+In the context of right-censored competing risk data, the
+high-dimensional nuisance parameters needed to be estimated
+will typically include the cause-specific cumulative hazard
+functions and the cumulative hazard function for censoring
+\citep[e.g.,][]{van2003unified,rytgaard2021estimation,rytgaard2022targeted}.
+The state learner is hence particularly well suited for
+targeted and debiased machine learning as it immediately
+provides estimates of these nuisance parameters. While it was
+our original motivation for developing the state learning, we
+leave the study of the state learner in the context of
 targeted and debiased machine learning for a future paper.
@@
 
-For all super learners we use a library consisting of three learners:
-The Kaplan-Meier estimator
-\citep{kaplan1958nonparametric,Gerds_2019prodlim}, a Cox model with
-main effects \citep{cox1972regression, survival-package}, and a random
-survival forest \citep{ishwaran2008random,rfsrc-paclage}. We use the
-same library to learn the outcome distribution and the censoring
-distribution. Note that the three learners in our library of learners
-can be used to learn the cumulative hazard functions of the 
-outcome and the censoring distribution. The latter works by
-training the learner on the data set \( \data_n^c \), where
-\( \data_n^c = \{O_i^c\}_{i=1}^n \) with
-\( O_i^c = (\tilde{T}_i, 1-\Delta, X_i) \). When we say that we use a
-learner for the cumulative hazard function of the outcome to learn the
-cumulative hazard function of the censoring time, we mean that the
-learner is trained on \( \data_n^c \).
+For all super learners we use a library consisting of three
+learners: The Kaplan-Meier estimator
+\citep{kaplan1958nonparametric,Gerds_2019prodlim}, a Cox model
+with main effects \citep{cox1972regression, survival-package},
+and a random survival forest
+\citep{ishwaran2008random,rfsrc-paclage}. We use the same
+library to learn the outcome distribution and the censoring
+distribution. Note that the three learners in our library of
+learners can be used to learn the cumulative hazard functions
+of the outcome and the censoring distribution. The latter
+works by training the learner on the data set \( \data_n^c \),
+where \( \data_n^c = \{O_i^c\}_{i=1}^n \) with
+\( O_i^c = (\tilde{T}_i, 1-\Delta_i, X_i) \). When we say that
+we use a learner for the cumulative hazard function of the
+outcome to learn the cumulative hazard function of the
+censoring time, we mean that the learner is trained on
+\( \data_n^c \).
 

### 2025-05-28 manuscript/manuscript-one.tex
rephrasing

@@
 
-Accurate risk prediction from time-to-event data is a central
-topic in many areas of medical research and public health.
-Super learning \citep{van2007super}, also known as ensemble
-learning or stacked regression
+Accurate risk prediction from time-to-event data is a
+important problem in many areas of medical research and public
+health. Super learning \citep{van2007super}, also known as
+ensemble learning or stacked regression
 \citep{wolpert1992stacked,breiman1996stacked}, provides a
-powerful approach by combining multiple candidate prediction
-models. In the context of survival analysis, a super learner
-can for instance combine Cox regression models and random
-survival forests \citep[][Section 8.4]{gerds2021medical}.
+powerful approach to this problem by combining multiple
+candidate prediction models. In the context of survival
+analysis, a super learner can for instance combine Cox
+regression models and random survival forests \citep[][Section
+8.4]{gerds2021medical}.
 
@@
 they predict the state occupations of this system across time,
-based on baseline covariates. We demonstrate how to construct
+based on baseline covariate information. We demonstrate how to construct
 a library for the state learner using existing survival
@@
 ensemble. To analyse the theoretical properties of the state
-learner we focus on the discrete super learner which pick the
+learner we focus on the discrete super learner which picks the
 model in the library with best estimated performance
 \citep{van2007super}. We show that the discrete state learner
-is consistent under mild conditions and prove a finite-sample
+is consistent under natural conditions and prove a finite-sample
 oracle inequality.
@@
 
-There are many existing super learners for right-censored
-data. Machine learning based on right-censored data commonly
-uses the partial log-likelihood as a loss function
+Super learning requires a library of candidate models, or
+\textit{learners}, and a loss function for evaluating
+performance on hold-out samples. Machine learning based on
+right-censored data commonly uses the partial log-likelihood
+as a loss function
 \citep[e.g.,][]{li2016regularized,yao2017deep,lee2018deephit,katzman2018deepsurv,gensheimer2019scalable,lee2021boosted,kvamme2021continuous}.
@@
 is assumed to be observed in discrete time. However, the
-partial log-likelihood loss does not work well with data
-splitting (cross-validation) in continuous time. The reason is
-that the partial log-likelihood loss assigns an infinite value
-for any learner that predicts piece-wise constant cumulative
-hazard functions when there are event times in the test set
-which do not occur in the learning set. This problem occurs
-with prominent survival learners including the Kaplan-Meier
-estimator, the random survival forest, and the semi-parametric
-Cox regression model and these learners cannot be included in
-the library of the super learner proposed by
-\cite{polley2011-sl-cens}. When a proportional hazards model
-is assumed, the baseline hazard function can be profiled out
-of the likelihood \citep{cox1972regression}. The
+partial log-likelihood loss does not work well as a measure of
+performance in hold-out samples observed in continuous time.
+The reason is that the partial log-likelihood loss assigns an
+infinite value for any learner that predicts piece-wise
+constant cumulative hazard functions when there are event
+times in the test set which do not occur in the learning set.
+This problem occurs with prominent survival learners including
+the Kaplan-Meier estimator, random survival forests, and
+semi-parametric Cox regression models and these learners
+cannot be included in the library of the super learner
+proposed by \cite{polley2011-sl-cens}. When a proportional
+hazards model is assumed, the baseline hazard function can be
+profiled out of the likelihood \citep{cox1972regression}. The
 cross-validated partial log-likelihood loss
@@
 Section~\ref{sec:framework}. Section~\ref{sec:super-learning}
-defines general super learning for right-censored data.
-Section~\ref{sec:super-learner-simple} presents the state
-learner. In Section~\ref{sec:use-cases-state}, we demonstrate
-how the state learner can be used to obtain risk predictions
-and how it can be applied in the context of causal inference.
-Section~\ref{sec:theor-results-prop} provides theoretical
-guarantees. Section~\ref{sec:numer-exper} reports the results
-of numerical experiments, and Section~\ref{sec:real-data-appl}
-illustrates the method on prostate cancer data. We conclude
-with a discussion in Section~\ref{sec:discussion}. Proofs are
-included in Appendix~\ref{sec:proofs}. Code and an
-implementation of the state learner are available at
+defines general super learning for right-censored data. In
+Section~\ref{sec:super-learner-simple} we define the state
+learner, while Section~\ref{sec:theor-results-prop} provides
+theoretical guarantees. In Section~\ref{sec:use-cases-state}
+we demonstrate how the state learner can be used to obtain
+risk predictions and how it can be applied in the context of
+causal inference. Section~\ref{sec:numer-exper} reports the
+results of numerical experiments, and
+Section~\ref{sec:real-data-appl} illustrates the method on
+prostate cancer data. We conclude with a discussion in
+Section~\ref{sec:discussion}. Proofs are collected in
+Appendix~\ref{sec:proofs}. Code and an implementation of the
+state learner are available at
 \url{https://github.com/amnudn/statelearner}.
@@
 
-In a competing risk framework \citep{andersen2012statistical}, let
-\( T\) be a time to event variable, \(D\in\{1,2\}\) the cause of the
-event, and $X \in \mathcal{X}$ a vector of baseline covariates taking
-values in a bounded subset \( \mathcal{X} \subset \R^p \),
-\( p\in\N \). Let $\tau< \infty$ be the prediction horizon. We use
-\( \mathcal{Q} \) to denote the collection of all probability measures
-on \( [0,\tau] \times \{1,2\}\times \mathcal{X} \) such that
-\( (T, D, X) \sim Q \) for some unknown \( Q \in \mathcal{Q} \). For
-\(j\in\{1,2\}\), the cause-specific conditional cumulative hazard
-functions are defined by
-\( \Lambda_{j} \colon [0, \tau] \times \mathcal{X} \rightarrow \R_+ \)
-such that
+In a competing risk framework \citep{andersen2012statistical},
+let \( T\) be a time to event variable, \(D\in\{1,2\}\) the
+cause of the event, and $X \in \mathcal{X}$ a vector of
+baseline covariates taking values in a bounded subset
+\( \mathcal{X} \subset \R^p \), \( p\in\N \). Let
+$\tau< \infty$ be the prediction horizon. We use
+\( \mathcal{Q} \) to denote the collection of all probability
+measures on \( [0,\tau] \times \{1,2\}\times \mathcal{X} \)
+such that \( (T, D, X) \sim Q \) for some unknown
+\( Q \in \mathcal{Q} \). For \(j\in\{1,2\}\), the
+cause-specific conditional cumulative hazard functions
+\( \Lambda_{j} \colon [0, \tau] \times \mathcal{X} \rightarrow
+\R_+ \) are defined as
 \begin{equation*}
@@
 For ease of presentation we assume throughout that the map
-\( t\mapsto \Lambda_j(t \mid x) \) is continuous for all \( x \) and
-\( j \). This is not a limitation: All arguments carry over directly
-to the general case. We denote by \(S\) the conditional event-free
-survival function:
+\( t\mapsto \Lambda_j(t \mid x) \) is continuous for all
+\( x \) and \( j \). All arguments carry over directly to the
+general case. We denote by \(S\) the conditional event-free
+survival function,
 \begin{equation}
@@
 
-We consider the right-censored setting in which we observe the coarsened data
-\(O = (\tilde{T},\tilde D, X)\), where $\tilde T = \min(T,C)$ for a
-right-censoring time \(C\), $\Delta = \1{\{T \leq C\}}$, and
-\(\tilde D=\Delta D\). Let \(\mathcal{P}\) denote a set of probability
-measures on the sample space
-\(\mathcal{O} = [0, \tau] \times \{0, 1, 2\} \times \mathcal{X}\) such
-that \(O \sim P \) for some unknown \(P\in \mathcal{P}\). We assume
-that the event times and the censoring times are conditionally
-independent given covariates, \( T \independent C \mid X \). This
-implies that any distribution \( P \in \mathcal{P} \) is characterised
-by a distribution \( Q \in \mathcal{Q} \) and a conditional cumulative
-hazard function for \( C \) given \( X \)
-\citep[c.f.,][]{begun1983information,gill1997coarsening}. We use
-\(\Gamma\in\mathcal{M}_{\tau}\) to denote the conditional cumulative hazard
-function for censoring. For ease of presentation we now also assume
-that \( \Gamma(\blank \mid x) \) is continuous for all \( x \). We let
-\((t,x)\mapsto G(t \mid x)=\exp\left\{-\Gamma(t \mid x)\right\}\) denote the
-survival function of the conditional censoring distribution. The
-distribution \( P \) is characterised by
+We consider the right-censored setting in which we observe
+\(O = (\tilde{T},\tilde D, X)\), where $\tilde T = \min(T,C)$
+for a right-censoring time \(C\), $\Delta = \1{\{T \leq C\}}$,
+and \(\tilde D=\Delta D\). Let \(\mathcal{P}\) denote a set of
+probability measures on the sample space
+\(\mathcal{O} = [0, \tau] \times \{0, 1, 2\} \times
+\mathcal{X}\) such that \(O \sim P \) for some unknown
+\(P\in \mathcal{P}\). We assume that the event times and the
+censoring times are conditionally independent given
+covariates, \( T \independent C \mid X \). This implies that
+any distribution \( P \in \mathcal{P} \) is characterised by a
+distribution \( Q \in \mathcal{Q} \) and a conditional
+cumulative hazard function for \( C \) given \( X \)
+\citep[c.f.,][]{begun1983information,gill1997coarsening}. We
+use \(\Gamma\in\mathcal{M}_{\tau}\) to denote the conditional
+cumulative hazard function for censoring. For ease of
+presentation we now also assume that
+\( \Gamma(\blank \mid x) \) is continuous for all \( x \). We
+let
+\((t,x)\mapsto G(t \mid x)=\exp\left\{-\Gamma(t \mid
+  x)\right\}\) denote the survival function of the conditional
+censoring distribution. The distribution \( P \) is
+characterised by
 \begin{equation}\label{eq:parametrizeP}
@@
 jointly model the observed data. To formally introduce the
-state learner, we define the multi-state process
+state learner, we define the artificial multi-state process
 \begin{equation*}
@@
 \end{equation*}
+We use the phrase `artificial' because multi-state models
+should not usually include censoring as a state.
 At time \(t\), we observe that each individual is in one of four mutually
@@
 \( x \in \mathcal{X} \). The function \( F \) describes the
-conditional state occupation probabilities of the multi-state
-process \(\eta\). We construct a super learner for \( F \).
-The target of this super learner is the function-valued
-parameter $\theta(P) = F$ which is identified through
-equation~(\ref{eq:F-def}). Under conditional independent
-censoring each tuple $(\Lambda_{1}, \Lambda_{2}, \Gamma, H)$
-characterises a distribution \(P\in\mathcal P\), c.f.\
+conditional state occupation probabilities of the process
+\(\eta\).
+
+We define the state learner as a super learner for the
+function-valued parameter $\theta(P) = F$ which is identified
+through equation~(\ref{eq:F-def}). Under conditional
+independent censoring each tuple
+$(\Lambda_{1}, \Lambda_{2}, \Gamma, H)$ characterises a
+distribution \(P\in\mathcal P\), c.f.\
 equation~\eqref{eq:parametrizeP}, which in turn determines
@@
 \end{equation}
-The state learner requires three libraries of learners:
+Equation~\eqref{eq:transition} implies that a library for the
+state learner can by build from three libraries of learners:
 \(\mathcal{A}_1\), \( \mathcal{A}_2 \), and \( \mathcal{B} \),
@@
 \(\mathcal{B}\) contains learners for the conditional
-cumulative hazard function of the censoring
-distribution. % We further
-% define \(\mathcal{H}\) as the set of all probability distributions on
-% \( \mathcal{X} \) and
-% \begin{equation*}
-%   \mathcal{H}_n=\left\{h\colon\mathcal{O}^n\longrightarrow\mathcal{H} \midd
-%     h(\data_n)=H_n = \frac 1 n \sum_{i=1}^n \delta_{X_i} \right\}
-% \end{equation*}
-% as the library of learners which consists of a single learner, the empirical
-% distribution function.
-Based on the Cartesian product of
-libraries of learners for \((\Lambda_1,\Lambda_2,\Gamma)\) we construct a library
-$\mathcal{F}$ of learners
-for \( F \):
+cumulative hazard function of the censoring distribution.
+Based on the Cartesian product of libraries of learners for
+\((\Lambda_1,\Lambda_2,\Gamma)\) we construct a library
+$\mathcal{F}$ of learners for \( F \):
 \begin{align*}
@@
 \end{align*}
-To evaluate how well a function \( F \) predicts the observed
-multi-state process we use the integrated Brier score
-\( \bar B_\tau( F,O) = \int_0^{\tau} B_t(F,O) \diff t \), where \( B_t \) is the
-Brier score \citep{brier1950verification} at time \( t \in [0, \tau] \),
+Importantly, the libraries \( \mathcal{A}_1 \),
+\( \mathcal{A}_2 \), and \( \mathcal{B} \) can be constructed
+using standard survival analysis software.
+
+To evaluate how well a function \( F \) predicts an observed
+process $\eta$ we use the integrated Brier score
+\( \bar B_\tau( F,O) = \int_0^{\tau} B_t(F,O) \diff t \),
+where \( B_t \) is the Brier score
+\citep{brier1950verification} at time \( t \in [0, \tau] \),
 \begin{equation*}
@@
 \end{align*}
+
+\section{Theoretical results for the state learner}
+\label{sec:theor-results-prop}
+
+In this section we establish theoretical guarantees for the
+state learner. We show that the state learner is consistent if
+its library contains a consistent learner, and we establish a
+finite sample inequality for the excess risk of the state
+learner compared to the oracle.
+
+\subsection{Consistency}
+\label{sec:consistency}
+
+% We let \( \E_P \) denote
+% expectation under \( P \) and use \( \Theta \) to denote the
+% collection of all conditional state-occupation probability functions.
+The following result can be derived from the fact
+that the integrated Brier score (also called the continuous
+ranked probability score) is a strictly proper scoring rule
+\citep{gneiting2007strictly}. This implies that if we minimise
+the average loss of the integrated Brier score, we recover the
+parameters of the data-generating distribution. Specifically,
+this implies that the oracle of a state learner is consistent
+if the library of learners contains at least one learner that
+is consistent for estimation of \( F \). Recall that the
+function \(F\) implicitly depends on the data-generating
+probability measure \(P\in\mathcal P\) but that this was
+suppressed in the notation. We now make this dependence
+explicit by writing \(F_P\) for the function determined by a
+given \(P \in\mathcal{P}\) in accordance with equation
+equation~(\ref{eq:F-def}). In the following we let
+\( \mathcal{H}_{\mathcal{P}} = \{F_P : P \in \mathcal{P}\} \).
+
+\begin{proposition}
+  \label{prop:stric-prop}
+  If \(P \in\mathcal{P}\) then
+  \begin{equation*}
+    F_P = \argmin_{F \in \mathcal{H}_{\mathcal{P}}} P{[\bar{B}_\tau(F, \blank)]},
+  \end{equation*}
+  \( P \)-almost surely.
+\end{proposition}
+\begin{proof}
+  See Appendix~\ref{sec:consistency-proof}.
+\end{proof}
+
+\subsection{Oracle inequalities}
+\label{sec:finite-sample-oracle}
+
+We establish a finite sample oracle result for the state
+learner. Our Proposition~\ref{prop:oracle-prop} is in essence a
+special case of Theorem 2.3 in \citep{vaart2006oracle}. We
+assume that we split the data into equally sized folds, and
+for simplicity of presentation we take \( n \) to be such that
+\( |\data_n^{-k}| = n/K \) with \( K \) fixed. We will allow
+the number of learners to grow with \( n \) and write
+\( \mathcal{F}_n=\mathcal{F}(\mathcal{A}_{1,n},
+\mathcal{A}_{2,n}, \mathcal{B}_n)\) as short-hand notation and
+to emphasise the dependence on \( n \). In the following we
+let the space \( \mathcal{H}_{\mathcal{P}} \) be equipped with
+the norm \( \| \blank \|_{P} \) defined as
+\begin{equation}
+  \label{eq:norm}
+  \| F \|_{P} = 
+  \left\{
+    \sum_{j=-1}^{2}\int_{\mathcal{X}} \int_0^{\tau} F(t, j, x)^2 \diff t P( \diff x)
+  \right\}^{1/2}.
+\end{equation}
+
+\begin{proposition}
+  \label{prop:oracle-prop}
+  For all \(P\in\mathcal{P}\), \( n \in \N \), \( k \in \{1, \dots, K\} \),
+  and $\delta>0$,
+  \begin{align*}
+    \E_{P}{\left[ \Vert \hat{\phi}_n(\data_n^{-k}) - F_P \Vert_{P}^2 \right]}
+    & \leq (1 + 2\delta)
+      \E_{P}{\left[ \Vert \tilde{\phi}_n(\data_n^{-k}) - F_P \Vert_{P}^2 \right]}
+    \\
+    & \quad
+      + (1+ \delta) 16   K \tau
+      \left(
+      13 + \frac{12}{\delta}
+      \right)
+      \frac{\log(1 + |\mathcal{F}_n|)}{n}.
+  \end{align*}
+\end{proposition}
+\begin{proof}
+  See Appendix~\ref{sec:oracle-inequalities}.
+\end{proof}
+
+Proposition~\ref{prop:oracle-prop} has the following asymptotic consequences.
+
+\begin{corollary}
+  \label{cor:asymp-cons}
+  Assume that \( |\mathcal{F}_n| = \bigO(n^q)\), for some \( q \in \N \) and
+  that there exists a sequence \( \phi_n \in \mathcal{F}_n \), \( n \in \N \),
+  such that
+  \( \E_{P}{\left[ \Vert \phi_n(\data_n^{-k}) - F_P \Vert_{P}^2 \right]} =
+  \bigO(n^{-\alpha}) \), for some \( \alpha\leq 1 \).
+  \begin{enumerate}[label=(\alph*)]
+  \item If $\alpha=1$ then
+    \( \E_{P}{\left[ \Vert \hat{\phi}_n(\data_n^{-k}) - F_P \Vert_{P}^2
+      \right]} = \bigO(\log(n)n^{-1}) \).
+  \item If $\alpha<1$ then
+    \( \E_{P}{\left[ \Vert \hat{\phi}_n(\data_n^{-k}) - F_P \Vert_{P}^2 \right]} =
+    \bigO(n^{-\alpha}) \).
+  \end{enumerate}
+\end{corollary}
+\begin{proof}
+  See Appendix~\ref{sec:oracle-inequalities}.
+\end{proof}
+
+Proposition~\ref{prop:oracle-prop} provides a precise bound on
+the average price we pay for using cross-validation to select
+a learner from a library of candidates as compared to knowing
+the oracle learner in advance. Corollary~\ref{cor:asymp-cons}
+states that this price is asymptotically vanishing as long as
+the number of learners in the library grows with sample size
+at a polynomial rate.
 
@@
 A targeted or debiased estimator is constructed by first
-estimation high-dimensional nuisance parameter using machine
-learning, and then employing a debiasing or targeting step
+estimation high-dimensional nuisance parameters using machine
+learning, and then employing a debiasing or targeting step by
 using tools from semi-parametric efficiency theory
@@
 targeted and debiased machine learning as it immediately
-provides estimates of these nuisance parameters. While it was
-our original motivation for developing the state learning, we
-leave the study of the state learner in the context of
-targeted and debiased machine learning for a future paper.
-
-\section{Theoretical results for the state learner}
-\label{sec:theor-results-prop}
-
-In this section we establish theoretical guarantees for the
-state learner. We show that the state learner is consistent if
-its library contains a consistent learner, and we establish a
-finite sample inequality for the excess risk of the state
-learner compared to the oracle.
-
-\subsection{Consistency}
-\label{sec:consistency}
-
-% We let \( \E_P \) denote
-% expectation under \( P \) and use \( \Theta \) to denote the
-% collection of all conditional state-occupation probability functions.
-Proposition~\ref{prop:stric-prop} can be derived from the fact that the
-integrated Brier score (also called the continuous ranked probability score) is
-a strictly proper scoring rule \citep{gneiting2007strictly}. This implies that
-if we minimise the average loss of the integrated Brier score, we recover the
-parameters of the data-generating distribution. Specifically, this implies that
-the oracle of a state learner is consistent if the library of learners contains
-at least one learner that is consistent for estimation of \( F \). Recall that
-the function \(F\) implicitly depends on the data-generating probability measure
-\(P\in\mathcal P\) but that this was suppressed in the notation. We now make
-this dependence explicit by writing \(F_0\) for the function which is obtained
-by substituting a specific \(P_0\in\mathcal{P}\) for \(P\) in equation
-\eqref{eq:transition}. In the following we let
-\( \mathcal{H}_{\mathcal{P}} = \{F_P : P \in \mathcal{P}\} \) where \( F_P \) is defined as in
-equation~(\ref{eq:F-def}) using the measure \( P \in \mathcal{P} \).
-
-\begin{proposition}
-  \label{prop:stric-prop}
-  If \(P_0\in\mathcal{P}\) then
-  \begin{equation*}
-    F_0 = \argmin_{F \in \mathcal{H}_{\mathcal{P}}} P_0{[\bar{B}_\tau(F, \blank)]},
-  \end{equation*}
-  \( H \)-almost surely for any \( j\in \{-1,0,1,2\} \) and almost any
-  \( t \in [0, \tau]\).
-\end{proposition}
-\begin{proof}
-  See Appendix~\ref{sec:consistency-proof}.
-\end{proof}
-
-\subsection{Oracle inequalities}
-\label{sec:finite-sample-oracle}
-
-We establish a finite sample oracle result for the state learner. Our
-Corollary~\ref{cor:oracle-prop} is in essence a special case of
-Theorem 2.3 in \citep{vaart2006oracle}.  We assume that we split the
-data into equally sized folds, and for simplicity of presentation we
-take \( n \) to be such that \( |\data_n^{-k}| = n/K \) with \( K \)
-fixed. We will allow the number of learners to grow with \( n \) and
-write
-\( \mathcal{F}_n=\mathcal{F}(\mathcal{A}_{1,n}, \mathcal{A}_{2,n},
-\mathcal{B}_n)\) as short-hand notation and to emphasise the
-dependence on \( n \).
-% The discrete super $\hat{\phi}_n$ and the
-% oracle $\tilde{\phi}_n$ are defined as in Section~\ref{sec:framework}
-% but now with respect to the library \( \mathcal{F}_n \) and the loss
-% \( B \).
-In the following we let the space \( \mathcal{H}_{\mathcal{P}} \) be equipped with the norm
-\( \| \blank \|_{P_0} \) defined as
-\begin{equation}
-  \label{eq:norm}
-  \| F \|_{P_0} = 
-  \left\{
-    \sum_{j=-1}^{2}\int_{\mathcal{X}} \int_0^{\tau} F(t, j, x)^2 \diff t H_0( \diff x)
-  \right\}^{1/2}.
-\end{equation}
-
-\begin{corollary}
-  \label{cor:oracle-prop}
-  For all \(P_0\in\mathcal{P}\), \( n \in \N \), \( k \in \{1, \dots, K\} \),
-  and $\delta>0$,
-  \begin{align*}
-    \E_{P_0}{\left[ \Vert \hat{\phi}_n(\data_n^{-k}) - F_0 \Vert_{P_0}^2 \right]}
-    & \leq (1 + 2\delta)
-      \E_{P_0}{\left[ \Vert \tilde{\phi}_n(\data_n^{-k}) - F_0 \Vert_{P_0}^2 \right]}
-    \\
-    & \quad
-      + (1+ \delta) 16   K \tau
-      \left(
-      13 + \frac{12}{\delta}
-      \right)
-      \frac{\log(1 + |\mathcal{F}_n|)}{n}.
-  \end{align*}
-\end{corollary}
-\begin{proof}
-  See Appendix~\ref{sec:oracle-inequalities}.
-\end{proof}
-
-Corollary~\ref{cor:oracle-prop} has the following asymptotic consequences.
-
-\begin{corollary}
-  \label{cor:asymp-cons}
-  Assume that \( |\mathcal{F}_n| = \bigO(n^q)\), for some \( q \in \N \) and
-  that there exists a sequence \( \phi_n \in \mathcal{F}_n \), \( n \in \N \),
-  such that
-  \( \E_{P_0}{\left[ \Vert \phi_n(\data_n^{-k}) - F_{0} \Vert_{P_0}^2 \right]} =
-  \bigO(n^{-\alpha}) \), for some \( \alpha\leq 1 \).
-  \begin{enumerate}[label=(\alph*)]
-  \item If $\alpha=1$ then
-    \( \E_{P_0}{\left[ \Vert \hat{\phi}_n(\data_n^{-k}) - F_0 \Vert_{P_0}^2
-      \right]} = \bigO(\log(n)n^{-1}) \).
-  \item If $\alpha<1$ then
-    \( \E_{P_0}{\left[ \Vert \hat{\phi}_n(\data_n^{-k}) - F_0 \Vert_{P_0}^2 \right]} =
-    \bigO(n^{-\alpha}) \).
-  \end{enumerate}
-\end{corollary}
-\begin{proof}
-  See Appendix~\ref{sec:oracle-inequalities}.
-\end{proof}
+provides estimates of these nuisance parameters. Though it
+served as our original motivation for developing the state
+learner, we leave the study of the state learner in the
+context of targeted and debiased machine learning for a future
+paper.
 
@@
 
-Each super learner provides a learner for the cumulative hazard
-function for the outcome of interest. From the cumulative hazard
-function a risk prediction model can be obtained (c.f.,
-equation~(\ref{eq:surv-def}) with $\Lambda_2 = 0$). We measure the
+Each super learner provides a learner for the cumulative
+hazard function for the outcome of interest. From the
+cumulative hazard function a risk prediction model can be
+obtained as described in Section~\ref{sec:surv-risk-pred}
+(with the special case of $\Lambda_2 = 0$). We measure the
 performance of each super learner by calculating the index of
-prediction accuracy (IPA) \citep{kattan2018index} at a fixed time
-horizon (36 months) for the risk prediction model provided by the
-super learner. The IPA is 1 minus the ratio between the model's Brier
-score and the null model's Brier score, where the null model is the
-model that does not use any covariate information. The IPA is
-approximated using a large (\( n = 20,000 \)) independent data set of
-uncensored data. As a benchmark we calculate the performance of the
-risk prediction model chosen by the oracle selector, which uses the
-large data set of uncensored event times to select the learner with
-the highest IPA.
+prediction accuracy (IPA) \citep{kattan2018index} at a fixed
+time horizon (36 months) for the risk prediction model
+provided by the super learner. The IPA is 1 minus the ratio
+between the model's Brier score and the null model's Brier
+score, where the null model is the model that does not use any
+covariate information. The IPA is approximated using a large
+(\( n = 20,000 \)) independent data set of uncensored data. As
+a benchmark we calculate the performance of the risk
+prediction model chosen by the oracle selector, which uses the
+large data set of uncensored event times to select the learner
+with the highest IPA.
 
@@
 In this section we use the prostate cancer data of
-\cite{kattan2000pretreatment} to illustrate the use of the state
-learner in the presence of competing risks. We have introduced the
-data in Section~\ref{sec:numer-exper}. The data consists of 1,042
-patients who are followed from start of followup until tumor
-recurrence, death without tumor recurrence or end of followup
-(censored) whatever came first. ... We use the state
-learner to rank libraries of learners for the cause-specific
-cumulative hazard functions of tumor recurrence, death without tumor
-recurrence, and censoring.  The libraries of learners each include
-five learners: the Nelson-Aalen estimator, three Cox regression models
-(unpenalized, Lasso, Elastic net) each including additive effects of
-the 5 covariates (Section~\ref{sec:numer-exper}), and a random
-survival forest. We use the same set of learners to learn the
-cumulative hazard function of tumor recurrence \( \Lambda_1 \), the
-cumulative hazard function of death without tumor recurrence
-\( \Lambda_2 \), and the cumulative hazard function of the conditional
-censoring distribution $\Gamma$.
+\cite{kattan2000pretreatment} to illustrate the use of the
+state learner in the presence of competing risks. We have
+introduced the data in Section~\ref{sec:numer-exper}. The data
+consists of 1,042 patients who are followed from start of
+followup until tumor recurrence, death without tumor
+recurrence or end of followup (censored) whatever came first.
+We use the state learner to rank libraries of learners for the
+cause-specific cumulative hazard functions of tumor
+recurrence, death without tumor recurrence, and censoring. The
+libraries of learners each include five learners: the
+Nelson-Aalen estimator, three Cox regression models
+(unpenalized, Lasso, Elastic net) each including additive
+effects of the 5 covariates (c.f., Section~\ref{sec:numer-exper}),
+and a random survival forest. We use the same set of learners
+to learn the cumulative hazard function of tumor recurrence
+\( \Lambda_1 \), the cumulative hazard function of death
+without tumor recurrence \( \Lambda_2 \), and the cumulative
+hazard function of the conditional censoring distribution
+$\Gamma$.
 
@@
 right-censored data and competing events. Compared to existing
-IPCW-based methods, the advantage of the state learner is that it does
-not depend on a pre-specified estimator of the censoring distribution,
-but selects one automatically based on a library of learners for the
-censoring distribution. Furthermore, the state learner neither
-requires that the cause-specific cumulative hazard functions
-\( \Lambda_j \) can be written as integrals with respect to Lebesgue
-measure, nor does it assume a (semi-)parametric formula. In the
-remainder of this section we discuss the limitations of our proposal
-and avenues for further research.
+IPCW-based methods, the advantage of the state learner is that
+it does not depend on a pre-specified estimator of the
+censoring distribution, but selects one automatically based on
+a library of candidate learners. Furthermore, the state
+learner neither requires that the cause-specific cumulative
+hazard functions \( \Lambda_j \) can be written as integrals
+with respect to Lebesgue measure, nor does it assume a
+(semi)parametric formula. In the remainder of this section
+we discuss the limitations of our proposal and avenues for
+further research.
 
@@
 Specifically, the finite sample oracle inequality in
-Corollary~\ref{cor:oracle-prop} concerns the function \( F \), which
+Proposition~\ref{prop:oracle-prop} concerns the function \( F \), which
 is a feature of \( P \in \mathcal{P} \), while what we are typically
@@
 Define
-\( \bar{B}_{\tau,0}(F, o) = \bar{B}_{\tau}(F, o) - \bar{B}_{\tau}(F_0, o) \) and
-\( R_{0}(F) = P_0{[\bar{B}_{\tau,0}(F, \blank)]} \), where the integrated Brier
-score \( \bar{B}_{\tau} \) was defined in
-Section~\ref{sec:super-learner-simple}.
+\( \bar{B}_{\tau,P}(F, o) = \bar{B}_{\tau}(F, o) -
+\bar{B}_{\tau}(F_P, o) \) and
+\( R_{P}(F) = P{[\bar{B}_{\tau,P}(F, \blank)]} \), where the
+integrated Brier score \( \bar{B}_{\tau} \) was defined in
+Section~\ref{sec:super-learner-simple}. Recall the norm
+\( \Vert \blank \Vert_{P}\) defined in
+equation~(\ref{eq:norm}).
+
 \begin{lemma}
   \label{lemma:norm}
-  \( R_{0}(F) = \Vert F - F_0 \Vert_{P_0}^2 \), where \( \Vert \blank \Vert_{P_0}\) is defined
-  in equation~(\ref{eq:norm}).
+  \( R_{P}(F) = \Vert F - F_P \Vert_{P}^2 \).
 \end{lemma}
@@
   \begin{align*}
-    & \E_{P_0}{\left[ (F(t, j, X) - \1{\{\eta(t) = j \}})^2 \right]}
-    \\
-    & =    \E_{P_0}{\left[ (F(t, j, X) - F_0(t, j, X) + F_0(t, j, X) - \1{\{\eta(t) = j
+    & \E_{P}{\left[ (F(t, j, X) - \1{\{\eta(t) = j \}})^2 \right]}
+    \\
+    & =    \E_{P}{\left[ (F(t, j, X) - F_P(t, j, X) + F_P(t, j, X) - \1{\{\eta(t) = j
       \}})^2 \right]}
     \\
-    & =    \E_{P_0}{\left[ (F(t, j, X) - F_0(t, j, X))^2\right]}
-      + \E_{P_0}{\left[ (F_0(t, j, X) - \1{\{\eta(t) = j \}})^2\right]}
+    & =    \E_{P}{\left[ (F(t, j, X) - F_P(t, j, X))^2\right]}
+      + \E_{P}{\left[ (F_P(t, j, X) - \1{\{\eta(t) = j \}})^2\right]}
     \\
     & \quad
-      + 2\E_{P_0}{\left[ (F(t, j, X) - F_0(t, j, X))(F_0(t, j, X) - \1{\{\eta(t) = j
+      + 2\E_{P}{\left[ (F(t, j, X) - F_P(t, j, X))(F_P(t, j, X) - \1{\{\eta(t) = j
       \}})\right]}
     \\
-    & =    \E_{P_0}{\left[ (F(t, j, X) - F_0(t, j, X))^2\right]}
-      + \E_{P_0}{\left[ (F_0(t, j, X) - \1{\{\eta(t) = j \}})^2\right]},
+    & =    \E_{P}{\left[ (F(t, j, X) - F_P(t, j, X))^2\right]}
+      + \E_{P}{\left[ (F_P(t, j, X) - \1{\{\eta(t) = j \}})^2\right]},
   \end{align*}
@@
     P{[\bar{B}_{\tau}(F, \blank)]}
-    = \Vert F - F_0 \Vert_{P_0}^2 + P_0{[\bar{B}_{\tau}(F_0, \blank)]}.
+    = \Vert F - F_P \Vert_{P}^2 + P{[\bar{B}_{\tau}(F_P, \blank)]}.
   \end{equation*}
@@
 
-\begin{proof}[Proof of Corollary~\ref{cor:oracle-prop}]
+\begin{proof}[Proof of Proposition~\ref{prop:oracle-prop}]
   First note that minimising the loss \( \bar{B}_{\tau} \) is equivalent to
-  minimising the loss \( \bar{B}_{\tau,0} \), so the discrete super learner and
-  oracle according to \( \bar{B}_{\tau} \) and \( \bar{B}_{\tau,0} \) are
-  identical. By Lemma~\ref{lemma:norm}, \( R_0(F) \geq 0 \) for any
+  minimising the loss \( \bar{B}_{\tau,P} \), so the discrete super learner and
+  oracle according to \( \bar{B}_{\tau} \) and \( \bar{B}_{\tau,P} \) are
+  identical. By Lemma~\ref{lemma:norm}, \( R_P(F) \geq 0 \) for any
   \( F \in \mathcal{H}_{\mathcal{P}} \), and so using Theorem 2.3 from
@@
 \begin{align*}
-  & \E_{P_0}{\left[ R_0(\hat{\phi}_n(\data_n^{-k})) \right]}
+  & \E_{P}{\left[ R_P(\hat{\phi}_n(\data_n^{-k})) \right]}
   \\
   &  \quad \leq
-    (1+2\delta)\E_{P_0}{\left[ R_0(\tilde{\phi}_n(\data_n^{-k})) \right]}
+    (1+2\delta)\E_{P}{\left[ R_P(\tilde{\phi}_n(\data_n^{-k})) \right]}
   \\
@@
     \left\{
-    M(F) + \frac{v(F)}{R_0(F)}
+    M(F) + \frac{v(F)}{R_P(F)}
     \left(
@@
 where for each \( F \in \mathcal{H}_{\mathcal{P}} \), \( (M(F), v(F)) \) is some Bernstein pair for
-the function \(o \mapsto \bar{B}_{\tau,0}(F, o) \). As
-\( \bar{B}_{\tau,0}(F, \blank) \) is uniformly bounded by \( \tau \) for any
+the function \(o \mapsto \bar{B}_{\tau,P}(F, o) \). As
+\( \bar{B}_{\tau,P}(F, \blank) \) is uniformly bounded by \( \tau \) for any
 \( F \in \mathcal{H}_{\mathcal{P}} \), it follows from section 8.1 in \citep{vaart2006oracle} that
-\( (\tau, 1.5 P_0{[\bar{B}_{\tau,0}(F, \blank)^2]}) \) is a Bernstein pair for
-\( \bar{B}_{\tau,0}(F, \blank) \). Now, for any \( a,b,c \in \R \) we have
+\( (\tau, 1.5 P{[\bar{B}_{\tau,P}(F, \blank)^2]}) \) is a Bernstein pair for
+\( \bar{B}_{\tau,P}(F, \blank) \). Now, for any \( a,b,c \in \R \) we have
 \begin{align*}
@@
 \end{align*}
-so using this with \( a=F(t, j, x) \), \( b=F_0(t, j, x) \), and
+so using this with \( a=F(t, j, x) \), \( b=F_P(t, j, x) \), and
 \( c = \1{\{\eta(t) = j\}} \), we have by Jensen's inequality
 \begin{align*}
-  & P_0{[\bar{B}_{\tau,0}(F, \blank)^2]}
+  & P{[\bar{B}_{\tau,P}(F, \blank)^2]}
   \\
   & \leq
-    2\tau\E_{P_0}{\left[
+    2\tau\E_{P}{\left[
     \sum_{j=-1}^{2} \int_0^{\tau}
@@
     \left(
-    F_0(t, j, X) - \1{\{\eta(t) = j\}}
+    F_P(t, j, X) - \1{\{\eta(t) = j\}}
     \right)^2
@@
   & =2\tau
-    \E_{P_0}\Bigg[
+    \E_{P}\Bigg[
     \sum_{j=-1}^{2} \int_0^{\tau}
     \left(
-    F(t, j, X) - F_0(t, j, X)
+    F(t, j, X) - F_P(t, j, X)
     \right)^2
@@
     \left\{
-    F(t, j, X) +  F_0(t, j, X)-2 \1{\{\eta(t) = j\}}
+    F(t, j, X) +  F_P(t, j, X)-2 \1{\{\eta(t) = j\}}
     \right\}^2
@@
   & \leq
-    8\tau \E_{P_0}{\left[
+    8\tau \E_{P}{\left[
     \sum_{j=-1}^{2} \int_0^{\tau}
     \left(
-    F(t, j, X) - F_0(t, j, X)
+    F(t, j, X) - F_P(t, j, X)
     \right)^2
@@
   & =
-    8\tau \Vert F - F_0 \Vert_{P_0}^2.
+    8\tau \Vert F - F_P \Vert_{P}^2.
 \end{align*}
-Thus when \( v(F) = 1.5 P_0{[\bar{B}_{\tau,0}(F, \blank)^2]} \) we have by
+Thus when \( v(F) = 1.5 P{[\bar{B}_{\tau,P}(F, \blank)^2]} \) we have by
 Lemma~\ref{lemma:norm}
 \begin{equation*}
-  \frac{v(F)}{R_0(F)}
-  = 1.5 \frac{P_0{[\bar{B}_{\tau,0}(F, \blank)^2]}}{P_0{[\bar{B}_{\tau,0}(F, \blank)]}}
+  \frac{v(F)}{R_P(F)}
+  = 1.5 \frac{P{[\bar{B}_{\tau,P}(F, \blank)^2]}}{P{[\bar{B}_{\tau,P}(F, \blank)]}}
   \leq 12 \tau,
 \end{equation*}
-and so using the Bernstein pairs \( (\tau, 1.5 P_0{[\bar{B}_{\tau,0}(F, \blank)^2]}) \) we have
+and so using the Bernstein pairs \( (\tau, 1.5 P{[\bar{B}_{\tau,P}(F, \blank)^2]}) \) we have
 \begin{equation*}
@@
   \left\{
-    M(F) + \frac{v(F)}{R_0(F)}
+    M(F) + \frac{v(F)}{R_P(F)}
     \left(
@@
     13 + \frac{12}{\delta}
-  \right),
+  \right).
 \end{equation*}
@@
 \begin{align*}
-  \E_{P_0}{\left[ R_0(\hat{\phi}_n(\data_n^{-k})) \right]}
+  \E_{P}{\left[ R_P(\hat{\phi}_n(\data_n^{-k})) \right]}
   \leq
-  &(1+2\delta)\E_{P_0}{\left[ R_0(\tilde{\phi}_n(\data_n^{-k})) \right]}
+  &(1+2\delta)\E_{P}{\left[ R_P(\tilde{\phi}_n(\data_n^{-k})) \right]}
   \\
@@
   \begin{equation*}
-    \E_{P_0}{\left[ \Vert \tilde{\phi}_n(\data_n^{-k}) - F_0 \Vert_{P_0}^2
-      \right]} \leq \E_{P_0}{\left[ \Vert \phi_n(\data_n^{-k}) - F_0 \Vert_{P_0}^2
+    \E_{P}{\left[ \Vert \tilde{\phi}_n(\data_n^{-k}) - F_P \Vert_{P}^2
+      \right]} \leq \E_{P}{\left[ \Vert \phi_n(\data_n^{-k}) - F_P \Vert_{P}^2
       \right]}  
@@
   for all \( n \in \N \). The results then follows from
-  Corollary~\ref{cor:oracle-prop}.
+  Proposition~\ref{prop:oracle-prop}.
 \end{proof}

### 2025-05-28 manuscript/manuscript-one.tex
typos

@@
 
-Accurate risk prediction from time-to-event data is a
+Accurate risk prediction from time-to-event data is an
 important problem in many areas of medical research and public
@@
 methods. Our approach also allows for full flexibility in the
-choice of models in the library, contrasting with earlier
+choice of models in the library, contrasting with other
 proposals that restrict the library to specific model classes
@@
 the Kaplan-Meier estimator, random survival forests, and
-semi-parametric Cox regression models and these learners
+semi-parametric Cox regression models, and these learners
 cannot be included in the library of the super learner
@@
 \cite{westling2021inference} circumvents the need to
-pre-specify a censoring model by iterating
-between estimation of the outcome and censoring models. However, it seems that this procedure is in general not
-guaranteed to converge to the true data-generating mechanism
+pre-specify a censoring model by iterating between estimation
+of the outcome and censoring models. However, this procedure
+is in general not guaranteed to converge to the true
+data-generating mechanism
 \citep[][Appendix~A.4]{munch2024thesis}.
@@
 we demonstrate how the state learner can be used to obtain
-risk predictions and how it can be applied in the context of
-causal inference. Section~\ref{sec:numer-exper} reports the
-results of numerical experiments, and
-Section~\ref{sec:real-data-appl} illustrates the method on
-prostate cancer data. We conclude with a discussion in
-Section~\ref{sec:discussion}. Proofs are collected in
-Appendix~\ref{sec:proofs}. Code and an implementation of the
-state learner are available at
+risk predictions, and we briefly mention how it can be applied
+in the context of causal inference.
+Section~\ref{sec:numer-exper} reports the results of numerical
+experiments, and Section~\ref{sec:real-data-appl} illustrates
+the method on prostate cancer data. We conclude with a
+discussion in Section~\ref{sec:discussion}. Proofs are
+collected in Appendix~\ref{sec:proofs}. Code and an
+implementation of the state learner are available at
 \url{https://github.com/amnudn/statelearner}.
@@
   \end{equation*}
-  \( P \)-almost surely.
+  for all \( j \in \{-1, 0, 1, 2 \} \), almost all
+  \( t \in [0,\tau] \), and \( P \)-almost all
+  \( x \in \mathcal{X} \).
 \end{proposition}
@@
 
-Proposition~\ref{prop:oracle-prop} provides a precise bound on
-the average price we pay for using cross-validation to select
-a learner from a library of candidates as compared to knowing
+Proposition~\ref{prop:oracle-prop} provides a bound on the
+average price we pay for using cross-validation to select a
+learner from a library of candidates as compared to knowing
 the oracle learner in advance. Corollary~\ref{cor:asymp-cons}
-states that this price is asymptotically vanishing as long as
-the number of learners in the library grows with sample size
-at a polynomial rate.
+states that this price is asymptotically vanishing, up to
+logarithmic terms, as long as the number of learners in the
+library grows with sample size at a polynomial rate.
 
@@
 
-\subsection{Targeted and debiased machine learning}
+\subsection{Targeted machine learning and causal inference}
 \label{sec:targeted-learning}
@@
 estimation high-dimensional nuisance parameters using machine
-learning, and then employing a debiasing or targeting step by
-using tools from semi-parametric efficiency theory
+learning, and then employing a debiasing or targeting step
+that is motivated from semi-parametric efficiency theory
 \citep{pfanzagl1985contributions,bickel1993efficient,van2003unified,tsiatis2007semiparametric,kennedy2022semiparametric}.
@@
 \citep[e.g.,][]{van2003unified,rytgaard2021estimation,rytgaard2022targeted}.
-The state learner is hence particularly well suited for
-targeted and debiased machine learning as it immediately
-provides estimates of these nuisance parameters. Though it
-served as our original motivation for developing the state
-learner, we leave the study of the state learner in the
-context of targeted and debiased machine learning for a future
-paper.
+The state learner immediately provides estimates of these
+nuisance parameters and is hence particularly well suited for
+targeted and debiased machine learning. Though it served as
+our original motivation for developing the state learner, we
+leave the study of the state learner in the context of
+targeted and debiased machine learning for a future paper.
 

### 2025-06-05 manuscript/manuscript-one.tex
tags suggestions and marginnotes

@@
 \setlength{\parindent}{0em}
-
+\usepackage{marginnote}
 % Handling space after sections
@@
 
-\begin{abstract}
-  In survival analysis, prediction models are needed as
-  stand-alone tools and in applications of causal inference to
-  estimate nuisance parameters. The super learner is a machine
-  learning algorithm which combines a library of prediction
-  models into a meta learner based on cross-validated loss. In
-  right-censored data, the choice of the loss function and the
-  estimation of the expected loss need careful consideration.
-  We introduce the state learner, a new super learner for
-  survival analysis, which simultaneously evaluates libraries
-  of prediction models for the event of interest and the
-  censoring distribution. The state learner places no
-  restrictions on the algorithms that can be included in the
-  library, works in the presence of competing risks, and does
-  not require a single pre-specified estimator of the
-  conditional censoring distribution. We establish an oracle
-  inequality for the state learner and investigate its
-  performance through numerical experiments. We illustrate the
-  application of the state learner with prostate cancer data.
+\begin{abstract} Risk prediction models are widely used to guide
+real-world decision-making in areas such as healthcare and economics,
+and they also play a key role in estimating nuisance parameters in
+semiparametric inference. The super learner is a machine learning
+framework that combines a library of prediction algorithms into a
+meta-learner using cross-validated loss. In the context of
+right-censored data, careful consideration must be given to both the
+choice of loss function and the estimation of expected loss. Moreover,
+estimators such as inverse probability of censoring weighting (IPCW)
+require accurate modeling and estimates of the censoring distribution.
+
+We propose a novel approach to super learning for survival analysis
+that jointly evaluates candidate learners for both the event time
+distribution and the censoring distribution. Our method imposes no
+restrictions on the algorithms included in the library, accommodates
+competing risks, and does not rely on a single pre-specified estimator
+of the censoring distribution. We establish an oracle inequality for
+our method, assess its performance through simulation studies, and
+demonstrate its practical utility using prostate cancer data.
 \end{abstract}
@@
 
-Accurate risk prediction from time-to-event data is an
-important problem in many areas of medical research and public
-health. Super learning \citep{van2007super}, also known as
-ensemble learning or stacked regression
-\citep{wolpert1992stacked,breiman1996stacked}, provides a
-powerful approach to this problem by combining multiple
-candidate prediction models. In the context of survival
-analysis, a super learner can for instance combine Cox
-regression models and random survival forests \citep[][Section
-8.4]{gerds2021medical}.
-
-In this paper we propose the state learner, a new super
-learner designed to handle the specific challenges of ensemble
-learning with right-censored data. The state learner estimates
-the performance of a library of candidate models for both the
-event-time and censoring distributions simultaneously. This
-joint estimation framework avoids the need for pre-specifying
-a censoring model, which is a limitation of many existing
-methods. Our approach also allows for full flexibility in the
-choice of models in the library, contrasting with other
-proposals that restrict the library to specific model classes
-\citep{polley2011-sl-cens,golmakani2020super}. Furthermore,
-unlike iterative approaches such as those in
-\cite{han2021inverse} and \cite{westling2021inference}, we can
-provide theoretical guarantees for the performance of the
-state learner.
-
-The state learner is based on the simple idea of considering
-right-censored competing risks data as an artificial
-multi-state system where censoring is included as a state of
-its own. Candidate models are then assessed based on how well
-they predict the state occupations of this system across time,
-based on baseline covariate information. We demonstrate how to construct
-a library for the state learner using existing survival
-models, and how to obtain risk predictions from the resulting
-ensemble. To analyse the theoretical properties of the state
-learner we focus on the discrete super learner which picks the
-model in the library with best estimated performance
-\citep{van2007super}. We show that the discrete state learner
-is consistent under natural conditions and prove a finite-sample
-oracle inequality.
-
-\subsection{Related work}
-\label{sec:related-work}
-
-Super learning requires a library of candidate models, or
-\textit{learners}, and a loss function for evaluating
-performance on hold-out samples. Machine learning based on
-right-censored data commonly uses the partial log-likelihood
-as a loss function
-\citep[e.g.,][]{li2016regularized,yao2017deep,lee2018deephit,katzman2018deepsurv,gensheimer2019scalable,lee2021boosted,kvamme2021continuous}.
-This loss function is also suggested for super learning with
-right-censored data by \cite{polley2011-sl-cens}, where data
-is assumed to be observed in discrete time. However, the
-partial log-likelihood loss does not work well as a measure of
-performance in hold-out samples observed in continuous time.
-The reason is that the partial log-likelihood loss assigns an
-infinite value for any learner that predicts piece-wise
-constant cumulative hazard functions when there are event
-times in the test set which do not occur in the learning set.
-This problem occurs with prominent survival learners including
-the Kaplan-Meier estimator, random survival forests, and
-semi-parametric Cox regression models, and these learners
-cannot be included in the library of the super learner
-proposed by \cite{polley2011-sl-cens}. When a proportional
-hazards model is assumed, the baseline hazard function can be
-profiled out of the likelihood \citep{cox1972regression}. The
-cross-validated partial log-likelihood loss
-\citep{verweij1993cross} has therefore been suggested as a
-loss function for super learning by \cite{golmakani2020super}.
-This choice of loss function restricts the library of learners
-to include only Cox proportional hazards models, and hence
-excludes many learners such as, e.g., random survival forests,
-additive hazards models, and accelerated failure time models.
-
-Alternative approaches for super learning with right-censored
-data use an inverse probability of censoring weighted (IPCW)
-loss function
-\citep{graf1999assessment,van2003unicv,molinaro2004tree,keles2004asymptotically,hothorn2006survival,gerds2006consistent,gonzalez2021stacked},
-censoring unbiased transformations
-\citep{fan1996local,steingrimsson2019censoring}, or
-pseudo-values
-\citep{andersen2003generalised,mogensen2013random,sachs2019ensemble}.
-All these methods rely on an estimator of the censoring
-distribution, and their drawback is that this estimator has to
-be pre-specified.
-
-More recent work by \cite{han2021inverse} and
-\cite{westling2021inference} circumvents the need to
-pre-specify a censoring model by iterating between estimation
-of the outcome and censoring models. However, this procedure
-is in general not guaranteed to converge to the true
-data-generating mechanism
-\citep[][Appendix~A.4]{munch2024thesis}.
+Accurately predicting risk from time-to-event data is a central
+challenge in diverse research fields including epidemiology, economics
+and weather forecasting with applications in clinical
+decision making and policy interventions. For instance, in prostate
+cancer management, clinicians often need to estimate a patient’s risk
+of disease progression and mortality over time to make informed
+decisions about treatment strategies such as active surveillance
+versus immediate intervention. Reliable time-to-event risk predictions
+can help tailor care to individual patients, avoid overtreatment, and
+allocate healthcare resources more effectively. Super learning
+\citep{van2007super}, also known as ensemble learning or stacked
+regression \citep{wolpert1992stacked,breiman1996stacked}, provides a
+powerful approach to this problem which by combining multiple
+candidate prediction models reduces the risk of bias incurred by a
+single potentially mispecified model. In survival analysis, a super
+learner may combine a stack of Cox regression models with a stack of
+random survival forests \citep[][Section 8.4]{gerds2021medical}. Such
+a strategy has recently produced KDpredict
+(\url{https://kdpredict.com/}) a model which jointly predicts the
+risks of kidney failure and all-cause mortality at multiple time
+horizons based on different sets of covariates
+\citep{liu2024predicting}. To evaluate the prediction performance of
+the learners the super learner behind KDpredict uses inverse
+probability of censoring weighting (IPCW) where the censoring
+distribution is estimated under the restrictive assumption that the
+censoring distribution does not depend on the covariates. This is a
+potential source of bias which is difficult to overcome with the
+currently available methods.
+
+In this paper we propose the state learner, a new super learner
+designed to handle the specific challenges of ensemble learning with
+right-censored data. The state learner simultaneously learns
+prediction models for the event-time and censoring distributions. The
+state learner is based on the simple idea of an artificial competing
+risk model where censoring is included as a state of its
+own. Candidate learners of the event time hazard and censoring hazard
+functions are then assessed based on how well they predict the state
+occupation probabilities of the artificial competing risk model across
+time based on baseline covariate information. Our estimation framework
+allows for competing risks, avoids restrictive assumptions on the
+censoring distribution, and is also fully flexible with respect to the
+choice of learners. The latter is in contrast to other proposals which
+restrict the library of learners to specific model classes
+\citep{polley2011-sl-cens,golmakani2020super}, see Section
+\ref{sec:choice-of-loss-function}. Recent work by
+\cite{han2021inverse} and \cite{westling2021inference} also
+circumvents the need to pre-specify a censoring model by iterating
+between estimation of the outcome and censoring models. However, this
+procedure is in general not guaranteed to converge to the true
+data-generating mechanism \citep[][Appendix~A.4]{munch2024thesis}. To
+analyse the theoretical properties of the state learner we focus on
+the discrete super learner which picks the model in the library with
+best estimated performance \citep{van2007super}. Here we provide
+theoretical guarantees for the performance of the state learner and in
+particular show that the discrete state learner is consistent under
+natural conditions and prove a finite-sample oracle inequality. We
+demonstrate how to construct a library for the state learner using
+common survival models, and how to obtain risk predictions from the
+resulting ensemble.\marginnote{is the term cross-fitting correct?}
+The state learner can be applied to produce a risk
+prediction model and can also be used for cross-fitting in semiparametric inference.
 
@@
 Section~\ref{sec:framework}. Section~\ref{sec:super-learning}
-defines general super learning for right-censored data. In
-Section~\ref{sec:super-learner-simple} we define the state
-learner, while Section~\ref{sec:theor-results-prop} provides
-theoretical guarantees. In Section~\ref{sec:use-cases-state}
-we demonstrate how the state learner can be used to obtain
-risk predictions, and we briefly mention how it can be applied
-in the context of causal inference.
-Section~\ref{sec:numer-exper} reports the results of numerical
-experiments, and Section~\ref{sec:real-data-appl} illustrates
-the method on prostate cancer data. We conclude with a
-discussion in Section~\ref{sec:discussion}. Proofs are
-collected in Appendix~\ref{sec:proofs}. Code and an
-implementation of the state learner are available at
-\url{https://github.com/amnudn/statelearner}.
-
-% A super learner is a machine learning algorithm that combines
-% a finite set of learners into a meta learner by estimating
-% prediction performance in hold-out samples using a
-% pre-specified loss function \citep{van2007super}. When the aim
-% is to make a prediction model, super learners combine strong
-% learners, such as Cox regression models and random survival
-% forests \citep[][Section 8.4]{gerds2021medical}. While the
-% general idea of combining strong learners based on
-% cross-validation stems from earlier work
-% \citep{wolpert1992stacked,breiman1996stacked}, the name super
-% learner is justified by an oracle inequality
-% \citep{van2003unicv,vaart2006oracle}.
-
-% We define the state learner, a new super learner for
-% right-censored data, which simultaneously estimates the
-% expected loss of learners of the event time distribution and
-% the censoring distribution. Compared to the super learners of
-% \cite{polley2011-sl-cens} and \cite{golmakani2020super}, the
-% state learner proposed here poses no restrictions on the type
-% of learners that can be included in the library. Compared to
-% super learners that rely on inverse probability of censoring
-% weighting, censoring unbiased transformations, or
-% pseudo-values, the state learner does not require a
-% pre-specified censoring distribution. Compared to the super
-% learners by \cite{han2021inverse} and
-% \cite{westling2021inference}, the state learner has
-% theoretical guarantees which these methods do not have.
-
-% The loss function underlying the state learner operates on
-% coarsened data. In the right-censored survival setting the
-% coarsened data consist of the minimum and the order of the
-% censoring time and the event time as well as the baseline
-% covariates. The state learner works by simultaneously
-% assessing how well learners of the event time distribution and
-% the censoring distribution predict the states of the coarsened
-% data. To analyse the theoretical properties of the state
-% learner we focus on the discrete super learner which combines
-% the library of learners by picking the one that minimises the
-% cross-validated loss \citep{van2007super}. In the presence of
-% competing risks, our algorithm uses separate libraries of
-% learners for the cumulative hazard functions for each of the
-% competing risks and for the censoring distribution. We show
-% that the oracle selector of the state learner is consistent if
-% all libraries contain a consistent learner and prove a finite
-% sample oracle inequality.
-
-% The state learner algorithm can output a medical risk
-% prediction model \citep{gerds2021medical} which predicts the
-% probability of an event based on covariates in the presence of
-% competing risks. Another important application, which
-% motivated our contribution, is in targeted learning where
-% conditional event probabilities occur as high-dimensional
-% nuisance parameters which need to be estimated at a certain
-% rate \citep{van2011targeted, rytgaard2021estimation,
-%   rytgaard2022targeted}. The state learner is particularly
-% well-suited for this application as it simultaneously provides
-% estimates for both the outcome and the censoring, both of
-% which are needed for targeted learning. For the asymptotic
-% bias term of a targeted estimator, which uses the state
-% learner to estimate nuisance parameters, we show that a second
-% order product structure holds. We illustrate both applications
-% of the state learner with prostate cancer data
-% \citep{kattan2000pretreatment}.
+introduces loss-based super learning. In
+Section~\ref{sec:super-learner-simple} we define the state learner,
+while Section~\ref{sec:theor-results-prop} provides theoretical
+guarantees. In Section~\ref{sec:use-cases-state} we demonstrate how
+the state learner can be used to obtain risk predictions, and we
+briefly mention how it can be applied in the context of causal
+inference.  Section~\ref{sec:numer-exper} reports the results of
+numerical experiments, and Section~\ref{sec:real-data-appl}
+illustrates the method on prostate cancer data. We conclude with a
+discussion in Section~\ref{sec:discussion}. Proofs are collected in
+Appendix~\ref{sec:proofs}. Code and an implementation of the state
+learner are available at \url{https://github.com/amnudn/statelearner}.
 
@@
 
-In a competing risk framework \citep{andersen2012statistical},
-let \( T\) be a time to event variable, \(D\in\{1,2\}\) the
-cause of the event, and $X \in \mathcal{X}$ a vector of
-baseline covariates taking values in a bounded subset
-\( \mathcal{X} \subset \R^p \), \( p\in\N \). Let
-$\tau< \infty$ be the prediction horizon. We use
-\( \mathcal{Q} \) to denote the collection of all probability
-measures on \( [0,\tau] \times \{1,2\}\times \mathcal{X} \)
-such that \( (T, D, X) \sim Q \) for some unknown
-\( Q \in \mathcal{Q} \). For \(j\in\{1,2\}\), the
-cause-specific conditional cumulative hazard functions
-\( \Lambda_{j} \colon [0, \tau] \times \mathcal{X} \rightarrow
-\R_+ \) are defined as
+In a competing risk framework \citep{andersen2012statistical}, let \(
+T\) be a time to event variable, \(D\in\{1,2\}\) the cause of the
+event, and $X \in \mathcal{X}$ a vector of baseline covariates taking
+values in a bounded subset \( \mathcal{X} \subset \R^p \), \( p\in\N
+\). Let $\tau< \infty$ be a fixed prediction horizon. We use \(
+\mathcal{Q} \) to denote the collection of all probability measures on
+\( [0,\tau] \times \{1,2\}\times \mathcal{X} \) such that \( (T, D, X)
+\sim Q \) for some unknown \( Q \in \mathcal{Q} \). For
+\(j\in\{1,2\}\), the cause-specific conditional cumulative hazard
+functions \( \Lambda_{j} \colon [0, \tau] \times \mathcal{X}
+\rightarrow \R_+ \) are defined as
 \begin{equation*}
@@
   \Lambda_{j}(t \mid x) = \int_0^t\frac{  Q(T \in \diff s, D=j \mid X=x )}{Q(T \geq s \mid X=x )}.
-\end{equation*}
-For ease of presentation we assume throughout that the map
-\( t\mapsto \Lambda_j(t \mid x) \) is continuous for all
-\( x \) and \( j \). All arguments carry over directly to the
-general case. We denote by \(S\) the conditional event-free
-survival function,
+\end{equation*} For ease of presentation we assume throughout that the
+map \( t\mapsto \Lambda_j(t \mid x) \) is continuous for all \( x \)
+and \( j \), however, 
+all technical arguments extend naturally to the general case \citep{andersen2012statistical}.
+The event-free survival function conditional on covariates is given by
 \begin{equation}
@@
 
-We consider the right-censored setting in which we observe
-\(O = (\tilde{T},\tilde D, X)\), where $\tilde T = \min(T,C)$
-for a right-censoring time \(C\), $\Delta = \1{\{T \leq C\}}$,
-and \(\tilde D=\Delta D\). Let \(\mathcal{P}\) denote a set of
-probability measures on the sample space
-\(\mathcal{O} = [0, \tau] \times \{0, 1, 2\} \times
-\mathcal{X}\) such that \(O \sim P \) for some unknown
-\(P\in \mathcal{P}\). We assume that the event times and the
-censoring times are conditionally independent given
-covariates, \( T \independent C \mid X \). This implies that
-any distribution \( P \in \mathcal{P} \) is characterised by a
-distribution \( Q \in \mathcal{Q} \) and a conditional
-cumulative hazard function for \( C \) given \( X \)
-\citep[c.f.,][]{begun1983information,gill1997coarsening}. We
-use \(\Gamma\in\mathcal{M}_{\tau}\) to denote the conditional
-cumulative hazard function for censoring. For ease of
-presentation we now also assume that
-\( \Gamma(\blank \mid x) \) is continuous for all \( x \). We
-let
-\((t,x)\mapsto G(t \mid x)=\exp\left\{-\Gamma(t \mid
-  x)\right\}\) denote the survival function of the conditional
-censoring distribution. The distribution \( P \) is
-characterised by
+We consider the right-censored setting in which we observe \(O =
+(\tilde{T},\tilde D, X)\), where $\tilde T = \min(T,C)$ for a
+right-censoring time \(C\), $\Delta = \1{\{T \leq C\}}$, and \(\tilde
+D=\Delta D\). Let \(\mathcal{P}\) denote a set of probability measures
+on the sample space \(\mathcal{O} = [0, \tau] \times \{0, 1, 2\}
+\times \mathcal{X}\) such that \(O \sim P \) for some unknown \(P\in
+\mathcal{P}\). We assume that the event times and the censoring times
+are conditionally independent given covariates, \( T \independent C
+\mid X \). This implies that any distribution \( P \in \mathcal{P} \)
+is characterised by a distribution \( Q \in \mathcal{Q} \) and a
+conditional cumulative hazard function for \( C \) given \( X \)
+\citep[c.f.,][]{begun1983information,gill1997coarsening}. We use
+\(\Gamma\in\mathcal{M}_{\tau}\) to denote the cumulative hazard
+function of the conditional censoring distribution given
+covariates. For ease of presentation we assume that \(t\mapsto
+\Gamma(t \mid x) \) is continuous for all \( x \). We let
+\((t,x)\mapsto G(t \mid x)=\exp\left\{-\Gamma(t \mid x)\right\}\)
+denote the survival function of the conditional censoring
+distribution. The distribution \( P \) is characterised by
 \begin{equation}\label{eq:parametrizeP}
@@
 P(\tilde T>t \mid X=x) = S(t \mid x)G(t \mid x) = \exp\left\{-\Lambda_{1}(t \mid x)-\Lambda_{2}(t \mid x)-\Gamma(t \mid x) \right\}.
-\end{equation*}
-We further assume that there exists \(\kappa<\infty\) such that
-\(\Lambda_{j}(\tau- \mid x)<\kappa \), for \(j\in\{1,2\}\), and
-\(\Gamma(\tau- \mid x)<\kappa\) for almost all \(x\in\mathcal X\). Note that this
-implies that \(G(\tau- \mid x)\) is bounded away from zero for almost all \(x\in\mathcal X\).
-Under these assumptions, the conditional cumulative hazard functions
-\(\Lambda_{j}\) and \(\Gamma\) can be identified from \(P\) by
+\end{equation*} We assume that there exists \(\kappa<\infty\) such
+that \(\Lambda_{j}(\tau- \mid x)<\kappa \), for \(j\in\{1,2\}\), and
+\(\Gamma(\tau- \mid x)<\kappa\) for almost all \(x\in\mathcal
+X\). Note that this implies that \(G(\tau- \mid x)\) is bounded away
+from zero for almost all \(x\in\mathcal X\).  Under these assumptions,
+the conditional cumulative hazard functions \(\Lambda_{j}\) and
+\(\Gamma\) can be identified from \(P\) by
 \begin{align}
@@
 
-\section{Super learning}
+\section{Loss-based super learning}
 \label{sec:super-learning}
 
-In survival analysis, a super learner can be used to estimate
-a parameter $\theta \in \Theta$ which can be identified from
-the observed data distribution \(P\in\mathcal P\). A super
-learner typically estimates a function, and thus the parameter
-space $\Theta$ is most often a class of functions. For
-example, $\theta$ could be a conditional survival function or
-a cumulative hazard function. In
-Section~\ref{sec:super-learner-simple} we consider a specific
-choice of parameter as target for our super learner, but in
-this section we describe super learning for a general
-parameter $\theta$ with values in some parameter space
-$\Theta$.
-
-As input to a super learner we need a data set
-\( \data_n=\{O_i\}_{i=1}^n \) of i.i.d.\ observations from
-\( P \in \mathcal{P} \) and a collection of candidate models
-or learners $\mathcal{A}$. Each learner \(a \in \mathcal{A}\)
-is a map \( a \colon \mathcal{O}^n \rightarrow \Theta \) which
-takes a data set as input and returns an estimate
-$a(\data_n) \in \Theta$ of $\theta$.
-% Formally, the domain of $a$ depends on \(n\) but we suppress this in the notation.
-In what follows, we use the short-hand notation
-\(P[f] = \int f(o) P(\diff o) \). A super learner evaluates the
-performance of \(a \in \mathcal{A}\) with a loss function
-\(L\colon \Theta \times \mathcal{O} \rightarrow \R_+\) and
-estimates the expected loss \(P[L(a(\data_n), \blank)]\) using
-cross-validation. Specifically, the expected loss of $a\in\mathcal A$
-is estimated by splitting the data set $\data_n$ into $K$ disjoint
-approximately equally sized subsets
+\marginnote{formally super learning may work without loss function}
+Loss-based super learning requires a library of candidate models, or
+\textit{learners}, a cross-validation algorithm, and a loss function
+for evaluating prediction performance on hold-out samples. In this
+section we provide an abstract definition of loss-based super
+learning.  As input to a super learner we need a data set \(
+\data_n=\{O_i\}_{i=1}^n \) of i.i.d.\ observations from \( P \in
+\mathcal{P} \) and a collection of candidate models or learners
+$\mathcal{A}$. Let \(\Theta\) be a suitable parameter space such as
+\([0,1]\) for risk predictions, or the space of cadlag functions on
+\([0,\tau]\) for predictions of survival functions.  Each learner \(a
+\in \mathcal{A}\) is a map \( a \colon \mathcal{O}^n \rightarrow
+\Theta \) which takes a data set as input and returns an estimate
+$a(\data_n)$ of $\theta$.  \marginnote{we do not use Pf throughout but
+also use integral and E notation!? would be good to reduce the amount
+of abstract notation and settle for one way} In what follows, we use
+the short-hand notation \(P[f] = \int f(o) P(\diff o) \).  Let
+\(L\colon \Theta \times \mathcal{O}^n \rightarrow \R_+\) be a loss
+function and denote the expected loss by \(P[L(a(\data_n), \blank)]\)
+where the expectation is taken across all possible data sets of size
+\(n\).
+
+The expected loss of a learner is estimated by splitting the data set
+$\data_n$ into $K$ disjoint approximately equally sized subsets
 \(\data_n^1, \data_n^2, \dots, \data_n^K \) and then calculating the
@@
 and that the oracle learner is a function of the data and the unknown
-data-generating distribution. However, these dependencies are
-suppressed in the notation.
+data-generating distribution.
+% However, these dependencies are suppressed in the notation.
 
@@
 
-The main idea of the state learner is to jointly use learners
-for \( \Lambda_1 \), \( \Lambda_2 \), and \( \Gamma \), and
-the relations in equation~(\ref{eq:parametrizeP}), to learn a
-feature of the observed data distribution \( P \). The
-discrete state learner ranks a tuple of learners for the tuple
-of the cumulative hazard functions
-\( (\Lambda_1, \Lambda_2, \Gamma) \) based on how well they
-jointly model the observed data. To formally introduce the
-state learner, we define the artificial multi-state process
+The main idea of the state learner is to jointly use learners for \(
+\Lambda_1 \), \( \Lambda_2 \), and \( \Gamma \), and the relations in
+equation~(\ref{eq:parametrizeP}), to learn a feature of the observed
+data distribution \( P \). The discrete state learner ranks a tuple of
+learners for the tuple of the cumulative hazard functions \(
+(\Lambda_1, \Lambda_2, \Gamma) \) based on how well they jointly model
+the observed data. To formally introduce the state learner, we define
+the process 
 \begin{equation*}
@@
   D=2\} - \1\{\tilde{T} \leq t, \tilde D=0\},
-  \quad \text{for} \quad t \in [0, \tau].
-\end{equation*}
-We use the phrase `artificial' because multi-state models
-should not usually include censoring as a state.
-At time \(t\), we observe that each individual is in one of four mutually
-exclusive states: \( 0 \), \( 1 \), \( 2 \), or \( -1 \).
-The conditional distribution of the process \( \eta(t) \) given
-baseline covariates \( X \) is determined by the function
+  \quad \text{for} \quad t \in [0, \tau]
+\end{equation*} which takes values in \( \{-1,0,1,2\}\) which
+represent four mutually exclusive states. Specifically value \( 0 \)
+represents the state where the individual is still event-free and
+uncensored, value \( 1\) the state where the event of interest has
+occured, value \( 2\) the state where a competing risk has occured and
+value \( -1\) the state where the observation is right censored.  The
+state occupation probabilities given baseline covariates \( X \) are
+given by the function\marginnote{cannot use k because k is used in
+previous section}
 \begin{equation}
@@
 for all \( t \in [0,\tau] \), \( k \in \{-1,0,1,2\} \), and
-\( x \in \mathcal{X} \). The function \( F \) describes the
-conditional state occupation probabilities of the process
-\(\eta\).
-
-We define the state learner as a super learner for the
-function-valued parameter $\theta(P) = F$ which is identified
-through equation~(\ref{eq:F-def}). Under conditional
-independent censoring each tuple
-$(\Lambda_{1}, \Lambda_{2}, \Gamma, H)$ characterises a
+\( x \in \mathcal{X} \).
+
+The state learner is a super learner for the function-valued parameter
+$\theta(P) = F$ which is identified through
+equation~(\ref{eq:F-def}). Under conditional independent censoring
+each tuple $(\Lambda_{1}, \Lambda_{2}, \Gamma, H)$ characterises a
 distribution \(P\in\mathcal P\), c.f.\
-equation~\eqref{eq:parametrizeP}, which in turn determines
-\( (F, H) \). Hence, a learner for \( F \) can be constructed
-from learners for \( \Lambda_1 \), \( \Lambda_2 \), and
-$\Gamma$ as follows:
+equation~\eqref{eq:parametrizeP}, which in turn determines \( (F, H)
+\). Hence, a learner for \( F \) can be constructed from learners for
+\( \Lambda_1 \), \( \Lambda_2 \), and $\Gamma$ as follows:
 \begin{equation}\label{eq:transition}
@@
   &= \int_0^t \phi_{a_1,a_2, b}(\data_n)(s-,0,x)  b(\data_n)(\diff s \mid x).
-\end{align*}
-Importantly, the libraries \( \mathcal{A}_1 \),
-\( \mathcal{A}_2 \), and \( \mathcal{B} \) can be constructed
-using standard survival analysis software.
-
-To evaluate how well a function \( F \) predicts an observed
-process $\eta$ we use the integrated Brier score
+\end{align*} Notably, the libraries \( \mathcal{A}_1 \), \(
+\mathcal{A}_2 \), and \( \mathcal{B} \) can be constructed using
+common methods for survival analysis.\marginnote{Elaborate on this last statement with a concrete example?}
+
+To evaluate how well a function \( F \) predicts the
+process $\eta$ we use the integrated Brier score \citep{graf1999assessment}
 \( \bar B_\tau( F,O) = \int_0^{\tau} B_t(F,O) \diff t \),
@@
 \end{equation*}
+\marginnote{Say that Brier score is across all 4 states because this is not commonly done}
 Based on a split of a data set \(\data_n\) into $K$ disjoint
@@
 
-\section{Theoretical results for the state learner}
+\section{Theoretical guarantees}
 \label{sec:theor-results-prop}
 
-In this section we establish theoretical guarantees for the
-state learner. We show that the state learner is consistent if
-its library contains a consistent learner, and we establish a
-finite sample inequality for the excess risk of the state
-learner compared to the oracle.
+\marginnote{depending on the journal both subsections may benefit from explanations about the usefulness of these results} 
+In this section we show that the state learner is consistent if its
+library contains a consistent learner, and we establish a finite
+sample inequality for the excess risk of the state learner compared to
+the oracle.
 
@@
 
-\subsection{Oracle inequalities}
+\subsection{Oracle inequality}
 \label{sec:finite-sample-oracle}
 
-We establish a finite sample oracle result for the state
-learner. Our Proposition~\ref{prop:oracle-prop} is in essence a
-special case of Theorem 2.3 in \citep{vaart2006oracle}. We
-assume that we split the data into equally sized folds, and
-for simplicity of presentation we take \( n \) to be such that
-\( |\data_n^{-k}| = n/K \) with \( K \) fixed. We will allow
-the number of learners to grow with \( n \) and write
-\( \mathcal{F}_n=\mathcal{F}(\mathcal{A}_{1,n},
-\mathcal{A}_{2,n}, \mathcal{B}_n)\) as short-hand notation and
-to emphasise the dependence on \( n \). In the following we
-let the space \( \mathcal{H}_{\mathcal{P}} \) be equipped with
-the norm \( \| \blank \|_{P} \) defined as
+To establish a finite sample oracle result for the state learner we
+assume a split of the data into equally sized folds, and for
+simplicity of presentation we take \( n \) to be such that \(
+|\data_n^{-k}| = n/K \) with \( K \) fixed. We will allow the number
+of learners to grow with \( n \) and write \(
+\mathcal{F}_n=\mathcal{F}(\mathcal{A}_{1,n}, \mathcal{A}_{2,n},
+\mathcal{B}_n)\) as short-hand notation emphasing the
+dependence on the sample size. In the following we let the space \(
+\mathcal{H}_{\mathcal{P}}= \{F_P : P \in \mathcal{P}\} \) be equipped with the norm \( \| \blank
+\|_{P} \) defined by
 \begin{equation}
@@
 \end{proof}
-
-Proposition~\ref{prop:oracle-prop} has the following asymptotic consequences.
+\marginnote{if it is a special case, why is a proof needed?}
+Proposition~\ref{prop:oracle-prop} is in essence a
+special case of Theorem 2.3 in \citep{vaart2006oracle}. It 
+has the following asymptotic consequences.
 
@@
 \label{sec:use-cases-state}
-
-The state learner estimates the parameter \( F \) which is a
-feature of the observed right-censored data distribution
-\( P \). In particular, \( F \) depends on the censoring
-distribution and is typically not of direct interest in
-itself. In the following, we describe how an estimate of
-\( F \), as provided by the state learner, can be used to
-estimate more relevant parameters of interest, such as, e.g.,
-survival probabilities.
+\marginnote{both subsections would benefit from examples. could we merge with section 8?} 
+The state learner estimates the function \( F \) which depends on the
+censoring distribution and is typically not of direct interest in
+itself. In the following, we describe how an estimate of \( F \), as
+provided by the state learner, can be used to predict risks over time.
 
@@
 
+\subsection{Choice of loss function}
+\label{sec:choice-of-loss-function}
+
+Machine learning based on right-censored data
+commonly uses the partial log-likelihood as a loss function
+\citep[e.g.,][]{li2016regularized,yao2017deep,lee2018deephit,katzman2018deepsurv,gensheimer2019scalable,lee2021boosted,kvamme2021continuous}.
+This loss function is also suggested for super learning with
+right-censored data by \cite{polley2011-sl-cens}, where data is
+assumed to be observed in discrete time. However, the partial
+log-likelihood loss does not work well as a measure of performance in
+hold-out samples observed in continuous time.  The reason is that the
+partial log-likelihood loss assigns an infinite value for any learner
+that predicts piece-wise constant cumulative hazard functions when
+there are event times in the test set which do not occur in the
+learning set.  This problem occurs with prominent survival learners
+including the Kaplan-Meier estimator, random survival forests, and
+semi-parametric Cox regression models, and these learners cannot be
+included in the library of the super learner proposed by
+\cite{polley2011-sl-cens}. When a proportional hazards model is
+assumed, the baseline hazard function can be profiled out of the
+likelihood \citep{cox1972regression}. The cross-validated partial
+log-likelihood loss \citep{verweij1993cross} has therefore been
+suggested as a loss function for super learning by
+\cite{golmakani2020super}.  This choice of loss function restricts the
+library of learners to include only Cox proportional hazards models,
+and hence excludes many learners such as, e.g., random survival
+forests, additive hazards models, and accelerated failure time models.
+
+Alternative approaches for super learning with right-censored
+data use an inverse probability of censoring weighted (IPCW)
+loss function
+\citep{graf1999assessment,van2003unicv,molinaro2004tree,keles2004asymptotically,hothorn2006survival,gerds2006consistent,gonzalez2021stacked},
+censoring unbiased transformations
+\citep{fan1996local,steingrimsson2019censoring}, or
+pseudo-values
+\citep{andersen2003generalised,mogensen2013random,sachs2019ensemble}.
+All these methods rely on an estimator of the censoring
+distribution, and their drawback is that this estimator has to
+be pre-specified.
+
 \section*{Conflict of interest}

### 2025-06-06 manuscript/manuscript-one.tex
incorporate tag changes

@@
 regression \citep{wolpert1992stacked,breiman1996stacked}, provides a
-powerful approach to this problem which by combining multiple
-candidate prediction models reduces the risk of bias incurred by a
+powerful approach to this problem by combining multiple
+candidate prediction models to reduce the risk of bias incurred by a
 single potentially mispecified model. In survival analysis, a super
@@
 \citep{polley2011-sl-cens,golmakani2020super}, see Section
-\ref{sec:choice-of-loss-function}. Recent work by
+\ref{sec:choice-of-loss-function}. Like our proposal, recent work by
 \cite{han2021inverse} and \cite{westling2021inference} also
 circumvents the need to pre-specify a censoring model by iterating
-between estimation of the outcome and censoring models. However, this
+between estimation of the outcome and censoring models.
+However, this iterative
 procedure is in general not guaranteed to converge to the true
-data-generating mechanism \citep[][Appendix~A.4]{munch2024thesis}. To
-analyse the theoretical properties of the state learner we focus on
-the discrete super learner which picks the model in the library with
-best estimated performance \citep{van2007super}. Here we provide
-theoretical guarantees for the performance of the state learner and in
-particular show that the discrete state learner is consistent under
-natural conditions and prove a finite-sample oracle inequality. We
-demonstrate how to construct a library for the state learner using
-common survival models, and how to obtain risk predictions from the
-resulting ensemble.\marginnote{is the term cross-fitting correct?}
-The state learner can be applied to produce a risk
-prediction model and can also be used for cross-fitting in semiparametric inference.
-
-\subsection{Outline of the paper}
-\label{sec:outline-paper}
-
+data-generating mechanism
+\citep[][Appendix~A.4]{munch2024thesis}.
+
+To analyse the theoretical properties of the state learner we
+focus on the discrete super learner which picks the model in
+the library with best estimated performance
+\citep{van2007super}. We provide theoretical guarantees
+for the performance of the state learner and in particular
+show that the discrete state learner is consistent under
+natural conditions and prove a finite-sample oracle
+inequality. We demonstrate how to construct a library for the
+state learner using common survival models, and how to obtain
+risk predictions from the resulting ensemble.
+
+The rest of the paper is organized as follows.
 We introduce our notation and framework in
@@
   D=2\} - \1\{\tilde{T} \leq t, \tilde D=0\},
-  \quad \text{for} \quad t \in [0, \tau]
-\end{equation*} which takes values in \( \{-1,0,1,2\}\) which
-represent four mutually exclusive states. Specifically value \( 0 \)
-represents the state where the individual is still event-free and
-uncensored, value \( 1\) the state where the event of interest has
-occured, value \( 2\) the state where a competing risk has occured and
-value \( -1\) the state where the observation is right censored.  The
-state occupation probabilities given baseline covariates \( X \) are
-given by the function\marginnote{cannot use k because k is used in
-previous section}
+  \quad \text{for} \quad t \in [0, \tau],
+\end{equation*}
+which takes values in \( \{-1,0,1,2\}\). The four values
+represent four mutually exclusive states. Specifically, value
+\( 0 \) represents the state where the individual is still
+event-free and uncensored, value \( 1\) the state where the
+event of interest has occurred, value \( 2\) the state where a
+competing risk has occurred, and value \( -1\) the state where
+the observation is right-censored. The state occupation
+probabilities given baseline covariates \( X \) are given by
+the function
 \begin{equation}
   \label{eq:F-def}
-  F(t, k, x) = P(\eta(t) = k \mid X=x),
+  F(t, l, x) = P(\eta(t) = l \mid X=x),
 \end{equation}
-for all \( t \in [0,\tau] \), \( k \in \{-1,0,1,2\} \), and
+for all \( t \in [0,\tau] \), \( l \in \{-1,0,1,2\} \), and
 \( x \in \mathcal{X} \).
@@
 \mathcal{A}_2 \), and \( \mathcal{B} \) can be constructed using
-common methods for survival analysis.\marginnote{Elaborate on this last statement with a concrete example?}
+standard software for survival analysis. In
+\texttt{R}, for instance, we can construct Cox models as
+learners using the \texttt{survival}-package
+\citep{survival-package}, and we can construct random survival
+forests as learners using the \texttt{randomForestSRC}-package \citep{randomForestSRC}.
 
@@
 \begin{equation*}
-  B_t(F,O) = \sum_{j=-1}^{2}
+  B_t(F,O) = \sum_{l=-1}^{2}
   \left(
-      F(t,j,X) - \1{\{\eta(t)=j\}}
+      F(t,l,X) - \1{\{\eta(t)=l\}}
   \right)^2.
@@
   \end{equation*}
-  for all \( j \in \{-1, 0, 1, 2 \} \), almost all
+  for all \( l \in \{-1, 0, 1, 2 \} \), almost all
   \( t \in [0,\tau] \), and \( P \)-almost all
@@
   \left\{
-    \sum_{j=-1}^{2}\int_{\mathcal{X}} \int_0^{\tau} F(t, j, x)^2 \diff t P( \diff x)
+    \sum_{l=-1}^{2}\int_{\mathcal{X}} \int_0^{\tau} F(t, l, x)^2 \diff t P( \diff x)
   \right\}^{1/2}.
@@
 \end{proof}
-\marginnote{if it is a special case, why is a proof needed?}
-Proposition~\ref{prop:oracle-prop} is in essence a
-special case of Theorem 2.3 in \citep{vaart2006oracle}. It 
-has the following asymptotic consequences.
+
+Proposition~\ref{prop:oracle-prop} has the following
+asymptotic consequences.
 
@@
 and a random survival forest
-\citep{ishwaran2008random,rfsrc-paclage}. We use the same
+\citep{ishwaran2008random,randomForestSRC}. We use the same
 library to learn the outcome distribution and the censoring
@@
 \begin{proof}[Proof]
-  For any \( t \in [0, \tau] \) and \( j\in \{-1,0,1,2\} \) we have
+  For any \( t \in [0, \tau] \) and \( l\in \{-1,0,1,2\} \) we have
   \begin{align*}
-    & \E_{P}{\left[ (F(t, j, X) - \1{\{\eta(t) = j \}})^2 \right]}
-    \\
-    & =    \E_{P}{\left[ (F(t, j, X) - F_P(t, j, X) + F_P(t, j, X) - \1{\{\eta(t) = j
+    & \E_{P}{\left[ (F(t, l, X) - \1{\{\eta(t) = l \}})^2 \right]}
+    \\
+    & =    \E_{P}{\left[ (F(t, l, X) - F_P(t, l, X) + F_P(t, l, X) - \1{\{\eta(t) = l
       \}})^2 \right]}
     \\
-    & =    \E_{P}{\left[ (F(t, j, X) - F_P(t, j, X))^2\right]}
-      + \E_{P}{\left[ (F_P(t, j, X) - \1{\{\eta(t) = j \}})^2\right]}
+    & =    \E_{P}{\left[ (F(t, l, X) - F_P(t, l, X))^2\right]}
+      + \E_{P}{\left[ (F_P(t, l, X) - \1{\{\eta(t) = l \}})^2\right]}
     \\
     & \quad
-      + 2\E_{P}{\left[ (F(t, j, X) - F_P(t, j, X))(F_P(t, j, X) - \1{\{\eta(t) = j
+      + 2\E_{P}{\left[ (F(t, l, X) - F_P(t, l, X))(F_P(t, l, X) - \1{\{\eta(t) = l
       \}})\right]}
     \\
-    & =    \E_{P}{\left[ (F(t, j, X) - F_P(t, j, X))^2\right]}
-      + \E_{P}{\left[ (F_P(t, j, X) - \1{\{\eta(t) = j \}})^2\right]},
+    & =    \E_{P}{\left[ (F(t, l, X) - F_P(t, l, X))^2\right]}
+      + \E_{P}{\left[ (F_P(t, l, X) - \1{\{\eta(t) = l \}})^2\right]},
   \end{align*}
@@
 \end{align*}
-so using this with \( a=F(t, j, x) \), \( b=F_P(t, j, x) \), and
-\( c = \1{\{\eta(t) = j\}} \), we have by Jensen's inequality
+so using this with \( a=F(t, l, x) \), \( b=F_P(t, l, x) \), and
+\( c = \1{\{\eta(t) = l\}} \), we have by Jensen's inequality
 \begin{align*}
@@
     2\tau\E_{P}{\left[
-    \sum_{j=-1}^{2} \int_0^{\tau}
+    \sum_{l=-1}^{2} \int_0^{\tau}
     \left\{
     \left(
-    F(t, j, X) - \1{\{\eta(t) = j\}}
+    F(t, l, X) - \1{\{\eta(t) = l\}}
     \right)^2
@@
     \left(
-    F_P(t, j, X) - \1{\{\eta(t) = j\}}
+    F_P(t, l, X) - \1{\{\eta(t) = l\}}
     \right)^2
@@
     \E_{P}\Bigg[
-    \sum_{j=-1}^{2} \int_0^{\tau}
+    \sum_{l=-1}^{2} \int_0^{\tau}
     \left(
-    F(t, j, X) - F_P(t, j, X)
+    F(t, l, X) - F_P(t, l, X)
     \right)^2
@@
     \left\{
-    F(t, j, X) +  F_P(t, j, X)-2 \1{\{\eta(t) = j\}}
+    F(t, l, X) +  F_P(t, l, X)-2 \1{\{\eta(t) = l\}}
     \right\}^2
@@
     8\tau \E_{P}{\left[
-    \sum_{j=-1}^{2} \int_0^{\tau}
+    \sum_{l=-1}^{2} \int_0^{\tau}
     \left(
-    F(t, j, X) - F_P(t, j, X)
+    F(t, l, X) - F_P(t, l, X)
     \right)^2
